### 2015-05-27 3dpas_intro.tex
initial files

% cpedoc.tex V2.0, 13 May 2010

\documentclass[times]{cpeauth}

\usepackage{moreverb}

\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}

\newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08em
T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\def\volumeyear{2012} %\documentclass[acmjacm,acmnow]{acmtrans2m}
%\documentclass[conference,final]{IEEEtran}
%\documentclass[10pt,letterpaper]{article}
%\documentclass[preprint,12pt]{article}
%\documentclass[11pt,letterpaper]{article}

\usepackage{graphicx} \usepackage{url} \usepackage{color} \usepackage{enumerate}
\usepackage{longtable} %\usepackage{authblk} % Is used internally within the CPE
Class \usepackage[firstpage]{draftwatermark}

%\SetWatermarkLightness{0.7} %\SetWatermarkScale{4}

%\usepackage{pdfsync} % \acmVolume{} % \acmNumber{} % \acmYear{} % \acmMonth{}

%\usepackage{sectsty} %\usepackage{setspace}

% \usepackage{ifpdf} % \usepackage{wrapfig} % %\usepackage[cm]{fullpage} %
\usepackage{textcomp} % \usepackage{srcltx} % \usepackage{fancyhdr} %
\usepackage{setspace} \usepackage{lscape} % \usepackage{longtable} %
\usepackage{paralist}

%\usepackage{hyperref} \usepackage[novbox]{pdfsync}

% \usepackage{ams}

% \ifpdf % \usepackage[pdftex, % colorlinks=true, % linkcolor=red, %
citecolor=red, % filecolor=red, % ]{hyperref} % \else %
\usepackage[hypertex]{hyperref} % \fi

% Space saving %\usepackage[small,compact]{titlesec}
%\usepackage[small,it]{caption}

% \renewcommand\floatpagefraction{.9} % \renewcommand\topfraction{.9} %
\renewcommand\bottomfraction{.9} %\renewcommand\textfrare{.1}

\setcounter{totalnumber}{50} \setcounter{topnumber}{50}
\setcounter{bottomnumber}{50}

%\textwidth = 6in %\textheight = 8.5 in %\oddsidemargin = 0.0 in
%\evensidemargin = 0.0 in

% \topmargin = 0.0 in % \headheight = 0.0 in % \headsep = 0.0 in
%\parskip = 0.0in
%\parindent = 0.5cm
\parskip = 0.04in
% \parindent = 0.0cm % \textfloatsep = 0.1in

% \usepackage[draft,pdftex,colorlinks=true, linkcolor=blue,citecolor=blue, %
urlcolor=blue]{hyperref}

% \ifpdf % \pdfinfo{ /Author (Shantenu Jha et al.)  /Title (Programming
Abstractions for Large-Scale Distributed Applications) } % \fi

%\newcommand{\projectnamefull}{\textit{Programming Abstractions for %
Large-Scale Distributed Applications} }

\newcommand{\upup}{\vspace*{-0.5em}} \newcommand{\up}{\vspace*{-0.25em}}
\newcommand{\I}[1]{\textit{#1}} \newcommand{\B}[1]{\textbf{#1}}
\newcommand{\T}[1]{\texttt{#1}} \newcommand{\BI}[1]{\B{\I{#1}}}

 \newenvironment{shortlist}{ \vspace*{-0.8em}
  \begin{itemize} \setlength{\itemsep}{-0.3em} }{
  \end{itemize} \vspace*{-0.6em} }

 \newenvironment{shortenum}{ \vspace*{-0.8em}
  \begin{enumerate} \setlength{\itemsep}{-0.3em} }{
  \end{enumerate} \vspace*{-0.6em} }

% \ifpdf % \DeclareGraphicsExtensions{.pdf, .png, .jpg} % \else %
\DeclareGraphicsExtensions{.ps, .eps} % \fi

\definecolor{orange}{rgb}{1.0,0.3,0.0} \definecolor{violet}{rgb}{0.75,0,1}
\definecolor{darkgreen}{rgb}{0,0.6,0} \definecolor{cyan}{rgb}{0.2,0.7,0.7}
\definecolor{blueish}{rgb}{0.2,0.2,0.8}

\long\def\comment#1{{\bf \textcolor{magenta}{\bf #1}}} \long\def\ccomment#1{{\bf
\textcolor{blue}{\bf #1}}} \newcommand{\C}{\comment} \newcommand{\CC}{\ccomment}

\newcommand{\yes}{$\bullet$}

\long\def\finalcomment#1{{\bf \textcolor{red}{\bf #1}}}
\newcommand{\CF}{\finalcomment}

\newif\ifdraft \drafttrue \ifdraft \newcommand{\katznote}[1]{
{\textcolor{magenta} { ***Dan: #1 }}} \newcommand{\rananote}[1]{
{\textcolor{cyan} { ***Omer: #1 }}} \newcommand{\omernote}[1]{ {\textcolor{cyan}
{ ***Omer: #1 }}} \newcommand{\note}[1]{ {\textcolor{red} { #1 }}}
\newcommand{\jhanote}[1]{ {\textcolor{blue} { ***Shantenu: #1 }}}
\newcommand{\jonnote}[1]{ {\textcolor{red} { ***Jon: #1 }}}
\newcommand{\alnote}[1]{ {\textcolor{darkgreen} { ***Andre: #1 }}}
\newcommand{\nchnote}[1]{ {\textcolor{orange} { ***Neil: #1 }}}
\newcommand{\ysnote}[1]{ {\textcolor{violet} { ***Yogesh: #1 }}}
\newcommand{\sdnote}[1]{ {\textcolor{blueish} { ***Simon: #1 }}} %
\usepackage[pdftex,colorlinks=true, linkcolor=blue,citecolor=blue, %
urlcolor=blue]{hyperref} \else \newcommand{\katznote}[1]{}
\newcommand{\rananote}[1]{} \newcommand{\note}[1]{} \newcommand{\jhanote}[1]{}
\newcommand{\jonnote}[1]{} \newcommand{\alnote}[1]{} \newcommand{\nchnote}[1]{}
\newcommand{\ysnote}[1]{} \fi

\begin{document} \runningheads{S.~Jha et al.}{Understanding Applications and
Infrastructure for D3 Science}

%\title{3DPAS: Distributed Dynamic Data-intensive Programming % Abstractions and
Systems for Large-Scale Science and Engineering}

\title{Introducing Distributed Dynamic Data-intensive (D3) Science:
Understanding Applications and Infrastructure}

% An Applications-Driven Analysis}

%\title{Towards Distributed Dynamic Data-intensive (D3) Science: A % Survey and
Analysis of D3 Science Applications and % Cyberinfrastructure}

\author{Shantenu Jha\affil{1,2}\corrauth\, Neil Chue Hong\affil{2}, Simon
Dobson\affil{3}, Daniel S. Katz\affil{4,5}, Andre~Luckow\affil{1}, Omer
Rana\affil{6}, Yogesh Simmhan\affil{7}}

\address{\affilnum{1}Rutgers University\break \affilnum{2}University of
Edinburgh\break \affilnum{3}University of St.~Andrews\break
\affilnum{4}University of Chicago\break \affilnum{5}Argonne National
Laboratory\break \affilnum{6}Cardiff University\break \affilnum{7}Indian
Institute of Science\break }

%\affilnum{9}University of Minnesota\break

\corraddr{Rutgers University, NJ, USA. E-mail: shantenu.jha@rutgers.edu}

% \author{Shantenu Jha} % \address{Rutgers University and University of
Edinburgh; \texttt{shantenu.jha@rutgers.edu}} % \author{Neil Chue Hong} %
\address{University of Edinburgh; \texttt{N.ChueHong@epcc.ed.ac.uk}} %
\author{Simon Dobson} % \address{University of St.~Andrews;
\texttt{simon.dobson@st-andrews.ac.uk}} % \author{Daniel S. Katz} %
\address{University of Chicago and Argonne National Laboratory;
\texttt{d.katz@ieee.org}} % \author{Andre Luckow} % \address{Louisiana State
University; \texttt{luckow@cs.uni-potsdam.de}} % \author{Omer Rana} %
\address{Cardiff University; \texttt{O.F.Rana@cs.cardiff.ac.uk}} %
\author{Yogesh Simmhan} % \address{University of Southern California;
\texttt{simmhan@usc.edu}} % \author{Jon Weissman} % \address{University of
Minnesota; \texttt{jon@cs.umn.edu}}

\keywords{dynamic, distributed, data intensive, scientific applications}

\begin{abstract} % Many problems at the forefront of science, engineering,
medicine, % and the social sciences are increasingly complex and %
interdisciplinary due to the plethora of data sources and % computational
methods available today.  A common feature across many % of these problems is
the amount and diversity of data and % computation that must be integrated to
yield insights. Such data is % increasingly large-scale and distributed: the
datasets, their % locations, and associated scheduling decisions are %
time-dependent.

% Applications need access % to these disparate data-sets and to the methods
that operate on % them.  Data-intensive applications need first-class support
for % distributed heterogeneous platform execution.

The 3DPAS Research Theme (funded by the eSI, Edinburgh and US NSF) investigated
applications that have these characteristics; in other words, it operated at the
triple point of {\it distributed} and {\it dynamic} and {\it data-intensive}
(D3) attributes.  This document represents the conclusions of the 3DPAS Theme.
\end{abstract}
\maketitle

\ifdraft
\newpage
\tableofcontents
%\newpage %\listoftables
\newpage \fi

%\input{notes}

\section{Introduction: Context, Scope and Outline}

\jhanote{NCH points out that we are inconsistent in the order of the ``D''s. SJ
to fix globally. Good catch Neil!}

The landscape of scientific and enterprise computing is being fundamentally
altered by the current surge in data. The resulting upheaval, grandiosely named
BigData science by the Lilliputians in the data science community, represents a
new reality where the volumes of data produced by scientific applications are
increasing rapidly, driven by advancing technologies (e.g., increasing compute
capacity and higher resolution sensors) and decreasing costs for data
acquisition and storage~\cite{tony-propaganda}. With the increase in data
volumes, types, and sources of data, an important concomitant development has
been an increase in the requirement for and complexity in the processing of
data. Data are often naturally distributed and have to be processed in situ, and
occasionally data that are localized have to be distributed in order to be
processed, either for reasons of performance or for collaborative analytics.
Furthermore when processing large volumes of data, data source/destination, data
production and transfer rates, and data processing capacity can all vary over
time. Thus properties such as resource scheduling, data placement, and transfer
decisions, that were statically determinable at small scales, emerge as dynamic
and important considerations at large scales of data and distribution.  The
BigData science problem is therefore only partly about large volumes of data.

Our focus is on a subspace of the BigData problem space, namely that which is
characterized by the distributed and dynamic aspects of data-intensive
science. We introduce the concept of Dynamic Distributed Data-intensive science
and refer to this subspace of BigData science as D3 science.  In order to
appreciate the D3 subspace we will briefly discuss the importance and role of
distribution and the pervasiveness of dynamical aspects.

% \begin{itemize}
% \item Centralized: data is at one location
% \item Decentralized: data is stored at multiple locations
% \end{itemize}

% \jhanote{I intend to convert these bullet points into a paragraph % about the
importance of treating distributed data}

Although the capacity to store data centrally is growing faster than Moore's
Law, it is not sufficient in all cases to be able to ``dump the data'' in one
place (or in the cloud.)  Oftentimes the costs (data transfer, storage,
consequences of catastrophic failure) prevent the centralization of data, and
decentralization---storing or place data in multiple locations---become prudent.

Conversely, not all production of data is centralized; in some cases there are
multiple producers (and consumers) of data, not to mention one-of-a-kind data
that must be shared across multiple consumers.  Furthermore in collaborative
science, data output from one analytical engine are often the input for the next
stage of the analytical pipeline.  Last but not least, distribution via
replication is a tested and effective approach to reliability and
fault-tolerance; we will argue that it is also potentially a great performance
enhancer and a means to ensure data consistency too.

Whether data are a priori distributed or localized, a critical challenge is to
find an efficient and effective means to either distribute the initially
localized data, or manage the distribution (or localization) of a priori
distributed data.

% \begin{itemize}
% \item Costs (data transfer, storage costs) for centralized data too % high
% \item data locally required for analysis (store data where it is used)
% \item volume too large to store centrally or move around
% \item required for scaling the system (load balancing, local reads are %
cheaper)
% \item reliability and fault tolerance (no single point of failure, %
redundancy)
% \item data consistency
% \item data security
% \end{itemize}

% \ysnote{distributed to ensure security also? Privacy % controls on moving data
out of a location or country.}

The challenges of distributed data-intensive applications span beyond the
management and curation of large volumes of data. A critical requirement is to
make sense of this data, i.e., transform the data into information en route to
assembling knowledge. Among other things, the transformation from data to
information requires extensive amounts of computational processing of the data;
thus the data-intensive nature of the BigData science problem is strongly
coupled with compute-intensive requirements.  As the data are often distributed,
a consequence is that many attributes of distributed systems permeate to
data. Thus almost naturally, distributed data have attributes of dynamism, both
at the data level as well as the infrastructure level.

That is not to say all data-intensive applications have elements of both data
and compute dynamism and are always distributed, but it is a reflection that at
extreme scales, dynamism and and distribution become increasingly important and
often correlated issues.  Thus, analyzing the applications along an analytical
space spanned by the three ``D''s of data-intensive, distributed, and dynamic
provides a general and effective framework.  The diverse set of requirements to
support D3 scenarios points to the need for a careful understanding of D3
characteristics as well as sophisticated, extensible, and skillfully architected
infrastructure and programming systems.

With recent attention on ``Big Data'' problems, several classification schemes
have been proposed; one that has garnered attention is the
3Vs~\cite{gartner3v}. These three dimensions (Volume, Velocity and Variety) just
describe the characteristics of the data. %, rather than describing the
interplay between Big %Data and applications that operate over them.
\jhanote{Does someone have the authentic reference for the 4Vs?} \katznote{see
http://www.villanovau.com/university-online-programs/what-is-big-data/ but it's
not `authoritative'} In contrast, D3 emphasizes the nature of scientific
applications as characterized by their data attributes. Furthermore the D3
attributes are considered as a collective property, as opposed to different
disjoint attributes.  Thus we believe there is a fundamental distinction between
the D3 and 3Vs classification schemes.

%\ysnote{Since we're clearly positioning around BigData, % one question that is
going to come up is how are the 3D's related to % the 3V's -- volume, velocity
and variety -- of Big Data. We may want % to explicitly address that early
on. For e.g. ``While the three % dimensions of Volume, Velocity and Variety of %
BigData~\cite{gartner3v}, or 3V's, are well recognized in industry, % these
narrowly consider the characteristics of the data rather than % the interplay
between Big Data and applications that operate over % them as we do through the
3D's. Furthermore, our emphasis is also on % the role of Big Data in science,
which is less studied (and arguably % more diverse and complex) than business
applications. That said, the % volume and velocity dimensions of the 3V's
partially capture the % data-intensive and dynamic aspects of our study.''}
\jhanote{we do not discuss explicitly the 4th D, i.e. diversity (courtesy AL) it
is assumed but not discussed}

%\subsubsection*{Terminology}

%\alnote{Should we introduce a degree of distribution? machine, % cluster,
data-center, WAN, Wireless} \jhanote{A figure akin to our % the figure in our
pilot-data paper maybe Ok to help establish the % vocabulary. Not sure this
would be a conflict or self-plagiarism?  % Either way, this should be in the
sub-section on ``Scope''}

Distribution can occur on many levels.  %, as shown in
%\figurename~\ref{fig:figures_terminology}.  In this paper, we use {\em locally
distributed} to indicate that data/compute resides on multiple nodes within one
data center.  {\em Geographically distributed} describes the distribution of
data and/or compute across multiple data centers. Further, we define the degree
as the number of data sources and sites that are used for storing and processing
the data. A distributed application is an application that is composed of
entities (i.e., either storage/compute resource or a data source) from at least
two physical locations. In this work, our focus is almost exclusively on
geographically distributed data and applications.

%\jhanote{terminology defined elsewhere in the paper isn't defined % here.. so a
bit misleading to call this part terminology. AL points % out ``dynamic'' as an
example.}

%\katznote{AL thinking about changes to the figure, according to SJ - % now SJ
to remove this figure completely.}

%\katznote{where are the data sources %in this figure?}\alnote{added}

%\begin{figure}[htbp] % \centering
% \includegraphics[width=0.7\textwidth]{figures/terminology.pdf}
% \caption{Levels of Distribution}
% \label{fig:figures_terminology}
%\end{figure}

% (\S\ref{sec:prog_infra}). \katznote{removed this ref, as it seemed % out of
place to have a ref to a subsection in the intro, as the only such ref}

% \subsection*{Distributed Data} % One specific attribute that we believe is
pervasive but not % necessarily obvious is ``dynamic data.''  Additionally the
requirement % of many applications impose a requirement that data is dynamic.

% \jhanote{I do not know why floating sensors was a subsection? As an % examplar
for D3 Science? If so, made it a subsubsection..}

% \katznote{as I wrote in the title, it got moved from apps and won't be there %
any more, but there was some thought that it might be useful as an example %
application here.  Feel free to delete it if you don't think it adds anything
here.}

% \subsubsection{Floating Sensors -- material than can be used in 1 as % an
example application that is not going to be discussed in %
\S\ref{sec:scenarios}}

% This {\em traditional application} constructs real-time hydrodynamics models,
and forecasts the behavior of a complicated water system: the Sacramento % - San
Joaquin River delta.  It begins after data generation and includes data
transformation \& decisions, storage, and processing.

% \subsubsection*{Application description}

% The application integrates both static and dynamic sensors streaming %
real-time data with a high-resolution hydrodynamics model running on % the NERSC
Magellan cloud testbed. The model uses sensor data to % initialize and shape the
run, and the model output is in turn used to % drive the real-time positioning
of the dynamic sensors. The output of % this process is a map of the current and
future behavior of the % modeled watershed.

% Custom built dynamic sensors are deployed in the watershed along with %
traditional statically placed sensors that provide data. These sensors %
communicate over a variety of different network infrastructures, e.g., % GSM,
CDMA, 802.11, wired ethernet, etc.  and the application acquires % the data over
the internet from the sensor platforms.  The application % processes the sensor
data in real-time as it is streamed.

% \subsubsection*{Big data aspects}

% Not known.

% \subsubsection*{Distributed aspects}

% Not known.

% \subsubsection*{Dynamic aspects}

% The sensor are all streaming real-time data to the application. The position
of the dynamic sensors % themselves change over time.

% \subsubsection*{Other important issues}

% The first full, end-to-end, field test of the system is scheduled for % May
2011.

%\jhanote{If some one knows a bit more about Bayesian %
approaches/inference/statistics and its 'implication' on dynamic % data via
dymamic models, i'd love to hear / learn more}

\jhanote{refine next two paragraph...} Analogous to the distinction between
programming in the large versus system level programming, there is a distinction
between the ``cyberinfrastructure/systems considerations/design at the
large-scale'' versus ``CI considerations at the small-scale''. This paper is
primarily about CI/systems considerations at the large-scale.  Searching for
common solutions and patterns at the large-scale is likely to be more fruitful
than at small-scales where customized tools and optimization of low-level
systems features is likely to return greater yield.

Thus, subjecting a range of otherwise distinct applications to a common
analytical framework provides advantages to the infrastructure scientists and
cyberinfrastructure experts, namely suggestions for where effort is both needed
and likely to yield maximum returns. In other words, as D3 applications evolve,
our analysis can provide guidance on where to focus effort on infrastructure.
Last but not least, the advantages of providing a common but extensible
vocabulary that can be used by the entire community, application end-users,
developers and infrastructure scientists alike, in an emerging field cannot be
overstated. One may argue that such vocabulary and terminology has been missing
from the practice of distributed cyberinfrastructure in general, with less than
desirable consequences.

%\katznote{some of this is repeating ideas from the para ``An important %
motivation...''  Perhaps these two paras should be % combined?}\jhanote{DSK,
this should not be an issue now, given that % this has now moved upstream}

\subsection{Scope}

% Before a discussion of the applications, infrastructure and research %
methodology, a brief mention of the motivation and thereby scope of % this work
is in order.

% Without a doubt, the era of {\it BigData} is here. However, BigData % is not
just an issue of large volumes of data, it is also about the

%One motivation of this work is to understand the complexity and
%characteristics of data that emerge as a consequence of scale, namely %the
dynamic properties and the distribution of applications, and the %complex
interplay between them that arises.

One motivation of this work is to understand the dynamic properties and the
distribution of applications, and the complex interplay between them that arises
as a consequence of scale.  Additionally, we are motivated by an attempt to
understand infrastructure---existing capabilities, trends, and
limitations---available to the complexity, challenges and characteristics of
dynamic and distributed data at scale.

We would like the reader to keep in mind three primary context and scope-setting
points:

%\katznote{a lot of this paragraph seems too repetitive of the previous %
section} \jhanote{Dan, I fixed the previous paragraph. OK?}

\begin{itemize}

\item This {\it survey} is not meant to be complete, either in scope of domains
covered or types of infrastructure and application surveyed.  They by definition
represent the interests and experience of the authors.

\item The focus is on applications and infrastructure that arise from science
and engineering projects in academia; this is not exclusive but is more a
guiding principle.  While we are cognizant of enterprise applications and
infrastructure, and recognize the impressive advances made therein, we also
acknowledge that the factors such as scale, distribution, open-source, and
interoperability influence the solutions employed and infrastructures used; it
is important to therefore focus on problems that emerge as in the space defined
by these ``academic constraints''.

\item Almost by definition, any attempt to capture the state-of-the-art of such
a dynamic and emerging domain is bound to be fraught with limitations and issues
of scoping.  A complete and rigorous survey of the landscape of BigData is bound
to be obsolete by the time it emerges; so given the rate of change, we asked
what could be done. What we present is a partial analysis motivated by a set of
existing and active D3 applications, their trials and tribulations, successes
and solutions. We believe there is merit in capturing the state-of-play as is,
deriving a research agenda and set of specific yet crosscutting issues, and
presenting these to the community.
\end{itemize}

%\jhanote{I think these two points can be ignored. To be discussed. But % either
way this should go in the scope subsection} % \katznote{Dan and Shantenu request
that Neil and Omer think % about the next two points, and either flesh them out
with text % or remove them.}
%
%\begin{itemize}
%\item How does this differ from Scientific Databases? How does it % relate to
Scientific Data Management and Scientific data flow
%\item Reference to EU 2030 Objectives (part of the Project Europe 2030 %
Report) -- underpins ``Data Intensive Research'' from an EU % perspective
(strategy) + relate this to NSF ``CIF21'' vision
%\end{itemize}

%\jhanote{some points to ensure we've addressed. I cut and paste some % comments
we received from the DPA survey paper that we should be % ensure at addressed --
implicitly or explicitly}
%
%\begin{verbatim}
%
%(i) It is difficult to argue about the completeness of the presented %material.
%
%(ii) Picking a handful of distributed applications to study does not form a
%basis of a taxonomy.
%
%(iii) Furthermore, it is unclear if the applications %are indeed representative
of any urgent need to execute across %distributed systems as the authors claim.
%
%(iv) At the same time, there is not enough description of the computer %science
behind the applications.
%
%(v) Several applications seem similar to each other and, %as far as I can tell,
do not add anything new.
%
%(v) The section is more a %laundry list of applications - I would have expected
to see a %synthesis of different application and, for example, a classification
%into different classes. No general requirements of applications are %derived. A
smaller number of more representative sample applications %(not covering nine
pages) would have been better.
%
%(vi) My main problem with the paper is the research methodology.  The paper
%uses 6 hand-picked applications to derive a vector of attributes for %each of
them, derive some patterns from observing these vectors, and %uses these
patterns to perform a distributed computing tools gap %analysis.
%
%(vii) It is unclear how important the applications used in this %study are.
%
%(viii) Your lack of depth in explaining the applications leaves the %reader
empathetic.  It's important that the reader feels that they can %learn something
from reading this paper.  How is what you are saying %applicable to me (a
computer scientist or computational domain scientist)?
%
%(ix) Have you considered adopting a different approach, like conducting a
%survey to glean the likes/dislikes of distributed application %developers?  In
essence focus on the development process, as opposed %to applications.  With a
large enough pool of responses, you may come %up with statistically significant
conclusions that will be of interest %to the wider community.
%
%\end{verbatim}

\subsection{Methodology}

This paper originated in work at the UK e-Science Institute (eSI) at the
University of Edinburgh, in a research theme examining Dynamic Distributed
Data-intensive Programming Abstractions and Systems, called
3DPAS~\cite{3dpas-theme}.  The description of this theme's work was:

\begin{quote} Many problems at the forefront of science, engineering, medicine,
and the social sciences, are increasingly complex and interdisciplinary due to
the plethora of data sources and computational methods available today. A common
feature across many of these problem domains is the amount and diversity of data
and computation that must be integrated to yield insights.

For many complex data-intensive applications, moving the data may have
restrictions. Increasingly important type of data-intensive applications are
data-driven applications. For example, data is increasingly large-scale,
distributed arising from sensors, scientific instruments \& simulations. Such
data-driven applications will involve computational activities triggered as a
consequence of independent data creation; thus it is imperative for an
application to be able to respond to unplanned changes in data load or
content. Understanding how to support dynamic computations is a fundamental, but
currently a critical missing element in data-intensive computing.

The 3DPAS theme seeks to understand the landscape of dynamic, distributed,
data-intensive computing: the programming models and abstractions, the run-time
and middleware services, and the computational infrastructure. It will analyze
existing tools and services, identify missing pieces and new abstractions, and
propose practical solutions and best practices.
\end{quote}

The theme held three workshops, and in each, application scientists and
computing technologists were invited to give presentations on their work, and
then to be involved in discussions to organize and understand the presented
materials.  Between the workshops, the theme organizers also drafted outlines of
a report that was a predecessor of this paper, for discussion in the next
workshop.  Overall, there were about 50 workshop attendees.

In order to select applications and gather information about them, during the
various workshop discussions, when an application that ``felt'' new was
mentioned, we asked the person who mentioned it to provide a written set of
answers to the following questions about the application:

\begin{enumerate}
\item What is the purpose of the application?
\item How is the application used to do this?
\item What infrastructure is used? (including compute, data, network,
instruments, etc.)
\item What dynamic data is used in the application?
\begin{enumerate}
\item What are the types of data,
\item What is the size of the data set(s)?
\end{enumerate}
\item How does the application get the data?
\item What are the time (or quality) constraints on the application?
\item How much diverse data integration is involved?
\item How diverse is the data?
\item Please feel free to also talk about the current state of the application,
if it exists today, and any specific gaps that you know need to be overcome.
\end{enumerate}

We then used the text of these answers to describe each of the applications (in
\S\ref{sec:scenarios}) in terms of:

\begin{itemize}
\item What does the application do? What problem does it solve?
\item How does it do it? How does the application run?  What infrastructure does
it need/use?
\item What are the big data aspects of the application?
\item What are the distributed aspects of the application?
\item What are the dynamic aspects of the application?
\item What else is important about the application?  What does it do well?
Poorly? How was the app written? What tools does it use?
\end{itemize}

After the workshops, the theme organizers and a set of attendees who were
interested in participating started discussing how to organize and analyze the
applications and the technologies, including infrastructure, programming systems
and abstractions.  These participants are the current authors of this paper, and
this paper is the result of that process.

As part of the theme, two 3DPAS workshops~\cite{3dapas,D3-escience} were
organized at major conferences: HPDC 2011 and IEEE eScience 2011.  These
workshops had both peer-reviewed and invited papers.

A combination of invitation-only focused discussion meetings and broader
community engagement events were used to ensure that the applications surveyed
and ideas captured in this paper are representative of current community
thinking as well as of intellectual relevance.

\subsection{Outline of this paper}

The rest of this paper consists of the application scenarios
(\S\ref{sec:scenarios}), a discussion of the aspects of distribution and
dynamism in the applications (\S\ref{sec:distdyndata}), a description of the
software infrastructure required for D3 applications % Yogesh: This section's
description seemed more verbose than others, hence commenting out details % ,
such as % software frameworks, tools, and middleware and platform services
(\S\ref{sec:infrastructure}), analysis and characterization of the applications
and conclusions (\S\ref{sec:conclusions}).

\section{D3 Application Scenarios \label{sec:scenarios}}

Starting with the experience of the attendees of the 3DPAS workshops, we have
assembled a set of D3 ``applications'' that we describe in this section.  We
realize that we have overloaded the term application here.  In this document,
some of our applications are really {\em traditional applications}, meaning a
program that is developed by a user or project to try to answer a science
question, usually with some input data needed and some output data produced.
Applications in this category are: NGS Analytics (\S\ref{bioSilvia}), CMB
(\S\ref{astroJulian}), Fusion (\S\ref{fusionScott}), Industrial Incident
Notification and Response (\S\ref{Kees}), MODIS Data Processing
(\S\ref{modisKeith}), and Distributed Network Intrusion Detection
(\S\ref{detectionJon}).

Another group of applications are {\em infrastructural applications}.
Applications of this type are collaborative, in that they depend on a context
that is often agreed upon by a community, most often in terms of how data is
stored by the community and how it is accessed, as well as, more often than not,
a weak consensus on the infrastructure to be used. %\jhanote{Dan: Please confirm
you are OK with this modified % defintion of infrastructural application} This
allows different sub-applications (that we again refer to as `applications') to
focus on different stages, such as generating data, or processing data.
Answering a science question now may involve a set of applications that need to
be run in series, perhaps in different stages, that may be run by different
groups that do not frequently interact.  This type of application includes:
ATLAS/WLCG (\S\ref{WLCGSteve}), LSST (\S\ref{astroAdam}), SOA Astronomy
(\S\ref{astroSOAAdam}), Sensor Network Application (\S\ref{sensorSimon}),
Climate (\S\ref{climateDan}), Interactive Exploration of Environmental Data
(\S\ref{envJon}), and Power Grids (\S\ref{power}).

In other words, multiple infrastructural applications are used together in
particular science domains.  Furthermore, infrastructural applications involve
more than just the end user (scientist), and typically require exerting control
at layers below, i.e., ``programming the system (infrastructure)'', as opposed
to a traditional applications, where end user effort and control is often
confined to a well-defined application kernel.

% \jhanote{Need better distinction between infrastructure and % traditional
applications. They are the same at the scale-up and % scale-out level, but
different in scale-across and people. Another % issue is that about ``control
and configure'', i.e traditional % applications only exert control and configure
the application layer, % whilst infrastructure applications need to configure
the lower % ie. infrastructural layer too}

% \katznote{I don't agree in general that more is needed, but if you % think so,
please give it a try.  I also don't agree specifically % that infrastructure
applications `configure' the lower layer.  % Rather, I would say they assume a
particular, community-defined % (perhaps de facto) layer.  Finally, while I
understand scale-up % and scale-out, and I don't know what `scale-across' is.}

Some of the applications we describe are really groups of applications, meaning
that we are grouping together a number of independent programs, written by
different authors, that are trying to answer the same type of science question
and thus, where a single user will only run one of them at a time.
%\katznote{Shantenu, please check %the preceding and see if this is ok, in the
context of the whole %paragraph and section.}  These codes and their developers
may either be competing or collaborating, and sometimes both, informally known
as collabetition. Specifically, this is the case for NGS Analytics
(\S\ref{bioSilvia}), SOA Astronomy (\S\ref{astroSOAAdam}), and Interactive
Exploration of Environmental Data (\S\ref{envJon}).  (Note that this issue is
orthogonal to traditional vs. infrastructural applications: NGS Analytics is
traditional, where SOA Astronomy and Interactive Exploration of Environmental
Data are infrastructural.)  In each of these examples, we have chosen one
specific application that is meant to represent a possibly large number of
comparable applications.

%\jhanote{Dan: This paragraph seems to suggest some % similarity with
infrastructural applications, but doesn't quite do % so. Can we make a bit more
explicit the intent. For example, If we % break into ``collaborating'' vs
``competing'' and put NGS into a % ``competing'' category the remainder of the
paragraph above would be % a useful and fair elaboration on infrastructural
applications?  % Reading forward, Section 5.1, it appears to be consistent with
the % ``observations'' section. For example, `` Some of the applications % we
describe are really groups of applications, meaning that we are % grouping
together a number of independent programs, written by % different authors.
These independent programs might be competing % means to answer the same type of
science question and thus are not % intended to run together, or they could be a
set of smaller % applications collaborating to answer the same overall science %
question.''}  \katznote{some changes made above.  I'm not sure why % NGS is
competition and the others are collaboration - to me they are % all a bit of
both.} \jhanote{NGS is classified as a traditional % application; others as
infrastructural and as having multiple % distinct applications. Thus it is
natural to have collaborative.  % But as NGS is said to be a traditional,
different applications can % be perceived to be competitive. I agree that the
term competitive is % not perfect} \katznote{I've added a note to say that this
(applications % selected to represent groups) is orthogonal to traditional vs %
infrastructural.  (This is actually why this is a different paragraph.)  % Does
this help?}

%\katznote{not sure which category is right for: Power %Grids (\S\ref{power})
Yogesh, what do you think?}  %\ysnote{The infra category is appropriate. i.e. no
change required}

%\alnote{After reading the definition a couple of times, I still have a
%difficult time to distinguish between infrastructural and archetypical
%applications. The definitions sound very similar.}  %\katznote{this section has
been re-written.  I've removed archetypical applications, and added a paragraph
at the end of this section (2.0) to mention this idea without naming it.  Please
see if this is now better.}  %\alnote{reads very well!}

The discussions in the workshops and our analysis of the applications led us to
examine the use of data in different parts of the applications, where we found
some common patterns across applications.  For example, applications that
involve streaming data collect data from sensors, transform and filter the data,
store the results, and then later process (analyze) the stored results.  In
general, we think that all of the applications we have studied can be mapped to
a set of stages or phases.  This is illustrated in
\figurename~\ref{fig:figures_application-stages}.

The question of where the application, as opposed to the entire system, %(where
the system is defined as all of \figurename~\ref{fig:figures_terminology}),
actually starts is somewhat difficult.  In this paper, we have decided not to
treat hardware data sources (e.g., sensors, telescopes) as part of the
applications.  Therefore, these applications generally start with the data
movement that follows the actual data generation.

Unlike the sensor applications, another set of applications generate data
computationally, for example, from a model or simulation.  These applications
then do start with data generation, and they often have just three stages, where
data is generated in the first stage, stored in a second stage, and then
processed in a third stage (i.e., the transformation \& decisions stage in
\figurename~\ref{fig:figures_application-stages} does not occur).

Other applications may operate on stored data (i.e., they only contain the
processing stage and its associate data movement).  %\alnote{We should refer to
these application stages from the % application descriptions. Currently, this
classification is not % used.}  \katznote{have now done, but this was harder to
do than I % thought it would be, particularly to define where the application %
starts - for example, LSST as an app does not include data % generation, but
starts after this (data transformation \& % decisions), and SOA Astronomy does
not include data storage, but % does include the movement from the storage.
Perhaps I need to go % back and make the fact that many apps start with data
movement more % clear.}\alnote{ok}

\begin{figure}[htbp] \centering
        \includegraphics[width=1.0\textwidth]{figures/application-stages-new.pdf}
        \caption{Application Stages: The work in data-intensive applications can
be put into four common application stages: data generation, transformation \&
decisions, storage, and processing, though not all applications have all stages.
% \alnote{We have some applications (e.g. sensor networks) that give feedback to
the % data generation stage. Do we want to refer to this possibility here?}  %
\alnote{data transformation and decision making sounds to me similar to
processing} % \katznote{what do you suggest?}\alnote{The proposal would be to
just call it Data Processing as before. Since % there are many forms of
processing (transformation being just a specific form of processing)}
\katznote{I want to distinguish between the processing that happens before
storage % and the processing that happens after storage, so I am thinking of not
changing this.  % Would this really bother you?}
                \label{fig:figures_application-stages}}
\end{figure}

%\katznote{Dan to clean up the description of app phases in the intro and the
apps, %explain where apps start, with data movement, data generation, data
storage, etc. -- %then ask Shantenu to read and comment on this.}

%\katznote{think about linking this with 4.1}

%Application Scenario Characteristics for Dynamic Data (Various)
%
%Identify: (1) what is the case scenario within the application; (2) what data
is dynamic; (3) how the data is being used in the context of each of these
scenarios.
%
%\katznote{proposed structure for rewriting these entries:\\ %1. what does the
application do - what problem does it solve (question 1 below) \\ %2. how does
it do it (questions 2,3,5,6 below, also address cyberinfrastructure (where does
the app run)) -- perhaps have a diagram here of what happens in the app \\
%3. what are the big data aspects (question 4b below) \\ %4. what are the
distributed aspects (question 3 below)\\ %5. what are the dynamic aspects
(question 4a below)\\ %6. what else is important (questions 7,8 below, also
address programming systems (how was the app written; what tools does it use))\\
%}
%
%
%Currently trying to do this by asking the following questions:
%\begin{enumerate}
%\item[1] What is the purpose of the application?
%\item[2] How is the application used to do this?
%\item[3] What infrastructure is used? (including compute, data, network,
instruments, etc.)
%\item[4] What dynamic data is used in the application?
%\begin{enumerate}
%\item[4a] What are the types of data,
%\item[4b] What is the size of the data set(s)?
%\end{enumerate}
%\item[5] How does the application get the data?
%\item[6] What are the time (or quality) constraints on the application?
%\item[7] How much diverse data integration is involved?
%\item[8] How diverse is the data?
%\end{enumerate}
%
%Please feel free to also talk about the current state of the %application, if
it exists today, and any specific gaps that you know %need to be overcome.

%\subsection{Silvia's Biosciences app: NGS, medical imaging
(Silvia)\label{bioSilvia}}

\subsection{Next Generation Sequencing (NGS) Analytics \label{bioSilvia}}

The (grand) challenge in bioinformatics is the need to provide support for the
analysis of genome sequence data that is becoming available due to the abundance
and high-throughput capability of the next-generation sequencing devices. The
rate of growth of data from the NGS machines humbles Moore's law; it has
increased by more than three orders of magnitude in barely a decade.

There are many different types and components of analyses of NGS data.  However,
fundamentally important to all types is the ability to efficiently and
effectively map/align the short reads to a reference genome.

Alignment underpins multiple scientific questions such as comparing metagenomic
(DNA sequence) data sets against each other, which could be used to distinguish
human disease biomarkers among different individuals, determine different soil
(or water or air or ...)  microbial composition from different samples, etc.

This subsection describes a set of {\em traditional applications} that are used
in the processing of next generation sequencing data.  The applications have two
of the possible stages: data storage and data processing.
%\katznote{Shantenu/Yogesh, can you confirm this or fix it?  % I'm sure data
generation is not part of this app, but am not sure % about transformation \&
decisions. - see % Figure~\ref{fig:figures_application-stages}.} \jhanote{Dan:
I'm % confused. Not sure I understand what you mean by data-generation is % not
part of this app?}  % \katznote{the sequencer is not part of the application -
the application % starts once the data has been generated, thus it starts on a
data movement % arrow.  I'm just not sure which one.}\alnote{I think the phases
storage and % processing are fine for NGS}

\subsubsection*{Application description}

There are multiple tools (programs) that can perform the mapping/alignment
phase, e.g., BWA, BFAST, Bowtie, MAQ, etc. Each has its own strength and its own
native data support/capability; they also implement different alignment
algorithms. Thus, based upon different design objectives and trade-offs, they
typically have different constraints and efficiencies.

Several of these tools are also composed together as a linear pipeline for
performing sequencing and analysis. The pipeline may go through alignment, SNP
calling, haplotyping, and so on, with different tools used in each stage. These
pipelines can often be performed independently on each chromosome to provide
simple parallelism.

All algorithms, architectures, storage, and networks have broken or will break
soon.  Some NGS analytics are not good fits for core-rich and memory-poor modern
architectures. Some problems (e.g., comparing k-mers across metagenomic data
sets) need terabyte-scale memory, not thousands of cores.~\cite{ngs-gap}

Individual data sets, which will be terabyte-scale, are generated by NGS
machines at a variety of locations. There could be many thousands of such sets
to compare. Having them all local may not be feasible (the transfer time from
where they are generated may be prohibitive).

\subsubsection*{Big data aspects}

Individual data sets will be terabyte-scale. For example, for a standard genomic
sequence, there are billions of short (35-100 bases) reads of DNA.  And there
could be many thousands of such sets to compare for a given problem.  Getting
more accuracy for sequencing may require multiple coverage (e.g. 10x), that can
cause sequence sizes to increase linearly.  This makes NGS data management
fundamentally a big data problem.  NGS analytics, in addition to being a big
data problem, is also a computationally demanding and distributed computing
problem.  The computational demands arise from the often complex and intensive
analysis that has to be performed on data, which in turn arise from algorithms
that are designed to account for repetitions, errors and incomplete information.

\subsubsection*{Distributed aspects}

The type of NGS research being conducted can distinguish the distributed
properties of the application. Researcher groups that own sequencing machines
need to deal with large dataset sizes that need to be resequenced and reduced
for further analysis. Other researchers may analyze large catalogs of sequenced
genome databases are maintained, and perform pattern matching. Further, clinical
may be interested in in-depth analysis of an individual sequence for disease
diagnosis or visualization.

The distributed computing aspects arise at multiple levels: for example, the
simple act of having to move data from source (generation) to the destination
where computing (analysis) will occur is a challenge due to the large volumes
involved. Trade-offs exists between the cost/challenges in distributing data
versus I/O saturation or memory bottlenecks.

When coupled with the compute intensive nature of the problem, it soon emerges
that a fundamental challenge is not only whether to distribute, but where to
distribute, what to distribute (should the computing move to the data, or the
data move to the compute), and how to distribute (what tools and infrastructure
to use).

\subsubsection*{Dynamic aspects}

The dynamic challenges of NGS data are more subtle; the data themselves are not
dynamic -- the sequences are acquired once and analyzed many times. However the
execution of these applications on distributed infrastructure have dynamic
aspects when optimized resource usage is considered. For example, workload
decomposition and distribution must be determined dynamically in order to
optimally use resources, e.g., compute resource selection can be based on data
location, processing profile and/or network capacity.

\subsubsection*{Other important issues}

There are several toolkits that have emerged for NGS providing incrementally
better algorithms for sequencing and analysis \cite{SAMtools, BioJava, BioPerl,
.NETBio}.  However, as a consequence of the broad range of infrastructural
requirements, there isn't currently a common ``standard'' cyberinfrastructure
used for NGS analytics; researchers use what is easily available to them, or try
to use COTS infrastructure. Historically, leaders in the field have developed
infrastructure for the larger community to build-upon, e.g., QCDOC or
MD-GRAPE. But given the broad range of requirements, a hardware-only solution is
unlikely for a broad range of problem instances, and a combination of hardware
and software approaches will be employed. This reinforces the need for flexible
programming systems and associated runtime environments that support collective
(hardware-software) optimization.

Nowadays, NGS sequencers are even being shipped with servers or small clusters
for performing the alignment and resequencing as a pipeline right after the
reads have been generated~\cite{ilm}.  %\alnote{What kind of codes are used %in
these clusters? What are the characteristics of these codes?}  %But any
additional analysis is not supported.  %\alnote{What would be an additional
analysis}
%
Scientific workflow systems like Taverna~\cite{Taverna:2006},
BioKepler~\cite{biokepler:2011}, and Galaxy~\cite{galaxy} (though Galaxy does
not support distributed systems currently) have been successful in integrating
genome analysis tools for bioinformatics. Script based pipelines that combine
executables are also used often. Very few workflow systems have explicit support
for distributed data or dynamic data handling.  Therefore reusable approaches at
extending existing tools to support distributed and dynamic data are likely to
be useful.  %\katznote{Shantenu/Yogesh/Andre, please read this paragraph and
tell me if changes %should be made to this section at this point}\alnote{ok with
me}

%\alnote{Some details on how distributed/dynamic data is handled in % these
workflows would be good.}

%\katznote{from Silvia Delgado Olabarriaga, Tom Slezak, and Shantenu}
%
%\katznote{summary below assigned to Shantenu}

%\subsubsection{new summary}

%\katznote{Please make a new title for this section}

% \katznote{I was assuming you would leave the questions here, so that % we
could pull some of this material into a table later...}

% 1. {\bf What does the application do? What problem does it solve?}

% \katznote{from answer to question 1}

%2. {\bf How does it do it?}

% \katznote{from answers to questions 2,3,5,6 % below, also address
cyberinfrastructure (where does the app run) -- % perhaps have a diagram here of
what happens in the app}

%\ysnote{The first para of this section reads more like ``what does the
application do'' rather than % ``how''. Should it move above? Also, the
description of cores/memory above can move here as it % relates to CI. }
%\ysnote{This section will benefit from describing the (often) linear pipeline
of % sequencing, with individual chromosomes that can be processed
independently.}  %\katznote{Yogesh or Shantenu, can you do address Andre's
comments in this section?}  %\ysnote{I've taken a pass. Shantenu, do you care to
make changes?}  % 3. {\bf What are the big data aspects?}  \katznote{from answer
to % question 4b below} 4. {\bf What are the distributed aspects?}  %
\katznote{from answer to question 3 below} \ysnote{distinguish % between
individual research project where compute can move to % data; clinical testlabs
that sequence and transmit sequenced data % to end user for post processing;
collaboratory projects like % 1000-genome where data has to be aggregated
centrally. Sequencing % done remotely but visualization done locally.}  %5. {\bf
What are the dynamic aspects?}  %\katznote{from answer to question 4a below}
%6. {\bf What else is important?}

%\katznote{from answers to questions % 7,8 below, also address programming
systems (how was the app % written; what tools does it use)}

%\ysnote{See biohpc.net, bioperl, biojava, bio.net} %\katznote{I'm not sure what
action goes with this comment.  Yogesh?}  % It is unlikely that researchers or
NGS teams will build integrated % customised infrastructure for no
cyberinfrastructure to solve this % problem. The solution to

% \katznote{do we want to retain the figure? - currently commented out}

% In general this is a large (for biology -- don't think in terms of %
exascale!) data problem with large (all-vs-all comparison) % computing. There
are strategies to make this more scalable that will % reduce the computing
volume.

% i) large (exascale) data and large (exascale) computing (e.g., data % mining,
analytics, bioinformatics)

% ii) large (exascale) data and modest computing (e.g., search)

% iii) modest data and large (exascale) computing (e.g., % simulation/modeling)

% Potentially peta-bytes or exa-bytes of data. Not sure how to quantify % the
computing. (Huge memories will be potentially more valuable than % huge numbers
of cores -- although this is heresy to the physics % crowd.)

% Computing might then have to be % distributed. Biology is wrestling with this
now.

% Although only the alignment step is computation-intensive, it is % more
practical to perform all the others on the grid to minimize data % transport
to/from the user's workstation, which is normally on low % speed networks. That
is, the computation is statically and manually % taken to the data in this
case. This can lead to sub-optimal resource % usage.

% Data is not really `dynamic' in this case -- the sequences are % acquired once
and analyzed many times. However the execution of this % application on a
distributed infrastructure has dynamic aspects when % optimized resource usage
is considered:

%\begin{itemize}
%\item Split/merge the data according to available resources
%\item Compute resource selection based on data location, processing profile
and/or network capacity
%\item Resource selection based on security constraints (access control, trust)
%\item Collection and processing of monitoring information to steer grid
enactment.
%\end{itemize}
%
%
%\subsubsection{old material follows}
%
%\begin{itemize}
%\item Infrastructure monitoring for the execution of workflows to % manage
failure
%\item the application is sequence alignment (split data (this is where there is
dynamic decision making) to execute the alignment in parallel)
%\item DTI Imaging: two cases are interesting: DTI atlas (split, run, merge,
split, merge - reduction from 10000 tasks to 1 output); PCA for classification
of patients/control
%\end{itemize}
%
%\subsubsection{Silvia's contribution}
%
%NGS sequence alignment on the AMC e-bioscience infrastructure
%
%1. What is the purpose of the application?
%
%The application performs analysis of DNA sequencing data for various %goals in
life science research, e.g., Mutation screening, Virus %discovery, Genome-wide
Associations (GWA), Linkage analysis, %Comparison of bacterial genomes, microRNA
expression and Exome %sequencing. The data analysis is composed of various
methods that are %implemented by a bioinformatician, whereas the data is
typically owned %by the life scientist.
%
%2. How is the application used to do this?
%
%Note: Here we focus on data analysis from the perspective of the
%bioinformatician, excluding steps related to data acquisition %(transformation
of raw images into sequences of amino acids) and %posterior biostatistics and
epidemiological analysis.
%
%The data analysis is implemented as a pipeline of generic and %customized
components, mostly third-party public software tools. The %pipeline may differ
for each specific study, depending on the data %acquisition and the research
question. Typically it includes data %preparation (file conversion, reformatting
and filtering); alignment %(comparison) with some reference database as the
human genome; %post-processing (output conversion, reformatting, statistical
%analysis) and visualization. Sequence alignment is the most
%computing-intensive step, which became prohibitive for sequential %execution in
next generation sequence experiments. All data is stored %in files that are
often completely read in memory for fast %processing. Compressed formats are
usually adopted. Although the files %are large, the data can be easily
decomposed because the processing is %done on each sequence (`string') at a
time.
%
%3. What infrastructure is used? (including compute, data, network,
%instruments, etc.)
%
%\begin{figure*}[htbp]
% \begin{center}
% \includegraphics[width=\textwidth]{figures/silvia.pdf}
% \caption{Figure from Silvia}
% \label{Fig:silvia}
% \end{center}
%\end{figure*}
%
%see Figure~\ref{Fig:silvia}.
%
%Data:
%
%Data acquisition: Roche, Solid and Illumina next generation sequencers
%
%Local storage: dedicated server at the AMC network (ftp)
%
%Grid storage: Storage Elements of the Dutch e-Science Grid %(www.biggrid.nl)
%
%Data transport:
%
%From sequencer to server: AMC network (XXX) or off-line media
%
%Between data server and grid storage: public network (XXX); lightpath %(near
future)
%
%Between grid storage and worker nodes: public (XXX) or site (XXX) %network
%
%Computing:
%
%Local Computing: clusters and servers at the AMC and workstations %(laptops,
desktops) anywhere Grid Computing: Dutch Grid %(www.biggrid.nl), part of EGI
%
%Originally all steps were performed in some local server or on the %user's
workstation. The visualization is still performed in the user's %(windows)
workstation, but all the other steps are currently performed %on the Dutch grid
infrastructure. The pipeline is described as a grid %workflow using the GWENDIA
language %(\url{http://dx.doi.org/10.1145/1645164.1645171}) and enacted on the
%Dutch Grid using the MOTEUR2 workflow engine
%(\url{http://modalis.polytech.unice.fr/softwares/moteur/start}).
%
%PS: although only the alignment step is computation-intensive, it is %more
practical to perform all the others on the grid to minimize data %transport
to/from the user's workstation, which is normally on low %speed networks. That
is, the computation is statically and manually %taken to the data in this
case. This can lead to sub-optimal resource %usage.
%
%4. What dynamic data is used in the application?
%
%Note: data is not really `dynamic' in this case -- the sequences are %acquired
once and analyzed many times. However the execution of this %application on a
distributed infrastructure has dynamic aspects when %optimized resource usage is
considered:
%
%\begin{itemize}
%\item Split/merge the data according to available resources
%\item Compute resource selection based on data location, processing profile
and/or network capacity
%\item Resource selection based on security constraints (access control, trust)
%\item Collection and processing of monitoring information to steer grid
enactment.
%\end{itemize}
%
%(a) What are the types of data,
%
%DNA sequences (strings) stored in open formats (FASTA, SFF, SAM, BAM) %Bundles
of results from alignment tools (custom format, binary?)
%
%(b) What is the size of the data set(s)?
%
%one example for BWA alignment tool (short sequences) follows:
%
%input data: Sequencing Data in the csFasta format, normally 25-35~GB; %Quality
files in the .qual format, normally 50-80~GB; Reference DB in %the Fasta BS
format: 3.2~GB (human genome) 140~MB (one chromosome)
%
%Intermediate files: Reference BWA index, normally 4.5GB (human genome) %240~MB
(one chromosome); Sequencing Data in the FastQ format %(fastq.gz), normally
20-30~GB
%
%Output: Results in .sai format (direct output of BWA), normally %2-3~GB;
Results in .sam format, normally 55-75~GB; Results in .bam %format, normally
20-30~GB
%
%5. How does the application get the data?
%
%From a local file. The legacy code is wrapped to stage the inputs and %outputs
in/out the worker node.
%
%All data is stored in files that are often completely read in memory %for fast
processing. Compressed formats are usually adopted. Although %the files are
large, the data can be easily decomposed because the %processing is done on each
sequence (`string') at a time
%
%6. What are the time (or quality) constraints on the application?
%
%In research there are no strict constraints on the time to complete %the
analysis.  In patient care, time could determine life, death or %money, but we
have no concrete case yet.
%
%Quality in this case refers to the accuracy of the results %obtained. This is
essential for research and patient care.

\subsection{ATLAS (an Example of Use of the WLCG) \label{WLCGSteve}}

%\katznote{from Steve Fisher}
%
%\katznote{summary below assigned to Dan}
%\subsubsection{new summary}

%1. {\bf What does the application do? What problem does it solve?}

%\katznote{from answer to question 1}

The Large Hadron Collider (LHC)~\cite{lhc} at CERN in Geneva is producing a
large amount of data across a number of experiments, including
ATLAS~\cite{atlas}. The application discussed in this section is an {\em
infrastructural application} (meaning that it consists of multiple stages that
are run by different people, in this case data acquisition, storage,
distribution, and analysis). It has the goal of allowing a physics group or a
user to analyze data to understand a specific physics channel (underlying
physical process) as recorded by ATLAS.  The set of available data grows
steadily over time as more data are collected (by the experiment) and
reconstructed. This application thus contains two stages: data storage and data
processing.  %\alnote{prev. also acquisition was %part of the application}
\katznote{I don't see that it should be.  Doesn't the app %start after the data
is in the Tier 1 archive?}

\subsubsection*{Application description} %2. {\bf How does it do it?}

%\katznote{from answers to questions 2,3,5,6 below, also address
cyberinfrastructure (where does the app run) -- perhaps have a diagram here of
what happens in the app}

Both physics groups and individuals run jobs on the WLCG (Worldwide LHC
Computing Grid)~\cite{lhcb} as they see fit, either reading the reconstructed
data or reading intermediate datasets created by themselves or their
colleagues. If, in the future, resources should prove inadequate then
adjustments to working practices to coordinate processing may be made to
increase physics output.

The WLCG has a hierarchical architecture. CERN (the `Tier~0' site) is connected
to eleven `Tier~1' (national) sites by a dedicated optical network. This may be
extended in the future.  All data is stored at CERN, and as needed, various
parts of the dataset are copied (and cached, i.e., replicated) to the Tier~1
sites, and from there to `Tier 2' (regional) sites and to `Tier~3' (university
or group or individual) sites.

Before data is analyzed, the file to be analyzed is copied from the closest site
to the SE (Storage Element) at the WLCG site where the job will run. From there
it is either copied to a local disk or read from the SE.

There is no particular time constraint on the processing, although physics
groups are eager to understand the ATLAS experiments and thus want to process
data as quickly as possible.  In the event some data are missed during
processing, the only effect is a reduction of statistical precision in the final
analysis. Accurate bookkeeping is therefore important in order to know which
data have been processed.

 \subsubsection*{Big data aspects} %3. {\bf What are the big data aspects?}

%\katznote{from answer to question 4b below}

Roughly 20~TB of data are collected every day. Multiple generations of processed
data are kept. Many data files are replicated on multiple sites; this is done
both by copying it to where it is likely to be useful and partly dynamically as
the need arises.  The bulk of the data is individual event data at different
stages of refinement. This is held in an in-house representation of C++
serialized objects. Though this format is not standard it is used by all LHC
experiments.

 \subsubsection*{Distributed aspects} %4. {\bf What are the distributed
aspects?}

%\katznote{from answer to question 3 below}

The WLCG has 250,000 cores distributed over 140 sites and with 100~PB of
disk. It uses a mixture of gLite software, experiment specific software and
physics group specific software.  The data is distributed and replicated as
described previously.

 \subsubsection*{Dynamic aspects} %5. {\bf What are the dynamic aspects?}

%\katznote{from answer to question 4a below}

The data being processed can be considered as dynamic as it grows steadily as
new raw data are collected and the basic processing program (called
reconstruction) is run to produce information about specific particles involved
in a collision. This reconstruction is rerun periodically (two or three times
per year is expected) as calibrations of the detector are
improved. Reconstruction algorithms are also improved as are the physics groups'
selection codes.  %\alnote{Is the reconstruction code the fundamental entity
that is required to deal with new data? Sounds rather like an auxiliary tool.}
%\katznote{reconstruction is a second type of processing.}\alnote{should we
explicitly state that?}

 \subsubsection*{Other important issues} %6. {\bf What else is important?}

%\katznote{from answers to questions 7,8 below, also address programming systems
(how was the app written; what tools does it use)}

Data analysis is really a pleasingly parallel task, where various events are
analyzed in parallel and then a summary is created of all the analyses.

The key challenge here is really in the infrastructure -- building a system that
can store massive amounts of data and move them to where the processing is to be
done by a large diverse group of users.

%\subsubsection{old material follows}
%
%
%\begin{itemize}
%\item Most of the dynamism is in infrastructure, not in the application
(e.g. use of PilotJob)
%\item Application dynamism: location of file (e.g. if file turns out to be not
local, look up in registry-- but not do this again)
%\item No specific QoS issue concerned with data transfer times
%\end{itemize}
%
%\subsubsection{New answers from Steve}
%
%3DPAS - Particle Physics - the example of Atlas physics channel selection.
%
% 1.  What is the purpose of the application?
%
%A physics group is interested in a specific physics channel (underlying
physical process) as recorded by the Atlas experiment at the LHC (Large Hadron
Collider) at CERN in Geneva. The set of available data grows steadily over time
as more data are collected and reconstructed.
%
% 2.  How is the application used to do this?
%
% Both physics groups and individuals run jobs as they see fit, either reading
the reconstructed data or reading intermediate datasets created by themselves or
their colleagues. If, in the future, resources should prove inadequate then
adjustments to working practices to coordinate processing may be made to
increase physics output.
%
% 3.  What infrastructure is used? (including compute, data, network,
instruments, etc.)
%
%The WLCG (Worldwide LHC (Large Hadron Collider) Computing Grid) is used with a
mixture of gLite software, experiment specific software and physics group
specific software. WLCG has 250,000 cores distributed over 140 sites and with
100~PB of disk. CERN is connected to the eleven `Tier 1' national sites by a
dedicated optical network. This may be extended in the future.
%
% 4.  What dynamic data is used in the application?
%
%The data being processed can be considered as dynamic as it grows steadily as
new raw data are collected and the basic reconstruction program is run to
produce information about specific particles involved in a collision. This
reconstruction is rerun periodically (two or three times per year is expected)
as calibrations of the detector are improved. Reconstruction algorithms are also
improved as are the physics groups' selection codes. One output of the
reconstruction is `tag' data - summary information written to a database to
facilitate event selection. This is not yet used very much. Calibration data are
held on an RDBMS with Frontier/Squid front ends.
%
% a.  What are the types of data,
%
% The bulk of the data is individual event data at different stages of
refinement. This is held in a in-house representation of C++ serialized
objects. Though this format is not standard it is used by all LHC experiments.
%
% b.  What is the size of the data set(s)?
%
% Roughly 20~TB of data are collected every day. Multiple generations of
processed data are kept. Many data files are replicated on multiple sites; this
is done both by copying it to where it is likely to be useful and partly
dynamically as the need arises.
%
% 5.  How does the application get the data?
%
%A copy of the file is made to the SE (Storage Element) at the site where the
job will run. From there it is either copied to a local disk or read from the
SE.
%
% 6.  What are the time (or quality) constraints on the application?
%
%If some data are missed during processing the only effect is a reduction of
statistical precision in the final analysis. Accurate book keeping is important
to know which data have been processed. If the selection code has been improved
then naturally the physics groups would like to see the results as soon as
possible.
%
%
%\subsubsection{Old answers from Steve}
%
% 1.  What is the purpose of the application?
%
%A physics group is interested in a specific physics channel (underlying
physical process) as recorded by the Atlas experiment at the LHC (Large Hadron
Collider) at CERN in Geneva. The set of available data grows steadily over time
as more data are collected and reconstructed. It would not be efficient to allow
individual physics groups to process this large volume of data in an
uncoordinated manner as it requires reading all available data.
%
% 2.  How is the application used to do this?
%
%Physics groups register their analysis code to be run on the next scheduled
analysis ``train''. The train leaves every two weeks and reads through all the
data, runs the physics selection code of each group and writes out the selected
events and output from the computation to produce relatively small data sets for
each physics group. The groups then use these selections to perform more ad-hoc
analysis.
%
% 3.  What infrastructure is used? (including compute, data, network,
instruments, etc.)
%
%The WLCG (Worldwide LHC (Large Hadron Collider) Computing Grid) is used with a
mixture of gLite software, experiment specific software and physics group
specific software. WLCG has 250,000 cores distributed over 140 sites and with
100~PB of disk. There is no wide area dedicated network infrastructure.
%
% 4.  What dynamic data is used in the application?
%
%The data being processed can be considered as dynamic as it grows steadily as
new raw data are collected and the basic reconstruction program is run to
produce information about specific particles involved in a collision. This
reconstruction is rerun periodically (two or three times per year is expected)
as calibrations of the detector are improved. Reconstruction algorithms are also
improved as are the physics groups' selection code. Calibration data are held on
an RDBMS with Frontier/Squid front ends.
%
% a What are the types of data,
%
% The bulk of the data is individual event data at different stages of
refinement. The is held in a non-standard representation of C++ serialized
objects.
%
% b.  What is the size of the data set(s)?
%
% Currently the reconstructed data (excluding copies) is about m TB and is
growing at about n TB per month. Many data files are replicated on multiple
sites. This is done both by copying it to where it is likely to be useful and
partly dynamically as the need arises.
%
% 5.  How does the application get the data?
%
%This is mostly by copying the file to where the job will run - though some
remote file access is also used.
%
% 6.  What are the time (or quality) constraints on the application?
%
%If some data are missed during processing the only effect is a reduction of
statistical precision in the final analysis. If the selection code has been
improved then naturally the physics groups would like to see the results as soon
as possible.

%\subsection{Astrophysics -- Virtual Observatory (Bob) \label{astroBob}}
%
%\katznote{this will be merged into the LSST and SOA entries}
%
%
%\begin{itemize}
%\item Analysis of data to generate event streams $\rightarrow$ lead to the %
positioning of telescopes
%\end{itemize}
%
%\subsubsection{Answers from Bob}
%
%I'll assume that the ``application'' in question is the sum of the two
%activities, despite the fact that they are different in many respects, %and
I'll try and distinguish between them where appropriate.
%
%Let me know if you want clarification or further information - in some cases I
wasn't quite sure what you were after.
%
%1. What is the purpose of the application?
%
%An increasingly important feature of astronomical research is the existence of
%systematic sky surveys covering the whole sky (or large fractions thereof) in
%many different wavebands. The principal outcome from these surveys are
%astronomical source catalogues, which contain values for attributes
%characterizing all the celestial sources detected in the sky survey dataset.
%These are now typically implemented in relational databases, with SQL
interfaces %implemented through web pages, and the principal goal of the Virtual
Observatory %(VO) is to standardize access to these, and other astronomical data
resources, %so that astronomers can readily perform multi-wavelength analyses,
combining all %the extant data on particular sources.
%
% 2. How is the application used to do this?
%
%The VO includes a registry that contains metadata describing all the data
%resources published using VO data access standards, enabling the discovery of
%datasets relevant to a particular analysis. Web service implementations of data
%access protocols enable users to access data in these repositories
%programmatically, and distributed scratch space is provided for the storage of
%intermediate result sets, so the user does not need to route large data flows
%through his/her own workstation. In the future, more and more data analysis
%software will be made available through web services compliant with VO
%standards, enabling more of the data integration and analysis process to be
%combined in workflow.
%
% 3. What infrastructure is used? (including compute, data, network, %
instruments, etc.)
%
%The data resources published to the VO exist in a number of data centers
%distributed internationally. Major datasets---e.g. those from large sky
%surveys---are typically implemented in relational databases running on
high-spec servers %or clusters thereof. The compute requirements of the VO are
currently modest, %but are likely to increase over time. Astronomers typically
use normal academic %networks, but more specialized networking is required in
some cases: e.g., sky %survey data is moved from Cambridge to Edinburgh using
the UKLight project's %dedicated optical network, and radio interferometers
require high-bandwidth %fibre connections from each antenna to the correlator.
%
% 4. What dynamic data is used in the application?  %(a) What are the types of
data, (b) What is the size of the data set(s)?
%
%The VO currently includes little dynamic data. The one exception is VOEvent
%messages, which are notifications of transient events which can be used to
%trigger follow-up observations on other telescopes. These are small XML
messages %containing basic information about the position of a transient source
and a %small amount of other metadata which the operators of a given telescope
can use %to define triggers for follow-up observations of interest to them or
their %community.
%
% 5. How does the application get the data?
%
%Assuming this refers solely to the dynamic data, then a number of telescopes
%undertaking real-time detection of transient events emit VOEvent messages
%through feeds to which other parties can subscribe.
%
%6. What are the time (or quality) constraints on the application?
%
%The time constraints differ for different types of transient source, but in
some %cases notification must be made within a few minutes if it is to be
useful; the %overheads associated with setting up a follow-up observation mean
that much %shorter time scales are practically irrelevant, although they are
still %scientifically interesting. The quality constraints centre on providing a
%telescope with good enough information about a potential transient source that
%it can make a reliable decision about whether to interrupt its existing
%observing program to follow-up the event. In most cases, the candidate
%transient will be some sort of noise event, so fairly sophisticated filtering
is %required to bring the false positive rate down to something acceptable,
given %the cost of observing time on cutting edge telescopes.

\subsection{LSST \label{astroAdam}}

%\katznote{from Adam Barker and Bob Mann}
%
%\katznote{summary below assigned to Dan}
%\subsubsection{new summary}

% 1. {\bf What does the application do? What problem does it solve?}

%\katznote{from answer to question 1}

This application, which similarly to the previous application is an {\em
infrastructural application}, wants to find and study variable objects and
moving objects, using a telescope that takes images each night. This is a
scenario that may be used by Large Synoptic Survey Telescope (LSST)~\cite{lsst},
and is somewhat similar to that done by the Palomar Transient Factory
(PTF)~\cite{ptf,ptf2,ptf3} and by Pan-STARRS~\cite{ps1,ps2}.  Overall, this
application is a leading-edge example of the work needed to build a set of data
(catalog and images) that then can be used by others for analysis (an example of
which is the next application, in \S\ref{astroSOAAdam}.)  This application
contains three stages: data transformation \& decisions, data storage.
%\alnote{Is the telescope not %part of the application?} \katznote{I don't know
- what do you think?  This %is a general issue I'm struggling with in this and
other applications.} \alnote{I struggle as well - it is just confusing that the
term telescope is part of the name} (Data generation comes from the telescope
itself, and processing is done outside of this application, e.g. using the next
application.)

\subsubsection*{Application description} %2. {\bf How does it do it?}

%\katznote{from answers to questions 2,3,5,6 below, also address
cyberinfrastructure (where does the app run) -- perhaps have a diagram here of
what happens in the app} %\ysnote{This section feels like it has too much
details/steps. Maybe we just need to stress the dynamic and distributed parts
e.g. \#1-6 can be collapsed into a couple of bullets, and \#7 into another.}

Processing of data (data analysis and detection of new sources) from each image
should be done while the next image is being taken by the telescope.  This means
that if anything unusual is detected, normal observation can be
interrupted. Other observing resources can also be notified instantly, so they
can observe the same event. As data is collected, it is added to all the data
previously detected from the same location of sky to create a very deep master
image. LSST will also build up a database of all known moving objects.

Every time a new image of the sky is obtained, the master image will be
subtracted from it. The result is a subtracted image that contains the
difference between the current sky image and its average state.

This subtracted image is then processed by a cluster of computers with 3 main
steps:
\begin{enumerate}[(i)]

\item Using the existing object catalog (containing the known orbits of all
known objects), find objects which are expected to appear in subtracted image,
given the area of sky and time of day.  Cross-match expected objects with
sources in subtracted image, resulting in the unmatched-source catalog, which
contains sources that cannot be matched with a previously known object.  If
these sources can be tracked over a few images, an orbit can be determined for
them and they can be added to the object catalog.  Also, the orbit catalog can
be updated with redetections of known objects; each rediscovery provides
information that improves the known orbits.

\item Attempt to classify all entries in the orbit catalog (i.e., all the
objects which now have known orbits). If any Near Earth Objects are detected to
be passing close to the earth, alerts are generated to astronomers so that
follow up observations can be scheduled.

\item If an unknown object is found that cannot be classified locally, at least
two possible options exist. One is a coordinated effort of comparison with other
observatories, where the local observatory issues a call for participation to a
set of additional possible observatories, each of which can accept or reject the
call.  The local observatory then chooses one of the accepting observatories
from which to obtain data, and that observatory takes data and returns it. The
local observatory then decides what to do: classify the object, contact a human,
or call for more observations (automatically to other observatories that decide
whether or not to accept, and when)

A second option is less coordinated, where this observatory would send VOEvent
messages, notifications of transient events which can be used to trigger
follow-up observations on other telescopes. These are small XML messages
containing basic information about the position of a transient source and a
small amount of other metadata which the operators of a given telescope can use
to define triggers for follow-up observations of interest to them or their
community.

\end{enumerate}

 \subsubsection*{Big data aspects} %3. {\bf What are the big data aspects?}

%\katznote{from answer to question 4b below} %\ysnote{The FITS images themselves
may be less interesting since they may only be used ocassionally % (e.g. for
re-processing images using a new algorithm) after the initial object extraction
into % databases. The object databases sizes may be more relevant, and they can
be 3-orders of magnitude smaller % than the images.}

LSST will generate 36~GB of data every 30 seconds, and over a 10-hour winter
night, will collect up to 30~TB.  Data will be stored as FITS image, and in
large databases containing analyses of the discoveries.

 \subsubsection*{Distributed aspects} %4. {\bf What are the distributed
aspects?}

%\katznote{from answer to question 3 below}

The overall system will contain a central telescope with computing data and
resources, linked to a wide area network of observatories and data and computing
resources.  %\alnote{are these data and computing resources still part %of the
application LSST?} \katznote{I'm not sure - how do you think we should handle
this?} \alnote{not sure, but since we say that processing and storage is part of
the app, the resource for this need to be at some place}

 \subsubsection*{Dynamic aspects} %5. {\bf What are the dynamic aspects?}

%\katznote{from answer to question 4a below}

The time constraints differ for different types of transient sources, but in
some cases notification of remote telescopes must be made within a few minutes
if it is to be useful; the overheads associated with setting up a follow-up
observation mean that much shorter time scales are practically irrelevant,
although they are still scientifically interesting. The quality constraints
center on providing a remote telescope with good enough information about a
potential transient source that it can make a reliable decision about whether to
interrupt its existing observing program to follow-up the event. In most cases,
the candidate transient will be some sort of noise event, so fairly
sophisticated filtering is required to bring the false positive rate down to
something acceptable, given the cost of observing time on cutting edge
telescopes.

 \subsubsection*{Other important issues} %6. {\bf What else is important?}

%\katznote{from answers to questions 7,8 below, also address programming systems
(how was the app written; what tools does it use)}

%\alnote{Is the workflow part still part of the app as defined before with the
stages: data transformation \& decision and storage?} \katznote{I think this is
data transformation.}\alnote{ok}

Data will initially be analyzed at the telescope; discoveries will be sent to
other centers.

Pan-STARRS, a contemporary sky survey with similar goals as LSST, uses a
scripted programming model on a cluster for image processing of telescope frames
to extract object attributes as a CSV file.  This is followed by followed by
scientific workflows, based on Trident, that load the files into individual
databases daily, and merges them with a master distributed-database on a weekly
basis on an HPC cluster~\cite{Simmhan:Building:2009}. Data extraction and mining
operations by astronomers are composed using a pseudo-SQL language with user
defined functions and submitted for execution on the master database using a
batch system.

%\subsubsection{old material follows}
%
%\begin{enumerate}
%\item What is the science/engineering problem? What is the purpose of the
application?
%
%Find and study variable objects and moving objects.
%
%2 How will the application solve the problem?
%
%Processing of data (data analysis and detection of new sources) from each image
should be done while the next image is being taken by the telescope.  This means
that if anything unusual is detected, normal observation can be
interrupted. Other observing resources can also be notified instantly, so they
can observe the same event. As data are collected, they will be added to all the
data previously detected from the same location of sky to create a very deep
master image. LSST will also build up a database of all known moving objects.
%
%Every time a new image of the sky is obtained, the master image will be
subtracted from it. The result is a subtracted image that contains the
difference between the current sky image and its average state.
%
%This subtracted image is then processed by a cluster of computers with the
following steps:
%\begin{enumerate}
%
%\item Query object catalog (knows orbits of all known objects) to find objects
which are expected to appear in subtracted image, given the area of sky, time of
day.
%
%\item Cross-match expected objects with sources in subtracted image. The result
is the unmatched-source catalog, which contains sources that can't be matched
with a previously known object
%
%\item Use unmatched-source catalogue to compute pairs of detections separated
by short time intervals, called tracklets. A tracklet is an observable, short
section of orbit. In order to create pairs of detections, all objects in the
unmatched-source catalog are queried against the orbit catalog, in an attempt to
obtain data about these objects from earlier observations. If earlier
observations and current observations can be linked to form tracklets, they are
stored in the tracklet catalog.
%
%\item Use tracklet catalog to attempt to link these tracklets over a larger
time window. This is achieved by again querying the orbit catalog, this time
looking for observations of the same objects even further back in time. If
matches can be found and the tracklets can be extrapolated out into longer
sections of orbit, then they are added as new orbits to the orbit catalog.
%
%\item Update orbit catalog with redetections of known objects. Each rediscovery
provides information that improves the known orbits.
%
%\item Attempt to classify all entries in the orbit catalog (i.e., all the
objects which now have known orbits). If any Near Earth Objects are detected to
be passing close to the earth, alerts are generated to astronomers so that
follow up observations can be scheduled.
%
%\item Compare with other observatories: If an unknown object is found that
cannot be classified locally
%\begin{enumerate}
%\item issue call for participation to set of possible observatories
%\item each observatory accepts or rejects the call
%\item local observatory chooses from accepting observatories one to obtain data
from
%\item that observatory takes data and sends it to local observatory
%\item local observatory decides what to do - classify the object, contact a
human, call for more observations (automatically to other observatories, that
decide whether or not to accept, and when)
%\end{enumerate}
%
%\end{enumerate}
%
%
%2a Which of the following describes the application
%
%large (exascale) data and large (exascale) computing (e.g., data mining,
analytics, bioinformatics)
%
%2b Can you quantify the amount of computing vs the amount of data?
%
%LSST will generate 36~GB of data every 30 seconds, and over a 10-hour winter
night, will collect up to 30~TB.
%
%3 What infrastructure will be used? (including compute, data, network,
instruments, etc.)
%
%A linked network of observatories and data and computing resources.
%
%(a) What will be distributed?  data, compute, both?
%
%both
%
%4 What data be used in the application?
%
%Fits images, and large databases for analysis of the discoveries.
%
% (a) What are the types of data
%
% (b) What is the size of the data set(s)?
%
% (c) Will any data be dynamic?
%
%5 How will the application get the data?
%
%Some will be analyzed at the mountain top, discoveries will be sent to other
centers.
%
%6 Are there time (or quality) constraints on the application?  (If so, what are
they?)
%
%Time component is critical. As real-time as possible.
%
%7 Please feel free to also talk about the current state of the application, if
it exists today, and any specific gaps that you know will need to be overcome.
%
%
%also, the Palomar Transient Factory is the best example to date:
%
%See:
%
%\url{http://www.astro.caltech.edu/ptf/}
%
%\url{http://xxx.lanl.gov/abs/0906.5350}
%
%\url{http://xxx.lanl.gov/abs/0906.5355}
%
%\url{http://arxiv.org/abs/1011.2199}
%

\subsection{SOA Astronomy \label{astroSOAAdam}}

%\katznote{from Adam Barker}
%
%\katznote{summary below assigned to Dan}
%\subsubsection{new summary}

% 1. {\bf What does the application do? What problem does it solve?}

%\katznote{from answer to question 1}

An increasingly important feature of astronomical research is the existence of
systematic sky surveys covering the whole sky (or large fractions thereof) in
many different wavebands. The principal outcome from these surveys are
astronomical source catalogues, which contain values for attributes
characterizing all the celestial sources detected in the sky survey dataset.
These are now typically implemented in relational databases, with SQL interfaces
implemented through web pages, and the principal goal of the Virtual Observatory
(VO) being developed by the International Virtual Observatory Alliance
(IVOA)~\cite{ivoa} is to standardize access to these, and other astronomical
data resources, so that astronomers can readily perform multi-wavelength
analyses, combining all the extant data on particular sources.  The previous
application (\S\ref{astroAdam}) was an example of a ``generator'' of such data
resources, this application is a typical ``user'' of them.  As such, it can
generally be considered as having just the last application stage: data
processing.  Thus it is an {\em infrastructural application}.  The particular
example described here calculates the photometric redshift for a given area of
sky using two tools: HyperZ and ANNz, both accessible via web services
interfaces~\cite{barker}.  However, it represents many service-oriented
architecture applications that are designed to work with the VO.

\subsubsection*{Application description} %2. {\bf How does it do it?}

%\katznote{from answers to questions 2,3,5,6 below, also address
cyberinfrastructure (where does the app run) -- perhaps have a diagram here of
what happens in the app}

The VO includes a registry that contains metadata describing all the data
resources published using VO data access standards, enabling the discovery of
datasets relevant to a particular analysis. Web service implementations of data
access protocols enable users to access data in these repositories
programmatically, and distributed scratch space is provided for the storage of
intermediate result sets, so the user does not need to route large data flows
through his/her own workstation. In the future, more and more data analysis
software will be made available through web services compliant with VO
standards, enabling more of the data integration and analysis process to be
combined in workflow.

The application orchestrates a set of services through a pipeline in order to
compute the redshift of a given area of sky.  It contains a set of components.
1) The Wide Field Survey Archive (WFS), which is an image catalog. 2) A
spectroscopic database, used to provide spectroscopic data for an area of
sky. 3) The SExtractor tool, which is an image processing that extracts all
objects of interest (stars, galaxies, etc.)  4) A cross matching tool that
compiles one table of all objects of interest over the five wavebands.  5)
Hyperz, the first photometric redshift estimation algorithm.  6) ANNz, the
second photometric redshift estimation algorithm, and 7) mySpace, the AstroGrid
storage service.

 \subsubsection*{Big data aspects} %3. {\bf What are the big data aspects?}

%\katznote{from answer to question 4b below}

Images will be O(1) GB files.

 \subsubsection*{Distributed aspects} %4. {\bf What are the distributed
aspects?}

The data resources published to the VO exist in a number of data centers
distributed internationally. Major datasets---e.g. those from large sky
surveys---are typically implemented in relational databases running on high-spec
servers or clusters thereof.  The services in this application are also
distributed, with web service interfaces used to link them.

%\katznote{from answer to question 3 below}

 \subsubsection*{Dynamic aspects} %5. {\bf What are the dynamic aspects?}

%\katznote{from answer to question 4a below}

The WFS catalog will be updated regularly, therefore a query at time $x$ will
not provide the same data as a query at time $y$.  In addition, the specific
queries that will be issued by users are unknown, and will change over time.

 \subsubsection*{Other important issues} %6. {\bf What else is important?}

%\katznote{from answers to questions 7,8 below, also address programming systems
(how was the app written; what tools does it use)}

Enough data needs to be available for a given area of sky from both the WFS
archive and the spectroscopic archive.

%\subsubsection{old material follows}
%
%
%1. What is the purpose of the application?
%
%Calculates the photometric redshift for a given area of sky [RA, DEC] using two
tools: HyperZ and ANNz, both accessible via WS interfaces.
%
%2. How is the application used to do this?
%
%The application is a service-oriented workflow. Services are orchestrated
through a pipeline in order to compute the redshift of a given area of sky.
%
%3. What infrastructure is used? (including compute, data, network, instruments,
etc.)
%
%The Wide Field Survey Archive (WFS): image catalogue %Spectroscopic database:
used to provide spectroscopic data for an area of sky
%
%SExtractor tool: image processing, extracts all objects of interest (stars,
galaxies etc.)  %Cross matching tool: compiles one table of all objects of
interest over the 5 wavebands
%
%Hyperz: photometric redshift estimation algorithm 1 %ANNz: photometric redshift
estimation algorithm 2
%
%mySpace: AstroGrid storage service
%
%\katznote{checking more on this point}
%
%4. What dynamic data is used in the application?
%
%The WFS catalogue will be updated regularly, therefore a query at time x will
not provide the same data as a query at time y.
%
%(a) What are the types of data, (b) What is the size of the data set(s)?
%
%Data sets (images) will be in the GB range, I can attempt to clarify these
details.
%
%5. How does the application get the data?
%
%WS interfaces.
%
%6. What are the time (or quality) constraints on the application?
%
%Enough data needs to be available for a given area of sky from both the WFS
archive and the spectroscopic archive.

\subsection{Understand the Cosmic Microwave Background \label{astroJulian}}

%\katznote{from Julian Borrill}
%
%\katznote{summary below assigned to Dan}
%\subsubsection{new summary}

% 1. {\bf What does the application do? What problem does it solve?}

%\katznote{from answer to question 1}

This {\em traditional application} performs data simulation and analysis to
understand the Cosmic Microwave Background (CMB)~\cite{cmb}, which is an image
of the Universe as it was 400,000 years after the Big Bang. Tiny fluctuations in
the CMB temperature and polarization encode the fundamental parameters of
cosmology and, using the Big Bang as the ultimate particle accelerator,
ultra-high energy physics. Extracting this information from the data gathered by
current and anticipated CMB observations is an extremely computationally
intensive endeavor for which massively parallel tools have been developed.  This
application contains just one stage: data processing.

\subsubsection*{Application description} %2. {\bf How does it do it?}

%\katznote{from answers to questions 2,3,5,6 below, also address
cyberinfrastructure (where does the app run) -- perhaps have a diagram here of
what happens in the app}

The CMB community wants to extract cosmology and fundamental physics from the
observations gathered by the detectors as time-ordered sequences of O($10^{12}$
- $10^{15}$). These observations are reduced first to a map of O($10^6$ -
$10^8$) sky pixels, then to O($10^3$ - $10^4$) angular power spectrum
coefficients, and finally to O(10) cosmological parameters~\cite{cmb2}.

The central computing challenge for any CMB dataset is the simulation and
analysis of O($10^4$) synthetic observations, used to correct for biases and
quantify uncertainties in the analysis of the real data.  Preconditioned
conjugate gradient techniques are used to solve for the maximum likelihood sky
map given the input data (obtained by scanning the sky with hundreds to
thousands of detectors for weeks to years) and its piecewise stationary noise
statistics.  To avoid the I/O bottleneck inherent in the traditional
simulate/write/read/map paradigm, all simulations are performed on the fly only
when requested by the map-making code.

Going from the map to the angular power spectrum is the computationally most
expensive step. The exact solution scales with the cube of the number of pixels
(in the map), so going from map to angular power spectrum is ruled out now (but
was possible earlier for much smaller observations). The approximate solution
requires sets of O($10^4$) Monte Carlo realizations of the observed sky to
remove biases and quantify uncertainties, each of which involves simulating and
mapping the time-ordered data.

The map-making application is therefore applied to both real and simulated data,
but many more times to simulated data (which requires us to use the on-the-fly
simulation module too).

%The computing scales with the number of observation samples, while the
communication and I/O scale with the number of pixels in the map. For the
current Planck satellite experiment these are O($10^{12}$) and O($10^8$)
respectively; for the proposed CMBpol satellite they will be O($10^{15}$) and
O($10^8$); that is, a 1000x increase in calculation cost is expected over the
next 15 years (including intermediate suborbital pathfinder experiments) but the
communication and I/O costs will be fixed.

 \subsubsection*{Big data aspects} %3. {\bf What are the big data aspects?}

%\katznote{from answer to question 4b below}

There is O(1 - 10) TB input data and a similar volume of output data.

 \subsubsection*{Distributed aspects} %4. {\bf What are the distributed
aspects?}

%\katznote{from answer to question 3 below}

This application is targeting the largest supercomputers available: Hopper,
Tianhe, Blue Waters, etc. It is all about the cycles, although as the
concurrency increases previously solved I/O and communication scaling
bottlenecks typically re-emerge.

The application currently uses a single HPC systems, but the scientists have
discussed using distributed systems, with a model of remote systems being used
for the data simulations.  Each simulation would be launched from the central
system that is building the map, and output data from the simulations would be
asynchronously delivered back to that central system as files that would be
incorporated in the map as they are produced.

 \subsubsection*{Dynamic aspects} %5. {\bf What are the dynamic aspects?}

%\katznote{from answer to question 4a below}

Overall, this application can make use of whatever computing it can access in
order to run the simulations needed to flesh out the observed data (which can be
read from local disk).  The simulations are dynamic in time and in location.

 \subsubsection*{Other important issues} %6. {\bf What else is important?}

%\katznote{from answers to questions 7,8 below, also address programming systems
(how was the app written; what tools does it use)}

A goal is to simulate and map O($10^2$) realizations of an experiment's data in
O(1) wallclock hour to provide 10\% errors during the early analysis stages, and
O($10^4$) realizations in O(100) wallclock hours for the final definitive
analysis.

%\subsubsection{old material follows}
%
%1. What is the science/engineering problem? What is the purpose of the
application?
%
%Cosmic Microwave Background (CMB) data simulation and analysis. The CMB is an
image of the Universe as it was only 400,000 years after the Big Bang. Tiny
fluctuations in the CMB temperature and polarization encode the fundamental
parameters of cosmology and, using the Big Bang as the ultimate particle
accelerator, ultra-high energy physics. Extracting this information from the
data gathered by current and anticipated CMB observations is an extremely
computationally intensive endeavor for which we have developed our massively
parallel tools.
%
%2. How will the application solve the problem?
%
%The central computing challenge for any CMB dataset is the simulation and
analysis of O($10^4$) synthetic observations, used to correct for biases and
quantify uncertainties in the analysis of the real data. We use preconditioned
conjugate gradient techniques to solve for the maximum likelihood sky map given
the input data (obtained by scanning the sky with hundreds to thousands of
detectors for weeks to years) and its piecewise stationary noise statistics. To
avoid the IO bottleneck inherent in the traditional simulate/write/read/map
paradigm, all simulations are performed on the fly only when requested by the
map-making code.
%
%(a) Which of the following describes the application
%
% i) large (exascale) data and large (exascale) computing (e.g., data mining,
analytics, bioinformatics)
%
% ii) large (exascale) data and modest computing (e.g., search)
%
% iii) modest data and large (exascale) computing (e.g., simulation/modeling)
%
%(iii) O(1 - 10) TB input data and a similar volume output.
%
%(b) Can you quantify the amount of computing vs the amount of data?
%
%The computing scales with the number of observation samples, while the
communication and IO scale with the number of pixels in the map. For the current
Planck satellite experiment these are O($10^12$) and O($10^8$) respectively; for
the proposed CMBpol satellite they will be O($10^15$) and O($10^8$) - that is,
we expect a 1000x increase in calculation cost over the next 15 years (including
intermediate suborbital pathfinder experiments) but fixed communication/IO
costs.
%
%3. What infrastructure will be used? (including compute, data, network,
instruments, etc.)
%
%We are targeting the largest supercomputers available - Hopper, Tianhe, Blue
Waters, etc. It's all about the cycles, although as the concurrency we work at
increases previously solved IO and communication scaling bottlenecks typically
re-emerge.
%
%(a) What will be distributed?  data, compute, both?
%
%Both.
%
%4. What data will be used in the application?
%
%Real and simulated CMB data.
%
%(a) What are the types of data
%
%(b) What is the size of the data set(s)?
%
%(c) Will any data be dynamic?
%
%See above.
%
%5. How will the application get the data?
%
%Real data are read from disk; simulated data are generated on the fly.
%
%6. Are there time (or quality) constraints on the application?  (If so, what
are they?)
%
%Typically we want to be able to simulate and map O($10^2$) realizations of an
experiment's data in O(1) wallclock hour to provide 10\% errors during the early
analysis stages, and O($10^4$) realizations in O(100) wallclock hours for the
final definitive analysis.
%
%7. Please feel free to also talk about the current state of the application, if
it exists today, and any specific gaps that you know will need to be overcome.
%
%Moving to the current and future generations of heterogeneous many-core systems
will require a significant re-working of our analysis codes, both to deal with
communication over hundreds of thousands of cores and to take advantage of
accelerator opportunities such as GPUs.

\subsection{Sensor Network Application \label{sensorSimon}}

%\katznote{from Simon Dobson}
%
%\katznote{summary below assigned to Yogesh}
%\subsubsection{new summary}

% 1. {\bf What does the application do? What problem does it solve?}

%\katznote{from answer to question 1}

Marine sensing covers a number of applications, including environmental
monitoring, remote exploration, marine life surveys, and habitat assessment. In
each case, the main challenge is collecting data reliably from a difficult
working environment, without interfering with the natural behaviors the animals
exhibit.

A canonical example is the monitoring of seal and other sea mammal populations
by the Scottish Oceans Institute (SOI), which tags animals with sensor packages
that can record dive behavior, speed, and movement~\cite{SMSSeal,SealContact}.
These datasets are then analyzed offline using traditional statistical
techniques and integrated with Google Earth to visualize animal tracks on a
large scale.  This is an {\em infrastructural application}. It begins after data
generation, and it includes the data storage and processing stages.
%\alnote{data transformation?} \katznote{I don't think so - I think all data %is
stored and then processed as users want} \alnote{I think most sensors do some
form of filtering/aggregation before transmitting things to a
backend...}\katznote{but I would argue that the application %starts after the
filtering/aggregation}

\subsubsection*{Application description} %2. {\bf How does it do it?}

%\katznote{from answers to questions 2,3,5,6 below, also address
cyberinfrastructure (where does the app run) -- perhaps have a diagram here of
what happens in the app}

The sensor packages on the marine mammals report back to base when the animal
crawls up a beach and comes within range of a cellular network. Data are
collected remotely and brought to a central site for analysis. Analysis includes
statistical applications and visualization is done using Google Earth.

 \subsubsection*{Big data aspects} %3. {\bf What are the big data aspects?}

%\katznote{from answer to question 4b below}

The data sets collected are modest in size by scientific-data standards. They
include positions and motion vectors for sea mammals, and concentrations and
gradients for environmental missions.

 \subsubsection*{Distributed aspects} %4. {\bf What are the distributed
aspects?}

%\katznote{from answer to question 3 below}

Time series data is published from distributed sensors, but they are stored in a
central repository. The data is analyzed locally and visualized by distributed
users.

 \subsubsection*{Dynamic aspects} %5. {\bf What are the dynamic aspects?}

%\katznote{from answer to question 4a below}

Time series data is published dynamically from the sensors when they are within
communication distance of cellular towers. So the time series data arrives at
the central repository asynchronously relative to the time of collection.

There is some basic adaptation to the resolution of the data collected to
account for different resolutions of the sensor packs.  Larger-scale
environmental sensing applications may make use of more structured and extensive
adaptation, such as changing the sampling frequency and other management
characteristics in response to the data being observed.

 \subsubsection*{Other important issues} %6. {\bf What else is important?}

%\katznote{from answers to questions 7,8 below, also address programming systems
(how was the app written; what tools does it use)}

Collecting data reliably is challenging due to the distributed deployment of
sensors in hostile environments.

% \subsubsection{old material follows}

% \begin{itemize}
% \item Environmental (Marine) -- sensing within a hostile environment, % data
quality + availability of communication infrastructure (Simon)
% \item Patient monitoring -- data quality, type of data to transmit % (Omer)
% \end{itemize}

% \subsubsection{Answers from Simon}

% 1. What is the purpose of the application?

% Marine sensing covers a range of different applications, including %
environmental monitoring, remote exploration, marine life surveys and % habitat
assessment. In each case, the main challenges are collecting % data reliably in
what is an extremely challenging technical % environment---far more so than in
air---and (in the case of biological % surveillance), without interfering with
the natural behaviors the % animals exhibit.

% 2. How is the application used to do this?

% At present most applications are quite small-scale. A canonical example % is
the monitoring of seal and other sea mammal populations by the % Scottish Oceans
Institute (SOI), which works by tagging animals with sensor % packages that can
analyze dive behavior, speed, movement and the like, % and report back to base
when the animal comes within range of a cellular % network.

% 3. What infrastructure is used? (including compute, data, network,
instruments, etc.)

% The data sets collected are not enormous by scientific-data standards, % and
are analyzed offline using traditional statistical techniques.  % Recent work by
SOI has included integration with Google Earth to allow % animal tracks to be
visualized on a large scale.

% Data are collected remotely, then brought to a central site for analysis.

% 4. What dynamic data is used in the application?

% In the sea mammal case the data isn't all that dynamic, being mainly a % time
series collected with only basic adaptation to the resolution etc.  %
Larger-scale applications in environmental sensing in particular would % make
use of more structured and extensive adaptation (e.g., % changing the sampling
frequency and other management characteristics in % response to the data being
observed).

% (a) What are the types of data, (b) What is the size of the data set(s)?

% For sea mammals: positions, motion vectors. For environmental missions: %
concentrations and gradients. I'd have to enquire about dataset sizes, % but not
excessive.

% 5. How does the application get the data?

% For the sea mammals the animal crawls up a beach, the sensor packages % sees
the cellular network and transmits. For environmental missions % we're still
wondering about that ourselves :-)

% 6. What are the time (or quality) constraints on the application?

% Not sure I see what you mean here. Generally it's just a matter of % capturing
at whatever resolution device the sensor pack can support.

% I could probably get more complete -- perhaps *way* too complete :-) -- %
answers from Mike Fedak in Scottish Oceans if that'd help.

%\subsection{Bioinformatics Applications (Tom Slezak) \label{bioTom}}
%
%\katznote{summary below assigned to Shantenu - perhaps merge this with Sylvia's
app's section}
%\subsubsection{new summary}
%
%1. {\bf What does the application do? What problem does it solve?}
%
%\katznote{from answer to question 1}
%
%2. {\bf How does it do it?}
%
%\katznote{from answers to questions 2,3,5,6 below, also address
cyberinfrastructure (where does the app run) -- perhaps have a diagram here of
what happens in the app}
%
%3. {\bf What are the big data aspects?}
%
%\katznote{from answer to question 4b below}
%
%4. {\bf What are the distributed aspects?}
%
%\katznote{from answer to question 3 below}
%
%5. {\bf What are the dynamic aspects?}
%
%\katznote{from answer to question 4a below}
%
%6. {\bf What else is important?}
%
%\katznote{from answers to questions 7,8 below, also address programming systems
(how was the app written; what tools does it use)}
%
%
%\subsubsection{old material follows}
%
%1. What is the science/engineering problem? What is the purpose of the
application?
%
%Comparing metagenomic (DNA sequence) data sets against each %other. This could
be to distinguish human disease biomarkers among %different individuals,
determine different soil (or water or air or %...) microbial composition from
different samples, etc.
%
%2. How will the application solve the problem?
%
%In general this is a large (for biology -- don't think in terms of exascale!)
data problem with large (all-vs-all comparison) computing. There are strategies
to make this more scalable that will reduce the computing volume.
%
%(a) Which of the following describes the application
%
% i) large (exascale) data and large (exascale) computing (e.g., data mining,
analytics, bioinformatics)
%
% ii) large (exascale) data and modest computing (e.g., search)
%
% iii) modest data and large (exascale) computing (e.g., simulation/modeling)
%
%(b) Can you quantify the amount of computing vs the amount of data?
%
%Potentially peta-bytes or exa-bytes of data. Not sure how to quantify the
computing. (Huge memories will be potentially more valuable than huge numbers of
cores -- although this is heresy to the physics crowd.)
%
%3. What infrastructure will be used? (including compute, data, network,
instruments, etc.)
%
%(a) What will be distributed?  data, compute, both?
%
%Individual data sets will be terabyte-scale. There could be many thousands of
such sets to compare. Having them all local may not be feasible (the transfer
time from where they are generated by others may be prohibitive. Computing might
then have to be distributed. Biology is wrestling with this now.
%
%4. What data be used in the application?
%
%(a) What are the types of data
%
%Genomic sequence, billions of short (35-100?) bases of DNA; unassembled
%
%(b) What is the size of the data set(s)?
%
%Terabyte scale
%
%(c) Will any data be dynamic?
%
%Not for this problem
%
%5. How will the application get the data?
%
%That is a major part of the problem! Data will (in general) be generated from
other places around the world. Getting to the data will be an issue.
%
%6. Are there time (or quality) constraints on the application?  (If so, what
are they?)
%
%You'd like answers in under a day, generally.
%
%7. Please feel free to also talk about the current state of the application, if
it exists today, and any specific gaps that you know will need to be overcome.
%
%Biology is wrestling now with about 3 orders of magnitude increase in DNA
sequence data generation capability in the past 2 years. All algorithms,
architectures, storage, and networks have broken or will break soon. Some
biology problems (not necessarily the one I describe here) are not good fits for
core-rich and memory-poor modern architectures. For some problems (comparing
k-mers across metagenomic data sets, for instance) you need terabyte-scale
memory, not thousands of cores.

\subsection{Climate \label{climateDan}}

%\katznote{from Don Middleton, mostly}
%
%\katznote{summary below assigned to Dan}
%\subsubsection{new summary}

% 1. {\bf What does the application do? What problem does it solve?}

%\katznote{from answer to question 1}

This {\em infrastructural application} is aimed at supporting international
CMIP/IPCC~\cite{intergovernmental2007fourth,cmip} intercomparison activity,
which involves producing, storing, and analyzing data from multiple climate
simulations.  It includes data generation, storage, and processing stages.

\subsubsection*{Application description} %2. {\bf How does it do it?}

%\katznote{from answers to questions 2,3,5,6 below, also address
cyberinfrastructure (where does the app run) -- perhaps have a diagram here of
what happens in the app}

The overall system is comprised of three stages.  In the first stage, climate
centers run a prescribed set of common experiments that produce 2-10 PB of data.
Data can also come from sensors, in which case the center would still
post-process the data before it would be published.  Centers can then publish
their own output data, or send it to another center to publish.  The data is
generated over a roughly 2-year period (then post-processed and published over a
few more months).  This stage is characterized by being both distributed-compute
and distributed-data intensive.

The second stage is data storage.  The Earth System Grid Federation
(ESGF)~\cite{esgf} develops and deploys a federated network of gateways and
associated data nodes.  As models run at each center, the output data is
post-processed into packages with common formats and prescribed metadata
conventions.  Most centers deploy the ESGF data node software stack, and using
this, they manage the data from their experiments.  The data node software stack
scans the data, makes sure it has the right metadata fields, does QA/QC, and
builds a set of catalogs of the prepared data.  Minimally, the catalog provides
HTTP links to data elements, and it can also provide GridFTP endpoints, and/or
product services that abstract the dataset in other ways: get a whole file,
subset, browse, etc.  When the center is happy with the data/catalog in the data
node, they publish it to a host/affiliated gateway.  This submits the catalog to
the gateway.  The gateway then shares this catalog with other gateways so that
all gateways have a consistent view of all the published data.  In general, most
data is replicated at several sites.  The replication activity is manually
requested/initiated by a gateway owner.  The properties of this stage are that
it is distributed data-intensive (with distributed gateways and data nodes) and
that it is dynamic (data appears in the system over time).

The third stage is data analysis.  The approximately 20,000 users (as of
Dec. 2010) can browse/search a catalog at any gateway and locate data, which
might be hosted by a data node affiliated with another gateway. (This can output
a `wget' script that can later fetch the data.)  They can also authenticate, and
thus gain access to a group, which might have private data.  They can also
download data via http, GridFTP, or access product services (the latter of which
uses ESGF's data retrieval syntax---DRS---and which can be scripted.)  Some
users will analyze data from a single model.  However, many applications are
multi-model analyses, where many users want to look at the same parts of the
output of some/all of the models, to understand if and how the models differ.
Some centers will gather some/all of the core archive (plus more, perhaps) on
local systems for local users to perform ``power analyses.''  Each user's data
analyses are almost always done on a single system.  Because there is no
distributed computing infrastructure, this stage is search, access, and transfer
intensive.

 \subsubsection*{Big data aspects} %3. {\bf What are the big data aspects?}

%\katznote{from answer to question 4b below}

The overall data generated and stored in the ESGF is 2-10 PB.  This includes a
core archive, which is 1-2 PB in size and contains the most popular data.

 \subsubsection*{Distributed aspects} %4. {\bf What are the distributed
aspects?}

%\katznote{from answer to question 3 below}

The data is generated by a distributed set of climate centers.  It is stored in
a distributed set of federated archives.  And it is used by a distributed set of
users, who either run data analyses on a climate center with which they are
associated, or they gather data from the ESGF to a local system for their
analyses.

 \subsubsection*{Dynamic aspects} %5. {\bf What are the dynamic aspects?}

%\katznote{from answer to question 4a below}

Data is generated over time, so the data in the ESGF changes.  One might imagine
a future version of ESGF where the launching of data analysis jobs is automated,
in which case the location of those jobs would be dynamic, and the system might
also be able to respond (or optimize for) various types of applications.

 \subsubsection*{Other important issues} %6. {\bf What else is important?}

%\katznote{from answers to questions 7,8 below, also address programming systems
(how was the app written; what tools does it use)}

This application/infrastructure serves a large group of users, and involves a
large amount of political and technical consensus to work.

%\subsubsection{old material follows}
%
%\subsubsection{CDAT} (from Dan)
%
%The Climate Data Analysis Tools (CDAT)~\cite{CDAT} are a scriptable %(Python)
set of packages and tools that are currently heavily used for %analysis of
climate data.  CDAT has been developed by LLNL's Program %for Climate Model
Diagnosis and Intercomparison (PCMDI), and is often %used with data from the
Earth System Grid (ESG)~\cite{ESG}.  The most %common current usage of these
tools with the data is that a user %downloads a set of data from ESG to a local
resource, then uses the %CDAT to analyze the data.  However, the PCMDI group is
now planning to %support a new usage mode, where data transfer can be scripted.
%Initial development of this capability is planned for February 2011, %using the
Globus online data transfer service~\cite{globusOnline} as a %scriptable
component.
%
%\subsubsection{ESGF} (from Don and Dan)
%
%Aimed at supporting international CMIP/IPCC intercomparison activity.  %Three
stages:
%
%{\bf Data Acquisition (Modeling)}
%
%\begin{itemize}
%\item Climate centers run prescribed set of common experiments; produce 2-10 PB
of data
%\begin{itemize}
%\item data can also come from sensors - here, the data would still be
post-processed before it would be published.
%\end{itemize}
%\item Centers can publish their own output data or send it to another center to
publish data generated over \~{}2 year period (then post-processed and published
over a few more months)
%\item Properties: data-intensive, distributed centers, lots of computing
%\end{itemize}
%
%
%
%{\bf Data storage (Infrastructure/system)}
%
%\begin{itemize}
%\item ESGF develops and deploys a federated network of gateways and associated
data nodes
%\item As models run at each center, output data is post-processed into packages
with common formats and prescribed metadata conventions
%\item Most centers will deploy data node software stack, and using this, they
manage the data from their experiments
% \begin{itemize}
% \item the data node software stack scans the data, makes sure it has the right
metadata fields, does QA/QC, build a set of catalogs of the prepared data
% \begin{itemize}
% \item minimally, catalog provides HTTP links to data elements
% \item also can provide GridFTP endpoints
% \item also can provide product services that abstract the dataset in other
ways - get whole file, subset, browse, etc.
% \end{itemize}
% \end{itemize}
% \item When the center is happy with the data/catalog in the data node, they
publish it to a host/affiliated gateway
%\begin{itemize}
% \item this submits the catalog to the gateway
% \item them the gateway shares this catalog with other gateways so that all
gateways have a consistent view of all the published data
%\end{itemize}
% \item Core archive (1-2 PB, most popular data)
%\begin{itemize}
% \item will be replicated at several sites
% \item replication activity manually requested/initiated by a gateway owner
% \item replication: copies catalog to replica site, do data transfer to new
data node, then publish catalog, push catalog to (this) gateway associated with
new data node
%\end{itemize}
% \item Currently building notification service
%\begin{itemize}
% \item track which users have downloaded data - if a problem is found in a data
set or if new data is added to a data set, notify users who have
downloaded/replicated it
%\end{itemize}
% \item Properties: data-intensive, distributed gateways and data nodes, data
appears in the system over time (dynamic)
%\end{itemize}
%
%
%
%{\bf Data processing (Application)}
%
%\begin{itemize}
% \item Users ($\sim$20000 as of Dec. 2010) can:
%\begin{itemize}
% \item browse/search a catalog at any gateway and locate data, which might be
hosted by a data node affiliated with another gateway. (can output a wget script
that can later fetch the data)
% \item authenticate, gain access to a group
% \item download data via http or GridFTP or access product services (uses data
retrieval syntax (DRS) - can be scripted)
% \end{itemize}
%\end{itemize}
%
%\begin{itemize}
% \item abstractions that can be used by a programmer: data is stored in NetCDF
format, uses CF metadata conventions, can be accessed using DRS
% \item Some users will analyze data from a single model
% \item Many applications are multi-model analyses - many users want to look at
the same parts of the output of some/all of the models, to understand if and how
the models differ
% \item Some centers will gather some/all of the core archive (plus more,
perhaps) on local systems for local users to perform ``power analysis''
%\end{itemize}

\subsection{Interactive Exploration of Environmental Data \label{envJon}}

%\katznote{from Jon Blower}
%
%\katznote{summary below assigned to Yogesh}
%\subsubsection{new summary}

% 1. {\bf What does the application do? What problem does it solve?}

%\katznote{from answer to question 1}

In the environmental sciences, the use of visualization techniques is vital for
understanding the ever-increasing volume and diversity of data that is being
produced by Earth observing systems and computer simulations.  The primary
purpose of visualization is to gain insight but the majority of scientific tools
generate static plots that neither the originating scientist nor the recipient
can easily customize to reveal new information. The issues with visualization of
a single dataset are compounded if multiple datasets are to be examined
simultaneously, as is common in environmental science for model validation,
``ground truthing,'' quality control, and data assimilation.

New, interactive modes of environmental data exploration and visualization based
on the principles of simplicity, open standards and user-friendliness are
required at all stages of scientific investigation.  The application described
here is an {\em infrastructural application} that uses stored data and consists
of one stage: data processing.  It represents a number of similar
applications~\cite{blower1,blower2}.

\subsubsection*{Application description} %2. {\bf How does it do it?}

%\katznote{from answers to questions 2,3,5,6 below, also address
cyberinfrastructure (where does the app run) -- perhaps have a diagram here of
what happens in the app}

The applications often take the form of graphical Geographic Information Systems
(GIS) tools, but with better support for large and multidimensional scientific
data. The general vision is of a map-based interface, onto which different
datasets can be overlain.  A processing step may be initiated by rubber-banding
an area of the map and selecting from a list of algorithms that process data
within the selected area, perhaps calculating statistics of the data.  In a
desktop application, this processing may take place on the user's desktop, but
in a web-based application the processing must take place on a server, which may
or may not be co-located with the datasets.  There is therefore the common
problem of moving large amounts of data around in a distributed system, to which
may be added concerns of security in cases in which the data in question are not
public.

 \subsubsection*{Big data aspects} %3. {\bf What are the big data aspects?}

%\katznote{from answer to question 4b below}

Data from instruments is small (1--100~MB) but model output may be very big
(GB--TB). The data is multi-dimensional in nature. Data aggregated over time
from instruments may need to be visualized (e.g. as an animation), causing the
visualized data size to grow.

 \subsubsection*{Distributed aspects} %4. {\bf What are the distributed
aspects?}

%\katznote{from answer to question 3 below}

 Increasingly, environmental data are distributed in multiple locations.  The
applications are driven by data served through distributed web services, that
may in turn be ``fed'' instruments or computer simulation generated
data. Compute services need to be integrated with these data services in an
efficient, easy-to-use, and transparent way, allowing fast data processing.  The
processing algorithms are generally simple and high overhead scheduling systems
can cause a loss of responsiveness.  The infrastructure has more in common with
the Web (or cloud) than the grid.

 \subsubsection*{Dynamic aspects} %5. {\bf What are the dynamic aspects?}

%\katznote{from answer to question 4a below}

Dynamic data are sometimes used, but may arise from real-time feeds from
instruments or from looking at live results from a model. Data hosted at
multiple locations may be frequently updated (often several times a day).  The
data services are designed to allow server-side subsetting of large datasets,
attempting to minimize the amount of unwanted data traveling across networks.
These make it hard to achieve scalability through caching.

In a Web environment, servers and proxy servers (which are sometimes beyond the
control of the data provider or the data user) may retain caches of data in a
strategy to reduce server load and network traffic.  For this reason, the HTTP
protocol provides mechanisms for defining expiration times for data resources,
specifying when a resource ought to be cleared from the cache.  In the
environmental and geospatial communities the Open Geospatial Consortium
specifications are being widely adopted alongside existing protocols such as
OPeNDAP~\cite{opendap}.  Unfortunately these protocols, although built atop
HTTP, do not make it easy for this versioning to be implemented correctly.  The
OPeNDAP protocol does not have the concept of a dataset version, meaning that
clients have no reliable or efficient means to detect that a dataset has
changed.  The OGC protocols, in general, \emph{do} provide versioning at the
level of the service endpoint, but provide no information on when a resource
should be considered expired.  Clients are therefore forced to poll the server
to check for updates.

 \subsubsection*{Other important issues} %6. {\bf What else is important?}

%\katznote{from answers to questions 7,8 below, also address programming systems
(how was the app written; what tools does it use)}

The application aims for near-real-time interaction with the data, i.e the user
should not wait more than a few seconds for some kind of response to a request,
so performance with low latency is a key challenge.  Security is another serious
concern: many environmental datasets are held under access control, and
different providers often have very different access control policies.  In some
infrastructures, these different policies are handled via a role-mapping
mechanism~\cite{nerc_data}.  Concerns of access control---as well as data
volume---also make it difficult for data to be replicated to different
geographic locations (and hence different access control regimes).  In a
distributed environment, it is common for machines to access data on behalf of
end users (e.g., a processing service might download its input data from a
remote store).  Finding a means for the user to delegate his/her authority to a
multi-web-service infrastructure is a key current challenge.  The MashMyData
project~\cite{mashmydata} is investigating two solutions, based on Grid proxy
certificates and OAuth.

% \subsubsection{old material follows}

% (Here's a synopsis of the chapter I'm writing for Malcolm's book on
Data-Intensive Research.  % I'm not sure how to merge this with the answers to
the pro-forma questions below.  Also, % many of these sentences are
cut-and-pasted from the book chapter - future copyright issue?)

% In the environmental sciences the use of visualization techniques is vital for
understanding % the ever-increasing volume and diversity of data that is being
produced by Earth observing % systems and computer simulations.  The primary
purpose of visualization is to gain insight % \cite{card_readings_1999}, but the
majority of current tools in wide use are not adequate % for achieving this.
The time spent to create visualizations is too great, usually due to the %
crippling overhead of low-level technical tasks such as file-format and metadata
interpretation.  % Most scientific tools generate static plots, which are shared
in documents or emails, but % neither the originating scientist nor the
recipient can easily customize or adapt the visualization % to reveal new
information.  The context of how the visualization was generated is usually lost
% in the publication/sharing process, making interpretation and reproduction
much more difficult.

% Visualization of a single dataset is therefore more tedious and less useful
than it should be.  % These problems are compounded if the scientist wishes to
examine multiple datasets simultaneously.  % This is an extremely common task in
environmental science, which forms an important part of model validation, %
``ground truthing'', quality control and data assimilation.

% There is much current interest in developing new, interactive modes of
environmental data exploration and visualization, which can % play a key role at
all stages of scientific investigation.  We consider that, for the case of
interactive % data exploration, concerns of performance, % intuitiveness and
user engagement tend to outweigh the desire for sophistication in data
processing.  % We therefore argue that a new generation of ``4D GIS'' tools,
based on the principles of simplicity, % open standards and user-friendliness is
required to gain greater insight from the environmental data % deluge.  These
new tools will not replace existing approaches, but rather complement them.

% Increasingly, environmental data are distributed in multiple locations, and
are frequently updated % (often several times a day).  Interactive exploration
of such large, dynamic, distributed data is therefore % a strong technical
challenge.  Data infrastructures must support the ability to quickly extract %
subsets of large data, and any data processing must happen quickly, with the
minimum of overhead.  % Web services are therefore playing an
increasingly-important role, and there is much interest in % cloud technologies
to provide the necessary scalability at the back end.  Grid technologies, being
essentially batch-mode % systems, are less suitable for this kind of application
as the scheduling overhead tends to dominate % the processing time for simple
jobs, seriously hampering interactivity.

% Performance is therefore a key challenge, exacerbated by the dynamic nature of
the data, which makes % it hard to achieve scalability through caching (see
section~\ref{sec:dynamicdataweb}).  [Note: could probably % merge material from
that section here.]  Security is another serious concern: many environmental
datasets % are held under access control.  In a distributed environment, it is
common for machines to access data % on behalf of end users (e.g., a processing
service might download its input data from a remote store).  % Finding a means
for the user to delegate his/her authority to a multi-web-service infrastructure
is % a key current challenge.  The MashMyData project
(http://www.mashmydata.org) is investigating two solutions, based % respectively
on Grid proxy certificates and OAuth.

% \subsubsection{quick answers from Jon}

% 1.  What is the purpose of the application?

% We are actually concerned with a class of applications (highly % interactive,
graphical ones) rather than a specific one.  But % generally, the applications
involve the intercomparison of different % environmental datasets for purposes
such as model validation, ground % trothing of observations and data
assimilation.

% 2.  How is the application used to do this?

% The applications often take the form of graphical tools that share a lot in
common with Geographic Information Systems (GIS). However these applications are
much better suited to scientific data, which are large and multidimensional.
The general vision is of a map-based interface, onto which different datasets
can be overlain.  A processing step may be initiated by rubber-banding an area
of the map and selecting from a list of algorithms that process data within the
selected area, perhaps calculating statistics of the data.

% 3.  What infrastructure is used? (including compute, data, network,
instruments, etc.)

% Usually the applications are driven by data served through web services of
various kinds.  These web services may be 'fed' by data coming from instruments
or from computer simulations.  A key challenge is how to integrate compute
services with these data services in an efficient, easy-to-use and transparent
way, allowing fast data processing.  The processing algorithms are generally
quite simple and so it is frustrating to succumb to the high overhead of some
scheduling systems.  The infrastructure probably has more in common with the Web
(or perhaps cloud) than the Grid.

% 4. What dynamic data is used in the application?  % (a) What are the types of
data, (b) What is the size of the data set(s)?

% Dynamic data are only sometimes used, but may arise from real-time feeds from
instruments or from looking at live results from a model.  Data from instruments
are small 1~MB-100~MB but model output may be very big (GB-TB).

% 4.  How does the application get the data?

% Generally through web services.  These services are designed to allow
server-side subsetting of large datasets, attempting to minimize the amount of
unwanted data traveling across networks.

% 5.  What are the time (or quality) constraints on the application?

% We try to aim for near-real time interaction with the data, i.e the user
should not be waiting more than a few seconds for some kind of response to a
request.  We have largely solved this problem for simple visualization, but for
linking data access and processing we are not there yet.

%\subsection{Dynamic Environmental Data on the Web} \label{sec:dynamicdataweb}

% Key contents now merged into above section.  Specifics on the NERC VO % are
probably not relevant at the moment as they are not yet implemented.

% \jhanote{I have moved this from erstwhile \S 2. Jon: please either (i) %
convert into a stand-alone application, or (ii) merge with the % ``Interactive
Exploration of Environmental Data'' example}

% \katznote{summary below assigned to Yogesh}

% 1. {\bf What does the application do? What problem does it solve?}

% The NERC\footnote{UK Natural Environment Research Council} Virtual %
Observatory\footnote{http://www.nerc.ac.uk/research/programmes/virtualobservatory/}
% is a new initiative to create a pilot cloud-based environment in which %
different practitioners will discover, visualize and process data % relating to
the UK's system of rivers and soils.  It will provide % access to many different
kinds of distributed hydrological data, % including sensor data and numerical
models, many of which produce % rapidly-changing, dynamic information.

% 2. {\bf How does it do it?}

% There are many scientific % challenges to be faced, notably including the
coupling of the models % and observations; this includes the possibility that
sensors can be % dynamically tasked in response to model forecasts, or that
models may % adapt dynamically to new information from sensors.

% 3. {\bf What are the big data aspects?}

% Not known.

% 4. {\bf What are the distributed aspects?}

% Hydrological data, % including sensor data and numerical models, are
distributed.  % Principles of the % Web and ``linked data'' may help to solve
some of the challenges in % informatics, and % Web Services are commonly used to
distribute data through HTTP.

% 5. {\bf What are the dynamic aspects?}

% Sensor data and numerical models produce % rapidly-changing, dynamic
information.  % A key characteristic of dynamic data, of course, is that it
changes with % time.  It is of great importance for the user to be confident
that she % is accessing the correct version of the data; this will often be the
% most up-to-date data, or may be a specific historical version.  In a % Web
environment, servers and proxy servers (which are sometimes beyond % the control
of the data provider or the data user) may retain caches % of data in a strategy
to reduce server load and network traffic.  For % this reason, the HTTP protocol
provides mechanisms for defining expiration % times for data resources,
specifying when a resource ought to be % cleared from the cache.

% In the environmental and geospatial communities the Open Geospatial %
Consortium specifications are being widely adopted alongside existing %
protocols such as OPeNDAP.  Unfortunately these protocols, although % built atop
HTTP, do not make it easy for this versioning to be % implemented correctly.
The OPeNDAP protocol does not have the concept % of a dataset version, meaning
that clients have no reliable or % efficient means to detect that a dataset has
changed.  The OGC % protocols in general %[TODO: look into this], % \emph{do}
provide % versioning at the level of the service endpoint, but provide no %
information on when a resource should be considered expired.  Clients % are
therefore forced to poll the server to check for updates.

% Many technical challenges will remain regarding the % handling of dynamic
data, including keeping careful records of its % provenance in order to record
the steps taken in making a decision.

% 6. {\bf What else is important?}

\subsection{Power Grids \label{power}}

%\katznote{from Yogesh}
%
%\katznote{summary below assigned to Yogesh}
%\subsubsection{new summary}

% 1. {\bf What does the application do? What problem does it solve?}

%\katznote{from answer to question 1}

Demand response (DR) optimization~\cite{ferc2011DR} in smart power grids deals
with ensuring an adequate supply of electricity by targeted curtailment of power
load by consumers during periods of peak load to avoid blackouts. One
application used for DR optimization is power demand forecasting at coarse and
fine temporal and spatial granularities to allow load curtailment operations to
be triggered in advance of a demand-supply mismatch situation.  This is a {\em
infrastructural application}.  It begins after data generation, and includes the
data transformation \& decisions, storage, and processing stages.

\subsubsection*{Application description} %2. {\bf How does it do it?}

%\katznote{from answers to questions 2,3,5,6 below, also address
cyberinfrastructure (where does the app run) -- perhaps have a diagram here of
what happens in the app}

Power usage information is aggregated from smart meters at consumer premises
into the utility's data center~\cite{Simmhan:buildsys:2011,simmhan2011TR}. Data
is collected at approximately 15-min intervals and transmitted to the utility
through cellular and wireless networks using proprietary protocols. Complex
event pattern matching over the power usage events and other information source
such as weather and scheduling data accessed using web services help perform
near term load forecasting. Long term power forecasting at the city-scale or at
a smaller micro-grid scale (e.g. industry complex, university campus) is done
using power usage models built using machine learning algorithms running at the
utility's or micro-grid's datacenter using historical usage
data~\cite{Aman:dddm:2011}. These applications run on private clouds and are
composed using DAG, Complex Event Processing and MapReduce
structures~\cite{zhou:2012:scepter}.

 \subsubsection*{Big data aspects} %3. {\bf What are the big data aspects?}

%\katznote{from answer to question 4b below}

Constructing forecast models using machine learning requires access to
historical data that can grow to large sizes~\cite{Yin:mapreduce:2012}. Several
hundred TB per year of smart meter power usage data can accumulate for a large
city with millions of customers. Other data used in predictive modeling, such as
event/people schedules and building/facility features, are smaller, on the order
of GBs per year. In future, social network data from consumers may also be mined
to learn about power usage patterns. These can also be on the order of GBs.

Real-time forecasting uses smaller data---events on the order of 1~KB, but
potentially from millions of sources---that are passed to complex event
pipelines or predictive models that run continuously. These pipelines can
process GBs of data per hour.

 \subsubsection*{Distributed aspects} %4. {\bf What are the distributed
aspects?}

%\katznote{from answer to question 3 below}

Data from domains such as power systems, building management systems, weather
and traffic, and social networks need to be integrated. While the data are
generated and/or stored in distributed locations, they are typically aggregated
at the utility's datacenter for analysis. Limited processing may happen locally
at the smart meters or local micro-grids with the results pushed as events to
the utility.

The applications themselves run on private cloud platforms across hundreds of
VMs. Communication between the utility and smart meters is bidirectional,
allowing control and pricing signals to be sent from the datacenter to the smart
meters. Integrated data will also need to be shared with external applications
running on, say, mobile platforms or portals, and with regulatory agencies.

 \subsubsection*{Dynamic aspects} %5. {\bf What are the dynamic aspects?}

%\katznote{from answer to question 4a below}

The data rates for events published by the sensors may change, either passively
due to local measurement limitations or actively throttled by the utility to
meet the application's accuracy needs/resource
constraints~\cite{Simmhan:sciencecloud:2011}. The data from smart meters can be
sent synchronously as they are generated or in batch mode once a day to conserve
bandwidth. The application processing itself is done elastic resources, scaling
up or down to meet latency requirements. Some of the forecasting models can also
decide to listen to additional data streams on-demand when higher accuracy is
required or special situations are encountered.

 \subsubsection*{Other important issues} %6. {\bf What else is important?}

%\katznote{from answers to questions 7,8 below, also address programming systems
(how was the app written; what tools does it use)}

Semantic information integration is being performed using an ontology composed
together from individual domains~\cite{Zhou:itng:2012}. Machine learning models
using Matlab, Weka and Hadoop are being used. Complex event processing is using
the ``Siddhi'' CEP engine with semantic support added~\cite{siddhi}. Stream
processing on Eucalyptus using IBM Infosphere Streams is being performed.

% \subsubsection{old material follows}

% Consider the energy informatics domain and smart power grids in %
particular. For data continuously arriving from 2 million smart meters % in
large city, households will soon need to be continuously analyzed % in order to
detect impending peak power usage in the smart power grid % and notify the
utility to respond by either spinning up additional % power sources or by
triggering load curtailment operations to reduce % the demand. This closed loop
cyber-physical application, modeled as a % workflow, needs to combine streaming
data arriving from sensors with % historic data available in file archives along
with structured % collections of weather forecast data that helps the large
scale % computational model make an energy use prediction in near real time.

% A workflow framework that supports this data model diversity, % including
streaming data, structured collections and files, and the % ability to execute
reliably and scalably on resources is required.

% In general, smart grids - respond to varying data loads, spinning up %
additional power sources or triggering load curtailment operations.  % E.g., of
a closed-loop cyber-physical system modeled as a workflow % needs to combine
streaming data arriving from sensors with historic % data available in archives,
along with structured collections of % weather forecast data. Need a workflow
model that supports such % requirements. Taken from ref (Towards Reliable
Performant Workflows % for Streaming Applications on Cloud Platforms).

% {\it Requirements: } (i) Applications will have to span desktops, %
workstations and clouds, (ii) streaming data models and stream % programming
abstractions that are more than TCP sockets, (iii) % reliability of VMs hosting
the workflow and avoid costly duplicate % movement of the same logical stream
(see paper in WORKS by Zinn et % al.)

% Scientific workflows are commonplace in eScience applications. Yet, % the lack
of integrated support for data models, including streaming % data, structured
collections and files, is limiting the ability of % workflows to support
emerging applications in energy informatics that % are stream oriented. This is
compounded by the absence of Cloud data % services that support reliable and
performant streams.  Need a % scientific workflow framework that supports
streams as first class % data, and is optimized for performant and reliable
execution across % desktop and Cloud platforms.

% \subsubsection{old version of answers}

% 1. What is the purpose of the application?

% Demand response optimization (DR) in smart power grids. Specifically, %
accurate forecasting of power demand at coarse and fine temporal and spatial %
granularities, and launching power load curtailment operations when a peak %
load situation is forecast.

% 2. How is the application used to do this?

% Several approaches exist for performing DR.

% (1) Perform realtime aggregation and analysis on power usage information %
arriving into the utility's data center from smart meters installed at %
consumer premises for low latency detection of current (peak) load;

% (2) Perform complex event pattern matching and mining over an event cloud %
that includes event streams from multiple information source types to detect %
pattern that forecast a peak load event or opportunity of energy % conservation.

% (3) Construct building and campus energy models using machine learning to %
predict future energy use.

% 3. What infrastructure is used? (including compute, data, network, %
instruments, etc.)

% Compute: 128-core Eucalyptus private cloud; Public Cloud (TBD)

% Data: Local disk, others (TBD)

% Software/Services: Eucalyptus Cloud fabric, IBM InfoSphere Streams, %
Esper/Siddhi Complex Event Processing (CEP) system, MatLab toolkit, Workflow %
system (TBD), custom research prototypes.

% Network: Ethernet, others (TBD) % Instruments: Smart meters (consumer power
usage), Power sensors (fan speed, % room temperature, thermostat setting,
appliance state, etc.), Facility % sensors (people counter, ambient light, etc.)

% 4. What dynamic data is used in the application?  % (a) What are the types of
data, (b) What is the size of the data set(s)?

% Current and expected data \& rates:

% Smart meter power usage data: 1.4M Consumers x $\sim$1 KB event/min
($\sim$700~TB/yr (max))

% Weather: KB/hr

% Traffic: KB/min

% Consumer Calendar Schedules: MB/hr

% Facility Calendar Schedule: MB/day

% Twitter/Facebook: MB/min

% 5. How does the application get the data?

% Web services, file transfer, proprietary network protocols

% 6. What are the time (or quality) constraints on the application?

% O(secs) to O(hrs) for load forecasting and response, depending on accuracy %
of prediction and how far ahead the prediction is. There is a tradeoff in % the
quality constraints based on available power capacity, data, % infrastructure
resources, granularity of prediction and response, etc. The % absolute quality
constraint is that the peak power load must {\bf not} exceed % the available
power capacity. So the power forecast, load curtailment % activation, and the
impact of the curtailment have to be in a closed loop % cycle whose equilibrium
is maintained.

% 7. How much diverse data integration is involved?

% 8. How diverse is the data?

% Data from domains such as power systems, building management, weather and %
traffic, and social networks need to be integrated. We use semantic %
information integration with an ontology composed together from individual %
domains.

% 9. Please feel free to also talk about the current state of the % application,
if it % exists today, and any specific gaps that you know need to be overcome.

% Currently, application for constructing Building and Campus energy models %
using machine learning and forecasting power usage exist. These over power %
usage data, building construction features, academic schedule, weather, etc.  %
Uniform information integration is currently underway to experiment with %
multiple models.

% Initial versions of stream processing on the private Cloud over campus level %
smart meter data using information from 100+ buildings exist with adaptive %
logic for stream rate control.

% Research into semantic CEP is underway, with domain patterns being defined.  %
Gaps include scalable stream and CEP processing over clouds (including % advance
resource provisioning and task scheduling), running ``online'' % workflows with
support for stream inputs that run continuously (to improve % forecast models
and make predictions), resource selection based on data % privacy and latency
constraints, and reliable execution with graceful % degradation of critical
analysis operations.

\subsection{Fusion (ITER) \label{fusionScott}}

%\katznote{from Scott Klasky}
%
%\katznote{summary below assigned to Dan}
%\subsubsection{new summary}

% 1. {\bf What does the application do? What problem does it solve?}

%\katznote{from answer to question 1}

With ITER~\cite{iter} scheduled to operate before 2020, plasma fusion has become
one of the highest priority research items for the US DOE. Scalability of
first-principles fusion codes is an important part of fusion research needed to
obtain predictive capability of plasma performance with first-principles physics
fidelity. The goal is to build a realistic full-distribution kinetic code, in
realistic ITER geometry, combining the strengths of each of the existing
gyrokinetic codes (GEM, GTS, XGC, and GTC), and utilizing the most up-to-date
computer science, hardware design, and applied mathematics technologies
developed. This {\em traditional application} is intended to be capable of 1)
discovering macroscopic magnetohydrodynamics activity in burning plasmas, 2)
understanding Energetic particle effects in fusion reactors, 3) understanding
the heating in a fusion reactor, and 4) understanding the interaction of the
plasma at the material boundary. This applications just has the data processing
stage, along with associated data movement.

\subsubsection*{Application description} %2. {\bf How does it do it?}

%\katznote{from answers to questions 2,3,5,6 below, also address
cyberinfrastructure (where does the app run) -- perhaps have a diagram here of
what happens in the app}

The applications that are currently run (XGC1, XGCP, GTC, GTS) run on all of the
DOE and TeraGrid leadership class facilities, and require very large resources
for the modest problems that are currently being solved. In the next generation
(exascale), not only will all of the computing power that will be available at
the time be needed to understand ITER, all of the data that will be generated
during the simulation also will have to be understood. The community's
techniques have been to examine and understand the data in situ, to reduce the
total amount of data produced in the simulation.

 \subsubsection*{Big data aspects} %3. {\bf What are the big data aspects?}

%\katznote{from answer to question 4b below}

Today the team has produced O(100) TB per week for a simulation using the ADIOS
middleware~\cite{adios} incorporated in the simulations. Using staging
techniques, they believe that they will only produce O(10) PB on an exascale
machine.

 \subsubsection*{Distributed aspects} %4. {\bf What are the distributed
aspects?}

%\katznote{from answer to question 3 below}

The team always runs on all available resources, so if they have computer time
across the country, then their data will be distributed, and so will the
computation.

 \subsubsection*{Dynamic aspects} %5. {\bf What are the dynamic aspects?}

%\katznote{from answer to question 4a below}

The primary aspect that is dynamic is that the set of computing resources that
are used for a given run are based on those that are available at the time of
that run.  As part of this, data is streamed from one running simulation to
another, and is transformed in-flight.

 \subsubsection*{Other important issues} %6. {\bf What else is important?}

%\katznote{from answers to questions 7,8 below, also address programming systems
(how was the app written; what tools does it use)}

The applications run on a distributed set of leadership-class facilities, using
advance reservations to co-schedule the simulations.  Each code reads and writes
data files, using ADIOS and HDF5.  Files output by each code are transformed and
transferred to be used as inputs by other codes, linking the codes into a single
coupled simulation.  Because the overall data generated is so large, it can not
all be written to disk for post-run analysis, so in-situ analysis and
visualization tools are being developed.

%\subsubsection{old material follows} %1. What is the science/engineering
problem? What is the purpose of the %application?
%
%With ITER scheduled to operate before 2020, plasma fusion has become %one of
the highest priority research items for DOE. Scalability of %first-principles
fusion codes is an important part of fusion research %needed to obtain
predictive capability of plasma performance with %first-principles physics
fidelity. We want to build a realistic %full-distribution kinetic code, in
realistic ITER geometry, combining %the strengths of each of the existing
gyrokinetic codes (GEM, GTS, %XGC, and GTC), and utilize the most up-to-date
computer science, %hardware design and applied mathematics technologies
developed. The %application that we will develop will be capable of 1)
discovering %Macroscopic Magneto Hydro Dynamics activity in burning plasmas, 2)
%Understand Energetic particle effects in fusion reactors, 3) %Understand the
heating in a fusion reactor, and 4) Understand the %interaction of the plasma at
the material boundary.
%
%2. How will the application solve the problem?
%
%(a) Which of the following describes the application
%
% i) large (exascale) data and large (exascale) computing (e.g., data mining,
analytics, bioinformatics)
%
% ii) large (exascale) data and modest computing (e.g., search)
%
% iii) modest data and large (exascale) computing (e.g., simulation/modeling)
%
%The applications we currently run (XGC1, XGCP, GTC, GTS) run on all of the
leadership class facilities, and require very large resources for the modest
problems which we are currently solving. In the next generation (exascale) we
will not only need all of the computing power that will be available at the time
to understand ITER, we will also need to understand all of the data that will be
generated during the simulation. Our techniques have been to examine and
understand the data in situ, to reduce the total amount of data produced in the
simulation. Today we have produced O(100 TB/week) for a simulation using our
ADIOS middleware incorporated in our simulations. Using our staging techniques,
we believe that we will only produce O(10 PB) on an exascale machine. (I don't
know if this is considered modest dat, or extreme size data)?
%
%(b) Can you quantify the amount of computing vs the amount of data?
%
%1 PB of data -- 10 PB of data, for an exascale size computation run in O(100
hours).  %Today we achieve about 6\% of peak with excellent scalability to 220K
%cores on the Cray XT systems. It is %difficult to predict how we will run on
%accelerators, and next generation machines.
%
%3. What infrastructure will be used? (including compute, data, %network,
instruments, etc.)
%
%For Validation, we will need to compare our data from the computer to %ITER (or
current fusion reactors). Typically the amount of data used %in comparison is
tiny by comparison (people only look at MB's form %experiments).
%
%(a) What will be distributed?  data, compute, both?
%
%We always run on all available resources, so if we have computer time %across
the country, then our data will be distributed, and so will our
%computation. ITER will have data that will be distributed, but it %still
unknown.
%
%4. What data be used in the application?
%
%I don't understand this question? We will input data (from perhaps a restart),
and this data is typically in the ADIOS-BP file format for large data. Smaller
data sets are typically stored in HDF5, and the smallest are stored in MDS+.
%
%(a) What are the types of data
%
%(b) What is the size of the data set(s)?
%
%Depends on what we want to understand --- For GTS/GTC we have typically %looked
at the particle data to compute fluxes, etc. and this data can %be large, but
due to the Monte Carlo error, and ways to generated 5D %distribution functions
from the particles, we reduce the particles to %a manageable amount (300 GB per
time step x 1 K time steps). Our %current I/O is capable of reading this in $<$
100 s per time step, which %allows us to analyze the data in about 1-2 days.
%
%(c) Will any data be dynamic?
%
%Don't understand the question. Features change and are dynamic. The physics
will get more complex as we continue to evolve our codes, and will be more
dynamic in nature.
%
%5. How will the application get the data?
%
%Reading the data from disk --- (Perhaps I don't understand the question).  It
will also couple to other codes, and the coupling will be in-memory data
exchange.
%
%6. Are there time (or quality) constraints on the application?  (If so, what
are they?)
%
%We always need to run as fast as possible since we have limited computing time
(our team does have about 100M hours FY 11, but this still puts severe
restrictions on what we run, so the faster we can run, the more science we can
get done.
%
%7. Please feel free to also talk about the current state of the application, if
it exists today, and any specific gaps that you know will need to be overcome.
%
%Our biggest gaps will be in accelerating the code on next generation exascale
platforms, and in processing as much data as possible in situ.

\subsection{Industrial Incident Notification and Response \label{Kees}}

%\katznote{from Kees}
%
%\katznote{summary below assigned to Yogesh}

%\katznote{Yogesh, some parts are repetitive - can you fix them?}

%\subsubsection{new summary}

% 1. {\bf What does the application do? What problem does it solve?}

%\katznote{from answer to question 1}

Applications such as large scale crisis management, maritime security, disease
and pollution monitoring, and environmental control need to fuse large amounts
of heterogeneous data to form virtual information sources that mitigate
noise~\cite{emergency1,emergency2,emergency3,emergency4}.  The area of greater
Rotterdam in the Netherlands is a chemical hub and houses more than 15\% of the
Dutch population, making safety and security a major concern.  Some leakage of
chemicals into the air is an everyday, unavoidable event.

The environmental protection agency (EVA) maintains a command and control call
center staffed by chemical hazard experts to (1) monitor the air quality and
sewage, (2) in case of an incident locate the source, (3) identify the threat
and (4) provide the public authorities, the emergency response organization, and
the public with situation awareness and suggested actions.  The information that
is shared is often incomplete, uncertain and inconclusive. Knowledge within the
EVA as well as from outside experts and industries must be accessed to
meaningfully use that information.  This subsection describes the {\em
traditional application} that the EVA uses for these tasks.  The application
starts after data generation, and includes data transformation \& decisions,
storage, and processing.

\subsubsection*{Application description} %2. {\bf How does it do it?}

%\katznote{from answers to questions 2,3,5,6 below, also address
cyberinfrastructure (where does the app run) -- perhaps have a diagram here of
what happens in the app}

For monitoring and incident detection, continuous streams of data are processed
and scanned for anomalies from a small number of physical sensors and air
analysis data (gas-chromatographs outputs).  If an incident or an anomaly is
detected, new streams of data become available from the general public and field
inspectors and must be processed. Static information on weather and about
chemicals are also needed.  These may be stored in different places, access may
not be granted automatically, and validity and update frequencies may not match
expectations. Thus specialized pre-processing is needed.

Different knowledge-based processing models must be used account for uncertainty
in the data and sensor behavior, and provide robust fusion (processing) for
situation awareness and decision support. Some confidential data must be
processed \emph{in situ}. New intermediate and final data that are generated may
have different retention characteristics and need to be stored and made
accessible for visualization and analysis in different workflows.

Processing in done in semi-real-time or in batch mode processing on a small,
private computer infrastructure. Therefore, scalability is lacking and the
dynamics of the workflows is kept under strict human control to match the
capabilities of a set of static resources.

 \subsubsection*{Big data aspects} %3. {\bf What are the big data aspects?}

%\katznote{from answer to question 4b below}

The data sizes are not very large.

 \subsubsection*{Distributed aspects} %4. {\bf What are the distributed
aspects?}

%\katznote{from answer to question 3 below}

A number of special data and sensor interface processes will be distributed over
various platforms and fixed in place.  A low frequency stream of reasonable
fidelity data from a small number of physical sensors and an even lower
frequency stream of air analysis data (gas-chromatographs outputs) is
continuously trickling in.  Low fidelity human provided data from local people
complaining to the call center, field inspectors calling in with observations
and measurements, information from social media, pictures from cell phones and
emails sent to a centralized email account are entered manually into a local
databases.  Static information such as current weather conditions and forecast
and information about chemicals are also needed.  These data are made available
via a distributed shared data space.  Some confidential data must be processed
remotely and only aggregated results communicated back to other entities. The
information processing algorithms are packaged as autonomous software agents
that are managed by a multi agent system infrastructure (middleware) and humans
(via GUIs). Processing algorithms need to access multiple concurrently and
remote data stores.

 \subsubsection*{Dynamic aspects} %5. {\bf What are the dynamic aspects?}

%\katznote{from answer to question 4a below}

Both the data and the processing are dynamic in nature. Data that cannot be
moved due to confidentiality have to be processed remotely by moving
applications to them.  A process integration framework will use agents that
provide autonomous and composable resources to dynamically construct workflows
for data driven processing.  If an incident happens or an anomaly is detected,
it triggers new streams of data about other anomalies that must be processed.
Another source of dynamic behavior is when the event is escalated and additional
management processes come into play, each requesting and adding their own sort
of processing to the mix.

 \subsubsection*{Other important issues} %6. {\bf What else is important?}

%\katznote{from answers to questions 7,8 below, also address programming systems
(how was the app written; what tools does it use)}

Quality of service aspects that are important for real world and real-time
applications are the dynamics of scheduling from an application perspective,
multi independent level of security support, processing service discovery
support, autonomous reconfiguration support and logging-for-traceability.  Most
of the dynamic data that these applications use has a short validity period,
i.e., it must be processed now or we process the next update coming in. The data
is uncertain, i.e., it may have dynamically changing signal to noise ratio and
although expected at some base frequency, may not be available for some period
of time.

% \subsubsection{old material follows}

% Efficient Processing in Real-Time Complex Distributed Information Systems

% Dr. Kees Nieuwenhuis, Thales Nederland, D-CIS Lab

% Acknowledgement: The use case and working prototype system discussed here are
% realized with the help of the DIADEM project grant from the EC (FP7 grant
224318).

% Introduction

% Contemporary net-centric applications, such as large scale crisis management,
% maritime security, disease and pollution monitoring, environmental control and
% management etc. cab greatly benefit from the fusion of large amounts of
heterogeneous % information, coming from static and mobile sensors as well as
intelligence obtained % from human observers, various databases and the World
Wide Web. This is because % new AI based techniques are available that help to
combine multiple information % modalities and thereby increase the amount of
information for (real-time) fusion % systems, which in turn can mitigate the
impact of noise and reduce ambiguities and % overcome the unavailability of
specific information sources.

% However, the compute and networking requirements for this new type of
applications % that we need to build are significantly different from the more
traditional massive data % processing or distributed computing applications that
are known and deployed in % professional business and scientific domains today.

% Use case description: DIADEM

% The area of greater Rotterdam in the Netherlands is a highly industrialized
area with the % largest harbor in Europe, very many chemical, petrochemical,
pharmaceutical and other % processing plants. Economically, it is the most
important area of the country. In addition, % the same area is home to more than
15\% of the Dutch population. Safety and security % are therefore a major
concern, especially with the type of hazards involved and % because major
evacuations are just not an option. Some leakage of chemicals into the % air
however, is an every day event and cannot be avoided. Therefore, the %
environmental protection agency (EVA) is actively monitoring air quality and
incidents % (that by law must be reported immediately) from a command and
control center which % includes a dedicated call center for the population. They
employ a number of chemical % and other hazard experts that can be deployed in
the field for inspection (prevention) % and measurements. Due to the sheer size
of the area and the (so far) inadequate % techniques and resources for
continuous data gathering, data sharing, data processing % and sensor
management, there are very few sensors in the area that the EVA has % continuous
access to.

% The tasks of the EVA are to 1) monitor the environment (mainly air quality and
sewage), % 2) in case of an incident locate the source, 3) identify the threat
and 4) provide the % public authorities, the emergency response organization and
the public with situation % awareness and possible courses of action. For that,
they must liaise and coordinate % with other expert organizations (health and
hazard experts, public order and safety % experts, the harbor authority, traffic
experts, etc.) and share and exchange information % and knowledge. An important
aspect of the latter is that the information that is available % is always
incomplete, uncertain and no single element is conclusive by itself. In order to
% process and use that information, inside and outside knowledge must be
accessed.

% Dynamic versus static data

% For continuous monitoring and incident detection continuous streams of data
are % necessary that need to be processed and scanned for anomalies. Typically,
a low % frequency stream of reasonable fidelity data from a small number of
physical sensors % and an even lower frequency stream of air analysis data
(gas-chromatographs outputs) % is trickling in. A much larger amount of data
from private sensors from hazard sources is % potentially available, but
currently not accessible. If an incident happens or an anomaly % is detected,
new streams of data become available and must be processed, such as low %
fidelity data from people in the area calling the call centre with complaints,
field(ed) % inspectors calling in with results of expert observations and
additional high fidelity % measurements, information from social media and maybe
pictures from cell phones % from individuals in the area (solicited or
not-solicited). In case of an incident, suddenly % also other more static
information such as current weather conditions and forecast and % information
about chemicals (hazards, risks, air-borne or water-borne behavior, health %
aspects, neutralizing methods and agents, etc.) is needed for further
processing. This % other information may be stored in different places and
access may or may not be % granted automatically and formats may be special and
validity and update frequencies % may not match expectations or needs and thus
demand for specialized pre-processing % emerges.

% Use of data

% The real-time data is mainly stochastic in nature. The fidelity is variable,
which means % that different knowledge based processing models must be used
concurrently to take % into account not only the uncertainty in the data /
measurements itself, but also to take % into account sensor behavior. Most of
the data are observations of the effects of the root % cause and therefore
robust fusion (processing) with the help of knowledge based % models is
necessary to provide situation awareness and decision support. Some fusion %
requires data that is confidential in nature or commercial in confidence and may
not % leave the place where it is stored. Therefore the special event based
processing % algorithms must be brought to the data and only aggregated results
can be % communicated to other processing entities. This means that both the
data and the % processing are dynamic in nature, which the application and its
computer infrastructure % must support. Relationships with static or semi-static
(e.g., weather) data are often built % in to the processing algorithms, which
means that these data sources need to be % accessed during real-time processing
also. This requires multiple simultaneous % accesses to remote data stores. The
processing will also produce new intermediate and % final data, sometimes with
different retention characteristics, which need to be stored % and made
accessible to special visualization and analysis applications in different %
workflows.

% Infrastructure use

% In current applications the three main network technologies used are the
telephone and % cell phone network and a centralized computer network with
internet connection (for % emails). Data is filtered and entered manually into
locally housed databases and files % and processed in semi-real-time or batch
mode processing on a private and limited % computer infrastructure. Therefore
scalability is lacking and dynamics of workflows is % kept under strict human
control to match the capabilities of a set of static resources.

% The need however, is to have more scalable solutions where resource usages can
% actually follow demand from an escalating (and de-escalating) incident and
information % availability point of view. This is complemented by the need to
support the dynamic % creation of workflows in a multi-stakeholder (and no
longer single stakeholder) setting.  % In addition, security is an important
aspects that drives a separate set of requirements % which impact data
availability and processing availability that must comply with a % dynamic
network centric processing model. That model must support secure access at % the
data-item level and without restrictions at the network or compute platform
level. A % mixture of the Cloud and Grid approaches may offer a solution here.

% Use of dynamic data

% The use of dynamic data has two origins: the first is the real world situation
that may % begin with a small scale event such as a minor gas leak. If such an
event is undetected % at first, it may escalate and trigger more and more
disturbances that add data about % different other anomalies to the data
streams. The second source of dynamic behavior % is that with the escalation of
the event, more and more special repression an % management process come into
play, each requesting and adding their own sort of % processing to the mix.

% How does the application get the data?

% In a typical application a number of special data and sensor interface
processes will be % part of the development. Whether packaged as software agents
or some other form of % software process, they will normally be distributed over
various platforms and fixed in % place. There task is to make the data available
via a distributed shared data space % (used here as a logical concept that can
be implemented in different ways) that % abstracts from physical storage. The
information processing algorithms are packaged % as autonomous processing
entities such as software agents, that are managed by a % multi agent system
infrastructure (middleware) and humans (via GUIs).

% QoS aspects

% Quality of service aspects that are important for real world and real time
applications are % the dynamics of scheduling from an application perspective,
multi independent level of % security support, processing service discovery
support, autonomous reconfiguration % support and
logging-for-traceability. Because of the nature of the application aspects %
like sustained communication bandwidth, processor availability,
checkpoint-restart etc.  % are not important. Most of the dynamic data that
these applications use has a short % validity period, i.e., it must be processed
now or we process the next update coming in.  % The data is uncertain, i.e., it
may have dynamically changing S/N ratio and although % expected at some base
frequency, may not be available for some period of time.

\subsection{MODIS Data Processing \label{modisKeith}}

%\katznote{from Keith Jackson}
%
%\katznote{summary below assigned to Yogesh}
%\subsubsection{new summary}

%\katznote{Yogesh, some parts are repetitive - can you fix them?}

% 1. {\bf What does the application do? What problem does it solve?}

%\katznote{from answer to question 1}

This {\em traditional application} allows the coordinated usage of
MODIS~\cite{modis} satellite imagery, and its integration with ground based
sensors and models for environmental applications.  It consists of data storage
and processing stages.

\subsubsection*{Application description} %2. {\bf How does it do it?}

%\katznote{from answers to questions 2,3,5,6 below, also address
cyberinfrastructure (where does the app run) -- perhaps have a diagram here of
what happens in the app}

The application~\cite{modisAzure,modis:ipdps:2010} is used to handle the processes of downloading
data from different sites and integrating them. The integration consists of
reprojecting the data from one map projection to another, spatial and temporal
resampling, and gap filling. The application itself is a loosely-coupled data
pipeline that coordinates these processes.

In addition, scientists can submit Matlab scripts for further custom processing
on subsets of the data that has been integrated. This includes some
summarization processing and plotting routines.  The only time constraints are
on the debugging cycle for debugging the scientist-supplied reduction
codes. These are purely driven by the users' patience.

The application has been implemented in both the Windows Azure cloud
environment~\cite{win_azure} and a standard HPC environment.

 \subsubsection*{Big data aspects} %3. {\bf What are the big data aspects?}

%\katznote{from answer to question 4b below}

There is approximately 25~TB of data for one global year.  The main data is the
various MODIS satellite data products, of which 11 are used by this
application. These are in approximately 3,000,000 files for one global year. A
smaller set of sensor data from the fluxnet data network is used as input to the
evapotransporation calculation, along with a set of 20-30 files containing
various metatdata like climate regions used by the calculation.

 \subsubsection*{Distributed aspects} %4. {\bf What are the distributed
aspects?}

%\katznote{from answer to question 3 below}

 The data is collected from a wide variety of FTP servers managed by different
groups and brought into the cloud/HPC environment. Custom software tooling
manages and validates this download process. Reliable cloud message queues are
used to manage the coordination between the different components in the
multi-stage pipeline. Simple NoSQL tabular storage is used for logging and fault
recovery. Data is stored in cloud blobs and NoSQL table storage. Application
scripts on the user's desktop can be uploaded into the cloud as post-processing
routines and analyses on the integrated data.

 \subsubsection*{Dynamic aspects} %5. {\bf What are the dynamic aspects?}

%\katznote{from answer to question 4a below}

 The data is currently static and the application acquires the data from a set
of FTP servers on a schedule or trigger. But as the pipeline is expanded to
handle other data, the usage of real-time streaming sensor data is also
expected. There is infrastructure dynamism through the use of cloud virtual
machines, acquired and released on-demand, to execute tasks in the pipeline and
to execute the user's application scripts. While the pipeline structure itself
is static, there is a limited form of application dynamism through the use of
user specified scripts pushed into the cloud for execution over the integrated
data.

 \subsubsection*{Other important issues} %6. {\bf What else is important?}

%\katznote{from answers to questions 7,8 below, also address programming systems
(how was the app written; what tools does it use)}

The data is in one of two map projections, and multiple temporal and spatial
resolutions. The fluxnet sensor data that is used required significant data
integration, but was handled in a separate project.

The quality constraints are driven by the needs of the scientific reduction
application. Significant effort is required to ensure that the data handled to
the scientific application meets these requirements.
% \subsubsection{old material follows}

% 1. What is the purpose of the application?

% The purpose is to allow the coordinated usage of MODIS satellite imagery, and
to allow its integration with ground based sensors, models, etc.

% 2. How is the application used to do this?

% The application is used to handle the processes of downloading data and data
integration. The data integration consists of the ability to reproject the data
from one map projection to another, spatial and temporal resampling, and gap
filling. The application consists of a loosely-coupled data pipeline that
coordinates this process.

% 3. What infrastructure is used? (including compute, data, network,
instruments, etc.)

% The application has been implemented in both the Windows Azure cloud
environment and a standard HPC environment. The data is collected from a wide
variety of ftp servers managed by different groups. Significant custom
infrastructure is required to manage and validate this download
process. Reliable message queues are used to manage the coordination between the
different components in the pipeline, and simple NoSql tabular storage is used
for logging and fault recovery.

% 4. What dynamic data is used in the application?

% The data is currently static, but as we expand the pipeline to handle other
data we expect to see the usage of real-time streaming sensor data.

% (a) What are the types of data

% The main data is the various MODIS data products, of which we use 11. These
are in approximately 3,000,000 files for one global year. A smaller set of
sensor data from the fluxnet data network is used as input to the
evapotransporation calculation, along with a set of 20-30 files containing
various metatdata like climate regions used by the calculation.

% (b) What is the size of the data set(s)?

% There is approximately 25~TB of data for one global year.

% 5. How does the application get the data?

% The application dynamically acquires the data from a set of ftp servers.

% 6. What are the time (or quality) constraints on the application?

% The only time constraints are on the debugging cycle for debugging the
scientist-supplied reduction codes. These are purely driven by the users'
patience. The quality constraints are driven by the needs of the scientific
reduction application. Significant effort is required to ensure that the data
handled to the scientific application meets these requirements.

% 7. How much diverse data integration is involved?

% The data is in one of two map projections, and multiple temporal and spatial
resolutions. The fluxnet sensor data that is used required significant data
integration, but was handled in a separate project.

% 8. How diverse is the data?

% 9. Please feel free to also talk about the current state of the application,
if it exists today, and any specific gaps that you know need to be overcome.

% \katznote{from talking to Keith: There is a pipeline of initial standardized
processing steps on the data.  Then, scientists can submit executables that do
further custom processing on subsets of the data, which likely include some
summarization processing (building graphs).}

%\subsection{Floating Sensors \label{floaterKeith}}
%
%%\katznote{from Keith Jackson}
%%
%%\katznote{summary below assigned to Yogesh}
%
%\katznote{Yogesh, some stuff is missing - can you fix any of it?}
%\ysnote{Unfortunately, thats all the info I have from Keith. No response to a
reminder I sent him earlier.}  %\katznote{I will talk to Alex, the PI of this
project, on Feb 2.}
%
%%\subsubsection{new summary}
%
%% 1. {\bf What does the application do? What problem does it solve?}
%
%%\katznote{from answer to question 1}
%
%This {\em traditional application} constructs real-time hydrodynamics models,
and forecasts the behavior of a complicated water system: the Sacramento %- San
Joaquin River delta.  It begins after data generation and includes data
transformation \& decisions, storage, and processing.
%
%\subsubsection*{Application description} %%2. {\bf How does it do it?}
%
%%\katznote{from answers to questions 2,3,5,6 below, also address
cyberinfrastructure (where does the app run) -- perhaps have a diagram here of
what happens in the app}
%
%The application integrates both static and dynamic sensors streaming %real-time
data with a high-resolution hydrodynamics model running on the NERSC Magellan
cloud %testbed. The model uses sensor data to initialize and shape the run, and
the model output is in turn %used to drive the real-time positioning of the
dynamic sensors. The output of this process is a map %of the current and future
behavior of the modeled watershed.
%
%Custom built dynamic sensors are deployed in the watershed along with
traditional statically placed %sensors that provide data. These sensors
communicate over a variety of different network %infrastructures, e.g., GSM,
CDMA, 802.11, wired ethernet, etc.  and the application acquires the %data over
the internet from the sensor platforms.  The application processes the sensor
data in %real-time as it is streamed.
%
% \subsubsection*{Big data aspects} %%3. {\bf What are the big data aspects?}
%
%
%%\katznote{from answer to question 4b below}
%
%Not known.
%
% \subsubsection*{Distributed aspects} % %4. {\bf What are the distributed
aspects?}
%
%
%%\katznote{from answer to question 3 below}
%
%Not known.
%
% \subsubsection*{Dynamic aspects} %%5. {\bf What are the dynamic aspects?}
%
%
%%\katznote{from answer to question 4a below}
%
%The sensor are all streaming real-time data to the application. The position of
the dynamic sensors %themselves change over time.
%
%
% \subsubsection*{Other important issues} %%6. {\bf What else is important?}
%
%
%% \katznote{from answers to questions 7,8 below, also address programming
systems (how was the app written; what tools does it use)}
%
%The first full, end-to-end, field test of the system is scheduled for May 2011.
%

% \subsubsection{old material follows}

% 1. What is the purpose of the application?

% To model in real-time, and forecast, the behavior of a complicated water
system like the Sacramento - San Joaquin River delta.

% 2. How is the application used to do this?

% The application integrated both static (Eulerian) and dynamic (Lagrangian)
sensors streaming real-time data with a high-resolution hydrodynamics model
running at the NERSC supercomputing center. The model uses sensor data to
initialize and shape the run, and model output is used to drive the real-time
positioning of the Lagrangian sensors. The output of this process is a map of
the current and future behavior of the modeled watershed.

% 3. What infrastructure is used? (including compute, data, network,
instruments, etc.)

% Custom built Lagrangian sensors are deployed in the watershed along with
traditional statically placed sensors. These sensors communicated over a variety
of different network infrastructures, e.g., GSM, CDMA, 802.11, wired ethernet,
etc. The model is running on the NERSC Magellan cloud testbed.

% 4. What dynamic data is used in the application?

% The sensor streams are all real-time streaming data.

% (a) What are the types of data % (b) What is the size of the data set(s)?

% \katznote{Keith: I need to chat with Alex to get you these answers Dan. Please
remind me if I don't get back to you on this.}

% 5. How does the application get the data?

% The application is acquired over the internet from the various sensor
platforms.

% 6. What are the time (or quality) constraints on the application?

% The application is processing the sensor data in real-time as it is streamed.

% 7. How much diverse data integration is involved?

% 8. How diverse is the data?

% \katznote{Keith: I need to chat with Alex about the data diversity also.}

% 9. Please feel free to also talk about the current state of the application,
if it exists today, and any specific gaps that you know need to be overcome.

% The first full, end-to-end, field test of the system is scheduled for May.

\subsection{Distributed Network Intrusion Detection \label{detectionJon}}

%\katznote{from Jon Weissman}
%
%\katznote{summary below assigned to Jon}

%\subsubsection{new summary}

% 1. {\bf What does the application do? What problem does it solve?}

%\katznote{from answer to question 1}

This {\em traditional application} performs detection of distributed network
intrusion and distributed denial of service attacks~\cite{jon}.  Network TCP/UDP
traffic data is analyzed/mined at each site. Potential alerts are collected
centrally and further analysis is done to see if an attack may be underway by
correlating alerts, and events are then communicated back to local sites. The
following features are extracted: 1) basic features such as source/destination
IP address pair, port number, and protocols and 2) some additional features such
as the number of flows from the same source to specific destination IP addresses
for a predetermined time period. The extracted features are input for the MINDS
system. While well-known intrusions are detected by the known attack detection
module, the anomaly detection module is applied to the remaining network
connections. During the process of anomaly detection, a training set is
generated if not given as input. The anomaly detection module scores each
network connection to reflect the level of anomalousness of the network
connection compared to the normal network traffic. The highly scored network
connections are analyzed by the association pattern analysis module to produce
summary and characterization of the possible attacks. This summary and
characterization is used to create new signatures and models for unknown
emerging attacks.  MINDS contains various modules for collecting and analyzing
massive amounts of network traffic.  Typical analyses include behavioral anomaly
detection, summarization, scan detection and profiling.  The application
includes all four stages: data generation, transformation \& decisions, storage,
and processing.

\subsubsection*{Application description} %2. {\bf How does it do it?}

%\katznote{from answers to questions 2,3,5,6 below, also address
cyberinfrastructure (where does the app run) -- perhaps have a diagram here of
what happens in the app}

MINDS is deployed at each site as a grid service. The analysis application
continuously scans local logs, and reports interesting events to a centralized
analyzer, and alerts are reported back to local sites.

Each site uses a data capturing device such as network flow tools or network
monitoring tools like tcpdump to collect the network traffic data from routers.
This collected data is first filtered to remove unimportant network connections
due to the large data volume.

All communication in the system is accomplished using messages on top of HTTP.

 \subsubsection*{Big data aspects} %3. {\bf What are the big data aspects?}

%\katznote{from answer to question 4b below}

For local data, this is site dependent. It can be MB/hour/site.

 \subsubsection*{Distributed aspects} %4. {\bf What are the distributed
aspects?}

%\katznote{from answer to question 3 below}

The system is deployed across clusters located at a set of participating sites
(University of Minnesota, University of Florida, and University of Illinois,
Chicago) connected by the public Internet. In addition, the data generated is
also produced (distributed) at this set of sites.

 \subsubsection*{Dynamic aspects} %5. {\bf What are the dynamic aspects?}

%\katznote{from answer to question 4a below}

Network traffic data is highly dynamic and constantly streaming in to the
application. The data consists of TCP flow and UDP records, which are very
structured.  They contain IP addresses, ports, various headers, and
payloads. Interesting events are propagated across sites. These are very small,
a few KB at most.

%\ysnote{Some of the applications here like behavior anomaly detection seems to
be triggered on demand, and they consume additional data. This may be similar to
the LSST case where the application is dynamic/steered based on incoming data or
situations.}  %\katznote{Yogesh, what is your suggested action to do with this
comment?}

 \subsubsection*{Other important issues} %6. {\bf What else is important?}

%\katznote{from answers to questions 7,8 below, also address programming systems
(how was the app written; what tools does it use)}

 The application is built using Globus Grid Services for wrapping the MINDS
analysis applications and for providing storage services for events. At the time
the application was written, the project had to develop tools such as remote
launching of Grid services themselves. MINDS was also a separate software
package that was utilized. The project also built an authorization service for
controlling access to data products. Other noteworthy points: real-time is
ultimate goal of the project and false positives are more tolerable than false
negatives.  Finally, the project found it to be very difficult to obtain network
data at each site due to privacy concerns.

 IDS systems also use pattern detection tools such as Bro and Snort, which
perform simple string and regular expression pattern matching on IP packet
headers and body. These are monolithic tools that run standalone, without any
comprehensive programming support to automate updating the patterns or taking
action upon pattern detection.

%\subsubsection{old material follows}
%
%
%1. What is the purpose of the application?
%
%Distributed network intrusion detection and distributed denial of service
attacks.  Network TCP/UDP traffic data is analyzed/mined at each site. Potential
alerts are collected centrally and further analysis is done to see if an attack
may be under way by correlating alerts and events are then communicated back to
local sites. The following features are extracted:1) basic features such as
source/destination IP address pair, port number, and protocols and 2) some
additional features such as the number of flows from the same source to specific
destination IP addresses for a predetermined time period. The extracted features
are input for MINDS system. While well-known intrusions are detected by the
known attack detection module, the anomaly detection module is applied to the
remaining network connections. During the process of anomaly detection, a
training set is generated if not given as input. The anomaly detection module
scores each network connection to reflect the level of anomalousness of the
network connection compared to the normal network traffic. The highly scored
network connections are analyzed by the association pattern analysis module to
produce summary and characterization of the possible attacks. This summary and
characterization is used to create new signatures and models for unknown
emerging attacks. MINDS contains various modules for collecting and analyzing
massive amounts of network traffic. Typical analyses include behavioral anomaly
detection, summarization, scan detection and profiling.
%
%2. How is the application used to do this?
%
%MINDS is deployed at each site as a service. The analysis application
continuously scans local logs, and reports interesting events to a centralized
analyzer, and alerts are reported back to local sites.  These communication
steps use messages.
%
%3. What infrastructure is used? (including compute, data, network, instruments,
etc.)
%
%Small clusters at a set of participating sites (UMn, UFL, UIC). Public Internet
connectivity.
%
%4. What dynamic data is used in the application?
%
%Network traffic data is highly dynamic.
%
%(a) What are the types of data
%
%TCP flow and UDP records -- very structured: contains IP addresses, ports,
various headers, and payloads. Interesting events are propagated across
sites. These are very small, a few KB at most.
%
%(b) What is the size of the data set(s)?
%
%For local data, this is site dependent. Can be MBs/hour/site.
%
%5. How does the application get the data?
%
%The data capturing device such as network flow tools or network monitoring
tools like tcpdump are used to collect the network traffic data from routers.
The collected data is first filtered to remove unimportant network connections.
%
%6. What are the time (or quality) constraints on the application?
%
%Real-time is ultimate goal. False positives are more tolerable than false
negatives.
%
%Gaps: The project had to develop tools such as remote launching of Grid
services themselves. They also found it to be very difficult to obtain network
data at each site due to privacy concerns.
%

%\subsection{Wikianalytics -- Yike}
%
%Wikianalytics offers a new methodology and platform for 3D applications.
%
%Wikianalytics is based on the concept of collective intelligence and social
computing.  %In Wikianalytics, complex and large scale data is analyzed via a
group of analysts %collectively in an interactive and collaborative way. The
data mining life cycle is not a %linear process but a dynamic composition of
concurrent interactive processes with %the joint effort of participating
analysts.  Each analyst contributes data, prior %knowledge, expertise,
algorithms to build, validate, improve, integrate models %(knowledge).  Also,
modern cloud computing and social computing model provides %an ideal platform
for Wikianalytics.
%
%Why is Wikianalytics suitable for 3D:
%
%Data intensive: Data intensive is not just about the quantity but also the
complexity.  %Given the human genome example, the pervasive use of human genome
for %healthcare requires the analysis to be done with medical knowledge.
Analyzing %human genome is not just the issue of dealing with the huge amount of
sequence %data but also the need of deep analysis the biological and medical
implication of the %data. To meet this challenge, collective effort in analysis
is essential.
%
%Distributed: Distributed feature has three means : 1) data is distributed 2)
data owner/ %analysts are distributed and 2) data analysis environment is
distributed.  All three %aspects requires the analysis to be a compositional
process. That is, distributed data %can be analyzed in a distributed way and the
analytical products ( information or %knowledge) can then be combined
(integrated). Such a composition of distributed %process is a ``communicating
parallel composition'' in the sense that for time to time %each individual
component process will communicate with others to exchange the %information and
coordinate the actions. Such a distributed analytics process can be %realized
well by the Wiki collective intelligence principle.
%
%Dynamical: Dynamical nature of data intensive computing comes the dynamism in
%data availability in data comprehension and also in the inherent dynamism in
the %applications.  From data analysis point of view, the dynamism is reflected
by the fact %that an analytical process cannot be captured by a static workflow,
rather it has to be %dynamically constructed based on the partial information
derived during the analysis.  %Such an adaptive nature of analysis is described
in Bayesian statistics as the %posterior probability where knowledge (belief) is
dynamically updated based on the %new available and analyzed data.  Wikianalysis
is not based on the concept of %organizing data via workflow, rather its
collective mechanism reflects precisely the %dynamism needed in analyzing
complex and dynamic data where the analysis %process requires to be built
adaptively based on the up-to-date analytical results gain %from the joint
effort.

\subsection{Application Summaries}

Here we provide short summaries of the computational (as opposed to scientific)
aspects of each application in preparation for the analysis that follows in the
rest of this document, which is motivated by and centered around these set of
applications.

{\bf NGS Analytics (\S\ref{bioSilvia})}: Data is processed through a customized
pipeline of analysis tools, such as BWA, BFAST, Bowtie; processing of each data
element is independent of other elements, so processing is
decomposed/parallelized over the data elements.  All data is stored in files.
Later analysis may want to operate on files from multiple data stores
distributed around the world, e.g., to compare O(1000s) of TB-scale data sets
(metagenomics) with each other.  Dynamic elements include the choice of
executables in each instance of the pipeline, the design of the pipeline itself,
and the choice of computing infrastructure to provide the best resource usage
and turnaround time.  The solution to this problem is not clear.  There is a
need for flexible programming systems and associated runtime environments for
hardware-software co-optimization.

{\bf ATLAS/WLCG (\S\ref{WLCGSteve})}: There is a hierarchy of systems. Data are
centrally stored, and locally cached (and copied to where they likely will be
used), perhaps at various levels of the hierarchy. Processing is done by
applications that are independent of each other. Processing of one data file is
independent of processing of another file, but groups of processing results are
collected to obtain statistical outputs about the data.

{\bf LSST (\S\ref{astroAdam})}: Data is taken by a telescope. Quick analysis is
done at the telescope site for interesting (urgent) events (which may involve
comparing new data with previous data).  The system can get more data from other
observatories if needed, request other observatories to take more data, or call
a human. Data (in files) is then transferred to an archive site, which may be at
the observatory. At the archive site, the data are analyzed, reduced, and
classified, some of which may be farmed out to grid resources.  Detailed
analysis of new data vs. archived data is performed.  Reanalysis of all data is
done periodically.  Data are stored in files and databases.

{\bf SOA Astronomy (\S\ref{astroSOAAdam})}: Services are orchestrated through a
pipeline, including a data retrieval service that is used to share data across
VO sites, as well as analysis and image processing services.  Data (files) are
moved through the pipeline, and intermediate and final products can be stored in
AstroGrid storage service, such as mySpace.  Scratch space is provided for
storage of intermediate results.

{\bf Cosmic Microwave Background (\S\ref{astroJulian})}: This application builds
a sky map.  It reads data from detectors, and obtains data from simulations that
are performed on-the-fly, when requested by the map-making component (in a
master-worker sense), one of a set of components that gradually reduces the data
from the detected and simulated data to a small number of cosmological
parameters.

{\bf Sensor Network Application (\S\ref{sensorSimon})}: Data are collected from
mobile sensors that periodically transmit when they come within communication
distance of receivers. Fault tolerance (collecting data in a hostile
environment) is an important component. The data, which are very diverse,
because they can involved different resolutions, sampling frequencies and other
management characteristics., are brought to a central site. Stored data are
analyzed using statistical techniques, then visualized with tools such as Google
Earth.

{\bf Climate (\S\ref{climateDan})}: Climate simulations are run by climate
centers.  Data (outputs of climate simulations) are distributed in multiple
federated stores.  Current data analysis and visualization tools require the
user to bring the data they want to analyze to a local system, then run the
tools locally.  New tools will include the capability to add data transfer.
This will allow the set of tools to iterate: transfer data, do analysis, etc.

{\bf Interactive Exploration of Environmental Data (\S\ref{envJon})}:
Applications are driven from map-based graphical tools.  The inputs are
geographic areas and algorithms to run on the data in those areas. Data can come
through web services from stored data, sensors, or simulations (possibly
running).  The services can perform server-side processing to reduce the size of
data files to be transferred. Responses are provided in near-real-time.
Security also supported through a delegation-based model.

{\bf Power Grids (\S\ref{power})}: Diverse streams arrive at a central utility
private cloud over cellular and wireless networks at dynamic rates controlled by
the application. Communication is bi-directional, with pricing and control
signals sent to customers power meters.  A real-time event detection pipeline,
involving complex event processing and machine learning/data analysis algorithms
and using other stored information such as weather data, can trigger load
curtailment operations. Data mining is performed on current and historical data
for forecasting. Partial application execution on remote micro-grid sites is
possible.

{\bf Fusion (\S\ref{fusionScott})}: Multiple computational intensive physics
simulation codes run concurrently on a distributed set of parallel computers.
Data from some codes are streamed to other codes to link them into a single
simulation.  Data are transformed in flight into needed inputs.  Data are also
examined in situ to understand the overall state of the simulation.  Some data
are stored for later data mining and visualization.

{\bf Industrial Incident Notification and Response (\S\ref{Kees})}: Data are
regularly streamed from diverse sources, and sometimes manually entered into the
system. Initial disaster detection causes additional information sources to be
requested from that region and one or more applications to be composed based on
available data. Some applications run on remote sites (where the data is stored)
for data privacy. Escalation can cause more humans in the loop and additional
operations.

{\bf MODIS Data Processing (\S\ref{modisKeith})}: Data are brought into system
from various FTP servers.  A pipeline of initial standardized processing steps
on data is done on clouds or HPC resources.  Scientists can then submit
executables and Matlab scripts that do further custom processing on subsets of
the data, which likely include some summarization processing (building graphs).
The application has been run on a cloud and in an HPC environment.

%Floating Sensors (\S\ref{floaterKeith}): Data from static and dynamic %sensors
are brought to a central location for processing by models. The results %of
modeling and forecasting can cause the sensor locations to change.

{\bf Distributed Network Intrusion Detection (\S\ref{detectionJon})}: Data are
analyzed or mined by a service running at a number of local sites. The results
are sent (via HTTP messages) to a central site for analysis. Events are then
communicated (via HTTP messages) back to local sites.

\section{Understanding Distributed Dynamic Data}\label{sec:distdyndata}

% \jhanote{AL and I agree the structure should be: 3.1. (i) Define % Dynamic,
(ii) Formalism/types (iii) % Applications/tables/discussions. 3.2 (i) Define
distributed, (ii) % discuss two stages (a) structuring (static) (b) processing %
(dynamic), (iii) Formalism/table, (iv) Application discussion; 3.3 % discussion}

In the previous section we described thirteen applications and analyzed them
along an analytical space spanned by the three ``D''s of data-intensive,
distributed, and dynamic behavior.

\emph{Distribution} refers to the presence of application data in different
physical or logical locations. Distributed data may arise naturally in that the
data sources themselves are geographically separated, such as from remotely
located sensors.  In some cases, data must remain distributed for privacy and
policy reasons, or due to constraints in sharing.  In other cases, a full
dataset may not fit in one location either on disk or in memory and must be
decomposed to enable storage and/or computation. Distributed data may be driven
by a need for performance, scalability localization, reliability, or
availability considerations. For example, copies of data may be generated and
distributed to maintain data availability or load balancing.  In general, a
\emph{distributed application} is one that needs, or would benefit from the use
of, multiple resource~\cite{dpa_surveypaper}. Example benefits include increased
throughput, decreased time to solution, and increased reliability.  A
distributed application rarely consist of a single component that is
distributed; most often a distributed application is composed of smaller and
functionally self-contained, if not entirely independent, components that can
execute on different distributed resources.

\emph{Dynamism} refers to changes in the behavior and characteristics of an
entity % \jhanote{``of an entity'', does this % sound right?} \katznote{fine
with me} over time and/or space.  Over a long enough timescale, almost all
infrastructure, and especially distributed infrastructure, will exhibit some
form of dynamism, as a consequence of failure, varying load, etc. However, we
are interested in dynamism on timescales that influence application behavior,
scheduling decisions, etc. Some prominent examples are resource fluctuation,
data availability, and change in application execution characteristics.

In this section we will try to extract commonalities in the dynamic behavior of
applications studied in the previous section, as well as properties and types of
distribution seen in these applications.

% \jhanote{needs fixing} After discussing the different kinds of % dynamism, we
will apply this framework to the applications.  This will % permit an extraction
of common features and possibly \katznote{either % we can do this or not...} a
derivation of requirements for multiple % applications.

\subsection{Types of Distribution \label{sec:distTypes}} %Distributed
\emph{data} occurs e.\,g.\

There are many scenarios under which distribution of data is either necessary or
desired. In addition to the reasons behind distribution presented at the
beginning of this section (viz., large volumes of data) and reasons that
constrain or localize data (viz., privacy/policy issues), another common
situation under which data is distributed occurs when data comes from multiple
sources, for example multiple sensor streams or data centers, but needs to be
processed collectively.

% and (ii) where a data % set is too large for a single storage module and thus
must be stored % in a distributed % way.  % \alnote{application
stages?}\jhanote{Andre, please clarify?}

%{\it Application structure and distribution of data:}

% One or multiple data sources (M) are expanded to one or more % locations (N),
M$ <$ N, for processing.  Scattering is commonly used % to facilitate
distributed processing of data.

There are many ways to address distributed data.  These include but are not
limited to: % Ways to manage \katznote{?} the % distribution include but are not
limited to:
\begin{itemize}
\item \textbf{Replication:} Data is duplicated on multiple nodes. When fully
replicated, each execution unit has access to the same data. Replication can be
partial, i.e., data can be replicated to a subset of resources. The degree of
replication quantifies the average of the number of copies of each data
item. Replication may be non-uniform, e.g., if only a subset of data (perhaps
``hot'' data) is replicated. Caching is a lightweight form of replication.
\item \textbf{Partitioning:} Data is divided and distributed across multiple
nodes. Different forms of partitioning exist.  For example, data can be
thematically partitioned with respect to its type, such as sensor data
vs. simulated data, or data can be partitioned according to a spatial or
temporal attributes. This can be an opaque (e.g., ID) and or a domain dependent
(e.g., spatial) attribute.
\item \textbf{Streaming}: A form of distributed data processing in which data
arrives from a source; often the stream is a steady, continuous flow, but often
it fluctuates both spatially and temporally. Also, typically, the stream of data
must be processed in near real-time. Streaming thus also has attributes of
dynamism, which is discussed in the next subsection.
\end{itemize}

Which of the above approaches is employed cannot be reduced to a simple
rule. Which option can or should be used often depends upon a plethora of
co-dependent factors, such as the data volumes, the {\it degree of distribution}
(which describes the number of data sources and sites used for storing and
processing the data), the availability of compute resources, or the desired
level of reliability. Data distribution can be difficult to manage for reasons
of performance: latencies and bandwidths may vary unpredictably.  Further,
issues associated with reliability may also need to considered. An example is
the occurrence of a partial failure, where some (but not all) of the required
data remains available, leads to applications having to decide whether to
continue with reduced data or exit without processing what remains available.

%\jhanote{I am not sure this is the best way to express co-dependence % of
reliability and performance}\alnote{refined}

Viewed from an application's perspective, there are at least two application
characteristics that influence the distribution of data. The first is the stage
and the second is the degree of coupling between components of a distributed
application. Distribution in the first stage is primarily determined by the
structure and layout of the data sources. For example, the placement of the data
generated in the climate application (\S\ref{climateDan}) is based on which site
was assigned to run the particular model that generates the data. But analysis
of the data from multiple models requires bringing the data together to a single
site.  Although it is difficult to formalize the difference due to the large
fluctuations between different applications, but in general for a given
application, there are differences between the distribution prior to the
application's processing stage and the distribution of data within the
processing stage, e.g., to aid data-parallel processing.

% \jhanote{can we % provide an example or some further application discussion
for the % previous point. it would help immensely}

The second application factor that determines distribution of data is the type
of separation and coupling of the components of the distributed
application. Different coupling types place different constraints on data
distribution, either due to latency, or due to data volumes that must be
coordinated. % Different types of coupling in % turn influences the distribution
of data.  \emph{Loosely-coupled} distribution involves coordination and
cooperation between largely independent application components, as might be
found in a workflow system. \emph{Tightly-coupled} applications involve greater
constraints between the components, either in latency tolerance, coordination
requirements or relative scheduling. \emph{Decoupled} distribution results in
the separation of application components elements of the same task, for example,
sharing a spatial data set between components that cooperate to perform a single
logical task~\cite{dpa_surveypaper,thebookthatchangedtheworld}.  Most of the
applications in \S\ref{sec:scenarios} are loosely coupled, with the CMB
(\S\ref{astroJulian}) and Fusion (\S\ref{fusionScott}) applications as examples
of tight coupling.

% \jhanote{again, can we relate the above general paragraph to % specific
application examples from the previous section} % \katznote{tried to come up
with an application example, but instead, % realized that I was very confused by
this paragraph.}\jhanote{Any % better? If you could be more specific that'll
help}

\begin{table}[h]
  \begin{scriptsize}
    \begin{center}
      \caption{Distributed Processing}\label{tab:app_distributed_processing}
      \begin{tabular}{|p{3.5cm}|p{10.2cm}|} \hline \textbf{Application}
&\textbf{Distributed Characteristics} \\
% \endhead \hline Next Generation Sequencing (NGS) Analytics & Data sources (NGS
machine storage) and compute not always co-located; multiple data sources in
case of metagenomics \\ \hline ATLAS (an example of use of the WLCG) &Source
data is generated by one central data source. Data needs to be distributed to
match distribution of computing resources. During processing data is copied from
closest replica to the resource where compute job is run. \\ \hline LSST
&Multiple different kinds of data sources (observatories, data storage);
multi-step workflow that processes and archives data, carried on distributed
compute resources.\\ \hline SOA Astronomy Applications &Data resides in multiple
data centers; services are placed closed to the data.\\ \hline Cosmic Microwave
Background &Replication of subset of data to compute resources for processing.\\
\hline Sensor Network Application & Data is moved from different data sources to
a central analysis system for processing\\ \hline Climate (ESGF) &A subset of
the data is commonly moved to compute (desktop or grid resource)\\ \hline
Interactive Exploration of Environmental Data &Data is brought to grid resource
or desktop \\ \hline PowerGrids &Data is brought to central private cloud for
analysis\\ \hline Fusion (ITER) &Simulations run wherever resources are
available; in-situ analysis of large output data; partitioning/replication of
data for data-parallel processing\\ \hline Industrial Incident Notification and
Response &Distributed data sources. Processing takes place on multiple
distributed resources. \\ \hline MODIS Data Processing & Data is copied
(replicated) from different sources for processing.  Processing can take place
in different locations, e.\,g.\ cloud and user desktop. \\ \hline Distributed
Network Intrusion Detection &Distributed data and compute resources.\\ \hline
     \end{tabular}
    \end{center}
  \end{scriptsize}
\end{table}

%\subsection{Characterizing Distributing Dynamic Data \label{sec:appChar}}

% \katznote{without the table that was here, we are now defining % parameters
($\alpha$, $\beta$) that we don't really do anything % with.  We should think
about what part(s) of this subsection make % sense to keep.} \jhanote{the tables
have been moved (and reference % to them fixed). Admittedly we don't use
$\alpha$ or $\beta$ % again..}  \katznote{then why do we define them?}

\subsection{Types of Dynamism \label{sec:dynTypes}}

\begin{table}[h]
  \begin{scriptsize}
    \begin{center}
      \caption{Understanding Dynamic Characteristics. This table highlights
different types and scenarios for dynamic data in the application set
investigated %\katznote{some explanation %here might
help}\jhanote{OK?}\katznote{yes} }
          \label{tab:app_dyn_char}
      \begin{tabular}{|p{4cm}|p{9.5cm}|} \hline \textbf{Application}
&\textbf{Dynamic Characteristics} \\
% \endhead \hline Next Generation Sequencing (NGS) Analytics & Data Dynamism:
different data formats and sources need to be processed\\ \hline ATLAS (an
Example of Use of the WLCG) & Data Dynamism: new raw data is steadily
collected\\ \hline LSST & Infrastructural Dynamism: Decision making based on
observed data\\ \hline SOA Astronomy Applications & Application and Data
Dynamism: data is continuously updated, dynamic queries\\ \hline Cosmic
Microwave Background Mapping & Application Dynamism: dynamic utilization of
resources\\ \hline Sensor Network Application & Data Dynamism: rate and volumes
of data dynamic \\ \hline Climate (ESGF) & Data \&Infrastructural Dynamism: new
data is generated over time. Dynamic distributed infrastructure for storing and
processing of data %\jhanote{please check} \\ \hline Interactive Exploration of
Environmental Data & Infrastructural and Data Dynamism: Various data sources,
varying input rates and different queries.\\ \hline PowerGrids & Application,
Infrastructural and Data Dynamism: data itself changes, data rates may change
(throttling), data sources change (traffic, weather, social networking), data
queries respond to seen data (e.g.  accuracy adjustment) \\ \hline Fusion (ITER)
& Application and Infrastructural Dynamism: data streamed between simulations
(multiple distinct data flows) \\ \hline Industrial Incident Notification and
Response & Application and Data Dynamism: data itself and data rate changes,
dynamic data sources, processing changes with data\\ % \jhanote{Omer needs to
clarify if application %dynamism too?} \\ \hline MODIS Data Processing &
Application and Data Dynamism: data that is operated upon changes,
infrastructure config changes\\ \hline Distributed Network Intrusion Detection &
Infrastructural, Application and Data Dynamism: data highly dynamically
changing, data rate variable, application responds to data, on-the-flight
processing\\ \hline
     \end{tabular}
    \end{center}
  \end{scriptsize}
\end{table}

% \jhanote{Modify this paragraph to reference table 1 and discuss the % kinds of
dynamism} There are often spatio-temporal variability in the values and
volumes~\cite{deRoos2011UnderstandingBigData} of data.  The applications in
\S\ref{sec:scenarios} have variability in the production/generation rate of
data, and in the configuration of data production/generation.  Data dynamism may
arise from extrinsic factors, for example because sensors begin to observe more
events in the same data stream. Contrariwise, data dynamism may occur because an
application component requested an increased sampling frequency.  Similarly
application dynamism may occur through managed replication of a
compute-intensive component or from the demands of a particularly complex
query. And infrastructure dynamism may arise because the infrastructure expands
to meet additional requirements, or contracts because of device failure.  A
structured analysis of the applications in \S\ref{sec:scenarios} suggests that
the sources of dynamism can be classified into three categories.
\begin{enumerate}

\item \emph{Data} dynamism arises when some property of the input data or its
delivery changes, for example in terms of arrival rate, provenance, burstiness,
or source.  There may be variability in the structure of the data, for example,
data schema, file formats, ontologies, etc.

  % \alnote{Not sure, but maybe changes in the structure of data would % better
fit into category 1 or 3 since the structure is typically % not related to the
infrastructure}\ysnote{I agree with Andre.}  % \jhanote{In order to % be
consistent with paragraph above, the third type of dynamism % should focus on
the infrastructure -- resources not on data any % more.}

\item % Dynamism in the Application, % Software/Services use of the data:
\emph{Application} dynamism involves changing the processes or components
applied to a data set, and might encapsulate changes in workflow, \textit{ad
hoc} queries, or parameterization.  There may be variability in the data queries
enacted by the application, or in the behavior and/or characteristics of the
application.  % \ysnote{this may affect % data access as well as generation of
derived data}

\item \emph{Infrastructure} dynamism occurs when the system wants to, or is
forced to, change its demands on its underlying platform, for example through
elastic computation or partial failure.  Changes in the structure and load of
the infrastructure can be associated with data.  Examples include variability in
the producer/provenance of the data, in the performance, and in the quality of
service of the infrastructure.

\end{enumerate}

Based upon this classification, in \tablename~\ref{tab:app_dyn_char}, we
describe the applications discussed in \S\ref{sec:scenarios}.  Many of these
dynamic aspects are related: an increase in the rate of data generation may lead
an application to increase its degree of parallelism to maintain throughput,
which may in turn trigger the allocation of new compute nodes. (In the other,
less attractive direction, decreasing computational capability might cause a
decrease in parallelism and a consequent need to drop data. Avoiding these cases
means treating dynamism equally in both directions.)

This should not be construed to imply that every data-intensive application has
all elements of data and compute dynamism and is always distributed, but is a
reflection that at extreme scales, dynamism and distribution become increasingly
important issues that are often correlated.  As alluded to, a high degree of
distribution often results in a high amount of infrastructure dynamism, %
\katznote{I don't really like this forward-looking link, or the one in % the
next paragraph}\jhanote{Not forward looking anymore} caused by, for example,
resource fluctuations and a higher failure probability.  Similarly, high levels
of dynamism, for example, in the data sources, data rates, and queries, usually
correlate with a high degree of distribution.  For example, the Interactive
Exploration of Environmental Data application (\S\ref{envJon}) manages various
sources of data, from archived data sets to real-time feeds.  Depending on the
spatiotemporal properties of data, computation can be carried out on a set of
distributed grid resources or the user's desktop, i.e., distributed processing
is used to efficiently handle dynamic data.

\subsection{Distributed Dynamic Data Observations}

%\jhanote{needs text to establish that this is the end of section 3} \katznote{I
think this should be a short subsection, as I've done here - can be removed...}

Reviewing \tablename{s}~\ref{tab:app_distributed_processing} and
\ref{tab:app_dyn_char}, which have summarized the distributed and dynamic
properties of the D3 applications, we observe:

%find that commonly re-occurring patterns %\jhanote{should we be calling the
following patterns? I don't think % they are patterns, but possibly
observations?} \katznote{yes, `observations'} in the data %distribution and
processing are:

\begin{itemize}
\item Data generation and consumption are commonly distributed from each other
if the volume permits.  The Fusion application is a counterexample, due to data
volumes and velocity.

\item Details of the storage system are often not known. Often data is stored in
a distributed way, e.g., on a distributed file system, which commonly fits into
the medium degree of distribution (the same data center).

%\item Mapping: The most common pattern is to copy data to an available %
compute resource. \katznote{I'm not sure about this bullet: 1. It's not a
sentence; % 2. I don't know what it's saying.}

\item Aggregation of data can either be the equivalent of gather or reduce
(using the language of collective communication). Gather is in many ways the
reverse process of distribution (called broadcast in the language of collective
communication). Similar to distribution, aggregation is a commonly occurring
pattern for data analysis.

\end{itemize}

%\subsubsection{Characterizing Dynamic Data \label{sec:dynDataChar}} %
\jhanote{AL: Next sentences OK to go?} Also, in lieu of quantitative %
estimates, we attempt classification of these ratios into categories % of
high/medium/low.  It is natural to expect that applications will % cluster
around certain {\it trends}.

% \jhanote{AL/SJ: The following should be deleted?} As a guide to how % the
metrics proposed may be useful, we characterize a couple of the % applications
investigated using $\alpha$ and $\beta$: % \katznote{actually, neither the
examples here nor the table mention % $\alpha$}

% \begin{itemize}
% \item NGS: $\beta_1$ is high, $\beta_2$ is high. Degree of % Distribution is
Medium. The number of distributed sites does not % scale with data.

% \item Atlas: $\beta_1$ is Medium, $\beta_2$ is Medium. Degree of %
Distribution is High. The number of distributed sites scales with % data.
% \end{itemize}

% \jhanote{Is LSST like NGS or Atlas or in its own class?}  % \jhanote{Is
PowerGrid more like NGS or Atlas or in its own class?}  % \note{metrics for
distribute data} \jhanote{No idea who put the % previous note there or why? Help
please?}

% \jhanote{not sure where this goes?} replicated $\rightarrow$ average %
replication factor of a data element/N ; coupled $\rightarrow$ M/N

% \jhanote{Not sure this is correct?}  % Degree of distribution: latency between
primary data source and data % processing element

% \alnote{Refine description, define view point} \note{Is the control of % the
sensors from the application a case of 5 or is it something % different?
Feedback to the data source (other areas, sources of % data, accuracy)?}
\note{most applications commonly operate in a % local setting, but can also
operate in a distributed settings.}  % \alnote{How to measure degree of
distribution?} \jhanote{these notes % should move}
% \begin{table}[h!]

% \begin{table}
% \begin{scriptsize}
% \begin{center}
% \caption{Dynamic and Distributed Characteristics of Example
Applications} \label{tab:app_characteristics}
% \begin{tabular}{|p{3cm}|p{5.0cm}|p{5.1cm}|} % %\begin{footnotesize}
% %% \begin{tabular}{|p{4cm}|p{5cm}|p{5cm}|}
% % \begin{longtable}[h!]{|p{3cm}|p{5.0cm}|p{5.1cm}|} % % \caption{Dynamic and
Distributed Characteristics of Example
Applications} \label{tab:app_characteristics} \\ % \hline % \textbf{Application}
&\textbf{Dynamic Properties} &\textbf{Distributed Properties} \\ % % \endhead %
\hline % Next Generation Sequencing (NGS) Analytics & Infrastructure dynamics
(2) % &degree-of-distribution: medium; $\beta_1$: high; $\beta_2$: high \\ %
\hline % ATLAS (an example of use of the WLCG) &New raw data is collected %
steadily (1) % &degree-of-distribution: high; $\beta_1$: medium; $\beta_2$:
medium \\ % \hline % LSST &DE decisions based on observed data (3)
&degree-of-distributed: % medium, multiple different kinds of data sources
(observatories, data % storage); $\beta_1$: ?; $\beta_2$: ? \\ % \hline % SOA
Astronomy Applications &data is continuously updated (1), thus % dynamic data
(pre-defined schedules or on-demand), queries change (3) %
&degree-of-distribution: high, multiple kinds of data sources\\ % \hline %
Cosmic Microwave Background &dynamic utilization of resources (2 or 3?)  %
&degree-of-distribution: low--medium\\ % \hline % Sensor Network Application
&data rate, volume dynamic (1) % & degree-of-distribution: high, multiple kinds
of data sources\\ % \hline % Climate (ESGF) &New data generated over time
(ESGF-wide - is this in % scope of the application?) (1)
&degree-of-distribution: high, multiple % kinds of data sources\\ % \hline %
Interactive Exploration of Environmental Data & data rates and % data
inputs/outputs change (1), queries change (3) % & degree-of-distribution: high
\\ % \hline % % Dynamic Environmental Data on the Web &data operated on changes
% % (realtime feeds) (2) % % &natural (sensor sources, simulations), (sub-set)
replication\\ % % \hline % PowerGrids &data itself changes, % data rates may
change (throttling) (1), % data sources change (traffic, weather, social %
networking) and infrastructure dynamics (2), response to data (e.g. accuracy
adjustment) (3) % &degree-of-distribution: high, large number of sensors;
$\beta_1$: medium; % $\beta_2$: medium \\ % \hline % Fusion (ITER) &data
streamed between simulations (multiple distinct data % flows) (1), on-the-fly
transformation (3) % &degree-of-distribution: medium\\ % \hline % Industrial
Incident Notification and Response &data itself and % data rate changes (1),
dynamic data sources (2), processing changes % based on data (3)
&degree-of-distribution: high\\ % \hline % MODIS Data Processing & dataset that
is operated upon changes (1), % infrastructure config changes (7) %
&degree-of-distribution: high, multiple data sources\\ % \hline % % Floating
Sensors &data itself (1), data rate (2), application causes % % data changes (5)
% % &degree-of-distribution: high, multiple kinds of data sources; % %
aggregation \\ % % \hline % Distributed Network Intrusion Detection &data %
highly dynamically changing, data rate variable (1), application % responds to
data, on-the-flight processing (3) % &degree-of-distribution: high\\ % \hline %
% % \end{table} % % \end{longtable} % %% \end{tabular} % %\end{footnotesize}
%
% \end{tabular}
% \end{center}
% \end{scriptsize}
% \end{table}

\section{Infrastructure}\label{sec:infrastructure}

%\alnote{relate infrastructures back to applications: in particular, how do
specific purpose built infrastructures as the LHCG or ESGF fit here?}

We define {\em infrastructure} as the hardware and software that is provided to
support an application, rather than being explicitly created or hosted by the
user themselves. In many cases the precise boundary between infrastructure and
application is blurred, e.g., a catalog is a component with functionality that
may be similar to that of similar components in other applications, but is often
implemented specifically by an application rather than being provided as part of
the infrastructure. This may also be the case when reconfiguring an
infrastructure service for an application, when it is more onerous than
reimplementing the service specifically for that application. We also introduce
the concept of an ``infrastructure'' application as a collection of
infrastructure components that are packaged so that they can be used by many
applications in defined ways. The difference between these and ``traditional''
applications is often delineated by who owns and administers the application,
with traditional applications ``owned'' entirely by the user, and infrastructure
applications owned by someone other than the user but provided for use by
others. Infrastructure can be specialist as opposed to generic, for instance the
purpose-built infrastructure in ESGF (\S\ref{climateDan}) that is provided to
support the climate community.

In this section, we focus on the {\em software} infrastructure that is provided
to support D3 applications, specifically components that deal with distributed
or dynamic data. The components that make up the infrastructure can be
categorized based on the function provided, namely (i) infrastructure that
supports {\it data management}, (ii) infrastructure that supports {\it data
  analysis}, and (iii) infrastructure that supports the \emph{coordination
  within applications}. Each of these components may be present at three
different levels of the infrastructure: programming frameworks, services and
tools (e.g., MapReduce, workflow engines); middleware and platform services
(e.g., databases, message queues); and system-level software capabilities (e.g.,
notifications, file system consistency).

\subsection{Infrastructure to Support Data Management}
\label{sec:infra_mgt}

% \katznote{we should think about how the items in this section map to the boxes
and arrows in Figure~\ref{fig:figures_application-stages}.  For data sources,
storage, access, and movement, it's fairly clear.  For discovery and
notification, it's less clear.}

{\em Data Sources} are used in the acquisition stage as inputs of data into the
application. Broadly speaking these sources include types such as files, servers,
sensors, instruments, simulations and models.  Additionally, databases are often
considered as an intermediate data source in types of system. The dynamic
characteristics are primarily due to availability. These are often accessed in
the Data Generation stage of the application
(\figurename~\ref{fig:figures_application-stages}).

{\em Data Storage} components are used to provide persistent access to data by
the application. They may be structured (e.g., databases, No-SQL tables) or
unstructured (e.g., collections of audio recordings). The dynamic
characteristics are primarily due to the level of persistence, e.g., temporary,
durable, permanent (archived or published), or replicated; and also in the
structure of the data stored, e.g., schema free tables as opposed to relational
tables. There may also be dynamic characteristics of storage due to rate of
growth of data, e.g., a field survey may bring in a lot of data during summer
but none during other times.  These are commonly used in the Data Storage stage
of the application (\figurename~\ref{fig:figures_application-stages}), but may
also be present at the boundaries between the stages for transient staging of
data.

{\em Data Access} components are gateways to Data Sources and Data Storage
components, and provide interfaces that often enforce security controls to
restrict access, provide dynamic information about the available data sources
and/or data analysis components, and may also perform logging of requests and
data movement (e.g., tracking provenance for auditing). A side effect of
auditing/logging may be the growth of provenance metadata simply due to access
transactions even if no new data is created. Data Access components usually
serve at the interface between the application stages, before and after data
movement.

{\em Data Movement} components are used for bulk or continuous movement of data
between parts of the application, or to and from the application. % (often
called staging) Data movement may be explicit and visible to the application
(e.g., staging), or implicit (e.g., automatic replication of files). Components
may operate on streams or blocks (files). Dynamic characteristics include
partitioning, scheduling and planning (e.g., edge caching of ``hot'' data, where
datasets that are being used frequently during some period are replicated close
to the application). Data movement can be considered to be real-time (i.e., as
it happens, e.g., sensor data), as real-time (e.g., a storage buffer has been
introduced into the transport system, but the data source is still real-time, as
with time-series data), or asynchronous (where the ordering of the data movement
is independent of the data itself).  Latency may be considered as an important
characteristic of the data movement infrastructure depending on the application
e.g. real-time processing of sensor data for disaster analysis. Other dynamic
aspects of data movement include data rates, their variability from data
sources, and the ``freshness'' of data. These infrastructure appear on the edges
between the application stages.

{\em Data Discovery} components assist the access to and use of data sources,
and include Catalogs, Information Services, and Metadata stores. The dynamic
characteristics pertain to the evolution or recalibration of information held in
these components. These components are orthogonal to the applications stages and
can be used by any of the stages to guide their selection or identification of
incoming data sources and outgoing data sinks.  %\ysnote{rate of growth of data
may be an issue.  %Catalogs that perform nightly ``crawling'' or data loading
(e.g., sky surveys) may not be suitable if the %latest data needs to be
accessible to applications. Catalog update will have to keep up with data
%update. However, batch processing applications can survive less frequent
updates.}

{\em Notification} components respond to changes in the application or data and
trigger other functionality. These include publish-subscribe message brokers,
and complex event processing. By nature, these components support the dynamic
aspects of an application.  These components again are orthogonal to the
applications stages and can be used for coordination of the stages and data
movement.

In each case, these components can be single or centralized, distributed,
partitioned or federated. These are also typically offered as middleware and
platform services, or system-level software capabilities.

%\nchnote{where do we put data publishing? Calibration? provenance?}\ysnote{They
can be incorporated in data discovery and data access}

\subsection{Infrastructure to Support Data Analysis}
\label{sec:infra_analysis}

% \alnote{Should we also include infrastructure for data exploration, data %
mining, machine learning, visualization?}  %\katznote{we should think about how
the items in this section map to the boxes and arrows in figure 1.  The first
three map to both the data transformation \& decisions and data processing
boxes.  I have no idea what should happen with workflow} % \alnote{should we
also talk about the computational aspects of the infrastructure?  % At some
point compute and data need to come together.}

Analysis components include functionality such as comparison, statistical
analysis, event analysis (i.e., analyzing information derived from data rather
than raw data) and visualization. These components leverage the computational
aspects of the infrastructure.  Based on the application stages outlined in
\figurename~\ref{fig:figures_application-stages}, we identify a number of
infrastructure services that can be associated with these stages.

{\em Conversion} components (associated with the Data Transformation and
Decisions stage) deal with the changing of data into a different form, including
file conversion, (re-)formatting, transformation and metadata
extraction. Heterogeneity of formats and representations is common in many
scientific disciplines and such components often provide essential
pre-processing of data prior to analysis. These are often referred to as adaptor
services (or ``shim'' services in the Taverna workflow system~\cite{Taverna}).

{\em Enrichment} components take data and attach additional detail to them,
which might include filtering, image processing, information integration, data
tagging, semantic annotation, data augmentation, or data fusion (e.g.,
OLAP). Many of these components may deal with distributed sources of data. It is
useful to note that these operations may not be fully automated and may involve
a human expert. These components would be associated with the Data
Transformation and Decisions stage.

%\ysnote{addresses Andre's commment} {\em Analytics and visualization}
components help extract domain knowledge and assist with exploration of the
data. These are often closely supervised by the end-user and may even be
interactive. The proliferation of ``big data'' has caused data mining libraries,
machine-learning toolboxes and visualization environments to become part of the
infrastructure, some of which are deployed for specific domains but others which
are more generic (e.g., Weka, GraphLab). These components are associated with
the Data Processing stage.

\subsection{Infrastructure to Support Coordination within Applications}
\label{sec:infra_coord}

Programming abstractions and orchestration frameworks cut across both data
management and analysis functions. While different in character from the other
infrastructure discussed in this section, they enable the coordination of the
different stages of the application, sometimes enforcing a particular constraint
to improve, e.g., scalability. These approaches can allow the composition of
sequential data processing pipelines and further offer the flexibility of
modeling more complex constructs.  Standardization of APIs and components within
these frameworks can allow interchangeability of components with similar
functionality, such as those described in \S\ref{sec:infra_mgt} and
\S\ref{sec:infra_analysis}.

{\em Orchestration} components handle the management and choreography of data
placement, code placement, communication, and messaging to allow complex
analyses to be performed on distributed data sources. They may also be dynamic
if they have the capability to adapt their behavior based on earlier parts of
the workflow or can be steered based on external notification received during
execution. In addition to supporting choreography, often making use of one or
more resource manager/scheduler components, an orchestration component may also
include specialist support for conversion and analysis components discussed
above.

{\em Bulk Data-parallel} frameworks use ``Bulk'' operators to express
computations over collections of data which are executed in a loosely-coupled,
data-parallel manner.  The {\em MapReduce} paradigm~\cite{MapReduce} and its
variants~\cite{twister,hmr} are the most prominent among these. MapReduce forces
programmers to model computation in two stages: a \emph{map} phase in which a
transformation is applied to every element of a collection independently of
every other; and a \emph{reduce} phase in which the transformed results are
aggregated through a binary function that is associative and commutative.  These
properties serve to guarantee that the MapReduce computation can be efficiently
parallelized, by performing the map phase concurrently and then reducing without
the reduction order affecting the final result.

{\em Stream processing} (sometimes called event processing) is an approach
modeled on signal processing, in which a stream of data is transformed
``on-the-fly'' to produce one or more further streams of values. As such it is a
programming abstraction suitable for continuous data feeds, or when results need
to be pushed to users of interactive applications. Since the data is not assumed
to be available \textit{en masse}, it can be used in situations where storing
the entire data set is undesirable, unnecessary, or impossible.

{\em Sensor networks} and other intelligent front-end data collection systems
are sometimes seen as an extension of stream processing and application
pipelines.  They often have the feature of being \emph{adaptive}, in that they
can vary their exact behavior according either to external control stimuli or
from observations made of the data they are themselves observing. Adaptation may
also be through users interactively changing the the pipeline, either by
changing parameters or by adding or removing tasks in the pipeline. This allows
decision logic to be initiated by the pipeline itself, or using an external
control signal that may come from users or an instrument. This form of
``computational steering'' is especially useful in systems being used to explore
a data set. In these cases, it is important to note that the infrastructure is
adaptive, rather than the scenario itself.

%Gomes et al.~\cite{gomes08} describe how a pipeline may be %dynamically
modified using structural and behavioral operators.  %This also provides an
end-to-end description of the entire %scientific process. This is valuable both
for documentation purposes, %but also in pursuing issues of component and data
re-use.

\subsection{Examples of Specific Cyberinfrastructure Usage}

We conclude this section by revisiting the application scenarios of
\S\ref{sec:scenarios} in terms of the infrastructure and coordination approaches
that they use currently, and discuss elements that could usefully be deployed.

The software infrastructure components we have classified above address
different characteristics of D3
applications. \tablename~\ref{tab:inf_app_mapping} summarizes the role of these
components in addressing dynamic and distributed characteristics of the D3
applications in \S\ref{sec:scenarios}. Each application makes use of different
components, though it can be seen that applications that utilize stream or event
processing %(see \ref{tab:progSystems} \katznote{this table name doesn't
resolve...}  have larger requirements on notification and orchestration
components.

\begin{table}[h]
  \begin{scriptsize}
    \begin{center}
      \caption{Infrastructure used to support dynamic and distributed properties
of application scenarios % \jhanote{Are all the terms % in parenthisis
describe/defined. I think not, e.g., % orchestration? Also, I removed analysis
as a % category.. needs fixing here too} \ysnote{fixed} }
      \label{tab:inf_app_mapping}
      \begin{tabular}{|p{6.7cm}|p{6.7cm}|}

\hline \multicolumn{2}{|l|}{{\textbf{Application Name}}} \\ \hline %\nchnote{For
example, workload decomposition and distribution must be determined dynamically
in order to optimally use resources, e.g., compute resource selection can be
based on data location, processing profile and/or network capacity. Workflows?
Resource selection?}  Infrastructure for Dynamic Properties & Infrastructure for
Distributed Properties \\ \hline
%\begin{footnotesize}
% \begin{longtable}[h!]{|p{6.7cm}|p{6.7cm}|}
% \caption{Infrastructure used to support dynamic and distributed properties of
application scenarios} \label{tab:inf_app_mapping} \\ % \hline %
\textbf{Infrastructure to support Dynamism} % &\textbf{Infrastructure to support
Distributedness} \\
% \endhead \hline \multicolumn{2}{|l|}{{\textbf{Next Generation Sequencing (NGS)
Analytics}}} \\ \hline %\nchnote{For example, workload decomposition and
distribution must be determined dynamically in order to optimally use resources,
e.g., compute resource selection can be based on data location, processing
profile and/or network capacity. Workflows? Resource selection?}  Dynamic
workload decomposition and distribution for resource optimization
\emph{(Orchestration)} & Distributing data from source (generation) to the
computing destination (analysis) \emph{(Data Discovery, Movement)} % No unifying
single requirement or single solution at present (or possible at all?)  %
Normally data local and CI doesn't easily support movement of data to compute %
\ysnote{We may want to remove the preceding lines since we're talking about
existing infrastructure, not missing ones at this point.}  % naturally
distributed: data source (NGS machine storage) and compute on % different
resources (typically not co-located); loosely coupled (many % machines,
different repositories) \\ \hline \multicolumn{2}{|l|}{{\textbf{ATLAS (an
example of use of the WLCG) Reconstruction app}}} \\ \hline Manage gradual
growth of data using SRM \emph{(Data Storage, Access)} % Management, subset
replication and partitioning to go inf components: data replication, (gLite),
data movement % specify SRM as standard and then they leave choice of
implementation to the organisations as part of consortium.  % New raw data is
collected % steadily (2) \note{timescale needs to worked out. What happens if
something is found?}  & Management, subset replication, data movement and
partitioning using gLite and SRM \emph{(Data Storage, Access, Movement,
Discovery)} % , data movement % natural, (sub-set) replicated, loosely coupled
\note{Is natural the % right word? Artificial % distribution (the source is
natural, but the intermediate points are % artificial), data not
partitioned}\ysnote{These details are not fully present in the actual
application description of its dynamism (Sec 2.2). Should we trim there and move
some content there? } \\ \hline \multicolumn{2}{|l|}{{\textbf{LSST}}} \\ \hline

Steering observations, Databases to manage data growth \emph{(Notification, Data
Storage)} % DE decisions based on observed data (2, 5) \ysnote{what do these
numbers mean? specific % characteristics?}  & Coordinate classification,
distributed databases, data movement to distributed observatories
\emph{(Notification, Data Access, Data Movement, Data Storage)} %Analysis, % ,
natural distribution of telescopes \\ \hline \multicolumn{2}{|l|}{{\textbf{SOA
Astronomy Applications}}} \\ \hline

Catalog contents updated, Query mix changes \emph{(Data Discovery)} % data is
continuously updated, thus dynamic data (pre-defined schedules or on-demand),
queries change (7)\note{timescale clarification!}  & Distributed VO Services for
data storage and access \emph{(Data Discovery, Data Access, Data Movement)} %
moving data sources, services and storage are distributed across institutions \\
\hline \multicolumn{2}{|l|}{{\textbf{Cosmic Microwave Background}}} \\ \hline
Dynamic resource selection %dynamic utilization of resources (4) & Compute on
distributed HPC centers \emph{(Data Movement)} % (sub-set) replication (during
application run) \note{What phases of the application do we consider? data
acquisition? app run?}\ysnote{Can any of the supercomputers be selected? If so,
the data is moved to distributed compute (or it is dynamic resource selection)}
\\ \hline \multicolumn{2}{|l|}{{\textbf{Sensor Network Application}}} \\ \hline
Notification of sensor data when in range, data rates and volumes \emph{(Data
Movement, Notification)} %data rate, volume dynamic (1-2) &Aggregating
distributed sensor data \emph{(Data Movement, Notification)} \\ \hline
\multicolumn{2}{|l|}{{\textbf{Climate (ESGF)}}} \\ \hline Manage data growth %
New data generated over time (ESGF-wide - is this in scope of the application?)
& Distributed, replicated catalogs using ESGF software stack (GridFTP,
wget/HTTP) \emph{(Data Access, Data Discovery, Data Movement, Data Conversion)}
\\ % multiple climate centers (natural), users can request data from ESGF
(sub-set replication), coupled % Can come in via website - cookie / OpenID %
Come in through command line script - grid cert \hline
\multicolumn{2}{|l|}{{\textbf{Interactive Exploration of Environmental Data}}}
\\ \hline On-demand data subsetting, caching and versioning, real-time
instrument data \emph{(Data Discovery, Data Movement, Data Sources)} % at moment
services are hosted on a single address. There may be farming for load balancing
in the future.  % data rates change (1), % data inputs/outputs change (3),
queries change (7) & Distributed Data Services, Access Restrictions \emph{(Data
Movement, Data Access)}

% data frequently located, but data sources themselves don't change
locations. Some replication for bigger community datasets. No replication
service - just another website!

% distribution of data has security constraints. Access rights which means you
can't just copy data. Data providers like to assert these rights at source.

% Increasingly user tokens being passed to data providers, but previously
application had tokens and trust required on part of data providers. Data
providers want to have user tokjens so they can tell usage, and whether its
researchers or commercial.

% Web server accessing data on your behalf. Not often directly accessing data -
going through an intermediary service. trying grid proxy certs and OAuth.

% If you had freely available data you could build a better system.

% As a response to query you get a subset of data % Virtualization and
aggregation of files - see single logical dataset rather than the thousands of
files its stored as.

% OpenDAP binary stream for data transport. logical OpenDAp data set may be
collection of files.

% If file changes (e.g. due to satellite reimaging) however difficult to tell
when data set has changed as versioning not well supported.

\\ \hline % Dynamic Environmental Data on the Web &data operated on changes %
(realtime feeds) (2) % &natural (sensor sources, simulations), (sub-set)
replication\\ % \hline \multicolumn{2}{|l|}{{\textbf{PowerGrids}}} \\ \hline

Stream processing over variable sources, rates and volumes, Event processing
over different patterns, Trigger analysis on-demand \emph{(Data Sources,
Notification, Enrichment, Orchestration)} %Analysis, % data changing at the
because application choosing more accurate sources % How is the selection made?
Is there a registry?  % There is a registry /catalog, it is distributed (sensors
catalogued by powergrid, buildings in campus database). Require distributed data
colletion to create a virtual single catalog.  % Using cloud computing to handle
scaling/elasticity for peaks. Not automated as yet but expected for the future.
& Dispersed event sources and sinks, Launch ensemble modeling on distributed
clusters and cloud \emph{(Data Sources, Notification, Orchestration)} %Analysis,

% There is a registry /catalog, it is distributed (sensors catalogued by
powergrid, buildings in campus database). Require distributed data colletion to
create a virtual single catalog.

% Output of computation itself is distributed - sent to various neighbourhoods /
locations.  % Some sent to human beings, sms, emails, many formats. Best effort
message delivery.

 \\

		\hline \multicolumn{2}{|l|}{{\textbf{Fusion (ITER)}}} \\ \hline
Stream and transform data between simulation sites on-demand \emph{(Data
Movement, Enrichment)} % (multiple distinct data % flows), on-the-fly %
transformation (8) & Run analysis on available, distributed resources
\emph{(Data Movement)} %Analysis % natural, coupled, transient sub-set
replicated \\ \hline \multicolumn{2}{|l|}{{\textbf{Industrial Incident
Notification and Response}}} \\ \hline Acquire different data sources and rates,
Processing determined by data \emph{(Data Sources, Data Access, Notification,
Orchestration)} %Analysis, % (1), % data itself changes (2), dynamic data
sources (3), processing changes % based on data %(5) & Distributed data sources,
Analysis by agents across institutions \emph{(Data Sources, Data Access)} % ,
Analysis % natural, coupled \\ \hline \multicolumn{2}{|l|}{{\textbf{MODIS Data
Processing}}} \\ \hline

% Satellite is already well known - static set of FTP sites.  -- % Pipeline is a
static structure - knows what it needs to do. Three different tasks looslely
coupled.  % Pipeline is setup for a community and once up and running will be
available to users for a while.

% How are users authorised to use the system? Portal, allows them to choose a
region. Basic authN?

		& Data from distributed FTP sources, Cloud queues for
coordination, Partitioned NoSQL data \emph{(Data movement, Data Storage,
Orchestration)} %Reliable message queues used as you don't want to have a black
spot in the sky - is it a real feature or a missing data. Try not to miss any
tiles!  \\ \hline %\multicolumn{3}{|l|}{{\textbf{Floating Sensors}}} \\ \hline %
&data rate (1), data itself (2), application causes % data changes (5) % &
spatial-temporal correlated, coupled \\ % \hline

\multicolumn{2}{|l|}{{\textbf{Distributed Network Intrusion Detection}}} \\
\hline Analyze dynamic network traffic rates \emph{(Data Movement)} % , Analysis
% data rate variable (1), data % highly dynamic changing (2), application
responds to data (5), % on-the-flight processing (8) & Run analysis at different
sites, Distribute generated data \emph{(Data Movement)} % Analysis, % natural,
data/compute tightly coupled \\ \hline
% \end{longtable}
%\end{footnotesize}

      \end{tabular}
    \end{center}
  \end{scriptsize}
\end{table}

% in the process pose two questions:
% \begin{enumerate}
% \item What programming systems and techniques are \emph{currently} % used in
implementing the application scenario?
% \item What elements identified in this section \emph{could} usefully % be
deployed or further explored in each application scenario?
% \end{enumerate}

% \jhanote{I think the second question is not addressed in the remainder % of
this subsection? Yogesh/Neil?}  % \ysnote{removed the "questions" and rephrased
it as a discussion}

\tablename~\ref{table:progSystems} summarizes the coordination approaches
currently dominant in each application case study. It is worth noting that, in
common with most long-lived systems, many of the cases have accreted substantial
code bases over time and so show little in the way of systematic
architecture. (One reason for the dominance of workflow pipelines may be that
such architectures are quite suitable for adding extra processing steps to
existing computations.)

\begin{table}
  \begin{scriptsize}
    \begin{center}
      \caption{Coordination approaches used by the
applications\label{table:progSystems}}
      \begin{tabular}{|p{6.3cm}|p{7.2cm}|} \hline Application & Current
approaches \\ \hline \hline NGS Analytics & pipelines that ship with
sequencers~\cite{illumina}; workflows~\cite{Taverna} \\ \hline ATLAS & custom
algorithms; job submission tied to WLCG \\ \hline LSST & custom layer; scripted
workflows; queries \\ \hline SOA Astronomy & web services \\ \hline Cosmic
Microwave Background & \textit{ad hoc} \\ \hline Sensor Network Application &
focused on data collection \\ \hline Climate & \textit{ad hoc} \\ \hline
Interactive Exploration of Environmental Data & visualization tools \\ \hline
Power Grids & stream processing; data mining \\ \hline Fusion (ITER) & scripted
pipeline \\ \hline Industrial Incident Notification and Response & agent-based
\\ \hline MODIS Data Processing & scripted pipeline \\ % \hline Floating Sensors
& \textit{ad hoc} \\ \hline Distributed Network Intrusion Detection & monolithic
packages \\ \hline
      \end{tabular}
    \end{center}
  \end{scriptsize}
\end{table}

The applications in \tablename~\ref{table:progSystems}, which make use of
scripted or \textit{ad hoc} pipelines, could in most cases be easily replaced
with a structured workflow engine. Doing so would bring immediate benefits in
terms of reproducibility, but also in terms of standardization of components and
interfaces.

%Standardization is not simply a software issue; it has %scientific significance
in the sense that data produced using %well-documented, well-used, and
well-debugged suites of components is %often more trusted -- and trustworthy --
than data produced using %locally-produced components that, while they may be
more innovative %and targeted, also run the risk of undetected assumptions and
errors.  %\jhanote{i think standardization of whatever data/component etc as an
% issue is independent of structured workflow engine. in fact i would % argue
that the causative arrow flows in the opposite direction: % structured workflow
engines do not help standardization but % standardizations does help structured
workflow engines..}  %\jhanote{IMO also still need to make it relevant to the
sub-section % aims?}

%\jhanote{NCH, we don't talk about standardization before this do we? so why %
further? Are we still trying to understand table IV? if yes, how % does this
relate?}  A further implication of standardization is that component reuse is
greatly improved. This is perhaps best illustrated by observing the increased
ability to automate the processing step over repeated runs.  The use of standard
component-based software techniques such as web services, coupled with web
technologies for metadata representation, assist in reuse, re-purposing and
integration, for instance into different workflow systems.

The {\em MapReduce} paradigm~\cite{MapReduce} and its
variants~\cite{twister,hmr} are the most prominent among the bulk data-parallel
frameworks which use bulk operators to express computations over collections of
data in an loosely coupled manner.  Applications such as Next Generation
Sequencing (NGS) (\S\ref{bioSilvia}) and Power Grids (\S\ref{power}) use
MapReduce for their
analytics~\cite{cloudburst,genometoolkit,Yin:mapreduce:2012}.

There are many examples of stream~\cite{chandrasekaran:telegraphcq2003,
abadi:borealis2005, liew:streamgraph2010} and event~\cite{ESPER, siddhi,
MicrosoftInsight} processing frameworks, and recent works have combined
streaming with more traditional file-based
workflows~\cite{ZinnCluster,Herath:CCGrid:2010}.  Applications such as the
Marine Sensor Network (\S\ref{sensorSimon}), Distributed Network Intrusion
Detection (\S\ref{detectionJon}), Power Grids (\S\ref{power}), and Industrial
Incident Notification and Response (\S\ref{Kees}) use stream and event
processing abstractions.

The view of scientific applications as pipelines hides significant differences
between (and within) domains, since the different stages of the pipeline receive
different emphases in different applications. Some applications must collect and
store all the incoming data, with only minor pre-processing to clean the data
ahead of the main processing. For others, a major constraint is to reduce the
incoming data stream as quickly as possible, perhaps discarding a large fraction
of the data before any processing occurs. In spite of these differences, we
believe the use of the pipeline abstraction is still useful.  Taking some of the
examples from \S\ref{sec:scenarios}, medical imaging and astrophysics often fall
into the first category, while particle physics (especially in the case of the
Large Hadron Collider) often falls dramatically into the second, to the extent
that principled data reduction forms one of the major scientific challenges for
the project.

There are also significant variations in where ``the pipeline'' begins and
ends. This represents the difficulty in differentiating between ``traditional''
applications and ``infrastructure'' applications. Many systems are driven by
data sets collected and shared widely, against which several different
applications may be applied.  Here the infrastructure application is the
collection of components used to allow the data sets to be accessed and analyzed
by the different applications.  Climate science is a good example of this
approach. In this case the data is ``pre-collected'' in the sense of existing
outside most of the applications, although its collection may have been a
complex task. The raw data required for gene sequencing (\S\ref{bioSilvia}), for
example, while being collected using a dedicated and very sophisticated
computerized processing system, would generally be regarded as ``given'' and not
part of the processing pipeline, since the pipeline would not affect the
collection task. Another way to look at this is that data is collected
\emph{independently} of its further processing.

Alternatively the application may be driven directly from a live data stream,
with processing happening on the fly. The front-end data collection subsystem is
becoming increasingly flexible and capable, with the introduction of sensor
networks to replace more passive approaches, and it may often make sense to
include the sensor network into the application workflow pipeline, such as for
Power Grids and Industrial Incident Response. All these choices have
implications for design and provisioning. For instance, the choice of which sort
of pipeline a particular application uses may well be fixed, or will change only
slowly as the experimental environment evolves. For example, we see some systems
that add simulation alongside experimental data collection to act as a
verification and validation step.

Revisiting the overall issue of infrastructure, which we defined as the hardware
and software needed to run the applications, we note that the specific hardware
isn't a primary issue.  The applications we've discussed all use a mix of the
systems that exist today, which have aspects of web, grid, and cloud, and can be
considered a hybrid of the three.  The key infrastructure components are really
the software and services on top of the hardware.  In this section, we've
discussed infrastructure to support data management and data analysis, as well
as the tools and systems used to program applications, and the specific
components used for the application in \S\ref{sec:scenarios}.  However, there is
not much commonality in the software and services that the applications use, so
the software and programming infrastructure we've described here is much more
broad than what is used by the applications we've studied.

%``This paper presents the design and the implementation of an %architecture for
the analysis of data streams in distributed %environments. In particular, data
stream analysis has been carried out %for the computation of items and itemsets
that exceed a frequency %threshold. The mining approach is hybrid, that is,
frequent items are %calculated with a single pass, using a sketch algorithm,
while %frequent itemsets are calculated offline by a further analysis. The
%architecture combines parallel and distributed processing to keep the %pace
with the rate of distributed data streams. In order to keep %computation close
to data, miners are distributed among the domains %where data streams are
generated. The paper also reports the %experiment results obtained with a
prototype of the architecture, %tested on a Grid composed of two domains
handling two different data %streams. Results confirm the scalability of the
approach and show that %it is possible to keep the pace of data production and
reduce the %communication overhead.''
%
%``.. In some cases streams of data must be analyzed in real time to %provide
information about trends, outlier values or regularities that %must be signaled
as soon as possible.  The need for online computation %is a notable challenge
with respect to classical data mining %algorithms [1], [2].  Important
application fields for stream mining %are as diverse as financial applications,
network monitoring, security %problems, telecommunication networks, Web
applications, sensor %networks, analysis of atmospheric data, etc.  A further
difficulty %originates when streams are distributed, and mining models must be
%derived not only for the data of a single stream, but for the %integration of
multiple and heterogeneous data streams [3]. This %scenario can occur in all the
application domains mentioned %before. For example, in a Content Distribution
Network, user requests %delivered to a Web system can be forwarded to any of
several servers %located in different and possibly distant places, in order to
serve %requests more efficiently and balance the load.''
%
%``..the architecture combines the parallel and distributed %paradigms, the
first to keep the pace with the rate of a %single data stream, by using multiple
miners (processors %or cores), the second to cope with the distributed %nature
of data streams. Miners are distributed among %the domains where data streams
are generated, in order %to keep computation close to data.''
%
%
%
%FlumeJava~\cite{flumeJava} is a distributed, reliable, and available %service
for efficiently collecting, aggregating, and moving large %amounts of log
data. It has a simple and flexible architecture based %on streaming data
flows. It is robust and fault tolerant with tunable %reliability mechanisms and
many failover and recovery mechanisms. The %system is centrally managed and
allows for intelligent dynamic %management. It uses a simple extensible data
model that allows for %online analytic application.

% Programming % Factored out by sd for various not-very-good reasons....
%\input{3dpas_programming}

\section{Discussion and Conclusions} \label{sec:conclusions}

% \jhanote{Andre L suggests a schematic to capture the main ideas from %
applications and the data properties. I think it is a good % idea. Omer: do you
want to think about this as you review/refine % this section}

%\jhanote{main message: (i) to application users/developers, you're not % alone,
(ii) infrastructure scientists: think of abstractions, and % how to build tools,
(iii) domain/application scientists message.}

%\jhanote{summarize contribution at front before 5.1 begins}

This section brings together some key observations from the applications and
infrastructure analysis carried out in this paper.  %as take away messages.
Section~\ref{sec:observations} summarize the dynamic and distributed
characteristics of data associated with the applications we've studied in this
work, attempting to infer more general characteristics that %can be applied may
apply to other similar applications.  In \S\ref{sec:discussion}, we suggest two
metrics that may be associated with such applications, primarily as a basis to
compare how compute and data processing (in terms of their number and
distribution) may be used to find common features between them. Such feature
analysis is useful, we believe, for two main reasons: (i) to identify
similarities between the applications covered in \S\ref{sec:scenarios} with
others that are in the design phase or under development, so that common
requirements can be identified; and (ii) to identify gaps in current software
tools and programming systems that such applications use, thereby leading to the
development of new tools and systems. As our application and infrastructure
coverage is, by necessity, limited, identifying common features also enables the
lessons learned to be generalized and applied more widely. We subsequently
discuss a set of architectures for D3 science, and pose some questions we
believe are pertinent as the community begins considering the next-generation
architectures for D3 applications. In \S\ref{sec:conclusion} we conclude by
providing a number of observations to inform applications and
cyberinfrastructure developers, based on common trends found in our study.

An important motivation for this work is to capture the state of the art in the
design and development of distributed dynamic applications and the associated
infrastructure currently used for their execution. We have aimed to try to
discern common approaches and challenges, and to identify any obvious gaps (or
opportunities). Using a common vocabulary and terminology to describe otherwise
distinct and unrelated applications allows potential application
users/developers to benefit from the insight that a common analytical framework
provides. For example, seeing how applications with similar patterns (e.g.,
pipelines) have addressed issues of dynamism and distribution enables common
solutions and facilitates the emergence of best practices even in the absence of
formal methods.  Similarly, our analysis sends a useful message to domain and
application scientists that the challenges and barriers they face are not unique
even if current solutions require a level of customization.

%\subsection{Common Characteristics}\label{sec:vectors}

% Given the application descriptions in \S\ref{sec:scenarios} and % subsequent
discussion about infrastructure characteristics in % \S\ref{sec:infrastructure},
we identify features that may be used to % compare these applications. Such
feature analysis is useful, we % believe, for two main reasons: (i) to identify
similarities between % the applications covered in \S\ref{sec:scenarios} with
others that are % in the design phase or under development, to enable common %
requirements to be identified; (ii) to identify gaps in current % software tools
and programming systems that such applications make use % of, thereby leading to
the development of new tools and systems. As % our application and
infrastructure coverage, by necessity, is limited, % identifying common features
also enables the lessons learned to be % generalized and applied more
widely. One important result is that we % can identify when particular
applications suit particular data-centric % programming models, e.g., stream
processing, MapReduce and Peer-2-Peer % networks.

%\item mechanism for supporting in-situ analysis at point of data % capture
(includes decision making to decide where analysis is % done)\note{move to
vector section.}
%\end{itemize}

%The first of the objectives above makes use of the four axes that %characterize
applications in the DPA survey paper, the intention of %which was to ascertain
how a number of real world applications made %use of distributed computing
infrastructure. Although we consider a %different set of applications here, the
aims are somewhat similar: to %both assess how existing real world applications
utilize distributed %computing infrastructure and also how they adapt to various
dynamic %characteristics that exist in their operating environment.

%Table~\ref{tab:characteristics} compares the applications across the four axes
identified at the beginning of this section.

\subsection{Observations}
\label{sec:observations}

In this subsection we summarize the source and nature of dynamic data that we
observe in the application surveyed in \S\ref{sec:scenarios}, and the types of
data distribution that we observe in such applications.

\subsubsection{Understanding Dynamic Data } %\alnote{in section 3 the three
classes for dynamism were different}

The applications in \S\ref{sec:scenarios} show that the dynamic characteristics
of many applications arise in two ways, due to data generation and placement or
due to the coordination mechanisms of the application.  The first of these
refers to data dynamism and the second to application dynamism (outlined in
\S\ref{sec:dynTypes}).

%\jhanote{Omer: this is inconsistent -- at least in terminology with % the 3
types of dynamism identified in Section 3: data, application % and
infrastructure} \rananote{Shantenu: this has been updated to % align with the
terminology in section 3. This is primarily referring % to the first two kinds
of dynamism mentioned there.}

\emph{Data generation, placement and consumption}: In many applications, data is
dynamically generated in response to a user request or when a particular
phenomenon of interest has been observed. Often this data is large in size and
must be placed close to the user undertaking the analysis. In some instances, it
may not be known where (and when) an analysis request may be made, and it is
therefore necessary to identify how replicas of the data could be produced and
placed across multiple possible locations.  In general, dynamic data arises due
to changing data properties, changing data volume (generation, consumption, or
processing), and changing data location or distribution.

% The coordination process of managing interaction between % components
undertaking data analysis, or the interaction of the % components and the
execution environment in which they are % deployed.

% Coordination mechanisms to support dynamic behavior remains an % important
feature across D3 Science applications.

\emph{Application Dynamism: } In many scientific applications the workflow
structure is well defined, but either the components used within this structure
or the location of the data that feeds the workflow may change dynamically.
Data coordination and planning (identifying how data is managed from ingest
through analysis to output) within such applications can range from statically
defined interactions between execution units and data sources to a fully
adaptable interaction using multi-agent systems.  As outlined in the Industrial
Incident Notification and Response application in (\S\ref{Kees}), the workflow
structure may also change based on a user request or the event being
observed. Data relating to the coordination of a components that make up the
application can therefore also be dynamic and change.  Exposing the data
planning used within an application makes it possible to modify or adapt it over
time.

%In the Industrial Incident Notification and Response
application\S\ref{sec:scenarios}.

%\jhanote{AL and I think this needs attention still. Either discuss % under data
planning or subsume dynamic coordination under % application dynamism}
\rananote{Agree with Dan's paragraph below as % a replacement -- and agree that
it should be subsumed under % application dynamism.}

%\katznote{SJ to move this paragraph to under application dynamism, and % merge
as needed.}\jhanote{done. NCH Can you check I have not mangled % the definition
of data planning please?}

%Exposing the data coordination/planning %used within an application (composed
of a number of different %components) to can be beneficial.  By exposing the
coordination %mechanism used, it becomes possible to modify or adapt it over
time.

%\jhanote{Omer: this is talking about changes in the infrastructure as % being
the cause and coordination changes as being the effect. The % last sentence of
the opening paragraph of this subsection, talks % about coordination being the
cause} \rananote{Shantenu: reference to % infrastructure has been removed to
avoid the confusion.}

%Data coordination within such applications can range from statically %defined
interactions between execution units and data sources to a %fully adaptable
interaction using multi-agent systems, as outlined in %the Industrial Incident
Notification and Response application in %\S\ref{sec:scenarios}.

%\jhanote{Omer: Multi-agent systems are not % called-out/discussed as such in
Section 2} \rananote{Shantenu: This % has been updated to refer to the
particular application when agents % have been described -- rather than the
section as a whole.}

% Workflow techniques and tools provide one of the most widely used % approaches
to expose such coordination to the user.

% \item An analysis of the characteristics of the D3 Science % applications in
\S\ref{sec:distdyndata} suggests that important % issues \jhanote{Important
issues of coordination? distribution? or % dynamism?}  become evident when
considering the ``data topology'' % of the application.

\subsubsection{Understanding Distributed Data}

From \S\ref{sec:distTypes} and \figurename~\ref{fig:figures_application}, we
find that applications with distributed data differ based on the amount or
volume of data distributed, the degree of distribution (scattered over one site
or many sites), and the dynamic aspects associated with distribution, e.g.,
latency and relative scheduling flexibility (and thus tolerance or constraints
on data transfer/aggregation).  Applications processing the largest volumes of
data (O(PB)) typically run on specialized data cyberinfrastructure (e.g., the
Atlas LHC Grid). Applications such as Fusion and NGS that run on smaller data
sets mainly utilize HPC environments. On the other end of the spectrum are
highly dynamic applications that process sensor data and make near-realtime
decisions based on streaming data, e.g., the power grid and sensor network
applications. These applications process much less data; often, sensor data is
also archived and aggregated for historical analyses, which leads to higher data
volumes.

\begin{figure}[ht] \centering
		\includegraphics[width=.7\textwidth]{figures/application.pdf}
                \caption{The distribution, dynamism, data volume characteristics
of the surveyed D3 applications. The clustering of applications with high
dynamism in the bottom right hand corner indicates a similar structure amongst
these applications, namely multiple (sensor-based) sources of data generation
with temporal variation in data generation. Although we do not explicitly
color/shade traditional and infrastructural applications, it is worth mentioning
that there is no correlation between application type and any of the three
``D''s; we attribute this to a strong dependence of degree-of-distribution
infrastructure available for both types of applications.
                  \label{fig:figures_application}}
\end{figure}

% \alnote{Think about how we can add a infrastructure dimension to this %
figure.} \jhanote{As AL and I discussed on a call, in addition to % AL's
suggestion, we need to also use this figure to integrate % section 2-4.}
\jhanote{Also, how would we provide intuitive % explanation of high versus low
dynamism?}  \jhanote{not to endlessly % quibble, but shouldn't ATLAS be more to
the right (and higher % dynamism?) and NGS more the left?}

%\jhanote{the following does not make sense to % me..}  Additional temporal
aspects pertain to the stage(s) of an %application in which data can be
distributed.  Because application stages often have a distribution that makes
sense to that stage, and because that distribution may not be appropriate for
all other phases, applications must either use one particular distribution that
may not work well in all phases or transform the data on input to some stages.
In particular, infrastructural applications require a specific distribution for
the data storage, and thus the stages that read and process the data will need
to make this choice.

As seen, how application data is distributed depends upon which stage of an
application is being considered. In spite of such variation, {\it data topology}
can be used to identify the flow of data through the application, with nodes
representing execution units and edges representing communication features.  %
The overall shape of such a data connectivity pattern gives a sense % of {\it
distributedness}; the thickness of edges gives a sense of % how data intensive
the application is; and the change in the shape % and thickness (and rate) and
number of edges gives a sense of the % dynamism of the application. \katznote{it
seems odd that we describe % a picture here but don't actually have any such
picture in the % paper.}  Additionally, how the application reacts to and
handles dynamism is an important factor and may also indicate that some aspects
of the data topology are unknown.

One aspect of managing data topology is ``data planning," which refers to how
the data is coordinated from ingest through analysis to output. In all cases,
there is a coordination of the data with code used to analyze or transform
it. In some cases these form clear, integrated stages where data flows from one
to another. In others the way that data enters and flows through the application
is more loosely defined and the execution of code is independent. Some examples
include the ATLAS experiments, which have a single point of primary data
generation which fans out to multiple analysis sites followed by mass exchange
of derived data, and sensor grids, which have multiple sources of data
generation collected to a single point of analysis followed by constrained point
of reuse.

We observe that pipelines provide a useful abstraction to represent the
application stages identified in
\figurename~\ref{fig:figures_application-stages}.  Many workflow tools also
enable simple data capture and processing pipelines to be defined, where each
pipeline stage may range from a complete application to a subroutine in Matlab
(for instance).  It is useful to note that pipelines may also be dynamically
constructed and adapted using context and monitoring information, as undertaken
in the Industrial Incident Notification and Response application studied in this
paper. In this instance, the actual pipeline stage to be invoked is triggered by
the availability of particular types of data, a feature also relevant in the
Power Grids application.  However, it is useful to note that different stages of
the pipeline receive different emphases in different applications. Some
applications must collect and store all incoming data, with only minor
pre-processing to clean the data ahead of the main processing. For others, a
major constraint is to reduce the incoming data stream as quickly as possible,
perhaps discarding a large fraction of the data before any processing
occurs. Reuse of pipeline stages across different applications must therefore be
undertaken with care.

Although dynamic and distributed properties have been characterized separately,
it is important to remember, as the above patterns indicate, that distribution
and dynamism become increasingly correlated at extreme scales.  For example,
applications that run on distributed infrastructures are often also highly
dynamic and vice versa.

% \jhanote{There is a problem here. I am not saying the following is %
incorrect, but ultimately coordination mechanisms are being touted % as a way to
support dynamic behaviour if not respond to dynamic % behaviour. However in
Section 3, this is not even hinted at. Needs % fixing IMO. The following should
be trivial: we need to relate which % dynamic behaviour -- infrastructure,
application or data dynamism, % is coordination influencing?}

% \jhanote{this is the first occurence of topology. can this be removed?  % SJ
to reduce last 2 paragraphs into a few sentences}

% \rananote{Update description of data topology -- relate it to % pipelines and
workflows mentioned earlier in the paper. Avoid % confusion in terminology}.
\jhanote{Also can this be related to % section 5.2?}

% Nonetheless, understanding the trend of say, $\beta_1$ as the degree % of
distribution changes can be insightful.

\subsection{Discussion}
\label{sec:discussion}

Now that we have discussed dynamic data and distributed data in the context of
specific applications, we can use this experience to understand the
characteristics that can be associated with D3 science applications, and the
characteristics of a computational infrastructure that may be most suited for D3
science applications.  We believe these are general trends that can be applied
to other applications we have not considered.

\subsubsection{Understanding Distributed Dynamic Applications: }
\label{sec:quantifyingdistapps}

In \S\ref{sec:distdyndata} we characterized dynamic and distributed properties
of the D3 Science applications. Our analysis was qualitative by necessity, due
to the fact that approaches to quantify dynamism and distributed properties were
often intimately tied to application/scenario definition and specific usage
modes.  Notwithstanding, to quantify better dynamic properties of an
application, we believe a term analogous to the use of Amdahl's
number~\cite{10.1109/ICDE.2000.839382} could be used to characterize `static'
data-intensive applications and systems:

\begin{itemize}
\item[] $\alpha$ = Num. of Compute Operations / Amount of Data (Consumed,
Generated or Transformed)
\end{itemize}

Note that this term is scale-invariant, i.e., a value of 1 could correspond to
very small or very large values of both numerator and denominator, and that it
does not capture the time scales over which data changes. However, we believe it
captures most of the non-temporal {\it dynamic} properties.

% \jhanote{we might consider removing the definitions. More text about % what is
$\alpha$ and $\beta$ and say it is early work..}

% defined above for dynamic characteristics,

Similarly, another term can be used to quantify the ``scale'' of distribution,
namely the degree of distribution as a function of the amount of data
transferred, where the degree of distribution can be defined as the number of
distinct data locations:

\begin{itemize}
\item[] $\beta$ = Amount of Data Transferred / Degree of Distribution
% \item[] $\beta_2$ = Number of Compute Operations / Degree of % Distribution
\end{itemize}

%Specifically, we propose $\beta$ as a ratio of the amount of data %transferred
to the number of distinct sites (i.e., degree) that it was %distributed over.

% either the computation involved or For a given application, the % degree of
distribution is often dependent or correlated with the % total number of
computational operations involved or the amount of % data transferred, e.g., for
small computational operations or % data-volumes there is typically a limited
amount of flexibility in % distribution before the overhead of distribution
(either source or % destination) dominates.

% and $\beta_2$ as a ratio of the number of compute operations to the % degree
of distribution; as can be expected, for low amounts of % compute units the
extent to which the tasks can be distributed is % limited, as overhead costs may
dominate.

%Notwithstanding, in this subsection, we provided a framework for

It has proven difficult to provide ``global'' quantitative values of these
ratios, as often the value of the numerators is subject to how an
application/scenario is defined, even more so for a ``class of applications'' as
opposed to a specific application. We believe, nonetheless, that these measures
suggest a possible way to quantify an application's dynamism and distribution
and thereby find possible common approaches solutions.  We plan to explore this
in our future work, and also hope others will consider building on these initial
concepts.

%and thus providing well-defined numbers is difficult,

\subsubsection{Towards an Architecture for D3 Science?}

\jhanote{Note to self: need smoothening}

A survey of data-intensive distributed applications on production
cyberinfrastructure leads us to posit that there are three primary macroscopic
architectures~\cite{pilot_clouds} and approaches for distributed data-intensive
applications: (i) localize all data into a ``cloud'' (or a large analytical
engine) -- the paradigm adopted by several genome projects and the cancer atlas,
(ii) decompose and distribute data to an appropriate number of
computing/analytical engines as available -- the paradigm employed by particle
physics for the discovery of the Higgs Boson (similar to the application in
\S\ref{WLCGSteve}), and (iii) a hybrid of the above two paradigms, wherein data
is decomposed and committed to several infrastructure, which in turn could be a
combination of either of the first two paradigms, similar to what is done in the
Power grids application (\S\ref{power}). \jhanote{improve description....replace
hybrid by bottom-up? }

It is obvious that the first architecture can be used for large volumes of data,
but surely there are limitations to the scalability or validity of this
model. How does this limitation change, qualitatively or quantitatively, if
dynamic data and/or applications are considered? This begets the question: how
fundamental an issue is {\it dynamic}?

The high-energy physics (HEP) community has self-organized towards the second
architectural approach.  Although distributed computing infrastructures
representing this approach, such as OSG and EGI, have been successful for this
community and application, broad uptake across other application types, modes,
and data volumes has not taken place. What lessons does the HEP experience teach
us for the other communities, such as the bioinformatics?  Can an infrastructure
conforming to a particular architecture and developed for a specific application
domain be generalized to another science domain, using the same underlying macro
architecture but by providing a different set of services and specialized data
cyberinfrastructure? Or is the architecture specific to a {\it fixed} range of
compute-data characteristics? If so, what are these characteristics?

%\subsection{Observations}
%
%\katznote{some text needed here.  Note that I promoted this from 5.1.5 to 5.2,
thinking that 5 might be best as discussion, observations, conclusions - how we
order 5.1 and 5.2 is still a question... Maybe call this `takeaway points'?}
%
%\katznote{Omer to go through and move the text/points to either Discussion or
Conclusions... This section will then empty/removed}

%\begin{itemize}

%\item In \S\ref{sec:scenarios}, we divided applications into groups: %
traditional and infrastructural applications.  Traditional % applications have
been built by one or more authors in order to % solve a particular problem or to
answer a science question.  % Infrastructural applications have been built by
groups or % communities to allow the solution of a set of problems, or to answer
% a set of science questions.  In both cases, we can observe that the % same set
of applications stages (as shown in %
\figurename~\ref{fig:figures_application-stages}) can be applied, % with APIs
defined as the inputs and outputs of those stages.  These % APIs are very
informal and loose for the traditional applications, % but more fixed and
conservative for the infrastructural % applications.  These APIs limit how new
paradigms can be used for % the infrastructural applications, unless those
paradigms happen to % fit the existing APIs.  Another way of saying this is that
the % abstractions for dynamism and distribution that can be used in the %
infrastructural applications are those abstractions that the % developers of the
original applications were aware of at the time % the infrastructural
applications were defined.  This leads to a % possible mismatch, where computer
scientists have new ideas that % could be applied to the class of problems that
an infrastructural % application was designed to solve, but that the
infrastructural % application doesn't allow, due to the fixed APIs.  Workflow
systems % are a possible way of solving this problem, as they are often %
sufficiently domain-independent that there does not seem to be a tie % of a
particular system to a particular infrastructural application.

%\item There exist patterns in the described applications; however, %
implementation and deployment specific details often mask these % patterns.  For
example, applications often differ in the way they % handle distribution and
dynamism, such as managing such as % data/compute localities. This makes it
difficult to support patterns % of distribution and dynamism in a
general-purpose fashion, for % anything but the simplest patterns.
%
%\item It is useful to observe that pipelines provide a useful % abstraction to
represent the application stages identified in %
\figurename~\ref{fig:figures_application-stages}.  Many workflow % tools also
enable simple data capture and processing pipelines to be % defined, where each
pipeline stage may range from a complete % application to a subroutine in Matlab
(for instance).  It is useful % to note that pipelines may also be dynamically
constructed and % adapted using context and monitoring information, as
undertaken in % the Industrial Incident Notification and Response application %
studied in this paper. In this instance, the actual pipeline stage % to be
invoked is triggered by the availability of particular types % of data, a
feature also relevant in the Power Grids application.  % However, it is useful
to note that different stages of the pipeline % receive different emphases in
different applications. Some % applications must collect and store all incoming
data, with only % minor pre-processing to clean the data ahead of the main %
processing. For others, a major constraint is to reduce the incoming % data
stream as quickly as possible, perhaps discarding a large % fraction of the data
before any processing occurs. Reuse of pipeline % stages across different
applications must therefore be undertaken % with care.

  % Managing and coordinating the behavior of pipeline execution, for %
supporting both file-based and streaming data, has been % extensively studied in
computer science (supporting synchronous % (blocking), asynchronous, concurrent,
threading semantics, for % instance).

%\item

%\jhanote{mismatch between abstraction levels, reference SAGA % paper ``leaky
abstractions''} \katznote{Shantenu to incorporate this % into the leaky
abstraction stuff above} \katznote{A possible problem is % that distributed
systems (infrastructures and perhaps applications) % are so different from each
other that there is no common system % abstraction that is both general enough
to work on all of them and % specific enough to be useful on any of them.}

%\item The trend toward more unstructured and more dynamic data sources lead to
%higher demands on data processing infrastructure.  % provide simple
abstractions or frameworks to % Higher-level abstraction need to provide
sufficient level of control % of infrastructure trade-offs (such as data/compute
localities), % while hiding a lot of the complexities (e.g.., the management of
% replicas).\jhanote{andre to try to link back to section 4 and % elaborate}

% This indicates the difficulty of % using existing abstractions and
infrastructures.

%\jhanote{Worth having in this section. Needs integration}

% \item Often the coupling between the various components/services that % are
part of the workflow can differ -- ranging from hardwired % interactions between
components to those that can be defined and % modified by an application
developer. We can therefore distinguish % between workflows that are embedded
within an application, to those % that are explicitly defined using a
composition tool (such as % Taverna, Triana, Kepler, etc). Workflow systems
often utilize a % variety of abstractions related to: (i) mapping of %
services/components to computational/data resources; (ii) data %
movement/migration; and (iii) coordination between various % components.  And, %
in many cases, they do not appear to be widely used where we might % think they
would be, perhaps due to lack of publicity, steep % learning curves, or other
issues.

%\end{itemize}

%\subsection{Points + Hypotheses}

\subsection{Conclusions}
\label{sec:conclusion}

Many fields, such as biology, that until recently were not characterized by
their intensive use of data are rapidly becoming data-driven.  The growth in
data has profound implications for the design and implementation of services and
infrastructure.  For example, there is a lifecycle to data that has often been
ignored until recently: the decreasing cost of storage has made it attractive to
archive \emph{everything}, just in case it becomes useful. This approach is
becoming untenable in many areas: some datasets are simply too big to archive,
and so must be summarized and their raw form discarded (possibly in real
time). Others are easier to regenerate than to store.  Still others are too
transient to justify their storage, and may in aggregate be too big to store
even when being small individually.

Additionally, it is useful to observe that the trend toward more unstructured
and more dynamic data sources lead to higher demands on data processing
infrastructure.~\cite{snia_2012} Any data science architecture needs to
explicitly recognize, capture, and manage the lifecycle requirements of
different datasets, within the constraints of available time and
storage. Associated with the growth in data volumes are increasing levels of
dynamism and distribution; sometimes distribution is logical, but sometimes the
data is physically distributed for reasons of performance or convenience. At
large scales, dynamism becomes intrinsic to the application and infrastructure;
sometimes it occurs because of changing experimental or analytic conditions,
sometimes due to the availability of infrastructure and resources.  %We close
out this paper with a brief %discussion of a range of issues associated with the
distribution and %dynamism of data.

We have identified many different application approaches, programming systems,
and infrastructure techniques applicable to dynamic data.  Consistent with
experience in other related fields, there seems no reason to expect that these
approaches and techniques will disappear in the near future. While a convergence
in approaches and techniques might be expected, this has proved to be elusive;
there remains a diversity in the design, implementation, and execution of
scientific codes.  It may even be argued that it is premature to have
convergence; after all, D3 Science is still an emerging domain.

From the previous discussion we can draw out some conclusions about the trends
within dynamic, distributed, data-intensive science. For instance, in
\S\ref{sec:scenarios}, we divided applications into groups: traditional and
infrastructural applications.  Traditional applications have been built by one
or more authors in order to solve a particular problem or to answer a science
question.  Infrastructural applications have been built by groups or communities
to allow the solution of a set of problems, or to answer a set of science
questions.  In both cases, we can observe that the same set of applications
stages (as shown in \figurename~\ref{fig:figures_application-stages}) can be
applied, with APIs defined as the inputs and outputs of those stages.  These
APIs are very informal and loose for the traditional applications, but more
fixed and conservative for the infrastructural applications.  % \jhanote{is
there a basis for these conjectures?} \katznote{just the % evidence of the
applications themselves} These APIs limit how new application paradigms
%\jhanote{new paradigms of what?}  can be introduced for the infrastructural
applications, unless those paradigms happen to fit the existing APIs.  Another
way of saying this is that the abstractions for dynamism and distribution that
can be used in the infrastructural applications are those abstractions that the
developers of the original applications were aware of at the time the
infrastructural applications were defined.  This leads to a possible mismatch,
where computer scientists have new ideas that could be applied to the class of
problems that an infrastructural application was designed to solve, but that the
infrastructural application doesn't allow, due to the fixed APIs.  New tools may
be developed to address a narrow set of features while ignoring the larger set
of problems that prevented earlier tools from being used.  These tools then may
be used be new applications as those applications are developed; however, they
are unlikely be used by the existing applications they were meant to help.

This is because applications (particularly infrastructural applications) aren't
developed in a vacuum, and that the constraints on the applications lead to new
tools that are or could be developed not being able to be applied in practice.
For example, while workflow systems are often sufficiently domain-independent
that there does not seem to be a tie of a particular system to a particular
infrastructural application, there are a very large number of workflow systems
that exist, and it is common in the authors' experience to hear about new ones
being developed from scratch because ``the ones that exist don't quite do what
is needed for my application or system.''

\katznote{previous 2 paragraphs to be re-read.  Old comments about them are
available as commented out comments.}  %\jhanote{Irrespective, I find this past
paragraph % somewhat speculative. And I don't see the point of it? WHat is the %
latter half trying to say: there are problems with APIs and % interfaces of 3D
applications and systems?} \katznote{It's saying % that applications aren't
developed in a vacuum, particularly % infrastructural applications, and that the
constraints on the % applications leads to new tools that are or could be
developed not % being able to be applied in practice.  In some ways, it's a call
for % an application abstraction or middleware, between the applications % and
the `infrastructure' as defined in \S4, much as perhaps AIMES is % trying to
build}\jhanote{Dan, I find your explanatory note more % convincing than the
sentence on workflow systems! Also there is no % mention of middleware as the
``impedance mismatch lowering layer'' % in the above. I propose we rewrite the
sentence starting, ``Workflow % systems are a possible way..} \katznote{ok, I
used some of my text % and turned around the sentence from Omer's optimistic
view to % my more pessimistic view.  If others disagree, we can change it back %
or compromise.}

As part of this survey, we also observe that there exist patterns in the
described applications; however, implementation and deployment specific details
can often mask these patterns.  For example, applications often differ in the
way they handle distribution and dynamism, such as managing such as data/compute
localities. This makes it difficult to support patterns of distribution and
dynamism in a general-purpose fashion, for anything but the simplest
patterns. In \S\ref{sec:quantifyingdistapps} we have attempted to initiate a
discussion in this context however by considering two ratios %(identified as
parameters $\alpha$ and $\beta$) to relate data volume and distribution with
compute operations performed by an application. We believe determining values
for these parameters would provide a useful starting point for comparing
infrastructure requirements of Big Data applications.  %\jhanote{I transposed
the last two paragraphs}

% \item Workflows seem to be a natural idea for expressing both % traditional
and infrastructural applications, and indeed, there are % a number of workflow
tools that are used in both cases.  It might % also seem that an infrastructural
application would have a specific % workflow system that is in common use, since
different % developers/users are all doing similar computational work, but this
% does not appear to be true, at least in the applications we studied.

  % \katznote{I would like to say that specific workflow systems are %
successful when embedded inside traditional applications, % because they then
become part of the application, but I am not % sure I have a good example here
from \S\ref{sec:scenarios}.  % Does anyone else?}

% \begin{itemize}

% \item \jhanote{Is the following observation, hypothesis and how do we % link
them?} \katznote{Can we say something about applications that % deal with files
vs those that deal with streams (stream-based are: % PowerGrid, Fusion, Incident
Response, network monitoring) PowerGrid % and Incident apps have centralized
stream processing;} Fusion and % network have distributed stream processing.
Not sure what to say % here.  Maybe this point isn't a point... \jhanote{Look at
tables % \ref{tab:app_distributed_processing} and %
\ref{table:progSystems}. From these tables can we say something % about
``common'' ways of say distribution for ``stream based % applications''}
\jhanote{analogy with abstractions for workflows: % num. of data elements, num
of streams, num. dynamic resources}

% \end{itemize}

\section*{Acknowledgements} This paper is an outcome of the UK e-Science
Institute sponsored Research Theme on Dynamic Distributed Data-Intensive
Programming Abstractions and Systems. This work is also sponsored by NSF
OCI-1059635.  We thank Jon Weissman and Jon Blower for useful initial
discussions, as well as 3DPAS workshop attendees. Discussions of the selected
applications benefited from the generous help of Adam Barker, Jon Blower, Julian
Borrill, Silvia Delgado Olabarriaga, Steve Fisher, Keith Jackson, Scott Klasky,
Bob Mann, Don Middleton, Kees Nieuwenhuis, Manish Parashar, Stephen Pascoe, and
Jon Weissman, though any errors are the responsibility of the authors.  Some of
the work by Katz was supported by the National Science Foundation while working
at the Foundation; any opinion, finding, and conclusions or recommendations
expressed in this material are those of the author(s) and do not necessarily
reflect the views of the National Science Foundation.

\bibliographystyle{unsrt} \bibliography{3dpas}

\end{document}

{\em Workflow} systems have already received significant attention in the
collaborative e-Science domain.  Such systems provide a useful means to
aggregate capability and subsequently deploy the resulting workflow over
multiple, distributed resources. Workflows can be ``abstract'', i.e., only the
connectivity graph is identified but is not bound to particular executables and
particular resources, or ``concrete'', i.e., the executable components are
identified along with their execution environment. Generally, an abstract
workflow needs to be mapped to a concrete workflow graph to enable
execution. This may be achieved dynamically using a variety of different
user-specified criteria, ranging from cost of execution (such as with the
availability of remotely accessed resources from Amazon.com~\cite{berriman11}),
Quality of Service, etc. This mapping can also be adaptive, especially if each
concrete workflow is dynamically composed based on event triggers from the
execution environment or constraints identified by a user. We may therefore
identify a range of possibilities:

\begin{itemize}

\item An abstract workflow graph is statically defined and mapped to a known
resource/infrastructure pool.

\item An abstract workflow graph is dynamically mapped to an infrastructure
based on user-specified constraints (related to cost of access, Quality of
Service, etc.)

\item An abstract workflow graph is dynamically enacted over a varying
infrastructure pool (often making use of an external resource management
system.)

\item An abstract workflow graph is used to dynamically compose a concrete
workflow using techniques from AI planning and model checking.

\end{itemize}

Workflow tools such as Taverna~\cite{Taverna}, Kepler~\cite{Kepler},
Trident~\cite{Trident} and others~\cite{iantaylorsbook} \jhanote{Are any of
these used by applications? If not I propose they not be mentioned} enable
reusable code for scientific analysis by expressing it as an explicit dataflow
or control flow composition of components, each serving a dedicated
function. This has simplified the development and the maintenance of complex
codes, both by allowing common component re-use and by highlighting the
structure of the processing pipeline. Workflow tools therefore provide both
infrastructure for supporting the construction and enactment/execution of
scientific applications and components that be used alongside application
specific infrastructure. The granularity of a workflow tool can therefore range
from simple components to complete infrastructures (i.e. enabling integration of
other components and supporting their enactment/execution).

### 2015-05-27 3dpas_intro.YS.tex
new files

--- 2015-05-27 3dpas_intro.tex
@@
 
-\def\volumeyear{2012} %\documentclass[acmjacm,acmnow]{acmtrans2m}
+\def\volumeyear{2012}
+%\documentclass[acmjacm,acmnow]{acmtrans2m}
 %\documentclass[conference,final]{IEEEtran}
@@
 
-\usepackage{graphicx} \usepackage{url} \usepackage{color} \usepackage{enumerate}
-\usepackage{longtable} %\usepackage{authblk} % Is used internally within the CPE
-Class \usepackage[firstpage]{draftwatermark}
-
-%\SetWatermarkLightness{0.7} %\SetWatermarkScale{4}
-
-%\usepackage{pdfsync} % \acmVolume{} % \acmNumber{} % \acmYear{} % \acmMonth{}
-
-%\usepackage{sectsty} %\usepackage{setspace}
-
-% \usepackage{ifpdf} % \usepackage{wrapfig} % %\usepackage[cm]{fullpage} %
-\usepackage{textcomp} % \usepackage{srcltx} % \usepackage{fancyhdr} %
-\usepackage{setspace} \usepackage{lscape} % \usepackage{longtable} %
-\usepackage{paralist}
-
-%\usepackage{hyperref} \usepackage[novbox]{pdfsync}
+\usepackage{graphicx}
+\usepackage{url}
+\usepackage{color}
+\usepackage{enumerate}
+\usepackage{longtable}
+%\usepackage{authblk} % Is used internally within the CPE Class
+%\usepackage[firstpage]{draftwatermark}
+
+%\SetWatermarkLightness{0.7}
+%\SetWatermarkScale{4}
+
+%\usepackage{pdfsync}
+% \acmVolume{}
+% \acmNumber{}
+% \acmYear{}
+% \acmMonth{}
+
+%\usepackage{sectsty}
+%\usepackage{setspace}
+
+% \usepackage{ifpdf}
+% \usepackage{wrapfig}
+% %\usepackage[cm]{fullpage}
+% \usepackage{textcomp}
+% \usepackage{srcltx}
+% \usepackage{fancyhdr}
+% \usepackage{setspace}
+\usepackage{lscape}
+% \usepackage{longtable}
+% \usepackage{paralist}
+
+%\usepackage{hyperref}
+\usepackage[novbox]{pdfsync}
 
@@
 
-% \ifpdf % \usepackage[pdftex, % colorlinks=true, % linkcolor=red, %
-citecolor=red, % filecolor=red, % ]{hyperref} % \else %
-\usepackage[hypertex]{hyperref} % \fi
-
-% Space saving %\usepackage[small,compact]{titlesec}
+% \ifpdf
+%              \usepackage[pdftex,
+%               colorlinks=true,
+%               linkcolor=red,
+%               citecolor=red,
+%               filecolor=red,
+%               ]{hyperref}
+% \else
+%   \usepackage[hypertex]{hyperref}
+% \fi
+
+% Space saving
+%\usepackage[small,compact]{titlesec}
 %\usepackage[small,it]{caption}
 
-% \renewcommand\floatpagefraction{.9} % \renewcommand\topfraction{.9} %
-\renewcommand\bottomfraction{.9} %\renewcommand\textfrare{.1}
-
-\setcounter{totalnumber}{50} \setcounter{topnumber}{50}
+% \renewcommand\floatpagefraction{.9}
+% \renewcommand\topfraction{.9}
+% \renewcommand\bottomfraction{.9}
+%\renewcommand\textfrare{.1}
+
+\setcounter{totalnumber}{50}
+\setcounter{topnumber}{50}
 \setcounter{bottomnumber}{50}
 
-%\textwidth = 6in %\textheight = 8.5 in %\oddsidemargin = 0.0 in
+%\textwidth = 6in
+%\textheight = 8.5 in
+%\oddsidemargin = 0.0 in
 %\evensidemargin = 0.0 in
 
-% \topmargin = 0.0 in % \headheight = 0.0 in % \headsep = 0.0 in
+% \topmargin = 0.0 in
+% \headheight = 0.0 in
+% \headsep = 0.0 in
 %\parskip = 0.0in
@@
 \parskip = 0.04in
-% \parindent = 0.0cm % \textfloatsep = 0.1in
-
-% \usepackage[draft,pdftex,colorlinks=true, linkcolor=blue,citecolor=blue, %
-urlcolor=blue]{hyperref}
-
-% \ifpdf % \pdfinfo{ /Author (Shantenu Jha et al.)  /Title (Programming
-Abstractions for Large-Scale Distributed Applications) } % \fi
-
-%\newcommand{\projectnamefull}{\textit{Programming Abstractions for %
-Large-Scale Distributed Applications} }
-
-\newcommand{\upup}{\vspace*{-0.5em}} \newcommand{\up}{\vspace*{-0.25em}}
-\newcommand{\I}[1]{\textit{#1}} \newcommand{\B}[1]{\textbf{#1}}
-\newcommand{\T}[1]{\texttt{#1}} \newcommand{\BI}[1]{\B{\I{#1}}}
-
- \newenvironment{shortlist}{ \vspace*{-0.8em}
-  \begin{itemize} \setlength{\itemsep}{-0.3em} }{
-  \end{itemize} \vspace*{-0.6em} }
-
- \newenvironment{shortenum}{ \vspace*{-0.8em}
-  \begin{enumerate} \setlength{\itemsep}{-0.3em} }{
-  \end{enumerate} \vspace*{-0.6em} }
-
-% \ifpdf % \DeclareGraphicsExtensions{.pdf, .png, .jpg} % \else %
-\DeclareGraphicsExtensions{.ps, .eps} % \fi
-
-\definecolor{orange}{rgb}{1.0,0.3,0.0} \definecolor{violet}{rgb}{0.75,0,1}
-\definecolor{darkgreen}{rgb}{0,0.6,0} \definecolor{cyan}{rgb}{0.2,0.7,0.7}
+% \parindent = 0.0cm
+% \textfloatsep = 0.1in
+
+% \usepackage[draft,pdftex,colorlinks=true, linkcolor=blue,citecolor=blue,
+%        urlcolor=blue]{hyperref}
+
+% \ifpdf
+% \pdfinfo{ /Author (Shantenu Jha et al.)  /Title (Programming Abstractions for Large-Scale Distributed Applications) }
+% \fi
+
+%\newcommand{\projectnamefull}{\textit{Programming Abstractions for
+%    Large-Scale Distributed Applications} }
+
+\newcommand{\upup}{\vspace*{-0.5em}}
+\newcommand{\up}{\vspace*{-0.25em}}
+\newcommand{\I}[1]{\textit{#1}}
+\newcommand{\B}[1]{\textbf{#1}}
+\newcommand{\T}[1]{\texttt{#1}}
+\newcommand{\BI}[1]{\B{\I{#1}}}
+
+ \newenvironment{shortlist}{
+        \vspace*{-0.8em}
+  \begin{itemize}
+  \setlength{\itemsep}{-0.3em}
+}{
+  \end{itemize}
+        \vspace*{-0.6em}
+}
+
+ \newenvironment{shortenum}{
+        \vspace*{-0.8em}
+  \begin{enumerate}
+  \setlength{\itemsep}{-0.3em}
+}{
+  \end{enumerate}
+        \vspace*{-0.6em}
+}
+
+% \ifpdf
+%   \DeclareGraphicsExtensions{.pdf, .png, .jpg}
+% \else
+%   \DeclareGraphicsExtensions{.ps, .eps}
+% \fi
+
+\definecolor{orange}{rgb}{1.0,0.3,0.0}
+\definecolor{violet}{rgb}{0.75,0,1}
+\definecolor{darkgreen}{rgb}{0,0.6,0}
+\definecolor{cyan}{rgb}{0.2,0.7,0.7}
 \definecolor{blueish}{rgb}{0.2,0.2,0.8}
 
-\long\def\comment#1{{\bf \textcolor{magenta}{\bf #1}}} \long\def\ccomment#1{{\bf
-\textcolor{blue}{\bf #1}}} \newcommand{\C}{\comment} \newcommand{\CC}{\ccomment}
+\long\def\comment#1{{\bf \textcolor{magenta}{\bf #1}}}
+\long\def\ccomment#1{{\bf \textcolor{blue}{\bf #1}}}
+\newcommand{\C}{\comment}
+\newcommand{\CC}{\ccomment}
 
@@
 
-\newif\ifdraft \drafttrue \ifdraft \newcommand{\katznote}[1]{
-{\textcolor{magenta} { ***Dan: #1 }}} \newcommand{\rananote}[1]{
-{\textcolor{cyan} { ***Omer: #1 }}} \newcommand{\omernote}[1]{ {\textcolor{cyan}
-{ ***Omer: #1 }}} \newcommand{\note}[1]{ {\textcolor{red} { #1 }}}
+\newif\ifdraft
+\drafttrue
+\ifdraft
+ \newcommand{\katznote}[1]{ {\textcolor{magenta}    { ***Dan:      #1 }}}
+ \newcommand{\rananote}[1]{ {\textcolor{cyan}    { ***Omer:     #1 }}}
+ \newcommand{\omernote}[1]{ {\textcolor{cyan}    { ***Omer:     #1 }}}
+ \newcommand{\note}[1]{ {\textcolor{red}    { #1 }}}
  \newcommand{\jhanote}[1]{  {\textcolor{blue}     { ***Shantenu: #1 }}}
@@
  \newcommand{\ysnote}[1]{  {\textcolor{violet}      { ***Yogesh: #1 }}}
-\newcommand{\sdnote}[1]{ {\textcolor{blueish} { ***Simon: #1 }}} %
-\usepackage[pdftex,colorlinks=true, linkcolor=blue,citecolor=blue, %
-urlcolor=blue]{hyperref} \else \newcommand{\katznote}[1]{}
-\newcommand{\rananote}[1]{} \newcommand{\note}[1]{} \newcommand{\jhanote}[1]{}
-\newcommand{\jonnote}[1]{} \newcommand{\alnote}[1]{} \newcommand{\nchnote}[1]{}
-\newcommand{\ysnote}[1]{} \fi
-
-\begin{document} \runningheads{S.~Jha et al.}{Understanding Applications and
-Infrastructure for D3 Science}
-
-%\title{3DPAS: Distributed Dynamic Data-intensive Programming % Abstractions and
-Systems for Large-Scale Science and Engineering}
+ \newcommand{\sdnote}[1]{  {\textcolor{blueish}      { ***Simon: #1 }}}
+% \usepackage[pdftex,colorlinks=true, linkcolor=blue,citecolor=blue,
+%       urlcolor=blue]{hyperref}
+\else
+ \newcommand{\katznote}[1]{}
+ \newcommand{\rananote}[1]{}
+ \newcommand{\note}[1]{}
+ \newcommand{\jhanote}[1]{}
+ \newcommand{\jonnote}[1]{}
+\newcommand{\alnote}[1]{}
+ \newcommand{\nchnote}[1]{}
+ \newcommand{\ysnote}[1]{}
+\fi
+
+\begin{document}
+\runningheads{S.~Jha et al.}{Understanding Applications and Infrastructure for D3 Science}
+
+%\title{3DPAS: Distributed Dynamic Data-intensive Programming
+%  Abstractions and Systems for Large-Scale Science and Engineering}
 
@@
 
-%\title{Towards Distributed Dynamic Data-intensive (D3) Science: A % Survey and
-Analysis of D3 Science Applications and % Cyberinfrastructure}
-
-\author{Shantenu Jha\affil{1,2}\corrauth\, Neil Chue Hong\affil{2}, Simon
-Dobson\affil{3}, Daniel S. Katz\affil{4,5}, Andre~Luckow\affil{1}, Omer
-Rana\affil{6}, Yogesh Simmhan\affil{7}}
-
-\address{\affilnum{1}Rutgers University\break \affilnum{2}University of
-Edinburgh\break \affilnum{3}University of St.~Andrews\break
-\affilnum{4}University of Chicago\break \affilnum{5}Argonne National
-Laboratory\break \affilnum{6}Cardiff University\break \affilnum{7}Indian
-Institute of Science\break }
+%\title{Towards Distributed Dynamic Data-intensive (D3) Science: A
+%  Survey and Analysis of D3 Science Applications and
+%  Cyberinfrastructure}
+
+\author{Shantenu Jha\affil{1,2}\corrauth\, Neil Chue Hong\affil{2},
+  Simon Dobson\affil{3}, Daniel S. Katz\affil{4,5},
+  Andre~Luckow\affil{1}, Omer Rana\affil{6}, Yogesh Simmhan\affil{7}}
+
+\address{\affilnum{1}Rutgers University\break
+\affilnum{2}University of Edinburgh\break
+\affilnum{3}University of St.~Andrews\break
+\affilnum{4}University of Chicago\break
+\affilnum{5}Argonne National Laboratory\break
+\affilnum{6}Cardiff University\break
+\affilnum{7}Indian Institute of Science\break
+}
 
@@
 
-% \author{Shantenu Jha} % \address{Rutgers University and University of
-Edinburgh; \texttt{shantenu.jha@rutgers.edu}} % \author{Neil Chue Hong} %
-\address{University of Edinburgh; \texttt{N.ChueHong@epcc.ed.ac.uk}} %
-\author{Simon Dobson} % \address{University of St.~Andrews;
-\texttt{simon.dobson@st-andrews.ac.uk}} % \author{Daniel S. Katz} %
-\address{University of Chicago and Argonne National Laboratory;
-\texttt{d.katz@ieee.org}} % \author{Andre Luckow} % \address{Louisiana State
-University; \texttt{luckow@cs.uni-potsdam.de}} % \author{Omer Rana} %
-\address{Cardiff University; \texttt{O.F.Rana@cs.cardiff.ac.uk}} %
-\author{Yogesh Simmhan} % \address{University of Southern California;
-\texttt{simmhan@usc.edu}} % \author{Jon Weissman} % \address{University of
-Minnesota; \texttt{jon@cs.umn.edu}}
+% \author{Shantenu Jha}
+% \address{Rutgers University and University of Edinburgh; \texttt{shantenu.jha@rutgers.edu}}
+% \author{Neil Chue Hong}
+% \address{University of Edinburgh; \texttt{N.ChueHong@epcc.ed.ac.uk}}
+% \author{Simon Dobson}
+% \address{University of St.~Andrews;  \texttt{simon.dobson@st-andrews.ac.uk}}
+% \author{Daniel S. Katz}
+% \address{University of Chicago and Argonne National Laboratory; \texttt{d.katz@ieee.org}}
+% \author{Andre Luckow}
+% \address{Louisiana State University; \texttt{luckow@cs.uni-potsdam.de}}
+% \author{Omer Rana}
+% \address{Cardiff University; \texttt{O.F.Rana@cs.cardiff.ac.uk}}
+% \author{Yogesh Simmhan}
+% \address{University of Southern California; \texttt{simmhan@usc.edu}}
+% \author{Jon Weissman}
+% \address{University of Minnesota; \texttt{jon@cs.umn.edu}}
 
@@
 
-\begin{abstract} % Many problems at the forefront of science, engineering,
-medicine, % and the social sciences are increasingly complex and %
-interdisciplinary due to the plethora of data sources and % computational
-methods available today.  A common feature across many % of these problems is
-the amount and diversity of data and % computation that must be integrated to
-yield insights. Such data is % increasingly large-scale and distributed: the
-datasets, their % locations, and associated scheduling decisions are %
-time-dependent.
-
-% Applications need access % to these disparate data-sets and to the methods
-that operate on % them.  Data-intensive applications need first-class support
-for % distributed heterogeneous platform execution.
-
-The 3DPAS Research Theme (funded by the eSI, Edinburgh and US NSF) investigated
-applications that have these characteristics; in other words, it operated at the
-triple point of {\it distributed} and {\it dynamic} and {\it data-intensive}
-(D3) attributes.  This document represents the conclusions of the 3DPAS Theme.
+\begin{abstract}
+%   Many problems at the forefront of science, engineering, medicine,
+%   and the social sciences are increasingly complex and
+%   interdisciplinary due to the plethora of data sources and
+%   computational methods available today.  A common feature across many
+%   of these problems is the amount and diversity of data and
+%   computation that must be integrated to yield insights. Such data is
+%   increasingly large-scale and distributed: the datasets, their
+%   locations, and associated scheduling decisions are
+%   time-dependent.
+
+%   Applications need access
+%   to these disparate data-sets and to the methods that operate on
+%   them.  Data-intensive applications need first-class support for
+%   distributed heterogeneous platform execution.
+
+The 3DPAS Research Theme (funded by the eSI, Edinburgh and US NSF)
+investigated applications that have these characteristics; in other
+words, it operated at the triple point of  {\it distributed} and {\it dynamic} and
+{\it data-intensive} (D3) attributes.  This document represents the conclusions
+of the 3DPAS Theme.
 \end{abstract}
@@
 \tableofcontents
-%\newpage %\listoftables
-\newpage \fi
+%\newpage
+%\listoftables
+\newpage
+\fi
 
@@
 
-\jhanote{NCH points out that we are inconsistent in the order of the ``D''s. SJ
-to fix globally. Good catch Neil!}
-
-The landscape of scientific and enterprise computing is being fundamentally
-altered by the current surge in data. The resulting upheaval, grandiosely named
-BigData science by the Lilliputians in the data science community, represents a
-new reality where the volumes of data produced by scientific applications are
-increasing rapidly, driven by advancing technologies (e.g., increasing compute
+\jhanote{NCH points out that we are inconsistent in the order of the
+  ``D''s. SJ to fix globally. Good catch Neil!}
+
+The landscape of scientific and enterprise
+computing is being fundamentally altered by the current surge in data. The
+resulting upheaval, grandiosely named BigData science by the Lilliputians
+in the data science community, represents a new reality where
+the volumes of data produced by scientific applications are increasing
+rapidly, driven by advancing technologies (e.g., increasing compute
 capacity and higher resolution sensors) and decreasing costs for data
 acquisition and storage~\cite{tony-propaganda}. With the increase in data
-volumes, types, and sources of data, an important concomitant development has
-been an increase in the requirement for and complexity in the processing of
-data. Data are often naturally distributed and have to be processed in situ, and
-occasionally data that are localized have to be distributed in order to be
-processed, either for reasons of performance or for collaborative analytics.
-Furthermore when processing large volumes of data, data source/destination, data
-production and transfer rates, and data processing capacity can all vary over
-time. Thus properties such as resource scheduling, data placement, and transfer
-decisions, that were statically determinable at small scales, emerge as dynamic
-and important considerations at large scales of data and distribution.  The
-BigData science problem is therefore only partly about large volumes of data.
-
-Our focus is on a subspace of the BigData problem space, namely that which is
-characterized by the distributed and dynamic aspects of data-intensive
-science. We introduce the concept of Dynamic Distributed Data-intensive science
-and refer to this subspace of BigData science as D3 science.  In order to
-appreciate the D3 subspace we will briefly discuss the importance and role of
-distribution and the pervasiveness of dynamical aspects.
+volumes, types, and sources of data, an important concomitant
+development has been an increase in the requirement for and complexity
+in the processing of data. Data are often naturally distributed and have
+to be processed in situ, and occasionally data that are localized
+have to be distributed in order to be processed, either for reasons
+of performance or for collaborative analytics.  Furthermore when
+processing large volumes of data, data source/destination, data
+production and transfer rates, and data processing capacity can
+all vary over time. Thus properties such as resource scheduling, data
+placement, and transfer decisions, that were statically determinable at
+small scales, emerge as dynamic and important considerations at
+large scales of data and distribution.  The BigData science problem
+is therefore only partly about large volumes of data.
+
+Our focus is on a subspace of the BigData problem space, namely that
+which is characterized by the distributed and dynamic aspects of
+data-intensive science. We introduce the concept of Dynamic
+Distributed Data-intensive science and refer to this subspace of
+BigData science as D3 science.  In order to appreciate the D3 subspace
+we will briefly discuss the importance and role of distribution and
+the pervasiveness of dynamical aspects.
 
@@
 
-% \jhanote{I intend to convert these bullet points into a paragraph % about the
-importance of treating distributed data}
-
-Although the capacity to store data centrally is growing faster than Moore's
-Law, it is not sufficient in all cases to be able to ``dump the data'' in one
-place (or in the cloud.)  Oftentimes the costs (data transfer, storage,
-consequences of catastrophic failure) prevent the centralization of data, and
+% \jhanote{I intend to convert these bullet points into a paragraph
+%   about the importance of treating distributed data}
+
+Although the capacity to store data centrally is
+growing faster than Moore's Law, it is not sufficient in all cases to
+be able to ``dump the data'' in one place (or in the cloud.)
+Oftentimes the costs (data transfer, storage, consequences of
+catastrophic failure) prevent the centralization of data, and
 decentralization---storing or place data in multiple locations---become prudent.
 
-Conversely, not all production of data is centralized; in some cases there are
-multiple producers (and consumers) of data, not to mention one-of-a-kind data
-that must be shared across multiple consumers.  Furthermore in collaborative
-science, data output from one analytical engine are often the input for the next
-stage of the analytical pipeline.  Last but not least, distribution via
-replication is a tested and effective approach to reliability and
-fault-tolerance; we will argue that it is also potentially a great performance
-enhancer and a means to ensure data consistency too.
-
-Whether data are a priori distributed or localized, a critical challenge is to
-find an efficient and effective means to either distribute the initially
-localized data, or manage the distribution (or localization) of a priori
-distributed data.
+Conversely, not all production of data is centralized; in some cases
+there are multiple producers (and consumers) of data, not to mention
+one-of-a-kind data that must be shared across multiple consumers.
+Furthermore in collaborative science, data output from one analytical
+engine are often the input for the next stage of the analytical
+pipeline.  Last but not least, distribution via replication is a
+tested and effective approach to reliability and fault-tolerance; we
+will argue that it is also potentially a great performance enhancer
+and a means to ensure data consistency too.
+
+Whether data are a priori distributed or localized, a critical
+challenge is to find an efficient and effective means to either
+distribute the initially localized data, or manage the distribution
+(or localization) of a priori distributed data.
 
 % \begin{itemize}
-% \item Costs (data transfer, storage costs) for centralized data too % high
+% \item Costs (data transfer, storage costs) for centralized data too
+%   high
 % \item data locally required for analysis (store data where it is used)
 % \item volume too large to store centrally or move around
-% \item required for scaling the system (load balancing, local reads are %
-cheaper)
-% \item reliability and fault tolerance (no single point of failure, %
-redundancy)
+% \item required for scaling the system (load balancing, local reads are
+%   cheaper)
+% \item reliability and fault tolerance (no single point of failure,
+%   redundancy)
 % \item data consistency
@@
 
-% \ysnote{distributed to ensure security also? Privacy % controls on moving data
-out of a location or country.}
-
-The challenges of distributed data-intensive applications span beyond the
-management and curation of large volumes of data. A critical requirement is to
-make sense of this data, i.e., transform the data into information en route to
-assembling knowledge. Among other things, the transformation from data to
-information requires extensive amounts of computational processing of the data;
-thus the data-intensive nature of the BigData science problem is strongly
-coupled with compute-intensive requirements.  As the data are often distributed,
-a consequence is that many attributes of distributed systems permeate to
-data. Thus almost naturally, distributed data have attributes of dynamism, both
-at the data level as well as the infrastructure level.
-
-That is not to say all data-intensive applications have elements of both data
-and compute dynamism and are always distributed, but it is a reflection that at
-extreme scales, dynamism and and distribution become increasingly important and
-often correlated issues.  Thus, analyzing the applications along an analytical
-space spanned by the three ``D''s of data-intensive, distributed, and dynamic
-provides a general and effective framework.  The diverse set of requirements to
-support D3 scenarios points to the need for a careful understanding of D3
-characteristics as well as sophisticated, extensible, and skillfully architected
-infrastructure and programming systems.
-
-With recent attention on ``Big Data'' problems, several classification schemes
-have been proposed; one that has garnered attention is the
-3Vs~\cite{gartner3v}. These three dimensions (Volume, Velocity and Variety) just
-describe the characteristics of the data. %, rather than describing the
-interplay between Big %Data and applications that operate over them.
-\jhanote{Does someone have the authentic reference for the 4Vs?} \katznote{see
-http://www.villanovau.com/university-online-programs/what-is-big-data/ but it's
-not `authoritative'} In contrast, D3 emphasizes the nature of scientific
-applications as characterized by their data attributes. Furthermore the D3
-attributes are considered as a collective property, as opposed to different
-disjoint attributes.  Thus we believe there is a fundamental distinction between
-the D3 and 3Vs classification schemes.
-
-%\ysnote{Since we're clearly positioning around BigData, % one question that is
-going to come up is how are the 3D's related to % the 3V's -- volume, velocity
-and variety -- of Big Data. We may want % to explicitly address that early
-on. For e.g. ``While the three % dimensions of Volume, Velocity and Variety of %
-BigData~\cite{gartner3v}, or 3V's, are well recognized in industry, % these
-narrowly consider the characteristics of the data rather than % the interplay
-between Big Data and applications that operate over % them as we do through the
-3D's. Furthermore, our emphasis is also on % the role of Big Data in science,
-which is less studied (and arguably % more diverse and complex) than business
-applications. That said, the % volume and velocity dimensions of the 3V's
-partially capture the % data-intensive and dynamic aspects of our study.''}
-\jhanote{we do not discuss explicitly the 4th D, i.e. diversity (courtesy AL) it
-is assumed but not discussed}
+% \ysnote{distributed to ensure security also? Privacy
+%           controls on moving data out of a location or country.}
+
+The challenges of distributed data-intensive applications span beyond
+the management and curation of large volumes of data. A critical
+requirement is to make sense of this data, i.e., transform the data
+into information en route to assembling knowledge. Among other things,
+the transformation from data to information requires extensive amounts
+of computational processing of the data; thus the data-intensive
+nature of the BigData science problem is strongly coupled with
+compute-intensive requirements.  As the data are often distributed, a
+consequence is that many attributes of distributed systems permeate to
+data. Thus almost naturally, distributed data have attributes of
+dynamism, both at the data level as well as the infrastructure
+level.
+
+That is not to say all data-intensive applications have elements of
+both data and compute dynamism and are always distributed, but it is a
+reflection that at extreme scales, dynamism and and distribution
+become increasingly important and often correlated issues.  Thus,
+analyzing the applications along an analytical space spanned by the
+three ``D''s of data-intensive, distributed, and dynamic provides a
+general and effective framework.  The diverse set of requirements to
+support D3 scenarios points to the need for a careful understanding of
+D3 characteristics as well as sophisticated, extensible, and
+skillfully architected infrastructure and programming systems.
+
+With recent attention on ``Big Data'' problems, several classification
+schemes have been proposed; one that has garnered attention is the
+3Vs~\cite{gartner3v}. These three
+dimensions (Volume, Velocity and Variety) just describe the
+characteristics of the data. %, rather than describing the interplay between Big
+%Data and applications that operate over them. 
+\jhanote{Does someone
+  have the authentic reference for the 4Vs?} \katznote{see http://www.villanovau.com/university-online-programs/what-is-big-data/ but it's not `authoritative'} In contrast,  D3 
+emphasizes the nature of scientific applications as characterized
+by their data attributes. Furthermore the D3 attributes are considered
+as a collective property, as opposed to  different disjoint attributes.
+Thus we believe there is a fundamental distinction between the D3 and
+3Vs classification schemes.
+
+%\ysnote{Since we're clearly positioning around BigData,
+%  one question that is going to come up is how are the 3D's related to
+%  the 3V's -- volume, velocity and variety -- of Big Data. We may want
+%  to explicitly address that early on. For e.g. ``While the three
+%  dimensions of Volume, Velocity and Variety of
+%  BigData~\cite{gartner3v}, or 3V's, are well recognized in industry,
+%  these narrowly consider the characteristics of the data rather than
+%  the interplay between Big Data and applications that operate over
+%  them as we do through the 3D's. Furthermore, our emphasis is also on
+%  the role of Big Data in science, which is less studied (and arguably
+%  more diverse and complex) than business applications. That said, the
+%  volume and velocity dimensions of the 3V's partially capture the
+%  data-intensive and dynamic aspects of our study.''}  
+\jhanote{we do
+  not discuss explicitly the 4th D, i.e. diversity (courtesy AL) it is
+  assumed but not discussed}
 
@@
 
-%\alnote{Should we introduce a degree of distribution? machine, % cluster,
-data-center, WAN, Wireless} \jhanote{A figure akin to our % the figure in our
-pilot-data paper maybe Ok to help establish the % vocabulary. Not sure this
-would be a conflict or self-plagiarism?  % Either way, this should be in the
-sub-section on ``Scope''}
-
-Distribution can occur on many levels.  %, as shown in
-%\figurename~\ref{fig:figures_terminology}.  In this paper, we use {\em locally
-distributed} to indicate that data/compute resides on multiple nodes within one
-data center.  {\em Geographically distributed} describes the distribution of
-data and/or compute across multiple data centers. Further, we define the degree
-as the number of data sources and sites that are used for storing and processing
-the data. A distributed application is an application that is composed of
-entities (i.e., either storage/compute resource or a data source) from at least
-two physical locations. In this work, our focus is almost exclusively on
-geographically distributed data and applications.
-
-%\jhanote{terminology defined elsewhere in the paper isn't defined % here.. so a
-bit misleading to call this part terminology. AL points % out ``dynamic'' as an
-example.}
-
-%\katznote{AL thinking about changes to the figure, according to SJ - % now SJ
-to remove this figure completely.}
-
-%\katznote{where are the data sources %in this figure?}\alnote{added}
-
-%\begin{figure}[htbp] % \centering
+%\alnote{Should we introduce a degree of distribution? machine,
+%  cluster, data-center, WAN, Wireless} \jhanote{A figure akin to our
+%  the figure in our pilot-data paper maybe Ok to help establish the
+%  vocabulary. Not sure this would be a conflict or self-plagiarism?
+%  Either way, this should be in the sub-section on ``Scope''}
+
+Distribution can occur on many levels.
+%, as shown in
+%\figurename~\ref{fig:figures_terminology}.
+In this paper, we use {\em locally distributed} to indicate that
+data/compute resides on multiple nodes within one data center.
+{\em Geographically distributed} describes the distribution of data and/or compute
+across multiple data centers. Further, we define the degree as the
+number of data sources and sites that are used for storing and
+processing the data. A distributed application is an application that
+is composed of entities (i.e., either storage/compute resource or a
+data source) from at least two physical locations. In this work, our
+focus is almost exclusively on geographically distributed data and
+applications.
+
+%\jhanote{terminology defined elsewhere in the paper isn't defined
+%  here.. so a bit misleading to call this part terminology. AL points
+%  out ``dynamic'' as an example.}
+
+%\katznote{AL thinking about changes to the figure, according to SJ -
+%  now SJ to remove this figure completely.}
+
+%\katznote{where are the data sources
+%in this figure?}\alnote{added}
+
+%\begin{figure}[htbp]
+%	\centering
 %		\includegraphics[width=0.7\textwidth]{figures/terminology.pdf}
@@
 
-% (\S\ref{sec:prog_infra}). \katznote{removed this ref, as it seemed % out of
-place to have a ref to a subsection in the intro, as the only such ref}
-
-% \subsection*{Distributed Data} % One specific attribute that we believe is
-pervasive but not % necessarily obvious is ``dynamic data.''  Additionally the
-requirement % of many applications impose a requirement that data is dynamic.
-
-% \jhanote{I do not know why floating sensors was a subsection? As an % examplar
-for D3 Science? If so, made it a subsubsection..}
-
-% \katznote{as I wrote in the title, it got moved from apps and won't be there %
-any more, but there was some thought that it might be useful as an example %
-application here.  Feel free to delete it if you don't think it adds anything
-here.}
-
-% \subsubsection{Floating Sensors -- material than can be used in 1 as % an
-example application that is not going to be discussed in %
-\S\ref{sec:scenarios}}
-
-% This {\em traditional application} constructs real-time hydrodynamics models,
-and forecasts the behavior of a complicated water system: the Sacramento % - San
-Joaquin River delta.  It begins after data generation and includes data
-transformation \& decisions, storage, and processing.
+% (\S\ref{sec:prog_infra}). \katznote{removed this ref, as it seemed
+% out of place to have a ref to a subsection in the intro, as the only such ref}
+
+% \subsection*{Distributed Data}
+% One specific attribute that we believe is pervasive but not
+% necessarily obvious is ``dynamic data.''  Additionally the requirement
+% of many applications impose a requirement that data is dynamic.
+
+% \jhanote{I do not know why floating sensors was a subsection? As an
+%   examplar for D3 Science? If so, made it a subsubsection..}
+
+%   \katznote{as I wrote in the title, it got moved from apps and won't be there
+%   any more, but there was some thought that it might be useful as an example
+%   application here.  Feel free to delete it if you don't think it adds anything here.}
+
+% \subsubsection{Floating Sensors -- material than can be used in 1 as
+%   an example application that is not going to be discussed in
+%   \S\ref{sec:scenarios}}
+
+% This {\em traditional application} constructs real-time hydrodynamics models, and forecasts the behavior of a complicated water system: the Sacramento
+% - San Joaquin River delta.  It begins after data generation and includes data transformation \& decisions, storage, and processing.
 
@@
 
-% The application integrates both static and dynamic sensors streaming %
-real-time data with a high-resolution hydrodynamics model running on % the NERSC
-Magellan cloud testbed. The model uses sensor data to % initialize and shape the
-run, and the model output is in turn used to % drive the real-time positioning
-of the dynamic sensors. The output of % this process is a map of the current and
-future behavior of the % modeled watershed.
-
-% Custom built dynamic sensors are deployed in the watershed along with %
-traditional statically placed sensors that provide data. These sensors %
-communicate over a variety of different network infrastructures, e.g., % GSM,
-CDMA, 802.11, wired ethernet, etc.  and the application acquires % the data over
-the internet from the sensor platforms.  The application % processes the sensor
-data in real-time as it is streamed.
+% The application integrates both static and dynamic sensors streaming
+% real-time data with a high-resolution hydrodynamics model running on
+% the NERSC Magellan cloud testbed. The model uses sensor data to
+% initialize and shape the run, and the model output is in turn used to
+% drive the real-time positioning of the dynamic sensors. The output of
+% this process is a map of the current and future behavior of the
+% modeled watershed.
+
+% Custom built dynamic sensors are deployed in the watershed along with
+% traditional statically placed sensors that provide data. These sensors
+% communicate over a variety of different network infrastructures, e.g.,
+% GSM, CDMA, 802.11, wired ethernet, etc.  and the application acquires
+% the data over the internet from the sensor platforms.  The application
+% processes the sensor data in real-time as it is streamed.
 
@@
 
-% The sensor are all streaming real-time data to the application. The position
-of the dynamic sensors % themselves change over time.
+% The sensor are all streaming real-time  data to the application. The position of the dynamic sensors
+% themselves change over time.
 
@@
 
-% The first full, end-to-end, field test of the system is scheduled for % May
-2011.
-
-%\jhanote{If some one knows a bit more about Bayesian %
-approaches/inference/statistics and its 'implication' on dynamic % data via
-dymamic models, i'd love to hear / learn more}
-
-\jhanote{refine next two paragraph...} Analogous to the distinction between
-programming in the large versus system level programming, there is a distinction
+%  The first full, end-to-end, field test of the system is scheduled for
+%  May 2011.
+
+%\jhanote{If some one knows a bit more about Bayesian
+%  approaches/inference/statistics and its 'implication' on dynamic
+%  data via dymamic models, i'd love to hear / learn more}
+
+\jhanote{refine next two paragraph...} Analogous to the distinction between programming in
+the large versus system level programming, there is a distinction
 between the ``cyberinfrastructure/systems considerations/design at the
-large-scale'' versus ``CI considerations at the small-scale''. This paper is
-primarily about CI/systems considerations at the large-scale.  Searching for
-common solutions and patterns at the large-scale is likely to be more fruitful
-than at small-scales where customized tools and optimization of low-level
-systems features is likely to return greater yield.
-
-Thus, subjecting a range of otherwise distinct applications to a common
-analytical framework provides advantages to the infrastructure scientists and
-cyberinfrastructure experts, namely suggestions for where effort is both needed
-and likely to yield maximum returns. In other words, as D3 applications evolve,
-our analysis can provide guidance on where to focus effort on infrastructure.
-Last but not least, the advantages of providing a common but extensible
-vocabulary that can be used by the entire community, application end-users,
-developers and infrastructure scientists alike, in an emerging field cannot be
-overstated. One may argue that such vocabulary and terminology has been missing
-from the practice of distributed cyberinfrastructure in general, with less than
-desirable consequences.
-
-%\katznote{some of this is repeating ideas from the para ``An important %
-motivation...''  Perhaps these two paras should be % combined?}\jhanote{DSK,
-this should not be an issue now, given that % this has now moved upstream}
+large-scale'' versus ``CI considerations at the small-scale''. This
+paper is primarily about CI/systems considerations at the large-scale.
+Searching for common solutions and patterns at the large-scale is
+likely to be more fruitful than at small-scales where customized tools
+and optimization of low-level systems features is likely to return
+greater yield.
+
+Thus, subjecting a range of otherwise distinct applications to a
+common analytical framework provides advantages to the infrastructure
+scientists and cyberinfrastructure experts, namely suggestions for
+where effort is both needed and likely to yield maximum returns. In
+other words, as D3 applications evolve, our analysis can provide
+guidance on where to focus effort on infrastructure.  Last but not
+least, the advantages of providing a common but extensible vocabulary
+that can be used by the entire community, application end-users,
+developers and infrastructure scientists alike, in an emerging field
+cannot be overstated. One may argue that such vocabulary and
+terminology has been missing from the practice of distributed
+cyberinfrastructure in general, with less than desirable consequences.
+
+%\katznote{some of this is repeating ideas from the para ``An important
+%  motivation...''  Perhaps these two paras should be
+%  combined?}\jhanote{DSK, this should not be an issue now, given that
+%  this has now moved upstream}
 
@@
 
-% Before a discussion of the applications, infrastructure and research %
-methodology, a brief mention of the motivation and thereby scope of % this work
-is in order.
-
-% Without a doubt, the era of {\it BigData} is here. However, BigData % is not
-just an issue of large volumes of data, it is also about the
+% Before a discussion of the applications, infrastructure and research
+% methodology, a brief mention of the motivation and thereby scope of
+% this work is in order.
+
+% Without a doubt, the era of {\it BigData} is here. However, BigData
+% is not just an issue of large volumes of data, it is also about the
 
 %One motivation of this work is to understand the complexity and
-%characteristics of data that emerge as a consequence of scale, namely %the
-dynamic properties and the distribution of applications, and the %complex
-interplay between them that arises.
-
-One motivation of this work is to understand the dynamic properties and the
-distribution of applications, and the complex interplay between them that arises
-as a consequence of scale.  Additionally, we are motivated by an attempt to
-understand infrastructure---existing capabilities, trends, and
-limitations---available to the complexity, challenges and characteristics of
-dynamic and distributed data at scale.
-
-We would like the reader to keep in mind three primary context and scope-setting
-points:
-
-%\katznote{a lot of this paragraph seems too repetitive of the previous %
-section} \jhanote{Dan, I fixed the previous paragraph. OK?}
+%characteristics of data that emerge as a consequence of scale, namely
+%the dynamic properties and the distribution of applications, and the
+%complex interplay between them that arises.
+
+One motivation of this work is to understand the dynamic properties
+and the distribution of applications, and the complex interplay
+between them that arises as a consequence of scale.  Additionally, we
+are motivated by an attempt to understand infrastructure---existing
+capabilities, trends, and limitations---available to the complexity,
+challenges and characteristics of dynamic and distributed data at
+scale.
+
+We would like the reader to keep in mind three primary context and
+scope-setting points:
+
+%\katznote{a lot of this paragraph seems too repetitive of the previous
+%  section} \jhanote{Dan, I fixed the previous paragraph. OK?}
 
@@
 
-\item This {\it survey} is not meant to be complete, either in scope of domains
-covered or types of infrastructure and application surveyed.  They by definition
-represent the interests and experience of the authors.
-
-\item The focus is on applications and infrastructure that arise from science
-and engineering projects in academia; this is not exclusive but is more a
-guiding principle.  While we are cognizant of enterprise applications and
-infrastructure, and recognize the impressive advances made therein, we also
-acknowledge that the factors such as scale, distribution, open-source, and
-interoperability influence the solutions employed and infrastructures used; it
-is important to therefore focus on problems that emerge as in the space defined
-by these ``academic constraints''.
-
-\item Almost by definition, any attempt to capture the state-of-the-art of such
-a dynamic and emerging domain is bound to be fraught with limitations and issues
-of scoping.  A complete and rigorous survey of the landscape of BigData is bound
-to be obsolete by the time it emerges; so given the rate of change, we asked
-what could be done. What we present is a partial analysis motivated by a set of
-existing and active D3 applications, their trials and tribulations, successes
-and solutions. We believe there is merit in capturing the state-of-play as is,
-deriving a research agenda and set of specific yet crosscutting issues, and
-presenting these to the community.
+\item This {\it survey} is not meant to be complete, either in scope
+  of domains covered or types of infrastructure and application
+  surveyed.  They by definition represent the interests and experience
+  of the authors.
+
+\item The focus is on applications and infrastructure that arise from
+  science and engineering projects in academia; this is not exclusive
+  but is more a guiding principle.  While we are cognizant of
+  enterprise applications and infrastructure, and recognize the
+  impressive advances made therein, we also acknowledge that the
+  factors such as scale, distribution, open-source, and
+  interoperability influence the solutions employed and infrastructures
+  used; it is important to therefore focus on problems that emerge as
+  in the space defined by these ``academic constraints''.
+
+\item Almost by definition, any attempt to capture the
+  state-of-the-art of such a dynamic and emerging domain is bound to
+  be fraught with limitations and issues of scoping.  A complete and
+  rigorous survey of the landscape of BigData is bound to be obsolete
+  by the time it emerges; so given the rate of change, we asked what
+  could be done. What we present is a partial analysis motivated by a
+  set of existing and active D3 applications, their trials and
+  tribulations, successes and solutions. We believe there is merit in
+  capturing the state-of-play as is, deriving a research agenda
+  and set of specific yet crosscutting issues, and presenting these
+  to the community.
 \end{itemize}
 
-%\jhanote{I think these two points can be ignored. To be discussed. But % either
-way this should go in the scope subsection} % \katznote{Dan and Shantenu request
-that Neil and Omer think % about the next two points, and either flesh them out
-with text % or remove them.}
+%\jhanote{I think these two points can be ignored. To be discussed. But
+%  either way this should go in the scope subsection}
+%  \katznote{Dan and Shantenu request that Neil and Omer think
+%  about the next two points, and either flesh them out with text
+%  or remove them.}
 %
 %\begin{itemize}
-%\item How does this differ from Scientific Databases? How does it % relate to
-Scientific Data Management and Scientific data flow
-%\item Reference to EU 2030 Objectives (part of the Project Europe 2030 %
-Report) -- underpins ``Data Intensive Research'' from an EU % perspective
-(strategy) + relate this to NSF ``CIF21'' vision
+%\item How does this differ from Scientific Databases? How does it
+%  relate to Scientific Data Management and Scientific data flow
+%\item Reference to EU 2030 Objectives (part of the Project Europe 2030
+%  Report) -- underpins ``Data Intensive Research'' from an EU
+%  perspective (strategy) + relate this to NSF ``CIF21'' vision
 %\end{itemize}
 
-%\jhanote{some points to ensure we've addressed. I cut and paste some % comments
-we received from the DPA survey paper that we should be % ensure at addressed --
-implicitly or explicitly}
+%\jhanote{some points to ensure we've addressed. I cut and paste some
+%  comments we received from the DPA survey paper that we should be
+%  ensure at addressed -- implicitly or explicitly}
 %
@@
 %
-%(i) It is difficult to argue about the completeness of the presented %material.
+%(i) It is difficult to argue about the completeness of the presented
+%material.
 %
@@
 %
-%(iii) Furthermore, it is unclear if the applications %are indeed representative
-of any urgent need to execute across %distributed systems as the authors claim.
-%
-%(iv) At the same time, there is not enough description of the computer %science
-behind the applications.
-%
-%(v) Several applications seem similar to each other and, %as far as I can tell,
-do not add anything new.
-%
-%(v) The section is more a %laundry list of applications - I would have expected
-to see a %synthesis of different application and, for example, a classification
-%into different classes. No general requirements of applications are %derived. A
-smaller number of more representative sample applications %(not covering nine
-pages) would have been better.
+%(iii) Furthermore, it is unclear if the applications
+%are indeed representative of any urgent need to execute across
+%distributed systems as the authors claim.
+%
+%(iv) At the same time, there is not enough description of the computer
+%science behind the applications.
+%
+%(v) Several applications seem similar to each other and,
+%as far as I can tell, do not add anything new.
+%
+%(v) The section is more a
+%laundry list of applications - I would have expected to see a
+%synthesis of different application and, for example, a classification
+%into different classes. No general requirements of applications are
+%derived. A smaller number of more representative sample applications
+%(not covering nine pages) would have been better.
 %
 %(vi) My main problem with the paper is the research methodology.  The paper
-%uses 6 hand-picked applications to derive a vector of attributes for %each of
-them, derive some patterns from observing these vectors, and %uses these
-patterns to perform a distributed computing tools gap %analysis.
-%
-%(vii) It is unclear how important the applications used in this %study are.
-%
-%(viii) Your lack of depth in explaining the applications leaves the %reader
-empathetic.  It's important that the reader feels that they can %learn something
-from reading this paper.  How is what you are saying %applicable to me (a
-computer scientist or computational domain scientist)?
+%uses 6 hand-picked applications to derive a vector of attributes for
+%each of them, derive some patterns from observing these vectors, and
+%uses these patterns to perform a distributed computing tools gap
+%analysis.
+%
+%(vii) It is unclear how important the applications used in this
+%study are.
+%
+%(viii) Your lack of depth in explaining the applications leaves the
+%reader empathetic.  It's important that the reader feels that they can
+%learn something from reading this paper.  How is what you are saying
+%applicable to me (a computer scientist or computational domain scientist)?
 %
 %(ix) Have you considered adopting a different approach, like conducting a
-%survey to glean the likes/dislikes of distributed application %developers?  In
-essence focus on the development process, as opposed %to applications.  With a
-large enough pool of responses, you may come %up with statistically significant
-conclusions that will be of interest %to the wider community.
+%survey to glean the likes/dislikes of distributed application
+%developers?  In essence focus on the development process, as opposed
+%to applications.  With a large enough pool of responses, you may come
+%up with statistically significant conclusions that will be of interest
+%to the wider community.
 %
@@
 
-This paper originated in work at the UK e-Science Institute (eSI) at the
-University of Edinburgh, in a research theme examining Dynamic Distributed
-Data-intensive Programming Abstractions and Systems, called
-3DPAS~\cite{3dpas-theme}.  The description of this theme's work was:
-
-\begin{quote} Many problems at the forefront of science, engineering, medicine,
-and the social sciences, are increasingly complex and interdisciplinary due to
-the plethora of data sources and computational methods available today. A common
-feature across many of these problem domains is the amount and diversity of data
-and computation that must be integrated to yield insights.
-
-For many complex data-intensive applications, moving the data may have
-restrictions. Increasingly important type of data-intensive applications are
-data-driven applications. For example, data is increasingly large-scale,
-distributed arising from sensors, scientific instruments \& simulations. Such
-data-driven applications will involve computational activities triggered as a
-consequence of independent data creation; thus it is imperative for an
-application to be able to respond to unplanned changes in data load or
-content. Understanding how to support dynamic computations is a fundamental, but
-currently a critical missing element in data-intensive computing.
-
-The 3DPAS theme seeks to understand the landscape of dynamic, distributed,
-data-intensive computing: the programming models and abstractions, the run-time
-and middleware services, and the computational infrastructure. It will analyze
-existing tools and services, identify missing pieces and new abstractions, and
-propose practical solutions and best practices.
+This paper originated in work at the UK e-Science Institute (eSI) at
+the University of Edinburgh, in a research theme examining Dynamic
+Distributed Data-intensive Programming Abstractions and Systems,
+called 3DPAS~\cite{3dpas-theme}.  The description of
+this theme's work was:
+
+\begin{quote}
+Many problems at the forefront of science, engineering, medicine, and the social sciences, are increasingly complex and interdisciplinary due to the plethora of data sources and computational methods available today.
+A common feature across many of these problem domains is the amount and diversity of data and computation that must be integrated to yield insights.
+
+For many complex data-intensive applications, moving the data may have restrictions.
+Increasingly important type of data-intensive applications are data-driven applications.
+For example, data is increasingly large-scale, distributed arising from sensors, scientific instruments \& simulations.
+Such data-driven applications will involve computational activities triggered as a consequence of independent data creation; thus it is imperative for an application to be able to respond to unplanned changes in data load or content.
+Understanding how to support dynamic computations is a fundamental, but currently a critical missing element in data-intensive computing.
+
+The 3DPAS theme seeks to understand the landscape of dynamic, distributed, data-intensive computing: the programming models and abstractions, the run-time and middleware services, and the computational infrastructure.
+It will analyze existing tools and services, identify missing pieces and new abstractions, and propose practical solutions and best practices.
 \end{quote}
 
-The theme held three workshops, and in each, application scientists and
-computing technologists were invited to give presentations on their work, and
-then to be involved in discussions to organize and understand the presented
-materials.  Between the workshops, the theme organizers also drafted outlines of
-a report that was a predecessor of this paper, for discussion in the next
-workshop.  Overall, there were about 50 workshop attendees.
-
-In order to select applications and gather information about them, during the
-various workshop discussions, when an application that ``felt'' new was
-mentioned, we asked the person who mentioned it to provide a written set of
-answers to the following questions about the application:
+The theme held three workshops, and in each, application scientists and computing technologists
+were invited to give presentations on their work, and then to be involved in discussions to
+organize and understand the presented materials.  Between the workshops, the theme
+organizers also drafted outlines of a report that was a predecessor of this paper, for
+discussion in the next workshop.  Overall, there were about 50 workshop attendees.
+
+In order to select applications and gather information about
+them, during the various workshop discussions, when an
+application that ``felt'' new was mentioned, we asked the person who
+mentioned it to provide a written set of answers to the following questions about the
+application:
 
@@
 \item How is the application used to do this?
-\item What infrastructure is used? (including compute, data, network,
-instruments, etc.)
+\item What infrastructure is used? (including compute, data, network, instruments, etc.)
 \item What dynamic data is used in the application?
@@
 \item How diverse is the data?
-\item Please feel free to also talk about the current state of the application,
-if it exists today, and any specific gaps that you know need to be overcome.
+\item Please feel free to also talk about the current state of the
+application, if it exists today, and any specific gaps that you know
+need to be overcome.
 \end{enumerate}
 
-We then used the text of these answers to describe each of the applications (in
-\S\ref{sec:scenarios}) in terms of:
+We then used the text of these answers to describe each of the applications (in \S\ref{sec:scenarios}) in terms of:
 
@@
 \item What does the application do? What problem does it solve?
-\item How does it do it? How does the application run?  What infrastructure does
-it need/use?
+\item How does it do it? How does the application run?  What infrastructure does it need/use?
 \item What are the big data aspects of the application?
@@
 \item What are the dynamic aspects  of the application?
-\item What else is important about the application?  What does it do well?
-Poorly? How was the app written? What tools does it use?
+\item What else is important about the application?  What does it do well?  Poorly? How was the app written? What tools does it use?
 \end{itemize}
 
-After the workshops, the theme organizers and a set of attendees who were
-interested in participating started discussing how to organize and analyze the
-applications and the technologies, including infrastructure, programming systems
-and abstractions.  These participants are the current authors of this paper, and
-this paper is the result of that process.
-
-As part of the theme, two 3DPAS workshops~\cite{3dapas,D3-escience} were
-organized at major conferences: HPDC 2011 and IEEE eScience 2011.  These
-workshops had both peer-reviewed and invited papers.
-
-A combination of invitation-only focused discussion meetings and broader
-community engagement events were used to ensure that the applications surveyed
-and ideas captured in this paper are representative of current community
-thinking as well as of intellectual relevance.
+After the workshops, the theme organizers and a set of attendees who
+were interested in participating started discussing how to organize
+and analyze the applications and the technologies, including
+infrastructure, programming systems and abstractions.  These
+participants are the current authors of this paper, and this paper is
+the result of that process.
+
+As part of the theme, two 3DPAS workshops~\cite{3dapas,D3-escience}
+were organized at major conferences: HPDC 2011 and IEEE eScience 2011.
+These workshops had both peer-reviewed and invited papers.
+
+A combination of invitation-only focused discussion meetings and
+broader community engagement events were used to ensure that the
+applications surveyed and ideas captured in this paper are
+representative of current community thinking as well as of
+intellectual relevance.
 
@@
 
-The rest of this paper consists of the application scenarios
-(\S\ref{sec:scenarios}), a discussion of the aspects of distribution and
-dynamism in the applications (\S\ref{sec:distdyndata}), a description of the
-software infrastructure required for D3 applications % Yogesh: This section's
-description seemed more verbose than others, hence commenting out details % ,
-such as % software frameworks, tools, and middleware and platform services
-(\S\ref{sec:infrastructure}), analysis and characterization of the applications
-and conclusions (\S\ref{sec:conclusions}).
+The rest of this paper consists of
+the application scenarios (\S\ref{sec:scenarios}),
+a discussion of the aspects of distribution and dynamism in the
+applications (\S\ref{sec:distdyndata}),
+a description of the software infrastructure required for D3 applications
+% Yogesh: This section's description seemed more verbose than others, hence commenting out details
+% , such as
+% software frameworks, tools, and middleware and platform services
+(\S\ref{sec:infrastructure}),
+analysis and characterization of the applications and
+conclusions (\S\ref{sec:conclusions}).
 
@@
 
-Starting with the experience of the attendees of the 3DPAS workshops, we have
-assembled a set of D3 ``applications'' that we describe in this section.  We
-realize that we have overloaded the term application here.  In this document,
-some of our applications are really {\em traditional applications}, meaning a
-program that is developed by a user or project to try to answer a science
-question, usually with some input data needed and some output data produced.
-Applications in this category are: NGS Analytics (\S\ref{bioSilvia}), CMB
-(\S\ref{astroJulian}), Fusion (\S\ref{fusionScott}), Industrial Incident
-Notification and Response (\S\ref{Kees}), MODIS Data Processing
-(\S\ref{modisKeith}), and Distributed Network Intrusion Detection
-(\S\ref{detectionJon}).
+Starting with the experience of the attendees of the 3DPAS workshops,
+we have assembled a set of D3 ``applications'' that we describe in
+this section.  We realize that we have overloaded the term application
+here.  In this document, some of our applications are really {\em
+  traditional applications}, meaning a program that is developed by a
+user or project to try to answer a science question, usually with some
+input data needed and some output data produced.  Applications in this
+category are: NGS Analytics (\S\ref{bioSilvia}), CMB
+(\S\ref{astroJulian}), Fusion (\S\ref{fusionScott}), Industrial
+Incident Notification and Response (\S\ref{Kees}), MODIS Data
+Processing (\S\ref{modisKeith}), and Distributed Network Intrusion
+Detection (\S\ref{detectionJon}).
 
 Another group of applications are {\em infrastructural applications}.
-Applications of this type are collaborative, in that they depend on a context
-that is often agreed upon by a community, most often in terms of how data is
-stored by the community and how it is accessed, as well as, more often than not,
-a weak consensus on the infrastructure to be used. %\jhanote{Dan: Please confirm
-you are OK with this modified % defintion of infrastructural application} This
-allows different sub-applications (that we again refer to as `applications') to
-focus on different stages, such as generating data, or processing data.
-Answering a science question now may involve a set of applications that need to
-be run in series, perhaps in different stages, that may be run by different
-groups that do not frequently interact.  This type of application includes:
-ATLAS/WLCG (\S\ref{WLCGSteve}), LSST (\S\ref{astroAdam}), SOA Astronomy
-(\S\ref{astroSOAAdam}), Sensor Network Application (\S\ref{sensorSimon}),
-Climate (\S\ref{climateDan}), Interactive Exploration of Environmental Data
+Applications of this type are collaborative, in that they depend on a
+context that is often agreed upon by a community, most often in terms
+of how data is stored by the community and how it is accessed, as well
+as, more often than not, a weak consensus on the infrastructure to be
+used. %\jhanote{Dan: Please confirm you are OK with this modified
+%  defintion of infrastructural application}
+This allows different
+sub-applications (that we again refer to as `applications') to focus
+on different stages, such as generating data, or processing data.
+Answering a science question now may involve a set of applications
+that need to be run in series, perhaps in different stages, that may
+be run by different groups that do not frequently interact.  This type
+of application includes: ATLAS/WLCG (\S\ref{WLCGSteve}), LSST
+(\S\ref{astroAdam}), SOA Astronomy (\S\ref{astroSOAAdam}), Sensor
+Network Application (\S\ref{sensorSimon}), Climate
+(\S\ref{climateDan}), Interactive Exploration of Environmental Data
 (\S\ref{envJon}), and Power Grids (\S\ref{power}).
 
-In other words, multiple infrastructural applications are used together in
-particular science domains.  Furthermore, infrastructural applications involve
-more than just the end user (scientist), and typically require exerting control
-at layers below, i.e., ``programming the system (infrastructure)'', as opposed
-to a traditional applications, where end user effort and control is often
+In other words, multiple infrastructural applications are used together in particular
+science domains.  Furthermore, infrastructural
+applications involve more than just the end user (scientist), and
+typically require exerting control at layers below, i.e.,
+``programming the system (infrastructure)'', as opposed to a
+traditional applications, where end user effort and control is often
 confined to a well-defined application kernel.
 
-% \jhanote{Need better distinction between infrastructure and % traditional
-applications. They are the same at the scale-up and % scale-out level, but
-different in scale-across and people. Another % issue is that about ``control
-and configure'', i.e traditional % applications only exert control and configure
-the application layer, % whilst infrastructure applications need to configure
-the lower % ie. infrastructural layer too}
-
-% \katznote{I don't agree in general that more is needed, but if you % think so,
-please give it a try.  I also don't agree specifically % that infrastructure
-applications `configure' the lower layer.  % Rather, I would say they assume a
-particular, community-defined % (perhaps de facto) layer.  Finally, while I
-understand scale-up % and scale-out, and I don't know what `scale-across' is.}
-
-Some of the applications we describe are really groups of applications, meaning
-that we are grouping together a number of independent programs, written by
-different authors, that are trying to answer the same type of science question
-and thus, where a single user will only run one of them at a time.
-%\katznote{Shantenu, please check %the preceding and see if this is ok, in the
-context of the whole %paragraph and section.}  These codes and their developers
-may either be competing or collaborating, and sometimes both, informally known
-as collabetition. Specifically, this is the case for NGS Analytics
-(\S\ref{bioSilvia}), SOA Astronomy (\S\ref{astroSOAAdam}), and Interactive
-Exploration of Environmental Data (\S\ref{envJon}).  (Note that this issue is
-orthogonal to traditional vs. infrastructural applications: NGS Analytics is
-traditional, where SOA Astronomy and Interactive Exploration of Environmental
-Data are infrastructural.)  In each of these examples, we have chosen one
-specific application that is meant to represent a possibly large number of
-comparable applications.
-
-%\jhanote{Dan: This paragraph seems to suggest some % similarity with
-infrastructural applications, but doesn't quite do % so. Can we make a bit more
-explicit the intent. For example, If we % break into ``collaborating'' vs
-``competing'' and put NGS into a % ``competing'' category the remainder of the
-paragraph above would be % a useful and fair elaboration on infrastructural
-applications?  % Reading forward, Section 5.1, it appears to be consistent with
-the % ``observations'' section. For example, `` Some of the applications % we
-describe are really groups of applications, meaning that we are % grouping
-together a number of independent programs, written by % different authors.
-These independent programs might be competing % means to answer the same type of
-science question and thus are not % intended to run together, or they could be a
-set of smaller % applications collaborating to answer the same overall science %
-question.''}  \katznote{some changes made above.  I'm not sure why % NGS is
-competition and the others are collaboration - to me they are % all a bit of
-both.} \jhanote{NGS is classified as a traditional % application; others as
-infrastructural and as having multiple % distinct applications. Thus it is
-natural to have collaborative.  % But as NGS is said to be a traditional,
-different applications can % be perceived to be competitive. I agree that the
-term competitive is % not perfect} \katznote{I've added a note to say that this
-(applications % selected to represent groups) is orthogonal to traditional vs %
-infrastructural.  (This is actually why this is a different paragraph.)  % Does
-this help?}
-
-%\katznote{not sure which category is right for: Power %Grids (\S\ref{power})
-Yogesh, what do you think?}  %\ysnote{The infra category is appropriate. i.e. no
-change required}
+% \jhanote{Need better distinction between infrastructure and
+%   traditional applications. They are the same at the scale-up and
+%   scale-out level, but different in scale-across and people. Another
+%   issue is that about ``control and configure'', i.e traditional
+%   applications only exert control and configure the application layer,
+%   whilst infrastructure applications need to configure the lower
+%   ie. infrastructural layer too}
+
+% \katznote{I don't agree in general that more is needed, but if you
+%   think so, please give it a try.  I also don't agree specifically
+%   that infrastructure applications `configure' the lower layer.
+%   Rather, I would say they assume a particular, community-defined
+%   (perhaps de facto) layer.  Finally, while I understand scale-up
+%   and scale-out, and I don't know what `scale-across' is.}
+
+Some of the applications we describe are really groups of
+applications, meaning that we are grouping together a number of
+independent programs, written by different authors, that are trying to
+answer the same type of science question and thus, where a single user
+will only run one of them at a time.  %\katznote{Shantenu, please check
+%the preceding and see if this is ok, in the context of the whole
+%paragraph and section.}
+These codes and their developers may either be
+competing or collaborating, and sometimes both, informally known as
+collabetition. Specifically, this is the case for NGS Analytics
+(\S\ref{bioSilvia}), SOA Astronomy (\S\ref{astroSOAAdam}), and
+Interactive Exploration of Environmental Data (\S\ref{envJon}).  (Note
+that this issue is orthogonal to traditional vs. infrastructural applications:
+NGS Analytics is traditional, where SOA Astronomy and Interactive
+Exploration of Environmental Data are infrastructural.)  In
+each of these examples, we have chosen one specific application that
+is meant to represent a possibly large number of comparable
+applications.
+
+%\jhanote{Dan: This paragraph seems to suggest some
+%  similarity with infrastructural applications, but doesn't quite do
+%  so. Can we make a bit more explicit the intent. For example, If we
+%  break into ``collaborating'' vs ``competing'' and put NGS into a
+%  ``competing'' category the remainder of the paragraph above would be
+%  a useful and fair elaboration on infrastructural applications?
+%  Reading forward, Section 5.1, it appears to be consistent with the
+%  ``observations'' section. For example, `` Some of the applications
+%  we describe are really groups of applications, meaning that we are
+%  grouping together a number of independent programs, written by
+%  different authors.  These independent programs might be competing
+%  means to answer the same type of science question and thus are not
+%  intended to run together, or they could be a set of smaller
+%  applications collaborating to answer the same overall science
+%  question.''}  \katznote{some changes made above.  I'm not sure why
+%  NGS is competition and the others are collaboration - to me they are
+%  all a bit of both.} \jhanote{NGS is classified as a traditional
+%  application; others as infrastructural and as having multiple
+%  distinct applications. Thus it is natural to have collaborative.
+%  But as NGS is said to be a traditional, different applications can
+%  be perceived to be competitive. I agree that the term competitive is
+%  not perfect} \katznote{I've added a note to say that this (applications
+%  selected to represent groups) is orthogonal to traditional vs
+%  infrastructural.  (This is actually why this is a different paragraph.)
+%  Does this help?}
+
+%\katznote{not sure which category is right for: Power
+%Grids (\S\ref{power})  Yogesh, what do you think?}
+%\ysnote{The infra category is appropriate. i.e. no change required}
 
@@
 %difficult time to distinguish between infrastructural and archetypical
-%applications. The definitions sound very similar.}  %\katznote{this section has
-been re-written.  I've removed archetypical applications, and added a paragraph
-at the end of this section (2.0) to mention this idea without naming it.  Please
-see if this is now better.}  %\alnote{reads very well!}
-
-The discussions in the workshops and our analysis of the applications led us to
-examine the use of data in different parts of the applications, where we found
-some common patterns across applications.  For example, applications that
-involve streaming data collect data from sensors, transform and filter the data,
-store the results, and then later process (analyze) the stored results.  In
-general, we think that all of the applications we have studied can be mapped to
-a set of stages or phases.  This is illustrated in
-\figurename~\ref{fig:figures_application-stages}.
-
-The question of where the application, as opposed to the entire system, %(where
-the system is defined as all of \figurename~\ref{fig:figures_terminology}),
-actually starts is somewhat difficult.  In this paper, we have decided not to
-treat hardware data sources (e.g., sensors, telescopes) as part of the
-applications.  Therefore, these applications generally start with the data
-movement that follows the actual data generation.
-
-Unlike the sensor applications, another set of applications generate data
-computationally, for example, from a model or simulation.  These applications
-then do start with data generation, and they often have just three stages, where
-data is generated in the first stage, stored in a second stage, and then
-processed in a third stage (i.e., the transformation \& decisions stage in
+%applications. The definitions sound very similar.}
+%\katznote{this section has been re-written.
+I've removed archetypical applications, and added a paragraph at the end of this section (2.0) to mention this idea without naming it.
+Please see if this is now better.}
+%\alnote{reads very well!}
+
+The discussions in the workshops and our analysis of the applications
+led us to examine the use of data in different parts of the
+applications, where we found some common patterns
+across applications.  For example, applications that involve
+streaming data collect data from sensors, transform and
+filter the data, store the results, and then later process (analyze)
+the stored results.  In general, we think that all of the applications
+we have studied can be mapped to a set of stages or phases.  This is
+illustrated in \figurename~\ref{fig:figures_application-stages}.
+
+The question of where the application, as opposed to the entire
+system, %(where the system is defined as all of \figurename~\ref{fig:figures_terminology}), 
+actually starts
+is somewhat difficult.  In this paper, we have decided not to treat
+hardware data sources (e.g., sensors, telescopes) as part of the
+applications.  Therefore, these applications generally start with the
+data movement that follows the actual data generation.
+
+Unlike the sensor applications, another set of applications generate
+data computationally, for example, from a model or simulation.  These
+applications then do start with data generation, and they often have
+just three stages, where data is generated in the first stage, stored
+in a second stage, and then processed in a third stage (i.e., the
+transformation \& decisions stage in
 \figurename~\ref{fig:figures_application-stages} does not occur).
 
-Other applications may operate on stored data (i.e., they only contain the
-processing stage and its associate data movement).  %\alnote{We should refer to
-these application stages from the % application descriptions. Currently, this
-classification is not % used.}  \katznote{have now done, but this was harder to
-do than I % thought it would be, particularly to define where the application %
-starts - for example, LSST as an app does not include data % generation, but
-starts after this (data transformation \& % decisions), and SOA Astronomy does
-not include data storage, but % does include the movement from the storage.
-Perhaps I need to go % back and make the fact that many apps start with data
-movement more % clear.}\alnote{ok}
-
-\begin{figure}[htbp] \centering
+Other
+applications may operate on stored data (i.e., they only contain the
+processing stage and its associate data movement).
+%\alnote{We should refer to these application stages from the
+%  application descriptions. Currently, this classification is not
+%  used.}  \katznote{have now done, but this was harder to do than I
+%  thought it would be, particularly to define where the application
+%  starts - for example, LSST as an app does not include data
+%  generation, but starts after this (data transformation \&
+%  decisions), and SOA Astronomy does not include data storage, but
+%  does include the movement from the storage.  Perhaps I need to go
+%  back and make the fact that many apps start with data movement more
+%  clear.}\alnote{ok}
+
+\begin{figure}[htbp]
+    \centering
         \includegraphics[width=1.0\textwidth]{figures/application-stages-new.pdf}
-        \caption{Application Stages: The work in data-intensive applications can
-be put into four common application stages: data generation, transformation \&
-decisions, storage, and processing, though not all applications have all stages.
-% \alnote{We have some applications (e.g. sensor networks) that give feedback to
-the % data generation stage. Do we want to refer to this possibility here?}  %
-\alnote{data transformation and decision making sounds to me similar to
-processing} % \katznote{what do you suggest?}\alnote{The proposal would be to
-just call it Data Processing as before. Since % there are many forms of
-processing (transformation being just a specific form of processing)}
-\katznote{I want to distinguish between the processing that happens before
-storage % and the processing that happens after storage, so I am thinking of not
-changing this.  % Would this really bother you?}
+        \caption{Application Stages: The work in data-intensive
+        applications can be put into four common application stages:
+		data generation, transformation \& decisions, storage, and processing, though not all
+		applications have all stages.
+%		\alnote{We have some applications (e.g. sensor networks) that give feedback to the
+%		data generation stage. Do we want to refer to this possibility here?}
+%		\alnote{data transformation and decision making sounds to me similar to processing}
+%		\katznote{what do you suggest?}\alnote{The proposal would be to just call it Data Processing as before. Since
+%		 there are many forms of processing (transformation being just a specific form of processing)} \katznote{I want to distinguish between the processing that happens before storage
+%		 and the processing that happens after storage, so I am thinking of not changing this.
+%		 Would this really bother you?}
                 \label{fig:figures_application-stages}}
@@
 
-%\katznote{Dan to clean up the description of app phases in the intro and the
-apps, %explain where apps start, with data movement, data generation, data
-storage, etc. -- %then ask Shantenu to read and comment on this.}
+%\katznote{Dan to clean up the description of app phases in the intro and the apps,
+%explain where apps start, with data movement, data generation, data storage, etc. --
+%then ask Shantenu to read and comment on this.}
 
@@
 %
-%Identify: (1) what is the case scenario within the application; (2) what data
-is dynamic; (3) how the data is being used in the context of each of these
-scenarios.
-%
-%\katznote{proposed structure for rewriting these entries:\\ %1. what does the
-application do - what problem does it solve (question 1 below) \\ %2. how does
-it do it (questions 2,3,5,6 below, also address cyberinfrastructure (where does
-the app run)) -- perhaps have a diagram here of what happens in the app \\
-%3. what are the big data aspects (question 4b below) \\ %4. what are the
-distributed aspects (question 3 below)\\ %5. what are the dynamic aspects
-(question 4a below)\\ %6. what else is important (questions 7,8 below, also
-address programming systems (how was the app written; what tools does it use))\\
+%Identify: (1) what is the case scenario within the application; (2) what data is dynamic; (3)  how the data is being used in the context of each of these scenarios.
+%
+%\katznote{proposed structure for rewriting these entries:\\
+%1. what does the application do - what problem does it solve (question 1 below) \\
+%2. how does it do it (questions 2,3,5,6 below, also address cyberinfrastructure (where does the app run)) -- perhaps have a diagram here of what happens in the app \\
+%3. what are the big data aspects (question 4b below) \\
+%4. what are the distributed aspects (question 3 below)\\
+%5. what are the dynamic aspects (question 4a below)\\
+%6. what else is important (questions 7,8 below, also address programming systems (how was the app written; what tools does it use))\\
 %}
@@
 %\item[2] How is the application used to do this?
-%\item[3] What infrastructure is used? (including compute, data, network,
-instruments, etc.)
+%\item[3] What infrastructure is used? (including compute, data, network, instruments, etc.)
 %\item[4] What dynamic data is used in the application?
@@
 %
-%Please feel free to also talk about the current state of the %application, if
-it exists today, and any specific gaps that you know %need to be overcome.
-
-%\subsection{Silvia's Biosciences app: NGS, medical imaging
-(Silvia)\label{bioSilvia}}
+%Please feel free to also talk about the current state of the
+%application, if it exists today, and any specific gaps that you know
+%need to be overcome.
+
+%\subsection{Silvia's Biosciences app: NGS, medical imaging (Silvia)\label{bioSilvia}}
 
@@
 
-The (grand) challenge in bioinformatics is the need to provide support for the
-analysis of genome sequence data that is becoming available due to the abundance
-and high-throughput capability of the next-generation sequencing devices. The
-rate of growth of data from the NGS machines humbles Moore's law; it has
-increased by more than three orders of magnitude in barely a decade.
-
-There are many different types and components of analyses of NGS data.  However,
-fundamentally important to all types is the ability to efficiently and
-effectively map/align the short reads to a reference genome.
-
-Alignment underpins multiple scientific questions such as comparing metagenomic
-(DNA sequence) data sets against each other, which could be used to distinguish
-human disease biomarkers among different individuals, determine different soil
-(or water or air or ...)  microbial composition from different samples, etc.
-
-This subsection describes a set of {\em traditional applications} that are used
-in the processing of next generation sequencing data.  The applications have two
-of the possible stages: data storage and data processing.
-%\katznote{Shantenu/Yogesh, can you confirm this or fix it?  % I'm sure data
-generation is not part of this app, but am not sure % about transformation \&
-decisions. - see % Figure~\ref{fig:figures_application-stages}.} \jhanote{Dan:
-I'm % confused. Not sure I understand what you mean by data-generation is % not
-part of this app?}  % \katznote{the sequencer is not part of the application -
-the application % starts once the data has been generated, thus it starts on a
-data movement % arrow.  I'm just not sure which one.}\alnote{I think the phases
-storage and % processing are fine for NGS}
+The (grand) challenge in bioinformatics is the need to provide support
+for the analysis of genome sequence data that is becoming available
+due to the abundance and high-throughput capability of the
+next-generation sequencing devices. The rate of growth of data from
+the NGS machines humbles Moore's law; it has increased by more than
+three orders of magnitude in barely a decade.
+
+There are many different types and components of analyses of NGS data.
+However, fundamentally important to all types is the ability to
+efficiently and effectively map/align the short reads to a reference
+genome.
+
+Alignment underpins multiple scientific questions such as comparing
+metagenomic (DNA sequence) data sets against each other, which could
+be used to distinguish human disease biomarkers among different
+individuals, determine different soil (or water or air or ...)
+microbial composition from different samples, etc.
+
+This subsection describes a set of {\em traditional applications} that
+are used in the processing of next generation sequencing data.  The
+applications have two of the possible stages: data storage and data
+processing.
+%\katznote{Shantenu/Yogesh, can you confirm this or fix it?
+%  I'm sure data generation is not part of this app, but am not sure
+%  about transformation \& decisions. - see
+%  Figure~\ref{fig:figures_application-stages}.} \jhanote{Dan: I'm
+%  confused. Not sure I understand what you mean by data-generation is
+%  not part of this app?}
+%  \katznote{the sequencer is not part of the application - the application
+%  starts once the data has been generated, thus it starts on a data movement
+%  arrow.  I'm just not sure which one.}\alnote{I think the phases storage and
+%  processing are fine for NGS}
 
@@
 There are multiple tools (programs) that can perform the mapping/alignment
-phase, e.g., BWA, BFAST, Bowtie, MAQ, etc. Each has its own strength and its own
-native data support/capability; they also implement different alignment
-algorithms. Thus, based upon different design objectives and trade-offs, they
-typically have different constraints and efficiencies.
-
-Several of these tools are also composed together as a linear pipeline for
-performing sequencing and analysis. The pipeline may go through alignment, SNP
-calling, haplotyping, and so on, with different tools used in each stage. These
-pipelines can often be performed independently on each chromosome to provide
-simple parallelism.
-
-All algorithms, architectures, storage, and networks have broken or will break
-soon.  Some NGS analytics are not good fits for core-rich and memory-poor modern
-architectures. Some problems (e.g., comparing k-mers across metagenomic data
-sets) need terabyte-scale memory, not thousands of cores.~\cite{ngs-gap}
-
-Individual data sets, which will be terabyte-scale, are generated by NGS
-machines at a variety of locations. There could be many thousands of such sets
-to compare. Having them all local may not be feasible (the transfer time from
-where they are generated may be prohibitive).
+phase, e.g., BWA, BFAST, Bowtie, MAQ, etc. Each has its own strength and
+its own native data support/capability; they also implement
+different alignment algorithms. Thus, based upon different design
+objectives and trade-offs, they typically have different constraints
+and efficiencies.
+
+Several of these tools are also composed together as a linear pipeline
+for performing sequencing and analysis. The pipeline may go through
+alignment, SNP calling, haplotyping, and so on, with different tools
+used in each stage. These pipelines can often be performed
+independently on each chromosome to provide simple parallelism.
+
+All algorithms, architectures, storage, and networks have broken or
+will break soon.  Some NGS analytics are not good fits for core-rich
+and memory-poor modern architectures. Some problems (e.g., comparing
+k-mers across metagenomic data sets) need terabyte-scale memory, not
+thousands of cores.~\cite{ngs-gap}
+
+Individual data sets, which will be terabyte-scale, are generated by
+NGS machines at a variety of locations. There could be many thousands
+of such sets to compare. Having them all local may not be feasible
+(the transfer time from where they are generated may be prohibitive).
 
@@
 
-Individual data sets will be terabyte-scale. For example, for a standard genomic
-sequence, there are billions of short (35-100 bases) reads of DNA.  And there
-could be many thousands of such sets to compare for a given problem.  Getting
-more accuracy for sequencing may require multiple coverage (e.g. 10x), that can
-cause sequence sizes to increase linearly.  This makes NGS data management
-fundamentally a big data problem.  NGS analytics, in addition to being a big
-data problem, is also a computationally demanding and distributed computing
-problem.  The computational demands arise from the often complex and intensive
-analysis that has to be performed on data, which in turn arise from algorithms
-that are designed to account for repetitions, errors and incomplete information.
+Individual data sets will be terabyte-scale. For example, for a
+standard genomic sequence, there are billions of short (35-100 bases)
+reads of DNA.  And there could be many thousands of such sets to
+compare for a given problem.  Getting more accuracy for sequencing may
+require multiple coverage (e.g. 10x), that can cause sequence sizes to
+increase linearly.  This makes NGS data management fundamentally a big
+data problem.  NGS analytics, in addition to being a big data problem,
+is also a computationally demanding and distributed computing problem.
+The computational demands arise from the often complex and intensive
+analysis that has to be performed on data, which in turn arise from
+algorithms that are designed to account for repetitions, errors and
+incomplete information.
 
@@
 
-The type of NGS research being conducted can distinguish the distributed
-properties of the application. Researcher groups that own sequencing machines
-need to deal with large dataset sizes that need to be resequenced and reduced
-for further analysis. Other researchers may analyze large catalogs of sequenced
-genome databases are maintained, and perform pattern matching. Further, clinical
-may be interested in in-depth analysis of an individual sequence for disease
+The type of NGS research being conducted can distinguish the
+distributed properties of the application. Researcher groups that
+own sequencing machines need to deal with large dataset sizes that
+need to be resequenced and reduced for further analysis. Other
+researchers may analyze large catalogs of sequenced genome databases
+are maintained, and perform pattern matching. Further, clinical may be
+interested in in-depth analysis of an individual sequence for disease
 diagnosis or visualization.
 
-The distributed computing aspects arise at multiple levels: for example, the
-simple act of having to move data from source (generation) to the destination
-where computing (analysis) will occur is a challenge due to the large volumes
-involved. Trade-offs exists between the cost/challenges in distributing data
-versus I/O saturation or memory bottlenecks.
-
-When coupled with the compute intensive nature of the problem, it soon emerges
-that a fundamental challenge is not only whether to distribute, but where to
-distribute, what to distribute (should the computing move to the data, or the
-data move to the compute), and how to distribute (what tools and infrastructure
-to use).
+The distributed computing aspects arise at multiple levels: for
+example, the simple act of having to move data from source
+(generation) to the destination where computing (analysis) will occur
+is a challenge due to the large volumes involved. Trade-offs exists
+between the cost/challenges in distributing data versus I/O
+saturation or memory bottlenecks.
+
+When coupled with the compute intensive nature of the problem, it soon
+emerges that a fundamental challenge is not only whether to
+distribute, but where to distribute, what to distribute (should the
+computing move to the data, or the data move to the compute), and how
+to distribute (what tools and infrastructure to use).
 
@@
 
-The dynamic challenges of NGS data are more subtle; the data themselves are not
-dynamic -- the sequences are acquired once and analyzed many times. However the
-execution of these applications on distributed infrastructure have dynamic
-aspects when optimized resource usage is considered. For example, workload
-decomposition and distribution must be determined dynamically in order to
-optimally use resources, e.g., compute resource selection can be based on data
+The dynamic challenges of NGS data are more subtle; the data
+themselves are not dynamic -- the sequences are acquired once and
+analyzed many times. However the execution of these applications on
+distributed infrastructure have dynamic aspects when optimized
+resource usage is considered. For example, workload decomposition and
+distribution must be determined dynamically in order to optimally use
+resources, e.g., compute resource selection can be based on data
 location, processing profile and/or network capacity.
@@
 
-There are several toolkits that have emerged for NGS providing incrementally
-better algorithms for sequencing and analysis \cite{SAMtools, BioJava, BioPerl,
-.NETBio}.  However, as a consequence of the broad range of infrastructural
-requirements, there isn't currently a common ``standard'' cyberinfrastructure
-used for NGS analytics; researchers use what is easily available to them, or try
-to use COTS infrastructure. Historically, leaders in the field have developed
-infrastructure for the larger community to build-upon, e.g., QCDOC or
-MD-GRAPE. But given the broad range of requirements, a hardware-only solution is
-unlikely for a broad range of problem instances, and a combination of hardware
-and software approaches will be employed. This reinforces the need for flexible
-programming systems and associated runtime environments that support collective
+There are several toolkits that have emerged for NGS providing
+incrementally better algorithms for sequencing and analysis
+\cite{SAMtools, BioJava, BioPerl, .NETBio}.  However, as a consequence
+of the broad range of infrastructural requirements, there isn't
+currently a common ``standard'' cyberinfrastructure used for NGS
+analytics; researchers use what is easily available to them, or try to
+use COTS infrastructure. Historically, leaders in the field have
+developed infrastructure for the larger community to build-upon, e.g.,
+QCDOC or MD-GRAPE. But given the broad range of requirements, a
+hardware-only solution is unlikely for a broad range of problem
+instances, and a combination of hardware and software approaches will
+be employed. This reinforces the need for flexible programming systems
+and associated runtime environments that support collective
 (hardware-software) optimization.
 
-Nowadays, NGS sequencers are even being shipped with servers or small clusters
-for performing the alignment and resequencing as a pipeline right after the
-reads have been generated~\cite{ilm}.  %\alnote{What kind of codes are used %in
-these clusters? What are the characteristics of these codes?}  %But any
-additional analysis is not supported.  %\alnote{What would be an additional
-analysis}
-%
-Scientific workflow systems like Taverna~\cite{Taverna:2006},
-BioKepler~\cite{biokepler:2011}, and Galaxy~\cite{galaxy} (though Galaxy does
-not support distributed systems currently) have been successful in integrating
-genome analysis tools for bioinformatics. Script based pipelines that combine
-executables are also used often. Very few workflow systems have explicit support
-for distributed data or dynamic data handling.  Therefore reusable approaches at
-extending existing tools to support distributed and dynamic data are likely to
-be useful.  %\katznote{Shantenu/Yogesh/Andre, please read this paragraph and
-tell me if changes %should be made to this section at this point}\alnote{ok with
-me}
-
-%\alnote{Some details on how distributed/dynamic data is handled in % these
-workflows would be good.}
+Nowadays, NGS sequencers are even being shipped with servers or small
+clusters for performing the alignment and resequencing as a pipeline
+right after the reads have been generated~\cite{ilm}.
+%\alnote{What kind of codes are used
+%in these clusters? What are the characteristics of these codes?}
+%But any additional analysis is not supported.
+%\alnote{What would be an additional analysis}
+%
+Scientific workflow systems like Taverna~\cite{Taverna:2006}, BioKepler~\cite{biokepler:2011}, and
+Galaxy~\cite{galaxy} (though Galaxy does not support distributed systems currently) have been successful in integrating genome
+analysis tools for bioinformatics. Script based pipelines that combine
+executables are also used often. Very few workflow systems have
+explicit support for distributed data or dynamic data handling.
+Therefore reusable approaches at extending existing tools to support
+distributed and dynamic data are likely to be useful.
+%\katznote{Shantenu/Yogesh/Andre, please read this paragraph and tell me if changes
+%should be made to this section at this point}\alnote{ok with me}
+
+%\alnote{Some details on how distributed/dynamic data is handled in
+%  these workflows would be good.}
 
@@
 
-% \katznote{I was assuming you would leave the questions here, so that % we
-could pull some of this material into a table later...}
+% \katznote{I was assuming you would leave the questions here, so that
+%   we could pull some of this material into a table later...}
 
@@
 
-% \katznote{from answers to questions 2,3,5,6 % below, also address
-cyberinfrastructure (where does the app run) -- % perhaps have a diagram here of
-what happens in the app}
-
-%\ysnote{The first para of this section reads more like ``what does the
-application do'' rather than % ``how''. Should it move above? Also, the
-description of cores/memory above can move here as it % relates to CI. }
-%\ysnote{This section will benefit from describing the (often) linear pipeline
-of % sequencing, with individual chromosomes that can be processed
-independently.}  %\katznote{Yogesh or Shantenu, can you do address Andre's
-comments in this section?}  %\ysnote{I've taken a pass. Shantenu, do you care to
-make changes?}  % 3. {\bf What are the big data aspects?}  \katznote{from answer
-to % question 4b below} 4. {\bf What are the distributed aspects?}  %
-\katznote{from answer to question 3 below} \ysnote{distinguish % between
-individual research project where compute can move to % data; clinical testlabs
-that sequence and transmit sequenced data % to end user for post processing;
-collaboratory projects like % 1000-genome where data has to be aggregated
-centrally. Sequencing % done remotely but visualization done locally.}  %5. {\bf
-What are the dynamic aspects?}  %\katznote{from answer to question 4a below}
+% \katznote{from answers to questions 2,3,5,6
+%   below, also address cyberinfrastructure (where does the app run) --
+%   perhaps have a diagram here of what happens in the app}
+
+%\ysnote{The first para of this section reads more like ``what does the application do'' rather than
+%  ``how''. Should it move above? Also, the description of cores/memory above can move here as it
+%  relates to CI. }
+%\ysnote{This section will benefit from describing the (often) linear pipeline of
+%  sequencing, with individual chromosomes that can be processed independently.}
+%\katznote{Yogesh or Shantenu, can you do address Andre's comments in this section?}
+%\ysnote{I've taken a pass. Shantenu, do you care to make changes?}
+% 3. {\bf What are the big data aspects?}  \katznote{from answer to
+%   question 4b below} 4. {\bf What are the distributed aspects?}
+% \katznote{from answer to question 3 below} \ysnote{distinguish
+%   between individual research project where compute can move to
+%   data; clinical testlabs that sequence and transmit sequenced data
+%   to end user for post processing; collaboratory projects like
+%   1000-genome where data has to be aggregated centrally. Sequencing
+%   done remotely but visualization done locally.}
+%5. {\bf What are the dynamic aspects?}
+%\katznote{from answer to question 4a below}
 %6. {\bf What else is important?}
 
-%\katznote{from answers to questions % 7,8 below, also address programming
-systems (how was the app % written; what tools does it use)}
-
-%\ysnote{See biohpc.net, bioperl, biojava, bio.net} %\katznote{I'm not sure what
-action goes with this comment.  Yogesh?}  % It is unlikely that researchers or
-NGS teams will build integrated % customised infrastructure for no
-cyberinfrastructure to solve this % problem. The solution to
+%\katznote{from answers to questions
+%  7,8 below, also address programming systems (how was the app
+%  written; what tools does it use)}
+
+%\ysnote{See biohpc.net, bioperl, biojava, bio.net}
+%\katznote{I'm not sure what action goes with this comment.  Yogesh?}
+% It is unlikely that researchers or NGS teams will build integrated
+% customised infrastructure for no cyberinfrastructure to solve this
+% problem. The solution to
 
@@
 
-% In general this is a large (for biology -- don't think in terms of %
-exascale!) data problem with large (all-vs-all comparison) % computing. There
-are strategies to make this more scalable that will % reduce the computing
-volume.
-
-% i) large (exascale) data and large (exascale) computing (e.g., data % mining,
-analytics, bioinformatics)
+% In general this is a large (for biology -- don't think in terms of
+% exascale!) data problem with large (all-vs-all comparison)
+% computing. There are strategies to make this more scalable that will
+% reduce the computing volume.
+
+% i) large (exascale) data and large (exascale) computing (e.g., data
+% mining, analytics, bioinformatics)
 
@@
 
-% iii) modest data and large (exascale) computing (e.g., % simulation/modeling)
-
-% Potentially peta-bytes or exa-bytes of data. Not sure how to quantify % the
-computing. (Huge memories will be potentially more valuable than % huge numbers
-of cores -- although this is heresy to the physics % crowd.)
-
-% Computing might then have to be % distributed. Biology is wrestling with this
-now.
-
-% Although only the alignment step is computation-intensive, it is % more
-practical to perform all the others on the grid to minimize data % transport
-to/from the user's workstation, which is normally on low % speed networks. That
-is, the computation is statically and manually % taken to the data in this
-case. This can lead to sub-optimal resource % usage.
-
-% Data is not really `dynamic' in this case -- the sequences are % acquired once
-and analyzed many times. However the execution of this % application on a
-distributed infrastructure has dynamic aspects when % optimized resource usage
-is considered:
+% iii) modest data and large (exascale) computing (e.g.,
+% simulation/modeling)
+
+% Potentially peta-bytes or exa-bytes of data. Not sure how to quantify
+% the computing. (Huge memories will be potentially more valuable than
+% huge numbers of cores -- although this is heresy to the physics
+% crowd.)
+
+% Computing might then have to be
+% distributed. Biology is wrestling with this now.
+
+% Although only the alignment step is computation-intensive, it is
+% more practical to perform all the others on the grid to minimize data
+% transport to/from the user's workstation, which is normally on low
+% speed networks. That is, the computation is statically and manually
+% taken to the data in this case. This can lead to sub-optimal resource
+% usage.
+
+% Data is not really `dynamic' in this case -- the sequences are
+% acquired once and analyzed many times. However the execution of this
+% application on a distributed infrastructure has dynamic aspects when
+% optimized resource usage is considered:
 
@@
 %\item Split/merge the data according to available resources
-%\item Compute resource selection based on data location, processing profile
-and/or network capacity
+%\item Compute resource selection based on data location, processing profile and/or network capacity
 %\item Resource selection based on security constraints (access control, trust)
-%\item Collection and processing of monitoring information to steer grid
-enactment.
+%\item Collection and processing of monitoring information to steer grid enactment.
 %\end{itemize}
@@
 %\begin{itemize}
-%\item Infrastructure monitoring for the execution of workflows to % manage
-failure
-%\item the application is sequence alignment (split data (this is where there is
-dynamic decision making) to execute the alignment in parallel)
-%\item DTI Imaging: two cases are interesting: DTI atlas (split, run, merge,
-split, merge - reduction from 10000 tasks to 1 output); PCA for classification
-of patients/control
+%\item Infrastructure monitoring for the execution of workflows to
+%  manage failure
+%\item the application is sequence alignment (split data (this is where there is dynamic decision making) to execute the alignment in parallel)
+%\item DTI Imaging: two cases are interesting: DTI atlas (split, run, merge, split, merge - reduction from 10000 tasks to 1 output); PCA for classification of patients/control
 %\end{itemize}
@@
 %
-%The application performs analysis of DNA sequencing data for various %goals in
-life science research, e.g., Mutation screening, Virus %discovery, Genome-wide
-Associations (GWA), Linkage analysis, %Comparison of bacterial genomes, microRNA
-expression and Exome %sequencing. The data analysis is composed of various
-methods that are %implemented by a bioinformatician, whereas the data is
-typically owned %by the life scientist.
+%The application performs analysis of DNA sequencing data for various
+%goals in life science research, e.g., Mutation screening, Virus
+%discovery, Genome-wide Associations (GWA), Linkage analysis,
+%Comparison of bacterial genomes, microRNA expression and Exome
+%sequencing. The data analysis is composed of various methods that are
+%implemented by a bioinformatician, whereas the data is typically owned
+%by the life scientist.
 %
@@
 %Note: Here we focus on data analysis from the perspective of the
-%bioinformatician, excluding steps related to data acquisition %(transformation
-of raw images into sequences of amino acids) and %posterior biostatistics and
-epidemiological analysis.
-%
-%The data analysis is implemented as a pipeline of generic and %customized
-components, mostly third-party public software tools. The %pipeline may differ
-for each specific study, depending on the data %acquisition and the research
-question. Typically it includes data %preparation (file conversion, reformatting
-and filtering); alignment %(comparison) with some reference database as the
-human genome; %post-processing (output conversion, reformatting, statistical
+%bioinformatician, excluding steps related to data acquisition
+%(transformation of raw images into sequences of amino acids) and
+%posterior biostatistics and epidemiological analysis.
+%
+%The data analysis is implemented as a pipeline of generic and
+%customized components, mostly third-party public software tools. The
+%pipeline may differ for each specific study, depending on the data
+%acquisition and the research question. Typically it includes data
+%preparation (file conversion, reformatting and filtering); alignment
+%(comparison) with some reference database as the human genome;
+%post-processing (output conversion, reformatting, statistical
 %analysis) and visualization. Sequence alignment is the most
-%computing-intensive step, which became prohibitive for sequential %execution in
-next generation sequence experiments. All data is stored %in files that are
-often completely read in memory for fast %processing. Compressed formats are
-usually adopted. Although the files %are large, the data can be easily
-decomposed because the processing is %done on each sequence (`string') at a
-time.
+%computing-intensive step, which became prohibitive for sequential
+%execution in next generation sequence experiments. All data is stored
+%in files that are often completely read in memory for fast
+%processing. Compressed formats are usually adopted. Although the files
+%are large, the data can be easily decomposed because the processing is
+%done on each sequence (`string') at a time.
 %
@@
 %
-%Grid storage: Storage Elements of the Dutch e-Science Grid %(www.biggrid.nl)
+%Grid storage: Storage Elements of the Dutch e-Science Grid
+%(www.biggrid.nl)
 %
@@
 %
-%Between data server and grid storage: public network (XXX); lightpath %(near
-future)
-%
-%Between grid storage and worker nodes: public (XXX) or site (XXX) %network
+%Between data server and grid storage: public network (XXX); lightpath
+%(near future)
+%
+%Between grid storage and worker nodes: public (XXX) or site (XXX)
+%network
 %
@@
 %
-%Local Computing: clusters and servers at the AMC and workstations %(laptops,
-desktops) anywhere Grid Computing: Dutch Grid %(www.biggrid.nl), part of EGI
-%
-%Originally all steps were performed in some local server or on the %user's
-workstation. The visualization is still performed in the user's %(windows)
-workstation, but all the other steps are currently performed %on the Dutch grid
-infrastructure. The pipeline is described as a grid %workflow using the GWENDIA
-language %(\url{http://dx.doi.org/10.1145/1645164.1645171}) and enacted on the
+%Local Computing: clusters and servers at the AMC and workstations
+%(laptops, desktops) anywhere Grid Computing: Dutch Grid
+%(www.biggrid.nl), part of EGI
+%
+%Originally all steps were performed in some local server or on the
+%user's workstation. The visualization is still performed in the user's
+%(windows) workstation, but all the other steps are currently performed
+%on the Dutch grid infrastructure. The pipeline is described as a grid
+%workflow using the GWENDIA language
+%(\url{http://dx.doi.org/10.1145/1645164.1645171}) and enacted on the
 %Dutch Grid using the MOTEUR2 workflow engine
@@
 %
-%PS: although only the alignment step is computation-intensive, it is %more
-practical to perform all the others on the grid to minimize data %transport
-to/from the user's workstation, which is normally on low %speed networks. That
-is, the computation is statically and manually %taken to the data in this
-case. This can lead to sub-optimal resource %usage.
+%PS: although only the alignment step is computation-intensive, it is
+%more practical to perform all the others on the grid to minimize data
+%transport to/from the user's workstation, which is normally on low
+%speed networks. That is, the computation is statically and manually
+%taken to the data in this case. This can lead to sub-optimal resource
+%usage.
 %
@@
 %
-%Note: data is not really `dynamic' in this case -- the sequences are %acquired
-once and analyzed many times. However the execution of this %application on a
-distributed infrastructure has dynamic aspects when %optimized resource usage is
-considered:
+%Note: data is not really `dynamic' in this case -- the sequences are
+%acquired once and analyzed many times. However the execution of this
+%application on a distributed infrastructure has dynamic aspects when
+%optimized resource usage is considered:
 %
@@
 %
-%DNA sequences (strings) stored in open formats (FASTA, SFF, SAM, BAM) %Bundles
-of results from alignment tools (custom format, binary?)
+%DNA sequences (strings) stored in open formats (FASTA, SFF, SAM, BAM)
+%Bundles of results from alignment tools (custom format, binary?)
 %
@@
 %
-%input data: Sequencing Data in the csFasta format, normally 25-35~GB; %Quality
-files in the .qual format, normally 50-80~GB; Reference DB in %the Fasta BS
-format: 3.2~GB (human genome) 140~MB (one chromosome)
-%
-%Intermediate files: Reference BWA index, normally 4.5GB (human genome) %240~MB
-(one chromosome); Sequencing Data in the FastQ format %(fastq.gz), normally
-20-30~GB
-%
-%Output: Results in .sai format (direct output of BWA), normally %2-3~GB;
-Results in .sam format, normally 55-75~GB; Results in .bam %format, normally
-20-30~GB
+%input data: Sequencing Data in the csFasta format, normally 25-35~GB;
+%Quality files in the .qual format, normally 50-80~GB; Reference DB in
+%the Fasta BS format: 3.2~GB (human genome) 140~MB (one chromosome)
+%
+%Intermediate files: Reference BWA index, normally 4.5GB (human genome)
+%240~MB (one chromosome); Sequencing Data in the FastQ format
+%(fastq.gz), normally 20-30~GB
+%
+%Output: Results in .sai format (direct output of BWA), normally
+%2-3~GB; Results in .sam format, normally 55-75~GB; Results in .bam
+%format, normally 20-30~GB
 %
@@
 %
-%From a local file. The legacy code is wrapped to stage the inputs and %outputs
-in/out the worker node.
-%
-%All data is stored in files that are often completely read in memory %for fast
-processing. Compressed formats are usually adopted. Although %the files are
-large, the data can be easily decomposed because the %processing is done on each
-sequence (`string') at a time
+%From a local file. The legacy code is wrapped to stage the inputs and
+%outputs in/out the worker node.
+%
+%All data is stored in files that are often completely read in memory
+%for fast processing. Compressed formats are usually adopted. Although
+%the files are large, the data can be easily decomposed because the
+%processing is done on each sequence (`string') at a time
 %
@@
 %
-%In research there are no strict constraints on the time to complete %the
-analysis.  In patient care, time could determine life, death or %money, but we
-have no concrete case yet.
-%
-%Quality in this case refers to the accuracy of the results %obtained. This is
-essential for research and patient care.
+%In research there are no strict constraints on the time to complete
+%the analysis.  In patient care, time could determine life, death or
+%money, but we have no concrete case yet.
+%
+%Quality in this case refers to the accuracy of the results
+%obtained. This is essential for research and patient care.
 
@@
 
-The Large Hadron Collider (LHC)~\cite{lhc} at CERN in Geneva is producing a
-large amount of data across a number of experiments, including
-ATLAS~\cite{atlas}. The application discussed in this section is an {\em
-infrastructural application} (meaning that it consists of multiple stages that
-are run by different people, in this case data acquisition, storage,
-distribution, and analysis). It has the goal of allowing a physics group or a
-user to analyze data to understand a specific physics channel (underlying
-physical process) as recorded by ATLAS.  The set of available data grows
-steadily over time as more data are collected (by the experiment) and
-reconstructed. This application thus contains two stages: data storage and data
-processing.  %\alnote{prev. also acquisition was %part of the application}
-\katznote{I don't see that it should be.  Doesn't the app %start after the data
-is in the Tier 1 archive?}
-
-\subsubsection*{Application description} %2. {\bf How does it do it?}
-
-%\katznote{from answers to questions 2,3,5,6 below, also address
-cyberinfrastructure (where does the app run) -- perhaps have a diagram here of
-what happens in the app}
-
-Both physics groups and individuals run jobs on the WLCG (Worldwide LHC
-Computing Grid)~\cite{lhcb} as they see fit, either reading the reconstructed
-data or reading intermediate datasets created by themselves or their
-colleagues. If, in the future, resources should prove inadequate then
-adjustments to working practices to coordinate processing may be made to
-increase physics output.
-
-The WLCG has a hierarchical architecture. CERN (the `Tier~0' site) is connected
-to eleven `Tier~1' (national) sites by a dedicated optical network. This may be
-extended in the future.  All data is stored at CERN, and as needed, various
-parts of the dataset are copied (and cached, i.e., replicated) to the Tier~1
-sites, and from there to `Tier 2' (regional) sites and to `Tier~3' (university
-or group or individual) sites.
-
-Before data is analyzed, the file to be analyzed is copied from the closest site
-to the SE (Storage Element) at the WLCG site where the job will run. From there
-it is either copied to a local disk or read from the SE.
-
-There is no particular time constraint on the processing, although physics
-groups are eager to understand the ATLAS experiments and thus want to process
-data as quickly as possible.  In the event some data are missed during
-processing, the only effect is a reduction of statistical precision in the final
-analysis. Accurate bookkeeping is therefore important in order to know which
-data have been processed.
-
- \subsubsection*{Big data aspects} %3. {\bf What are the big data aspects?}
+The Large Hadron Collider (LHC)~\cite{lhc} at CERN in Geneva is
+producing a large amount of data across a number of experiments,
+including ATLAS~\cite{atlas}. The application discussed in this
+section is an {\em infrastructural application} (meaning that it
+consists of multiple stages that are run by different people, in this
+case data acquisition, storage, distribution, and analysis). It has
+the goal of allowing a physics group or a user to analyze data to
+understand a specific physics channel (underlying physical process) as
+recorded by ATLAS.  The set of available data grows steadily over time
+as more data are collected (by the experiment) and reconstructed. This
+application thus contains two stages: data storage and data
+processing.
+%\alnote{prev. also acquisition was
+%part of the application} \katznote{I don't see that it should be.  Doesn't the app
+%start after the data is in the Tier 1 archive?}
+
+\subsubsection*{Application description}
+%2. {\bf How does it do it?}
+
+%\katznote{from answers to questions 2,3,5,6 below, also address cyberinfrastructure (where does the app run) -- perhaps have a diagram here of what happens in the app}
+
+Both physics groups and individuals run jobs on the WLCG (Worldwide LHC Computing Grid)~\cite{lhcb} as they see fit, either reading the reconstructed data or reading intermediate datasets created by themselves or their colleagues.
+If, in the future, resources should prove inadequate then adjustments to working practices to coordinate processing may be made to increase physics output.
+
+The WLCG has a hierarchical architecture.
+CERN (the `Tier~0' site) is connected to eleven `Tier~1' (national) sites by a dedicated optical network.
+This may be extended in the future.
+All data is stored at CERN, and as needed, various parts of the dataset are copied (and cached, i.e., replicated) to the Tier~1 sites, and from there to `Tier 2' (regional) sites and to `Tier~3' (university or group or individual) sites.
+
+Before data is analyzed, the file to be analyzed is copied from the closest site to the SE (Storage Element) at the WLCG site where the job will run.
+From there it is either copied to a local disk or read from the SE.
+
+There is no particular time constraint on the processing, although physics groups are eager to understand the ATLAS experiments and thus want to process data as quickly as possible.
+In the event some data are missed during processing, the only effect is a reduction of statistical precision in the final analysis.
+Accurate bookkeeping is therefore important in order to know which data have been processed.
+
+ \subsubsection*{Big data aspects}
+%3. {\bf What are the big data aspects?}
 
@@
 
-Roughly 20~TB of data are collected every day. Multiple generations of processed
-data are kept. Many data files are replicated on multiple sites; this is done
-both by copying it to where it is likely to be useful and partly dynamically as
-the need arises.  The bulk of the data is individual event data at different
-stages of refinement. This is held in an in-house representation of C++
-serialized objects. Though this format is not standard it is used by all LHC
-experiments.
-
- \subsubsection*{Distributed aspects} %4. {\bf What are the distributed
-aspects?}
+Roughly 20~TB of data are collected every day.
+Multiple generations of processed data are kept.
+Many data files are replicated on multiple sites; this is done both by copying it to where it is likely to be useful and partly dynamically as the need arises.
+The bulk of the data is individual event data at different stages of refinement.
+This is held in an in-house representation of C++ serialized objects.
+Though this format is not standard it is used by all LHC experiments.
+
+ \subsubsection*{Distributed aspects}
+ %4. {\bf What are the distributed aspects?}
 
@@
 
-The WLCG has 250,000 cores distributed over 140 sites and with 100~PB of
-disk. It uses a mixture of gLite software, experiment specific software and
-physics group specific software.  The data is distributed and replicated as
-described previously.
-
- \subsubsection*{Dynamic aspects} %5. {\bf What are the dynamic aspects?}
+The WLCG has 250,000 cores distributed over 140 sites and with 100~PB of disk.
+It uses a mixture of gLite software, experiment specific software and physics group specific software.
+The data is distributed and replicated as described previously.
+
+ \subsubsection*{Dynamic aspects}
+%5. {\bf What are the dynamic aspects?}
 
@@
 
-The data being processed can be considered as dynamic as it grows steadily as
-new raw data are collected and the basic processing program (called
-reconstruction) is run to produce information about specific particles involved
-in a collision. This reconstruction is rerun periodically (two or three times
-per year is expected) as calibrations of the detector are
-improved. Reconstruction algorithms are also improved as are the physics groups'
-selection codes.  %\alnote{Is the reconstruction code the fundamental entity
-that is required to deal with new data? Sounds rather like an auxiliary tool.}
-%\katznote{reconstruction is a second type of processing.}\alnote{should we
-explicitly state that?}
-
- \subsubsection*{Other important issues} %6. {\bf What else is important?}
-
-%\katznote{from answers to questions 7,8 below, also address programming systems
-(how was the app written; what tools does it use)}
-
-Data analysis is really a pleasingly parallel task, where various events are
-analyzed in parallel and then a summary is created of all the analyses.
-
-The key challenge here is really in the infrastructure -- building a system that
-can store massive amounts of data and move them to where the processing is to be
-done by a large diverse group of users.
+The data being processed can be considered as dynamic as it grows steadily as new raw data are collected and the basic processing program (called reconstruction) is run to produce information about specific particles involved in a collision.
+This reconstruction is rerun periodically (two or three times per year is expected) as calibrations of the detector are improved.
+Reconstruction algorithms are also improved as are the physics groups' selection codes.
+%\alnote{Is the reconstruction code the fundamental entity that is required to deal with new data? Sounds rather like an auxiliary tool.}
+%\katznote{reconstruction is a second type of processing.}\alnote{should we explicitly state that?}
+
+ \subsubsection*{Other important issues}
+%6. {\bf What else is important?}
+
+%\katznote{from answers to questions 7,8 below, also address programming systems (how was the app written; what tools does it use)}
+
+Data analysis is really a pleasingly parallel task, where various events are analyzed in parallel and then a summary is created of all the analyses.
+
+The key challenge here is really in the infrastructure -- building a system that can store massive amounts of data and move them to where the processing is to be done by a large diverse group of users.
 
@@
 %\begin{itemize}
-%\item Most of the dynamism is in infrastructure, not in the application
-(e.g. use of PilotJob)
-%\item Application dynamism: location of file (e.g. if file turns out to be not
-local, look up in registry-- but not do this again)
+%\item Most of the dynamism is in infrastructure, not in the application (e.g. use of PilotJob)
+%\item Application dynamism: location of file (e.g. if file turns out to be not local, look up in registry-- but not do this again)
 %\item No specific QoS issue concerned with data transfer times
@@
 %
-%A physics group is interested in a specific physics channel (underlying
-physical process) as recorded by the Atlas experiment at the LHC (Large Hadron
-Collider) at CERN in Geneva. The set of available data grows steadily over time
-as more data are collected and reconstructed.
+%A physics group is interested in a specific physics channel (underlying physical process) as recorded by the Atlas experiment at the LHC (Large Hadron Collider) at CERN in Geneva.
+The set of available data grows steadily over time as more data are collected and reconstructed.
 %
@@
 %
-% Both physics groups and individuals run jobs as they see fit, either reading
-the reconstructed data or reading intermediate datasets created by themselves or
-their colleagues. If, in the future, resources should prove inadequate then
-adjustments to working practices to coordinate processing may be made to
-increase physics output.
-%
-% 3.  What infrastructure is used? (including compute, data, network,
-instruments, etc.)
-%
-%The WLCG (Worldwide LHC (Large Hadron Collider) Computing Grid) is used with a
-mixture of gLite software, experiment specific software and physics group
-specific software. WLCG has 250,000 cores distributed over 140 sites and with
-100~PB of disk. CERN is connected to the eleven `Tier 1' national sites by a
-dedicated optical network. This may be extended in the future.
+% Both physics groups and individuals run jobs as they see fit, either reading the reconstructed data or reading intermediate datasets created by themselves or their colleagues.
+If, in the future, resources should prove inadequate then adjustments to working practices to coordinate processing may be made to increase physics output.
+%
+%   3.      What infrastructure is used? (including compute, data, network, instruments, etc.)
+%
+%The WLCG (Worldwide LHC (Large Hadron Collider) Computing Grid) is used with a mixture of gLite software, experiment specific software and physics group specific software.
+WLCG has 250,000 cores distributed over 140 sites and with 100~PB of disk.
+CERN is connected to the eleven `Tier 1' national sites by a dedicated optical network.
+This may be extended in the future.
 %
@@
 %
-%The data being processed can be considered as dynamic as it grows steadily as
-new raw data are collected and the basic reconstruction program is run to
-produce information about specific particles involved in a collision. This
-reconstruction is rerun periodically (two or three times per year is expected)
-as calibrations of the detector are improved. Reconstruction algorithms are also
-improved as are the physics groups' selection codes. One output of the
-reconstruction is `tag' data - summary information written to a database to
-facilitate event selection. This is not yet used very much. Calibration data are
-held on an RDBMS with Frontier/Squid front ends.
+%The data being processed can be considered as dynamic as it grows steadily as new raw data are collected and the basic reconstruction program is run to produce information about specific particles involved in a collision.
+This reconstruction is rerun periodically (two or three times per year is expected) as calibrations of the detector are improved.
+Reconstruction algorithms are also improved as are the physics groups' selection codes.
+One output of the reconstruction is `tag' data - summary information written to a database to facilitate event selection.
+This is not yet used very much.
+Calibration data are held on an RDBMS with Frontier/Squid front ends.
 %
@@
 %
-% The bulk of the data is individual event data at different stages of
-refinement. This is held in a in-house representation of C++ serialized
-objects. Though this format is not standard it is used by all LHC experiments.
+%          The bulk of the data is individual event data at different stages of refinement.
+This is held in a in-house representation of C++ serialized objects.
+Though this format is not standard it is used by all LHC experiments.
 %
@@
 %
-% Roughly 20~TB of data are collected every day. Multiple generations of
-processed data are kept. Many data files are replicated on multiple sites; this
-is done both by copying it to where it is likely to be useful and partly
-dynamically as the need arises.
+%          Roughly 20~TB of data are collected every day.
+Multiple generations of processed data are kept.
+Many data files are replicated on multiple sites; this is done both by copying it to where it is likely to be useful and partly dynamically as the need arises.
 %
@@
 %
-%A copy of the file is made to the SE (Storage Element) at the site where the
-job will run. From there it is either copied to a local disk or read from the
-SE.
+%A copy of the file is made to the SE (Storage Element) at the site where the job will run. From there it is either copied to a local disk or read from the SE.
 %
@@
 %
-%If some data are missed during processing the only effect is a reduction of
-statistical precision in the final analysis. Accurate book keeping is important
-to know which data have been processed. If the selection code has been improved
-then naturally the physics groups would like to see the results as soon as
-possible.
+%If some data are missed during processing the only effect is a reduction of statistical precision in the final analysis.
+Accurate book keeping is important to know which data have been processed.
+If the selection code has been improved then naturally the physics groups would like to see the results as soon as possible.
 %
@@
 %
-%A physics group is interested in a specific physics channel (underlying
-physical process) as recorded by the Atlas experiment at the LHC (Large Hadron
-Collider) at CERN in Geneva. The set of available data grows steadily over time
-as more data are collected and reconstructed. It would not be efficient to allow
-individual physics groups to process this large volume of data in an
-uncoordinated manner as it requires reading all available data.
+%A physics group is interested in a specific physics channel (underlying physical process) as recorded by the Atlas experiment at the LHC (Large Hadron Collider) at CERN in Geneva.
+The set of available data grows steadily over time as more data are collected and reconstructed.
+It would not be efficient to allow individual physics groups to process this large volume of data in an uncoordinated manner as it requires reading all available data.
 %
@@
 %
-%Physics groups register their analysis code to be run on the next scheduled
-analysis ``train''. The train leaves every two weeks and reads through all the
-data, runs the physics selection code of each group and writes out the selected
-events and output from the computation to produce relatively small data sets for
-each physics group. The groups then use these selections to perform more ad-hoc
-analysis.
-%
-% 3.  What infrastructure is used? (including compute, data, network,
-instruments, etc.)
-%
-%The WLCG (Worldwide LHC (Large Hadron Collider) Computing Grid) is used with a
-mixture of gLite software, experiment specific software and physics group
-specific software. WLCG has 250,000 cores distributed over 140 sites and with
-100~PB of disk. There is no wide area dedicated network infrastructure.
+%Physics groups register their analysis code to be run on the next scheduled analysis ``train''.
+The train leaves every two weeks and reads through all the data, runs the physics selection code of each group and writes out the selected events and output from the computation to produce relatively small data sets for each physics group.
+The groups then use these selections to perform more ad-hoc analysis.
+%
+%   3.      What infrastructure is used? (including compute, data, network, instruments, etc.)
+%
+%The WLCG (Worldwide LHC (Large Hadron Collider) Computing Grid) is used with a mixture of gLite software, experiment specific software and physics group specific software.
+WLCG has 250,000 cores distributed over 140 sites and with 100~PB of disk.
+There is no wide area dedicated network infrastructure.
 %
@@
 %
-%The data being processed can be considered as dynamic as it grows steadily as
-new raw data are collected and the basic reconstruction program is run to
-produce information about specific particles involved in a collision. This
-reconstruction is rerun periodically (two or three times per year is expected)
-as calibrations of the detector are improved. Reconstruction algorithms are also
-improved as are the physics groups' selection code. Calibration data are held on
-an RDBMS with Frontier/Squid front ends.
+%The data being processed can be considered as dynamic as it grows steadily as new raw data are collected and the basic reconstruction program is run to produce information about specific particles involved in a collision.
+This reconstruction is rerun periodically (two or three times per year is expected) as calibrations of the detector are improved.
+Reconstruction algorithms are also improved as are the physics groups' selection code.
+Calibration data are held on an RDBMS with Frontier/Squid front ends.
 %
@@
 %
-% The bulk of the data is individual event data at different stages of
-refinement. The is held in a non-standard representation of C++ serialized
-objects.
+%          The bulk of the data is individual event data at different stages of refinement. The is held in a non-standard representation of C++ serialized objects.
 %
@@
 %
-% Currently the reconstructed data (excluding copies) is about m TB and is
-growing at about n TB per month. Many data files are replicated on multiple
-sites. This is done both by copying it to where it is likely to be useful and
-partly dynamically as the need arises.
+%          Currently the reconstructed data (excluding copies) is about m TB and is growing at about  n TB per month.
+Many data files are replicated on multiple sites.
+This is done both by copying it to where it is likely to be useful and partly dynamically as the need arises.
 %
@@
 %
-%This is mostly by copying the file to where the job will run - though some
-remote file access is also used.
+%This is mostly by copying the file to where the job will run - though some remote file access is also used.
 %
@@
 %
-%If some data are missed during processing the only effect is a reduction of
-statistical precision in the final analysis. If the selection code has been
-improved then naturally the physics groups would like to see the results as soon
-as possible.
+%If some data are missed during processing the only effect is a reduction of statistical precision in the final analysis.
+If the selection code has been improved then naturally the physics groups would like to see the results as soon as possible.
 
@@
 %\begin{itemize}
-%\item Analysis of data to generate event streams $\rightarrow$ lead to the %
-positioning of telescopes
+%\item Analysis of data to generate event streams $\rightarrow$ lead to the
+%  positioning of telescopes
 %\end{itemize}
@@
 %I'll assume that the ``application'' in question is the sum of the two
-%activities, despite the fact that they are different in many respects, %and
-I'll try and distinguish between them where appropriate.
-%
-%Let me know if you want clarification or further information - in some cases I
-wasn't quite sure what you were after.
+%activities, despite the fact that they are different in many respects,
+%and I'll try and distinguish between them where appropriate.
+%
+%Let me know if you want clarification or further information - in some cases I wasn't quite sure what you were after.
 %
@@
 %characterizing all the celestial sources detected in the sky survey dataset.
-%These are now typically implemented in relational databases, with SQL
-interfaces %implemented through web pages, and the principal goal of the Virtual
-Observatory %(VO) is to standardize access to these, and other astronomical data
-resources, %so that astronomers can readily perform multi-wavelength analyses,
-combining all %the extant data on particular sources.
+%These are now typically implemented in relational databases, with SQL interfaces
+%implemented through web pages, and the principal goal of the Virtual Observatory
+%(VO) is to standardize access to these, and other astronomical data resources,
+%so that astronomers can readily perform multi-wavelength analyses, combining all
+%the extant data on particular sources.
 %
@@
 %
-% 3. What infrastructure is used? (including compute, data, network, %
-instruments, etc.)
+%      3. What infrastructure is used? (including compute, data, network,
+%      instruments, etc.)
 %
@@
 %distributed internationally. Major datasets---e.g. those from large sky
-%surveys---are typically implemented in relational databases running on
-high-spec servers %or clusters thereof. The compute requirements of the VO are
-currently modest, %but are likely to increase over time. Astronomers typically
-use normal academic %networks, but more specialized networking is required in
-some cases: e.g., sky %survey data is moved from Cambridge to Edinburgh using
-the UKLight project's %dedicated optical network, and radio interferometers
-require high-bandwidth %fibre connections from each antenna to the correlator.
-%
-% 4. What dynamic data is used in the application?  %(a) What are the types of
-data, (b) What is the size of the data set(s)?
+%surveys---are typically implemented in relational databases running on high-spec servers
+%or clusters thereof. The compute requirements of the VO are currently modest,
+%but are likely to increase over time. Astronomers typically use normal academic
+%networks, but more specialized networking is required in some cases: e.g., sky
+%survey data is moved from Cambridge to Edinburgh using the UKLight project's
+%dedicated optical network, and radio interferometers require high-bandwidth
+%fibre connections from each antenna to the correlator.
+%
+%      4. What dynamic data is used in the application?
+%(a) What are the types of data, (b) What is the size of the data set(s)?
 %
@@
 %messages, which are notifications of transient events which can be used to
-%trigger follow-up observations on other telescopes. These are small XML
-messages %containing basic information about the position of a transient source
-and a %small amount of other metadata which the operators of a given telescope
-can use %to define triggers for follow-up observations of interest to them or
-their %community.
+%trigger follow-up observations on other telescopes. These are small XML messages
+%containing basic information about the position of a transient source and a
+%small amount of other metadata which the operators of a given telescope can use
+%to define triggers for follow-up observations of interest to them or their
+%community.
 %
@@
 %
-%The time constraints differ for different types of transient source, but in
-some %cases notification must be made within a few minutes if it is to be
-useful; the %overheads associated with setting up a follow-up observation mean
-that much %shorter time scales are practically irrelevant, although they are
-still %scientifically interesting. The quality constraints centre on providing a
+%The time constraints differ for different types of transient source, but in some
+%cases notification must be made within a few minutes if it is to be useful; the
+%overheads associated with setting up a follow-up observation mean that much
+%shorter time scales are practically irrelevant, although they are still
+%scientifically interesting. The quality constraints centre on providing a
 %telescope with good enough information about a potential transient source that
@@
 %observing program to follow-up the event. In most cases, the candidate
-%transient will be some sort of noise event, so fairly sophisticated filtering
-is %required to bring the false positive rate down to something acceptable,
-given %the cost of observing time on cutting edge telescopes.
+%transient will be some sort of noise event, so fairly sophisticated filtering is
+%required to bring the false positive rate down to something acceptable, given
+%the cost of observing time on cutting edge telescopes.
 
@@
 
-This application, which similarly to the previous application is an {\em
-infrastructural application}, wants to find and study variable objects and
-moving objects, using a telescope that takes images each night. This is a
-scenario that may be used by Large Synoptic Survey Telescope (LSST)~\cite{lsst},
-and is somewhat similar to that done by the Palomar Transient Factory
-(PTF)~\cite{ptf,ptf2,ptf3} and by Pan-STARRS~\cite{ps1,ps2}.  Overall, this
-application is a leading-edge example of the work needed to build a set of data
-(catalog and images) that then can be used by others for analysis (an example of
-which is the next application, in \S\ref{astroSOAAdam}.)  This application
-contains three stages: data transformation \& decisions, data storage.
-%\alnote{Is the telescope not %part of the application?} \katznote{I don't know
-- what do you think?  This %is a general issue I'm struggling with in this and
-other applications.} \alnote{I struggle as well - it is just confusing that the
-term telescope is part of the name} (Data generation comes from the telescope
-itself, and processing is done outside of this application, e.g. using the next
+This application, which similarly to the previous application is an
+{\em infrastructural application}, wants to find and study variable
+objects and moving objects, using a telescope that takes images each
+night. This is a scenario that may be used by Large Synoptic Survey
+Telescope (LSST)~\cite{lsst}, and is somewhat similar to that done by
+the Palomar Transient Factory (PTF)~\cite{ptf,ptf2,ptf3} and by
+Pan-STARRS~\cite{ps1,ps2}.  Overall, this application is a
+leading-edge example of the work needed to build a set of data
+(catalog and images) that then can be used by others for analysis (an
+example of which is the next application, in \S\ref{astroSOAAdam}.)
+This application contains three stages: data transformation \&
+decisions, data storage.
+%\alnote{Is the telescope not
+%part of the application?} \katznote{I don't know - what do you think?  This
+%is a general issue I'm struggling with in this and other applications.} \alnote{I struggle as well - it is just confusing that the term telescope is part of the name}
+(Data generation comes from the telescope itself,
+and processing is done outside of this application, e.g. using the next
 application.)
 
-\subsubsection*{Application description} %2. {\bf How does it do it?}
-
-%\katznote{from answers to questions 2,3,5,6 below, also address
-cyberinfrastructure (where does the app run) -- perhaps have a diagram here of
-what happens in the app} %\ysnote{This section feels like it has too much
-details/steps. Maybe we just need to stress the dynamic and distributed parts
-e.g. \#1-6 can be collapsed into a couple of bullets, and \#7 into another.}
-
-Processing of data (data analysis and detection of new sources) from each image
-should be done while the next image is being taken by the telescope.  This means
-that if anything unusual is detected, normal observation can be
-interrupted. Other observing resources can also be notified instantly, so they
-can observe the same event. As data is collected, it is added to all the data
-previously detected from the same location of sky to create a very deep master
-image. LSST will also build up a database of all known moving objects.
-
-Every time a new image of the sky is obtained, the master image will be
-subtracted from it. The result is a subtracted image that contains the
-difference between the current sky image and its average state.
-
-This subtracted image is then processed by a cluster of computers with 3 main
-steps:
+\subsubsection*{Application description}
+%2. {\bf How does it do it?}
+
+%\katznote{from answers to questions 2,3,5,6 below, also address cyberinfrastructure (where does the app run) -- perhaps have a diagram here of what happens in the app}
+%\ysnote{This section feels like it has too much details/steps.
+Maybe we just need to stress the dynamic and distributed parts e.g.
+\#1-6 can be collapsed into a couple of bullets, and \#7 into another.}
+
+Processing of data (data analysis and detection of new sources) from each image should be done while the next image is being taken by the telescope.
+This means that if anything unusual is detected, normal observation can be interrupted.
+Other observing resources can also be notified instantly, so they can observe the same event.
+As data is collected, it is added to all the data previously detected from the same location of sky to create a very deep master image.
+LSST will also build up a database of all known moving objects.
+
+Every time a new image of the sky is obtained, the master image will be subtracted from it.
+The result is a subtracted image that contains the difference between the current sky image and its average state.
+
+This subtracted image is then processed by a cluster of computers with 3 main steps:
 \begin{enumerate}[(i)]
 
-\item Using the existing object catalog (containing the known orbits of all
-known objects), find objects which are expected to appear in subtracted image,
-given the area of sky and time of day.  Cross-match expected objects with
-sources in subtracted image, resulting in the unmatched-source catalog, which
-contains sources that cannot be matched with a previously known object.  If
-these sources can be tracked over a few images, an orbit can be determined for
-them and they can be added to the object catalog.  Also, the orbit catalog can
-be updated with redetections of known objects; each rediscovery provides
-information that improves the known orbits.
-
-\item Attempt to classify all entries in the orbit catalog (i.e., all the
-objects which now have known orbits). If any Near Earth Objects are detected to
-be passing close to the earth, alerts are generated to astronomers so that
-follow up observations can be scheduled.
-
-\item If an unknown object is found that cannot be classified locally, at least
-two possible options exist. One is a coordinated effort of comparison with other
-observatories, where the local observatory issues a call for participation to a
-set of additional possible observatories, each of which can accept or reject the
-call.  The local observatory then chooses one of the accepting observatories
-from which to obtain data, and that observatory takes data and returns it. The
-local observatory then decides what to do: classify the object, contact a human,
-or call for more observations (automatically to other observatories that decide
-whether or not to accept, and when)
+\item Using the existing object catalog (containing the known orbits of all known objects), find objects which are expected to appear in subtracted image, given the area of sky and time of day.
+Cross-match expected objects with sources in subtracted image, resulting in the unmatched-source catalog, which contains sources that cannot be matched with a previously known object.
+If these sources can be tracked over a few images, an orbit can be determined for them and they can be added to the object catalog.
+Also, the orbit catalog can be updated with redetections of known objects; each rediscovery provides information that improves the known orbits.
+
+\item Attempt to classify all entries in the orbit catalog (i.e., all the objects which now have known orbits).
+If any Near Earth Objects are detected to be passing close to the earth, alerts are generated to astronomers so that follow up observations can be scheduled.
+
+\item If an unknown object is found that cannot be classified locally, at least two possible options exist.
+One is a coordinated effort of comparison with other observatories, where the local observatory issues a call for participation to a set of additional possible observatories, each of which can accept or reject the call.
+The local observatory then chooses one of the accepting observatories from which to obtain data, and that observatory takes data and returns it.
+The local observatory then decides what to do: classify the object, contact a human, or call for more observations (automatically to other observatories that decide whether or not to accept, and when)
 
 A second option is less coordinated, where this observatory would send VOEvent
-messages, notifications of transient events which can be used to trigger
-follow-up observations on other telescopes. These are small XML messages
+messages, notifications of transient events which can be used to
+trigger follow-up observations on other telescopes. These are small XML messages
 containing basic information about the position of a transient source and a
@@
 
- \subsubsection*{Big data aspects} %3. {\bf What are the big data aspects?}
-
-%\katznote{from answer to question 4b below} %\ysnote{The FITS images themselves
-may be less interesting since they may only be used ocassionally % (e.g. for
-re-processing images using a new algorithm) after the initial object extraction
-into % databases. The object databases sizes may be more relevant, and they can
-be 3-orders of magnitude smaller % than the images.}
-
-LSST will generate 36~GB of data every 30 seconds, and over a 10-hour winter
-night, will collect up to 30~TB.  Data will be stored as FITS image, and in
-large databases containing analyses of the discoveries.
-
- \subsubsection*{Distributed aspects} %4. {\bf What are the distributed
-aspects?}
+ \subsubsection*{Big data aspects}
+%3. {\bf What are the big data aspects?}
+
+%\katznote{from answer to question 4b below}
+%\ysnote{The FITS images themselves may be less interesting since they may only be used ocassionally
+%  (e.g. for re-processing images using a new algorithm) after the initial object extraction into
+%  databases. The object databases sizes may be more relevant, and they can be 3-orders of magnitude smaller
+%  than the images.}
+
+LSST will generate 36~GB of data every 30 seconds, and over a 10-hour winter night, will collect up to 30~TB.
+Data will be stored as FITS image, and in large databases containing analyses of  the discoveries.
+
+ \subsubsection*{Distributed aspects}
+ %4. {\bf What are the distributed aspects?}
 
@@
 
-The overall system will contain a central telescope with computing data and
-resources, linked to a wide area network of observatories and data and computing
-resources.  %\alnote{are these data and computing resources still part %of the
-application LSST?} \katznote{I'm not sure - how do you think we should handle
-this?} \alnote{not sure, but since we say that processing and storage is part of
-the app, the resource for this need to be at some place}
-
- \subsubsection*{Dynamic aspects} %5. {\bf What are the dynamic aspects?}
+The overall system will contain a central telescope with computing data and resources, linked to a wide area network of observatories and data and computing resources.
+%\alnote{are these data and computing resources still part
+%of the application LSST?} \katznote{I'm not sure - how do you think we should handle this?} \alnote{not sure, but since we say that processing and storage is part of the app, the resource for this need to be at some place}
+
+ \subsubsection*{Dynamic aspects}
+%5. {\bf What are the dynamic aspects?}
 
@@
 
-The time constraints differ for different types of transient sources, but in
-some cases notification of remote telescopes must be made within a few minutes
-if it is to be useful; the overheads associated with setting up a follow-up
-observation mean that much shorter time scales are practically irrelevant,
-although they are still scientifically interesting. The quality constraints
-center on providing a remote telescope with good enough information about a
-potential transient source that it can make a reliable decision about whether to
-interrupt its existing observing program to follow-up the event. In most cases,
-the candidate transient will be some sort of noise event, so fairly
-sophisticated filtering is required to bring the false positive rate down to
-something acceptable, given the cost of observing time on cutting edge
-telescopes.
-
- \subsubsection*{Other important issues} %6. {\bf What else is important?}
-
-%\katznote{from answers to questions 7,8 below, also address programming systems
-(how was the app written; what tools does it use)}
-
-%\alnote{Is the workflow part still part of the app as defined before with the
-stages: data transformation \& decision and storage?} \katznote{I think this is
-data transformation.}\alnote{ok}
-
-Data will initially be analyzed at the telescope; discoveries will be sent to
-other centers.
-
-Pan-STARRS, a contemporary sky survey with similar goals as LSST, uses a
-scripted programming model on a cluster for image processing of telescope frames
-to extract object attributes as a CSV file.  This is followed by followed by
-scientific workflows, based on Trident, that load the files into individual
-databases daily, and merges them with a master distributed-database on a weekly
-basis on an HPC cluster~\cite{Simmhan:Building:2009}. Data extraction and mining
-operations by astronomers are composed using a pseudo-SQL language with user
-defined functions and submitted for execution on the master database using a
-batch system.
+The time constraints differ for different types of transient sources, but in some
+cases notification of remote telescopes must be made within a few minutes if it is to be useful; the
+overheads associated with setting up a follow-up observation mean that much
+shorter time scales are practically irrelevant, although they are still
+scientifically interesting. The quality constraints center on providing a
+remote telescope with good enough information about a potential transient source that
+it can make a reliable decision about whether to interrupt its existing
+observing program to follow-up the event. In most cases, the candidate
+transient will be some sort of noise event, so fairly sophisticated filtering is
+required to bring the false positive rate down to something acceptable, given
+the cost of observing time on cutting edge telescopes.
+
+ \subsubsection*{Other important issues}
+%6. {\bf What else is important?}
+
+%\katznote{from answers to questions 7,8 below, also address programming systems (how was the app written; what tools does it use)}
+
+Data will initially be analyzed at the telescope; discoveries will be sent to other centers.
+
+Pan-STARRS, a contemporary sky survey with similar goals as LSST,  uses a scripted programming model
+on a cluster for image processing of telescope frames to extract object attributes as a CSV
+file.
+%\alnote{Is the workflow part still part of the app as defined before with the stages: data transformation \& decision and storage?} \katznote{I think this is data transformation.}\alnote{ok}
+This is followed by followed by scientific workflows, based on Trident, that load the files
+into individual databases daily, and merges them with a master distributed-database on a weekly basis on an
+HPC cluster~\cite{Simmhan:Building:2009}. Data extraction and mining operations by astronomers are composed using a pseudo-SQL
+language with user defined functions and submitted for execution on the master database using a batch system.
 
@@
 %\begin{enumerate}
-%\item What is the science/engineering problem? What is the purpose of the
-application?
+%\item What is the science/engineering problem? What is the purpose of the application?
 %
@@
 %
-%Processing of data (data analysis and detection of new sources) from each image
-should be done while the next image is being taken by the telescope.  This means
-that if anything unusual is detected, normal observation can be
-interrupted. Other observing resources can also be notified instantly, so they
-can observe the same event. As data are collected, they will be added to all the
-data previously detected from the same location of sky to create a very deep
-master image. LSST will also build up a database of all known moving objects.
-%
-%Every time a new image of the sky is obtained, the master image will be
-subtracted from it. The result is a subtracted image that contains the
-difference between the current sky image and its average state.
-%
-%This subtracted image is then processed by a cluster of computers with the
-following steps:
+%Processing of data (data analysis and detection of new sources) from each image should be done while the next image is being taken by the telescope.
+This means that if anything unusual is detected, normal observation can be interrupted.
+Other observing resources can also be notified instantly, so they can observe the same event.
+As data are collected, they will be added to all the data previously detected from the same location of sky to create a very deep master image.
+LSST will also build up a database of all known moving objects.
+%
+%Every time a new image of the sky is obtained, the master image will be subtracted from it.
+The result is a subtracted image that contains the difference between the current sky image and its average state.
+%
+%This subtracted image is then processed by a cluster of computers with the following steps:
 %\begin{enumerate}
 %
-%\item Query object catalog (knows orbits of all known objects) to find objects
-which are expected to appear in subtracted image, given the area of sky, time of
-day.
-%
-%\item Cross-match expected objects with sources in subtracted image. The result
-is the unmatched-source catalog, which contains sources that can't be matched
-with a previously known object
-%
-%\item Use unmatched-source catalogue to compute pairs of detections separated
-by short time intervals, called tracklets. A tracklet is an observable, short
-section of orbit. In order to create pairs of detections, all objects in the
-unmatched-source catalog are queried against the orbit catalog, in an attempt to
-obtain data about these objects from earlier observations. If earlier
-observations and current observations can be linked to form tracklets, they are
-stored in the tracklet catalog.
-%
-%\item Use tracklet catalog to attempt to link these tracklets over a larger
-time window. This is achieved by again querying the orbit catalog, this time
-looking for observations of the same objects even further back in time. If
-matches can be found and the tracklets can be extrapolated out into longer
-sections of orbit, then they are added as new orbits to the orbit catalog.
-%
-%\item Update orbit catalog with redetections of known objects. Each rediscovery
-provides information that improves the known orbits.
-%
-%\item Attempt to classify all entries in the orbit catalog (i.e., all the
-objects which now have known orbits). If any Near Earth Objects are detected to
-be passing close to the earth, alerts are generated to astronomers so that
-follow up observations can be scheduled.
-%
-%\item Compare with other observatories: If an unknown object is found that
-cannot be classified locally
+%\item Query object catalog (knows orbits of all known objects) to find objects which are expected to appear in subtracted image, given the area of sky, time of day.
+%
+%\item Cross-match expected objects with sources in subtracted image. The result is the unmatched-source catalog, which contains sources that can't be matched with a previously known object
+%
+%\item Use unmatched-source catalogue to compute pairs of detections separated by short time intervals, called tracklets.
+A tracklet is an observable, short section of orbit.
+In order to create pairs of detections, all objects in the unmatched-source catalog are queried against the orbit catalog, in an attempt to obtain data about these objects from earlier observations.
+If earlier observations and current observations can be linked to form tracklets, they are stored in the tracklet catalog.
+%
+%\item Use tracklet catalog to attempt to link these tracklets over a larger time window.
+This is achieved by again querying the orbit catalog, this time looking for observations of the same objects even further back in time.
+If matches can be found and the tracklets can be extrapolated out into longer sections of orbit, then they are added as new orbits to the orbit catalog.
+%
+%\item Update orbit catalog with redetections of known objects. Each rediscovery provides information that improves the known orbits.
+%
+%\item Attempt to classify all entries in the orbit catalog (i.e., all the objects which now have known orbits).
+If any Near Earth Objects are detected to be passing close to the earth, alerts are generated to astronomers so that follow up observations can be scheduled.
+%
+%\item Compare with other observatories: If an unknown object is found that cannot be classified locally
 %\begin{enumerate}
@@
 %\item each observatory accepts or rejects the call
-%\item local observatory chooses from accepting observatories one to obtain data
-from
+%\item local observatory chooses from accepting observatories one to obtain data from
 %\item that observatory takes data and sends it to local observatory
-%\item local observatory decides what to do - classify the object, contact a
-human, call for more observations (automatically to other observatories, that
-decide whether or not to accept, and when)
+%\item local observatory decides what to do - classify the object, contact a human, call for more observations (automatically to other observatories, that decide whether or not to accept, and when)
 %\end{enumerate}
@@
 %
-%large (exascale) data and large (exascale) computing (e.g., data mining,
-analytics, bioinformatics)
+%large (exascale) data and large (exascale) computing (e.g., data mining, analytics, bioinformatics)
 %
@@
 %
-%LSST will generate 36~GB of data every 30 seconds, and over a 10-hour winter
-night, will collect up to 30~TB.
-%
-%3 What infrastructure will be used? (including compute, data, network,
-instruments, etc.)
+%LSST will generate 36~GB of data every 30 seconds, and over a 10-hour winter night, will collect up to 30~TB.
+%
+%3 What infrastructure will be used? (including compute, data, network, instruments, etc.)
 %
@@
 %
-%Some will be analyzed at the mountain top, discoveries will be sent to other
-centers.
-%
-%6 Are there time (or quality) constraints on the application?  (If so, what are
-they?)
+%Some will be analyzed at the mountain top, discoveries will be sent to other centers.
+%
+%6 Are there time (or quality) constraints on the application?  (If so, what are they?)
 %
@@
 %
-%7 Please feel free to also talk about the current state of the application, if
-it exists today, and any specific gaps that you know will need to be overcome.
+%7 Please feel free to also talk about the current state of the application, if it exists today, and any specific gaps that you know will need to be overcome.
 %
@@
 
-An increasingly important feature of astronomical research is the existence of
-systematic sky surveys covering the whole sky (or large fractions thereof) in
-many different wavebands. The principal outcome from these surveys are
-astronomical source catalogues, which contain values for attributes
-characterizing all the celestial sources detected in the sky survey dataset.
-These are now typically implemented in relational databases, with SQL interfaces
-implemented through web pages, and the principal goal of the Virtual Observatory
+An increasingly important feature of astronomical research is the
+existence of systematic sky surveys covering the whole sky (or large
+fractions thereof) in many different wavebands. The principal outcome
+from these surveys are astronomical source catalogues, which contain
+values for attributes characterizing all the celestial sources
+detected in the sky survey dataset.  These are now typically
+implemented in relational databases, with SQL interfaces implemented
+through web pages, and the principal goal of the Virtual Observatory
 (VO) being developed by the International Virtual Observatory Alliance
-(IVOA)~\cite{ivoa} is to standardize access to these, and other astronomical
-data resources, so that astronomers can readily perform multi-wavelength
-analyses, combining all the extant data on particular sources.  The previous
-application (\S\ref{astroAdam}) was an example of a ``generator'' of such data
-resources, this application is a typical ``user'' of them.  As such, it can
-generally be considered as having just the last application stage: data
-processing.  Thus it is an {\em infrastructural application}.  The particular
-example described here calculates the photometric redshift for a given area of
+(IVOA)~\cite{ivoa} is to standardize access to these, and other
+astronomical data resources, so that astronomers can readily perform
+multi-wavelength analyses, combining all the extant data on particular
+sources.  The previous application (\S\ref{astroAdam}) was an example
+of a ``generator'' of such data resources, this application is a
+typical ``user'' of them.  As such, it can generally be considered as
+having just the last application stage: data processing.  Thus it is
+an {\em infrastructural application}.  The particular example
+described here calculates the photometric redshift for a given area of
 sky using two tools: HyperZ and ANNz, both accessible via web services
-interfaces~\cite{barker}.  However, it represents many service-oriented
-architecture applications that are designed to work with the VO.
-
-\subsubsection*{Application description} %2. {\bf How does it do it?}
-
-%\katznote{from answers to questions 2,3,5,6 below, also address
-cyberinfrastructure (where does the app run) -- perhaps have a diagram here of
-what happens in the app}
+interfaces~\cite{barker}.  However, it represents many
+service-oriented architecture applications that are designed to work
+with the VO.
+
+\subsubsection*{Application description}
+%2. {\bf How does it do it?}
+
+%\katznote{from answers to questions 2,3,5,6 below, also address cyberinfrastructure (where does the app run) -- perhaps have a diagram here of what happens in the app}
 
@@
 
-The application orchestrates a set of services through a pipeline in order to
-compute the redshift of a given area of sky.  It contains a set of components.
-1) The Wide Field Survey Archive (WFS), which is an image catalog. 2) A
-spectroscopic database, used to provide spectroscopic data for an area of
-sky. 3) The SExtractor tool, which is an image processing that extracts all
-objects of interest (stars, galaxies, etc.)  4) A cross matching tool that
-compiles one table of all objects of interest over the five wavebands.  5)
-Hyperz, the first photometric redshift estimation algorithm.  6) ANNz, the
-second photometric redshift estimation algorithm, and 7) mySpace, the AstroGrid
-storage service.
-
- \subsubsection*{Big data aspects} %3. {\bf What are the big data aspects?}
+The application orchestrates a set of services through a pipeline in order to compute the redshift of a given area of sky.
+It contains a set of components.
+1) The Wide Field Survey Archive (WFS), which is an image catalog.
+2) A spectroscopic database, used to provide spectroscopic data for an area of sky.
+3) The SExtractor tool, which is an image processing that extracts all objects of interest (stars, galaxies, etc.)  4) A cross matching tool that compiles one table of all objects of interest over the five wavebands.
+5) Hyperz, the first photometric redshift estimation algorithm.
+6) ANNz, the second photometric redshift estimation algorithm, and 7) mySpace, the AstroGrid storage service.
+
+ \subsubsection*{Big data aspects}
+%3. {\bf What are the big data aspects?}
 
@@
 
- \subsubsection*{Distributed aspects} %4. {\bf What are the distributed
-aspects?}
+ \subsubsection*{Distributed aspects}
+ %4. {\bf What are the distributed aspects?}
 
@@
 distributed internationally. Major datasets---e.g. those from large sky
-surveys---are typically implemented in relational databases running on high-spec
-servers or clusters thereof.  The services in this application are also
-distributed, with web service interfaces used to link them.
+surveys---are typically implemented in relational databases running on high-spec servers
+or clusters thereof.  The services in this application are also distributed, with web service interfaces used to link them.
 
@@
 
- \subsubsection*{Dynamic aspects} %5. {\bf What are the dynamic aspects?}
+ \subsubsection*{Dynamic aspects}
+%5. {\bf What are the dynamic aspects?}
 
@@
 
-The WFS catalog will be updated regularly, therefore a query at time $x$ will
-not provide the same data as a query at time $y$.  In addition, the specific
-queries that will be issued by users are unknown, and will change over time.
-
- \subsubsection*{Other important issues} %6. {\bf What else is important?}
-
-%\katznote{from answers to questions 7,8 below, also address programming systems
-(how was the app written; what tools does it use)}
-
-Enough data needs to be available for a given area of sky from both the WFS
-archive and the spectroscopic archive.
+The WFS catalog will be updated regularly, therefore a query at time $x$ will not provide the same data as a query at time $y$.
+In addition, the specific queries that will be issued by users are unknown, and will change over time.
+
+ \subsubsection*{Other important issues}
+%6. {\bf What else is important?}
+
+%\katznote{from answers to questions 7,8 below, also address programming systems (how was the app written; what tools does it use)}
+
+Enough data needs to be available for a given area of sky from both the WFS archive and the spectroscopic archive.
 
@@
 %
-%Calculates the photometric redshift for a given area of sky [RA, DEC] using two
-tools: HyperZ and ANNz, both accessible via WS interfaces.
+%Calculates the photometric redshift for a given area of sky [RA, DEC] using two tools: HyperZ and ANNz, both accessible via WS interfaces.
 %
@@
 %
-%The application is a service-oriented workflow. Services are orchestrated
-through a pipeline in order to compute the redshift of a given area of sky.
-%
-%3. What infrastructure is used? (including compute, data, network, instruments,
-etc.)
-%
-%The Wide Field Survey Archive (WFS): image catalogue %Spectroscopic database:
-used to provide spectroscopic data for an area of sky
-%
-%SExtractor tool: image processing, extracts all objects of interest (stars,
-galaxies etc.)  %Cross matching tool: compiles one table of all objects of
-interest over the 5 wavebands
-%
-%Hyperz: photometric redshift estimation algorithm 1 %ANNz: photometric redshift
-estimation algorithm 2
+%The application is a service-oriented workflow. Services are orchestrated through a pipeline in order to compute the redshift of a given area of sky.
+%
+%3. What infrastructure is used? (including compute, data, network, instruments, etc.)
+%
+%The Wide Field Survey Archive (WFS):  image catalogue
+%Spectroscopic database: used to provide spectroscopic data for an area of sky
+%
+%SExtractor tool: image processing, extracts all objects of interest (stars, galaxies etc.)
+%Cross matching tool: compiles one table of all objects of interest over the 5 wavebands
+%
+%Hyperz: photometric redshift estimation algorithm 1
+%ANNz: photometric redshift estimation algorithm 2
 %
@@
 %
-%The WFS catalogue will be updated regularly, therefore a query at time x will
-not provide the same data as a query at time y.
+%The WFS catalogue will be updated regularly, therefore a query at time x will not provide the same data as a query at time y.
 %
@@
 %
-%Data sets (images) will be in the GB range, I can attempt to clarify these
-details.
+%Data sets (images) will be in the GB range, I can attempt to clarify these details.
 %
@@
 %
-%Enough data needs to be available for a given area of sky from both the WFS
-archive and the spectroscopic archive.
+%Enough data needs to be available for a given area of sky from both the WFS archive and the spectroscopic archive.
 
@@
 
-This {\em traditional application} performs data simulation and analysis to
-understand the Cosmic Microwave Background (CMB)~\cite{cmb}, which is an image
-of the Universe as it was 400,000 years after the Big Bang. Tiny fluctuations in
-the CMB temperature and polarization encode the fundamental parameters of
-cosmology and, using the Big Bang as the ultimate particle accelerator,
-ultra-high energy physics. Extracting this information from the data gathered by
-current and anticipated CMB observations is an extremely computationally
-intensive endeavor for which massively parallel tools have been developed.  This
-application contains just one stage: data processing.
-
-\subsubsection*{Application description} %2. {\bf How does it do it?}
-
-%\katznote{from answers to questions 2,3,5,6 below, also address
-cyberinfrastructure (where does the app run) -- perhaps have a diagram here of
-what happens in the app}
-
-The CMB community wants to extract cosmology and fundamental physics from the
-observations gathered by the detectors as time-ordered sequences of O($10^{12}$
-- $10^{15}$). These observations are reduced first to a map of O($10^6$ -
-$10^8$) sky pixels, then to O($10^3$ - $10^4$) angular power spectrum
-coefficients, and finally to O(10) cosmological parameters~\cite{cmb2}.
-
-The central computing challenge for any CMB dataset is the simulation and
-analysis of O($10^4$) synthetic observations, used to correct for biases and
-quantify uncertainties in the analysis of the real data.  Preconditioned
-conjugate gradient techniques are used to solve for the maximum likelihood sky
-map given the input data (obtained by scanning the sky with hundreds to
-thousands of detectors for weeks to years) and its piecewise stationary noise
-statistics.  To avoid the I/O bottleneck inherent in the traditional
-simulate/write/read/map paradigm, all simulations are performed on the fly only
-when requested by the map-making code.
-
-Going from the map to the angular power spectrum is the computationally most
-expensive step. The exact solution scales with the cube of the number of pixels
-(in the map), so going from map to angular power spectrum is ruled out now (but
-was possible earlier for much smaller observations). The approximate solution
-requires sets of O($10^4$) Monte Carlo realizations of the observed sky to
-remove biases and quantify uncertainties, each of which involves simulating and
-mapping the time-ordered data.
-
-The map-making application is therefore applied to both real and simulated data,
-but many more times to simulated data (which requires us to use the on-the-fly
-simulation module too).
-
-%The computing scales with the number of observation samples, while the
-communication and I/O scale with the number of pixels in the map. For the
-current Planck satellite experiment these are O($10^{12}$) and O($10^8$)
-respectively; for the proposed CMBpol satellite they will be O($10^{15}$) and
-O($10^8$); that is, a 1000x increase in calculation cost is expected over the
-next 15 years (including intermediate suborbital pathfinder experiments) but the
-communication and I/O costs will be fixed.
-
- \subsubsection*{Big data aspects} %3. {\bf What are the big data aspects?}
+This {\em traditional application} performs data simulation and analysis to understand the Cosmic Microwave Background (CMB)~\cite{cmb}, which is an image of the Universe as it was 400,000 years after the Big Bang.
+Tiny fluctuations in the CMB temperature and polarization encode the fundamental parameters of cosmology and, using the Big Bang as the ultimate particle accelerator, ultra-high energy physics.
+Extracting this information from the data gathered by current and anticipated CMB observations is an extremely computationally intensive endeavor for which massively parallel tools have been developed.
+This application contains just one stage: data processing.
+
+\subsubsection*{Application description}
+%2. {\bf How does it do it?}
+
+%\katznote{from answers to questions 2,3,5,6 below, also address cyberinfrastructure (where does the app run) -- perhaps have a diagram here of what happens in the app}
+
+The CMB community wants to extract cosmology and fundamental physics
+from the observations gathered by the detectors as time-ordered
+sequences of O($10^{12}$ - $10^{15}$). These observations are reduced
+first to a map of O($10^6$ - $10^8$) sky pixels, then to O($10^3$ -
+$10^4$) angular power spectrum coefficients, and finally to O(10)
+cosmological parameters~\cite{cmb2}.
+
+The central computing challenge for any CMB dataset is the simulation and analysis of O($10^4$) synthetic observations, used to correct for biases and quantify uncertainties in the analysis of the real data.
+Preconditioned conjugate gradient techniques are used to solve for the maximum likelihood sky map given the input data (obtained by scanning the sky with hundreds to thousands of detectors for weeks to years) and its piecewise stationary noise statistics.
+To avoid the I/O bottleneck inherent in the traditional simulate/write/read/map paradigm, all simulations are performed on the fly only when requested by the map-making code.
+
+Going from the map to the angular power spectrum is the
+computationally most expensive step. The exact solution scales with
+the cube of the number of pixels (in the map), so going from map to
+angular power spectrum is ruled out now (but was possible earlier for
+much smaller observations). The approximate solution requires sets of
+O($10^4$) Monte Carlo realizations of the observed sky to remove
+biases and quantify uncertainties, each of which involves simulating
+and mapping the time-ordered data.
+
+The map-making application is therefore applied to both real and
+simulated data, but many more times to simulated data (which requires
+us to use the on-the-fly simulation module too).
+
+%The computing scales with the number of observation samples, while the communication and I/O scale with the number of pixels in the map.
+For the current Planck satellite experiment these are O($10^{12}$) and O($10^8$) respectively; for the proposed CMBpol satellite they will be O($10^{15}$) and O($10^8$); that is, a 1000x increase in calculation cost is expected over the next 15 years (including intermediate suborbital pathfinder experiments) but the communication and I/O costs will be fixed.
+
+ \subsubsection*{Big data aspects}
+%3. {\bf What are the big data aspects?}
 
@@
 
-This application is targeting the largest supercomputers available: Hopper,
-Tianhe, Blue Waters, etc. It is all about the cycles, although as the
-concurrency increases previously solved I/O and communication scaling
-bottlenecks typically re-emerge.
-
-The application currently uses a single HPC systems, but the scientists have
-discussed using distributed systems, with a model of remote systems being used
-for the data simulations.  Each simulation would be launched from the central
-system that is building the map, and output data from the simulations would be
-asynchronously delivered back to that central system as files that would be
-incorporated in the map as they are produced.
-
- \subsubsection*{Dynamic aspects} %5. {\bf What are the dynamic aspects?}
+This application is targeting the largest supercomputers available: Hopper, Tianhe, Blue Waters, etc.
+It is all about the cycles, although as the concurrency increases previously solved I/O and communication scaling bottlenecks typically re-emerge.
+
+The application currently uses a single HPC systems, but the scientists have discussed using distributed systems, with a model of remote systems being used for the data simulations.
+Each simulation would be launched from the central system that is building the map, and
+output data from the simulations would be asynchronously delivered back to that central
+system as files that would be incorporated in the map as they are produced.
+
+ \subsubsection*{Dynamic aspects}
+%5. {\bf What are the dynamic aspects?}
 
@@
 
-Overall, this application can make use of whatever computing it can access in
-order to run the simulations needed to flesh out the observed data (which can be
-read from local disk).  The simulations are dynamic in time and in location.
-
- \subsubsection*{Other important issues} %6. {\bf What else is important?}
-
-%\katznote{from answers to questions 7,8 below, also address programming systems
-(how was the app written; what tools does it use)}
-
-A goal is to simulate and map O($10^2$) realizations of an experiment's data in
-O(1) wallclock hour to provide 10\% errors during the early analysis stages, and
-O($10^4$) realizations in O(100) wallclock hours for the final definitive
-analysis.
+Overall, this application can make use of whatever computing it can access in order to run the simulations needed to flesh out the observed data (which can be read from local disk).
+The simulations are dynamic in time and in location.
+
+ \subsubsection*{Other important issues}
+%6. {\bf What else is important?}
+
+%\katznote{from answers to questions 7,8 below, also address programming systems (how was the app written; what tools does it use)}
+
+A goal is to simulate and map O($10^2$) realizations of an experiment's data in O(1) wallclock hour to provide 10\% errors during the early analysis stages, and O($10^4$) realizations in O(100) wallclock hours for the final definitive analysis.
 
@@
 %
-%1. What is the science/engineering problem? What is the purpose of the
-application?
-%
-%Cosmic Microwave Background (CMB) data simulation and analysis. The CMB is an
-image of the Universe as it was only 400,000 years after the Big Bang. Tiny
-fluctuations in the CMB temperature and polarization encode the fundamental
-parameters of cosmology and, using the Big Bang as the ultimate particle
-accelerator, ultra-high energy physics. Extracting this information from the
-data gathered by current and anticipated CMB observations is an extremely
-computationally intensive endeavor for which we have developed our massively
-parallel tools.
+%1. What is the science/engineering problem? What is the purpose of the application?
+%
+%Cosmic Microwave Background (CMB) data simulation and analysis.
+The CMB is an image of the Universe as it was only 400,000 years after the Big Bang.
+Tiny fluctuations in the CMB temperature and polarization encode the fundamental parameters of cosmology and, using the Big Bang as the ultimate particle accelerator, ultra-high energy physics.
+Extracting this information from the data gathered by current and anticipated CMB observations is an extremely computationally intensive endeavor for which we have developed our massively parallel tools.
 %
@@
 %
-%The central computing challenge for any CMB dataset is the simulation and
-analysis of O($10^4$) synthetic observations, used to correct for biases and
-quantify uncertainties in the analysis of the real data. We use preconditioned
-conjugate gradient techniques to solve for the maximum likelihood sky map given
-the input data (obtained by scanning the sky with hundreds to thousands of
-detectors for weeks to years) and its piecewise stationary noise statistics. To
-avoid the IO bottleneck inherent in the traditional simulate/write/read/map
-paradigm, all simulations are performed on the fly only when requested by the
-map-making code.
+%The central computing challenge for any CMB dataset is the simulation and analysis of O($10^4$) synthetic observations, used to correct for biases and quantify uncertainties in the analysis of the real data.
+We use preconditioned conjugate gradient techniques to solve for the maximum likelihood sky map given the input data (obtained by scanning the sky with hundreds to thousands of detectors for weeks to years) and its piecewise stationary noise statistics.
+To avoid the IO bottleneck inherent in the traditional simulate/write/read/map paradigm, all simulations are performed on the fly only when requested by the map-making code.
 %
@@
 %
-% i) large (exascale) data and large (exascale) computing (e.g., data mining,
-analytics, bioinformatics)
+%     i) large (exascale) data and large (exascale) computing (e.g., data mining, analytics, bioinformatics)
 %
@@
 %
-%The computing scales with the number of observation samples, while the
-communication and IO scale with the number of pixels in the map. For the current
-Planck satellite experiment these are O($10^12$) and O($10^8$) respectively; for
-the proposed CMBpol satellite they will be O($10^15$) and O($10^8$) - that is,
-we expect a 1000x increase in calculation cost over the next 15 years (including
-intermediate suborbital pathfinder experiments) but fixed communication/IO
-costs.
-%
-%3. What infrastructure will be used? (including compute, data, network,
-instruments, etc.)
-%
-%We are targeting the largest supercomputers available - Hopper, Tianhe, Blue
-Waters, etc. It's all about the cycles, although as the concurrency we work at
-increases previously solved IO and communication scaling bottlenecks typically
-re-emerge.
+%The computing scales with the number of observation samples, while the communication and IO scale with the number of pixels in the map.
+For the current Planck satellite experiment these are O($10^12$) and O($10^8$) respectively; for the proposed CMBpol satellite they will be O($10^15$) and O($10^8$) - that is, we expect a 1000x increase in calculation cost over the next 15 years (including intermediate suborbital pathfinder experiments) but fixed communication/IO costs.
+%
+%3. What infrastructure will be used? (including compute, data, network, instruments, etc.)
+%
+%We are targeting the largest supercomputers available - Hopper, Tianhe, Blue Waters, etc.
+It's all about the cycles, although as the concurrency we work at increases previously solved IO and communication scaling bottlenecks typically re-emerge.
 %
@@
 %
-%6. Are there time (or quality) constraints on the application?  (If so, what
-are they?)
-%
-%Typically we want to be able to simulate and map O($10^2$) realizations of an
-experiment's data in O(1) wallclock hour to provide 10\% errors during the early
-analysis stages, and O($10^4$) realizations in O(100) wallclock hours for the
-final definitive analysis.
-%
-%7. Please feel free to also talk about the current state of the application, if
-it exists today, and any specific gaps that you know will need to be overcome.
-%
-%Moving to the current and future generations of heterogeneous many-core systems
-will require a significant re-working of our analysis codes, both to deal with
-communication over hundreds of thousands of cores and to take advantage of
-accelerator opportunities such as GPUs.
+%6. Are there time (or quality) constraints on the application?  (If so, what are they?)
+%
+%Typically we want to be able to simulate and map O($10^2$) realizations of an experiment's data in O(1) wallclock hour to provide 10\% errors during the early analysis stages, and O($10^4$) realizations in O(100) wallclock hours for the final definitive analysis.
+%
+%7. Please feel free to also talk about the current state of the application, if it exists today, and any specific gaps that you know will need to be overcome.
+%
+%Moving to the current and future generations of heterogeneous many-core systems will require a significant re-working of our analysis codes, both to deal with communication over hundreds of thousands of cores and to take advantage of accelerator opportunities such as GPUs.
 
@@
 
-Marine sensing covers a number of applications, including environmental
-monitoring, remote exploration, marine life surveys, and habitat assessment. In
-each case, the main challenge is collecting data reliably from a difficult
-working environment, without interfering with the natural behaviors the animals
-exhibit.
-
-A canonical example is the monitoring of seal and other sea mammal populations
-by the Scottish Oceans Institute (SOI), which tags animals with sensor packages
-that can record dive behavior, speed, and movement~\cite{SMSSeal,SealContact}.
-These datasets are then analyzed offline using traditional statistical
-techniques and integrated with Google Earth to visualize animal tracks on a
-large scale.  This is an {\em infrastructural application}. It begins after data
-generation, and it includes the data storage and processing stages.
-%\alnote{data transformation?} \katznote{I don't think so - I think all data %is
-stored and then processed as users want} \alnote{I think most sensors do some
-form of filtering/aggregation before transmitting things to a
-backend...}\katznote{but I would argue that the application %starts after the
-filtering/aggregation}
-
-\subsubsection*{Application description} %2. {\bf How does it do it?}
-
-%\katznote{from answers to questions 2,3,5,6 below, also address
-cyberinfrastructure (where does the app run) -- perhaps have a diagram here of
-what happens in the app}
-
-The sensor packages on the marine mammals report back to base when the animal
-crawls up a beach and comes within range of a cellular network. Data are
-collected remotely and brought to a central site for analysis. Analysis includes
-statistical applications and visualization is done using Google Earth.
-
- \subsubsection*{Big data aspects} %3. {\bf What are the big data aspects?}
+Marine sensing covers a number of applications, including
+environmental monitoring, remote exploration, marine life surveys, and
+habitat assessment. In each case, the main challenge is collecting
+data reliably from a difficult working
+environment, without interfering with the natural behaviors the
+animals exhibit.
+
+A canonical example is the monitoring of seal and other sea mammal
+populations by the Scottish Oceans Institute (SOI), which tags animals
+with sensor packages that can record dive behavior, speed, and
+movement~\cite{SMSSeal,SealContact}.  These datasets are then analyzed
+offline using traditional statistical techniques and integrated with
+Google Earth to visualize animal tracks on a large scale.  This is an
+{\em infrastructural application}. It begins after data generation,
+and it includes the data storage and processing stages.
+%\alnote{data transformation?} \katznote{I don't think so - I think all data
+%is stored and then processed as users want} \alnote{I think most sensors do some form of filtering/aggregation before transmitting things to a backend...}\katznote{but I would argue that the application
+%starts after the filtering/aggregation}
+
+\subsubsection*{Application description}
+%2. {\bf How does it do it?}
+
+%\katznote{from answers to questions 2,3,5,6 below, also address cyberinfrastructure (where does the app run) -- perhaps have a diagram here of what happens in the app}
+
+The sensor packages on the marine mammals report back to base when the animal crawls up a beach and comes within range of a cellular network.
+Data are collected remotely and brought to a central site for analysis.
+Analysis includes statistical applications and visualization is done using Google Earth.
+
+ \subsubsection*{Big data aspects}
+%3. {\bf What are the big data aspects?}
 
@@
 
-The data sets collected are modest in size by scientific-data standards. They
-include positions and motion vectors for sea mammals, and concentrations and
-gradients for environmental missions.
-
- \subsubsection*{Distributed aspects} %4. {\bf What are the distributed
-aspects?}
+The data sets collected are modest in size by scientific-data standards. They include positions and motion vectors for sea mammals, and concentrations and gradients for environmental missions.
+
+ \subsubsection*{Distributed aspects}
+ %4. {\bf What are the distributed aspects?}
 
@@
 
-Time series data is published from distributed sensors, but they are stored in a
-central repository. The data is analyzed locally and visualized by distributed
-users.
-
- \subsubsection*{Dynamic aspects} %5. {\bf What are the dynamic aspects?}
+Time series data is published from distributed sensors, but they are stored in a central repository. The data is analyzed locally and visualized by distributed users.
+
+ \subsubsection*{Dynamic aspects}
+%5. {\bf What are the dynamic aspects?}
 
@@
 
-Time series data is published dynamically from the sensors when they are within
-communication distance of cellular towers. So the time series data arrives at
-the central repository asynchronously relative to the time of collection.
-
-There is some basic adaptation to the resolution of the data collected to
-account for different resolutions of the sensor packs.  Larger-scale
-environmental sensing applications may make use of more structured and extensive
-adaptation, such as changing the sampling frequency and other management
-characteristics in response to the data being observed.
-
- \subsubsection*{Other important issues} %6. {\bf What else is important?}
-
-%\katznote{from answers to questions 7,8 below, also address programming systems
-(how was the app written; what tools does it use)}
-
-Collecting data reliably is challenging due to the distributed deployment of
-sensors in hostile environments.
+Time series data is published dynamically from the sensors when they are within communication distance of cellular towers.
+So the time series data arrives at the central repository asynchronously relative to the time of collection.
+
+There is some basic adaptation to the resolution of the data collected to account for different resolutions of the sensor packs.
+Larger-scale environmental sensing applications may
+make use of more structured and extensive adaptation, such as
+changing the sampling frequency and other management characteristics in
+response to the data being observed.
+
+ \subsubsection*{Other important issues}
+%6. {\bf What else is important?}
+
+%\katznote{from answers to questions 7,8 below, also address programming systems (how was the app written; what tools does it use)}
+
+Collecting data reliably is challenging due to the distributed deployment of sensors in hostile environments.
 
@@
 % \begin{itemize}
-% \item Environmental (Marine) -- sensing within a hostile environment, % data
-quality + availability of communication infrastructure (Simon)
-% \item Patient monitoring -- data quality, type of data to transmit % (Omer)
+% \item Environmental (Marine) -- sensing within a hostile environment,
+%   data quality + availability of communication infrastructure (Simon)
+% \item Patient monitoring -- data quality, type of data to transmit
+%   (Omer)
 % \end{itemize}
@@
 
-% Marine sensing covers a range of different applications, including %
-environmental monitoring, remote exploration, marine life surveys and % habitat
-assessment. In each case, the main challenges are collecting % data reliably in
-what is an extremely challenging technical % environment---far more so than in
-air---and (in the case of biological % surveillance), without interfering with
-the natural behaviors the % animals exhibit.
+% Marine sensing covers a range of different applications, including
+% environmental monitoring, remote exploration, marine life surveys and
+% habitat assessment. In each case, the main challenges are collecting
+% data reliably in what is an extremely challenging technical
+% environment---far more so than in air---and (in the case of biological
+% surveillance), without interfering with the natural behaviors the
+% animals exhibit.
 
@@
 
-% At present most applications are quite small-scale. A canonical example % is
-the monitoring of seal and other sea mammal populations by the % Scottish Oceans
-Institute (SOI), which works by tagging animals with sensor % packages that can
-analyze dive behavior, speed, movement and the like, % and report back to base
-when the animal comes within range of a cellular % network.
-
-% 3. What infrastructure is used? (including compute, data, network,
-instruments, etc.)
-
-% The data sets collected are not enormous by scientific-data standards, % and
-are analyzed offline using traditional statistical techniques.  % Recent work by
-SOI has included integration with Google Earth to allow % animal tracks to be
-visualized on a large scale.
+% At present most applications are quite small-scale. A canonical example
+% is the monitoring of seal and other sea mammal populations by the
+% Scottish Oceans Institute (SOI), which works by tagging animals with sensor
+% packages that can analyze dive behavior, speed, movement and the like,
+% and report back to base when the animal comes within range of a cellular
+% network.
+
+% 3. What infrastructure is used? (including compute, data, network, instruments, etc.)
+
+% The data sets collected are not enormous by scientific-data standards,
+% and are analyzed offline using traditional statistical techniques.
+% Recent work by SOI has included integration with Google Earth to allow
+% animal tracks to be visualized on a large scale.
 
@@
 
-% In the sea mammal case the data isn't all that dynamic, being mainly a % time
-series collected with only basic adaptation to the resolution etc.  %
-Larger-scale applications in environmental sensing in particular would % make
-use of more structured and extensive adaptation (e.g., % changing the sampling
-frequency and other management characteristics in % response to the data being
-observed).
+% In the sea mammal case the data isn't all that dynamic, being mainly a
+% time series collected with only basic adaptation to the resolution etc.
+% Larger-scale applications in environmental sensing in particular would
+% make use of more structured and extensive adaptation (e.g.,
+% changing the sampling frequency and other management characteristics in
+% response to the data being observed).
 
@@
 
-% For sea mammals: positions, motion vectors. For environmental missions: %
-concentrations and gradients. I'd have to enquire about dataset sizes, % but not
-excessive.
+% For sea mammals: positions, motion vectors. For environmental missions:
+% concentrations and gradients. I'd have to enquire about dataset sizes,
+% but not excessive.
 
@@
 
-% For the sea mammals the animal crawls up a beach, the sensor packages % sees
-the cellular network and transmits. For environmental missions % we're still
-wondering about that ourselves :-)
+% For the sea mammals the animal crawls up a beach, the sensor packages
+% sees the cellular network and transmits. For environmental missions
+% we're still wondering about that ourselves :-)
 
@@
 
-% Not sure I see what you mean here. Generally it's just a matter of % capturing
-at whatever resolution device the sensor pack can support.
-
-% I could probably get more complete -- perhaps *way* too complete :-) -- %
-answers from Mike Fedak in Scottish Oceans if that'd help.
+% Not sure I see what you mean here. Generally it's just a matter of
+% capturing at whatever resolution device the sensor pack can support.
+
+% I could probably get more complete -- perhaps *way* too complete :-) --
+% answers from Mike Fedak in Scottish Oceans if that'd help.
 
@@
 %
-%\katznote{summary below assigned to Shantenu - perhaps merge this with Sylvia's
-app's section}
+%\katznote{summary below assigned to Shantenu - perhaps merge this with Sylvia's app's section}
 %\subsubsection{new summary}
@@
 %
-%\katznote{from answers to questions 2,3,5,6 below, also address
-cyberinfrastructure (where does the app run) -- perhaps have a diagram here of
-what happens in the app}
+%\katznote{from answers to questions 2,3,5,6 below, also address cyberinfrastructure (where does the app run) -- perhaps have a diagram here of what happens in the app}
 %
@@
 %
-%\katznote{from answers to questions 7,8 below, also address programming systems
-(how was the app written; what tools does it use)}
+%\katznote{from answers to questions 7,8 below, also address programming systems (how was the app written; what tools does it use)}
 %
@@
 %
-%1. What is the science/engineering problem? What is the purpose of the
-application?
-%
-%Comparing metagenomic (DNA sequence) data sets against each %other. This could
-be to distinguish human disease biomarkers among %different individuals,
-determine different soil (or water or air or %...) microbial composition from
-different samples, etc.
+%1. What is the science/engineering problem? What is the purpose of the application?
+%
+%Comparing metagenomic (DNA sequence) data sets against each
+%other. This could be to distinguish human disease biomarkers among
+%different individuals, determine different soil (or water or air or
+%...) microbial composition from different samples, etc.
 %
@@
 %
-%In general this is a large (for biology -- don't think in terms of exascale!)
-data problem with large (all-vs-all comparison) computing. There are strategies
-to make this more scalable that will reduce the computing volume.
+%In general this is a large (for biology -- don't think in terms of exascale!) data problem with large (all-vs-all comparison) computing.
+There are strategies to make this more scalable that will reduce the computing volume.
 %
@@
 %
-% i) large (exascale) data and large (exascale) computing (e.g., data mining,
-analytics, bioinformatics)
+%         i) large (exascale) data and large (exascale) computing (e.g., data mining, analytics, bioinformatics)
 %
@@
 %
-%Potentially peta-bytes or exa-bytes of data. Not sure how to quantify the
-computing. (Huge memories will be potentially more valuable than huge numbers of
-cores -- although this is heresy to the physics crowd.)
-%
-%3. What infrastructure will be used? (including compute, data, network,
-instruments, etc.)
+%Potentially peta-bytes or exa-bytes of data.
+Not sure how to quantify the computing.
+(Huge memories will be potentially more valuable than huge numbers of cores -- although this is heresy to the physics crowd.)
+%
+%3. What infrastructure will be used? (including compute, data, network, instruments, etc.)
 %
@@
 %
-%Individual data sets will be terabyte-scale. There could be many thousands of
-such sets to compare. Having them all local may not be feasible (the transfer
-time from where they are generated by others may be prohibitive. Computing might
-then have to be distributed. Biology is wrestling with this now.
+%Individual data sets will be terabyte-scale.
+There could be many thousands of such sets to compare.
+Having them all local may not be feasible (the transfer time from where they are generated by others may be prohibitive.
+Computing might then have to be distributed.
+Biology is wrestling with this now.
 %
@@
 %
-%That is a major part of the problem! Data will (in general) be generated from
-other places around the world. Getting to the data will be an issue.
-%
-%6. Are there time (or quality) constraints on the application?  (If so, what
-are they?)
+%That is a major part of the problem! Data will (in general) be generated from other places around the world. Getting to the data will be an issue.
+%
+%6. Are there time (or quality) constraints on the application?  (If so, what are they?)
 %
@@
 %
-%7. Please feel free to also talk about the current state of the application, if
-it exists today, and any specific gaps that you know will need to be overcome.
-%
-%Biology is wrestling now with about 3 orders of magnitude increase in DNA
-sequence data generation capability in the past 2 years. All algorithms,
-architectures, storage, and networks have broken or will break soon. Some
-biology problems (not necessarily the one I describe here) are not good fits for
-core-rich and memory-poor modern architectures. For some problems (comparing
-k-mers across metagenomic data sets, for instance) you need terabyte-scale
-memory, not thousands of cores.
+%7. Please feel free to also talk about the current state of the application, if it exists today, and any specific gaps that you know will need to be overcome.
+%
+%Biology is wrestling now with  about 3 orders of magnitude increase in DNA sequence data generation capability in the past 2 years.
+All algorithms, architectures, storage, and networks have broken or will break soon.
+Some biology problems (not necessarily the one I describe here) are not good fits for core-rich and memory-poor modern architectures.
+For some problems (comparing k-mers across metagenomic data sets, for instance) you need terabyte-scale memory, not thousands of cores.
 
@@
 
-This {\em infrastructural application} is aimed at supporting international
-CMIP/IPCC~\cite{intergovernmental2007fourth,cmip} intercomparison activity,
-which involves producing, storing, and analyzing data from multiple climate
-simulations.  It includes data generation, storage, and processing stages.
-
-\subsubsection*{Application description} %2. {\bf How does it do it?}
-
-%\katznote{from answers to questions 2,3,5,6 below, also address
-cyberinfrastructure (where does the app run) -- perhaps have a diagram here of
-what happens in the app}
-
-The overall system is comprised of three stages.  In the first stage, climate
-centers run a prescribed set of common experiments that produce 2-10 PB of data.
-Data can also come from sensors, in which case the center would still
-post-process the data before it would be published.  Centers can then publish
-their own output data, or send it to another center to publish.  The data is
-generated over a roughly 2-year period (then post-processed and published over a
-few more months).  This stage is characterized by being both distributed-compute
-and distributed-data intensive.
+This {\em infrastructural application} is aimed at supporting
+international CMIP/IPCC~\cite{intergovernmental2007fourth,cmip}
+intercomparison activity, which involves producing, storing, and
+analyzing data from multiple climate simulations.  It includes data
+generation, storage, and processing stages.
+
+\subsubsection*{Application description}
+%2. {\bf How does it do it?}
+
+%\katznote{from answers to questions 2,3,5,6 below, also address cyberinfrastructure (where does the app run) -- perhaps have a diagram here of what happens in the app}
+
+The overall system is comprised of three stages.  In the first stage,
+climate centers run a prescribed set of common experiments that
+produce 2-10 PB of data.  Data can also come from sensors, in which
+case the center would still post-process the data before it would be
+published.  Centers can then publish their own output data, or send it
+to another center to publish.  The data is generated over a roughly
+2-year period (then post-processed and published over a few more
+months).  This stage is characterized by being both
+distributed-compute and distributed-data intensive.
 
@@
 (ESGF)~\cite{esgf} develops and deploys a federated network of gateways and
-associated data nodes.  As models run at each center, the output data is
-post-processed into packages with common formats and prescribed metadata
-conventions.  Most centers deploy the ESGF data node software stack, and using
-this, they manage the data from their experiments.  The data node software stack
-scans the data, makes sure it has the right metadata fields, does QA/QC, and
-builds a set of catalogs of the prepared data.  Minimally, the catalog provides
-HTTP links to data elements, and it can also provide GridFTP endpoints, and/or
-product services that abstract the dataset in other ways: get a whole file,
-subset, browse, etc.  When the center is happy with the data/catalog in the data
-node, they publish it to a host/affiliated gateway.  This submits the catalog to
-the gateway.  The gateway then shares this catalog with other gateways so that
-all gateways have a consistent view of all the published data.  In general, most
-data is replicated at several sites.  The replication activity is manually
-requested/initiated by a gateway owner.  The properties of this stage are that
-it is distributed data-intensive (with distributed gateways and data nodes) and
-that it is dynamic (data appears in the system over time).
-
-The third stage is data analysis.  The approximately 20,000 users (as of
-Dec. 2010) can browse/search a catalog at any gateway and locate data, which
-might be hosted by a data node affiliated with another gateway. (This can output
-a `wget' script that can later fetch the data.)  They can also authenticate, and
-thus gain access to a group, which might have private data.  They can also
-download data via http, GridFTP, or access product services (the latter of which
-uses ESGF's data retrieval syntax---DRS---and which can be scripted.)  Some
-users will analyze data from a single model.  However, many applications are
-multi-model analyses, where many users want to look at the same parts of the
-output of some/all of the models, to understand if and how the models differ.
-Some centers will gather some/all of the core archive (plus more, perhaps) on
-local systems for local users to perform ``power analyses.''  Each user's data
-analyses are almost always done on a single system.  Because there is no
-distributed computing infrastructure, this stage is search, access, and transfer
-intensive.
-
- \subsubsection*{Big data aspects} %3. {\bf What are the big data aspects?}
+associated data nodes.  As models run at each center, the output data
+is post-processed into packages with common formats and prescribed
+metadata conventions.  Most centers deploy the ESGF data node software
+stack, and using this, they manage the data from their experiments.
+The data node software stack scans the data, makes sure it has the
+right metadata fields, does QA/QC, and builds a set of catalogs of the
+prepared data.  Minimally, the catalog provides HTTP links to data
+elements, and it can also provide GridFTP endpoints, and/or product
+services that abstract the dataset in other ways: get a whole file,
+subset, browse, etc.  When the center is happy with the data/catalog
+in the data node, they publish it to a host/affiliated gateway.  This
+submits the catalog to the gateway.  The gateway then shares this
+catalog with other gateways so that all gateways have a consistent
+view of all the published data.  In general, most data is replicated
+at several sites.  The replication activity is manually
+requested/initiated by a gateway owner.  The properties of this stage
+are that it is distributed data-intensive (with distributed gateways
+and data nodes) and that it is dynamic (data appears in the system
+over time).
+
+The third stage is data analysis.
+The approximately 20,000 users (as of Dec.
+2010) can browse/search a catalog at any gateway and locate data, which might be hosted by a data node affiliated with another gateway.
+(This can output a `wget' script that can later fetch the data.)  They can also authenticate, and thus gain access to a group, which might have private data.
+They can also download data via http, GridFTP, or access product services (the latter of which uses ESGF's data retrieval syntax---DRS---and which can be scripted.)
+Some users will analyze data from a single model.
+However, many applications are multi-model analyses, where many users want to look at the same parts of the output of some/all of the models, to understand if and how the models differ.
+Some centers will gather some/all of the core archive (plus more, perhaps) on local systems for local users to perform ``power analyses.''
+Each user's data analyses are almost always done on a single system.  Because there is no distributed computing infrastructure, this stage is search, access, and transfer intensive.
+
+ \subsubsection*{Big data aspects}
+%3. {\bf What are the big data aspects?}
 
@@
 
-The overall data generated and stored in the ESGF is 2-10 PB.  This includes a
-core archive, which is 1-2 PB in size and contains the most popular data.
-
- \subsubsection*{Distributed aspects} %4. {\bf What are the distributed
-aspects?}
+The overall data generated and stored in the ESGF is 2-10 PB.  This includes a  core archive, which is 1-2 PB in size and contains the most popular data.
+
+ \subsubsection*{Distributed aspects}
+ %4. {\bf What are the distributed aspects?}
 
@@
 
-The data is generated by a distributed set of climate centers.  It is stored in
-a distributed set of federated archives.  And it is used by a distributed set of
-users, who either run data analyses on a climate center with which they are
-associated, or they gather data from the ESGF to a local system for their
-analyses.
-
- \subsubsection*{Dynamic aspects} %5. {\bf What are the dynamic aspects?}
+The data is generated by a distributed set of climate centers.
+It is stored in a distributed set of federated archives.
+And it is used by a distributed set of users, who either run data analyses on a climate center with which they are associated, or they gather data from the ESGF to a local system for their analyses.
+
+ \subsubsection*{Dynamic aspects}
+%5. {\bf What are the dynamic aspects?}
 
@@
 
-Data is generated over time, so the data in the ESGF changes.  One might imagine
-a future version of ESGF where the launching of data analysis jobs is automated,
-in which case the location of those jobs would be dynamic, and the system might
-also be able to respond (or optimize for) various types of applications.
-
- \subsubsection*{Other important issues} %6. {\bf What else is important?}
-
-%\katznote{from answers to questions 7,8 below, also address programming systems
-(how was the app written; what tools does it use)}
-
-This application/infrastructure serves a large group of users, and involves a
-large amount of political and technical consensus to work.
+Data is generated over time, so the data in the ESGF changes.  One might imagine a future
+version of ESGF where the launching of data analysis jobs is automated, in which case  the
+location of those jobs would be dynamic, and the system might also be able to respond
+(or optimize for) various types of applications.
+
+ \subsubsection*{Other important issues}
+%6. {\bf What else is important?}
+
+%\katznote{from answers to questions 7,8 below, also address programming systems (how was the app written; what tools does it use)}
+
+This application/infrastructure serves a large group of users, and involves a large amount of political and technical consensus to work.
 
@@
 %
-%The Climate Data Analysis Tools (CDAT)~\cite{CDAT} are a scriptable %(Python)
-set of packages and tools that are currently heavily used for %analysis of
-climate data.  CDAT has been developed by LLNL's Program %for Climate Model
-Diagnosis and Intercomparison (PCMDI), and is often %used with data from the
-Earth System Grid (ESG)~\cite{ESG}.  The most %common current usage of these
-tools with the data is that a user %downloads a set of data from ESG to a local
-resource, then uses the %CDAT to analyze the data.  However, the PCMDI group is
-now planning to %support a new usage mode, where data transfer can be scripted.
-%Initial development of this capability is planned for February 2011, %using the
-Globus online data transfer service~\cite{globusOnline} as a %scriptable
-component.
+%The Climate Data Analysis Tools (CDAT)~\cite{CDAT} are a scriptable
+%(Python) set of packages and tools that are currently heavily used for
+%analysis of climate data.  CDAT has been developed by LLNL's Program
+%for Climate Model Diagnosis and Intercomparison (PCMDI), and is often
+%used with data from the Earth System Grid (ESG)~\cite{ESG}.  The most
+%common current usage of these tools with the data is that a user
+%downloads a set of data from ESG to a local resource, then uses the
+%CDAT to analyze the data.  However, the PCMDI group is now planning to
+%support a new usage mode, where data transfer can be scripted.
+%Initial development of this capability is planned for February 2011,
+%using the Globus online data transfer service~\cite{globusOnline} as a
+%scriptable component.
 %
@@
 %
-%Aimed at supporting international CMIP/IPCC intercomparison activity.  %Three
-stages:
+%Aimed at supporting international CMIP/IPCC intercomparison activity.
+%Three stages:
 %
@@
 %\begin{itemize}
-%\item Climate centers run prescribed set of common experiments; produce 2-10 PB
-of data
+%\item Climate centers run prescribed set of common experiments; produce 2-10 PB of data
 %\begin{itemize}
-%\item data can also come from sensors - here, the data would still be
-post-processed before it would be published.
+%\item data can also come from sensors - here, the data would still be post-processed before it would be published.
 %\end{itemize}
-%\item Centers can publish their own output data or send it to another center to
-publish data generated over \~{}2 year period (then post-processed and published
-over a few more months)
+%\item Centers can publish their own output data or send it to another center to publish data generated over \~{}2 year period (then post-processed and published over a few more months)
 %\item Properties: data-intensive, distributed centers, lots of computing
@@
 %\begin{itemize}
-%\item ESGF develops and deploys a federated network of gateways and associated
-data nodes
-%\item As models run at each center, output data is post-processed into packages
-with common formats and prescribed metadata conventions
-%\item Most centers will deploy data node software stack, and using this, they
-manage the data from their experiments
+%\item ESGF develops and deploys a federated network of gateways and associated data nodes
+%\item As models run at each center, output data is post-processed into packages with common formats and prescribed metadata conventions
+%\item Most centers will deploy data node software stack, and using this, they manage the data from their experiments
 %  \begin{itemize}
-% \item the data node software stack scans the data, makes sure it has the right
-metadata fields, does QA/QC, build a set of catalogs of the prepared data
+%  \item the data node software stack scans the data, makes sure it has the right metadata fields, does QA/QC, build a set of catalogs of the prepared data
 %  \begin{itemize}
@@
 %                \item also can provide GridFTP endpoints
-% \item also can provide product services that abstract the dataset in other
-ways - get whole file, subset, browse, etc.
+%                \item also can provide product services that abstract the dataset in other ways - get whole file, subset, browse, etc.
 %  \end{itemize}
 %  \end{itemize}
-% \item When the center is happy with the data/catalog in the data node, they
-publish it to a host/affiliated gateway
+%    \item When the center is happy with the data/catalog in the data node, they publish it to a host/affiliated gateway
 %\begin{itemize}
 %         \item  this submits the catalog to the gateway
-% \item them the gateway shares this catalog with other gateways so that all
-gateways have a consistent view of all the published data
+%          \item  them the gateway shares this catalog with other gateways so that all gateways have a consistent view of all the published data
 %\end{itemize}
@@
 %         \item  replication activity manually requested/initiated by a gateway owner
-% \item replication: copies catalog to replica site, do data transfer to new
-data node, then publish catalog, push catalog to (this) gateway associated with
-new data node
+%          \item  replication: copies catalog to replica site, do data transfer to new data node, then publish  catalog, push catalog to (this) gateway associated with new data node
 %\end{itemize}
@@
 %\begin{itemize}
-% \item track which users have downloaded data - if a problem is found in a data
-set or if new data is added to a data set, notify users who have
-downloaded/replicated it
+%          \item  track which users have downloaded data - if a problem is found in a data set or if new data is added to a data set, notify users who have downloaded/replicated it
 %\end{itemize}
-% \item Properties: data-intensive, distributed gateways and data nodes, data
-appears in the system over time (dynamic)
+%    \item  Properties: data-intensive, distributed gateways and data nodes, data appears in the system over time (dynamic)
 %\end{itemize}
@@
 %\begin{itemize}
-% \item browse/search a catalog at any gateway and locate data, which might be
-hosted by a data node affiliated with another gateway. (can output a wget script
-that can later fetch the data)
+%         \item  browse/search a catalog at any gateway and locate data, which might be hosted by a data node affiliated with another gateway. (can output a wget script that can later fetch the data)
 %         \item  authenticate, gain access to a group
-% \item download data via http or GridFTP or access product services (uses data
-retrieval syntax (DRS) - can be scripted)
+%        \item  download data via http or GridFTP or access product services (uses data retrieval syntax (DRS) - can be scripted)
 %  \end{itemize}
@@
 %\begin{itemize}
-% \item abstractions that can be used by a programmer: data is stored in NetCDF
-format, uses CF metadata conventions, can be accessed using DRS
+%    \item  abstractions that can be used by a programmer: data is stored in NetCDF format, uses CF metadata conventions, can be accessed using DRS
 %    \item  Some users will analyze data from a single model
-% \item Many applications are multi-model analyses - many users want to look at
-the same parts of the output of some/all of the models, to understand if and how
-the models differ
-% \item Some centers will gather some/all of the core archive (plus more,
-perhaps) on local systems for local users to perform ``power analysis''
+%    \item  Many applications are multi-model analyses - many users want to look at the same parts of the output of some/all of the models, to understand if and how the models differ
+%    \item  Some centers will gather some/all of the core archive (plus more, perhaps) on local systems for local users to perform ``power analysis''
 %\end{itemize}
@@
 
-In the environmental sciences, the use of visualization techniques is vital for
-understanding the ever-increasing volume and diversity of data that is being
-produced by Earth observing systems and computer simulations.  The primary
-purpose of visualization is to gain insight but the majority of scientific tools
-generate static plots that neither the originating scientist nor the recipient
-can easily customize to reveal new information. The issues with visualization of
-a single dataset are compounded if multiple datasets are to be examined
-simultaneously, as is common in environmental science for model validation,
-``ground truthing,'' quality control, and data assimilation.
-
-New, interactive modes of environmental data exploration and visualization based
-on the principles of simplicity, open standards and user-friendliness are
-required at all stages of scientific investigation.  The application described
-here is an {\em infrastructural application} that uses stored data and consists
-of one stage: data processing.  It represents a number of similar
+In the environmental sciences, the use of visualization techniques is
+vital for understanding the ever-increasing volume and diversity of
+data that is being produced by Earth observing systems and computer
+simulations.  The primary purpose of visualization is to gain insight
+but the majority of scientific tools generate static plots that
+neither the originating scientist nor the recipient can easily
+customize to reveal new information. The issues with visualization of
+a single dataset are compounded if multiple datasets are to be
+examined simultaneously, as is common in environmental science for
+model validation, ``ground truthing,'' quality control, and data
+assimilation.
+
+New, interactive modes of environmental data exploration and
+visualization based on the principles of simplicity, open standards
+and user-friendliness are required at all stages of scientific
+investigation.  The application described here is an {\em
+  infrastructural application} that uses stored data and consists of
+one stage: data processing.  It represents a number of similar
 applications~\cite{blower1,blower2}.
 
-\subsubsection*{Application description} %2. {\bf How does it do it?}
-
-%\katznote{from answers to questions 2,3,5,6 below, also address
-cyberinfrastructure (where does the app run) -- perhaps have a diagram here of
-what happens in the app}
-
-The applications often take the form of graphical Geographic Information Systems
-(GIS) tools, but with better support for large and multidimensional scientific
-data. The general vision is of a map-based interface, onto which different
-datasets can be overlain.  A processing step may be initiated by rubber-banding
-an area of the map and selecting from a list of algorithms that process data
-within the selected area, perhaps calculating statistics of the data.  In a
-desktop application, this processing may take place on the user's desktop, but
-in a web-based application the processing must take place on a server, which may
-or may not be co-located with the datasets.  There is therefore the common
-problem of moving large amounts of data around in a distributed system, to which
-may be added concerns of security in cases in which the data in question are not
-public.
-
- \subsubsection*{Big data aspects} %3. {\bf What are the big data aspects?}
+\subsubsection*{Application description}
+%2. {\bf How does it do it?}
+
+%\katznote{from answers to questions 2,3,5,6 below, also address cyberinfrastructure (where does the app run) -- perhaps have a diagram here of what happens in the app}
+
+The applications often take the form of graphical Geographic
+Information Systems (GIS) tools, but with better support for large and multidimensional scientific data. The general vision is of a map-based interface, onto which
+different datasets can be overlain.  A processing step may be initiated by rubber-banding an area of
+the map and selecting from a list of algorithms that process data within the selected area, perhaps
+calculating statistics of the data.  In a desktop application, this processing may take place on the user's desktop,
+but in a web-based application the processing must take place on a server, which may or may not be
+co-located with the datasets.  There is therefore the common problem of moving large amounts of data around in a distributed
+system, to which may be added concerns of security in cases in which the data in question are not public.
+
+ \subsubsection*{Big data aspects}
+%3. {\bf What are the big data aspects?}
 
@@
 
-Data from instruments is small (1--100~MB) but model output may be very big
-(GB--TB). The data is multi-dimensional in nature. Data aggregated over time
-from instruments may need to be visualized (e.g. as an animation), causing the
-visualized data size to grow.
-
- \subsubsection*{Distributed aspects} %4. {\bf What are the distributed
-aspects?}
+Data from instruments is small (1--100~MB) but model output may be very big (GB--TB). The data is
+multi-dimensional in nature. Data aggregated over time from instruments may need to be visualized
+(e.g. as an animation), causing the visualized data size to grow.
+
+ \subsubsection*{Distributed aspects}
+ %4. {\bf What are the distributed aspects?}
 
@@
 
- Increasingly, environmental data are distributed in multiple locations.  The
-applications are driven by data served through distributed web services, that
-may in turn be ``fed'' instruments or computer simulation generated
-data. Compute services need to be integrated with these data services in an
-efficient, easy-to-use, and transparent way, allowing fast data processing.  The
-processing algorithms are generally simple and high overhead scheduling systems
-can cause a loss of responsiveness.  The infrastructure has more in common with
-the Web (or cloud) than the grid.
-
- \subsubsection*{Dynamic aspects} %5. {\bf What are the dynamic aspects?}
+ Increasingly, environmental data are distributed in multiple
+ locations.  The applications are driven by data served through
+ distributed web services, that may in turn be ``fed'' instruments or
+ computer simulation generated data. Compute services need to be
+ integrated with these data services in an efficient, easy-to-use, and
+ transparent way, allowing fast data processing.  The processing
+ algorithms are generally simple and high overhead scheduling systems
+ can cause a loss of responsiveness.  The infrastructure has more in
+ common with the Web (or cloud) than the grid.
+
+ \subsubsection*{Dynamic aspects}
+%5. {\bf What are the dynamic aspects?}
 
@@
 instruments or from looking at live results from a model. Data hosted at
-multiple locations may be frequently updated (often several times a day).  The
-data services are designed to allow server-side subsetting of large datasets,
-attempting to minimize the amount of unwanted data traveling across networks.
-These make it hard to achieve scalability through caching.
-
-In a Web environment, servers and proxy servers (which are sometimes beyond the
-control of the data provider or the data user) may retain caches of data in a
-strategy to reduce server load and network traffic.  For this reason, the HTTP
-protocol provides mechanisms for defining expiration times for data resources,
-specifying when a resource ought to be cleared from the cache.  In the
-environmental and geospatial communities the Open Geospatial Consortium
-specifications are being widely adopted alongside existing protocols such as
-OPeNDAP~\cite{opendap}.  Unfortunately these protocols, although built atop
-HTTP, do not make it easy for this versioning to be implemented correctly.  The
-OPeNDAP protocol does not have the concept of a dataset version, meaning that
-clients have no reliable or efficient means to detect that a dataset has
-changed.  The OGC protocols, in general, \emph{do} provide versioning at the
-level of the service endpoint, but provide no information on when a resource
-should be considered expired.  Clients are therefore forced to poll the server
-to check for updates.
-
- \subsubsection*{Other important issues} %6. {\bf What else is important?}
-
-%\katznote{from answers to questions 7,8 below, also address programming systems
-(how was the app written; what tools does it use)}
-
-The application aims for near-real-time interaction with the data, i.e the user
-should not wait more than a few seconds for some kind of response to a request,
-so performance with low latency is a key challenge.  Security is another serious
-concern: many environmental datasets are held under access control, and
-different providers often have very different access control policies.  In some
-infrastructures, these different policies are handled via a role-mapping
-mechanism~\cite{nerc_data}.  Concerns of access control---as well as data
-volume---also make it difficult for data to be replicated to different
-geographic locations (and hence different access control regimes).  In a
-distributed environment, it is common for machines to access data on behalf of
-end users (e.g., a processing service might download its input data from a
-remote store).  Finding a means for the user to delegate his/her authority to a
-multi-web-service infrastructure is a key current challenge.  The MashMyData
-project~\cite{mashmydata} is investigating two solutions, based on Grid proxy
-certificates and OAuth.
+multiple locations may be frequently updated (often several times a day).
+The data services are designed to allow server-side subsetting of large
+datasets, attempting to minimize the amount of unwanted data traveling
+across networks.  These make it hard to achieve scalability through caching.
+
+In a Web environment, servers and proxy servers (which are sometimes beyond
+the control of the data provider or the data user) may retain caches
+of data in a strategy to reduce server load and network traffic.  For
+this reason, the HTTP protocol provides mechanisms for defining expiration
+times for data resources, specifying when a resource ought to be
+cleared from the cache.  In the environmental and geospatial communities the Open Geospatial
+Consortium specifications are being widely adopted alongside existing
+protocols such as OPeNDAP~\cite{opendap}.  Unfortunately these protocols, although
+built atop HTTP, do not make it easy for this versioning to be
+implemented correctly.  The OPeNDAP protocol does not have the concept
+of a dataset version, meaning that clients have no reliable or
+efficient means to detect that a dataset has changed.  The OGC
+protocols, in general, \emph{do} provide
+versioning at the level of the service endpoint, but provide no
+information on when a resource should be considered expired.  Clients
+are therefore forced to poll the server to check for updates.
+
+ \subsubsection*{Other important issues}
+%6. {\bf What else is important?}
+
+%\katznote{from answers to questions 7,8 below, also address programming systems (how was the app written; what tools does it use)}
+
+The application aims for near-real-time interaction with the data, i.e the user should not wait more than a few seconds for some kind of response to a request, so performance with low latency is a key challenge.
+Security is another serious concern: many environmental datasets
+are held under access control, and different providers often have very different access control policies.
+In some infrastructures, these different policies are handled via a role-mapping mechanism~\cite{nerc_data}.
+Concerns of access control---as well as data volume---also make it difficult for data to be replicated
+to different geographic locations (and hence different access control regimes).
+In a distributed environment, it is common for machines to access data
+on behalf of end users (e.g., a processing service might download its input data from a remote store).
+Finding a means for the user to delegate his/her authority to a multi-web-service infrastructure is
+a key current challenge.  The MashMyData project~\cite{mashmydata} is investigating two solutions, based on Grid proxy certificates and OAuth.
 
@@
 
-% (Here's a synopsis of the chapter I'm writing for Malcolm's book on
-Data-Intensive Research.  % I'm not sure how to merge this with the answers to
-the pro-forma questions below.  Also, % many of these sentences are
-cut-and-pasted from the book chapter - future copyright issue?)
-
-% In the environmental sciences the use of visualization techniques is vital for
-understanding % the ever-increasing volume and diversity of data that is being
-produced by Earth observing % systems and computer simulations.  The primary
-purpose of visualization is to gain insight % \cite{card_readings_1999}, but the
-majority of current tools in wide use are not adequate % for achieving this.
-The time spent to create visualizations is too great, usually due to the %
-crippling overhead of low-level technical tasks such as file-format and metadata
-interpretation.  % Most scientific tools generate static plots, which are shared
-in documents or emails, but % neither the originating scientist nor the
-recipient can easily customize or adapt the visualization % to reveal new
-information.  The context of how the visualization was generated is usually lost
-% in the publication/sharing process, making interpretation and reproduction
-much more difficult.
-
-% Visualization of a single dataset is therefore more tedious and less useful
-than it should be.  % These problems are compounded if the scientist wishes to
-examine multiple datasets simultaneously.  % This is an extremely common task in
-environmental science, which forms an important part of model validation, %
-``ground truthing'', quality control and data assimilation.
-
-% There is much current interest in developing new, interactive modes of
-environmental data exploration and visualization, which can % play a key role at
-all stages of scientific investigation.  We consider that, for the case of
-interactive % data exploration, concerns of performance, % intuitiveness and
-user engagement tend to outweigh the desire for sophistication in data
-processing.  % We therefore argue that a new generation of ``4D GIS'' tools,
-based on the principles of simplicity, % open standards and user-friendliness is
-required to gain greater insight from the environmental data % deluge.  These
-new tools will not replace existing approaches, but rather complement them.
-
-% Increasingly, environmental data are distributed in multiple locations, and
-are frequently updated % (often several times a day).  Interactive exploration
-of such large, dynamic, distributed data is therefore % a strong technical
-challenge.  Data infrastructures must support the ability to quickly extract %
-subsets of large data, and any data processing must happen quickly, with the
-minimum of overhead.  % Web services are therefore playing an
-increasingly-important role, and there is much interest in % cloud technologies
-to provide the necessary scalability at the back end.  Grid technologies, being
-essentially batch-mode % systems, are less suitable for this kind of application
-as the scheduling overhead tends to dominate % the processing time for simple
-jobs, seriously hampering interactivity.
-
-% Performance is therefore a key challenge, exacerbated by the dynamic nature of
-the data, which makes % it hard to achieve scalability through caching (see
-section~\ref{sec:dynamicdataweb}).  [Note: could probably % merge material from
-that section here.]  Security is another serious concern: many environmental
-datasets % are held under access control.  In a distributed environment, it is
-common for machines to access data % on behalf of end users (e.g., a processing
-service might download its input data from a remote store).  % Finding a means
-for the user to delegate his/her authority to a multi-web-service infrastructure
-is % a key current challenge.  The MashMyData project
-(http://www.mashmydata.org) is investigating two solutions, based % respectively
-on Grid proxy certificates and OAuth.
+% (Here's a synopsis of the chapter I'm writing for Malcolm's book on Data-Intensive Research.
+% I'm not sure how to merge this with the answers to the pro-forma questions below.  Also,
+% many of these sentences are cut-and-pasted from the book chapter - future copyright issue?)
+
+% In the environmental sciences the use of visualization techniques is vital for understanding
+% the ever-increasing volume and diversity of data that is being produced by Earth observing
+% systems and computer simulations.  The primary purpose of visualization is to gain insight
+% \cite{card_readings_1999}, but the majority of current tools in wide use are not adequate
+% for achieving this.  The time spent to create visualizations is too great, usually due to the
+% crippling overhead of low-level technical tasks such as file-format and metadata interpretation.
+% Most scientific tools generate static plots, which are shared in documents or emails, but
+% neither the originating scientist nor the recipient can easily customize or adapt the visualization
+% to reveal new information.  The context of how the visualization was generated is usually lost
+% in the publication/sharing process, making interpretation and reproduction much more difficult.
+
+% Visualization of a single dataset is therefore more tedious and less useful than it should be.
+% These problems are compounded if the scientist wishes to examine multiple datasets simultaneously.
+% This is an extremely common task in environmental science, which forms an important part of model validation,
+% ``ground truthing'', quality control and data assimilation.
+
+% There is much current interest in developing new, interactive modes of environmental data exploration and visualization, which can
+% play a key role at all stages of scientific investigation.   We consider that, for the case of interactive
+% data exploration, concerns of performance,
+% intuitiveness and user engagement tend to outweigh the desire for sophistication in data processing.
+% We therefore argue that a new generation of ``4D GIS'' tools, based on the principles of simplicity,
+% open standards and user-friendliness is required to gain greater insight from the environmental data
+% deluge.  These new tools will not replace existing approaches, but rather complement them.
+
+% Increasingly, environmental data are distributed in multiple locations, and are frequently updated
+% (often several times a day).  Interactive exploration of such large, dynamic, distributed data is therefore
+% a strong technical challenge.  Data infrastructures must support the ability to quickly extract
+% subsets of large data, and any data processing must happen quickly, with the minimum of overhead.
+% Web services are therefore playing an increasingly-important role, and there is much interest in
+% cloud technologies to provide the necessary scalability at the back end.  Grid technologies, being essentially batch-mode
+% systems, are less suitable for this kind of application as the scheduling overhead tends to dominate
+% the processing time for simple jobs, seriously hampering interactivity.
+
+% Performance is therefore a key challenge, exacerbated by the dynamic nature of the data, which makes
+% it hard to achieve scalability through caching (see section~\ref{sec:dynamicdataweb}).  [Note: could probably
+% merge material from that section here.]  Security is another serious concern: many environmental datasets
+% are held under access control.  In a distributed environment, it is common for machines to access data
+% on behalf of end users (e.g., a processing service might download its input data from a remote store).
+% Finding a means for the user to delegate his/her authority to a multi-web-service infrastructure is
+% a key current challenge.  The MashMyData project (http://www.mashmydata.org) is investigating two solutions, based
+% respectively on Grid proxy certificates and OAuth.
 
@@
 
-% We are actually concerned with a class of applications (highly % interactive,
-graphical ones) rather than a specific one.  But % generally, the applications
-involve the intercomparison of different % environmental datasets for purposes
-such as model validation, ground % trothing of observations and data
-assimilation.
+% We are actually concerned with a class of applications (highly
+% interactive, graphical ones) rather than a specific one.  But
+% generally, the applications involve the intercomparison of different
+% environmental datasets for purposes such as model validation, ground
+% trothing of observations and data assimilation.
 
@@
 
-% The applications often take the form of graphical tools that share a lot in
-common with Geographic Information Systems (GIS). However these applications are
-much better suited to scientific data, which are large and multidimensional.
-The general vision is of a map-based interface, onto which different datasets
-can be overlain.  A processing step may be initiated by rubber-banding an area
-of the map and selecting from a list of algorithms that process data within the
-selected area, perhaps calculating statistics of the data.
-
-% 3.  What infrastructure is used? (including compute, data, network,
-instruments, etc.)
-
-% Usually the applications are driven by data served through web services of
-various kinds.  These web services may be 'fed' by data coming from instruments
-or from computer simulations.  A key challenge is how to integrate compute
-services with these data services in an efficient, easy-to-use and transparent
-way, allowing fast data processing.  The processing algorithms are generally
-quite simple and so it is frustrating to succumb to the high overhead of some
-scheduling systems.  The infrastructure probably has more in common with the Web
-(or perhaps cloud) than the Grid.
-
-% 4. What dynamic data is used in the application?  % (a) What are the types of
-data, (b) What is the size of the data set(s)?
-
-% Dynamic data are only sometimes used, but may arise from real-time feeds from
-instruments or from looking at live results from a model.  Data from instruments
-are small 1~MB-100~MB but model output may be very big (GB-TB).
+% The applications often take the form of graphical tools that share a lot in common with Geographic Information Systems (GIS).
+However these applications are much better suited to scientific data, which are large and multidimensional.
+The general vision is of a map-based interface, onto which different datasets can be overlain.
+A processing step may be initiated by rubber-banding an area of the map and selecting from a list of algorithms that process data within the selected area, perhaps calculating statistics of the data.
+
+% 3.     What infrastructure is used? (including compute, data, network, instruments, etc.)
+
+% Usually the applications are driven by data served through web services of various kinds.
+These web services may be 'fed' by data coming from instruments or from computer simulations.
+A key challenge is how to integrate compute services with these data services in an efficient, easy-to-use and transparent way, allowing fast data processing.
+The processing algorithms are generally quite simple and so it is frustrating to succumb to the high overhead of some scheduling systems.
+The infrastructure probably has more in common with the Web (or perhaps cloud) than the Grid.
+
+%      4. What dynamic data is used in the application?
+% (a) What are the types of data, (b) What is the size of the data set(s)?
+
+% Dynamic data are only sometimes used, but may arise from real-time feeds from instruments or from looking at live results from a model.
+Data from instruments are small 1~MB-100~MB but model output may be very big (GB-TB).
 
@@
 
-% Generally through web services.  These services are designed to allow
-server-side subsetting of large datasets, attempting to minimize the amount of
-unwanted data traveling across networks.
+% Generally through web services.  These services are designed to allow server-side subsetting of large datasets, attempting to minimize the amount of unwanted data traveling across networks.
 
@@
 
-% We try to aim for near-real time interaction with the data, i.e the user
-should not be waiting more than a few seconds for some kind of response to a
-request.  We have largely solved this problem for simple visualization, but for
-linking data access and processing we are not there yet.
+% We try to aim for near-real time interaction with the data, i.e the user should not be waiting more than a few seconds for some kind of response to a request.
+We have largely solved this problem for simple visualization, but for linking data access and processing we are not there yet.
 
@@
 
-% Key contents now merged into above section.  Specifics on the NERC VO % are
-probably not relevant at the moment as they are not yet implemented.
-
-% \jhanote{I have moved this from erstwhile \S 2. Jon: please either (i) %
-convert into a stand-alone application, or (ii) merge with the % ``Interactive
-Exploration of Environmental Data'' example}
+% Key contents now merged into above section.  Specifics on the NERC VO
+% are probably not relevant at the moment as they are not yet implemented.
+
+% \jhanote{I have moved this from erstwhile \S 2. Jon: please either (i)
+  % convert into a stand-alone application, or (ii) merge with the
+  % ``Interactive Exploration of Environmental Data'' example}
 
@@
 
-% The NERC\footnote{UK Natural Environment Research Council} Virtual %
-Observatory\footnote{http://www.nerc.ac.uk/research/programmes/virtualobservatory/}
-% is a new initiative to create a pilot cloud-based environment in which %
-different practitioners will discover, visualize and process data % relating to
-the UK's system of rivers and soils.  It will provide % access to many different
-kinds of distributed hydrological data, % including sensor data and numerical
-models, many of which produce % rapidly-changing, dynamic information.
+% The NERC\footnote{UK Natural Environment Research Council} Virtual
+% Observatory\footnote{http://www.nerc.ac.uk/research/programmes/virtualobservatory/}
+% is a new initiative to create a pilot cloud-based environment in which
+% different practitioners will discover, visualize and process data
+% relating to the UK's system of rivers and soils.  It will provide
+% access to many different kinds of distributed hydrological data,
+% including sensor data and numerical models, many of which produce
+% rapidly-changing, dynamic information.
 
@@
 
-% There are many scientific % challenges to be faced, notably including the
-coupling of the models % and observations; this includes the possibility that
-sensors can be % dynamically tasked in response to model forecasts, or that
-models may % adapt dynamically to new information from sensors.
+% There are many scientific
+% challenges to be faced, notably including the coupling of the models
+% and observations; this includes the possibility that sensors can be
+% dynamically tasked in response to model forecasts, or that models may
+% adapt dynamically to new information from sensors.
 
@@
 
-% Hydrological data, % including sensor data and numerical models, are
-distributed.  % Principles of the % Web and ``linked data'' may help to solve
-some of the challenges in % informatics, and % Web Services are commonly used to
-distribute data through HTTP.
+% Hydrological data,
+% including sensor data and numerical models, are distributed.
+% Principles of the
+% Web and ``linked data'' may help to solve some of the challenges in
+% informatics, and
+% Web Services are commonly used to distribute data through HTTP.
 
@@
 
-% Sensor data and numerical models produce % rapidly-changing, dynamic
-information.  % A key characteristic of dynamic data, of course, is that it
-changes with % time.  It is of great importance for the user to be confident
-that she % is accessing the correct version of the data; this will often be the
-% most up-to-date data, or may be a specific historical version.  In a % Web
-environment, servers and proxy servers (which are sometimes beyond % the control
-of the data provider or the data user) may retain caches % of data in a strategy
-to reduce server load and network traffic.  For % this reason, the HTTP protocol
-provides mechanisms for defining expiration % times for data resources,
-specifying when a resource ought to be % cleared from the cache.
-
-% In the environmental and geospatial communities the Open Geospatial %
-Consortium specifications are being widely adopted alongside existing %
-protocols such as OPeNDAP.  Unfortunately these protocols, although % built atop
-HTTP, do not make it easy for this versioning to be % implemented correctly.
-The OPeNDAP protocol does not have the concept % of a dataset version, meaning
-that clients have no reliable or % efficient means to detect that a dataset has
-changed.  The OGC % protocols in general %[TODO: look into this], % \emph{do}
-provide % versioning at the level of the service endpoint, but provide no %
-information on when a resource should be considered expired.  Clients % are
-therefore forced to poll the server to check for updates.
-
-% Many technical challenges will remain regarding the % handling of dynamic
-data, including keeping careful records of its % provenance in order to record
-the steps taken in making a decision.
+% Sensor data and numerical models produce
+% rapidly-changing, dynamic information.
+% A key characteristic of dynamic data, of course, is that it changes with
+% time.  It is of great importance for the user to be confident that she
+% is accessing the correct version of the data; this will often be the
+% most up-to-date data, or may be a specific historical version.  In a
+% Web environment, servers and proxy servers (which are sometimes beyond
+% the control of the data provider or the data user) may retain caches
+% of data in a strategy to reduce server load and network traffic.  For
+% this reason, the HTTP protocol provides mechanisms for defining expiration
+% times for data resources, specifying when a resource ought to be
+% cleared from the cache.
+
+% In the environmental and geospatial communities the Open Geospatial
+% Consortium specifications are being widely adopted alongside existing
+% protocols such as OPeNDAP.  Unfortunately these protocols, although
+% built atop HTTP, do not make it easy for this versioning to be
+% implemented correctly.  The OPeNDAP protocol does not have the concept
+% of a dataset version, meaning that clients have no reliable or
+% efficient means to detect that a dataset has changed.  The OGC
+% protocols in general %[TODO: look into this],
+% \emph{do} provide
+% versioning at the level of the service endpoint, but provide no
+% information on when a resource should be considered expired.  Clients
+% are therefore forced to poll the server to check for updates.
+
+% Many technical challenges will remain regarding the
+% handling of dynamic data, including keeping careful records of its
+% provenance in order to record the steps taken in making a decision.
 
@@
 
-Demand response (DR) optimization~\cite{ferc2011DR} in smart power grids deals
-with ensuring an adequate supply of electricity by targeted curtailment of power
-load by consumers during periods of peak load to avoid blackouts. One
-application used for DR optimization is power demand forecasting at coarse and
-fine temporal and spatial granularities to allow load curtailment operations to
-be triggered in advance of a demand-supply mismatch situation.  This is a {\em
-infrastructural application}.  It begins after data generation, and includes the
-data transformation \& decisions, storage, and processing stages.
-
-\subsubsection*{Application description} %2. {\bf How does it do it?}
-
-%\katznote{from answers to questions 2,3,5,6 below, also address
-cyberinfrastructure (where does the app run) -- perhaps have a diagram here of
-what happens in the app}
-
-Power usage information is aggregated from smart meters at consumer premises
-into the utility's data center~\cite{Simmhan:buildsys:2011,simmhan2011TR}. Data
-is collected at approximately 15-min intervals and transmitted to the utility
-through cellular and wireless networks using proprietary protocols. Complex
-event pattern matching over the power usage events and other information source
-such as weather and scheduling data accessed using web services help perform
-near term load forecasting. Long term power forecasting at the city-scale or at
-a smaller micro-grid scale (e.g. industry complex, university campus) is done
-using power usage models built using machine learning algorithms running at the
-utility's or micro-grid's datacenter using historical usage
-data~\cite{Aman:dddm:2011}. These applications run on private clouds and are
-composed using DAG, Complex Event Processing and MapReduce
-structures~\cite{zhou:2012:scepter}.
-
- \subsubsection*{Big data aspects} %3. {\bf What are the big data aspects?}
+Demand response (DR) optimization~\cite{ferc2011DR} in smart power
+grids deals with ensuring an adequate supply of electricity by
+targeted curtailment of power load by consumers during periods of peak
+load to avoid blackouts. One application used for DR optimization is
+power demand forecasting at coarse and fine temporal and spatial
+granularities to allow load curtailment operations to be triggered in
+advance of a demand-supply mismatch situation.  This is a {\em
+  infrastructural application}.  It begins after data generation, and
+includes the data transformation \& decisions, storage, and processing
+stages.
+
+\subsubsection*{Application description}
+%2. {\bf How does it do it?}
+
+%\katznote{from answers to questions 2,3,5,6 below, also address cyberinfrastructure (where does the app run) -- perhaps have a diagram here of what happens in the app}
+
+Power usage information is aggregated from smart meters at consumer premises into the utility's data center~\cite{Simmhan:buildsys:2011,simmhan2011TR}.
+Data is collected at approximately 15-min intervals and transmitted to the utility through cellular and wireless networks using proprietary protocols.
+Complex event pattern matching over the power usage events and other information source such as weather and scheduling data accessed using web services help perform near term load forecasting.
+Long term power forecasting at the city-scale or at a smaller micro-grid scale (e.g.
+industry complex, university campus) is done using power usage models built using machine learning algorithms running at the utility's or micro-grid's datacenter using historical usage data~\cite{Aman:dddm:2011}.
+These applications run on private clouds and are composed using DAG, Complex Event Processing and MapReduce structures~\cite{zhou:2012:scepter}.
+
+ \subsubsection*{Big data aspects}
+%3. {\bf What are the big data aspects?}
 
@@
 
-Constructing forecast models using machine learning requires access to
-historical data that can grow to large sizes~\cite{Yin:mapreduce:2012}. Several
-hundred TB per year of smart meter power usage data can accumulate for a large
-city with millions of customers. Other data used in predictive modeling, such as
-event/people schedules and building/facility features, are smaller, on the order
-of GBs per year. In future, social network data from consumers may also be mined
-to learn about power usage patterns. These can also be on the order of GBs.
-
-Real-time forecasting uses smaller data---events on the order of 1~KB, but
-potentially from millions of sources---that are passed to complex event
-pipelines or predictive models that run continuously. These pipelines can
-process GBs of data per hour.
-
- \subsubsection*{Distributed aspects} %4. {\bf What are the distributed
-aspects?}
+Constructing forecast models using machine learning requires access to historical data that can grow to large sizes~\cite{Yin:mapreduce:2012}.
+Several hundred TB per year of smart meter power usage data  can accumulate for a large city with millions of customers.
+Other data used in predictive modeling, such as event/people schedules and building/facility features, are smaller, on the order of GBs per year.
+In future, social network data from consumers may also be mined to learn about power usage patterns.
+These can also be on the order of GBs.
+
+Real-time forecasting uses smaller data---events on the order of 1~KB, but potentially from millions of sources---that are passed to complex event pipelines or predictive models that run continuously.
+These pipelines can process GBs of data per hour.
+
+ \subsubsection*{Distributed aspects}
+ %4. {\bf What are the distributed aspects?}
 
@@
 
-Data from domains such as power systems, building management systems, weather
-and traffic, and social networks need to be integrated. While the data are
-generated and/or stored in distributed locations, they are typically aggregated
-at the utility's datacenter for analysis. Limited processing may happen locally
-at the smart meters or local micro-grids with the results pushed as events to
-the utility.
-
-The applications themselves run on private cloud platforms across hundreds of
-VMs. Communication between the utility and smart meters is bidirectional,
-allowing control and pricing signals to be sent from the datacenter to the smart
-meters. Integrated data will also need to be shared with external applications
-running on, say, mobile platforms or portals, and with regulatory agencies.
-
- \subsubsection*{Dynamic aspects} %5. {\bf What are the dynamic aspects?}
+Data from domains such as power systems, building management systems, weather and
+traffic, and social networks need to be integrated. While the
+data are generated and/or stored in distributed locations, they are typically aggregated at the utility's datacenter for analysis.
+Limited processing may happen locally at the smart meters or local micro-grids with the results pushed as events to the utility.
+
+The applications themselves run on private cloud platforms across hundreds of VMs.
+Communication between the utility and smart meters is bidirectional, allowing control and pricing signals to be sent from the datacenter to the smart meters.
+Integrated data will also need to be shared with external applications running on, say, mobile platforms or portals, and with regulatory agencies.
+
+ \subsubsection*{Dynamic aspects}
+%5. {\bf What are the dynamic aspects?}
 
@@
 
-The data rates for events published by the sensors may change, either passively
-due to local measurement limitations or actively throttled by the utility to
-meet the application's accuracy needs/resource
-constraints~\cite{Simmhan:sciencecloud:2011}. The data from smart meters can be
-sent synchronously as they are generated or in batch mode once a day to conserve
-bandwidth. The application processing itself is done elastic resources, scaling
-up or down to meet latency requirements. Some of the forecasting models can also
-decide to listen to additional data streams on-demand when higher accuracy is
-required or special situations are encountered.
-
- \subsubsection*{Other important issues} %6. {\bf What else is important?}
-
-%\katznote{from answers to questions 7,8 below, also address programming systems
-(how was the app written; what tools does it use)}
-
-Semantic information integration is being performed using an ontology composed
-together from individual domains~\cite{Zhou:itng:2012}. Machine learning models
-using Matlab, Weka and Hadoop are being used. Complex event processing is using
-the ``Siddhi'' CEP engine with semantic support added~\cite{siddhi}. Stream
-processing on Eucalyptus using IBM Infosphere Streams is being performed.
+The data rates for events published by the sensors may change, either passively due to local measurement limitations or actively throttled by the utility to meet the application's accuracy needs/resource constraints~\cite{Simmhan:sciencecloud:2011}.
+The data from smart meters can be sent synchronously as they are generated or in batch mode once a day to conserve bandwidth.
+The application processing itself is done elastic resources, scaling up or down to meet latency requirements.
+Some of the forecasting models can also decide to listen to additional data streams on-demand when higher accuracy is required or special situations are encountered.
+
+ \subsubsection*{Other important issues}
+%6. {\bf What else is important?}
+
+%\katznote{from answers to questions 7,8 below, also address programming systems (how was the app written; what tools does it use)}
+
+Semantic
+information integration is being performed using an ontology composed together from individual
+domains~\cite{Zhou:itng:2012}.
+Machine learning models using Matlab, Weka and Hadoop are being used.
+Complex event processing is using the ``Siddhi'' CEP engine with semantic support added~\cite{siddhi}.
+Stream processing on Eucalyptus using IBM Infosphere Streams is being performed.
 
@@
 
-% Consider the energy informatics domain and smart power grids in %
-particular. For data continuously arriving from 2 million smart meters % in
-large city, households will soon need to be continuously analyzed % in order to
-detect impending peak power usage in the smart power grid % and notify the
-utility to respond by either spinning up additional % power sources or by
-triggering load curtailment operations to reduce % the demand. This closed loop
-cyber-physical application, modeled as a % workflow, needs to combine streaming
-data arriving from sensors with % historic data available in file archives along
-with structured % collections of weather forecast data that helps the large
-scale % computational model make an energy use prediction in near real time.
-
-% A workflow framework that supports this data model diversity, % including
-streaming data, structured collections and files, and the % ability to execute
-reliably and scalably on resources is required.
-
-% In general, smart grids - respond to varying data loads, spinning up %
-additional power sources or triggering load curtailment operations.  % E.g., of
-a closed-loop cyber-physical system modeled as a workflow % needs to combine
-streaming data arriving from sensors with historic % data available in archives,
-along with structured collections of % weather forecast data. Need a workflow
-model that supports such % requirements. Taken from ref (Towards Reliable
-Performant Workflows % for Streaming Applications on Cloud Platforms).
-
-% {\it Requirements: } (i) Applications will have to span desktops, %
-workstations and clouds, (ii) streaming data models and stream % programming
-abstractions that are more than TCP sockets, (iii) % reliability of VMs hosting
-the workflow and avoid costly duplicate % movement of the same logical stream
-(see paper in WORKS by Zinn et % al.)
-
-% Scientific workflows are commonplace in eScience applications. Yet, % the lack
-of integrated support for data models, including streaming % data, structured
-collections and files, is limiting the ability of % workflows to support
-emerging applications in energy informatics that % are stream oriented. This is
-compounded by the absence of Cloud data % services that support reliable and
-performant streams.  Need a % scientific workflow framework that supports
-streams as first class % data, and is optimized for performant and reliable
-execution across % desktop and Cloud platforms.
+% Consider the energy informatics domain and smart power grids in
+% particular. For data continuously arriving from 2 million smart meters
+% in large city, households will soon need to be continuously analyzed
+% in order to detect impending peak power usage in the smart power grid
+% and notify the utility to respond by either spinning up additional
+% power sources or by triggering load curtailment operations to reduce
+% the demand. This closed loop cyber-physical application, modeled as a
+% workflow, needs to combine streaming data arriving from sensors with
+% historic data available in file archives along with structured
+% collections of weather forecast data that helps the large scale
+% computational model make an energy use prediction in near real time.
+
+% A workflow framework that supports this data model diversity,
+% including streaming data, structured collections and files, and the
+% ability to execute reliably and scalably on resources is required.
+
+% In general, smart grids - respond to varying data loads, spinning up
+% additional power sources or triggering load curtailment operations.
+% E.g., of a closed-loop cyber-physical system modeled as a workflow
+% needs to combine streaming data arriving from sensors with historic
+% data available in archives, along with structured collections of
+% weather forecast data. Need a workflow model that supports such
+% requirements. Taken from ref (Towards Reliable Performant Workflows
+% for Streaming Applications on Cloud Platforms).
+
+% {\it Requirements: } (i) Applications will have to span desktops,
+% workstations and clouds, (ii) streaming data models and stream
+% programming abstractions that are more than TCP sockets, (iii)
+% reliability of VMs hosting the workflow and avoid costly duplicate
+% movement of the same logical stream (see paper in WORKS by Zinn et
+% al.)
+
+% Scientific workflows are commonplace in eScience applications. Yet,
+% the lack of integrated support for data models, including streaming
+% data, structured collections and files, is limiting the ability of
+% workflows to support emerging applications in energy informatics that
+% are stream oriented. This is compounded by the absence of Cloud data
+% services that support reliable and performant streams.  Need a
+% scientific workflow framework that supports streams as first class
+% data, and is optimized for performant and reliable execution across
+% desktop and Cloud platforms.
 
@@
 
-% Demand response optimization (DR) in smart power grids. Specifically, %
-accurate forecasting of power demand at coarse and fine temporal and spatial %
-granularities, and launching power load curtailment operations when a peak %
-load situation is forecast.
+% Demand response optimization (DR) in smart power grids. Specifically,
+% accurate forecasting of power demand at coarse and fine temporal and spatial
+% granularities, and launching power load curtailment operations when a peak
+% load situation is forecast.
 
@@
 
-% (1) Perform realtime aggregation and analysis on power usage information %
-arriving into the utility's data center from smart meters installed at %
-consumer premises for low latency detection of current (peak) load;
-
-% (2) Perform complex event pattern matching and mining over an event cloud %
-that includes event streams from multiple information source types to detect %
-pattern that forecast a peak load event or opportunity of energy % conservation.
-
-% (3) Construct building and campus energy models using machine learning to %
-predict future energy use.
-
-% 3. What infrastructure is used? (including compute, data, network, %
-instruments, etc.)
+% (1) Perform realtime aggregation and analysis on power usage information
+% arriving into the utility's data center from smart meters installed at
+% consumer premises for low latency detection of current (peak) load;
+
+% (2) Perform complex event pattern matching and mining over an event cloud
+% that includes event streams from multiple information source types to detect
+% pattern that forecast a peak load event or opportunity of energy
+% conservation.
+
+% (3) Construct building and campus energy models using machine learning to
+% predict future energy use.
+
+% 3. What infrastructure is used? (including compute, data, network,
+% instruments, etc.)
 
@@
 
-% Software/Services: Eucalyptus Cloud fabric, IBM InfoSphere Streams, %
-Esper/Siddhi Complex Event Processing (CEP) system, MatLab toolkit, Workflow %
-system (TBD), custom research prototypes.
-
-% Network: Ethernet, others (TBD) % Instruments: Smart meters (consumer power
-usage), Power sensors (fan speed, % room temperature, thermostat setting,
-appliance state, etc.), Facility % sensors (people counter, ambient light, etc.)
-
-% 4. What dynamic data is used in the application?  % (a) What are the types of
-data, (b) What is the size of the data set(s)?
+% Software/Services: Eucalyptus Cloud fabric, IBM InfoSphere Streams,
+% Esper/Siddhi Complex Event Processing (CEP) system, MatLab toolkit, Workflow
+% system (TBD), custom research prototypes.
+
+% Network: Ethernet, others (TBD)
+% Instruments: Smart meters (consumer power usage), Power sensors (fan speed,
+% room temperature, thermostat setting, appliance state, etc.), Facility
+% sensors (people counter, ambient light, etc.)
+
+% 4. What dynamic data is used in the application?
+% (a) What are the types of data, (b) What is the size of the data set(s)?
 
@@
 
-% Smart meter power usage data: 1.4M Consumers x $\sim$1 KB event/min
-($\sim$700~TB/yr (max))
+% Smart meter power usage data: 1.4M Consumers x $\sim$1 KB event/min ($\sim$700~TB/yr (max))
 
@@
 
-% O(secs) to O(hrs) for load forecasting and response, depending on accuracy %
-of prediction and how far ahead the prediction is. There is a tradeoff in % the
-quality constraints based on available power capacity, data, % infrastructure
-resources, granularity of prediction and response, etc. The % absolute quality
-constraint is that the peak power load must {\bf not} exceed % the available
-power capacity. So the power forecast, load curtailment % activation, and the
-impact of the curtailment have to be in a closed loop % cycle whose equilibrium
-is maintained.
+% O(secs) to O(hrs) for load forecasting and response, depending on accuracy
+% of prediction and how far ahead the prediction is. There is a tradeoff in
+% the quality constraints based on available power capacity, data,
+% infrastructure resources, granularity of prediction and response, etc. The
+% absolute quality constraint is that the peak power load must {\bf not} exceed
+% the available power capacity. So the power forecast, load curtailment
+% activation, and the impact of the curtailment have to be in a closed loop
+% cycle whose equilibrium is maintained.
 
@@
 
-% Data from domains such as power systems, building management, weather and %
-traffic, and social networks need to be integrated. We use semantic %
-information integration with an ontology composed together from individual %
-domains.
-
-% 9. Please feel free to also talk about the current state of the % application,
-if it % exists today, and any specific gaps that you know need to be overcome.
-
-% Currently, application for constructing Building and Campus energy models %
-using machine learning and forecasting power usage exist. These over power %
-usage data, building construction features, academic schedule, weather, etc.  %
-Uniform information integration is currently underway to experiment with %
-multiple models.
-
-% Initial versions of stream processing on the private Cloud over campus level %
-smart meter data using information from 100+ buildings exist with adaptive %
-logic for stream rate control.
-
-% Research into semantic CEP is underway, with domain patterns being defined.  %
-Gaps include scalable stream and CEP processing over clouds (including % advance
-resource provisioning and task scheduling), running ``online'' % workflows with
-support for stream inputs that run continuously (to improve % forecast models
-and make predictions), resource selection based on data % privacy and latency
-constraints, and reliable execution with graceful % degradation of critical
-analysis operations.
+% Data from domains such as power systems, building management, weather and
+% traffic, and social networks need to be integrated. We use semantic
+% information integration with an ontology composed together from individual
+% domains.
+
+% 9. Please feel free to also talk about the current state of the
+% application, if it
+% exists today, and any specific gaps that you know need to be overcome.
+
+% Currently, application for constructing Building and Campus energy models
+% using machine learning and forecasting power usage exist. These over power
+% usage data, building construction features, academic schedule, weather, etc.
+% Uniform information integration is currently underway to experiment with
+% multiple models.
+
+% Initial versions of stream processing on the private Cloud over campus level
+% smart meter data using information from 100+ buildings exist with adaptive
+% logic for stream rate control.
+
+% Research into semantic CEP is underway, with domain patterns being defined.
+% Gaps include scalable stream and CEP processing over clouds (including
+% advance resource provisioning and task scheduling), running ``online''
+% workflows with support for stream inputs that run continuously (to improve
+% forecast models and make predictions), resource selection based on data
+% privacy and latency constraints, and reliable execution with graceful
+% degradation of critical analysis operations.
 
@@
 one of the highest priority research items for the US DOE. Scalability of
-first-principles fusion codes is an important part of fusion research needed to
-obtain predictive capability of plasma performance with first-principles physics
-fidelity. The goal is to build a realistic full-distribution kinetic code, in
-realistic ITER geometry, combining the strengths of each of the existing
-gyrokinetic codes (GEM, GTS, XGC, and GTC), and utilizing the most up-to-date
-computer science, hardware design, and applied mathematics technologies
-developed. This {\em traditional application} is intended to be capable of 1)
-discovering macroscopic magnetohydrodynamics activity in burning plasmas, 2)
-understanding Energetic particle effects in fusion reactors, 3) understanding
-the heating in a fusion reactor, and 4) understanding the interaction of the
-plasma at the material boundary. This applications just has the data processing
-stage, along with associated data movement.
-
-\subsubsection*{Application description} %2. {\bf How does it do it?}
-
-%\katznote{from answers to questions 2,3,5,6 below, also address
-cyberinfrastructure (where does the app run) -- perhaps have a diagram here of
-what happens in the app}
-
-The applications that are currently run (XGC1, XGCP, GTC, GTS) run on all of the
-DOE and TeraGrid leadership class facilities, and require very large resources
-for the modest problems that are currently being solved. In the next generation
-(exascale), not only will all of the computing power that will be available at
-the time be needed to understand ITER, all of the data that will be generated
-during the simulation also will have to be understood. The community's
-techniques have been to examine and understand the data in situ, to reduce the
-total amount of data produced in the simulation.
-
- \subsubsection*{Big data aspects} %3. {\bf What are the big data aspects?}
+first-principles fusion codes is an important part of fusion research
+needed to obtain predictive capability of plasma performance with
+first-principles physics fidelity. The goal is to build a realistic
+full-distribution kinetic code, in realistic ITER geometry, combining
+the strengths of each of the existing gyrokinetic codes (GEM, GTS,
+XGC, and GTC), and utilizing the most up-to-date computer science,
+hardware design, and applied mathematics technologies developed. This
+{\em traditional application} is intended to be capable of 1) discovering
+macroscopic magnetohydrodynamics activity in burning plasmas, 2)
+understanding Energetic particle effects in fusion reactors, 3)
+understanding the heating in a fusion reactor, and 4) understanding the
+interaction of the plasma at the material boundary. This applications just
+has the data processing stage, along with associated data movement.
+
+\subsubsection*{Application description}
+%2. {\bf How does it do it?}
+
+%\katznote{from answers to questions 2,3,5,6 below, also address cyberinfrastructure (where does the app run) -- perhaps have a diagram here of what happens in the app}
+
+The applications that are currently run (XGC1, XGCP, GTC, GTS) run on all of the DOE and TeraGrid leadership class facilities, and require very large resources for the modest problems that are currently being solved.
+In the next generation (exascale), not only will all of the computing power that will be available at the time be needed to understand ITER, all of the data that will be generated during the simulation also will have to be understood.
+The community's techniques have been to examine and understand the data in situ, to reduce the total amount of data produced in the simulation.
+
+ \subsubsection*{Big data aspects}
+%3. {\bf What are the big data aspects?}
 
@@
 
-Today the team has produced O(100) TB per week for a simulation using the ADIOS
-middleware~\cite{adios} incorporated in the simulations. Using staging
-techniques, they believe that they will only produce O(10) PB on an exascale
-machine.
-
- \subsubsection*{Distributed aspects} %4. {\bf What are the distributed
-aspects?}
+Today the team has produced O(100) TB per week for a simulation using the ADIOS middleware~\cite{adios} incorporated in the simulations.
+Using staging techniques, they believe that they will only produce O(10) PB on an exascale machine.
+
+ \subsubsection*{Distributed aspects}
+ %4. {\bf What are the distributed aspects?}
 
@@
 
-The primary aspect that is dynamic is that the set of computing resources that
-are used for a given run are based on those that are available at the time of
-that run.  As part of this, data is streamed from one running simulation to
-another, and is transformed in-flight.
-
- \subsubsection*{Other important issues} %6. {\bf What else is important?}
-
-%\katznote{from answers to questions 7,8 below, also address programming systems
-(how was the app written; what tools does it use)}
-
-The applications run on a distributed set of leadership-class facilities, using
-advance reservations to co-schedule the simulations.  Each code reads and writes
-data files, using ADIOS and HDF5.  Files output by each code are transformed and
-transferred to be used as inputs by other codes, linking the codes into a single
-coupled simulation.  Because the overall data generated is so large, it can not
-all be written to disk for post-run analysis, so in-situ analysis and
-visualization tools are being developed.
-
-%\subsubsection{old material follows} %1. What is the science/engineering
-problem? What is the purpose of the %application?
-%
-%With ITER scheduled to operate before 2020, plasma fusion has become %one of
-the highest priority research items for DOE. Scalability of %first-principles
-fusion codes is an important part of fusion research %needed to obtain
-predictive capability of plasma performance with %first-principles physics
-fidelity. We want to build a realistic %full-distribution kinetic code, in
-realistic ITER geometry, combining %the strengths of each of the existing
-gyrokinetic codes (GEM, GTS, %XGC, and GTC), and utilize the most up-to-date
-computer science, %hardware design and applied mathematics technologies
-developed. The %application that we will develop will be capable of 1)
-discovering %Macroscopic Magneto Hydro Dynamics activity in burning plasmas, 2)
-%Understand Energetic particle effects in fusion reactors, 3) %Understand the
-heating in a fusion reactor, and 4) Understand the %interaction of the plasma at
-the material boundary.
+The primary aspect that is dynamic is that the set of computing resources that are used
+for a given run are based on those that are available at the time of that run.  As part of this,
+data is streamed from one running simulation to another, and is transformed in-flight.
+
+ \subsubsection*{Other important issues}
+%6. {\bf What else is important?}
+
+%\katznote{from answers to questions 7,8 below, also address programming systems (how was the app written; what tools does it use)}
+
+The applications run on a distributed set of leadership-class
+facilities, using advance reservations to co-schedule the simulations.
+Each code reads and writes data files, using ADIOS and HDF5.  Files
+output by each code are transformed and transferred to be used as
+inputs by other codes, linking the codes into a single coupled
+simulation.  Because the overall data generated is so large, it can
+not all be written to disk for post-run analysis, so in-situ analysis
+and visualization tools are being developed.
+
+%\subsubsection{old material follows}
+%1. What is the science/engineering problem? What is the purpose of the
+%application?
+%
+%With ITER scheduled to operate before 2020, plasma fusion has become
+%one of the highest priority research items for DOE. Scalability of
+%first-principles fusion codes is an important part of fusion research
+%needed to obtain predictive capability of plasma performance with
+%first-principles physics fidelity. We want to build a realistic
+%full-distribution kinetic code, in realistic ITER geometry, combining
+%the strengths of each of the existing gyrokinetic codes (GEM, GTS,
+%XGC, and GTC), and utilize the most up-to-date computer science,
+%hardware design and applied mathematics technologies developed. The
+%application that we will develop will be capable of 1) discovering
+%Macroscopic Magneto Hydro Dynamics activity in burning plasmas, 2)
+%Understand Energetic particle effects in fusion reactors, 3)
+%Understand the heating in a fusion reactor, and 4) Understand the
+%interaction of the plasma at the material boundary.
 %
@@
 %
-% i) large (exascale) data and large (exascale) computing (e.g., data mining,
-analytics, bioinformatics)
+%        i) large (exascale) data and large (exascale) computing (e.g., data mining, analytics, bioinformatics)
 %
@@
 %
-%The applications we currently run (XGC1, XGCP, GTC, GTS) run on all of the
-leadership class facilities, and require very large resources for the modest
-problems which we are currently solving. In the next generation (exascale) we
-will not only need all of the computing power that will be available at the time
-to understand ITER, we will also need to understand all of the data that will be
-generated during the simulation. Our techniques have been to examine and
-understand the data in situ, to reduce the total amount of data produced in the
-simulation. Today we have produced O(100 TB/week) for a simulation using our
-ADIOS middleware incorporated in our simulations. Using our staging techniques,
-we believe that we will only produce O(10 PB) on an exascale machine. (I don't
-know if this is considered modest dat, or extreme size data)?
+%The applications we currently run (XGC1, XGCP, GTC, GTS) run on all of the leadership class facilities, and require very large resources for the modest problems which we are currently solving.
+In the next generation (exascale) we will not only need all of the computing power that will be available at the time to understand ITER, we will also need to understand all of the data that will be generated during the simulation.
+Our techniques have been to examine and understand the data in situ, to reduce the total amount of data produced in the simulation.
+Today we have produced O(100 TB/week) for a simulation using our ADIOS middleware incorporated in our simulations.
+Using our staging techniques, we believe that we will only produce O(10 PB) on an exascale machine.
+(I don't know if this is considered modest dat, or extreme size data)?
 %
@@
 %
-%1 PB of data -- 10 PB of data, for an exascale size computation run in O(100
-hours).  %Today we achieve about 6\% of peak with excellent scalability to 220K
-%cores on the Cray XT systems. It is %difficult to predict how we will run on
+%1 PB of data -- 10 PB of data, for an exascale size computation run in O(100 hours).
+%Today we achieve about 6\% of peak with excellent scalability to 220K
+%cores on the Cray XT systems. It is
+%difficult to predict how we will run on
 %accelerators, and next generation machines.
 %
-%3. What infrastructure will be used? (including compute, data, %network,
-instruments, etc.)
-%
-%For Validation, we will need to compare our data from the computer to %ITER (or
-current fusion reactors). Typically the amount of data used %in comparison is
-tiny by comparison (people only look at MB's form %experiments).
+%3. What infrastructure will be used? (including compute, data,
+%network, instruments, etc.)
+%
+%For Validation, we will need to compare our data from the computer to
+%ITER (or current fusion reactors). Typically the amount of data used
+%in comparison is tiny by comparison (people only look at MB's form
+%experiments).
 %
@@
 %
-%We always run on all available resources, so if we have computer time %across
-the country, then our data will be distributed, and so will our
-%computation. ITER will have data that will be distributed, but it %still
-unknown.
+%We always run on all available resources, so if we have computer time
+%across the country, then our data will be distributed, and so will our
+%computation. ITER will have data that will be distributed, but it
+%still unknown.
 %
@@
 %
-%I don't understand this question? We will input data (from perhaps a restart),
-and this data is typically in the ADIOS-BP file format for large data. Smaller
-data sets are typically stored in HDF5, and the smallest are stored in MDS+.
+%I don't understand this question?
+We will input data (from perhaps a restart), and this data is typically in the ADIOS-BP file format for large data.
+Smaller data sets are typically stored in HDF5, and the smallest are stored in MDS+.
 %
@@
 %
-%Depends on what we want to understand --- For GTS/GTC we have typically %looked
-at the particle data to compute fluxes, etc. and this data can %be large, but
-due to the Monte Carlo error, and ways to generated 5D %distribution functions
-from the particles, we reduce the particles to %a manageable amount (300 GB per
-time step x 1 K time steps). Our %current I/O is capable of reading this in $<$
-100 s per time step, which %allows us to analyze the data in about 1-2 days.
+%Depends on what we want to understand --- For GTS/GTC we have typically
+%looked at the particle data to compute fluxes, etc. and this data can
+%be large, but due to the Monte Carlo error, and ways to generated 5D
+%distribution functions from the particles, we reduce the particles to
+%a manageable amount (300 GB per time step x 1 K time steps). Our
+%current I/O is capable of reading this in $<$ 100 s per time step, which
+%allows us to analyze the data in about 1-2 days.
 %
@@
 %
-%Don't understand the question. Features change and are dynamic. The physics
-will get more complex as we continue to evolve our codes, and will be more
-dynamic in nature.
+%Don't understand the question. Features change and are dynamic. The physics will get more complex as we continue to evolve our codes, and will be more dynamic in nature.
 %
@@
 %
-%Reading the data from disk --- (Perhaps I don't understand the question).  It
-will also couple to other codes, and the coupling will be in-memory data
-exchange.
-%
-%6. Are there time (or quality) constraints on the application?  (If so, what
-are they?)
-%
-%We always need to run as fast as possible since we have limited computing time
-(our team does have about 100M hours FY 11, but this still puts severe
-restrictions on what we run, so the faster we can run, the more science we can
-get done.
-%
-%7. Please feel free to also talk about the current state of the application, if
-it exists today, and any specific gaps that you know will need to be overcome.
-%
-%Our biggest gaps will be in accelerating the code on next generation exascale
-platforms, and in processing as much data as possible in situ.
+%Reading the data from disk --- (Perhaps I don't understand the question).  It will also couple to other codes, and the coupling will be in-memory data exchange.
+%
+%6. Are there time (or quality) constraints on the application?  (If so, what are they?)
+%
+%We always need to run as fast as possible since we have limited computing time (our team does have about 100M hours FY 11, but this still puts severe restrictions on what we run, so the faster we can run, the more science we can get done.
+%
+%7. Please feel free to also talk about the current state of the application, if it exists today, and any specific gaps that you know will need to be overcome.
+%
+%Our biggest gaps will be in accelerating the code on next generation exascale platforms, and in processing as much data as possible in situ.
 
@@
 
-Applications such as large scale crisis management, maritime security, disease
-and pollution monitoring, and environmental control need to fuse large amounts
-of heterogeneous data to form virtual information sources that mitigate
-noise~\cite{emergency1,emergency2,emergency3,emergency4}.  The area of greater
-Rotterdam in the Netherlands is a chemical hub and houses more than 15\% of the
-Dutch population, making safety and security a major concern.  Some leakage of
-chemicals into the air is an everyday, unavoidable event.
-
-The environmental protection agency (EVA) maintains a command and control call
-center staffed by chemical hazard experts to (1) monitor the air quality and
-sewage, (2) in case of an incident locate the source, (3) identify the threat
-and (4) provide the public authorities, the emergency response organization, and
-the public with situation awareness and suggested actions.  The information that
-is shared is often incomplete, uncertain and inconclusive. Knowledge within the
-EVA as well as from outside experts and industries must be accessed to
-meaningfully use that information.  This subsection describes the {\em
-traditional application} that the EVA uses for these tasks.  The application
-starts after data generation, and includes data transformation \& decisions,
-storage, and processing.
-
-\subsubsection*{Application description} %2. {\bf How does it do it?}
-
-%\katznote{from answers to questions 2,3,5,6 below, also address
-cyberinfrastructure (where does the app run) -- perhaps have a diagram here of
-what happens in the app}
-
-For monitoring and incident detection, continuous streams of data are processed
-and scanned for anomalies from a small number of physical sensors and air
-analysis data (gas-chromatographs outputs).  If an incident or an anomaly is
-detected, new streams of data become available from the general public and field
-inspectors and must be processed. Static information on weather and about
-chemicals are also needed.  These may be stored in different places, access may
-not be granted automatically, and validity and update frequencies may not match
-expectations. Thus specialized pre-processing is needed.
-
-Different knowledge-based processing models must be used account for uncertainty
-in the data and sensor behavior, and provide robust fusion (processing) for
-situation awareness and decision support. Some confidential data must be
-processed \emph{in situ}. New intermediate and final data that are generated may
-have different retention characteristics and need to be stored and made
-accessible for visualization and analysis in different workflows.
-
-Processing in done in semi-real-time or in batch mode processing on a small,
-private computer infrastructure. Therefore, scalability is lacking and the
-dynamics of the workflows is kept under strict human control to match the
-capabilities of a set of static resources.
-
- \subsubsection*{Big data aspects} %3. {\bf What are the big data aspects?}
+Applications such as large scale crisis management, maritime security, disease and pollution
+monitoring, and environmental control need to fuse large amounts of heterogeneous data to form virtual
+information sources that mitigate noise~\cite{emergency1,emergency2,emergency3,emergency4}.  The area of greater Rotterdam in the Netherlands is a
+chemical hub and houses more than 15\% of the Dutch population, making safety and security a major
+concern.  Some leakage of chemicals into the air is an everyday, unavoidable event.
+
+The environmental protection agency (EVA) maintains a command and control call center staffed by
+chemical hazard experts to (1) monitor the air quality and sewage, (2) in case
+of an incident locate the source, (3) identify the threat and (4) provide the public authorities, the
+emergency response organization, and the public with situation awareness and suggested
+actions.  The information that is shared is often incomplete, uncertain and
+inconclusive.
+Knowledge within the EVA as well as from outside experts and industries must be accessed to meaningfully use that information.
+This subsection describes the {\em traditional application} that the EVA uses for these tasks.
+The application starts after data generation, and includes data transformation \& decisions, storage, and processing.
+
+\subsubsection*{Application description}
+%2. {\bf How does it do it?}
+
+%\katznote{from answers to questions 2,3,5,6 below, also address cyberinfrastructure (where does the app run) -- perhaps have a diagram here of what happens in the app}
+
+For monitoring and incident detection, continuous streams of data are processed and scanned for
+anomalies from a small number of physical sensors and air analysis data (gas-chromatographs
+outputs).  If an incident or an anomaly is detected, new streams of data become available from the general public and field inspectors and must be processed. Static information on weather
+and about chemicals are also needed.  These may be stored in
+different places, access may not be granted automatically, and validity and update frequencies may not
+match expectations. Thus specialized pre-processing is needed.
+
+Different knowledge-based processing models must be used account for uncertainty in
+the data and sensor behavior, and provide robust fusion (processing) for situation awareness and
+decision support. Some confidential data must be processed \emph{in situ}. New
+intermediate and final data that are generated may have different retention characteristics and
+need to be stored and made accessible for visualization and analysis in different workflows.
+
+Processing in done in semi-real-time or in batch mode processing on a
+small, private computer infrastructure. Therefore, scalability is
+lacking and the dynamics of the workflows is kept under strict human
+control to match the capabilities of a set of static resources.
+
+ \subsubsection*{Big data aspects}
+%3. {\bf What are the big data aspects?}
 
@@
 
-A number of special data and sensor interface processes will be distributed over
-various platforms and fixed in place.  A low frequency stream of reasonable
-fidelity data from a small number of physical sensors and an even lower
-frequency stream of air analysis data (gas-chromatographs outputs) is
-continuously trickling in.  Low fidelity human provided data from local people
-complaining to the call center, field inspectors calling in with observations
-and measurements, information from social media, pictures from cell phones and
-emails sent to a centralized email account are entered manually into a local
-databases.  Static information such as current weather conditions and forecast
-and information about chemicals are also needed.  These data are made available
-via a distributed shared data space.  Some confidential data must be processed
-remotely and only aggregated results communicated back to other entities. The
-information processing algorithms are packaged as autonomous software agents
-that are managed by a multi agent system infrastructure (middleware) and humans
-(via GUIs). Processing algorithms need to access multiple concurrently and
-remote data stores.
-
- \subsubsection*{Dynamic aspects} %5. {\bf What are the dynamic aspects?}
+A number of special data and sensor interface processes will be distributed over various platforms
+and fixed in place.  A low frequency stream of reasonable fidelity data from a small number of
+physical sensors and an even lower frequency stream of air analysis data (gas-chromatographs
+outputs) is continuously trickling in.  Low fidelity human provided data from local people complaining to the call center,
+field inspectors calling in with observations and measurements, information from
+social media, pictures from cell phones and emails sent to a centralized email account are entered manually into a local databases.  Static information such as current
+weather conditions and forecast and information about chemicals are also needed.  These data are
+made available via a distributed shared data space.
+Some confidential data must be processed remotely and only aggregated results communicated back to other entities.
+The information processing algorithms are packaged as autonomous software agents that are managed by a multi agent system infrastructure
+(middleware) and humans (via GUIs). Processing algorithms need to access multiple concurrently and remote data stores.
+
+ \subsubsection*{Dynamic aspects}
+%5. {\bf What are the dynamic aspects?}
 
@@
 
-Both the data and the processing are dynamic in nature. Data that cannot be
-moved due to confidentiality have to be processed remotely by moving
-applications to them.  A process integration framework will use agents that
-provide autonomous and composable resources to dynamically construct workflows
-for data driven processing.  If an incident happens or an anomaly is detected,
-it triggers new streams of data about other anomalies that must be processed.
-Another source of dynamic behavior is when the event is escalated and additional
-management processes come into play, each requesting and adding their own sort
-of processing to the mix.
-
- \subsubsection*{Other important issues} %6. {\bf What else is important?}
-
-%\katznote{from answers to questions 7,8 below, also address programming systems
-(how was the app written; what tools does it use)}
+Both the data and the processing are dynamic in nature. Data that cannot be moved due to
+confidentiality have to be processed remotely by moving applications to them.  A process integration
+framework will use agents that provide autonomous and composable resources to dynamically construct
+workflows for data driven processing.  If an incident happens or an anomaly is detected, it triggers
+new streams of data about other anomalies that must be processed.  Another source of dynamic
+behavior is when the event is escalated and additional management processes come into play, each
+requesting and adding their own sort of processing to the mix.
+
+ \subsubsection*{Other important issues}
+%6. {\bf What else is important?}
+
+%\katznote{from answers to questions 7,8 below, also address programming systems (how was the app written; what tools does it use)}
 
@@
 applications are the dynamics of scheduling from an application perspective,
-multi independent level of security support, processing service discovery
-support, autonomous reconfiguration support and logging-for-traceability.  Most
-of the dynamic data that these applications use has a short validity period,
-i.e., it must be processed now or we process the next update coming in. The data
-is uncertain, i.e., it may have dynamically changing signal to noise ratio and
-although expected at some base frequency, may not be available for some period
-of time.
+multi independent level of security support, processing service discovery support, autonomous reconfiguration support and logging-for-traceability.
+Most of the dynamic data that these applications use has a short validity
+period, i.e., it must be processed now or we process the next update coming
+in. The data is uncertain, i.e., it may have dynamically changing signal to noise ratio
+and although expected at some base frequency, may not be available for some
+period of time.
 
@@
 % Acknowledgement: The use case and working prototype system discussed here are
-% realized with the help of the DIADEM project grant from the EC (FP7 grant
-224318).
+% realized with the help of the DIADEM project grant from the EC (FP7 grant 224318).
 
@@
 % maritime security, disease and pollution monitoring, environmental control and
-% management etc. cab greatly benefit from the fusion of large amounts of
-heterogeneous % information, coming from static and mobile sensors as well as
-intelligence obtained % from human observers, various databases and the World
-Wide Web. This is because % new AI based techniques are available that help to
-combine multiple information % modalities and thereby increase the amount of
-information for (real-time) fusion % systems, which in turn can mitigate the
-impact of noise and reduce ambiguities and % overcome the unavailability of
-specific information sources.
-
-% However, the compute and networking requirements for this new type of
-applications % that we need to build are significantly different from the more
-traditional massive data % processing or distributed computing applications that
-are known and deployed in % professional business and scientific domains today.
+% management etc. cab greatly benefit from the fusion of large amounts of heterogeneous
+% information, coming from static and mobile sensors as well as intelligence obtained
+% from human observers, various databases and the World Wide Web. This is because
+% new AI based techniques are available that help to combine multiple information
+% modalities and thereby increase the amount of information for (real-time) fusion
+% systems, which in turn can mitigate the impact of noise and reduce ambiguities and
+% overcome the unavailability of specific information sources.
+
+% However, the compute and networking requirements for this new type of applications
+% that we need to build are significantly different from the more traditional massive data
+% processing or distributed computing applications that are known and deployed in
+% professional business and scientific domains today.
 
@@
 
-% The area of greater Rotterdam in the Netherlands is a highly industrialized
-area with the % largest harbor in Europe, very many chemical, petrochemical,
-pharmaceutical and other % processing plants. Economically, it is the most
-important area of the country. In addition, % the same area is home to more than
-15\% of the Dutch population. Safety and security % are therefore a major
-concern, especially with the type of hazards involved and % because major
-evacuations are just not an option. Some leakage of chemicals into the % air
-however, is an every day event and cannot be avoided. Therefore, the %
-environmental protection agency (EVA) is actively monitoring air quality and
-incidents % (that by law must be reported immediately) from a command and
-control center which % includes a dedicated call center for the population. They
-employ a number of chemical % and other hazard experts that can be deployed in
-the field for inspection (prevention) % and measurements. Due to the sheer size
-of the area and the (so far) inadequate % techniques and resources for
-continuous data gathering, data sharing, data processing % and sensor
-management, there are very few sensors in the area that the EVA has % continuous
-access to.
-
-% The tasks of the EVA are to 1) monitor the environment (mainly air quality and
-sewage), % 2) in case of an incident locate the source, 3) identify the threat
-and 4) provide the % public authorities, the emergency response organization and
-the public with situation % awareness and possible courses of action. For that,
-they must liaise and coordinate % with other expert organizations (health and
-hazard experts, public order and safety % experts, the harbor authority, traffic
-experts, etc.) and share and exchange information % and knowledge. An important
-aspect of the latter is that the information that is available % is always
-incomplete, uncertain and no single element is conclusive by itself. In order to
-% process and use that information, inside and outside knowledge must be
-accessed.
+% The area of greater Rotterdam in the Netherlands is a highly industrialized area with the
+% largest harbor in Europe, very many chemical, petrochemical, pharmaceutical and other
+% processing plants. Economically, it is the most important area of the country. In addition,
+% the same area is home to more than 15\% of the Dutch population. Safety and security
+% are therefore a major concern, especially with the type of hazards involved and
+% because major evacuations are just not an option. Some leakage of chemicals into the
+% air however, is an every day event and cannot be avoided. Therefore, the
+% environmental protection agency (EVA) is actively monitoring air quality and incidents
+% (that by law must be reported immediately) from a command and control center which
+% includes a dedicated call center for the population. They employ a number of chemical
+% and other hazard experts that can be deployed in the field for inspection (prevention)
+% and measurements. Due to the sheer size of the area and the (so far) inadequate
+% techniques and resources for continuous data gathering, data sharing, data processing
+% and sensor management, there are very few sensors in the area that the EVA has
+% continuous access to.
+
+% The tasks of the EVA are to 1) monitor the environment (mainly air quality and sewage),
+% 2) in case of an incident locate the source, 3) identify the threat and 4) provide the
+% public authorities, the emergency response organization and the public with situation
+% awareness and possible courses of action. For that, they must liaise and coordinate
+% with other expert organizations (health and hazard experts, public order and safety
+% experts, the harbor authority, traffic experts, etc.) and share and exchange information
+% and knowledge. An important aspect of the latter is that the information that is available
+% is always incomplete, uncertain and no single element is conclusive by itself. In order to
+% process and use that information, inside and outside knowledge must be accessed.
 
@@
 
-% For continuous monitoring and incident detection continuous streams of data
-are % necessary that need to be processed and scanned for anomalies. Typically,
-a low % frequency stream of reasonable fidelity data from a small number of
-physical sensors % and an even lower frequency stream of air analysis data
-(gas-chromatographs outputs) % is trickling in. A much larger amount of data
-from private sensors from hazard sources is % potentially available, but
-currently not accessible. If an incident happens or an anomaly % is detected,
-new streams of data become available and must be processed, such as low %
-fidelity data from people in the area calling the call centre with complaints,
-field(ed) % inspectors calling in with results of expert observations and
-additional high fidelity % measurements, information from social media and maybe
-pictures from cell phones % from individuals in the area (solicited or
-not-solicited). In case of an incident, suddenly % also other more static
-information such as current weather conditions and forecast and % information
-about chemicals (hazards, risks, air-borne or water-borne behavior, health %
-aspects, neutralizing methods and agents, etc.) is needed for further
-processing. This % other information may be stored in different places and
-access may or may not be % granted automatically and formats may be special and
-validity and update frequencies % may not match expectations or needs and thus
-demand for specialized pre-processing % emerges.
+% For continuous monitoring and incident detection continuous streams of data are
+% necessary that need to be processed and scanned for anomalies. Typically, a low
+% frequency stream of reasonable fidelity data from a small number of physical sensors
+% and an even lower frequency stream of air analysis data (gas-chromatographs outputs)
+% is trickling in. A much larger amount of data from private sensors from hazard sources is
+% potentially available, but currently not accessible. If an incident happens or an anomaly
+% is detected, new streams of data become available and must be processed, such as low
+% fidelity data from people in the area calling the call centre with complaints, field(ed)
+% inspectors calling in with results of expert observations and additional high fidelity
+% measurements, information from social media and maybe pictures from cell phones
+% from individuals in the area (solicited or not-solicited). In case of an incident, suddenly
+% also other more static information such as current weather conditions and forecast and
+% information about chemicals (hazards, risks, air-borne or water-borne behavior, health
+% aspects, neutralizing methods and agents, etc.) is needed for further processing. This
+% other information may be stored in different places and access may or may not be
+% granted automatically and formats may be special and validity and update frequencies
+% may not match expectations or needs and thus demand for specialized pre-processing
+% emerges.
 
@@
 
-% The real-time data is mainly stochastic in nature. The fidelity is variable,
-which means % that different knowledge based processing models must be used
-concurrently to take % into account not only the uncertainty in the data /
-measurements itself, but also to take % into account sensor behavior. Most of
-the data are observations of the effects of the root % cause and therefore
-robust fusion (processing) with the help of knowledge based % models is
-necessary to provide situation awareness and decision support. Some fusion %
-requires data that is confidential in nature or commercial in confidence and may
-not % leave the place where it is stored. Therefore the special event based
-processing % algorithms must be brought to the data and only aggregated results
-can be % communicated to other processing entities. This means that both the
-data and the % processing are dynamic in nature, which the application and its
-computer infrastructure % must support. Relationships with static or semi-static
-(e.g., weather) data are often built % in to the processing algorithms, which
-means that these data sources need to be % accessed during real-time processing
-also. This requires multiple simultaneous % accesses to remote data stores. The
-processing will also produce new intermediate and % final data, sometimes with
-different retention characteristics, which need to be stored % and made
-accessible to special visualization and analysis applications in different %
-workflows.
+% The real-time data is mainly stochastic in nature. The fidelity is variable, which means
+% that different knowledge based processing models must be used concurrently to take
+% into account not only the uncertainty in the data / measurements itself, but also to take
+% into account sensor behavior. Most of the data are observations of the effects of the root
+% cause and therefore robust fusion (processing) with the help of knowledge based
+% models is necessary to provide situation awareness and decision support. Some fusion
+% requires data that is confidential in nature or commercial in confidence and may not
+% leave the place where it is stored. Therefore the special event based processing
+% algorithms must be brought to the data and only aggregated results can be
+% communicated to other processing entities. This means that both the data and the
+% processing are dynamic in nature, which the application and its computer infrastructure
+% must support. Relationships with static or semi-static (e.g., weather) data are often built
+% in to the processing algorithms, which means that these data sources need to be
+% accessed during real-time processing also. This requires multiple simultaneous
+% accesses to remote data stores. The processing will also produce new intermediate and
+% final data, sometimes with different retention characteristics, which need to be stored
+% and made accessible to special visualization and analysis applications in different
+% workflows.
 
@@
 
-% In current applications the three main network technologies used are the
-telephone and % cell phone network and a centralized computer network with
-internet connection (for % emails). Data is filtered and entered manually into
-locally housed databases and files % and processed in semi-real-time or batch
-mode processing on a private and limited % computer infrastructure. Therefore
-scalability is lacking and dynamics of workflows is % kept under strict human
-control to match the capabilities of a set of static resources.
+% In current applications the three main network technologies used are the telephone and
+% cell phone network and a centralized computer network with internet connection (for
+% emails). Data is filtered and entered manually into locally housed databases and files
+% and processed in semi-real-time or batch mode processing on a private and limited
+% computer infrastructure. Therefore scalability is lacking and dynamics of workflows is
+% kept under strict human control to match the capabilities of a set of static resources.
 
 % The need however, is to have more scalable solutions where resource usages can
-% actually follow demand from an escalating (and de-escalating) incident and
-information % availability point of view. This is complemented by the need to
-support the dynamic % creation of workflows in a multi-stakeholder (and no
-longer single stakeholder) setting.  % In addition, security is an important
-aspects that drives a separate set of requirements % which impact data
-availability and processing availability that must comply with a % dynamic
-network centric processing model. That model must support secure access at % the
-data-item level and without restrictions at the network or compute platform
-level. A % mixture of the Cloud and Grid approaches may offer a solution here.
+% actually follow demand from an escalating (and de-escalating) incident and information
+% availability point of view. This is complemented by the need to support the dynamic
+% creation of workflows in a multi-stakeholder (and no longer single stakeholder) setting.
+% In addition, security is an important aspects that drives a separate set of requirements
+% which impact data availability and processing availability that must comply with a
+% dynamic network centric processing model. That model must support secure access at
+% the data-item level and without restrictions at the network or compute platform level. A
+% mixture of the Cloud and Grid approaches may offer a solution here.
 
@@
 
-% The use of dynamic data has two origins: the first is the real world situation
-that may % begin with a small scale event such as a minor gas leak. If such an
-event is undetected % at first, it may escalate and trigger more and more
-disturbances that add data about % different other anomalies to the data
-streams. The second source of dynamic behavior % is that with the escalation of
-the event, more and more special repression an % management process come into
-play, each requesting and adding their own sort of % processing to the mix.
+% The use of dynamic data has two origins: the first is the real world situation that may
+% begin with a small scale event such as a minor gas leak. If such an event is undetected
+% at first, it may escalate and trigger more and more disturbances that add data about
+% different other anomalies to the data streams. The second source of dynamic behavior
+% is that with the escalation of the event, more and more special repression an
+% management process come into play, each requesting and adding their own sort of
+% processing to the mix.
 
@@
 
-% In a typical application a number of special data and sensor interface
-processes will be % part of the development. Whether packaged as software agents
-or some other form of % software process, they will normally be distributed over
-various platforms and fixed in % place. There task is to make the data available
-via a distributed shared data space % (used here as a logical concept that can
-be implemented in different ways) that % abstracts from physical storage. The
-information processing algorithms are packaged % as autonomous processing
-entities such as software agents, that are managed by a % multi agent system
-infrastructure (middleware) and humans (via GUIs).
+% In a typical application a number of special data and sensor interface processes will be
+% part of the development. Whether packaged as software agents or some other form of
+% software process, they will normally be distributed over various platforms and fixed in
+% place. There task is to make the data available via a distributed shared data space
+% (used here as a logical concept that can be implemented in different ways) that
+% abstracts from physical storage. The information processing algorithms are packaged
+% as autonomous processing entities such as software agents, that are managed by a
+% multi agent system infrastructure (middleware) and humans (via GUIs).
 
@@
 
-% Quality of service aspects that are important for real world and real time
-applications are % the dynamics of scheduling from an application perspective,
-multi independent level of % security support, processing service discovery
-support, autonomous reconfiguration % support and
-logging-for-traceability. Because of the nature of the application aspects %
-like sustained communication bandwidth, processor availability,
-checkpoint-restart etc.  % are not important. Most of the dynamic data that
-these applications use has a short % validity period, i.e., it must be processed
-now or we process the next update coming in.  % The data is uncertain, i.e., it
-may have dynamically changing S/N ratio and although % expected at some base
-frequency, may not be available for some period of time.
+% Quality of service aspects that are important for real world and real time applications are
+% the dynamics of scheduling from an application perspective, multi independent level of
+% security support, processing service discovery support, autonomous reconfiguration
+% support and logging-for-traceability. Because of the nature of the application aspects
+% like sustained communication bandwidth, processor availability, checkpoint-restart etc.
+% are not important. Most of the dynamic data that these applications use has a short
+% validity period, i.e., it must be processed now or we process the next update coming in.
+% The data is uncertain, i.e., it may have dynamically changing S/N ratio and although
+% expected at some base frequency, may not be available for some period of time.
 
@@
 
-This {\em traditional application} allows the coordinated usage of
-MODIS~\cite{modis} satellite imagery, and its integration with ground based
-sensors and models for environmental applications.  It consists of data storage
-and processing stages.
-
-\subsubsection*{Application description} %2. {\bf How does it do it?}
-
-%\katznote{from answers to questions 2,3,5,6 below, also address
-cyberinfrastructure (where does the app run) -- perhaps have a diagram here of
-what happens in the app}
-
-The application~\cite{modisAzure,modis:ipdps:2010} is used to handle the processes of downloading
-data from different sites and integrating them. The integration consists of
-reprojecting the data from one map projection to another, spatial and temporal
-resampling, and gap filling. The application itself is a loosely-coupled data
+This {\em traditional application} allows the coordinated usage of MODIS~\cite{modis}
+satellite imagery, and its integration with ground based sensors and models for
+environmental applications.  It consists of data storage and processing stages.
+
+\subsubsection*{Application description}
+%2. {\bf How does it do it?}
+
+%\katznote{from answers to questions 2,3,5,6 below, also address cyberinfrastructure (where does the app run) -- perhaps have a diagram here of what happens in the app}
+
+The application~\cite{modisAzure, modis:ipdps:2010} is used to handle the processes of downloading data from different
+sites and integrating them. The integration consists of reprojecting the data from one map projection to another,
+spatial and temporal resampling, and gap filling. The application itself is a loosely-coupled data
 pipeline that coordinates these processes.
 
-In addition, scientists can submit Matlab scripts for further custom processing
-on subsets of the data that has been integrated. This includes some
-summarization processing and plotting routines.  The only time constraints are
-on the debugging cycle for debugging the scientist-supplied reduction
-codes. These are purely driven by the users' patience.
+In addition, scientists can submit Matlab scripts for further custom processing on subsets of
+the data that has been integrated. This includes some summarization processing and plotting
+routines.  The only time constraints are on the debugging cycle for debugging the scientist-supplied
+reduction codes. These are purely driven by the users' patience.
 
@@
 
- \subsubsection*{Big data aspects} %3. {\bf What are the big data aspects?}
+ \subsubsection*{Big data aspects}
+%3. {\bf What are the big data aspects?}
 
@@
 
-There is approximately 25~TB of data for one global year.  The main data is the
-various MODIS satellite data products, of which 11 are used by this
-application. These are in approximately 3,000,000 files for one global year. A
-smaller set of sensor data from the fluxnet data network is used as input to the
-evapotransporation calculation, along with a set of 20-30 files containing
-various metatdata like climate regions used by the calculation.
-
- \subsubsection*{Distributed aspects} %4. {\bf What are the distributed
-aspects?}
+There is approximately 25~TB of data for one global year.  The main data is the various MODIS satellite data
+products, of which 11 are used by this application. These are in approximately 3,000,000 files for one global year. A
+smaller set of sensor data from the fluxnet data network is used as input to the evapotransporation
+calculation, along with a set of 20-30 files containing various metatdata like climate regions used
+by the calculation.
+
+ \subsubsection*{Distributed aspects}
+ %4. {\bf What are the distributed aspects?}
 
@@
 
- The data is collected from a wide variety of FTP servers managed by different
-groups and brought into the cloud/HPC environment. Custom software tooling
-manages and validates this download process. Reliable cloud message queues are
-used to manage the coordination between the different components in the
-multi-stage pipeline. Simple NoSQL tabular storage is used for logging and fault
-recovery. Data is stored in cloud blobs and NoSQL table storage. Application
-scripts on the user's desktop can be uploaded into the cloud as post-processing
-routines and analyses on the integrated data.
-
- \subsubsection*{Dynamic aspects} %5. {\bf What are the dynamic aspects?}
+ The data is collected from a wide variety of FTP servers managed by
+ different groups and brought into the cloud/HPC environment. Custom
+ software tooling manages and validates this download
+ process. Reliable cloud message queues are used to manage the
+ coordination between the different components in the multi-stage
+ pipeline. Simple NoSQL tabular storage is used for logging and fault
+ recovery. Data is stored in cloud blobs and NoSQL table
+ storage. Application scripts on the user's desktop can be uploaded
+ into the cloud as post-processing routines and analyses on the
+ integrated data.
+
+ \subsubsection*{Dynamic aspects}
+%5. {\bf What are the dynamic aspects?}
 
@@
 
- The data is currently static and the application acquires the data from a set
-of FTP servers on a schedule or trigger. But as the pipeline is expanded to
-handle other data, the usage of real-time streaming sensor data is also
-expected. There is infrastructure dynamism through the use of cloud virtual
-machines, acquired and released on-demand, to execute tasks in the pipeline and
-to execute the user's application scripts. While the pipeline structure itself
-is static, there is a limited form of application dynamism through the use of
-user specified scripts pushed into the cloud for execution over the integrated
-data.
-
- \subsubsection*{Other important issues} %6. {\bf What else is important?}
-
-%\katznote{from answers to questions 7,8 below, also address programming systems
-(how was the app written; what tools does it use)}
-
-The data is in one of two map projections, and multiple temporal and spatial
-resolutions. The fluxnet sensor data that is used required significant data
-integration, but was handled in a separate project.
-
-The quality constraints are driven by the needs of the scientific reduction
-application. Significant effort is required to ensure that the data handled to
-the scientific application meets these requirements.
+ The data is currently static and the application acquires the data
+ from a set of FTP servers on a schedule or trigger. But as the
+ pipeline is expanded to handle other data, the usage of real-time
+ streaming sensor data is also expected. There is infrastructure
+ dynamism through the use of cloud virtual machines, acquired and
+ released on-demand, to execute tasks in the pipeline and to execute
+ the user's application scripts. While the pipeline structure itself
+ is static, there is a limited form of application dynamism through
+ the use of user specified scripts pushed into the cloud for execution
+ over the integrated data.
+
+ \subsubsection*{Other important issues}
+%6. {\bf What else is important?}
+
+%\katznote{from answers to questions 7,8 below, also address programming systems (how was the app written; what tools does it use)}
+
+The data is in one of two map projections, and multiple temporal and spatial resolutions. The
+fluxnet sensor data that is used required significant data integration, but was handled in a
+separate project.
+
+The quality constraints are driven by the needs of the scientific reduction application. Significant
+effort is required to ensure that the data handled to the scientific application meets these
+requirements.
 % \subsubsection{old material follows}
@@
 
-% The purpose is to allow the coordinated usage of MODIS satellite imagery, and
-to allow its integration with ground based sensors, models, etc.
+% The purpose is to allow the coordinated usage of MODIS satellite imagery, and to allow its integration with ground based sensors, models, etc.
 
@@
 
-% The application is used to handle the processes of downloading data and data
-integration. The data integration consists of the ability to reproject the data
-from one map projection to another, spatial and temporal resampling, and gap
-filling. The application consists of a loosely-coupled data pipeline that
-coordinates this process.
-
-% 3. What infrastructure is used? (including compute, data, network,
-instruments, etc.)
-
-% The application has been implemented in both the Windows Azure cloud
-environment and a standard HPC environment. The data is collected from a wide
-variety of ftp servers managed by different groups. Significant custom
-infrastructure is required to manage and validate this download
-process. Reliable message queues are used to manage the coordination between the
-different components in the pipeline, and simple NoSql tabular storage is used
-for logging and fault recovery.
+% The application is used to handle the processes of downloading data and data integration.
+The data integration consists of the ability to reproject the data from one map projection to another, spatial and temporal resampling, and gap filling.
+The application consists of a loosely-coupled data pipeline that coordinates this process.
+
+% 3. What infrastructure is used? (including compute, data, network, instruments, etc.)
+
+% The application has been implemented in both the Windows Azure cloud environment and a standard HPC environment.
+The data is collected from a wide variety of ftp servers managed by different groups.
+Significant custom infrastructure is required to manage and validate this download process.
+Reliable message queues are used to manage the coordination between the different components in the pipeline, and simple NoSql tabular storage is used for logging and fault recovery.
 
@@
 
-% The data is currently static, but as we expand the pipeline to handle other
-data we expect to see the usage of real-time streaming sensor data.
+% The data is currently static, but as we expand the pipeline to handle other data we expect to see the usage of real-time streaming sensor data.
 
@@
 
-% The main data is the various MODIS data products, of which we use 11. These
-are in approximately 3,000,000 files for one global year. A smaller set of
-sensor data from the fluxnet data network is used as input to the
-evapotransporation calculation, along with a set of 20-30 files containing
-various metatdata like climate regions used by the calculation.
+% The main data is the various MODIS data products, of which we use 11.
+These are in approximately 3,000,000 files for one global year.
+A smaller set of sensor data from the fluxnet data network is used as input to the evapotransporation calculation, along with a set of 20-30 files containing various metatdata like climate regions used by the calculation.
 
@@
 
-% The only time constraints are on the debugging cycle for debugging the
-scientist-supplied reduction codes. These are purely driven by the users'
-patience. The quality constraints are driven by the needs of the scientific
-reduction application. Significant effort is required to ensure that the data
-handled to the scientific application meets these requirements.
+% The only time constraints are on the debugging cycle for debugging the scientist-supplied reduction codes.
+These are purely driven by the users' patience.
+The quality constraints are driven by the needs of the scientific reduction application.
+Significant effort is required to ensure that the data handled to the scientific application meets these requirements.
 
@@
 
-% The data is in one of two map projections, and multiple temporal and spatial
-resolutions. The fluxnet sensor data that is used required significant data
-integration, but was handled in a separate project.
+% The data is in one of two map projections, and multiple temporal and spatial resolutions.
+The fluxnet sensor data that is used required significant data integration, but was handled in a separate project.
 
@@
 
-% 9. Please feel free to also talk about the current state of the application,
-if it exists today, and any specific gaps that you know need to be overcome.
-
-% \katznote{from talking to Keith: There is a pipeline of initial standardized
-processing steps on the data.  Then, scientists can submit executables that do
-further custom processing on subsets of the data, which likely include some
-summarization processing (building graphs).}
+% 9. Please feel free to also talk about the current state of the application, if it exists today, and any specific gaps that you know need to be overcome.
+
+% \katznote{from talking to Keith: There is a pipeline of initial standardized processing steps on the data.
+Then, scientists can submit executables that do further custom processing on subsets of the data, which likely include some summarization processing (building graphs).}
 
@@
 %\katznote{Yogesh, some stuff is missing - can you fix any of it?}
-%\ysnote{Unfortunately, thats all the info I have from Keith. No response to a
-reminder I sent him earlier.}  %\katznote{I will talk to Alex, the PI of this
-project, on Feb 2.}
+%\ysnote{Unfortunately, thats all the info I have from Keith. No response to a reminder I sent him earlier.}
+%\katznote{I will talk to Alex, the PI of this project, on Feb 2.}
 %
@@
 %
-%This {\em traditional application} constructs real-time hydrodynamics models,
-and forecasts the behavior of a complicated water system: the Sacramento %- San
-Joaquin River delta.  It begins after data generation and includes data
-transformation \& decisions, storage, and processing.
-%
-%\subsubsection*{Application description} %%2. {\bf How does it do it?}
-%
-%%\katznote{from answers to questions 2,3,5,6 below, also address
-cyberinfrastructure (where does the app run) -- perhaps have a diagram here of
-what happens in the app}
-%
-%The application integrates both static and dynamic sensors streaming %real-time
-data with a high-resolution hydrodynamics model running on the NERSC Magellan
-cloud %testbed. The model uses sensor data to initialize and shape the run, and
-the model output is in turn %used to drive the real-time positioning of the
-dynamic sensors. The output of this process is a map %of the current and future
-behavior of the modeled watershed.
-%
-%Custom built dynamic sensors are deployed in the watershed along with
-traditional statically placed %sensors that provide data. These sensors
-communicate over a variety of different network %infrastructures, e.g., GSM,
-CDMA, 802.11, wired ethernet, etc.  and the application acquires the %data over
-the internet from the sensor platforms.  The application processes the sensor
-data in %real-time as it is streamed.
-%
-% \subsubsection*{Big data aspects} %%3. {\bf What are the big data aspects?}
+%This {\em traditional application} constructs real-time hydrodynamics models, and forecasts the behavior of a complicated water system: the Sacramento
+%- San Joaquin River delta.  It begins after data generation and includes data transformation \& decisions, storage, and processing.
+%
+%\subsubsection*{Application description}
+%%2. {\bf How does it do it?}
+%
+%%\katznote{from answers to questions 2,3,5,6 below, also address cyberinfrastructure (where does the app run) -- perhaps have a diagram here of what happens in the app}
+%
+%The application integrates both static and dynamic sensors streaming
+%real-time data with a high-resolution hydrodynamics model running on the NERSC Magellan cloud
+%testbed. The model uses sensor data to initialize and shape the run, and the model output is in turn
+%used to drive the real-time positioning of the dynamic sensors. The output of this process is a map
+%of the current and future behavior of the modeled watershed.
+%
+%Custom built dynamic sensors are deployed in the watershed along with traditional statically placed
+%sensors that provide data. These sensors communicate over a variety of different network
+%infrastructures, e.g., GSM, CDMA, 802.11, wired ethernet, etc.  and the application acquires the
+%data over the internet from the sensor platforms.  The application processes the sensor data in
+%real-time as it is streamed.
+%
+% \subsubsection*{Big data aspects}
+%%3. {\bf What are the big data aspects?}
 %
@@
 %
-% \subsubsection*{Distributed aspects} % %4. {\bf What are the distributed
-aspects?}
+% \subsubsection*{Distributed aspects}
+% %4. {\bf What are the distributed aspects?}
 %
@@
 %
-% \subsubsection*{Dynamic aspects} %%5. {\bf What are the dynamic aspects?}
+% \subsubsection*{Dynamic aspects}
+%%5. {\bf What are the dynamic aspects?}
 %
@@
 %
-%The sensor are all streaming real-time data to the application. The position of
-the dynamic sensors %themselves change over time.
-%
-%
-% \subsubsection*{Other important issues} %%6. {\bf What else is important?}
-%
-%
-%% \katznote{from answers to questions 7,8 below, also address programming
-systems (how was the app written; what tools does it use)}
+%The sensor are all streaming real-time  data to the application. The position of the dynamic sensors
+%themselves change over time.
+%
+%
+% \subsubsection*{Other important issues}
+%%6. {\bf What else is important?}
+%
+%
+%% \katznote{from answers to questions 7,8 below, also address programming systems (how was the app written; what tools does it use)}
 %
@@
 
-% To model in real-time, and forecast, the behavior of a complicated water
-system like the Sacramento - San Joaquin River delta.
+% To model in real-time, and forecast, the behavior of a complicated water system like the Sacramento - San Joaquin River delta.
 
@@
 
-% The application integrated both static (Eulerian) and dynamic (Lagrangian)
-sensors streaming real-time data with a high-resolution hydrodynamics model
-running at the NERSC supercomputing center. The model uses sensor data to
-initialize and shape the run, and model output is used to drive the real-time
-positioning of the Lagrangian sensors. The output of this process is a map of
-the current and future behavior of the modeled watershed.
-
-% 3. What infrastructure is used? (including compute, data, network,
-instruments, etc.)
-
-% Custom built Lagrangian sensors are deployed in the watershed along with
-traditional statically placed sensors. These sensors communicated over a variety
-of different network infrastructures, e.g., GSM, CDMA, 802.11, wired ethernet,
-etc. The model is running on the NERSC Magellan cloud testbed.
+% The application integrated both static (Eulerian) and dynamic (Lagrangian) sensors streaming real-time data with a high-resolution hydrodynamics model running at the NERSC supercomputing center.
+The model uses sensor data to initialize and shape the run, and model output is used to drive the real-time positioning of the Lagrangian sensors.
+The output of this process is a map of the current and future behavior of the modeled watershed.
+
+% 3. What infrastructure is used? (including compute, data, network, instruments, etc.)
+
+% Custom built Lagrangian sensors are deployed in the watershed along with traditional statically placed sensors.
+These sensors communicated over a variety of different network infrastructures, e.g., GSM, CDMA, 802.11, wired ethernet, etc.
+The model is running on the NERSC Magellan cloud testbed.
 
@@
 
-% (a) What are the types of data % (b) What is the size of the data set(s)?
-
-% \katznote{Keith: I need to chat with Alex to get you these answers Dan. Please
-remind me if I don't get back to you on this.}
+% (a) What are the types of data
+% (b) What is the size of the data set(s)?
+
+% \katznote{Keith: I need to chat with Alex to get you these answers Dan. Please remind me if I don't get back to you on this.}
 
@@
 
-% The application is acquired over the internet from the various sensor
-platforms.
+% The application is acquired over the internet from the various sensor platforms.
 
@@
 
-% 9. Please feel free to also talk about the current state of the application,
-if it exists today, and any specific gaps that you know need to be overcome.
+% 9. Please feel free to also talk about the current state of the application, if it exists today, and any specific gaps that you know need to be overcome.
 
@@
 
-This {\em traditional application} performs detection of distributed network
-intrusion and distributed denial of service attacks~\cite{jon}.  Network TCP/UDP
-traffic data is analyzed/mined at each site. Potential alerts are collected
-centrally and further analysis is done to see if an attack may be underway by
-correlating alerts, and events are then communicated back to local sites. The
-following features are extracted: 1) basic features such as source/destination
-IP address pair, port number, and protocols and 2) some additional features such
-as the number of flows from the same source to specific destination IP addresses
-for a predetermined time period. The extracted features are input for the MINDS
-system. While well-known intrusions are detected by the known attack detection
-module, the anomaly detection module is applied to the remaining network
-connections. During the process of anomaly detection, a training set is
-generated if not given as input. The anomaly detection module scores each
-network connection to reflect the level of anomalousness of the network
-connection compared to the normal network traffic. The highly scored network
-connections are analyzed by the association pattern analysis module to produce
-summary and characterization of the possible attacks. This summary and
-characterization is used to create new signatures and models for unknown
-emerging attacks.  MINDS contains various modules for collecting and analyzing
-massive amounts of network traffic.  Typical analyses include behavioral anomaly
-detection, summarization, scan detection and profiling.  The application
-includes all four stages: data generation, transformation \& decisions, storage,
-and processing.
-
-\subsubsection*{Application description} %2. {\bf How does it do it?}
-
-%\katznote{from answers to questions 2,3,5,6 below, also address
-cyberinfrastructure (where does the app run) -- perhaps have a diagram here of
-what happens in the app}
-
-MINDS is deployed at each site as a grid service. The analysis application
-continuously scans local logs, and reports interesting events to a centralized
-analyzer, and alerts are reported back to local sites.
-
-Each site uses a data capturing device such as network flow tools or network
-monitoring tools like tcpdump to collect the network traffic data from routers.
-This collected data is first filtered to remove unimportant network connections
-due to the large data volume.
+This {\em traditional application} performs detection of distributed network intrusion and distributed denial of service attacks~\cite{jon}.  Network TCP/UDP
+traffic data is analyzed/mined at each site. Potential alerts are collected centrally and further
+analysis is done to see if an attack may be underway by correlating alerts, and events are then
+communicated back to local sites. The following features are extracted: 1) basic features such as
+source/destination IP address pair, port number, and protocols and 2) some additional features
+such as the number of flows from the same source to specific destination IP addresses for a
+predetermined time period. The extracted features are input for the MINDS system. While well-known
+intrusions are detected by the known attack detection module, the anomaly detection module is
+applied to the remaining network connections. During the process of anomaly detection, a training
+set is generated if not given as input. The anomaly detection module scores each network
+connection to reflect the level of anomalousness of the network connection compared to the normal
+network traffic. The highly scored network connections are analyzed by the association pattern
+analysis module to produce summary and characterization of the possible attacks. This summary
+and characterization is used to create new signatures and models for unknown emerging attacks.
+MINDS contains various modules for collecting and analyzing massive amounts of network traffic.
+Typical analyses include behavioral anomaly detection, summarization, scan detection and
+profiling.  The application includes all four stages: data generation, transformation \& decisions, storage, and processing.
+
+\subsubsection*{Application description}
+%2. {\bf How does it do it?}
+
+%\katznote{from answers to questions 2,3,5,6 below, also address cyberinfrastructure (where does the app run) -- perhaps have a diagram here of what happens in the app}
+
+MINDS is deployed at each site as a grid service.
+The analysis application continuously scans local logs, and reports interesting events to a centralized analyzer, and alerts are reported back to local sites.
+
+Each site uses a data capturing device such as network flow tools or network monitoring tools like tcpdump  to collect the network traffic data from routers.
+This collected data is first filtered to remove unimportant network connections due to the large data volume.
 
@@
 
-The system is deployed across clusters located at a set of participating sites
-(University of Minnesota, University of Florida, and University of Illinois,
-Chicago) connected by the public Internet. In addition, the data generated is
-also produced (distributed) at this set of sites.
-
- \subsubsection*{Dynamic aspects} %5. {\bf What are the dynamic aspects?}
+The system is deployed across clusters
+located  at a set of participating sites (University of Minnesota, University of Florida, and University of Illinois, Chicago) connected by the public Internet.
+In addition, the data generated is also produced (distributed) at this set of sites.
+
+ \subsubsection*{Dynamic aspects}
+%5. {\bf What are the dynamic aspects?}
 
@@
 
-Network traffic data is highly dynamic and constantly streaming in to the
-application. The data consists of TCP flow and UDP records, which are very
-structured.  They contain IP addresses, ports, various headers, and
-payloads. Interesting events are propagated across sites. These are very small,
-a few KB at most.
-
-%\ysnote{Some of the applications here like behavior anomaly detection seems to
-be triggered on demand, and they consume additional data. This may be similar to
-the LSST case where the application is dynamic/steered based on incoming data or
-situations.}  %\katznote{Yogesh, what is your suggested action to do with this
-comment?}
-
- \subsubsection*{Other important issues} %6. {\bf What else is important?}
-
-%\katznote{from answers to questions 7,8 below, also address programming systems
-(how was the app written; what tools does it use)}
-
- The application is built using Globus Grid Services for wrapping the MINDS
-analysis applications and for providing storage services for events. At the time
-the application was written, the project had to develop tools such as remote
-launching of Grid services themselves. MINDS was also a separate software
-package that was utilized. The project also built an authorization service for
-controlling access to data products. Other noteworthy points: real-time is
-ultimate goal of the project and false positives are more tolerable than false
-negatives.  Finally, the project found it to be very difficult to obtain network
-data at each site due to privacy concerns.
-
- IDS systems also use pattern detection tools such as Bro and Snort, which
-perform simple string and regular expression pattern matching on IP packet
-headers and body. These are monolithic tools that run standalone, without any
-comprehensive programming support to automate updating the patterns or taking
-action upon pattern detection.
+Network traffic data is highly dynamic and constantly streaming in to the application.
+The data consists of TCP flow and UDP records, which are very structured.
+They contain IP addresses, ports, various headers, and payloads.
+Interesting events are propagated across sites.
+These are very small, a few KB at most.
+
+%\ysnote{Some of the applications here like behavior anomaly detection seems to be triggered on demand, and they consume additional data.
+This may be similar to the LSST case where the application is dynamic/steered based on incoming data or situations.}
+%\katznote{Yogesh, what is your suggested action to do with this comment?}
+
+ \subsubsection*{Other important issues}
+%6. {\bf What else is important?}
+
+%\katznote{from answers to questions 7,8 below, also address programming systems (how was the app written; what tools does it use)}
+
+ The application is built using Globus Grid Services for wrapping the
+ MINDS analysis applications and for providing storage services for
+ events. At the time the application was written, the project had to
+ develop tools such as remote launching of Grid services
+ themselves. MINDS was also a separate software package that was
+ utilized. The project also built an authorization service for
+ controlling access to data products. Other noteworthy points:
+ real-time is ultimate goal of the project and false positives are
+ more tolerable than false negatives.  Finally, the project found it
+ to be very difficult to obtain network data at each site due to
+ privacy concerns.
+
+ IDS systems also use pattern detection tools such as Bro and Snort,
+ which perform simple string and regular expression pattern matching on
+ IP packet headers and body. These are monolithic tools that run
+ standalone, without any comprehensive programming support to automate
+ updating the patterns or taking action upon pattern detection.
 
@@
 %
-%Distributed network intrusion detection and distributed denial of service
-attacks.  Network TCP/UDP traffic data is analyzed/mined at each site. Potential
-alerts are collected centrally and further analysis is done to see if an attack
-may be under way by correlating alerts and events are then communicated back to
-local sites. The following features are extracted:1) basic features such as
-source/destination IP address pair, port number, and protocols and 2) some
-additional features such as the number of flows from the same source to specific
-destination IP addresses for a predetermined time period. The extracted features
-are input for MINDS system. While well-known intrusions are detected by the
-known attack detection module, the anomaly detection module is applied to the
-remaining network connections. During the process of anomaly detection, a
-training set is generated if not given as input. The anomaly detection module
-scores each network connection to reflect the level of anomalousness of the
-network connection compared to the normal network traffic. The highly scored
-network connections are analyzed by the association pattern analysis module to
-produce summary and characterization of the possible attacks. This summary and
-characterization is used to create new signatures and models for unknown
-emerging attacks. MINDS contains various modules for collecting and analyzing
-massive amounts of network traffic. Typical analyses include behavioral anomaly
-detection, summarization, scan detection and profiling.
+%Distributed network intrusion detection and distributed denial of service attacks.
+Network TCP/UDP traffic data is analyzed/mined at each site.
+Potential alerts are collected centrally and further analysis is done to see if an attack may be under way by correlating alerts and events are then communicated back to local sites.
+The following features are extracted:1) basic features such as source/destination IP address pair, port number, and protocols and 2) some additional features such as the number of flows from the same source to specific destination IP addresses for a predetermined time period.
+The extracted features are input for MINDS system.
+While well-known intrusions are detected by the known attack detection module, the anomaly detection module is applied to the remaining network connections.
+During the process of anomaly detection, a training set is generated if not given as input.
+The anomaly detection module scores each network connection to reflect the level of anomalousness of the network connection compared to the normal network traffic.
+The highly scored network connections are analyzed by the association pattern analysis module to produce summary and characterization of the possible attacks.
+This summary and characterization is used to create new signatures and models for unknown emerging attacks.
+MINDS contains various modules for collecting and analyzing massive amounts of network traffic.
+Typical analyses include behavioral anomaly detection, summarization, scan detection and profiling.
 %
@@
 %
-%MINDS is deployed at each site as a service. The analysis application
-continuously scans local logs, and reports interesting events to a centralized
-analyzer, and alerts are reported back to local sites.  These communication
-steps use messages.
-%
-%3. What infrastructure is used? (including compute, data, network, instruments,
-etc.)
-%
-%Small clusters at a set of participating sites (UMn, UFL, UIC). Public Internet
-connectivity.
+%MINDS is deployed at each site as a service.
+The analysis application continuously scans local logs, and reports interesting events to a centralized analyzer, and alerts are reported back to local sites.
+These communication steps use messages.
+%
+%3. What infrastructure is used? (including compute, data, network, instruments, etc.)
+%
+%Small clusters at a set of participating sites (UMn, UFL, UIC). Public Internet connectivity.
 %
@@
 %
-%TCP flow and UDP records -- very structured: contains IP addresses, ports,
-various headers, and payloads. Interesting events are propagated across
-sites. These are very small, a few KB at most.
+%TCP flow and UDP records -- very structured: contains IP addresses, ports, various headers, and payloads. Interesting events are propagated across sites. These are very small, a few KB at most.
 %
@@
 %
-%The data capturing device such as network flow tools or network monitoring
-tools like tcpdump are used to collect the network traffic data from routers.
+%The data capturing device such as network flow tools or network monitoring tools like tcpdump are used to collect the network traffic data from routers.
 The collected data is first filtered to remove unimportant network connections.
@@
 %
-%Real-time is ultimate goal. False positives are more tolerable than false
-negatives.
-%
-%Gaps: The project had to develop tools such as remote launching of Grid
-services themselves. They also found it to be very difficult to obtain network
-data at each site due to privacy concerns.
+%Real-time is ultimate goal. False positives are more tolerable than false negatives.
+%
+%Gaps:  The project had to develop tools such as remote launching of Grid services themselves. They also found it to be very difficult to obtain network data at each site due to privacy concerns.
 %
@@
 %
-%Wikianalytics is based on the concept of collective intelligence and social
-computing.  %In Wikianalytics, complex and large scale data is analyzed via a
-group of analysts %collectively in an interactive and collaborative way. The
-data mining life cycle is not a %linear process but a dynamic composition of
-concurrent interactive processes with %the joint effort of participating
-analysts.  Each analyst contributes data, prior %knowledge, expertise,
-algorithms to build, validate, improve, integrate models %(knowledge).  Also,
-modern cloud computing and social computing model provides %an ideal platform
-for Wikianalytics.
+%Wikianalytics is based on the concept of collective intelligence and social computing.
+%In Wikianalytics, complex and large scale data is analyzed via a group of analysts
+%collectively  in an interactive and collaborative way. The data mining life cycle is not a
+%linear process  but a dynamic composition of  concurrent interactive processes  with
+%the joint effort of participating analysts.  Each analyst contributes data, prior
+%knowledge, expertise,  algorithms to build, validate, improve, integrate models
+%(knowledge).  Also, modern cloud computing and social computing model provides
+%an ideal platform for Wikianalytics.
 %
@@
 %
-%Data intensive: Data intensive is not just about the quantity but also the
-complexity.  %Given the human genome example, the pervasive use of human genome
-for %healthcare requires the analysis to be done with medical knowledge.
-Analyzing %human genome is not just the issue of dealing with the huge amount of
-sequence %data but also the need of deep analysis the biological and medical
-implication of the %data. To meet this challenge, collective effort in analysis
-is essential.
-%
-%Distributed: Distributed feature has three means : 1) data is distributed 2)
-data owner/ %analysts are distributed and 2) data analysis environment is
-distributed.  All three %aspects requires the analysis to be a compositional
-process. That is, distributed data %can be analyzed in a distributed way and the
-analytical products ( information or %knowledge) can then be combined
-(integrated). Such a composition of distributed %process is a ``communicating
-parallel composition'' in the sense that for time to time %each individual
-component process will communicate with others to exchange the %information and
-coordinate the actions. Such a distributed analytics process can be %realized
-well by the Wiki collective intelligence principle.
+%Data intensive: Data intensive is not just about the quantity but also the complexity.
+%Given the human genome example,  the pervasive use of human genome for
+%healthcare requires the analysis to be done with medical knowledge.  Analyzing
+%human genome is not just the issue of dealing with the huge amount of sequence
+%data but also the need of deep analysis the biological and medical implication of the
+%data. To meet this challenge, collective effort in analysis is essential.
+%
+%Distributed: Distributed feature has three means : 1) data is distributed 2) data owner/
+%analysts are distributed and 2) data analysis environment is distributed.  All three
+%aspects requires the analysis to be a compositional process. That is, distributed data
+%can be analyzed in a distributed way and the analytical products ( information or
+%knowledge) can then be combined (integrated). Such a composition of distributed
+%process is a ``communicating parallel composition'' in the sense that for time to time
+%each individual component process will communicate with others to exchange the
+%information and coordinate the actions. Such a distributed analytics process can be
+%realized well by the Wiki collective intelligence principle.
 %
 %Dynamical:  Dynamical nature of data intensive computing comes the dynamism in
-%data availability in data comprehension and also in the inherent dynamism in
-the %applications.  From data analysis point of view, the dynamism is reflected
-by the fact %that an analytical process cannot be captured by a static workflow,
-rather it has to be %dynamically constructed based on the partial information
-derived during the analysis.  %Such an adaptive nature of analysis is described
-in Bayesian statistics as the %posterior probability where knowledge (belief) is
-dynamically updated based on the %new available and analyzed data.  Wikianalysis
-is not based on the concept of %organizing data via workflow, rather its
-collective mechanism reflects precisely the %dynamism needed in analyzing
-complex and dynamic data where the analysis %process requires to be built
-adaptively based on the up-to-date analytical results gain %from the joint
-effort.
+%data availability in data comprehension and also in the inherent dynamism in the
+%applications.  From data analysis point of view, the dynamism is reflected by the fact
+%that an analytical process cannot be captured by a static workflow, rather it has to be
+%dynamically constructed based on the partial information derived during the analysis.
+%Such an adaptive nature of analysis is described in Bayesian statistics as the
+%posterior probability where knowledge (belief) is dynamically updated based on the
+%new available and analyzed data.  Wikianalysis  is not based on the concept of
+%organizing data  via workflow, rather its collective mechanism reflects precisely the
+%dynamism needed in analyzing complex and dynamic data where the analysis
+%process requires to be built adaptively based on the up-to-date analytical results gain
+%from the joint effort.
 
@@
 
-Here we provide short summaries of the computational (as opposed to scientific)
-aspects of each application in preparation for the analysis that follows in the
-rest of this document, which is motivated by and centered around these set of
-applications.
-
-{\bf NGS Analytics (\S\ref{bioSilvia})}: Data is processed through a customized
-pipeline of analysis tools, such as BWA, BFAST, Bowtie; processing of each data
-element is independent of other elements, so processing is
-decomposed/parallelized over the data elements.  All data is stored in files.
-Later analysis may want to operate on files from multiple data stores
-distributed around the world, e.g., to compare O(1000s) of TB-scale data sets
-(metagenomics) with each other.  Dynamic elements include the choice of
-executables in each instance of the pipeline, the design of the pipeline itself,
-and the choice of computing infrastructure to provide the best resource usage
-and turnaround time.  The solution to this problem is not clear.  There is a
-need for flexible programming systems and associated runtime environments for
-hardware-software co-optimization.
-
-{\bf ATLAS/WLCG (\S\ref{WLCGSteve})}: There is a hierarchy of systems. Data are
-centrally stored, and locally cached (and copied to where they likely will be
-used), perhaps at various levels of the hierarchy. Processing is done by
-applications that are independent of each other. Processing of one data file is
-independent of processing of another file, but groups of processing results are
-collected to obtain statistical outputs about the data.
-
-{\bf LSST (\S\ref{astroAdam})}: Data is taken by a telescope. Quick analysis is
-done at the telescope site for interesting (urgent) events (which may involve
-comparing new data with previous data).  The system can get more data from other
-observatories if needed, request other observatories to take more data, or call
-a human. Data (in files) is then transferred to an archive site, which may be at
-the observatory. At the archive site, the data are analyzed, reduced, and
-classified, some of which may be farmed out to grid resources.  Detailed
-analysis of new data vs. archived data is performed.  Reanalysis of all data is
-done periodically.  Data are stored in files and databases.
-
-{\bf SOA Astronomy (\S\ref{astroSOAAdam})}: Services are orchestrated through a
-pipeline, including a data retrieval service that is used to share data across
-VO sites, as well as analysis and image processing services.  Data (files) are
-moved through the pipeline, and intermediate and final products can be stored in
-AstroGrid storage service, such as mySpace.  Scratch space is provided for
-storage of intermediate results.
-
-{\bf Cosmic Microwave Background (\S\ref{astroJulian})}: This application builds
-a sky map.  It reads data from detectors, and obtains data from simulations that
-are performed on-the-fly, when requested by the map-making component (in a
-master-worker sense), one of a set of components that gradually reduces the data
-from the detected and simulated data to a small number of cosmological
-parameters.
-
-{\bf Sensor Network Application (\S\ref{sensorSimon})}: Data are collected from
-mobile sensors that periodically transmit when they come within communication
-distance of receivers. Fault tolerance (collecting data in a hostile
-environment) is an important component. The data, which are very diverse,
-because they can involved different resolutions, sampling frequencies and other
-management characteristics., are brought to a central site. Stored data are
-analyzed using statistical techniques, then visualized with tools such as Google
-Earth.
-
-{\bf Climate (\S\ref{climateDan})}: Climate simulations are run by climate
-centers.  Data (outputs of climate simulations) are distributed in multiple
-federated stores.  Current data analysis and visualization tools require the
-user to bring the data they want to analyze to a local system, then run the
-tools locally.  New tools will include the capability to add data transfer.
+Here we provide short summaries of the computational (as
+opposed to scientific) aspects of each
+application in preparation for the analysis that
+follows in the rest of this document, which is motivated by and
+centered around these set of applications.
+
+{\bf NGS Analytics (\S\ref{bioSilvia})}: Data is processed through a
+customized pipeline of analysis tools, such as BWA, BFAST, Bowtie;
+processing of each data element is independent of other elements, so
+processing is decomposed/parallelized over the data elements.  All
+data is stored in files.  Later analysis may want to operate on files
+from multiple data stores distributed around the world, e.g., to
+compare O(1000s) of TB-scale data sets (metagenomics) with each other.
+Dynamic elements include the choice of executables in each instance of
+the pipeline, the design of the pipeline itself, and the choice of
+computing infrastructure to provide the best resource usage and
+turnaround time.  The solution to this problem is not clear.  There is
+a need for flexible programming systems and associated runtime
+environments for hardware-software co-optimization.
+
+{\bf ATLAS/WLCG (\S\ref{WLCGSteve})}: There is a hierarchy of systems.
+Data are centrally stored, and locally cached (and copied to where they likely will be used), perhaps at various levels of the hierarchy.
+Processing is done by applications that are independent of each other.
+Processing of one data file is independent of processing of another file, but groups of processing results are collected to obtain statistical outputs about the data.
+
+{\bf LSST (\S\ref{astroAdam})}: Data is taken by a telescope. Quick analysis is done at the telescope site for interesting (urgent) events (which may involve comparing new data with previous data).
+The system can get more data from other observatories if needed, request other observatories to take more data, or call a human.
+Data (in files) is then transferred to an archive site, which may be at the observatory.
+At the archive site, the data are analyzed, reduced, and classified, some of which may be farmed out to grid resources.
+Detailed analysis of new data vs.
+archived data is performed.
+Reanalysis of all data is done periodically.
+Data are stored in files and databases.
+
+{\bf SOA Astronomy (\S\ref{astroSOAAdam})}: Services are orchestrated through a pipeline, including a data retrieval service that is used to share data across VO sites, as well as analysis and image processing services.
+Data (files) are moved through the pipeline, and intermediate and final products can be stored in AstroGrid storage service, such as mySpace.
+Scratch space is provided for storage of intermediate results.
+
+{\bf Cosmic Microwave Background (\S\ref{astroJulian})}: This application builds a sky map.
+It reads data from detectors, and obtains data from simulations that are performed on-the-fly, when requested by the map-making component (in a master-worker sense), one of a set of components that gradually reduces the data from the detected and simulated data to a small number of cosmological parameters.
+
+{\bf Sensor Network Application (\S\ref{sensorSimon})}: Data are collected from mobile sensors that periodically transmit when they come within communication distance of receivers.
+Fault tolerance (collecting data in a hostile environment) is an important component.
+The data, which are very diverse, because they can involved different resolutions, sampling frequencies and other management characteristics., are brought to a central site.
+Stored data are analyzed using statistical techniques, then visualized with tools such as Google Earth.
+
+{\bf Climate (\S\ref{climateDan})}: Climate simulations are run by climate centers.
+Data (outputs of climate simulations) are distributed in multiple federated stores.
+Current data analysis and visualization tools require the user to bring the data they want to analyze to a local system, then run the tools locally.
+New tools will include the capability to add data transfer.
 This will allow the set of tools to iterate: transfer data, do analysis, etc.
 
-{\bf Interactive Exploration of Environmental Data (\S\ref{envJon})}:
-Applications are driven from map-based graphical tools.  The inputs are
-geographic areas and algorithms to run on the data in those areas. Data can come
-through web services from stored data, sensors, or simulations (possibly
-running).  The services can perform server-side processing to reduce the size of
-data files to be transferred. Responses are provided in near-real-time.
+{\bf Interactive Exploration of Environmental Data (\S\ref{envJon})}: Applications are driven from map-based graphical tools.
+The inputs are geographic areas and algorithms to run on the data in those areas.
+Data can come through web services from stored data, sensors, or simulations (possibly running).
+The services can perform server-side processing to reduce the size of data files to be transferred.
+Responses are provided in near-real-time.
 Security also supported through a delegation-based model.
 
-{\bf Power Grids (\S\ref{power})}: Diverse streams arrive at a central utility
-private cloud over cellular and wireless networks at dynamic rates controlled by
-the application. Communication is bi-directional, with pricing and control
-signals sent to customers power meters.  A real-time event detection pipeline,
-involving complex event processing and machine learning/data analysis algorithms
-and using other stored information such as weather data, can trigger load
-curtailment operations. Data mining is performed on current and historical data
-for forecasting. Partial application execution on remote micro-grid sites is
-possible.
-
-{\bf Fusion (\S\ref{fusionScott})}: Multiple computational intensive physics
-simulation codes run concurrently on a distributed set of parallel computers.
-Data from some codes are streamed to other codes to link them into a single
-simulation.  Data are transformed in flight into needed inputs.  Data are also
-examined in situ to understand the overall state of the simulation.  Some data
-are stored for later data mining and visualization.
-
-{\bf Industrial Incident Notification and Response (\S\ref{Kees})}: Data are
-regularly streamed from diverse sources, and sometimes manually entered into the
-system. Initial disaster detection causes additional information sources to be
-requested from that region and one or more applications to be composed based on
-available data. Some applications run on remote sites (where the data is stored)
-for data privacy. Escalation can cause more humans in the loop and additional
+{\bf Power Grids (\S\ref{power})}: Diverse streams arrive at a central utility private cloud over cellular and wireless networks at dynamic rates controlled by the application.
+Communication is bi-directional, with pricing and control signals sent to customers power meters.
+A real-time event detection pipeline, involving complex event processing and machine learning/data analysis algorithms and using other stored information such as weather data, can trigger load curtailment operations.
+Data mining is performed on current and historical data for forecasting.
+Partial application execution on remote micro-grid sites is possible.
+
+{\bf Fusion (\S\ref{fusionScott})}: Multiple computational intensive
+physics simulation codes run concurrently on a distributed set of
+parallel computers.  Data from some codes are streamed to other codes
+to link them into a single simulation.  Data are transformed in flight
+into needed inputs.  Data are also examined in situ to understand the
+overall state of the simulation.  Some data are stored for later data
+mining and visualization.
+
+{\bf Industrial Incident Notification and Response (\S\ref{Kees})}:
+Data are regularly streamed from diverse sources, and sometimes
+manually entered into the system. Initial disaster detection causes
+additional information sources to be requested from that region and
+one or more applications to be composed based on available data. Some
+applications run on remote sites (where the data is stored) for data
+privacy. Escalation can cause more humans in the loop and additional
 operations.
 
-{\bf MODIS Data Processing (\S\ref{modisKeith})}: Data are brought into system
-from various FTP servers.  A pipeline of initial standardized processing steps
-on data is done on clouds or HPC resources.  Scientists can then submit
-executables and Matlab scripts that do further custom processing on subsets of
-the data, which likely include some summarization processing (building graphs).
-The application has been run on a cloud and in an HPC environment.
-
-%Floating Sensors (\S\ref{floaterKeith}): Data from static and dynamic %sensors
-are brought to a central location for processing by models. The results %of
-modeling and forecasting can cause the sensor locations to change.
-
-{\bf Distributed Network Intrusion Detection (\S\ref{detectionJon})}: Data are
-analyzed or mined by a service running at a number of local sites. The results
-are sent (via HTTP messages) to a central site for analysis. Events are then
-communicated (via HTTP messages) back to local sites.
+{\bf MODIS Data Processing (\S\ref{modisKeith})}: Data are brought
+into system from various FTP servers.  A pipeline of initial
+standardized processing steps on data is done on clouds or HPC
+resources.  Scientists can then submit executables and Matlab scripts
+that do further custom processing on subsets of the data, which likely
+include some summarization processing (building graphs).  The
+application has been run on a cloud and in an HPC environment.
+
+%Floating Sensors (\S\ref{floaterKeith}): Data from static and dynamic
+%sensors are brought to a central location for processing by models. The results
+%of modeling and forecasting can cause the sensor locations to change.
+
+{\bf Distributed Network Intrusion Detection (\S\ref{detectionJon})}:
+Data are analyzed or mined by a service running at a number of local
+sites. The results are sent (via HTTP messages) to a central site for
+analysis. Events are then communicated (via HTTP messages) back to
+local sites.
 
@@
 
-% \jhanote{AL and I agree the structure should be: 3.1. (i) Define % Dynamic,
-(ii) Formalism/types (iii) % Applications/tables/discussions. 3.2 (i) Define
-distributed, (ii) % discuss two stages (a) structuring (static) (b) processing %
-(dynamic), (iii) Formalism/table, (iv) Application discussion; 3.3 % discussion}
-
-In the previous section we described thirteen applications and analyzed them
-along an analytical space spanned by the three ``D''s of data-intensive,
-distributed, and dynamic behavior.
-
-\emph{Distribution} refers to the presence of application data in different
-physical or logical locations. Distributed data may arise naturally in that the
-data sources themselves are geographically separated, such as from remotely
-located sensors.  In some cases, data must remain distributed for privacy and
-policy reasons, or due to constraints in sharing.  In other cases, a full
-dataset may not fit in one location either on disk or in memory and must be
-decomposed to enable storage and/or computation. Distributed data may be driven
-by a need for performance, scalability localization, reliability, or
-availability considerations. For example, copies of data may be generated and
-distributed to maintain data availability or load balancing.  In general, a
-\emph{distributed application} is one that needs, or would benefit from the use
-of, multiple resource~\cite{dpa_surveypaper}. Example benefits include increased
+% \jhanote{AL and I agree the structure should be: 3.1. (i) Define
+%   Dynamic, (ii) Formalism/types (iii)
+%   Applications/tables/discussions. 3.2 (i) Define distributed, (ii)
+%   discuss two stages (a) structuring (static) (b) processing
+%   (dynamic), (iii) Formalism/table, (iv) Application discussion; 3.3
+%   discussion}
+
+In the previous section we described thirteen applications and
+analyzed them along an analytical space
+spanned by the three ``D''s of data-intensive, distributed, and
+dynamic behavior.
+
+\emph{Distribution} refers to the presence of application data in
+different physical or logical locations. Distributed data may arise
+naturally in that the data sources themselves are geographically
+separated, such as from remotely located sensors.  In some cases, data
+must remain distributed for privacy and policy reasons, or due to
+constraints in sharing.  In other cases, a full dataset may not fit in
+one location either on disk or in memory and must be decomposed to
+enable storage and/or computation. Distributed data may be driven by a
+need for performance, scalability localization, reliability, or
+availability considerations. For example, copies of data may be
+generated and distributed to maintain data availability or load
+balancing.  In general, a \emph{distributed application} is one that
+needs, or would benefit from the use of, multiple
+resource~\cite{dpa_surveypaper}. Example benefits include increased
 throughput, decreased time to solution, and increased reliability.  A
 distributed application rarely consist of a single component that is
-distributed; most often a distributed application is composed of smaller and
-functionally self-contained, if not entirely independent, components that can
-execute on different distributed resources.
-
-\emph{Dynamism} refers to changes in the behavior and characteristics of an
-entity % \jhanote{``of an entity'', does this % sound right?} \katznote{fine
-with me} over time and/or space.  Over a long enough timescale, almost all
-infrastructure, and especially distributed infrastructure, will exhibit some
-form of dynamism, as a consequence of failure, varying load, etc. However, we
-are interested in dynamism on timescales that influence application behavior,
-scheduling decisions, etc. Some prominent examples are resource fluctuation,
-data availability, and change in application execution characteristics.
-
-In this section we will try to extract commonalities in the dynamic behavior of
-applications studied in the previous section, as well as properties and types of
-distribution seen in these applications.
-
-% \jhanote{needs fixing} After discussing the different kinds of % dynamism, we
-will apply this framework to the applications.  This will % permit an extraction
-of common features and possibly \katznote{either % we can do this or not...} a
-derivation of requirements for multiple % applications.
-
-\subsection{Types of Distribution \label{sec:distTypes}} %Distributed
-\emph{data} occurs e.\,g.\
-
-There are many scenarios under which distribution of data is either necessary or
-desired. In addition to the reasons behind distribution presented at the
-beginning of this section (viz., large volumes of data) and reasons that
-constrain or localize data (viz., privacy/policy issues), another common
-situation under which data is distributed occurs when data comes from multiple
-sources, for example multiple sensor streams or data centers, but needs to be
-processed collectively.
-
-% and (ii) where a data % set is too large for a single storage module and thus
-must be stored % in a distributed % way.  % \alnote{application
-stages?}\jhanote{Andre, please clarify?}
+distributed; most often a distributed application is composed of
+smaller and functionally self-contained, if not entirely independent,
+components that can execute on different distributed resources.
+
+\emph{Dynamism} refers to changes in the behavior and characteristics
+of an entity % \jhanote{``of an entity'', does this
+% sound right?} \katznote{fine with me}
+over time and/or space.  Over a long enough timescale, almost all
+infrastructure, and especially distributed infrastructure, will
+exhibit some form of dynamism, as a consequence of failure, varying
+load, etc. However, we are interested in dynamism on timescales that
+influence application behavior, scheduling decisions, etc. Some
+prominent examples are resource fluctuation, data availability, and
+change in application execution characteristics.
+
+In this section we will try to extract commonalities in the dynamic
+behavior of applications studied in the previous section, as well as
+properties and types of distribution seen in these applications.
+
+% \jhanote{needs fixing} After discussing the different kinds of
+% dynamism, we will apply this framework to the applications.  This will
+% permit an extraction of common features and possibly \katznote{either
+%   we can do this or not...} a derivation of requirements for multiple
+% applications.
+
+\subsection{Types of Distribution \label{sec:distTypes}}
+%Distributed \emph{data} occurs e.\,g.\
+
+There are many scenarios under which distribution of data is either
+necessary or desired. In addition to the reasons behind distribution
+presented at the beginning of this section (viz., large volumes of
+data) and reasons that constrain or localize data (viz.,
+privacy/policy issues), another common situation under which data is
+distributed occurs when data comes from multiple sources, for example
+multiple sensor streams or data centers, but needs to be processed
+collectively.
+
+%  and (ii) where a data
+% set is too large for a single storage module and thus must be stored
+% in a distributed
+% way.
+% \alnote{application stages?}\jhanote{Andre, please clarify?}
 
@@
 
-% One or multiple data sources (M) are expanded to one or more % locations (N),
-M$ <$ N, for processing.  Scattering is commonly used % to facilitate
-distributed processing of data.
-
-There are many ways to address distributed data.  These include but are not
-limited to: % Ways to manage \katznote{?} the % distribution include but are not
-limited to:
+% One or multiple data sources (M) are expanded to one or more
+% locations (N), M$ <$ N, for processing.  Scattering is commonly used
+% to facilitate distributed processing of data.
+
+There are many ways to address distributed data.  These include but
+are not limited to:
+% Ways to manage \katznote{?} the
+% distribution include but are not limited to:
 \begin{itemize}
-\item \textbf{Replication:} Data is duplicated on multiple nodes. When fully
-replicated, each execution unit has access to the same data. Replication can be
-partial, i.e., data can be replicated to a subset of resources. The degree of
-replication quantifies the average of the number of copies of each data
-item. Replication may be non-uniform, e.g., if only a subset of data (perhaps
-``hot'' data) is replicated. Caching is a lightweight form of replication.
-\item \textbf{Partitioning:} Data is divided and distributed across multiple
-nodes. Different forms of partitioning exist.  For example, data can be
-thematically partitioned with respect to its type, such as sensor data
-vs. simulated data, or data can be partitioned according to a spatial or
-temporal attributes. This can be an opaque (e.g., ID) and or a domain dependent
-(e.g., spatial) attribute.
-\item \textbf{Streaming}: A form of distributed data processing in which data
-arrives from a source; often the stream is a steady, continuous flow, but often
-it fluctuates both spatially and temporally. Also, typically, the stream of data
-must be processed in near real-time. Streaming thus also has attributes of
-dynamism, which is discussed in the next subsection.
+\item \textbf{Replication:} Data is duplicated on multiple nodes. When
+  fully replicated, each execution unit has access to the same
+  data. Replication can be partial, i.e., data can be replicated to a
+  subset of resources. The degree of replication quantifies the
+  average of the number of copies of each data item. Replication may
+  be non-uniform, e.g., if only a subset of data (perhaps ``hot''
+  data) is replicated. Caching is a lightweight form of replication.
+\item \textbf{Partitioning:} Data is divided and distributed across
+  multiple nodes. Different forms of partitioning exist.  For example,
+  data can be thematically partitioned with respect to its type, such
+  as sensor data vs. simulated data, or data can be partitioned
+  according to a spatial or temporal attributes. This can be an opaque
+  (e.g., ID) and or a domain dependent (e.g., spatial) attribute.
+\item \textbf{Streaming}: A form of distributed data processing in
+  which data arrives from a source; often the stream is a steady,
+  continuous flow, but often it fluctuates both spatially and
+  temporally. Also, typically, the stream of data must be processed in
+  near real-time. Streaming thus also has attributes of dynamism, which is
+  discussed in the next subsection.
 \end{itemize}
@@
 Which of the above approaches is employed cannot be reduced to a simple
-rule. Which option can or should be used often depends upon a plethora of
-co-dependent factors, such as the data volumes, the {\it degree of distribution}
-(which describes the number of data sources and sites used for storing and
-processing the data), the availability of compute resources, or the desired
-level of reliability. Data distribution can be difficult to manage for reasons
-of performance: latencies and bandwidths may vary unpredictably.  Further,
-issues associated with reliability may also need to considered. An example is
-the occurrence of a partial failure, where some (but not all) of the required
-data remains available, leads to applications having to decide whether to
-continue with reduced data or exit without processing what remains available.
-
-%\jhanote{I am not sure this is the best way to express co-dependence % of
-reliability and performance}\alnote{refined}
-
-Viewed from an application's perspective, there are at least two application
-characteristics that influence the distribution of data. The first is the stage
-and the second is the degree of coupling between components of a distributed
-application. Distribution in the first stage is primarily determined by the
-structure and layout of the data sources. For example, the placement of the data
-generated in the climate application (\S\ref{climateDan}) is based on which site
-was assigned to run the particular model that generates the data. But analysis
-of the data from multiple models requires bringing the data together to a single
-site.  Although it is difficult to formalize the difference due to the large
-fluctuations between different applications, but in general for a given
-application, there are differences between the distribution prior to the
-application's processing stage and the distribution of data within the
-processing stage, e.g., to aid data-parallel processing.
-
-% \jhanote{can we % provide an example or some further application discussion
-for the % previous point. it would help immensely}
-
-The second application factor that determines distribution of data is the type
-of separation and coupling of the components of the distributed
-application. Different coupling types place different constraints on data
-distribution, either due to latency, or due to data volumes that must be
-coordinated. % Different types of coupling in % turn influences the distribution
-of data.  \emph{Loosely-coupled} distribution involves coordination and
-cooperation between largely independent application components, as might be
-found in a workflow system. \emph{Tightly-coupled} applications involve greater
-constraints between the components, either in latency tolerance, coordination
-requirements or relative scheduling. \emph{Decoupled} distribution results in
-the separation of application components elements of the same task, for example,
-sharing a spatial data set between components that cooperate to perform a single
-logical task~\cite{dpa_surveypaper,thebookthatchangedtheworld}.  Most of the
-applications in \S\ref{sec:scenarios} are loosely coupled, with the CMB
-(\S\ref{astroJulian}) and Fusion (\S\ref{fusionScott}) applications as examples
-of tight coupling.
-
-% \jhanote{again, can we relate the above general paragraph to % specific
-application examples from the previous section} % \katznote{tried to come up
-with an application example, but instead, % realized that I was very confused by
-this paragraph.}\jhanote{Any % better? If you could be more specific that'll
-help}
+rule. Which option can or should be used often depends upon a plethora
+of co-dependent factors, such as the data volumes, the {\it degree of
+distribution} (which describes the number of data sources and sites used for
+storing and processing the data), the availability of compute resources, or the
+desired level of reliability. Data distribution can be difficult to manage for
+reasons of performance: latencies and bandwidths may vary unpredictably.
+Further, issues associated with reliability may also need to considered. An
+example is the occurrence of a partial failure, where some (but not all) of
+the required data remains available, leads to applications having to decide
+whether to continue with reduced data or exit without processing what remains
+available.
+
+%\jhanote{I am not sure this is the best way to express co-dependence
+%  of reliability and performance}\alnote{refined}
+
+Viewed from an application's perspective, there are at least two
+application characteristics that influence the distribution of
+data. The first is the stage and the second is the degree of coupling
+between components of a distributed application. Distribution in the
+first stage is primarily determined by the structure and layout of the
+data sources. For example, the placement of the data generated in the
+climate application (\S\ref{climateDan}) is based on which site was
+assigned to run the particular model that generates the data. But
+analysis of the data from multiple models requires bringing the data
+together to a single site.  Although it is difficult to formalize the
+difference due to the large fluctuations between different
+applications, but in general for a given application, there are
+differences between the distribution prior to the application's
+processing stage and the distribution of data within the processing
+stage, e.g., to aid data-parallel processing.
+
+% \jhanote{can we
+%  provide an example or some further application discussion for the
+%  previous point. it would help immensely}
+
+The second application factor that determines distribution of data is
+the type of separation and coupling of the components of the
+distributed application. Different coupling types place different
+constraints on data distribution, either due to latency, or due to
+data volumes that must be coordinated. % Different types of coupling in
+% turn influences the distribution of data.
+\emph{Loosely-coupled} distribution involves coordination and
+cooperation between largely independent application components, as
+might be found in a workflow system. \emph{Tightly-coupled}
+applications involve greater constraints between the components,
+either in latency tolerance, coordination requirements or relative
+scheduling. \emph{Decoupled} distribution results in the separation of
+application components elements of the same task, for example, sharing
+a spatial data set between components that cooperate to perform a
+single logical task~\cite{dpa_surveypaper,thebookthatchangedtheworld}.
+Most of the applications in \S\ref{sec:scenarios} are loosely coupled,
+with the CMB (\S\ref{astroJulian}) and Fusion (\S\ref{fusionScott})
+applications as examples of tight coupling.
+
+% \jhanote{again, can we relate the above general paragraph to
+%   specific application examples from the previous section}
+% \katznote{tried to come up with an application example, but instead,
+%   realized that I was very confused by this paragraph.}\jhanote{Any
+%   better? If you could be more specific that'll help}
 
@@
       \caption{Distributed Processing}\label{tab:app_distributed_processing}
-      \begin{tabular}{|p{3.5cm}|p{10.2cm}|} \hline \textbf{Application}
-&\textbf{Distributed Characteristics} \\
-% \endhead \hline Next Generation Sequencing (NGS) Analytics & Data sources (NGS
-machine storage) and compute not always co-located; multiple data sources in
-case of metagenomics \\ \hline ATLAS (an example of use of the WLCG) &Source
-data is generated by one central data source. Data needs to be distributed to
-match distribution of computing resources. During processing data is copied from
-closest replica to the resource where compute job is run. \\ \hline LSST
-&Multiple different kinds of data sources (observatories, data storage);
-multi-step workflow that processes and archives data, carried on distributed
-compute resources.\\ \hline SOA Astronomy Applications &Data resides in multiple
-data centers; services are placed closed to the data.\\ \hline Cosmic Microwave
-Background &Replication of subset of data to compute resources for processing.\\
-\hline Sensor Network Application & Data is moved from different data sources to
-a central analysis system for processing\\ \hline Climate (ESGF) &A subset of
-the data is commonly moved to compute (desktop or grid resource)\\ \hline
-Interactive Exploration of Environmental Data &Data is brought to grid resource
-or desktop \\ \hline PowerGrids &Data is brought to central private cloud for
-analysis\\ \hline Fusion (ITER) &Simulations run wherever resources are
-available; in-situ analysis of large output data; partitioning/replication of
-data for data-parallel processing\\ \hline Industrial Incident Notification and
-Response &Distributed data sources. Processing takes place on multiple
-distributed resources. \\ \hline MODIS Data Processing & Data is copied
-(replicated) from different sources for processing.  Processing can take place
-in different locations, e.\,g.\ cloud and user desktop. \\ \hline Distributed
-Network Intrusion Detection &Distributed data and compute resources.\\ \hline
+      \begin{tabular}{|p{3.5cm}|p{10.2cm}|}
+		\hline
+		\textbf{Application}  &\textbf{Distributed Characteristics} \\		
+%		 \endhead
+		\hline Next Generation Sequencing (NGS) Analytics &
+                Data sources (NGS machine storage) and compute not
+                always co-located; multiple data sources in case of
+                metagenomics
+		\\
+		\hline ATLAS (an example of use of the WLCG) &Source
+                data is generated by one central data source. Data
+                needs to be distributed to match distribution of
+                computing resources. During processing data is copied
+                from closest replica to the resource where
+		compute job is run. \\
+		\hline LSST &Multiple different kinds of data sources
+                (observatories, data storage); multi-step workflow
+                that processes and archives data,
+		carried on distributed compute resources.\\
+		\hline SOA Astronomy Applications &Data resides in
+                multiple data centers; services are placed closed to
+		the data.\\
+		\hline Cosmic Microwave Background &Replication of
+                subset of data
+		to compute resources for processing.\\
+		\hline Sensor Network Application & Data is moved from
+                different data sources to a central analysis
+		system for processing\\
+		\hline Climate (ESGF)
+                &A subset of the data is commonly moved to compute (desktop or grid resource)\\
+		\hline Interactive Exploration of Environmental Data
+		&Data is brought to grid resource or desktop \\
+		\hline PowerGrids
+		&Data is brought to central private cloud for analysis\\
+		\hline Fusion (ITER) &Simulations run wherever
+                resources are available; in-situ analysis of large
+                output data; partitioning/replication of data for
+		data-parallel processing\\
+		\hline Industrial Incident Notification and Response
+                &Distributed data
+                sources. Processing takes place on multiple distributed resources. \\
+		\hline MODIS Data Processing & Data is copied
+                (replicated) from different sources for processing.
+                Processing can take place in different locations,
+                e.\,g.\ cloud and
+		user desktop. \\
+		\hline
+                Distributed Network Intrusion Detection  &Distributed data and compute resources.\\
+		\hline
      \end{tabular}
@@
 
-% \katznote{without the table that was here, we are now defining % parameters
-($\alpha$, $\beta$) that we don't really do anything % with.  We should think
-about what part(s) of this subsection make % sense to keep.} \jhanote{the tables
-have been moved (and reference % to them fixed). Admittedly we don't use
-$\alpha$ or $\beta$ % again..}  \katznote{then why do we define them?}
+% \katznote{without the table that was here, we are now defining
+%   parameters ($\alpha$, $\beta$) that we don't really do anything
+%   with.  We should think about what part(s) of this subsection make
+%   sense to keep.} \jhanote{the tables have been moved (and reference
+%   to them fixed). Admittedly we don't use $\alpha$ or $\beta$
+%   again..}  \katznote{then why do we define them?}
 
@@
     \begin{center}
-      \caption{Understanding Dynamic Characteristics. This table highlights
-different types and scenarios for dynamic data in the application set
-investigated %\katznote{some explanation %here might
-help}\jhanote{OK?}\katznote{yes} }
+      \caption{Understanding Dynamic Characteristics. This table
+        highlights different types and scenarios for dynamic data in
+        the application set investigated %\katznote{some explanation
+          %here might help}\jhanote{OK?}\katznote{yes}
+          }
           \label{tab:app_dyn_char}
-      \begin{tabular}{|p{4cm}|p{9.5cm}|} \hline \textbf{Application}
-&\textbf{Dynamic Characteristics} \\
-% \endhead \hline Next Generation Sequencing (NGS) Analytics & Data Dynamism:
-different data formats and sources need to be processed\\ \hline ATLAS (an
-Example of Use of the WLCG) & Data Dynamism: new raw data is steadily
-collected\\ \hline LSST & Infrastructural Dynamism: Decision making based on
-observed data\\ \hline SOA Astronomy Applications & Application and Data
-Dynamism: data is continuously updated, dynamic queries\\ \hline Cosmic
-Microwave Background Mapping & Application Dynamism: dynamic utilization of
-resources\\ \hline Sensor Network Application & Data Dynamism: rate and volumes
-of data dynamic \\ \hline Climate (ESGF) & Data \&Infrastructural Dynamism: new
-data is generated over time. Dynamic distributed infrastructure for storing and
-processing of data %\jhanote{please check} \\ \hline Interactive Exploration of
-Environmental Data & Infrastructural and Data Dynamism: Various data sources,
-varying input rates and different queries.\\ \hline PowerGrids & Application,
-Infrastructural and Data Dynamism: data itself changes, data rates may change
-(throttling), data sources change (traffic, weather, social networking), data
-queries respond to seen data (e.g.  accuracy adjustment) \\ \hline Fusion (ITER)
-& Application and Infrastructural Dynamism: data streamed between simulations
-(multiple distinct data flows) \\ \hline Industrial Incident Notification and
-Response & Application and Data Dynamism: data itself and data rate changes,
-dynamic data sources, processing changes with data\\ % \jhanote{Omer needs to
-clarify if application %dynamism too?} \\ \hline MODIS Data Processing &
-Application and Data Dynamism: data that is operated upon changes,
-infrastructure config changes\\ \hline Distributed Network Intrusion Detection &
-Infrastructural, Application and Data Dynamism: data highly dynamically
-changing, data rate variable, application responds to data, on-the-flight
-processing\\ \hline
+      \begin{tabular}{|p{4cm}|p{9.5cm}|}
+		\hline
+		\textbf{Application} &\textbf{Dynamic Characteristics} \\		
+%		 \endhead
+		\hline		
+		Next Generation Sequencing (NGS) Analytics
+		& Data Dynamism: different data formats and sources need to be processed\\
+		\hline
+                ATLAS (an Example of Use of the WLCG)
+		& Data Dynamism: new raw data is steadily collected\\
+		\hline
+                LSST & Infrastructural Dynamism: Decision making based on observed data\\
+		\hline
+                SOA Astronomy Applications
+		&  Application and Data Dynamism: data is continuously updated, dynamic queries\\
+		\hline
+                Cosmic Microwave Background Mapping
+		& Application Dynamism: dynamic utilization of resources\\
+		\hline
+                Sensor Network Application
+                & Data Dynamism: rate and volumes of data dynamic \\
+		\hline
+                Climate (ESGF) & Data \&Infrastructural Dynamism: new data is generated over time. Dynamic distributed infrastructure for storing and processing of data
+                %\jhanote{please check}
+                \\
+		\hline
+                Interactive Exploration of Environmental Data
+		& Infrastructural and Data Dynamism: Various data sources, varying
+		input rates and different queries.\\
+		\hline
+                PowerGrids
+		& Application, Infrastructural and Data Dynamism: data itself changes,
+		data rates may change (throttling), data sources change (traffic,
+		weather, social networking), data queries respond to seen data (e.g.
+		accuracy adjustment) \\
+		\hline
+                Fusion (ITER)
+                & Application and Infrastructural Dynamism: data streamed
+				between simulations (multiple distinct data flows)  \\
+		\hline
+                Industrial Incident Notification and Response & Application
+				and Data Dynamism: data itself and data rate changes, dynamic
+				data sources, processing changes with data\\ % \jhanote{Omer needs to clarify if application
+                  %dynamism too?} \\
+		\hline
+                MODIS Data Processing
+		& Application and Data Dynamism: data that is operated upon changes,
+		infrastructure config changes\\
+		\hline
+                Distributed Network Intrusion Detection &
+                Infrastructural, Application and Data Dynamism: data highly
+				dynamically changing, data rate variable,
+				application responds to data, on-the-flight processing\\
+		\hline
      \end{tabular}
@@
 
-% \jhanote{Modify this paragraph to reference table 1 and discuss the % kinds of
-dynamism} There are often spatio-temporal variability in the values and
-volumes~\cite{deRoos2011UnderstandingBigData} of data.  The applications in
-\S\ref{sec:scenarios} have variability in the production/generation rate of
-data, and in the configuration of data production/generation.  Data dynamism may
-arise from extrinsic factors, for example because sensors begin to observe more
-events in the same data stream. Contrariwise, data dynamism may occur because an
-application component requested an increased sampling frequency.  Similarly
-application dynamism may occur through managed replication of a
-compute-intensive component or from the demands of a particularly complex
-query. And infrastructure dynamism may arise because the infrastructure expands
-to meet additional requirements, or contracts because of device failure.  A
-structured analysis of the applications in \S\ref{sec:scenarios} suggests that
-the sources of dynamism can be classified into three categories.
+% \jhanote{Modify this paragraph to reference table 1 and discuss the
+%   kinds of dynamism}
+There are often spatio-temporal variability in the values and
+volumes~\cite{deRoos2011UnderstandingBigData} of data.  The
+applications in \S\ref{sec:scenarios} have variability in the
+production/generation rate of data, and in the configuration of data
+production/generation.  Data dynamism may arise from extrinsic
+factors, for example because sensors begin to observe more events in
+the same data stream. Contrariwise, data dynamism may occur because an
+application component requested an increased sampling frequency.
+Similarly application dynamism may occur through managed replication
+of a compute-intensive component or from the demands of a particularly
+complex query. And infrastructure dynamism may arise because the
+infrastructure expands to meet additional requirements, or contracts
+because of device failure.  A structured analysis of the applications
+in \S\ref{sec:scenarios} suggests that the sources of dynamism can be
+classified into three categories.
 \begin{enumerate}
 
-\item \emph{Data} dynamism arises when some property of the input data or its
-delivery changes, for example in terms of arrival rate, provenance, burstiness,
-or source.  There may be variability in the structure of the data, for example,
-data schema, file formats, ontologies, etc.
-
-  % \alnote{Not sure, but maybe changes in the structure of data would % better
-fit into category 1 or 3 since the structure is typically % not related to the
-infrastructure}\ysnote{I agree with Andre.}  % \jhanote{In order to % be
-consistent with paragraph above, the third type of dynamism % should focus on
-the infrastructure -- resources not on data any % more.}
-
-\item % Dynamism in the Application, % Software/Services use of the data:
-\emph{Application} dynamism involves changing the processes or components
-applied to a data set, and might encapsulate changes in workflow, \textit{ad
-hoc} queries, or parameterization.  There may be variability in the data queries
-enacted by the application, or in the behavior and/or characteristics of the
-application.  % \ysnote{this may affect % data access as well as generation of
-derived data}
-
-\item \emph{Infrastructure} dynamism occurs when the system wants to, or is
-forced to, change its demands on its underlying platform, for example through
-elastic computation or partial failure.  Changes in the structure and load of
-the infrastructure can be associated with data.  Examples include variability in
-the producer/provenance of the data, in the performance, and in the quality of
-service of the infrastructure.
+\item \emph{Data} dynamism arises when some property of the input data
+  or its delivery changes, for example in terms of arrival rate,
+  provenance, burstiness, or source.  There may be variability in the
+  structure of the data, for example, data schema, file formats,
+  ontologies, etc.
+
+  % \alnote{Not sure, but maybe changes in the structure of data would
+  %   better fit into category 1 or 3 since the structure is typically
+  %   not related to the infrastructure}\ysnote{I agree with Andre.}
+  % \jhanote{In order to
+  %   be consistent with paragraph above, the third type of dynamism
+  %   should focus on the infrastructure -- resources not on data any
+  %   more.}
+
+\item  % Dynamism in the Application,
+  % Software/Services use of the data:
+  \emph{Application} dynamism
+  involves changing the processes or components applied to a data set,
+  and might encapsulate changes in workflow, \textit{ad hoc} queries,
+  or parameterization.  There may be variability in the data queries
+  enacted by the application, or in the behavior
+  and/or characteristics of the
+  application.  % \ysnote{this may affect
+%     data access as well as generation of derived data}
+
+\item \emph{Infrastructure} dynamism occurs when the system wants to,
+  or is forced to, change its demands on its underlying platform, for
+  example through elastic computation or partial failure.  Changes in
+  the structure and load of the infrastructure can be associated with data.
+  Examples include variability in the producer/provenance of the data,
+  in the performance, and in the quality of service of the
+  infrastructure.
 
@@
 Based upon this classification, in \tablename~\ref{tab:app_dyn_char}, we
-describe the applications discussed in \S\ref{sec:scenarios}.  Many of these
-dynamic aspects are related: an increase in the rate of data generation may lead
-an application to increase its degree of parallelism to maintain throughput,
-which may in turn trigger the allocation of new compute nodes. (In the other,
-less attractive direction, decreasing computational capability might cause a
-decrease in parallelism and a consequent need to drop data. Avoiding these cases
-means treating dynamism equally in both directions.)
-
-This should not be construed to imply that every data-intensive application has
-all elements of data and compute dynamism and is always distributed, but is a
-reflection that at extreme scales, dynamism and distribution become increasingly
-important issues that are often correlated.  As alluded to, a high degree of
-distribution often results in a high amount of infrastructure dynamism, %
-\katznote{I don't really like this forward-looking link, or the one in % the
-next paragraph}\jhanote{Not forward looking anymore} caused by, for example,
-resource fluctuations and a higher failure probability.  Similarly, high levels
-of dynamism, for example, in the data sources, data rates, and queries, usually
-correlate with a high degree of distribution.  For example, the Interactive
-Exploration of Environmental Data application (\S\ref{envJon}) manages various
-sources of data, from archived data sets to real-time feeds.  Depending on the
-spatiotemporal properties of data, computation can be carried out on a set of
-distributed grid resources or the user's desktop, i.e., distributed processing
-is used to efficiently handle dynamic data.
+describe the applications discussed in \S\ref{sec:scenarios}.  Many of
+these dynamic aspects are related: an increase in the rate of data
+generation may lead an application to increase its degree of
+parallelism to maintain throughput, which may in turn trigger the
+allocation of new compute nodes. (In the other, less attractive
+direction, decreasing computational capability might cause a decrease
+in parallelism and a consequent need to drop data. Avoiding these
+cases means treating dynamism equally in both directions.)
+
+This should not be construed to imply that every data-intensive
+application has all elements of data and compute dynamism and is
+always distributed, but is a reflection that at extreme scales,
+dynamism and distribution become increasingly important issues that
+are often correlated.  As alluded to, a high degree of distribution
+often results in a high amount of infrastructure dynamism,
+% \katznote{I don't really like this forward-looking link, or the one in
+%   the next paragraph}\jhanote{Not forward looking anymore}
+caused by, for example, resource fluctuations and a higher failure
+probability.  Similarly, high levels of dynamism, for example, in the
+data sources, data rates, and queries, usually correlate with a high
+degree of distribution.  For example, the Interactive Exploration of
+Environmental Data application (\S\ref{envJon}) manages various
+sources of data, from archived data sets to real-time feeds.
+Depending on the spatiotemporal properties of data, computation can be
+carried out on a set of distributed grid resources or the user's
+desktop, i.e., distributed processing is used to efficiently handle
+dynamic data.
 
@@
 
-%\jhanote{needs text to establish that this is the end of section 3} \katznote{I
-think this should be a short subsection, as I've done here - can be removed...}
-
-Reviewing \tablename{s}~\ref{tab:app_distributed_processing} and
-\ref{tab:app_dyn_char}, which have summarized the distributed and dynamic
-properties of the D3 applications, we observe:
-
-%find that commonly re-occurring patterns %\jhanote{should we be calling the
-following patterns? I don't think % they are patterns, but possibly
-observations?} \katznote{yes, `observations'} in the data %distribution and
-processing are:
+%\jhanote{needs text to establish that this is the end of section 3} \katznote{I think this should be a short subsection, as I've done here - can be removed...}
+
+Reviewing \tablename{s}~\ref{tab:app_distributed_processing} and \ref{tab:app_dyn_char},
+which have summarized the distributed and dynamic properties of the D3
+applications, we observe:
+
+%find that commonly re-occurring patterns
+%\jhanote{should we be calling the following patterns? I don't think
+%  they are patterns, but possibly observations?} \katznote{yes, `observations'} in the data
+%distribution and processing are:
 
 \begin{itemize}
-\item Data generation and consumption are commonly distributed from each other
-if the volume permits.  The Fusion application is a counterexample, due to data
-volumes and velocity.
-
-\item Details of the storage system are often not known. Often data is stored in
-a distributed way, e.g., on a distributed file system, which commonly fits into
-the medium degree of distribution (the same data center).
-
-%\item Mapping: The most common pattern is to copy data to an available %
-compute resource. \katznote{I'm not sure about this bullet: 1. It's not a
-sentence; % 2. I don't know what it's saying.}
-
-\item Aggregation of data can either be the equivalent of gather or reduce
-(using the language of collective communication). Gather is in many ways the
-reverse process of distribution (called broadcast in the language of collective
-communication). Similar to distribution, aggregation is a commonly occurring
-pattern for data analysis.
+\item Data generation and consumption are commonly distributed from
+  each other if the volume permits.  The Fusion application is a counterexample,
+  due to data volumes and velocity.
+
+\item Details of the storage system are often not known. Often data is
+  stored in a distributed way, e.g., on a distributed file system,
+  which commonly fits into the medium degree of distribution (the same
+  data center).
+
+%\item Mapping: The most common pattern is to copy data to an available
+%  compute resource. \katznote{I'm not sure about this bullet:  1. It's not a sentence;
+%  2. I don't know what it's saying.}
+
+\item Aggregation of data can either be the equivalent of gather or
+reduce (using the language of collective communication). Gather is
+in many ways the reverse process of distribution (called broadcast in the
+language of collective communication). Similar to distribution, aggregation is a
+commonly occurring pattern for data analysis.
 
@@
 
-%\subsubsection{Characterizing Dynamic Data \label{sec:dynDataChar}} %
-\jhanote{AL: Next sentences OK to go?} Also, in lieu of quantitative %
-estimates, we attempt classification of these ratios into categories % of
-high/medium/low.  It is natural to expect that applications will % cluster
-around certain {\it trends}.
-
-% \jhanote{AL/SJ: The following should be deleted?} As a guide to how % the
-metrics proposed may be useful, we characterize a couple of the % applications
-investigated using $\alpha$ and $\beta$: % \katznote{actually, neither the
-examples here nor the table mention % $\alpha$}
+%\subsubsection{Characterizing Dynamic Data \label{sec:dynDataChar}}
+% \jhanote{AL: Next sentences OK to go?} Also, in lieu of quantitative
+% estimates, we attempt classification of these ratios into categories
+% of high/medium/low.  It is natural to expect that applications will
+% cluster around certain {\it trends}.
+
+% \jhanote{AL/SJ: The following should be deleted?} As a guide to how
+% the metrics proposed may be useful, we characterize a couple of the
+% applications investigated using $\alpha$ and $\beta$:
+% \katznote{actually, neither the examples here nor the table mention
+%   $\alpha$}
 
 % \begin{itemize}
-% \item NGS: $\beta_1$ is high, $\beta_2$ is high. Degree of % Distribution is
-Medium. The number of distributed sites does not % scale with data.
-
-% \item Atlas: $\beta_1$ is Medium, $\beta_2$ is Medium. Degree of %
-Distribution is High. The number of distributed sites scales with % data.
+% \item NGS: $\beta_1$ is high, $\beta_2$ is high. Degree of
+%   Distribution is Medium. The number of distributed sites does not
+%   scale with data.
+
+% \item Atlas: $\beta_1$ is Medium, $\beta_2$ is Medium. Degree of
+%   Distribution is High. The number of distributed sites scales with
+%   data.
 % \end{itemize}
 
-% \jhanote{Is LSST like NGS or Atlas or in its own class?}  % \jhanote{Is
-PowerGrid more like NGS or Atlas or in its own class?}  % \note{metrics for
-distribute data} \jhanote{No idea who put the % previous note there or why? Help
-please?}
-
-% \jhanote{not sure where this goes?} replicated $\rightarrow$ average %
-replication factor of a data element/N ; coupled $\rightarrow$ M/N
-
-% \jhanote{Not sure this is correct?}  % Degree of distribution: latency between
-primary data source and data % processing element
-
-% \alnote{Refine description, define view point} \note{Is the control of % the
-sensors from the application a case of 5 or is it something % different?
-Feedback to the data source (other areas, sources of % data, accuracy)?}
-\note{most applications commonly operate in a % local setting, but can also
-operate in a distributed settings.}  % \alnote{How to measure degree of
-distribution?} \jhanote{these notes % should move}
+% \jhanote{Is LSST like NGS or Atlas or in its own class?}
+% \jhanote{Is PowerGrid more like NGS or Atlas or in its own class?}
+% \note{metrics for distribute data} \jhanote{No idea who put the
+%   previous note there or why? Help please?}
+
+% \jhanote{not sure where this goes?} replicated $\rightarrow$ average
+% replication factor of a data element/N ; coupled $\rightarrow$ M/N
+
+% \jhanote{Not sure this is correct?}
+% Degree of distribution: latency between primary data source and data
+% processing element
+
+% \alnote{Refine description, define view point} \note{Is the control of
+%   the sensors from the application a case of 5 or is it something
+%   different? Feedback to the data source (other areas, sources of
+%   data, accuracy)?}  \note{most applications commonly operate in a
+%   local setting, but can also operate in a distributed settings.}
+% \alnote{How to measure degree of distribution?} \jhanote{these notes
+%   should move}
 % \begin{table}[h!]
@@
 %     \begin{center}
-% \caption{Dynamic and Distributed Characteristics of Example
-Applications} \label{tab:app_characteristics}
-% \begin{tabular}{|p{3cm}|p{5.0cm}|p{5.1cm}|} % %\begin{footnotesize}
+%       \caption{Dynamic and Distributed Characteristics of Example Applications} \label{tab:app_characteristics}
+%       \begin{tabular}{|p{3cm}|p{5.0cm}|p{5.1cm}|}
+% %\begin{footnotesize}
 % %% 	\begin{tabular}{|p{4cm}|p{5cm}|p{5cm}|}
-% % \begin{longtable}[h!]{|p{3cm}|p{5.0cm}|p{5.1cm}|} % % \caption{Dynamic and
-Distributed Characteristics of Example
-Applications} \label{tab:app_characteristics} \\ % \hline % \textbf{Application}
-&\textbf{Dynamic Properties} &\textbf{Distributed Properties} \\ % % \endhead %
-\hline % Next Generation Sequencing (NGS) Analytics & Infrastructure dynamics
-(2) % &degree-of-distribution: medium; $\beta_1$: high; $\beta_2$: high \\ %
-\hline % ATLAS (an example of use of the WLCG) &New raw data is collected %
-steadily (1) % &degree-of-distribution: high; $\beta_1$: medium; $\beta_2$:
-medium \\ % \hline % LSST &DE decisions based on observed data (3)
-&degree-of-distributed: % medium, multiple different kinds of data sources
-(observatories, data % storage); $\beta_1$: ?; $\beta_2$: ? \\ % \hline % SOA
-Astronomy Applications &data is continuously updated (1), thus % dynamic data
-(pre-defined schedules or on-demand), queries change (3) %
-&degree-of-distribution: high, multiple kinds of data sources\\ % \hline %
-Cosmic Microwave Background &dynamic utilization of resources (2 or 3?)  %
-&degree-of-distribution: low--medium\\ % \hline % Sensor Network Application
-&data rate, volume dynamic (1) % & degree-of-distribution: high, multiple kinds
-of data sources\\ % \hline % Climate (ESGF) &New data generated over time
-(ESGF-wide - is this in % scope of the application?) (1)
-&degree-of-distribution: high, multiple % kinds of data sources\\ % \hline %
-Interactive Exploration of Environmental Data & data rates and % data
-inputs/outputs change (1), queries change (3) % & degree-of-distribution: high
-\\ % \hline % % Dynamic Environmental Data on the Web &data operated on changes
-% % (realtime feeds) (2) % % &natural (sensor sources, simulations), (sub-set)
-replication\\ % % \hline % PowerGrids &data itself changes, % data rates may
-change (throttling) (1), % data sources change (traffic, weather, social %
-networking) and infrastructure dynamics (2), response to data (e.g. accuracy
-adjustment) (3) % &degree-of-distribution: high, large number of sensors;
-$\beta_1$: medium; % $\beta_2$: medium \\ % \hline % Fusion (ITER) &data
-streamed between simulations (multiple distinct data % flows) (1), on-the-fly
-transformation (3) % &degree-of-distribution: medium\\ % \hline % Industrial
-Incident Notification and Response &data itself and % data rate changes (1),
-dynamic data sources (2), processing changes % based on data (3)
-&degree-of-distribution: high\\ % \hline % MODIS Data Processing & dataset that
-is operated upon changes (1), % infrastructure config changes (7) %
-&degree-of-distribution: high, multiple data sources\\ % \hline % % Floating
-Sensors &data itself (1), data rate (2), application causes % % data changes (5)
-% % &degree-of-distribution: high, multiple kinds of data sources; % %
-aggregation \\ % % \hline % Distributed Network Intrusion Detection &data %
-highly dynamically changing, data rate variable (1), application % responds to
-data, on-the-flight processing (3) % &degree-of-distribution: high\\ % \hline %
-% % \end{table} % % \end{longtable} % %% \end{tabular} % %\end{footnotesize}
+% %	 \begin{longtable}[h!]{|p{3cm}|p{5.0cm}|p{5.1cm}|}
+% %		\caption{Dynamic and Distributed Characteristics of Example Applications} \label{tab:app_characteristics} \\
+% 		\hline
+% 		\textbf{Application} &\textbf{Dynamic Properties} &\textbf{Distributed Properties} \\		
+% %		 \endhead
+% 		\hline		
+% 		Next Generation Sequencing (NGS) Analytics & Infrastructure dynamics (2)
+% 		&degree-of-distribution: medium; $\beta_1$: high; $\beta_2$: high  \\
+% 		\hline
+% 	    ATLAS (an example of use of the WLCG) &New raw data is collected
+% 	    steadily (1)
+% 		&degree-of-distribution: high; $\beta_1$: medium; $\beta_2$: medium  \\
+% 		\hline
+% 	    LSST &DE decisions based on observed data (3) &degree-of-distributed:
+% 		medium, multiple different kinds of data sources (observatories, data
+% 		storage); $\beta_1$: ?; $\beta_2$: ? \\
+% 		\hline
+% 	    SOA Astronomy Applications &data is continuously updated (1), thus
+% 	    dynamic data (pre-defined schedules or on-demand), queries change (3)
+% 		&degree-of-distribution: high, multiple kinds of data sources\\
+% 		\hline
+% 	    Cosmic Microwave Background &dynamic utilization of resources (2 or 3?)
+% 		&degree-of-distribution: low--medium\\
+% 		\hline
+% 	    Sensor Network Application &data rate, volume dynamic (1)
+% 	    & degree-of-distribution: high, multiple kinds of data sources\\
+% 		\hline
+% 	    Climate (ESGF) &New data generated over time (ESGF-wide - is this in
+% 		scope of the application?) (1) &degree-of-distribution: high, multiple
+% 		kinds of data sources\\
+% 		\hline
+% 	    Interactive Exploration of Environmental Data  & data rates and
+% 	    data inputs/outputs change (1), queries change (3)
+% 		& degree-of-distribution: high \\
+% 		\hline
+% 	    % Dynamic Environmental Data on the Web &data operated on changes
+% 	    %     (realtime feeds) (2)
+% 	    %     &natural (sensor sources, simulations), (sub-set) replication\\
+% 	    %     \hline
+% 	    PowerGrids &data itself changes,
+% 					data rates may change (throttling) (1),
+% 	    data sources change (traffic, weather, social
+% 		networking) and infrastructure dynamics (2), response to data (e.g. accuracy adjustment) (3)
+% 		&degree-of-distribution: high, large number of sensors; $\beta_1$: medium;
+% 		$\beta_2$: medium \\
+% 		\hline
+% 	    Fusion (ITER) &data streamed between simulations (multiple distinct data
+% 	    flows) (1), on-the-fly  transformation (3)
+% 	    &degree-of-distribution: medium\\
+% 		\hline
+% 	    Industrial Incident Notification and Response &data itself and
+% 	    data rate changes (1), dynamic data sources (2), processing changes
+% 	    based on data (3) &degree-of-distribution: high\\
+% 		\hline
+% 	    MODIS Data Processing & dataset that is operated upon changes (1),
+% 		infrastructure config changes (7)
+% 		&degree-of-distribution: high, multiple data sources\\
+% 		\hline
+% %	    Floating Sensors &data itself (1), data rate (2), application causes
+% %	    data changes (5)
+% %		&degree-of-distribution: high, multiple kinds of data sources;
+% %		aggregation \\
+% %		\hline
+% 	    Distributed Network Intrusion Detection &data
+% 	    highly dynamically changing, data rate variable (1),  application
+% 		responds to data, 	  on-the-flight processing (3)
+%  		&degree-of-distribution: high\\
+% 		\hline
+% %		% \end{table}
+% %	 \end{longtable}
+% %% 	\end{tabular}
+% %\end{footnotesize}
 %
@@
 
-%\alnote{relate infrastructures back to applications: in particular, how do
-specific purpose built infrastructures as the LHCG or ESGF fit here?}
-
-We define {\em infrastructure} as the hardware and software that is provided to
-support an application, rather than being explicitly created or hosted by the
-user themselves. In many cases the precise boundary between infrastructure and
-application is blurred, e.g., a catalog is a component with functionality that
-may be similar to that of similar components in other applications, but is often
-implemented specifically by an application rather than being provided as part of
+%\alnote{relate infrastructures back to applications: in particular, how do specific purpose built infrastructures as the LHCG or ESGF fit here?}
+
+We define {\em infrastructure} as the hardware and software that is
+provided to support an application, rather than being explicitly
+created or hosted by the user themselves. In many cases the precise boundary between
+infrastructure and application is blurred, e.g., a catalog is
+a component with functionality that may be similar to that of similar
+components in other applications, but is often implemented
+specifically by an application rather than being provided as part of
 the infrastructure. This may also be the case when reconfiguring an
 infrastructure service for an application, when it is more onerous than
-reimplementing the service specifically for that application. We also introduce
-the concept of an ``infrastructure'' application as a collection of
-infrastructure components that are packaged so that they can be used by many
-applications in defined ways. The difference between these and ``traditional''
-applications is often delineated by who owns and administers the application,
-with traditional applications ``owned'' entirely by the user, and infrastructure
-applications owned by someone other than the user but provided for use by
-others. Infrastructure can be specialist as opposed to generic, for instance the
-purpose-built infrastructure in ESGF (\S\ref{climateDan}) that is provided to
-support the climate community.
-
-In this section, we focus on the {\em software} infrastructure that is provided
-to support D3 applications, specifically components that deal with distributed
-or dynamic data. The components that make up the infrastructure can be
-categorized based on the function provided, namely (i) infrastructure that
-supports {\it data management}, (ii) infrastructure that supports {\it data
-  analysis}, and (iii) infrastructure that supports the \emph{coordination
-  within applications}. Each of these components may be present at three
-different levels of the infrastructure: programming frameworks, services and
-tools (e.g., MapReduce, workflow engines); middleware and platform services
-(e.g., databases, message queues); and system-level software capabilities (e.g.,
-notifications, file system consistency).
+reimplementing the service specifically for that
+application. We also introduce the concept of an ``infrastructure'' application as a collection
+of infrastructure components that are packaged so that they can be used by many
+applications in defined ways. The difference between these and ``traditional'' applications
+is often delineated by who owns and administers the application, with traditional applications
+``owned'' entirely by the user, and infrastructure applications owned by someone other than the
+user but provided for use by others. Infrastructure can be specialist as opposed to generic, for instance
+the purpose-built infrastructure in ESGF (\S\ref{climateDan}) that is provided to support the climate community.
+
+In this section, we focus on the {\em software} infrastructure that is
+provided to support D3 applications, specifically components that deal
+with distributed or dynamic data. The components that make up the infrastructure can be categorized
+based on the function provided, namely (i) infrastructure that supports {\it data management},
+(ii) infrastructure that supports {\it data analysis}, and (iii) infrastructure that supports the \emph{coordination within applications}.
+Each of these components may be present at three different levels of the infrastructure: programming frameworks, services and tools (e.g., MapReduce, workflow engines); middleware and platform services (e.g., databases, message queues); and system-level
+software capabilities (e.g., notifications, file system consistency).
 
@@
 
-% \katznote{we should think about how the items in this section map to the boxes
-and arrows in Figure~\ref{fig:figures_application-stages}.  For data sources,
-storage, access, and movement, it's fairly clear.  For discovery and
-notification, it's less clear.}
-
-{\em Data Sources} are used in the acquisition stage as inputs of data into the
-application. Broadly speaking these sources include types such as files, servers,
-sensors, instruments, simulations and models.  Additionally, databases are often
-considered as an intermediate data source in types of system. The dynamic
-characteristics are primarily due to availability. These are often accessed in
-the Data Generation stage of the application
+% \katznote{we should think about how the items in this section map to the boxes and arrows in Figure~\ref{fig:figures_application-stages}.
+For data sources, storage, access, and movement, it's fairly clear.
+For discovery and notification, it's less clear.}
+
+{\em Data Sources} are used in the acquisition stage as inputs of data
+into the application. Broadly speaking these sources include types such as
+files, servers, sensors, instruments, simulations and models.
+Additionally, databases are often considered as an intermediate data
+source in some systems. The dynamic characteristics are primarily
+due to availability. These are often accessed in the Data Generation stage of the application
 (\figurename~\ref{fig:figures_application-stages}).
 
-{\em Data Storage} components are used to provide persistent access to data by
-the application. They may be structured (e.g., databases, No-SQL tables) or
-unstructured (e.g., collections of audio recordings). The dynamic
-characteristics are primarily due to the level of persistence, e.g., temporary,
-durable, permanent (archived or published), or replicated; and also in the
-structure of the data stored, e.g., schema free tables as opposed to relational
-tables. There may also be dynamic characteristics of storage due to rate of
-growth of data, e.g., a field survey may bring in a lot of data during summer
-but none during other times.  These are commonly used in the Data Storage stage
-of the application (\figurename~\ref{fig:figures_application-stages}), but may
-also be present at the boundaries between the stages for transient staging of
-data.
-
-{\em Data Access} components are gateways to Data Sources and Data Storage
-components, and provide interfaces that often enforce security controls to
-restrict access, provide dynamic information about the available data sources
-and/or data analysis components, and may also perform logging of requests and
-data movement (e.g., tracking provenance for auditing). A side effect of
-auditing/logging may be the growth of provenance metadata simply due to access
-transactions even if no new data is created. Data Access components usually
-serve at the interface between the application stages, before and after data
-movement.
-
-{\em Data Movement} components are used for bulk or continuous movement of data
-between parts of the application, or to and from the application. % (often
-called staging) Data movement may be explicit and visible to the application
-(e.g., staging), or implicit (e.g., automatic replication of files). Components
-may operate on streams or blocks (files). Dynamic characteristics include
-partitioning, scheduling and planning (e.g., edge caching of ``hot'' data, where
-datasets that are being used frequently during some period are replicated close
-to the application). Data movement can be considered to be real-time (i.e., as
-it happens, e.g., sensor data), as real-time (e.g., a storage buffer has been
-introduced into the transport system, but the data source is still real-time, as
-with time-series data), or asynchronous (where the ordering of the data movement
-is independent of the data itself).  Latency may be considered as an important
-characteristic of the data movement infrastructure depending on the application
-e.g. real-time processing of sensor data for disaster analysis. Other dynamic
-aspects of data movement include data rates, their variability from data
-sources, and the ``freshness'' of data. These infrastructure appear on the edges
-between the application stages.
-
-{\em Data Discovery} components assist the access to and use of data sources,
-and include Catalogs, Information Services, and Metadata stores. The dynamic
-characteristics pertain to the evolution or recalibration of information held in
-these components. These components are orthogonal to the applications stages and
-can be used by any of the stages to guide their selection or identification of
-incoming data sources and outgoing data sinks.  %\ysnote{rate of growth of data
-may be an issue.  %Catalogs that perform nightly ``crawling'' or data loading
-(e.g., sky surveys) may not be suitable if the %latest data needs to be
-accessible to applications. Catalog update will have to keep up with data
-%update. However, batch processing applications can survive less frequent
-updates.}
-
-{\em Notification} components respond to changes in the application or data and
-trigger other functionality. These include publish-subscribe message brokers,
-and complex event processing. By nature, these components support the dynamic
-aspects of an application.  These components again are orthogonal to the
-applications stages and can be used for coordination of the stages and data
-movement.
-
-In each case, these components can be single or centralized, distributed,
-partitioned or federated. These are also typically offered as middleware and
-platform services, or system-level software capabilities.
-
-%\nchnote{where do we put data publishing? Calibration? provenance?}\ysnote{They
-can be incorporated in data discovery and data access}
+{\em Data Storage} components are used to provide persistent access to
+data by the application. They may be structured (e.g., databases,
+No-SQL tables) or unstructured (e.g., collections of audio
+recordings). The dynamic characteristics are primarily due to the
+level of persistence, e.g., temporary, durable, permanent (archived or
+published), or replicated; and also in the structure of the data
+stored, e.g., schema free tables as opposed to relational tables. There
+may also be dynamic characteristics of storage due to rate of growth
+of data, e.g., a field survey may bring in a lot of data during summer
+but none during other times. Some storage components like files and databases may also serve as data
+sources. These are commonly used in the Data Storage stage of the application (\figurename~\ref{fig:figures_application-stages}),
+but may also be present at the boundaries between the stages for transient staging of data.
+
+{\em Data Access} components are gateways to Data Sources and Data
+Storage components, and provide interfaces that often enforce security
+controls to restrict access, provide dynamic information about the
+available data sources and/or data analysis components, and may also
+perform logging of requests and data movement (e.g., tracking
+provenance for auditing). A side effect of auditing/logging may be the
+growth of provenance metadata simply due to access transactions even
+if no new data is created. Data Access components usually serve at the interface between the application stages, before
+and after data movement.
+
+{\em Data Movement} components are used for bulk or continuous
+movement of data between parts of the application, or to and from the
+application, % (often called staging) Data movement may be explicit
+and visible to the application (e.g., staging), or implicit (e.g.,
+automatic replication of files). Components may operate on streams or
+blocks (files). Dynamic characteristics include partitioning,
+scheduling and planning (e.g., edge caching of ``hot'' data, where
+datasets that are being used frequently during some period are
+replicated close to the application). Data movement can be considered
+to be real-time (i.e., as it happens, e.g., sensor data), `as real-time'
+(e.g., a storage buffer has been introduced into the transport system,
+but the data source is still real-time, as with time-series data), or
+asynchronous (where the ordering of the data movement is independent
+of the data itself).  Latency may be considered as an important
+characteristic of the data movement infrastructure depending on the
+application, e.g., real-time processing of sensor data for disaster
+analysis. Other dynamic aspects of data movement include data rates,
+their variability from data sources, and the ``freshness'' of
+data. These infrastructure appear on the edges between the application
+stages.
+
+{\em Data Discovery} components assist the access to and use of data
+sources, and include Catalogs, Information Services, and Metadata
+stores. The dynamic characteristics pertain to the evolution or
+recalibration of information held in these components. These components are orthogonal to the
+applications stages and can be used by any of the
+stages to guide their selection or identification of incoming data sources and outgoing data sinks
+or storage.
+%\ysnote{rate of growth of data may be an issue.
+%Catalogs that perform nightly ``crawling'' or data loading (e.g., sky surveys) may not be suitable if the
+%latest data needs to be accessible to applications. Catalog update will have to keep up with data
+%update. However, batch processing applications can survive less frequent updates.}
+
+{\em Notification} components respond to changes in the application or
+data and trigger other functionality. These include publish-subscribe message brokers,
+and complex event processing.
+By nature, these components support the dynamic aspects of an application.
+These components again are orthogonal to the applications stages and can be used for coordination of the stages and data movement.
+
+In each case, these components can be single or centralized,
+distributed, partitioned or federated. These are also typically
+offered as middleware and platform services, or system-level software
+capabilities.
+
+%\nchnote{where do we put data publishing? Calibration? provenance?}\ysnote{They can be incorporated in data discovery and data access}
 
@@
 
-% \alnote{Should we also include infrastructure for data exploration, data %
-mining, machine learning, visualization?}  %\katznote{we should think about how
-the items in this section map to the boxes and arrows in figure 1.  The first
-three map to both the data transformation \& decisions and data processing
-boxes.  I have no idea what should happen with workflow} % \alnote{should we
-also talk about the computational aspects of the infrastructure?  % At some
-point compute and data need to come together.}
-
-Analysis components include functionality such as comparison, statistical
-analysis, event analysis (i.e., analyzing information derived from data rather
-than raw data) and visualization. These components leverage the computational
-aspects of the infrastructure.  Based on the application stages outlined in
-\figurename~\ref{fig:figures_application-stages}, we identify a number of
-infrastructure services that can be associated with these stages.
-
-{\em Conversion} components (associated with the Data Transformation and
-Decisions stage) deal with the changing of data into a different form, including
-file conversion, (re-)formatting, transformation and metadata
-extraction. Heterogeneity of formats and representations is common in many
-scientific disciplines and such components often provide essential
-pre-processing of data prior to analysis. These are often referred to as adaptor
-services (or ``shim'' services in the Taverna workflow system~\cite{Taverna}).
-
-{\em Enrichment} components take data and attach additional detail to them,
-which might include filtering, image processing, information integration, data
-tagging, semantic annotation, data augmentation, or data fusion (e.g.,
-OLAP). Many of these components may deal with distributed sources of data. It is
-useful to note that these operations may not be fully automated and may involve
-a human expert. These components would be associated with the Data
+% \alnote{Should we also include infrastructure for data exploration, data
+% mining, machine learning, visualization?}
+%\katznote{we should think about how the items in this section map to the boxes and arrows in figure  1.
+The first three map to both the data transformation \& decisions and data processing boxes.
+I  have no idea what should happen with workflow}
+% \alnote{should we also talk about the computational aspects of the infrastructure?
+% At some point compute and data need to come together.}
+
+Analysis components include functionality such as comparison,
+statistical analysis, event analysis (i.e., analyzing information
+derived from data rather than raw data) and visualization. These components leverage
+the computational aspects of the infrastructure.
+Based on the application stages outlined in
+\figurename~\ref{fig:figures_application-stages}, we identify a number
+of infrastructure services for data analysis that can be associated with these stages.
+
+{\em Conversion} components (associated with the Data Transformation
+and Decisions stage) deal with the changing of data into a different
+form, including file conversion, (re-)formatting, transformation and
+metadata extraction. Heterogeneity of formats and representations is
+common in many scientific disciplines and such components often
+provide essential pre-processing of data prior to analysis. These are often
+referred to as adaptor services (or ``shim'' services in the Taverna workflow system~\cite{Taverna}).
+
+{\em Enrichment} components take data and attach additional detail to
+them, which might include filtering, image processing, information
+integration, data tagging, semantic annotation, data augmentation, or
+data fusion (e.g., OLAP). Many of these components may deal with
+distributed sources of data. It is useful to note that these
+operations may not be fully automated and may involve a human
+expert. These components would be associated with the Data
 Transformation and Decisions stage.
 
-%\ysnote{addresses Andre's commment} {\em Analytics and visualization}
-components help extract domain knowledge and assist with exploration of the
-data. These are often closely supervised by the end-user and may even be
-interactive. The proliferation of ``big data'' has caused data mining libraries,
-machine-learning toolboxes and visualization environments to become part of the
-infrastructure, some of which are deployed for specific domains but others which
-are more generic (e.g., Weka, GraphLab). These components are associated with
-the Data Processing stage.
+%\ysnote{addresses Andre's commment}
+{\em Analytics and visualization} components help extract domain knowledge and assist with
+exploration of the data. These are often closely supervised by the end-user and may even be
+interactive. The proliferation of ``big data'' has caused data mining libraries, machine-learning
+toolboxes and visualization environments to become part of the infrastructure, some of which are
+deployed for specific domains but others which are more generic (e.g., Weka, GraphLab, Apache Mahout). These components are associated
+with the Data Processing stage.
 
@@
 infrastructure discussed in this section, they enable the coordination of the
-different stages of the application, sometimes enforcing a particular constraint
-to improve, e.g., scalability. These approaches can allow the composition of
-sequential data processing pipelines and further offer the flexibility of
-modeling more complex constructs.  Standardization of APIs and components within
-these frameworks can allow interchangeability of components with similar
-functionality, such as those described in \S\ref{sec:infra_mgt} and
-\S\ref{sec:infra_analysis}.
-
-{\em Orchestration} components handle the management and choreography of data
-placement, code placement, communication, and messaging to allow complex
-analyses to be performed on distributed data sources. They may also be dynamic
-if they have the capability to adapt their behavior based on earlier parts of
-the workflow or can be steered based on external notification received during
-execution. In addition to supporting choreography, often making use of one or
-more resource manager/scheduler components, an orchestration component may also
-include specialist support for conversion and analysis components discussed
-above.
-
-{\em Bulk Data-parallel} frameworks use ``Bulk'' operators to express
-computations over collections of data which are executed in a loosely-coupled,
-data-parallel manner.  The {\em MapReduce} paradigm~\cite{MapReduce} and its
-variants~\cite{twister,hmr} are the most prominent among these. MapReduce forces
-programmers to model computation in two stages: a \emph{map} phase in which a
-transformation is applied to every element of a collection independently of
-every other; and a \emph{reduce} phase in which the transformed results are
-aggregated through a binary function that is associative and commutative.  These
-properties serve to guarantee that the MapReduce computation can be efficiently
-parallelized, by performing the map phase concurrently and then reducing without
-the reduction order affecting the final result.
-
-{\em Stream processing} (sometimes called event processing) is an approach
-modeled on signal processing, in which a stream of data is transformed
-``on-the-fly'' to produce one or more further streams of values. As such it is a
-programming abstraction suitable for continuous data feeds, or when results need
-to be pushed to users of interactive applications. Since the data is not assumed
-to be available \textit{en masse}, it can be used in situations where storing
-the entire data set is undesirable, unnecessary, or impossible.
-
-{\em Sensor networks} and other intelligent front-end data collection systems
-are sometimes seen as an extension of stream processing and application
-pipelines.  They often have the feature of being \emph{adaptive}, in that they
-can vary their exact behavior according either to external control stimuli or
-from observations made of the data they are themselves observing. Adaptation may
-also be through users interactively changing the the pipeline, either by
-changing parameters or by adding or removing tasks in the pipeline. This allows
-decision logic to be initiated by the pipeline itself, or using an external
-control signal that may come from users or an instrument. This form of
-``computational steering'' is especially useful in systems being used to explore
-a data set. In these cases, it is important to note that the infrastructure is
-adaptive, rather than the scenario itself.
-
-%Gomes et al.~\cite{gomes08} describe how a pipeline may be %dynamically
-modified using structural and behavioral operators.  %This also provides an
-end-to-end description of the entire %scientific process. This is valuable both
-for documentation purposes, %but also in pursuing issues of component and data
-re-use.
+different stages of the application, sometimes enforcing a particular constraint to improve, e.g., scalability.
+These approaches can allow the composition of sequential data processing pipelines and further offer the flexibility of modeling more complex constructs.
+Standardization of APIs and components within these frameworks can allow
+interchangeability of components with similar functionality, such as those described in \S\ref{sec:infra_mgt} and \S\ref{sec:infra_analysis}.
+
+{\em Orchestration} components handle the management and
+choreography of data placement, code placement, communication, and
+messaging to allow complex analyses to be performed on distributed
+data sources. These are often composed as workflows or Directed Acyclic Graphs (DAGs) that capture
+the execution dependency between
+components. They may also be dynamic if they have the capability to
+adapt their behavior based on earlier parts of the workflow or can be
+steered based on external notification received during execution. In
+addition to supporting choreography, while often making use of one or more
+resource manager/scheduler components, an orchestration component may also
+include specialist support for conversion and analysis components
+discussed above.
+
+{\em Bulk Data-parallel} frameworks use ``Bulk'' operators to express computations over 
+collections of data which are executed in a loosely-coupled, data-parallel manner.
+The {\em MapReduce} paradigm~\cite{MapReduce} and its variants~\cite{twister,hmr} are the most
+prominent among these, with recent ones like Pregel and Apace Giraph~\cite{pregel} supporting specific domains like
+graph analytics. MapReduce forces programmers to model computation in two stages: 
+a \emph{map} phase in which a transformation is applied to every element of a collection independently of every other;  and a \emph{reduce} phase in which the transformed results 
+are aggregated through a binary function that is associative and commutative. 
+These properties serve to guarantee that the MapReduce computation can be efficiently 
+parallelized, by performing the map phase concurrently and
+then reducing without the reduction order affecting the final result.
+
+{\em Stream processing} (sometimes called event processing) is an approach modeled 
+on signal processing, in which a stream of data is transformed ``on-the-fly'' to produce one or
+more further streams of values. As such it is a programming
+abstraction suitable for continuous data feeds, or when results need
+to be pushed to users of interactive applications. Since the data is
+not assumed to be available \textit{en masse}, it can be used in
+situations where storing the entire data set is undesirable,
+unnecessary, or impossible.
+
+{\em Sensor networks} and other intelligent front-end data collection systems are sometimes seen as an extension of stream processing and application pipelines.
+They often have the feature of being \emph{adaptive}, in that they can vary their exact behavior according either to external control stimuli or from observations made of the data they are 
+themselves observing. Adaptation may also be through users interactively changing the
+the pipeline, either by changing parameters or by adding or removing
+tasks in the pipeline. This allows decision logic to be initiated by
+the pipeline itself, or using an external control signal that may come
+from users or an instrument. This form of ``computational steering''
+is especially useful in systems being used to explore a data set. In these cases, it is important to 
+note that the infrastructure is adaptive, rather than the scenario itself.
+
+%Gomes et al.~\cite{gomes08} describe how a pipeline may be
+%dynamically modified using structural and behavioral operators.
+%This also provides an end-to-end description of the entire
+%scientific process. This is valuable both for documentation purposes,
+%but also in pursuing issues of component and data re-use.
 
@@
 We conclude this section by revisiting the application scenarios of
-\S\ref{sec:scenarios} in terms of the infrastructure and coordination approaches
-that they use currently, and discuss elements that could usefully be deployed.
-
-The software infrastructure components we have classified above address
-different characteristics of D3
-applications. \tablename~\ref{tab:inf_app_mapping} summarizes the role of these
-components in addressing dynamic and distributed characteristics of the D3
-applications in \S\ref{sec:scenarios}. Each application makes use of different
-components, though it can be seen that applications that utilize stream or event
-processing %(see \ref{tab:progSystems} \katznote{this table name doesn't
-resolve...}  have larger requirements on notification and orchestration
-components.
+\S\ref{sec:scenarios} in terms of the infrastructure and coordination approaches that they use
+currently, and discuss elements that could usefully be deployed.
+
+The software infrastructure components we have classified above
+address different characteristics of D3
+applications. \tablename~\ref{tab:inf_app_mapping} summarizes the role of
+these components in addressing dynamic and distributed characteristics
+of the D3 applications in \S\ref{sec:scenarios}.
+Each application makes use of different components, though it can be seen that applications that utilize stream or event processing %(see \ref{tab:progSystems} \katznote{this table name doesn't resolve...}
+have larger requirements on notification and orchestration components.
 
@@
     \begin{center}
-      \caption{Infrastructure used to support dynamic and distributed properties
-of application scenarios % \jhanote{Are all the terms % in parenthisis
-describe/defined. I think not, e.g., % orchestration? Also, I removed analysis
-as a % category.. needs fixing here too} \ysnote{fixed} }
+      \caption{Infrastructure used to support dynamic and distributed
+        properties of application scenarios % \jhanote{Are all the terms
+          % in parenthisis describe/defined. I think not, e.g.,
+          % orchestration? Also, I removed analysis as a
+          % category.. needs fixing here too} \ysnote{fixed}
+      }
       \label{tab:inf_app_mapping}
@@
 
-\hline \multicolumn{2}{|l|}{{\textbf{Application Name}}} \\ \hline %\nchnote{For
-example, workload decomposition and distribution must be determined dynamically
-in order to optimally use resources, e.g., compute resource selection can be
-based on data location, processing profile and/or network capacity. Workflows?
-Resource selection?}  Infrastructure for Dynamic Properties & Infrastructure for
-Distributed Properties \\ \hline
+\hline	
+\multicolumn{2}{|l|}{{\textbf{Application Name}}} \\ \hline	
+%\nchnote{For example, workload decomposition and distribution must be determined dynamically in order to optimally use resources, e.g., compute resource selection can be based on data location, processing profile and/or network capacity.
+Workflows?
+Resource selection?}
+Infrastructure to support Dynamic Properties & Infrastructure to support Distributed Properties
+ \\ \hline
 %\begin{footnotesize}
 %	 \begin{longtable}[h!]{|p{6.7cm}|p{6.7cm}|}
-% \caption{Infrastructure used to support dynamic and distributed properties of
-application scenarios} \label{tab:inf_app_mapping} \\ % \hline %
-\textbf{Infrastructure to support Dynamism} % &\textbf{Infrastructure to support
-Distributedness} \\
-% \endhead \hline \multicolumn{2}{|l|}{{\textbf{Next Generation Sequencing (NGS)
-Analytics}}} \\ \hline %\nchnote{For example, workload decomposition and
-distribution must be determined dynamically in order to optimally use resources,
-e.g., compute resource selection can be based on data location, processing
-profile and/or network capacity. Workflows? Resource selection?}  Dynamic
-workload decomposition and distribution for resource optimization
-\emph{(Orchestration)} & Distributing data from source (generation) to the
-computing destination (analysis) \emph{(Data Discovery, Movement)} % No unifying
-single requirement or single solution at present (or possible at all?)  %
-Normally data local and CI doesn't easily support movement of data to compute %
-\ysnote{We may want to remove the preceding lines since we're talking about
-existing infrastructure, not missing ones at this point.}  % naturally
-distributed: data source (NGS machine storage) and compute on % different
-resources (typically not co-located); loosely coupled (many % machines,
-different repositories) \\ \hline \multicolumn{2}{|l|}{{\textbf{ATLAS (an
-example of use of the WLCG) Reconstruction app}}} \\ \hline Manage gradual
-growth of data using SRM \emph{(Data Storage, Access)} % Management, subset
-replication and partitioning to go inf components: data replication, (gLite),
-data movement % specify SRM as standard and then they leave choice of
-implementation to the organisations as part of consortium.  % New raw data is
-collected % steadily (2) \note{timescale needs to worked out. What happens if
-something is found?}  & Management, subset replication, data movement and
-partitioning using gLite and SRM \emph{(Data Storage, Access, Movement,
-Discovery)} % , data movement % natural, (sub-set) replicated, loosely coupled
-\note{Is natural the % right word? Artificial % distribution (the source is
-natural, but the intermediate points are % artificial), data not
-partitioned}\ysnote{These details are not fully present in the actual
-application description of its dynamism (Sec 2.2). Should we trim there and move
-some content there? } \\ \hline \multicolumn{2}{|l|}{{\textbf{LSST}}} \\ \hline
-
-Steering observations, Databases to manage data growth \emph{(Notification, Data
-Storage)} % DE decisions based on observed data (2, 5) \ysnote{what do these
-numbers mean? specific % characteristics?}  & Coordinate classification,
-distributed databases, data movement to distributed observatories
-\emph{(Notification, Data Access, Data Movement, Data Storage)} %Analysis, % ,
-natural distribution of telescopes \\ \hline \multicolumn{2}{|l|}{{\textbf{SOA
-Astronomy Applications}}} \\ \hline
-
-Catalog contents updated, Query mix changes \emph{(Data Discovery)} % data is
-continuously updated, thus dynamic data (pre-defined schedules or on-demand),
-queries change (7)\note{timescale clarification!}  & Distributed VO Services for
-data storage and access \emph{(Data Discovery, Data Access, Data Movement)} %
-moving data sources, services and storage are distributed across institutions \\
-\hline \multicolumn{2}{|l|}{{\textbf{Cosmic Microwave Background}}} \\ \hline
-Dynamic resource selection %dynamic utilization of resources (4) & Compute on
-distributed HPC centers \emph{(Data Movement)} % (sub-set) replication (during
-application run) \note{What phases of the application do we consider? data
-acquisition? app run?}\ysnote{Can any of the supercomputers be selected? If so,
-the data is moved to distributed compute (or it is dynamic resource selection)}
-\\ \hline \multicolumn{2}{|l|}{{\textbf{Sensor Network Application}}} \\ \hline
-Notification of sensor data when in range, data rates and volumes \emph{(Data
-Movement, Notification)} %data rate, volume dynamic (1-2) &Aggregating
-distributed sensor data \emph{(Data Movement, Notification)} \\ \hline
-\multicolumn{2}{|l|}{{\textbf{Climate (ESGF)}}} \\ \hline Manage data growth %
-New data generated over time (ESGF-wide - is this in scope of the application?)
-& Distributed, replicated catalogs using ESGF software stack (GridFTP,
-wget/HTTP) \emph{(Data Access, Data Discovery, Data Movement, Data Conversion)}
-\\ % multiple climate centers (natural), users can request data from ESGF
-(sub-set replication), coupled % Can come in via website - cookie / OpenID %
-Come in through command line script - grid cert \hline
-\multicolumn{2}{|l|}{{\textbf{Interactive Exploration of Environmental Data}}}
-\\ \hline On-demand data subsetting, caching and versioning, real-time
-instrument data \emph{(Data Discovery, Data Movement, Data Sources)} % at moment
-services are hosted on a single address. There may be farming for load balancing
-in the future.  % data rates change (1), % data inputs/outputs change (3),
-queries change (7) & Distributed Data Services, Access Restrictions \emph{(Data
+%		\caption{Infrastructure used to support dynamic and distributed properties of application scenarios} \label{tab:inf_app_mapping} \\
+%		\hline
+%		\textbf{Infrastructure to support Dynamism}
+%                &\textbf{Infrastructure to support Distributedness} \\		
+%		 \endhead
+		\hline	
+\multicolumn{2}{|l|}{{\textbf{Next Generation Sequencing (NGS) Analytics} (\S\ref{bioSilvia})}} \\ \hline	
+%\nchnote{For example, workload decomposition and distribution must be determined dynamically in order to optimally use resources, e.g., compute resource selection can be based on data location, processing profile and/or network capacity.
+Workflows?
+Resource selection?}
+Dynamic workload decomposition and distribution for resource optimization \emph{(Orchestration)}
+		&
+Distributing data from source (generation) to the computing destination (analysis) \emph{(Data
+  Discovery, Movement)}
+% 		No unifying single requirement or single solution at present (or possible at all?)
+% 		Normally data local and CI doesn't easily support movement of data to compute
+% \ysnote{We may want to remove the preceding lines since we're talking about existing infrastructure, not missing ones at this point.}
+% naturally distributed: data source (NGS machine storage) and compute on
+% 		different resources (typically not co-located); loosely coupled (many
+%		machines, different repositories)
+\\
+		\hline
+\multicolumn{2}{|l|}{{\textbf{ATLAS (an example of use of the WLCG) Reconstruction app} (\S\ref{WLCGSteve})}} \\ \hline
+Manage gradual growth of data using SRM \emph{(Data Storage, Access)}
+% Management, subset replication and partitioning to go inf components: data replication, (gLite), data movement
+% specify SRM as standard and then they leave choice of implementation to the organisations as part of consortium.
+% New raw data is collected
+% 	    steadily (2) \note{timescale needs to worked out. What happens if something is found?}
+&
+Management, subset replication, data movement and partitioning using gLite and SRM \emph{(Data
+  Storage, Access, Movement, Discovery)}
+% , data movement
+% natural, (sub-set) replicated, loosely coupled \note{Is natural the
+% 	   right word? Artificial
+% 		distribution (the source is natural, but the intermediate points are
+% 		artificial), data not partitioned}\ysnote{These details are not fully present in the actual application description of its dynamism (Sec 2.2). Should we trim there and move some content there? }
+              \\
+		\hline
+\multicolumn{2}{|l|}{{\textbf{LSST} (\S\ref{astroAdam})}} \\ \hline
+
+Steering observations, Databases to manage data growth \emph{(Notification, Data Storage)}
+% DE decisions based on observed data (2, 5) \ysnote{what do these numbers mean? specific
+             %  characteristics?}
+            &
+Coordinate classification, distributed databases, data movement to distributed observatories
+\emph{(Notification, Data Access, Data Movement, Data Storage)} %Analysis, % , natural distribution of telescopes
+\\
+		\hline
+\multicolumn{2}{|l|}{{\textbf{SOA Astronomy Applications} (\S\ref{astroSOAAdam})}} \\ \hline
+
+Catalog contents updated, Query mix changes \emph{(Data Discovery)}
+% data is continuously updated, thus dynamic data (pre-defined schedules or on-demand), queries change (7)\note{timescale clarification!}
+             &
+Distributed VO Services for data storage and access \emph{(Data Discovery, Data Access, Data Movement)}
+% moving data sources, services and storage are distributed across institutions
+\\
+		\hline
+\multicolumn{2}{|l|}{{\textbf{Cosmic Microwave Background} (\S\ref{astroJulian})}} \\ \hline	
+	      Dynamic resource selection  \emph{(Data Discovery)}
+%dynamic utilization of resources (4)
+& Compute on distributed HPC centers \emph{(Data Movement)}
+% (sub-set) replication (during application run) \note{What phases of the application do we consider?
+data acquisition?
+app run?}\ysnote{Can any of the supercomputers be selected?
+If so, the data is moved to distributed compute (or it is dynamic resource selection)}
+\\
+		\hline
+\multicolumn{2}{|l|}{{\textbf{Sensor Network Application} (\S\ref{sensorSimon})}} \\ \hline	
+	     Notification of sensor data when in range, data rates and volumes \emph{(Data Movement, Notification)} %data rate, volume dynamic (1-2)
+&Aggregating distributed sensor data \emph{(Data Movement, Notification)}  \\
+		\hline
+\multicolumn{2}{|l|}{{\textbf{Climate (ESGF)} (\S\ref{climateDan})}} \\ \hline	
+	      Manage data growth % New data generated over time (ESGF-wide - is this in scope of the application?)
+             &
+Distributed, replicated catalogs using ESGF software stack (GridFTP, wget/HTTP) \emph{(Data Access,
+  Data Discovery, Data Movement, Data Conversion)} \\
+  % multiple climate centers (natural), users can request data from ESGF (sub-set replication), coupled
+% Can come in via website - cookie / OpenID
+% Come in through command line script - grid cert
+		\hline
+\multicolumn{2}{|l|}{{\textbf{Interactive Exploration of Environmental Data} (\S\ref{envJon})}} \\ \hline	
+On-demand data subsetting, caching and versioning, real-time instrument data \emph{(Data Discovery, Data
+  Movement, Data Sources)}
+% at moment services are hosted on a single address. There may be farming for load balancing in the future.
+% data rates change (1),
+% 	    data inputs/outputs change (3), queries change (7)
+		&
+Distributed Data Services, Access Restrictions \emph{(Data
   Movement, Data Access)}
 
-% data frequently located, but data sources themselves don't change
-locations. Some replication for bigger community datasets. No replication
-service - just another website!
-
-% distribution of data has security constraints. Access rights which means you
-can't just copy data. Data providers like to assert these rights at source.
-
-% Increasingly user tokens being passed to data providers, but previously
-application had tokens and trust required on part of data providers. Data
-providers want to have user tokjens so they can tell usage, and whether its
-researchers or commercial.
-
-% Web server accessing data on your behalf. Not often directly accessing data -
-going through an intermediary service. trying grid proxy certs and OAuth.
+% data frequently located, but data sources themselves don't change locations. Some replication for bigger community datasets. No replication service - just another website!
+
+% distribution of data has security constraints. Access rights which means you can't just copy data. Data providers like to assert these rights at source.
+
+% Increasingly user tokens being passed to data providers, but previously application had tokens and trust required on part of data providers.
+Data providers want to have user tokjens so they can tell usage, and whether its researchers or commercial.
+
+% Web server accessing data on your behalf. Not often directly accessing data - going through an intermediary service. trying grid proxy certs and OAuth.
 
@@
 
-% As a response to query you get a subset of data % Virtualization and
-aggregation of files - see single logical dataset rather than the thousands of
-files its stored as.
-
-% OpenDAP binary stream for data transport. logical OpenDAp data set may be
-collection of files.
-
-% If file changes (e.g. due to satellite reimaging) however difficult to tell
-when data set has changed as versioning not well supported.
-
-\\ \hline % Dynamic Environmental Data on the Web &data operated on changes %
-(realtime feeds) (2) % &natural (sensor sources, simulations), (sub-set)
-replication\\ % \hline \multicolumn{2}{|l|}{{\textbf{PowerGrids}}} \\ \hline
-
-Stream processing over variable sources, rates and volumes, Event processing
-over different patterns, Trigger analysis on-demand \emph{(Data Sources,
-Notification, Enrichment, Orchestration)} %Analysis, % data changing at the
-because application choosing more accurate sources % How is the selection made?
-Is there a registry?  % There is a registry /catalog, it is distributed (sensors
-catalogued by powergrid, buildings in campus database). Require distributed data
-colletion to create a virtual single catalog.  % Using cloud computing to handle
-scaling/elasticity for peaks. Not automated as yet but expected for the future.
-& Dispersed event sources and sinks, Launch ensemble modeling on distributed
-clusters and cloud \emph{(Data Sources, Notification, Orchestration)} %Analysis,
-
-% There is a registry /catalog, it is distributed (sensors catalogued by
-powergrid, buildings in campus database). Require distributed data colletion to
-create a virtual single catalog.
-
-% Output of computation itself is distributed - sent to various neighbourhoods /
-locations.  % Some sent to human beings, sms, emails, many formats. Best effort
-message delivery.
+% As a response to query you get a subset of data
+% Virtualization and aggregation of files - see single logical dataset rather than the thousands of files its stored as.
+
+% OpenDAP binary stream for data transport. logical OpenDAp data set may be collection of files.
+
+% If file changes (e.g. due to satellite reimaging) however difficult to tell when data set has changed as versioning not well supported.
 
 \\
-
-		\hline \multicolumn{2}{|l|}{{\textbf{Fusion (ITER)}}} \\ \hline
-Stream and transform data between simulation sites on-demand \emph{(Data
-Movement, Enrichment)} % (multiple distinct data % flows), on-the-fly %
-transformation (8) & Run analysis on available, distributed resources
-\emph{(Data Movement)} %Analysis % natural, coupled, transient sub-set
-replicated \\ \hline \multicolumn{2}{|l|}{{\textbf{Industrial Incident
-Notification and Response}}} \\ \hline Acquire different data sources and rates,
-Processing determined by data \emph{(Data Sources, Data Access, Notification,
-Orchestration)} %Analysis, % (1), % data itself changes (2), dynamic data
-sources (3), processing changes % based on data %(5) & Distributed data sources,
-Analysis by agents across institutions \emph{(Data Sources, Data Access)} % ,
-Analysis % natural, coupled \\ \hline \multicolumn{2}{|l|}{{\textbf{MODIS Data
-Processing}}} \\ \hline
-
-% Satellite is already well known - static set of FTP sites.  -- % Pipeline is a
-static structure - knows what it needs to do. Three different tasks looslely
-coupled.  % Pipeline is setup for a community and once up and running will be
-available to users for a while.
-
-% How are users authorised to use the system? Portal, allows them to choose a
-region. Basic authN?
-
-		& Data from distributed FTP sources, Cloud queues for
-coordination, Partitioned NoSQL data \emph{(Data movement, Data Storage,
-Orchestration)} %Reliable message queues used as you don't want to have a black
-spot in the sky - is it a real feature or a missing data. Try not to miss any
-tiles!  \\ \hline %\multicolumn{3}{|l|}{{\textbf{Floating Sensors}}} \\ \hline %
-&data rate (1), data itself (2), application causes % data changes (5) % &
-spatial-temporal correlated, coupled \\ % \hline
-
-\multicolumn{2}{|l|}{{\textbf{Distributed Network Intrusion Detection}}} \\
-\hline Analyze dynamic network traffic rates \emph{(Data Movement)} % , Analysis
-% data rate variable (1), data % highly dynamic changing (2), application
-responds to data (5), % on-the-flight processing (8) & Run analysis at different
-sites, Distribute generated data \emph{(Data Movement)} % Analysis, % natural,
-data/compute tightly coupled \\ \hline
+		\hline
+	    % Dynamic Environmental Data on the Web &data operated on changes
+	    %     (realtime feeds) (2)
+	    %     &natural (sensor sources, simulations), (sub-set) replication\\
+	    %     \hline
+\multicolumn{2}{|l|}{{\textbf{PowerGrids} (\S\ref{power})}} \\ \hline
+
+Stream processing over variable sources, rates and volumes, Event processing over different
+patterns, Trigger analysis on-demand  \emph{(Data
+  Sources, Notification, Enrichment,  Orchestration)} %Analysis,
+% data changing at the because application choosing more accurate sources
+% How is the selection made? Is there a registry?
+% There is a registry /catalog, it is distributed (sensors catalogued by powergrid, buildings in campus database). Require distributed data colletion to create a virtual single catalog.
+% Using cloud computing to handle scaling/elasticity for peaks. Not automated as yet but expected for the future.
+&
+Dispersed event sources and sinks, Launch ensemble modeling on distributed clusters and cloud
+\emph{(Data Sources, Notification,  Orchestration)} %Analysis,
+
+% There is a registry /catalog, it is distributed (sensors catalogued by powergrid, buildings in campus database). Require distributed data colletion to create a virtual single catalog.
+
+% Output of computation itself is distributed - sent to various neighbourhoods / locations.
+% Some sent to human beings, sms, emails, many formats. Best effort message delivery.
+
+ \\
+
+		\hline
+\multicolumn{2}{|l|}{{\textbf{Fusion (ITER)} (\S\ref{fusionScott})}} \\ \hline	
+	     Stream and transform data between simulation sites on-demand \emph{(Data Movement, Enrichment)}
+%  (multiple distinct data
+	    % flows), on-the-fly
+	    % transformation (8)
+	    &
+Run analysis on available, distributed resources \emph{(Data Movement)} %Analysis
+% natural, coupled, transient sub-set replicated
+            \\
+		\hline
+\multicolumn{2}{|l|}{{\textbf{Industrial Incident Notification and Response} (\S\ref{Kees})}} \\ \hline	
+	     Acquire different data sources and rates, Processing determined by data
+             \emph{(Data Sources, Data Access, Notification, Orchestration)}   %Analysis,  % (1),
+	    % data itself changes (2), dynamic data sources (3), processing changes
+	    % based on data
+            %(5)
+&
+Distributed data sources, Analysis by agents across institutions, Access control \emph{(Data Sources, Data
+  Access)} % , Analysis
+% natural, coupled
+            \\
+		\hline
+\multicolumn{2}{|l|}{{\textbf{MODIS Data Processing} (\S\ref{modisKeith})}} \\ \hline
+
+% Satellite is already well known - static set of FTP sites.
+--
+% Pipeline is a static structure - knows what it needs to do. Three different tasks looslely coupled.
+% Pipeline is setup for a community and once up and running will be available to users for a while.
+
+% How are users authorised to use the system? Portal, allows them to choose a region. Basic authN?
+
+		&
+Data from distributed FTP sources, Cloud queues for coordination, Partitioned NoSQL data \emph{(Data
+  movement, Data Storage, Orchestration)}
+%Reliable message queues used as you don't want to have a black spot in the sky - is it a real feature or a missing data. Try not to miss any tiles!
+\\
+		\hline
+%\multicolumn{3}{|l|}{{\textbf{Floating Sensors}}} \\ \hline	
+%	     &data rate (1), data itself (2), application causes
+%	    data changes (5)
+%		& spatial-temporal correlated, coupled \\
+%		\hline
+
+\multicolumn{2}{|l|}{{\textbf{Distributed Network Intrusion Detection} (\S\ref{detectionJon})}} \\ \hline	
+Analyze dynamic network traffic rates \emph{(Data Movement)} % , Analysis
+	    %  data rate variable (1), data
+	    % highly dynamic changing (2), application responds to data (5),
+	    % on-the-flight processing (8)
+            &
+Run analysis at different sites, Distribute generated data \emph{(Data Movement)} % Analysis,
+% natural, data/compute tightly coupled
+\\
+		\hline
 %	 \end{longtable}
@@
 % \begin{enumerate}
-% \item What programming systems and techniques are \emph{currently} % used in
-implementing the application scenario?
-% \item What elements identified in this section \emph{could} usefully % be
-deployed or further explored in each application scenario?
+% \item What programming systems and techniques are \emph{currently}
+%   used in implementing the application scenario?
+% \item What elements identified in this section \emph{could} usefully
+%   be deployed or further explored in each application scenario?
 % \end{enumerate}
 
-% \jhanote{I think the second question is not addressed in the remainder % of
-this subsection? Yogesh/Neil?}  % \ysnote{removed the "questions" and rephrased
-it as a discussion}
+% \jhanote{I think the second question is not addressed in the remainder
+%   of this subsection? Yogesh/Neil?}
+% \ysnote{removed the "questions" and rephrased it as a discussion}
 
 \tablename~\ref{table:progSystems} summarizes the coordination approaches
-currently dominant in each application case study. It is worth noting that, in
-common with most long-lived systems, many of the cases have accreted substantial
-code bases over time and so show little in the way of systematic
-architecture. (One reason for the dominance of workflow pipelines may be that
-such architectures are quite suitable for adding extra processing steps to
-existing computations.)
+currently dominant in each application case study. It is worth noting
+that, in common with most long-lived systems, many of the cases
+have accreted substantial code bases over time and so show little in
+the way of systematic architecture. (One reason for the dominance of
+workflow pipelines may be that such architectures are quite suitable for adding
+extra processing steps to existing computations.)
 
@@
     \begin{center}
-      \caption{Coordination approaches used by the
-applications\label{table:progSystems}}
-      \begin{tabular}{|p{6.3cm}|p{7.2cm}|} \hline Application & Current
-approaches \\ \hline \hline NGS Analytics & pipelines that ship with
-sequencers~\cite{illumina}; workflows~\cite{Taverna} \\ \hline ATLAS & custom
-algorithms; job submission tied to WLCG \\ \hline LSST & custom layer; scripted
-workflows; queries \\ \hline SOA Astronomy & web services \\ \hline Cosmic
-Microwave Background & \textit{ad hoc} \\ \hline Sensor Network Application &
-focused on data collection \\ \hline Climate & \textit{ad hoc} \\ \hline
-Interactive Exploration of Environmental Data & visualization tools \\ \hline
-Power Grids & stream processing; data mining \\ \hline Fusion (ITER) & scripted
-pipeline \\ \hline Industrial Incident Notification and Response & agent-based
-\\ \hline MODIS Data Processing & scripted pipeline \\ % \hline Floating Sensors
-& \textit{ad hoc} \\ \hline Distributed Network Intrusion Detection & monolithic
-packages \\ \hline
+      \caption{Coordination approaches used by the applications\label{table:progSystems}}
+      \begin{tabular}{|p{6.3cm}|p{7.2cm}|}
+        \hline Application  & Current approaches \\
+        \hline
+        \hline NGS Analytics & pipelines that ship with
+        sequencers~\cite{illumina}; workflows~\cite{Taverna} \\
+        \hline ATLAS & custom algorithms; job submission tied to WLCG \\
+        \hline LSST & custom layer; scripted workflows; queries \\
+        \hline SOA Astronomy & web services~\cite{barker} \\
+        \hline Cosmic Microwave Background & \textit{ad hoc} \\
+        \hline Sensor Network Application & focused on data collection \\
+        \hline Climate & \textit{ad hoc} \\
+        \hline Interactive Exploration of Environmental Data &
+        visualization tools~\cite{blower2} \\
+        \hline Power Grids & stream processing~\cite{Simmhan:sciencecloud:2011}; data mining~\cite{Yin:mapreduce:2012} \\
+        \hline Fusion (ITER) & scripted pipeline~\cite{adios} \\
+        \hline Industrial Incident Notification and Response &
+        agent-based~\cite{emergency4}  \\
+        \hline MODIS Data Processing & scripted pipeline~\cite{modis:ipdps:2010} \\
+%        \hline Floating Sensors & \textit{ad hoc} \\
+        \hline Distributed Network Intrusion Detection & monolithic
+        packages \\
+        \hline
       \end{tabular}
@@
 
-The applications in \tablename~\ref{table:progSystems}, which make use of
-scripted or \textit{ad hoc} pipelines, could in most cases be easily replaced
-with a structured workflow engine. Doing so would bring immediate benefits in
-terms of reproducibility, but also in terms of standardization of components and
-interfaces.
-
-%Standardization is not simply a software issue; it has %scientific significance
-in the sense that data produced using %well-documented, well-used, and
-well-debugged suites of components is %often more trusted -- and trustworthy --
-than data produced using %locally-produced components that, while they may be
-more innovative %and targeted, also run the risk of undetected assumptions and
-errors.  %\jhanote{i think standardization of whatever data/component etc as an
-% issue is independent of structured workflow engine. in fact i would % argue
-that the causative arrow flows in the opposite direction: % structured workflow
-engines do not help standardization but % standardizations does help structured
-workflow engines..}  %\jhanote{IMO also still need to make it relevant to the
-sub-section % aims?}
-
-%\jhanote{NCH, we don't talk about standardization before this do we? so why %
-further? Are we still trying to understand table IV? if yes, how % does this
-relate?}  A further implication of standardization is that component reuse is
-greatly improved. This is perhaps best illustrated by observing the increased
-ability to automate the processing step over repeated runs.  The use of standard
-component-based software techniques such as web services, coupled with web
-technologies for metadata representation, assist in reuse, re-purposing and
+The applications in \tablename~\ref{table:progSystems}, which make use
+of scripted or \textit{ad hoc} pipelines, could in most cases be easily
+replaced with a structured workflow engine. Doing so would bring
+immediate benefits in terms of reproducibility, but also in terms of
+standardization of components and interfaces.
+
+%Standardization is not simply a software issue; it has
+%scientific significance in the sense that data produced using
+%well-documented, well-used, and well-debugged suites of components is
+%often more trusted -- and trustworthy -- than data produced using
+%locally-produced components that, while they may be more innovative
+%and targeted, also run the risk of undetected assumptions and errors.
+%\jhanote{i think standardization of whatever data/component etc as an
+%  issue is independent of structured workflow engine. in fact i would
+%  argue that the causative arrow flows in the opposite direction:
+%  structured workflow engines do not help standardization but
+%  standardizations does help structured workflow engines..}
+%\jhanote{IMO also still need to make it relevant to the sub-section
+%  aims?}
+
+%\jhanote{NCH, we don't talk about standardization before this do we? so why
+%  further? Are we still trying to understand table IV? if yes, how
+%  does this relate?}
+A further implication of standardization is that
+component reuse is greatly improved. This is perhaps best illustrated
+by observing the increased ability to automate the processing step
+over repeated runs.  The use of standard component-based software
+techniques such as web services, coupled with web technologies for
+metadata representation, assist in reuse, re-purposing and
 integration, for instance into different workflow systems.
 
-The {\em MapReduce} paradigm~\cite{MapReduce} and its
-variants~\cite{twister,hmr} are the most prominent among the bulk data-parallel
-frameworks which use bulk operators to express computations over collections of
-data in an loosely coupled manner.  Applications such as Next Generation
-Sequencing (NGS) (\S\ref{bioSilvia}) and Power Grids (\S\ref{power}) use
-MapReduce for their
-analytics~\cite{cloudburst,genometoolkit,Yin:mapreduce:2012}.
+The {\em MapReduce} paradigm~\cite{MapReduce} and its variants~\cite{twister,hmr} are the most prominent among the bulk data-parallel frameworks which use bulk operators to express computations over collections of data in an loosely coupled manner.
+Applications such as Next Generation Sequencing (NGS) (\S\ref{bioSilvia}) and Power Grids (\S\ref{power}) use MapReduce for
+their analytics~\cite{cloudburst,genometoolkit,Yin:mapreduce:2012}.
 
 There are many examples of stream~\cite{chandrasekaran:telegraphcq2003,
-abadi:borealis2005, liew:streamgraph2010} and event~\cite{ESPER, siddhi,
-MicrosoftInsight} processing frameworks, and recent works have combined
-streaming with more traditional file-based
-workflows~\cite{ZinnCluster,Herath:CCGrid:2010}.  Applications such as the
-Marine Sensor Network (\S\ref{sensorSimon}), Distributed Network Intrusion
-Detection (\S\ref{detectionJon}), Power Grids (\S\ref{power}), and Industrial
-Incident Notification and Response (\S\ref{Kees}) use stream and event
-processing abstractions.
-
-The view of scientific applications as pipelines hides significant differences
-between (and within) domains, since the different stages of the pipeline receive
-different emphases in different applications. Some applications must collect and
-store all the incoming data, with only minor pre-processing to clean the data
-ahead of the main processing. For others, a major constraint is to reduce the
-incoming data stream as quickly as possible, perhaps discarding a large fraction
-of the data before any processing occurs. In spite of these differences, we
-believe the use of the pipeline abstraction is still useful.  Taking some of the
-examples from \S\ref{sec:scenarios}, medical imaging and astrophysics often fall
-into the first category, while particle physics (especially in the case of the
-Large Hadron Collider) often falls dramatically into the second, to the extent
-that principled data reduction forms one of the major scientific challenges for
-the project.
-
-There are also significant variations in where ``the pipeline'' begins and
-ends. This represents the difficulty in differentiating between ``traditional''
-applications and ``infrastructure'' applications. Many systems are driven by
-data sets collected and shared widely, against which several different
-applications may be applied.  Here the infrastructure application is the
-collection of components used to allow the data sets to be accessed and analyzed
-by the different applications.  Climate science is a good example of this
-approach. In this case the data is ``pre-collected'' in the sense of existing
-outside most of the applications, although its collection may have been a
-complex task. The raw data required for gene sequencing (\S\ref{bioSilvia}), for
-example, while being collected using a dedicated and very sophisticated
-computerized processing system, would generally be regarded as ``given'' and not
-part of the processing pipeline, since the pipeline would not affect the
-collection task. Another way to look at this is that data is collected
+abadi:borealis2005,
+liew:streamgraph2010} and event~\cite{ESPER, siddhi,
+MicrosoftInsight} processing frameworks, and recent works have
+combined streaming with more traditional file-based
+workflows~\cite{ZinnCluster,Herath:CCGrid:2010}.  Applications such as
+the Marine Sensor Network (\S\ref{sensorSimon}), Distributed Network
+Intrusion Detection (\S\ref{detectionJon}), Power Grids
+(\S\ref{power}), and Industrial Incident Notification and Response
+(\S\ref{Kees}) use stream and event processing abstractions.
+
+The view of scientific applications as pipelines hides significant
+differences between (and within) domains, since the different stages
+of the pipeline receive different emphases in different
+applications. Some applications must collect and store all the
+incoming data, with only minor pre-processing to clean the data ahead
+of the main processing. For others, a major constraint is to reduce
+the incoming data stream as quickly as possible, perhaps discarding a
+large fraction of the data before any processing occurs. In spite of these differences, we believe the use of the pipeline abstraction is still useful.
+Taking some of the examples from
+\S\ref{sec:scenarios}, medical imaging and astrophysics often fall
+into the first category, while particle physics (especially in the
+case of the Large Hadron Collider) often falls dramatically into the
+second, to the extent that principled data reduction forms one of the
+major scientific challenges for the project.
+
+There are also significant variations in where ``the pipeline'' begins
+and ends. This represents the difficulty in differentiating between ``traditional'' applications 
+and ``infrastructure'' applications. Many systems are driven by data sets collected and shared
+widely, against which several different applications may be applied. 
+Here the infrastructure application is the collection of components used to allow the 
+data sets to be accessed and analyzed by the different applications. 
+Climate science is a good example of this approach. In this
+case the data is ``pre-collected'' in the sense of existing outside
+most of the applications, although its collection may have been a
+complex task. The raw data required for gene sequencing
+(\S\ref{bioSilvia}), for example, while being collected using a
+dedicated and very sophisticated computerized processing system, would
+generally be regarded as ``given'' and not part of the processing
+pipeline, since the pipeline would not affect the collection
+task. Another way to look at this is that data is collected
 \emph{independently} of its further processing.
 
-Alternatively the application may be driven directly from a live data stream,
-with processing happening on the fly. The front-end data collection subsystem is
-becoming increasingly flexible and capable, with the introduction of sensor
-networks to replace more passive approaches, and it may often make sense to
-include the sensor network into the application workflow pipeline, such as for
-Power Grids and Industrial Incident Response. All these choices have
-implications for design and provisioning. For instance, the choice of which sort
-of pipeline a particular application uses may well be fixed, or will change only
-slowly as the experimental environment evolves. For example, we see some systems
-that add simulation alongside experimental data collection to act as a
-verification and validation step.
-
-Revisiting the overall issue of infrastructure, which we defined as the hardware
-and software needed to run the applications, we note that the specific hardware
-isn't a primary issue.  The applications we've discussed all use a mix of the
-systems that exist today, which have aspects of web, grid, and cloud, and can be
-considered a hybrid of the three.  The key infrastructure components are really
-the software and services on top of the hardware.  In this section, we've
-discussed infrastructure to support data management and data analysis, as well
-as the tools and systems used to program applications, and the specific
-components used for the application in \S\ref{sec:scenarios}.  However, there is
-not much commonality in the software and services that the applications use, so
-the software and programming infrastructure we've described here is much more
-broad than what is used by the applications we've studied.
-
-%``This paper presents the design and the implementation of an %architecture for
-the analysis of data streams in distributed %environments. In particular, data
-stream analysis has been carried out %for the computation of items and itemsets
-that exceed a frequency %threshold. The mining approach is hybrid, that is,
-frequent items are %calculated with a single pass, using a sketch algorithm,
-while %frequent itemsets are calculated offline by a further analysis. The
-%architecture combines parallel and distributed processing to keep the %pace
-with the rate of distributed data streams. In order to keep %computation close
-to data, miners are distributed among the domains %where data streams are
-generated. The paper also reports the %experiment results obtained with a
-prototype of the architecture, %tested on a Grid composed of two domains
-handling two different data %streams. Results confirm the scalability of the
-approach and show that %it is possible to keep the pace of data production and
-reduce the %communication overhead.''
-%
-%``.. In some cases streams of data must be analyzed in real time to %provide
-information about trends, outlier values or regularities that %must be signaled
-as soon as possible.  The need for online computation %is a notable challenge
-with respect to classical data mining %algorithms [1], [2].  Important
-application fields for stream mining %are as diverse as financial applications,
-network monitoring, security %problems, telecommunication networks, Web
-applications, sensor %networks, analysis of atmospheric data, etc.  A further
-difficulty %originates when streams are distributed, and mining models must be
-%derived not only for the data of a single stream, but for the %integration of
-multiple and heterogeneous data streams [3]. This %scenario can occur in all the
-application domains mentioned %before. For example, in a Content Distribution
-Network, user requests %delivered to a Web system can be forwarded to any of
-several servers %located in different and possibly distant places, in order to
-serve %requests more efficiently and balance the load.''
-%
-%``..the architecture combines the parallel and distributed %paradigms, the
-first to keep the pace with the rate of a %single data stream, by using multiple
-miners (processors %or cores), the second to cope with the distributed %nature
-of data streams. Miners are distributed among %the domains where data streams
-are generated, in order %to keep computation close to data.''
-%
-%
-%
-%FlumeJava~\cite{flumeJava} is a distributed, reliable, and available %service
-for efficiently collecting, aggregating, and moving large %amounts of log
-data. It has a simple and flexible architecture based %on streaming data
-flows. It is robust and fault tolerant with tunable %reliability mechanisms and
-many failover and recovery mechanisms. The %system is centrally managed and
-allows for intelligent dynamic %management. It uses a simple extensible data
-model that allows for %online analytic application.
-
-% Programming % Factored out by sd for various not-very-good reasons....
+Alternatively the application may be driven directly
+from a live data stream, with processing happening on the
+fly. The front-end data collection subsystem is becoming increasingly
+flexible and capable, with the introduction of sensor networks to
+replace more passive approaches, and it may often make sense to
+include the sensor network into the application workflow pipeline, such as for Power Grids and
+Industrial Incident Response. All
+these choices have implications for design and provisioning. For instance,
+the choice of which sort of pipeline a particular application uses may
+well be fixed, or will change only slowly as the experimental
+environment evolves. For example, we see some systems that add simulation alongside experimental data collection to act
+as a verification and validation step.
+
+Revisiting the overall issue of infrastructure, which we defined as
+the hardware and software needed to run the applications, we
+note that the specific hardware is not a primary issue.  The
+applications we have discussed all use a mix of the systems that exist
+today, which have aspects of web, grid, and cloud, and can be
+considered a hybrid of the three.  The key infrastructure components
+are really the software and services on top of the hardware.  In this
+section, we have discussed infrastructure to support data management and
+data analysis, as well as the tools and systems used to program
+applications, and the specific components used for the application in
+\S\ref{sec:scenarios}.  However, there is not much commonality in the
+software and services that the applications use, so the software and programming infrastructure
+we have described here is much more broad than what is used by the
+individual applications we have studied.
+
+%``This paper presents the design and the implementation of an
+%architecture for the analysis of data streams in distributed
+%environments. In particular, data stream analysis has been carried out
+%for the computation of items and itemsets that exceed a frequency
+%threshold. The mining approach is hybrid, that is, frequent items are
+%calculated with a single pass, using a sketch algorithm, while
+%frequent itemsets are calculated offline by a further analysis. The
+%architecture combines parallel and distributed processing to keep the
+%pace with the rate of distributed data streams. In order to keep
+%computation close to data, miners are distributed among the domains
+%where data streams are generated. The paper also reports the
+%experiment results obtained with a prototype of the architecture,
+%tested on a Grid composed of two domains handling two different data
+%streams. Results confirm the scalability of the approach and show that
+%it is possible to keep the pace of data production and reduce the
+%communication overhead.''
+%
+%``.. In some cases streams of data must be analyzed in real time to
+%provide information about trends, outlier values or regularities that
+%must be signaled as soon as possible.  The need for online computation
+%is a notable challenge with respect to classical data mining
+%algorithms [1], [2].  Important application fields for stream mining
+%are as diverse as financial applications, network monitoring, security
+%problems, telecommunication networks, Web applications, sensor
+%networks, analysis of atmospheric data, etc.  A further difficulty
+%originates when streams are distributed, and mining models must be
+%derived not only for the data of a single stream, but for the
+%integration of multiple and heterogeneous data streams [3]. This
+%scenario can occur in all the application domains mentioned
+%before. For example, in a Content Distribution Network, user requests
+%delivered to a Web system can be forwarded to any of several servers
+%located in different and possibly distant places, in order to serve
+%requests more efficiently and balance the load.''
+%
+%``..the architecture combines the parallel and distributed
+%paradigms, the first to keep the pace with the rate of a
+%single data stream, by using multiple miners (processors
+%or cores), the second to cope with the distributed
+%nature of data streams. Miners are distributed among
+%the domains where data streams are generated, in order
+%to keep computation close to data.''
+%
+%
+%
+%FlumeJava~\cite{flumeJava} is a distributed, reliable, and available
+%service for efficiently collecting, aggregating, and moving large
+%amounts of log data. It has a simple and flexible architecture based
+%on streaming data flows. It is robust and fault tolerant with tunable
+%reliability mechanisms and many failover and recovery mechanisms. The
+%system is centrally managed and allows for intelligent dynamic
+%management. It uses a simple extensible data model that allows for
+%online analytic application.
+
+% Programming
+% Factored out by sd for various not-very-good reasons....
 %\input{3dpas_programming}
@@
 
-% \jhanote{Andre L suggests a schematic to capture the main ideas from %
-applications and the data properties. I think it is a good % idea. Omer: do you
-want to think about this as you review/refine % this section}
-
-%\jhanote{main message: (i) to application users/developers, you're not % alone,
-(ii) infrastructure scientists: think of abstractions, and % how to build tools,
-(iii) domain/application scientists message.}
+% \jhanote{Andre L suggests a schematic to capture the main ideas from
+%   applications and the data properties. I think it is a good
+%   idea. Omer: do you want to think about this as you review/refine
+%   this section}
+
+%\jhanote{main message: (i) to application users/developers, you're not
+%  alone, (ii) infrastructure scientists: think of abstractions, and
+%  how to build tools, (iii) domain/application scientists message.}
 
@@
 
-This section brings together some key observations from the applications and
-infrastructure analysis carried out in this paper.  %as take away messages.
-Section~\ref{sec:observations} summarize the dynamic and distributed
-characteristics of data associated with the applications we've studied in this
-work, attempting to infer more general characteristics that %can be applied may
-apply to other similar applications.  In \S\ref{sec:discussion}, we suggest two
-metrics that may be associated with such applications, primarily as a basis to
-compare how compute and data processing (in terms of their number and
-distribution) may be used to find common features between them. Such feature
-analysis is useful, we believe, for two main reasons: (i) to identify
-similarities between the applications covered in \S\ref{sec:scenarios} with
-others that are in the design phase or under development, so that common
-requirements can be identified; and (ii) to identify gaps in current software
-tools and programming systems that such applications use, thereby leading to the
-development of new tools and systems. As our application and infrastructure
-coverage is, by necessity, limited, identifying common features also enables the
-lessons learned to be generalized and applied more widely. We subsequently
-discuss a set of architectures for D3 science, and pose some questions we
-believe are pertinent as the community begins considering the next-generation
-architectures for D3 applications. In \S\ref{sec:conclusion} we conclude by
-providing a number of observations to inform applications and
+This section brings together some key observations
+from the applications and infrastructure
+analysis carried out in this paper.
+%as take away messages. 
+Section~\ref{sec:discussion}
+provides a discussion of dynamic and distributed characteristics of
+data associated with the applications we analyze in this work,
+attempting to infer more general characteristics that %can be applied
+may apply
+to other similar applications.  We also suggest two metrics that may
+be associated with such applications, primarily as a basis to compare
+how compute and data processing (in terms of their number and
+distribution) may be used to find common features between them. Such
+feature analysis is useful, we believe, for two main reasons: (i) to
+identify similarities between the applications covered in
+\S\ref{sec:scenarios} with others that are in the design phase or
+under development, so that common requirements can be identified; and
+(ii) to identify gaps in current software tools and programming
+systems that such applications use, thereby leading to the
+development of new tools and systems. As our application and
+infrastructure coverage is, by necessity, limited, identifying common
+features also enables the lessons learned to be generalized and
+applied more widely. We subsequently discuss a set of architectures
+for D3 science, and pose some questions we believe are pertinent as
+the community begins considering the next-generation
+architectures for D3 applications. In
+\S\ref{sec:conclusion} we conclude by providing a number of
+observations to inform applications and
 cyberinfrastructure developers, based on common trends found in our study.
 
-An important motivation for this work is to capture the state of the art in the
-design and development of distributed dynamic applications and the associated
-infrastructure currently used for their execution. We have aimed to try to
-discern common approaches and challenges, and to identify any obvious gaps (or
-opportunities). Using a common vocabulary and terminology to describe otherwise
-distinct and unrelated applications allows potential application
-users/developers to benefit from the insight that a common analytical framework
-provides. For example, seeing how applications with similar patterns (e.g.,
-pipelines) have addressed issues of dynamism and distribution enables common
-solutions and facilitates the emergence of best practices even in the absence of
-formal methods.  Similarly, our analysis sends a useful message to domain and
-application scientists that the challenges and barriers they face are not unique
-even if current solutions require a level of customization.
+An important motivation for this work is to capture the state of the
+art in the design and development of distributed dynamic applications
+and the associated infrastructure currently used for their
+execution. We have aimed to try to discern common
+approaches and challenges, and to identify any obvious gaps (or
+opportunities). Using a common vocabulary and terminology
+to describe otherwise distinct and unrelated applications allows
+potential application
+users/developers to benefit from the insight that a common
+analytical framework provides. For example, seeing how
+applications with similar patterns (e.g., pipelines) have addressed
+issues of dynamism and distribution enables common solutions and
+facilitates the emergence of best practices even in the absence of formal
+methods.  Similarly, our analysis sends a useful message to domain and
+application scientists that the challenges and barriers they face are not
+unique even if current solutions require a level of customization.
 
@@
 
-% Given the application descriptions in \S\ref{sec:scenarios} and % subsequent
-discussion about infrastructure characteristics in % \S\ref{sec:infrastructure},
-we identify features that may be used to % compare these applications. Such
-feature analysis is useful, we % believe, for two main reasons: (i) to identify
-similarities between % the applications covered in \S\ref{sec:scenarios} with
-others that are % in the design phase or under development, to enable common %
-requirements to be identified; (ii) to identify gaps in current % software tools
-and programming systems that such applications make use % of, thereby leading to
-the development of new tools and systems. As % our application and
-infrastructure coverage, by necessity, is limited, % identifying common features
-also enables the lessons learned to be % generalized and applied more
-widely. One important result is that we % can identify when particular
-applications suit particular data-centric % programming models, e.g., stream
-processing, MapReduce and Peer-2-Peer % networks.
-
-%\item mechanism for supporting in-situ analysis at point of data % capture
-(includes decision making to decide where analysis is % done)\note{move to
-vector section.}
+% Given the application descriptions in \S\ref{sec:scenarios} and
+% subsequent discussion about infrastructure characteristics in
+% \S\ref{sec:infrastructure}, we identify features that may be used to
+% compare these applications. Such feature analysis is useful, we
+% believe, for two main reasons: (i) to identify similarities between
+% the applications covered in \S\ref{sec:scenarios} with others that are
+% in the design phase or under development, to enable common
+% requirements to be identified; (ii) to identify gaps in current
+% software tools and programming systems that such applications make use
+% of, thereby leading to the development of new tools and systems. As
+% our application and infrastructure coverage, by necessity, is limited,
+% identifying common features also enables the lessons learned to be
+% generalized and applied more widely. One important result is that we
+% can identify when particular applications suit particular data-centric
+% programming models, e.g., stream processing, MapReduce and Peer-2-Peer
+% networks.
+
+%\item mechanism for supporting in-situ analysis at point of data
+%  capture (includes decision making to decide where analysis is
+%  done)\note{move to vector section.}
 %\end{itemize}
 
-%The first of the objectives above makes use of the four axes that %characterize
-applications in the DPA survey paper, the intention of %which was to ascertain
-how a number of real world applications made %use of distributed computing
-infrastructure. Although we consider a %different set of applications here, the
-aims are somewhat similar: to %both assess how existing real world applications
-utilize distributed %computing infrastructure and also how they adapt to various
-dynamic %characteristics that exist in their operating environment.
-
-%Table~\ref{tab:characteristics} compares the applications across the four axes
-identified at the beginning of this section.
-
-\subsection{Observations}
-\label{sec:observations}
-
-In this subsection we summarize the source and nature of dynamic data that we
-observe in the application surveyed in \S\ref{sec:scenarios}, and the types of
-data distribution that we observe in such applications.
-
-\subsubsection{Understanding Dynamic Data } %\alnote{in section 3 the three
-classes for dynamism were different}
-
-The applications in \S\ref{sec:scenarios} show that the dynamic characteristics
-of many applications arise in two ways, due to data generation and placement or
-due to the coordination mechanisms of the application.  The first of these
-refers to data dynamism and the second to application dynamism (outlined in
-\S\ref{sec:dynTypes}).
-
-%\jhanote{Omer: this is inconsistent -- at least in terminology with % the 3
-types of dynamism identified in Section 3: data, application % and
-infrastructure} \rananote{Shantenu: this has been updated to % align with the
-terminology in section 3. This is primarily referring % to the first two kinds
-of dynamism mentioned there.}
-
-\emph{Data generation, placement and consumption}: In many applications, data is
-dynamically generated in response to a user request or when a particular
-phenomenon of interest has been observed. Often this data is large in size and
-must be placed close to the user undertaking the analysis. In some instances, it
-may not be known where (and when) an analysis request may be made, and it is
-therefore necessary to identify how replicas of the data could be produced and
-placed across multiple possible locations.  In general, dynamic data arises due
-to changing data properties, changing data volume (generation, consumption, or
-processing), and changing data location or distribution.
-
-% The coordination process of managing interaction between % components
-undertaking data analysis, or the interaction of the % components and the
-execution environment in which they are % deployed.
-
-% Coordination mechanisms to support dynamic behavior remains an % important
-feature across D3 Science applications.
-
-\emph{Application Dynamism: } In many scientific applications the workflow
-structure is well defined, but either the components used within this structure
-or the location of the data that feeds the workflow may change dynamically.
-Data coordination and planning (identifying how data is managed from ingest
-through analysis to output) within such applications can range from statically
-defined interactions between execution units and data sources to a fully
-adaptable interaction using multi-agent systems.  As outlined in the Industrial
-Incident Notification and Response application in (\S\ref{Kees}), the workflow
-structure may also change based on a user request or the event being
-observed. Data relating to the coordination of a components that make up the
-application can therefore also be dynamic and change.  Exposing the data
-planning used within an application makes it possible to modify or adapt it over
-time.
-
-%In the Industrial Incident Notification and Response
-application\S\ref{sec:scenarios}.
-
-%\jhanote{AL and I think this needs attention still. Either discuss % under data
-planning or subsume dynamic coordination under % application dynamism}
-\rananote{Agree with Dan's paragraph below as % a replacement -- and agree that
-it should be subsumed under % application dynamism.}
-
-%\katznote{SJ to move this paragraph to under application dynamism, and % merge
-as needed.}\jhanote{done. NCH Can you check I have not mangled % the definition
-of data planning please?}
-
-%Exposing the data coordination/planning %used within an application (composed
-of a number of different %components) to can be beneficial.  By exposing the
-coordination %mechanism used, it becomes possible to modify or adapt it over
-time.
-
-%\jhanote{Omer: this is talking about changes in the infrastructure as % being
-the cause and coordination changes as being the effect. The % last sentence of
-the opening paragraph of this subsection, talks % about coordination being the
-cause} \rananote{Shantenu: reference to % infrastructure has been removed to
-avoid the confusion.}
-
-%Data coordination within such applications can range from statically %defined
-interactions between execution units and data sources to a %fully adaptable
-interaction using multi-agent systems, as outlined in %the Industrial Incident
-Notification and Response application in %\S\ref{sec:scenarios}.
-
-%\jhanote{Omer: Multi-agent systems are not % called-out/discussed as such in
-Section 2} \rananote{Shantenu: This % has been updated to refer to the
-particular application when agents % have been described -- rather than the
-section as a whole.}
-
-% Workflow techniques and tools provide one of the most widely used % approaches
-to expose such coordination to the user.
-
-% \item An analysis of the characteristics of the D3 Science % applications in
-\S\ref{sec:distdyndata} suggests that important % issues \jhanote{Important
-issues of coordination? distribution? or % dynamism?}  become evident when
-considering the ``data topology'' % of the application.
+%The first of the objectives above makes use of the four axes that
+%characterize applications in the DPA survey paper, the intention of
+%which was to ascertain how a number of real world applications made
+%use of distributed computing infrastructure. Although we consider a
+%different set of applications here, the aims are somewhat similar: to
+%both assess how existing real world applications utilize distributed
+%computing infrastructure and also how they adapt to various dynamic
+%characteristics that exist in their operating environment.
+
+%Table~\ref{tab:characteristics} compares the applications across the four axes identified at the beginning of this section.
+
+\subsection{Discussion}
+\label{sec:discussion}
+
+In this section we attempt to summarize four key themes from previous
+sections: (i) the source and nature of dynamic data that we observe in
+the application surveyed in \S\ref{sec:scenarios}; (ii) the types of
+data distribution that we observe in such applications; (iii) the
+general characteristics that can be associated with D3 science
+applications; (iv) the characteristics of a computational
+infrastructure that may be most suited for D3 science
+applications. Whereas (i) and (ii) are primarily observations that we
+can make on existing applications that we have discussed in
+\S\ref{sec:scenarios}, (iii) and (iv) attempt to synthesize general
+trends that can be applied to other applications we have not
+considered. \katznote{should this be split into two subsections then? Perhaps Observations \& Trends?}
+
+%\katznote{Some text is needed here to introduce the discussion subsection.}
+%\rananote{Added some text above -- does this make sense Dan?}
+
+\subsubsection{Understanding Dynamic Data }
+%\alnote{in section 3 the three classes for dynamism were different}
+
+The applications in \S\ref{sec:scenarios} show that the dynamic
+characteristics of many applications arise in two ways, due to data
+generation and placement or due to the coordination
+mechanisms of the application.
+The first of these refers to data dynamism and the second to
+application dynamism (outlined in \S\ref{sec:dynTypes}).
+
+%\jhanote{Omer: this is inconsistent -- at least in terminology with
+%  the 3 types of dynamism identified in Section 3: data, application
+%  and infrastructure} \rananote{Shantenu: this has been updated to
+%  align with the terminology in section 3. This is primarily referring
+%  to the first two kinds of dynamism mentioned there.}
+
+\emph{Data generation, placement and consumption}: In many
+applications, data is dynamically generated in response to a user
+request or when a particular phenomenon of interest has been
+observed. Often this data is large in size and must be placed close to
+the user undertaking the analysis. In some instances, it may not be
+known where (and when) an analysis request may be made, and it is
+therefore necessary to identify how replicas of the data could be
+produced and placed across multiple possible locations.  In general,
+dynamic data arises due to changing data properties, changing data
+volume (generation, consumption, or processing), and changing data
+location or distribution.
+
+% The coordination process of managing interaction between
+%   components undertaking data analysis, or the interaction of the
+%   components and the execution environment in which they are
+%   deployed.
+
+%   Coordination mechanisms to support dynamic behavior remains an
+%   important feature across D3 Science applications.
+
+\emph{Application Dynamism: } In many scientific applications the
+workflow structure is well defined, but either the components used
+within this structure or the location of the data that feeds the
+workflow may change dynamically.  Data coordination and planning
+(identifying how data is managed from ingest through analysis to
+output) within such applications can range from statically defined
+interactions between execution units and data sources to a fully
+adaptable interaction using multi-agent systems.  As outlined in the
+Industrial Incident Notification and Response application in
+(\S\ref{Kees}), the workflow structure may also change based on a user
+request or the event being observed. Data relating to the coordination
+of a components that make up the application can therefore also be
+dynamic and change.  Exposing the data planning used within an
+application makes it possible to modify or adapt it over time.
+
+%In the Industrial Incident Notification and Response application\S\ref{sec:scenarios}.
+
+%\jhanote{AL and I think this needs attention still. Either discuss
+%  under data planning or subsume dynamic coordination under
+%  application dynamism} \rananote{Agree with Dan's paragraph below as
+%  a replacement -- and agree that it should be subsumed under
+%  application dynamism.}
+
+%\katznote{SJ to move this paragraph to under application dynamism, and
+%  merge as needed.}\jhanote{done. NCH Can you check I have not mangled
+%  the definition of data planning please?}
+
+%Exposing the data coordination/planning
+%used within an application (composed of a number of different
+%components) to can be beneficial.  By exposing the coordination
+%mechanism used, it becomes possible to modify or adapt it over time.
+
+%\jhanote{Omer: this is talking about changes in the infrastructure as
+%  being the cause and coordination changes as being the effect. The
+%  last sentence of the opening paragraph of this subsection, talks
+%  about coordination being the cause} \rananote{Shantenu: reference to
+%  infrastructure has been removed to avoid the confusion.}
+
+%Data coordination within such applications can range from statically
+%defined interactions between execution units and data sources to a
+%fully adaptable interaction using multi-agent systems, as outlined in
+%the Industrial Incident Notification and Response application in
+%\S\ref{sec:scenarios}.
+
+%\jhanote{Omer: Multi-agent systems are not
+%  called-out/discussed as such in Section 2} \rananote{Shantenu: This
+%  has been updated to refer to the particular application when agents
+%  have been described -- rather than the section as a whole.}
+
+% Workflow techniques and tools provide one of the most widely used
+% approaches to expose such coordination to the user.
+
+% \item An analysis of the characteristics of the D3 Science
+%   applications in \S\ref{sec:distdyndata} suggests that important
+%   issues \jhanote{Important issues of coordination? distribution? or
+%     dynamism?}  become evident when considering the ``data topology''
+%   of the application.
 
@@
 
-From \S\ref{sec:distTypes} and \figurename~\ref{fig:figures_application}, we
-find that applications with distributed data differ based on the amount or
-volume of data distributed, the degree of distribution (scattered over one site
-or many sites), and the dynamic aspects associated with distribution, e.g.,
-latency and relative scheduling flexibility (and thus tolerance or constraints
-on data transfer/aggregation).  Applications processing the largest volumes of
-data (O(PB)) typically run on specialized data cyberinfrastructure (e.g., the
-Atlas LHC Grid). Applications such as Fusion and NGS that run on smaller data
-sets mainly utilize HPC environments. On the other end of the spectrum are
-highly dynamic applications that process sensor data and make near-realtime
-decisions based on streaming data, e.g., the power grid and sensor network
-applications. These applications process much less data; often, sensor data is
-also archived and aggregated for historical analyses, which leads to higher data
-volumes.
-
-\begin{figure}[ht] \centering
+From \S\ref{sec:distTypes} and
+\figurename~\ref{fig:figures_application}, we find that applications
+with distributed data differ based on the amount or volume of data
+distributed, the degree of distribution (scattered over one site or
+many sites), and the dynamic aspects associated with distribution,
+e.g., latency and relative scheduling flexibility (and thus tolerance
+or constraints on data transfer/aggregation).  Applications processing
+the largest volumes of data (O(PB)) typically run on specialized data
+cyberinfrastructure (e.g., the Atlas LHC Grid). Applications such as
+Fusion and NGS that run on smaller data sets mainly utilize HPC
+environments. On the other end of the spectrum are highly dynamic
+applications that process sensor data and make near-realtime
+decisions based on streaming data, e.g., the power grid and sensor
+network applications. These applications process much less data;
+often, sensor data is also archived and aggregated for historical
+analyses, which leads to higher data volumes. \katznote{not sure about the `other' part of this...}
+
+\begin{figure}[ht]
+	\centering
 		\includegraphics[width=.7\textwidth]{figures/application.pdf}
-                \caption{The distribution, dynamism, data volume characteristics
-of the surveyed D3 applications. The clustering of applications with high
-dynamism in the bottom right hand corner indicates a similar structure amongst
-these applications, namely multiple (sensor-based) sources of data generation
-with temporal variation in data generation. Although we do not explicitly
-color/shade traditional and infrastructural applications, it is worth mentioning
-that there is no correlation between application type and any of the three
-``D''s; we attribute this to a strong dependence of degree-of-distribution
-infrastructure available for both types of applications.
+                \caption{The distribution, dynamism, data volume
+                  characteristics of the surveyed D3 applications. The
+                  clustering of applications with high dynamism in the
+                  bottom right hand corner indicates a similar
+                  structure amongst these applications, namely
+                  multiple (sensor-based) sources of data
+                  generation with temporal variation in data
+                  generation. Although we do not explicitly
+                  color/shade traditional and infrastructural
+                  applications, it is worth mentioning that there is
+                  no correlation between application type and any of
+                  the three ``D''s; we attribute this to a strong
+                  dependence of degree-of-distribution infrastructure
+                  available for both types of applications.
                   \label{fig:figures_application}}
@@
 
-% \alnote{Think about how we can add a infrastructure dimension to this %
-figure.} \jhanote{As AL and I discussed on a call, in addition to % AL's
-suggestion, we need to also use this figure to integrate % section 2-4.}
-\jhanote{Also, how would we provide intuitive % explanation of high versus low
-dynamism?}  \jhanote{not to endlessly % quibble, but shouldn't ATLAS be more to
-the right (and higher % dynamism?) and NGS more the left?}
-
-%\jhanote{the following does not make sense to % me..}  Additional temporal
-aspects pertain to the stage(s) of an %application in which data can be
-distributed.  Because application stages often have a distribution that makes
-sense to that stage, and because that distribution may not be appropriate for
-all other phases, applications must either use one particular distribution that
-may not work well in all phases or transform the data on input to some stages.
-In particular, infrastructural applications require a specific distribution for
-the data storage, and thus the stages that read and process the data will need
-to make this choice.
-
-As seen, how application data is distributed depends upon which stage of an
-application is being considered. In spite of such variation, {\it data topology}
-can be used to identify the flow of data through the application, with nodes
-representing execution units and edges representing communication features.  %
-The overall shape of such a data connectivity pattern gives a sense % of {\it
-distributedness}; the thickness of edges gives a sense of % how data intensive
-the application is; and the change in the shape % and thickness (and rate) and
-number of edges gives a sense of the % dynamism of the application. \katznote{it
-seems odd that we describe % a picture here but don't actually have any such
-picture in the % paper.}  Additionally, how the application reacts to and
-handles dynamism is an important factor and may also indicate that some aspects
-of the data topology are unknown.
-
-One aspect of managing data topology is ``data planning," which refers to how
-the data is coordinated from ingest through analysis to output. In all cases,
-there is a coordination of the data with code used to analyze or transform
-it. In some cases these form clear, integrated stages where data flows from one
-to another. In others the way that data enters and flows through the application
-is more loosely defined and the execution of code is independent. Some examples
-include the ATLAS experiments, which have a single point of primary data
-generation which fans out to multiple analysis sites followed by mass exchange
-of derived data, and sensor grids, which have multiple sources of data
-generation collected to a single point of analysis followed by constrained point
-of reuse.
-
-We observe that pipelines provide a useful abstraction to represent the
-application stages identified in
-\figurename~\ref{fig:figures_application-stages}.  Many workflow tools also
-enable simple data capture and processing pipelines to be defined, where each
-pipeline stage may range from a complete application to a subroutine in Matlab
-(for instance).  It is useful to note that pipelines may also be dynamically
-constructed and adapted using context and monitoring information, as undertaken
-in the Industrial Incident Notification and Response application studied in this
-paper. In this instance, the actual pipeline stage to be invoked is triggered by
-the availability of particular types of data, a feature also relevant in the
-Power Grids application.  However, it is useful to note that different stages of
-the pipeline receive different emphases in different applications. Some
-applications must collect and store all incoming data, with only minor
-pre-processing to clean the data ahead of the main processing. For others, a
-major constraint is to reduce the incoming data stream as quickly as possible,
-perhaps discarding a large fraction of the data before any processing
-occurs. Reuse of pipeline stages across different applications must therefore be
-undertaken with care.
-
-Although dynamic and distributed properties have been characterized separately,
-it is important to remember, as the above patterns indicate, that distribution
-and dynamism become increasingly correlated at extreme scales.  For example,
-applications that run on distributed infrastructures are often also highly
-dynamic and vice versa.
-
-% \jhanote{There is a problem here. I am not saying the following is %
-incorrect, but ultimately coordination mechanisms are being touted % as a way to
-support dynamic behaviour if not respond to dynamic % behaviour. However in
-Section 3, this is not even hinted at. Needs % fixing IMO. The following should
-be trivial: we need to relate which % dynamic behaviour -- infrastructure,
-application or data dynamism, % is coordination influencing?}
-
-% \jhanote{this is the first occurence of topology. can this be removed?  % SJ
-to reduce last 2 paragraphs into a few sentences}
-
-% \rananote{Update description of data topology -- relate it to % pipelines and
-workflows mentioned earlier in the paper. Avoid % confusion in terminology}.
-\jhanote{Also can this be related to % section 5.2?}
-
-% Nonetheless, understanding the trend of say, $\beta_1$ as the degree % of
-distribution changes can be insightful.
-
-\subsection{Discussion}
-\label{sec:discussion}
-
-Now that we have discussed dynamic data and distributed data in the context of
-specific applications, we can use this experience to understand the
-characteristics that can be associated with D3 science applications, and the
-characteristics of a computational infrastructure that may be most suited for D3
-science applications.  We believe these are general trends that can be applied
-to other applications we have not considered.
+% \alnote{Think about how we can add a infrastructure dimension to this
+%   figure.} \jhanote{As AL and I discussed on a call, in addition to
+%   AL's suggestion, we need to also use this figure to integrate
+%   section 2-4.} \jhanote{Also, how would we provide intuitive
+%   explanation of high versus low dynamism?}  \jhanote{not to endlessly
+%   quibble, but shouldn't ATLAS be more to the right (and higher
+%   dynamism?) and NGS more the left?}
+
+%\jhanote{the following does not make sense to
+%  me..}  Additional temporal aspects pertain to the stage(s) of an
+%application in which data can be distributed.
+Because application stages often have a distribution that makes sense
+to that stage, and because that distribution may not be appropriate for all other
+phases, applications must either use one particular distribution
+that may not work well in all phases or transform the data on input to
+some stages.  In particular, infrastructural applications
+require a specific distribution for the data storage, and thus
+the stages that read and process the data will need to make this choice.
+
+As seen, how application data is distributed depends upon which stage
+of an application is being considered. In spite of such variation,
+{\it data topology} can be used to identify the flow of data through
+the application, with nodes representing execution units and edges
+representing communication features.
+% The overall shape of such a data connectivity pattern gives a sense
+% of {\it distributedness}; the thickness of edges gives a sense of
+% how data intensive the application is; and the change in the shape
+% and thickness (and rate) and number of edges gives a sense of the
+% dynamism of the application. \katznote{it seems odd that we describe
+%   a picture here but don't actually have any such picture in the
+%   paper.}
+Additionally, how the application reacts to and handles dynamism is an
+important factor and may also indicate that some aspects of the data
+topology are unknown.
+
+One aspect of managing data topology is ``data planning," which refers
+to how the data is coordinated from ingest through analysis to
+output. In all cases, there is a coordination of the data with code
+used to analyze or transform it. In some cases these form clear,
+integrated stages where data flows from one to another. In others the
+way that data enters and flows through the application is more loosely
+defined and the execution of code is independent. Some examples
+include the ATLAS experiments, which have a single point of primary
+data generation which fans out to multiple analysis sites followed by
+mass exchange of derived data, and sensor grids, which have multiple
+sources of data generation collected to a single point of analysis
+followed by constrained point of reuse.
+
+We observe that pipelines provide a useful abstraction to represent
+the application stages identified in
+\figurename~\ref{fig:figures_application-stages}.  Many workflow tools
+also enable simple data capture and processing pipelines to be
+defined, where each pipeline stage may range from a complete
+application to a subroutine in Matlab (for instance).  It is useful to
+note that pipelines may also be dynamically constructed and adapted
+using context and monitoring information, as undertaken in the
+Industrial Incident Notification and Response application studied in
+this paper. In this instance, the actual pipeline stage to be invoked
+is triggered by the availability of particular types of data, a
+feature also relevant in the Power Grids application.  However, it is
+useful to note that different stages of the pipeline receive different
+emphases in different applications. Some applications must collect and
+store all incoming data, with only minor pre-processing to clean the
+data ahead of the main processing. For others, a major constraint is
+to reduce the incoming data stream as quickly as possible, perhaps
+discarding a large fraction of the data before any processing
+occurs. Reuse of pipeline stages across different applications must
+therefore be undertaken with care.
+
+Although dynamic and distributed properties have been characterized
+separately, it is important to remember, as the above patterns
+indicate, that distribution and dynamism become increasingly
+correlated at extreme scales.  For example, applications that run on
+distributed infrastructures are often also highly dynamic and vice
+versa.
+
+% \jhanote{There is a problem here. I am not saying the following is
+%   incorrect, but ultimately coordination mechanisms are being touted
+%   as a way to support dynamic behaviour if not respond to dynamic
+%   behaviour. However in Section 3, this is not even hinted at. Needs
+%   fixing IMO. The following should be trivial: we need to relate which
+%   dynamic behaviour -- infrastructure, application or data dynamism,
+%   is coordination influencing?}
+
+% \jhanote{this is the first occurence of topology. can this be removed?
+%   SJ to reduce last 2 paragraphs into a few sentences}
+
+% \rananote{Update description of data topology -- relate it to
+%   pipelines and workflows mentioned earlier in the paper. Avoid
+%   confusion in terminology}.  \jhanote{Also can this be related to
+%   section 5.2?}
+
+% Nonetheless, understanding the trend of say, $\beta_1$ as the degree
+% of distribution changes can be insightful.
 
@@
 
-In \S\ref{sec:distdyndata} we characterized dynamic and distributed properties
-of the D3 Science applications. Our analysis was qualitative by necessity, due
-to the fact that approaches to quantify dynamism and distributed properties were
-often intimately tied to application/scenario definition and specific usage
-modes.  Notwithstanding, to quantify better dynamic properties of an
+In \S\ref{sec:distdyndata} we characterized dynamic and distributed
+properties of the D3 Science applications. Our analysis was
+qualitative by necessity, due to the fact that approaches to quantify
+dynamism and distributed properties were often intimately tied to
+application/scenario definition and specific usage modes.
+Notwithstanding, to quantify better dynamic properties of an
 application, we believe a term analogous to the use of Amdahl's
@@
 \begin{itemize}
-\item[] $\alpha$ = Num. of Compute Operations / Amount of Data (Consumed,
-Generated or Transformed)
+\item[] $\alpha$ = Num. of Compute Operations / Amount of Data
+  (Consumed, Generated or Transformed)
 \end{itemize}
 
-Note that this term is scale-invariant, i.e., a value of 1 could correspond to
-very small or very large values of both numerator and denominator, and that it
-does not capture the time scales over which data changes. However, we believe it
-captures most of the non-temporal {\it dynamic} properties.
-
-% \jhanote{we might consider removing the definitions. More text about % what is
-$\alpha$ and $\beta$ and say it is early work..}
+Note that this term is scale-invariant, i.e., a value of 1 could
+correspond to very small or very large values of both numerator and
+denominator, and that it does not capture the time scales over
+which data changes. However, we believe it captures most of the
+non-temporal {\it dynamic} properties.
+
+% \jhanote{we might consider removing the definitions. More text about
+%   what is $\alpha$ and $\beta$ and say it is early work..}
 
@@
 
-Similarly, another term can be used to quantify the ``scale'' of distribution,
-namely the degree of distribution as a function of the amount of data
-transferred, where the degree of distribution can be defined as the number of
-distinct data locations:
+Similarly, another term can be used to quantify the
+``scale'' of distribution, namely the degree of distribution as a
+function of the amount of data transferred, where the degree of
+distribution can be defined as the number of distinct data locations:
 
 \begin{itemize}
-\item[] $\beta$ = Amount of Data Transferred / Degree of Distribution
-% \item[] $\beta_2$ = Number of Compute Operations / Degree of % Distribution
+\item[] $\beta$ = Amount of Data Transferred / Degree of
+  Distribution
+% \item[] $\beta_2$ = Number of Compute Operations / Degree of
+%   Distribution
 \end{itemize}
 
-%Specifically, we propose $\beta$ as a ratio of the amount of data %transferred
-to the number of distinct sites (i.e., degree) that it was %distributed over.
-
-% either the computation involved or For a given application, the % degree of
-distribution is often dependent or correlated with the % total number of
-computational operations involved or the amount of % data transferred, e.g., for
-small computational operations or % data-volumes there is typically a limited
-amount of flexibility in % distribution before the overhead of distribution
-(either source or % destination) dominates.
-
-% and $\beta_2$ as a ratio of the number of compute operations to the % degree
-of distribution; as can be expected, for low amounts of % compute units the
-extent to which the tasks can be distributed is % limited, as overhead costs may
-dominate.
+%Specifically, we propose $\beta$ as a ratio of the amount of data
+%transferred to the number of distinct sites (i.e., degree) that it was
+%distributed over.
+
+% either the computation involved or For a given application, the
+% degree of distribution is often dependent or correlated with the
+% total number of computational operations involved or the amount of
+% data transferred, e.g., for small computational operations or
+% data-volumes there is typically a limited amount of flexibility in
+% distribution before the overhead of distribution (either source or
+% destination) dominates.
+
+% and $\beta_2$ as a ratio of the number of compute operations to the
+% degree of distribution; as can be expected, for low amounts of
+% compute units the extent to which the tasks can be distributed is
+% limited, as overhead costs may dominate.
 
@@
 
-It has proven difficult to provide ``global'' quantitative values of these
-ratios, as often the value of the numerators is subject to how an
-application/scenario is defined, even more so for a ``class of applications'' as
-opposed to a specific application. We believe, nonetheless, that these measures
-suggest a possible way to quantify an application's dynamism and distribution
-and thereby find possible common approaches solutions.  We plan to explore this
-in our future work, and also hope others will consider building on these initial
-concepts.
+It has proven difficult to provide ``global'' quantitative values of
+these ratios, as often the value of the numerators is
+subject to how an application/scenario is defined, even more so for a
+``class of applications'' as opposed to a specific application. We
+believe, nonetheless, that these measures suggest a possible way to
+quantify an application's dynamism and distribution and thereby find
+possible common approaches solutions.  We plan to explore this in our
+future work, and also hope others will consider building on these
+initial concepts.
 
@@
 A survey of data-intensive distributed applications on production
-cyberinfrastructure leads us to posit that there are three primary macroscopic
-architectures~\cite{pilot_clouds} and approaches for distributed data-intensive
-applications: (i) localize all data into a ``cloud'' (or a large analytical
-engine) -- the paradigm adopted by several genome projects and the cancer atlas,
-(ii) decompose and distribute data to an appropriate number of
-computing/analytical engines as available -- the paradigm employed by particle
-physics for the discovery of the Higgs Boson (similar to the application in
-\S\ref{WLCGSteve}), and (iii) a hybrid of the above two paradigms, wherein data
-is decomposed and committed to several infrastructure, which in turn could be a
-combination of either of the first two paradigms, similar to what is done in the
-Power grids application (\S\ref{power}). \jhanote{improve description....replace
-hybrid by bottom-up? }
-
-It is obvious that the first architecture can be used for large volumes of data,
-but surely there are limitations to the scalability or validity of this
-model. How does this limitation change, qualitatively or quantitatively, if
-dynamic data and/or applications are considered? This begets the question: how
-fundamental an issue is {\it dynamic}?
-
-The high-energy physics (HEP) community has self-organized towards the second
-architectural approach.  Although distributed computing infrastructures
-representing this approach, such as OSG and EGI, have been successful for this
-community and application, broad uptake across other application types, modes,
-and data volumes has not taken place. What lessons does the HEP experience teach
-us for the other communities, such as the bioinformatics?  Can an infrastructure
-conforming to a particular architecture and developed for a specific application
-domain be generalized to another science domain, using the same underlying macro
-architecture but by providing a different set of services and specialized data
-cyberinfrastructure? Or is the architecture specific to a {\it fixed} range of
-compute-data characteristics? If so, what are these characteristics?
+cyberinfrastructure leads us to posit that there are three primary
+macroscopic architectures~\cite{pilot_clouds} and approaches for distributed
+data-intensive applications: (i) localize all data into a ``cloud''
+(or a large analytical engine) -- the paradigm adopted by several
+genome projects and the cancer atlas, (ii) decompose and distribute
+data to an appropriate number of computing/analytical engines as
+available -- the paradigm employed by particle physics for the
+discovery of the Higgs Boson (similar to the application in
+\S\ref{WLCGSteve}), and (iii) a hybrid of the above two paradigms,
+wherein data is decomposed and committed to several infrastructure,
+which in turn could be a combination of either of the first two
+paradigms, similar to what is done in the Power grids application
+(\S\ref{power}). \jhanote{improve description....replace hybrid by
+  bottom-up? }
+
+It is obvious that the first architecture can be used for large
+volumes of data, but surely there are limitations to the scalability
+or validity of this model. How does this limitation change,
+qualitatively or quantitatively, if dynamic data and/or applications
+are considered? This begets the question: how fundamental an issue is
+{\it dynamic}?
+
+The high-energy physics (HEP) community has self-organized towards the
+second architectural approach.  Although distributed computing
+infrastructures representing this approach, such as OSG and EGI, have
+been successful for this community and application, broad uptake
+across other application types, modes, and data volumes has not taken
+place. What lessons does the HEP experience teach us for the other
+communities, such as the bioinformatics?  Can an infrastructure
+conforming to a particular architecture and developed for a specific
+application domain be generalized to another science domain, using the
+same underlying macro architecture but by providing a different set of
+services and specialized data cyberinfrastructure? Or is the
+architecture specific to a {\it fixed} range of compute-data
+characteristics? If so, what are these characteristics?
 
@@
 %
-%\katznote{some text needed here.  Note that I promoted this from 5.1.5 to 5.2,
-thinking that 5 might be best as discussion, observations, conclusions - how we
-order 5.1 and 5.2 is still a question... Maybe call this `takeaway points'?}
-%
-%\katznote{Omer to go through and move the text/points to either Discussion or
-Conclusions... This section will then empty/removed}
+%\katznote{some text needed here.
+Note that I promoted this from 5.1.5 to 5.2, thinking that 5 might be best as discussion, observations, conclusions - how we order 5.1 and 5.2 is still a question...
+Maybe call this `takeaway points'?}
+%
+%\katznote{Omer to go through and move the text/points to either Discussion or Conclusions... This section will then empty/removed}
 
@@
 
-%\item In \S\ref{sec:scenarios}, we divided applications into groups: %
-traditional and infrastructural applications.  Traditional % applications have
-been built by one or more authors in order to % solve a particular problem or to
-answer a science question.  % Infrastructural applications have been built by
-groups or % communities to allow the solution of a set of problems, or to answer
-% a set of science questions.  In both cases, we can observe that the % same set
-of applications stages (as shown in %
-\figurename~\ref{fig:figures_application-stages}) can be applied, % with APIs
-defined as the inputs and outputs of those stages.  These % APIs are very
-informal and loose for the traditional applications, % but more fixed and
-conservative for the infrastructural % applications.  These APIs limit how new
-paradigms can be used for % the infrastructural applications, unless those
-paradigms happen to % fit the existing APIs.  Another way of saying this is that
-the % abstractions for dynamism and distribution that can be used in the %
-infrastructural applications are those abstractions that the % developers of the
-original applications were aware of at the time % the infrastructural
-applications were defined.  This leads to a % possible mismatch, where computer
-scientists have new ideas that % could be applied to the class of problems that
-an infrastructural % application was designed to solve, but that the
-infrastructural % application doesn't allow, due to the fixed APIs.  Workflow
-systems % are a possible way of solving this problem, as they are often %
-sufficiently domain-independent that there does not seem to be a tie % of a
-particular system to a particular infrastructural application.
-
-%\item There exist patterns in the described applications; however, %
-implementation and deployment specific details often mask these % patterns.  For
-example, applications often differ in the way they % handle distribution and
-dynamism, such as managing such as % data/compute localities. This makes it
-difficult to support patterns % of distribution and dynamism in a
-general-purpose fashion, for % anything but the simplest patterns.
-%
-%\item It is useful to observe that pipelines provide a useful % abstraction to
-represent the application stages identified in %
-\figurename~\ref{fig:figures_application-stages}.  Many workflow % tools also
-enable simple data capture and processing pipelines to be % defined, where each
-pipeline stage may range from a complete % application to a subroutine in Matlab
-(for instance).  It is useful % to note that pipelines may also be dynamically
-constructed and % adapted using context and monitoring information, as
-undertaken in % the Industrial Incident Notification and Response application %
-studied in this paper. In this instance, the actual pipeline stage % to be
-invoked is triggered by the availability of particular types % of data, a
-feature also relevant in the Power Grids application.  % However, it is useful
-to note that different stages of the pipeline % receive different emphases in
-different applications. Some % applications must collect and store all incoming
-data, with only % minor pre-processing to clean the data ahead of the main %
-processing. For others, a major constraint is to reduce the incoming % data
-stream as quickly as possible, perhaps discarding a large % fraction of the data
-before any processing occurs. Reuse of pipeline % stages across different
-applications must therefore be undertaken % with care.
-
-  % Managing and coordinating the behavior of pipeline execution, for %
-supporting both file-based and streaming data, has been % extensively studied in
-computer science (supporting synchronous % (blocking), asynchronous, concurrent,
-threading semantics, for % instance).
+%\item In \S\ref{sec:scenarios}, we divided applications into groups:
+%  traditional and infrastructural applications.  Traditional
+%  applications have been built by one or more authors in order to
+%  solve a particular problem or to answer a science question.
+%  Infrastructural applications have been built by groups or
+%  communities to allow the solution of a set of problems, or to answer
+%  a set of science questions.  In both cases, we can observe that the
+%  same set of applications stages (as shown in
+%  \figurename~\ref{fig:figures_application-stages}) can be applied,
+%  with APIs defined as the inputs and outputs of those stages.  These
+%  APIs are very informal and loose for the traditional applications,
+%  but more fixed and conservative for the infrastructural
+%  applications.  These APIs limit how new paradigms can be used for
+%  the infrastructural applications, unless those paradigms happen to
+%  fit the existing APIs.  Another way of saying this is that the
+%  abstractions for dynamism and distribution that can be used in the
+%  infrastructural applications are those abstractions that the
+%  developers of the original applications were aware of at the time
+%  the infrastructural applications were defined.  This leads to a
+%  possible mismatch, where computer scientists have new ideas that
+%  could be applied to the class of problems that an infrastructural
+%  application was designed to solve, but that the infrastructural
+%  application doesn't allow, due to the fixed APIs.  Workflow systems
+%  are a possible way of solving this problem, as they are often
+%  sufficiently domain-independent that there does not seem to be a tie
+%  of a particular system to a particular infrastructural application.
+
+%\item There exist patterns in the described applications; however,
+%  implementation and deployment specific details often mask these
+%  patterns.  For example, applications often differ in the way they
+%  handle distribution and dynamism, such as managing such as
+%  data/compute localities. This makes it difficult to support patterns
+%  of distribution and dynamism in a general-purpose fashion, for
+%  anything but the simplest patterns.
+%
+%\item It is useful to observe that pipelines provide a useful
+%  abstraction to represent the application stages identified in
+%  \figurename~\ref{fig:figures_application-stages}.  Many workflow
+%  tools also enable simple data capture and processing pipelines to be
+%  defined, where each pipeline stage may range from a complete
+%  application to a subroutine in Matlab (for instance).  It is useful
+%  to note that pipelines may also be dynamically constructed and
+%  adapted using context and monitoring information, as undertaken in
+%  the Industrial Incident Notification and Response application
+%  studied in this paper. In this instance, the actual pipeline stage
+%  to be invoked is triggered by the availability of particular types
+%  of data, a feature also relevant in the Power Grids application.
+%  However, it is useful to note that different stages of the pipeline
+%  receive different emphases in different applications. Some
+%  applications must collect and store all incoming data, with only
+%  minor pre-processing to clean the data ahead of the main
+%  processing. For others, a major constraint is to reduce the incoming
+%  data stream as quickly as possible, perhaps discarding a large
+%  fraction of the data before any processing occurs. Reuse of pipeline
+%  stages across different applications must therefore be undertaken
+%  with care.
+
+  % Managing and coordinating the behavior of pipeline execution, for
+  % supporting both file-based and streaming data, has been
+  % extensively studied in computer science (supporting synchronous
+  % (blocking), asynchronous, concurrent, threading semantics, for
+  % instance).
 
@@
 
-%\jhanote{mismatch between abstraction levels, reference SAGA % paper ``leaky
-abstractions''} \katznote{Shantenu to incorporate this % into the leaky
-abstraction stuff above} \katznote{A possible problem is % that distributed
-systems (infrastructures and perhaps applications) % are so different from each
-other that there is no common system % abstraction that is both general enough
-to work on all of them and % specific enough to be useful on any of them.}
+%\jhanote{mismatch between abstraction levels, reference SAGA
+%  paper ``leaky abstractions''} \katznote{Shantenu to incorporate this
+%  into the leaky abstraction stuff above} \katznote{A possible problem is
+%  that distributed systems (infrastructures and perhaps applications)
+%  are so different from each other that there is no common system
+%  abstraction that is both general enough to work on all of them and
+%  specific enough to be useful on any of them.}
 
 %\item The trend toward more unstructured and more dynamic data sources lead to
-%higher demands on data processing infrastructure.  % provide simple
-abstractions or frameworks to % Higher-level abstraction need to provide
-sufficient level of control % of infrastructure trade-offs (such as data/compute
-localities), % while hiding a lot of the complexities (e.g.., the management of
-% replicas).\jhanote{andre to try to link back to section 4 and % elaborate}
-
-% This indicates the difficulty of % using existing abstractions and
-infrastructures.
+%higher demands on data processing infrastructure.
+%  provide simple abstractions or frameworks to
+%  Higher-level abstraction need to provide sufficient level of control
+%  of infrastructure trade-offs (such as data/compute localities),
+%  while hiding a lot of the complexities (e.g.., the management of
+%  replicas).\jhanote{andre to try to link back to section 4 and
+%    elaborate}
+
+% This indicates the difficulty of
+%  using existing abstractions and infrastructures.
 
@@
 
-% \item Often the coupling between the various components/services that % are
-part of the workflow can differ -- ranging from hardwired % interactions between
-components to those that can be defined and % modified by an application
-developer. We can therefore distinguish % between workflows that are embedded
-within an application, to those % that are explicitly defined using a
-composition tool (such as % Taverna, Triana, Kepler, etc). Workflow systems
-often utilize a % variety of abstractions related to: (i) mapping of %
-services/components to computational/data resources; (ii) data %
-movement/migration; and (iii) coordination between various % components.  And, %
-in many cases, they do not appear to be widely used where we might % think they
-would be, perhaps due to lack of publicity, steep % learning curves, or other
-issues.
+% \item Often the coupling between the various components/services that
+%   are part of the workflow can differ -- ranging from hardwired
+%   interactions between components to those that can be defined and
+%   modified by an application developer. We can therefore distinguish
+%   between workflows that are embedded within an application, to those
+%   that are explicitly defined using a composition tool (such as
+%   Taverna, Triana, Kepler, etc). Workflow systems often utilize a
+%   variety of abstractions related to: (i) mapping of
+%   services/components to computational/data resources; (ii) data
+%   movement/migration; and (iii) coordination between various
+%   components.  And,
+%   in many cases, they do not appear to be widely used where we might
+%   think they would be, perhaps due to lack of publicity, steep
+%   learning curves, or other issues.
 
@@
 
-Many fields, such as biology, that until recently were not characterized by
-their intensive use of data are rapidly becoming data-driven.  The growth in
-data has profound implications for the design and implementation of services and
-infrastructure.  For example, there is a lifecycle to data that has often been
-ignored until recently: the decreasing cost of storage has made it attractive to
-archive \emph{everything}, just in case it becomes useful. This approach is
-becoming untenable in many areas: some datasets are simply too big to archive,
-and so must be summarized and their raw form discarded (possibly in real
-time). Others are easier to regenerate than to store.  Still others are too
-transient to justify their storage, and may in aggregate be too big to store
-even when being small individually.
-
-Additionally, it is useful to observe that the trend toward more unstructured
-and more dynamic data sources lead to higher demands on data processing
-infrastructure.~\cite{snia_2012} Any data science architecture needs to
-explicitly recognize, capture, and manage the lifecycle requirements of
-different datasets, within the constraints of available time and
-storage. Associated with the growth in data volumes are increasing levels of
-dynamism and distribution; sometimes distribution is logical, but sometimes the
-data is physically distributed for reasons of performance or convenience. At
-large scales, dynamism becomes intrinsic to the application and infrastructure;
-sometimes it occurs because of changing experimental or analytic conditions,
-sometimes due to the availability of infrastructure and resources.  %We close
-out this paper with a brief %discussion of a range of issues associated with the
-distribution and %dynamism of data.
-
-We have identified many different application approaches, programming systems,
-and infrastructure techniques applicable to dynamic data.  Consistent with
-experience in other related fields, there seems no reason to expect that these
-approaches and techniques will disappear in the near future. While a convergence
-in approaches and techniques might be expected, this has proved to be elusive;
-there remains a diversity in the design, implementation, and execution of
-scientific codes.  It may even be argued that it is premature to have
-convergence; after all, D3 Science is still an emerging domain.
-
-From the previous discussion we can draw out some conclusions about the trends
-within dynamic, distributed, data-intensive science. For instance, in
-\S\ref{sec:scenarios}, we divided applications into groups: traditional and
-infrastructural applications.  Traditional applications have been built by one
-or more authors in order to solve a particular problem or to answer a science
-question.  Infrastructural applications have been built by groups or communities
-to allow the solution of a set of problems, or to answer a set of science
-questions.  In both cases, we can observe that the same set of applications
-stages (as shown in \figurename~\ref{fig:figures_application-stages}) can be
-applied, with APIs defined as the inputs and outputs of those stages.  These
-APIs are very informal and loose for the traditional applications, but more
-fixed and conservative for the infrastructural applications.  % \jhanote{is
-there a basis for these conjectures?} \katznote{just the % evidence of the
-applications themselves} These APIs limit how new application paradigms
-%\jhanote{new paradigms of what?}  can be introduced for the infrastructural
-applications, unless those paradigms happen to fit the existing APIs.  Another
-way of saying this is that the abstractions for dynamism and distribution that
-can be used in the infrastructural applications are those abstractions that the
-developers of the original applications were aware of at the time the
-infrastructural applications were defined.  This leads to a possible mismatch,
-where computer scientists have new ideas that could be applied to the class of
-problems that an infrastructural application was designed to solve, but that the
-infrastructural application doesn't allow, due to the fixed APIs.  New tools may
-be developed to address a narrow set of features while ignoring the larger set
-of problems that prevented earlier tools from being used.  These tools then may
-be used be new applications as those applications are developed; however, they
-are unlikely be used by the existing applications they were meant to help.
-
-This is because applications (particularly infrastructural applications) aren't
-developed in a vacuum, and that the constraints on the applications lead to new
-tools that are or could be developed not being able to be applied in practice.
-For example, while workflow systems are often sufficiently domain-independent
-that there does not seem to be a tie of a particular system to a particular
-infrastructural application, there are a very large number of workflow systems
-that exist, and it is common in the authors' experience to hear about new ones
-being developed from scratch because ``the ones that exist don't quite do what
-is needed for my application or system.''
-
-\katznote{previous 2 paragraphs to be re-read.  Old comments about them are
-available as commented out comments.}  %\jhanote{Irrespective, I find this past
-paragraph % somewhat speculative. And I don't see the point of it? WHat is the %
-latter half trying to say: there are problems with APIs and % interfaces of 3D
-applications and systems?} \katznote{It's saying % that applications aren't
-developed in a vacuum, particularly % infrastructural applications, and that the
-constraints on the % applications leads to new tools that are or could be
-developed not % being able to be applied in practice.  In some ways, it's a call
-for % an application abstraction or middleware, between the applications % and
-the `infrastructure' as defined in \S4, much as perhaps AIMES is % trying to
-build}\jhanote{Dan, I find your explanatory note more % convincing than the
-sentence on workflow systems! Also there is no % mention of middleware as the
-``impedance mismatch lowering layer'' % in the above. I propose we rewrite the
-sentence starting, ``Workflow % systems are a possible way..} \katznote{ok, I
-used some of my text % and turned around the sentence from Omer's optimistic
-view to % my more pessimistic view.  If others disagree, we can change it back %
-or compromise.}
-
-As part of this survey, we also observe that there exist patterns in the
-described applications; however, implementation and deployment specific details
-can often mask these patterns.  For example, applications often differ in the
-way they handle distribution and dynamism, such as managing such as data/compute
-localities. This makes it difficult to support patterns of distribution and
-dynamism in a general-purpose fashion, for anything but the simplest
-patterns. In \S\ref{sec:quantifyingdistapps} we have attempted to initiate a
-discussion in this context however by considering two ratios %(identified as
-parameters $\alpha$ and $\beta$) to relate data volume and distribution with
-compute operations performed by an application. We believe determining values
-for these parameters would provide a useful starting point for comparing
-infrastructure requirements of Big Data applications.  %\jhanote{I transposed
-the last two paragraphs}
-
-% \item Workflows seem to be a natural idea for expressing both % traditional
-and infrastructural applications, and indeed, there are % a number of workflow
-tools that are used in both cases.  It might % also seem that an infrastructural
-application would have a specific % workflow system that is in common use, since
-different % developers/users are all doing similar computational work, but this
+Many fields, such as biology, that until recently were not
+characterized by their intensive use of data are rapidly becoming
+data-driven.  The growth in data has profound implications for the
+design and implementation of services and infrastructure.  For
+example, there is a lifecycle to data that has often been ignored
+until recently: the decreasing cost of storage has made it attractive
+to archive \emph{everything}, just in case it becomes useful. This
+approach is becoming untenable in many areas: some datasets are simply
+too big to archive, and so must be summarized and their raw form
+discarded (possibly in real time). Others are easier to regenerate
+than to store.  Still others are too transient to justify their
+storage, and may in aggregate be too big to store even when being
+small individually.
+
+Additionally, it is useful to observe that the trend toward more
+unstructured and more dynamic data sources lead to higher demands on
+data processing infrastructure.~\cite{snia_2012}
+Any data science architecture needs to
+explicitly recognize, capture, and manage the lifecycle requirements
+of different datasets, within the constraints of available time and
+storage. Associated with the growth in data volumes are increasing
+levels of dynamism and distribution; sometimes distribution is
+logical, but sometimes the data is physically distributed for reasons
+of performance or convenience. At large scales, dynamism becomes
+intrinsic to the application and infrastructure; sometimes it occurs
+because of changing experimental or analytic conditions, sometimes due
+to the availability of infrastructure and resources.
+%We close out this paper with a brief
+%discussion of a range of issues associated with the distribution and
+%dynamism of data.
+
+We have identified many different application approaches, programming
+systems, and infrastructure techniques applicable to dynamic data.
+Consistent with experience in other related fields, there seems no
+reason to expect that these approaches and techniques will disappear
+in the near future. While a convergence in approaches and techniques
+might be expected, this has proved to be elusive; there remains a
+diversity in the design, implementation, and execution of scientific
+codes.  It may even be argued that it is premature to have convergence;
+after all, D3 Science is still an emerging domain.
+
+From the previous discussion we can draw out some conclusions about
+the trends within dynamic, distributed, data-intensive science. For
+instance, in \S\ref{sec:scenarios}, we divided applications into
+groups: traditional and infrastructural applications.  Traditional
+applications have been built by one or more authors in order to solve
+a particular problem or to answer a science question.  Infrastructural
+applications have been built by groups or communities to allow the
+solution of a set of problems, or to answer a set of science
+questions.  In both cases, we can observe that the same set of
+applications stages (as shown in
+\figurename~\ref{fig:figures_application-stages}) can be applied, with
+APIs defined as the inputs and outputs of those stages.  These APIs
+are very informal and loose for the traditional applications, but more
+fixed and conservative for the infrastructural applications.
+% \jhanote{is there a basis for these conjectures?} \katznote{just the
+%   evidence of the applications themselves}
+These APIs limit how new application paradigms %\jhanote{new paradigms of what?}
+can be introduced for the infrastructural applications, unless those
+paradigms happen to fit the existing APIs.  Another way of saying this
+is that the abstractions for dynamism and distribution that can be
+used in the infrastructural applications are those abstractions that
+the developers of the original applications were aware of at the time
+the infrastructural applications were defined.  This leads to a
+possible mismatch, where computer scientists have new ideas that could
+be applied to the class of problems that an infrastructural
+application was designed to solve, but that the infrastructural
+application doesn't allow, due to the fixed APIs.
+New tools may be developed to address a
+narrow set of features while ignoring the larger set of problems that
+prevented earlier tools from being used.  These tools then may be used
+be new applications as those applications are developed; however, they
+are unlikely be used by the existing applications they were meant to
+help.
+
+This is because applications
+(particularly infrastructural applications)
+aren't developed in a vacuum, and that the constraints on the
+applications lead to new tools that are or could be developed not
+being able to be applied in practice.  For example, while
+workflow systems are often sufficiently domain-independent that
+there does not seem to be a tie of a particular system to a particular
+infrastructural application, there are a very large number of workflow
+systems that exist, and it is common in the authors' experience to
+hear about new ones being developed from scratch because ``the
+ones that exist don't quite do what is needed for my application or system.''
+
+\katznote{previous 2 paragraphs to be re-read.  Old comments about them
+are available as commented out comments.}
+%\jhanote{Irrespective, I find this past paragraph
+%  somewhat speculative. And I don't see the point of it? WHat is the
+%  latter half trying to say: there are problems with APIs and
+%  interfaces of 3D applications and systems?} \katznote{It's saying
+%  that applications aren't developed in a vacuum, particularly
+%  infrastructural applications, and that the constraints on the
+%  applications leads to new tools that are or could be developed not
+%  being able to be applied in practice.  In some ways, it's a call for
+%  an application abstraction or middleware, between the applications
+%  and the `infrastructure' as defined in \S4, much as perhaps AIMES is
+%  trying to build}\jhanote{Dan, I find your explanatory note more
+%  convincing than the sentence on workflow systems! Also there is no
+%  mention of middleware as the ``impedance mismatch lowering layer''
+%  in the above. I propose we rewrite the sentence starting, ``Workflow
+%  systems are a possible way..} \katznote{ok, I used some of my text
+%  and turned around the sentence from Omer's optimistic view to
+%  my more pessimistic view.  If others disagree, we can change it back
+%  or compromise.}
+
+As part of this survey, we also observe that there exist patterns in
+the described applications; however, implementation and deployment
+specific details can often mask these patterns.  For example,
+applications often differ in the way they handle distribution and
+dynamism, such as managing such as data/compute localities. This makes
+it difficult to support patterns of distribution and dynamism in a
+general-purpose fashion, for anything but the simplest patterns. In
+\S\ref{sec:quantifyingdistapps} we have attempted to initiate a
+discussion in this context however by considering two
+ratios %(identified as parameters $\alpha$ and $\beta$)
+to relate data volume and distribution with compute operations
+performed by an application. We believe determining values for these
+parameters would provide a useful starting point for comparing
+infrastructure requirements of Big Data applications.
+%\jhanote{I transposed the last two paragraphs}
+
+% \item Workflows seem to be a natural idea for expressing both
+%   traditional and infrastructural applications, and indeed, there are
+%   a number of workflow tools that are used in both cases.  It might
+%   also seem that an infrastructural application would have a specific
+%   workflow system that is in common use, since different
+%   developers/users are all doing similar computational work, but this
 %   does not appear to be true, at least in the applications we studied.
 
-  % \katznote{I would like to say that specific workflow systems are %
-successful when embedded inside traditional applications, % because they then
-become part of the application, but I am not % sure I have a good example here
-from \S\ref{sec:scenarios}.  % Does anyone else?}
+  % \katznote{I would like to say that specific workflow systems are
+  %   successful when embedded inside traditional applications,
+  %   because they then become part of the application, but I am not
+  %   sure I have a good example here from \S\ref{sec:scenarios}.
+  %   Does anyone else?}
 
@@
 
-% \item \jhanote{Is the following observation, hypothesis and how do we % link
-them?} \katznote{Can we say something about applications that % deal with files
-vs those that deal with streams (stream-based are: % PowerGrid, Fusion, Incident
-Response, network monitoring) PowerGrid % and Incident apps have centralized
-stream processing;} Fusion and % network have distributed stream processing.
-Not sure what to say % here.  Maybe this point isn't a point... \jhanote{Look at
-tables % \ref{tab:app_distributed_processing} and %
-\ref{table:progSystems}. From these tables can we say something % about
-``common'' ways of say distribution for ``stream based % applications''}
-\jhanote{analogy with abstractions for workflows: % num. of data elements, num
-of streams, num. dynamic resources}
+% \item \jhanote{Is the following observation, hypothesis and how do we
+%   link them?} \katznote{Can we say something about applications that
+%   deal with files vs those that deal with streams (stream-based are:
+%   PowerGrid, Fusion, Incident Response, network monitoring) PowerGrid
+%   and Incident apps have centralized stream processing;} Fusion and
+%   network have distributed stream processing.  Not sure what to say
+%   here.  Maybe this point isn't a point... \jhanote{Look at tables
+%     \ref{tab:app_distributed_processing} and
+%     \ref{table:progSystems}. From these tables can we say something
+%     about ``common'' ways of say distribution for ``stream based
+%     applications''} \jhanote{analogy with abstractions for workflows:
+%     num. of data elements, num of streams, num. dynamic resources}
 
@@
 
-\section*{Acknowledgements} This paper is an outcome of the UK e-Science
-Institute sponsored Research Theme on Dynamic Distributed Data-Intensive
-Programming Abstractions and Systems. This work is also sponsored by NSF
+\section*{Acknowledgements}
+This paper is an outcome of the UK e-Science Institute sponsored
+Research Theme on Dynamic Distributed Data-Intensive Programming
+Abstractions and Systems. This work is also sponsored by NSF
 OCI-1059635.  We thank Jon Weissman and Jon Blower for useful initial
-discussions, as well as 3DPAS workshop attendees. Discussions of the selected
-applications benefited from the generous help of Adam Barker, Jon Blower, Julian
-Borrill, Silvia Delgado Olabarriaga, Steve Fisher, Keith Jackson, Scott Klasky,
-Bob Mann, Don Middleton, Kees Nieuwenhuis, Manish Parashar, Stephen Pascoe, and
-Jon Weissman, though any errors are the responsibility of the authors.  Some of
-the work by Katz was supported by the National Science Foundation while working
-at the Foundation; any opinion, finding, and conclusions or recommendations
-expressed in this material are those of the author(s) and do not necessarily
-reflect the views of the National Science Foundation.
-
-\bibliographystyle{unsrt} \bibliography{3dpas}
+discussions, as well as 3DPAS workshop attendees. Discussions of the
+selected applications benefited from the generous help of
+Adam Barker,
+Jon Blower,
+Julian Borrill,
+Silvia Delgado Olabarriaga,
+Steve Fisher,
+Keith Jackson,
+Scott Klasky,
+Bob Mann,
+Don Middleton,
+Kees Nieuwenhuis,
+Manish Parashar,
+Stephen Pascoe,
+and
+Jon Weissman,
+though any errors are the responsibility of the authors.
+Some of the work by Katz was
+supported by the National Science Foundation while working at the
+Foundation; any opinion, finding, and conclusions or recommendations
+expressed in this material are those of the author(s) and do not
+necessarily reflect the views of the National Science Foundation.
+
+\bibliographystyle{unsrt}
+\bibliography{3dpas}
 
@@
 
-{\em Workflow} systems have already received significant attention in the
-collaborative e-Science domain.  Such systems provide a useful means to
-aggregate capability and subsequently deploy the resulting workflow over
-multiple, distributed resources. Workflows can be ``abstract'', i.e., only the
-connectivity graph is identified but is not bound to particular executables and
-particular resources, or ``concrete'', i.e., the executable components are
-identified along with their execution environment. Generally, an abstract
-workflow needs to be mapped to a concrete workflow graph to enable
-execution. This may be achieved dynamically using a variety of different
-user-specified criteria, ranging from cost of execution (such as with the
-availability of remotely accessed resources from Amazon.com~\cite{berriman11}),
-Quality of Service, etc. This mapping can also be adaptive, especially if each
-concrete workflow is dynamically composed based on event triggers from the
-execution environment or constraints identified by a user. We may therefore
+{\em Workflow} systems have already received significant attention in
+the collaborative e-Science domain.  Such systems provide a useful
+means to aggregate capability and subsequently deploy the resulting
+workflow over multiple, distributed resources. Workflows can be
+``abstract'', i.e., only the connectivity graph is identified but is
+not bound to particular executables and particular resources, or
+``concrete'', i.e., the executable components are identified along
+with their execution environment. Generally, an abstract workflow
+needs to be mapped to a concrete workflow graph to enable
+execution. This may be achieved dynamically using a variety of
+different user-specified criteria, ranging from cost of execution
+(such as with the availability of remotely accessed resources from
+Amazon.com~\cite{berriman11}), Quality of Service, etc. This mapping
+can also be adaptive, especially if each concrete workflow is
+dynamically composed based on event triggers from the execution
+environment or constraints identified by a user. We may therefore
 identify a range of possibilities:
@@
 
-\item An abstract workflow graph is statically defined and mapped to a known
-resource/infrastructure pool.
-
-\item An abstract workflow graph is dynamically mapped to an infrastructure
-based on user-specified constraints (related to cost of access, Quality of
-Service, etc.)
-
-\item An abstract workflow graph is dynamically enacted over a varying
-infrastructure pool (often making use of an external resource management
-system.)
-
-\item An abstract workflow graph is used to dynamically compose a concrete
-workflow using techniques from AI planning and model checking.
+\item An abstract workflow graph is statically defined and mapped to a known resource/infrastructure pool.
+
+\item An abstract workflow graph is dynamically mapped to an infrastructure based on user-specified constraints (related to cost of access, Quality of Service, etc.)
+
+\item An abstract workflow graph is dynamically enacted over a varying infrastructure pool (often making use of an external resource management system.)
+
+\item An abstract workflow graph is used to dynamically compose a concrete workflow using techniques from AI planning and model checking.
 
@@
 Workflow tools such as Taverna~\cite{Taverna}, Kepler~\cite{Kepler},
-Trident~\cite{Trident} and others~\cite{iantaylorsbook} \jhanote{Are any of
-these used by applications? If not I propose they not be mentioned} enable
-reusable code for scientific analysis by expressing it as an explicit dataflow
-or control flow composition of components, each serving a dedicated
-function. This has simplified the development and the maintenance of complex
-codes, both by allowing common component re-use and by highlighting the
-structure of the processing pipeline. Workflow tools therefore provide both
-infrastructure for supporting the construction and enactment/execution of
-scientific applications and components that be used alongside application
-specific infrastructure. The granularity of a workflow tool can therefore range
-from simple components to complete infrastructures (i.e. enabling integration of
-other components and supporting their enactment/execution).
+Trident~\cite{Trident} and others~\cite{iantaylorsbook} \jhanote{Are
+  any of these used by applications? If not I propose they not be
+  mentioned} enable reusable code for scientific analysis by
+expressing it as an explicit dataflow or control flow composition of
+components, each serving a dedicated function. This has simplified the
+development and the maintenance of complex codes, both by allowing
+common component re-use and by highlighting the structure of the
+processing pipeline. Workflow tools therefore provide both
+infrastructure for supporting the construction and enactment/execution
+of scientific applications and components that be used alongside
+application specific infrastructure. The granularity of a workflow
+tool can therefore range from simple components to complete
+infrastructures (i.e. enabling integration of other components and
+supporting their enactment/execution).

### 2015-05-27 compare.tex
test for compare

--- 2015-05-27 3dpas_intro.YS.tex
@@
-% cpedoc.tex V2.0, 13 May 2010
-
-\documentclass[times]{cpeauth}
-
-\usepackage{moreverb}
-
-\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
-
-\newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08em
-T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
-
-\def\volumeyear{2012}
-%\documentclass[acmjacm,acmnow]{acmtrans2m}
-%\documentclass[conference,final]{IEEEtran}
-%\documentclass[10pt,letterpaper]{article}
-%\documentclass[preprint,12pt]{article}
-%\documentclass[11pt,letterpaper]{article}
-
-\usepackage{graphicx}
-\usepackage{url}
-\usepackage{color}
-\usepackage{enumerate}
-\usepackage{longtable}
-%\usepackage{authblk} % Is used internally within the CPE Class
-%\usepackage[firstpage]{draftwatermark}
-
-%\SetWatermarkLightness{0.7}
-%\SetWatermarkScale{4}
-
-%\usepackage{pdfsync}
-% \acmVolume{}
-% \acmNumber{}
-% \acmYear{}
-% \acmMonth{}
-
-%\usepackage{sectsty}
-%\usepackage{setspace}
-
-% \usepackage{ifpdf}
-% \usepackage{wrapfig}
-% %\usepackage[cm]{fullpage}
-% \usepackage{textcomp}
-% \usepackage{srcltx}
-% \usepackage{fancyhdr}
-% \usepackage{setspace}
-\usepackage{lscape}
-% \usepackage{longtable}
-% \usepackage{paralist}
-
-%\usepackage{hyperref}
-\usepackage[novbox]{pdfsync}
-
-% \usepackage{ams}
-
-% \ifpdf
-%              \usepackage[pdftex,
-%               colorlinks=true,
-%               linkcolor=red,
-%               citecolor=red,
-%               filecolor=red,
-%               ]{hyperref}
-% \else
-%   \usepackage[hypertex]{hyperref}
-% \fi
-
-% Space saving
-%\usepackage[small,compact]{titlesec}
-%\usepackage[small,it]{caption}
-
-% \renewcommand\floatpagefraction{.9}
-% \renewcommand\topfraction{.9}
-% \renewcommand\bottomfraction{.9}
-%\renewcommand\textfrare{.1}
-
-\setcounter{totalnumber}{50}
-\setcounter{topnumber}{50}
-\setcounter{bottomnumber}{50}
-
-%\textwidth = 6in
-%\textheight = 8.5 in
-%\oddsidemargin = 0.0 in
-%\evensidemargin = 0.0 in
-
-% \topmargin = 0.0 in
-% \headheight = 0.0 in
-% \headsep = 0.0 in
-%\parskip = 0.0in
-%\parindent = 0.5cm
-\parskip = 0.04in
-% \parindent = 0.0cm
-% \textfloatsep = 0.1in
-
-% \usepackage[draft,pdftex,colorlinks=true, linkcolor=blue,citecolor=blue,
-%        urlcolor=blue]{hyperref}
-
-% \ifpdf
-% \pdfinfo{ /Author (Shantenu Jha et al.)  /Title (Programming Abstractions for Large-Scale Distributed Applications) }
-% \fi
-
-%\newcommand{\projectnamefull}{\textit{Programming Abstractions for
-%    Large-Scale Distributed Applications} }
-
-\newcommand{\upup}{\vspace*{-0.5em}}
-\newcommand{\up}{\vspace*{-0.25em}}
-\newcommand{\I}[1]{\textit{#1}}
-\newcommand{\B}[1]{\textbf{#1}}
-\newcommand{\T}[1]{\texttt{#1}}
-\newcommand{\BI}[1]{\B{\I{#1}}}
-
- \newenvironment{shortlist}{
-        \vspace*{-0.8em}
-  \begin{itemize}
-  \setlength{\itemsep}{-0.3em}
-}{
-  \end{itemize}
-        \vspace*{-0.6em}
-}
-
- \newenvironment{shortenum}{
-        \vspace*{-0.8em}
-  \begin{enumerate}
-  \setlength{\itemsep}{-0.3em}
-}{
-  \end{enumerate}
-        \vspace*{-0.6em}
-}
-
-% \ifpdf
-%   \DeclareGraphicsExtensions{.pdf, .png, .jpg}
-% \else
-%   \DeclareGraphicsExtensions{.ps, .eps}
-% \fi
-
-\definecolor{orange}{rgb}{1.0,0.3,0.0}
-\definecolor{violet}{rgb}{0.75,0,1}
-\definecolor{darkgreen}{rgb}{0,0.6,0}
-\definecolor{cyan}{rgb}{0.2,0.7,0.7}
-\definecolor{blueish}{rgb}{0.2,0.2,0.8}
-
-\long\def\comment#1{{\bf \textcolor{magenta}{\bf #1}}}
-\long\def\ccomment#1{{\bf \textcolor{blue}{\bf #1}}}
-\newcommand{\C}{\comment}
-\newcommand{\CC}{\ccomment}
-
-\newcommand{\yes}{$\bullet$}
-
-\long\def\finalcomment#1{{\bf \textcolor{red}{\bf #1}}}
-\newcommand{\CF}{\finalcomment}
-
-\newif\ifdraft
-\drafttrue
-\ifdraft
- \newcommand{\katznote}[1]{ {\textcolor{magenta}    { ***Dan:      #1 }}}
- \newcommand{\rananote}[1]{ {\textcolor{cyan}    { ***Omer:     #1 }}}
- \newcommand{\omernote}[1]{ {\textcolor{cyan}    { ***Omer:     #1 }}}
- \newcommand{\note}[1]{ {\textcolor{red}    { #1 }}}
- \newcommand{\jhanote}[1]{  {\textcolor{blue}     { ***Shantenu: #1 }}}
- \newcommand{\jonnote}[1]{  {\textcolor{red}      { ***Jon: #1 }}}
- \newcommand{\alnote}[1]{  {\textcolor{darkgreen}      { ***Andre: #1 }}}
- \newcommand{\nchnote}[1]{  {\textcolor{orange}      { ***Neil: #1 }}}
- \newcommand{\ysnote}[1]{  {\textcolor{violet}      { ***Yogesh: #1 }}}
- \newcommand{\sdnote}[1]{  {\textcolor{blueish}      { ***Simon: #1 }}}
-% \usepackage[pdftex,colorlinks=true, linkcolor=blue,citecolor=blue,
-%       urlcolor=blue]{hyperref}
-\else
- \newcommand{\katznote}[1]{}
- \newcommand{\rananote}[1]{}
- \newcommand{\note}[1]{}
- \newcommand{\jhanote}[1]{}
- \newcommand{\jonnote}[1]{}
-\newcommand{\alnote}[1]{}
- \newcommand{\nchnote}[1]{}
- \newcommand{\ysnote}[1]{}
-\fi
+\include{header}
 
@@
 \affilnum{6}Cardiff University\break
-\affilnum{7}Indian Institute of Science\break
+\affilnum{7}University of Southern California\break
 }
@@
 into individual databases daily, and merges them with a master distributed-database on a weekly basis on an
-HPC cluster~\cite{Simmhan:Building:2009}. Data extraction and mining operations by astronomers are composed using a pseudo-SQL
+HPC cluster. Data extraction and mining operations by astronomers are composed using a pseudo-SQL
 language with user defined functions and submitted for execution on the master database using a batch system.
@@
 
-The application~\cite{modisAzure, modis:ipdps:2010} is used to handle the processes of downloading data from different
+The application~\cite{modisAzure} is used to handle the processes of downloading data from different
 sites and integrating them. The integration consists of reprojecting the data from one map projection to another,
@@
 based on the function provided, namely (i) infrastructure that supports {\it data management},
-(ii) infrastructure that supports {\it data analysis}, and (iii) infrastructure that supports the \emph{coordination within applications}.
+(ii) infrastructure that supports {\it data analysis}, and (iii) infrastructure that supports the structure of the applications.
 Each of these components may be present at three different levels of the infrastructure: programming frameworks, services and tools (e.g., MapReduce, workflow engines); middleware and platform services (e.g., databases, message queues); and system-level
@@
 {\em Data Sources} are used in the acquisition stage as inputs of data
-into the application. Broadly speaking these sources include types such as
+into the application. Broadly speaking they include types such as
 files, servers, sensors, instruments, simulations and models.
 Additionally, databases are often considered as an intermediate data
-source in some systems. The dynamic characteristics are primarily
+source in types of system. The dynamic characteristics are primarily
 due to availability. These are often accessed in the Data Generation stage of the application
@@
 of data, e.g., a field survey may bring in a lot of data during summer
-but none during other times. Some storage components like files and databases may also serve as data
-sources. These are commonly used in the Data Storage stage of the application (\figurename~\ref{fig:figures_application-stages}),
+but none during other times. These are commonly used in the Data Storage stage of the application (\figurename~\ref{fig:figures_application-stages}),
 but may also be present at the boundaries between the stages for transient staging of data.
@@
 movement of data between parts of the application, or to and from the
-application, % (often called staging) Data movement may be explicit
+application. % (often called staging) Data movement may be explicit
 and visible to the application (e.g., staging), or implicit (e.g.,
@@
 replicated close to the application). Data movement can be considered
-to be real-time (i.e., as it happens, e.g., sensor data), `as real-time'
+to be real-time (i.e., as it happens, e.g., sensor data), as real-time
 (e.g., a storage buffer has been introduced into the transport system,
@@
 characteristic of the data movement infrastructure depending on the
-application, e.g., real-time processing of sensor data for disaster
+application e.g. real-time processing of sensor data for disaster
 analysis. Other dynamic aspects of data movement include data rates,
@@
 applications stages and can be used by any of the
-stages to guide their selection or identification of incoming data sources and outgoing data sinks
-or storage.
+stages to guide their selection or identification of incoming data sources and outgoing data sinks.
 %\ysnote{rate of growth of data may be an issue.
@@
 \figurename~\ref{fig:figures_application-stages}, we identify a number
-of infrastructure services for data analysis that can be associated with these stages.
+of infrastructure services that can be associated with these stages.
 
@@
 toolboxes and visualization environments to become part of the infrastructure, some of which are
-deployed for specific domains but others which are more generic (e.g., Weka, GraphLab, Apache Mahout). These components are associated
+deployed for specific domains but others which are more generic (e.g., Weka, GraphLab). These components are associated
 with the Data Processing stage.
@@
 messaging to allow complex analyses to be performed on distributed
-data sources. These are often composed as workflows or Directed Acyclic Graphs (DAGs) that capture
-the execution dependency between
-components. They may also be dynamic if they have the capability to
+data sources. They may also be dynamic if they have the capability to
 adapt their behavior based on earlier parts of the workflow or can be
 steered based on external notification received during execution. In
-addition to supporting choreography, while often making use of one or more
+addition to supporting choreography, often making use of one or more
 resource manager/scheduler components, an orchestration component may also
@@
 collections of data which are executed in a loosely-coupled, data-parallel manner.
-The {\em MapReduce} paradigm~\cite{MapReduce} and its variants~\cite{twister,hmr} are the most
-prominent among these, with recent ones like Pregel and Apace Giraph~\cite{pregel} supporting specific domains like
-graph analytics. MapReduce forces programmers to model computation in two stages: 
+The {\em MapReduce} paradigm~\cite{MapReduce} and its variants~\cite{twister,hmr} are the most prominent among these. MapReduce forces programmers to model computation in two stages: 
 a \emph{map} phase in which a transformation is applied to every element of a collection independently of every other;  and a \emph{reduce} phase in which the transformed results 
@@
 Resource selection?}
-Infrastructure to support Dynamic Properties & Infrastructure to support Distributed Properties
+Infrastructure for Dynamic Properties & Infrastructure for Distributed Properties
  \\ \hline
@@
 		\hline	
-\multicolumn{2}{|l|}{{\textbf{Next Generation Sequencing (NGS) Analytics} (\S\ref{bioSilvia})}} \\ \hline	
+\multicolumn{2}{|l|}{{\textbf{Next Generation Sequencing (NGS) Analytics}}} \\ \hline	
 %\nchnote{For example, workload decomposition and distribution must be determined dynamically in order to optimally use resources, e.g., compute resource selection can be based on data location, processing profile and/or network capacity.
@@
 		\hline
-\multicolumn{2}{|l|}{{\textbf{ATLAS (an example of use of the WLCG) Reconstruction app} (\S\ref{WLCGSteve})}} \\ \hline
+\multicolumn{2}{|l|}{{\textbf{ATLAS (an example of use of the WLCG) Reconstruction app}}} \\ \hline
 Manage gradual growth of data using SRM \emph{(Data Storage, Access)}
@@
 		\hline
-\multicolumn{2}{|l|}{{\textbf{LSST} (\S\ref{astroAdam})}} \\ \hline
+\multicolumn{2}{|l|}{{\textbf{LSST}}} \\ \hline
 
@@
 		\hline
-\multicolumn{2}{|l|}{{\textbf{SOA Astronomy Applications} (\S\ref{astroSOAAdam})}} \\ \hline
+\multicolumn{2}{|l|}{{\textbf{SOA Astronomy Applications}}} \\ \hline
 
@@
 		\hline
-\multicolumn{2}{|l|}{{\textbf{Cosmic Microwave Background} (\S\ref{astroJulian})}} \\ \hline	
-	      Dynamic resource selection  \emph{(Data Discovery)}
+\multicolumn{2}{|l|}{{\textbf{Cosmic Microwave Background}}} \\ \hline	
+	      Dynamic resource selection
 %dynamic utilization of resources (4)
@@
 		\hline
-\multicolumn{2}{|l|}{{\textbf{Sensor Network Application} (\S\ref{sensorSimon})}} \\ \hline	
+\multicolumn{2}{|l|}{{\textbf{Sensor Network Application}}} \\ \hline	
 	     Notification of sensor data when in range, data rates and volumes \emph{(Data Movement, Notification)} %data rate, volume dynamic (1-2)
@@
 		\hline
-\multicolumn{2}{|l|}{{\textbf{Climate (ESGF)} (\S\ref{climateDan})}} \\ \hline	
+\multicolumn{2}{|l|}{{\textbf{Climate (ESGF)}}} \\ \hline	
 	      Manage data growth % New data generated over time (ESGF-wide - is this in scope of the application?)
@@
 		\hline
-\multicolumn{2}{|l|}{{\textbf{Interactive Exploration of Environmental Data} (\S\ref{envJon})}} \\ \hline	
+\multicolumn{2}{|l|}{{\textbf{Interactive Exploration of Environmental Data}}} \\ \hline	
 On-demand data subsetting, caching and versioning, real-time instrument data \emph{(Data Discovery, Data
@@
 	    %     \hline
-\multicolumn{2}{|l|}{{\textbf{PowerGrids} (\S\ref{power})}} \\ \hline
+\multicolumn{2}{|l|}{{\textbf{PowerGrids}}} \\ \hline
 
@@
 		\hline
-\multicolumn{2}{|l|}{{\textbf{Fusion (ITER)} (\S\ref{fusionScott})}} \\ \hline	
+\multicolumn{2}{|l|}{{\textbf{Fusion (ITER)}}} \\ \hline	
 	     Stream and transform data between simulation sites on-demand \emph{(Data Movement, Enrichment)}
@@
 		\hline
-\multicolumn{2}{|l|}{{\textbf{Industrial Incident Notification and Response} (\S\ref{Kees})}} \\ \hline	
+\multicolumn{2}{|l|}{{\textbf{Industrial Incident Notification and Response}}} \\ \hline	
 	     Acquire different data sources and rates, Processing determined by data
@@
 &
-Distributed data sources, Analysis by agents across institutions, Access control \emph{(Data Sources, Data
+Distributed data sources, Analysis by agents across institutions \emph{(Data Sources, Data
   Access)} % , Analysis
@@
 		\hline
-\multicolumn{2}{|l|}{{\textbf{MODIS Data Processing} (\S\ref{modisKeith})}} \\ \hline
+\multicolumn{2}{|l|}{{\textbf{MODIS Data Processing}}} \\ \hline
 
@@
 
-\multicolumn{2}{|l|}{{\textbf{Distributed Network Intrusion Detection} (\S\ref{detectionJon})}} \\ \hline	
+\multicolumn{2}{|l|}{{\textbf{Distributed Network Intrusion Detection}}} \\ \hline	
 Analyze dynamic network traffic rates \emph{(Data Movement)} % , Analysis
@@
         \hline LSST & custom layer; scripted workflows; queries \\
-        \hline SOA Astronomy & web services~\cite{barker} \\
+        \hline SOA Astronomy & web services \\
         \hline Cosmic Microwave Background & \textit{ad hoc} \\
@@
         \hline Interactive Exploration of Environmental Data &
-        visualization tools~\cite{blower2} \\
-        \hline Power Grids & stream processing~\cite{Simmhan:sciencecloud:2011}; data mining~\cite{Yin:mapreduce:2012} \\
-        \hline Fusion (ITER) & scripted pipeline~\cite{adios} \\
+        visualization tools \\
+        \hline Power Grids & stream processing; data mining \\
+        \hline Fusion (ITER) & scripted pipeline \\
         \hline Industrial Incident Notification and Response &
-        agent-based~\cite{emergency4}  \\
-        \hline MODIS Data Processing & scripted pipeline~\cite{modis:ipdps:2010} \\
+        agent-based  \\
+        \hline MODIS Data Processing & scripted pipeline \\
 %        \hline Floating Sensors & \textit{ad hoc} \\
@@
 the hardware and software needed to run the applications, we
-note that the specific hardware is not a primary issue.  The
-applications we have discussed all use a mix of the systems that exist
+note that the specific hardware isn't a primary issue.  The
+applications we've discussed all use a mix of the systems that exist
 today, which have aspects of web, grid, and cloud, and can be
@@
 are really the software and services on top of the hardware.  In this
-section, we have discussed infrastructure to support data management and
+section, we've discussed infrastructure to support data management and
 data analysis, as well as the tools and systems used to program
@@
 software and services that the applications use, so the software and programming infrastructure
-we have described here is much more broad than what is used by the
-individual applications we have studied.
+we've described here is much more broad than what is used by the
+applications we've studied.
 
@@
 %as take away messages. 
-Section~\ref{sec:discussion}
-provides a discussion of dynamic and distributed characteristics of
-data associated with the applications we analyze in this work,
+Section~\ref{sec:observations}
+summarize the dynamic and distributed characteristics of
+data associated with the applications we've studied in this work,
 attempting to infer more general characteristics that %can be applied
 may apply
-to other similar applications.  We also suggest two metrics that may
+to other similar applications.  In \S\ref{sec:discussion}, we
+suggest two metrics that may
 be associated with such applications, primarily as a basis to compare
@@
 
-\subsection{Discussion}
-\label{sec:discussion}
-
-In this section we attempt to summarize four key themes from previous
-sections: (i) the source and nature of dynamic data that we observe in
-the application surveyed in \S\ref{sec:scenarios}; (ii) the types of
-data distribution that we observe in such applications; (iii) the
-general characteristics that can be associated with D3 science
-applications; (iv) the characteristics of a computational
-infrastructure that may be most suited for D3 science
-applications. Whereas (i) and (ii) are primarily observations that we
-can make on existing applications that we have discussed in
-\S\ref{sec:scenarios}, (iii) and (iv) attempt to synthesize general
-trends that can be applied to other applications we have not
-considered. \katznote{should this be split into two subsections then? Perhaps Observations \& Trends?}
-
-%\katznote{Some text is needed here to introduce the discussion subsection.}
-%\rananote{Added some text above -- does this make sense Dan?}
+\subsection{Observations}
+\label{sec:observations}
+
+In this subsection we summarize the source and nature of dynamic data that we observe in
+the application surveyed in \S\ref{sec:scenarios}, and the types of
+data distribution that we observe in such applications.
 
@@
 often, sensor data is also archived and aggregated for historical
-analyses, which leads to higher data volumes. \katznote{not sure about the `other' part of this...}
+analyses, which leads to higher data volumes.
 
@@
 % of distribution changes can be insightful.
+
+\subsection{Discussion}
+\label{sec:discussion}
+
+Now that we have discussed dynamic data and distributed data in the
+context of specific applications, we can use this experience to
+understand the characteristics that can be associated with D3 science
+applications, and the characteristics of a computational
+infrastructure that may be most suited for D3 science
+applications. 
+We believe these are general
+trends that can be applied to other applications we have not
+considered.
 

### 2015-05-27 3dpas_intro.tex
merged in YS changes:
replaced <CR> with <LF>
affiliation
added citations
modifactions mainly to infrastructure section

@@
-% cpedoc.tex V2.0, 13 May 2010
-
-\documentclass[times]{cpeauth}
-
-\usepackage{moreverb}
-
-\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
-
-\newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08em
-T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
-
-\def\volumeyear{2012} %\documentclass[acmjacm,acmnow]{acmtrans2m}
-%\documentclass[conference,final]{IEEEtran}
-%\documentclass[10pt,letterpaper]{article}
-%\documentclass[preprint,12pt]{article}
-%\documentclass[11pt,letterpaper]{article}
-
-\usepackage{graphicx} \usepackage{url} \usepackage{color} \usepackage{enumerate}
-\usepackage{longtable} %\usepackage{authblk} % Is used internally within the CPE
-Class \usepackage[firstpage]{draftwatermark}
-
-%\SetWatermarkLightness{0.7} %\SetWatermarkScale{4}
-
-%\usepackage{pdfsync} % \acmVolume{} % \acmNumber{} % \acmYear{} % \acmMonth{}
-
-%\usepackage{sectsty} %\usepackage{setspace}
-
-% \usepackage{ifpdf} % \usepackage{wrapfig} % %\usepackage[cm]{fullpage} %
-\usepackage{textcomp} % \usepackage{srcltx} % \usepackage{fancyhdr} %
-\usepackage{setspace} \usepackage{lscape} % \usepackage{longtable} %
-\usepackage{paralist}
-
-%\usepackage{hyperref} \usepackage[novbox]{pdfsync}
-
-% \usepackage{ams}
-
-% \ifpdf % \usepackage[pdftex, % colorlinks=true, % linkcolor=red, %
-citecolor=red, % filecolor=red, % ]{hyperref} % \else %
-\usepackage[hypertex]{hyperref} % \fi
-
-% Space saving %\usepackage[small,compact]{titlesec}
-%\usepackage[small,it]{caption}
-
-% \renewcommand\floatpagefraction{.9} % \renewcommand\topfraction{.9} %
-\renewcommand\bottomfraction{.9} %\renewcommand\textfrare{.1}
-
-\setcounter{totalnumber}{50} \setcounter{topnumber}{50}
-\setcounter{bottomnumber}{50}
-
-%\textwidth = 6in %\textheight = 8.5 in %\oddsidemargin = 0.0 in
-%\evensidemargin = 0.0 in
-
-% \topmargin = 0.0 in % \headheight = 0.0 in % \headsep = 0.0 in
-%\parskip = 0.0in
-%\parindent = 0.5cm
-\parskip = 0.04in
-% \parindent = 0.0cm % \textfloatsep = 0.1in
-
-% \usepackage[draft,pdftex,colorlinks=true, linkcolor=blue,citecolor=blue, %
-urlcolor=blue]{hyperref}
-
-% \ifpdf % \pdfinfo{ /Author (Shantenu Jha et al.)  /Title (Programming
-Abstractions for Large-Scale Distributed Applications) } % \fi
-
-%\newcommand{\projectnamefull}{\textit{Programming Abstractions for %
-Large-Scale Distributed Applications} }
-
-\newcommand{\upup}{\vspace*{-0.5em}} \newcommand{\up}{\vspace*{-0.25em}}
-\newcommand{\I}[1]{\textit{#1}} \newcommand{\B}[1]{\textbf{#1}}
-\newcommand{\T}[1]{\texttt{#1}} \newcommand{\BI}[1]{\B{\I{#1}}}
-
- \newenvironment{shortlist}{ \vspace*{-0.8em}
-  \begin{itemize} \setlength{\itemsep}{-0.3em} }{
-  \end{itemize} \vspace*{-0.6em} }
-
- \newenvironment{shortenum}{ \vspace*{-0.8em}
-  \begin{enumerate} \setlength{\itemsep}{-0.3em} }{
-  \end{enumerate} \vspace*{-0.6em} }
-
-% \ifpdf % \DeclareGraphicsExtensions{.pdf, .png, .jpg} % \else %
-\DeclareGraphicsExtensions{.ps, .eps} % \fi
-
-\definecolor{orange}{rgb}{1.0,0.3,0.0} \definecolor{violet}{rgb}{0.75,0,1}
-\definecolor{darkgreen}{rgb}{0,0.6,0} \definecolor{cyan}{rgb}{0.2,0.7,0.7}
-\definecolor{blueish}{rgb}{0.2,0.2,0.8}
-
-\long\def\comment#1{{\bf \textcolor{magenta}{\bf #1}}} \long\def\ccomment#1{{\bf
-\textcolor{blue}{\bf #1}}} \newcommand{\C}{\comment} \newcommand{\CC}{\ccomment}
-
-\newcommand{\yes}{$\bullet$}
-
-\long\def\finalcomment#1{{\bf \textcolor{red}{\bf #1}}}
-\newcommand{\CF}{\finalcomment}
-
-\newif\ifdraft \drafttrue \ifdraft \newcommand{\katznote}[1]{
-{\textcolor{magenta} { ***Dan: #1 }}} \newcommand{\rananote}[1]{
-{\textcolor{cyan} { ***Omer: #1 }}} \newcommand{\omernote}[1]{ {\textcolor{cyan}
-{ ***Omer: #1 }}} \newcommand{\note}[1]{ {\textcolor{red} { #1 }}}
-\newcommand{\jhanote}[1]{ {\textcolor{blue} { ***Shantenu: #1 }}}
-\newcommand{\jonnote}[1]{ {\textcolor{red} { ***Jon: #1 }}}
-\newcommand{\alnote}[1]{ {\textcolor{darkgreen} { ***Andre: #1 }}}
-\newcommand{\nchnote}[1]{ {\textcolor{orange} { ***Neil: #1 }}}
-\newcommand{\ysnote}[1]{ {\textcolor{violet} { ***Yogesh: #1 }}}
-\newcommand{\sdnote}[1]{ {\textcolor{blueish} { ***Simon: #1 }}} %
-\usepackage[pdftex,colorlinks=true, linkcolor=blue,citecolor=blue, %
-urlcolor=blue]{hyperref} \else \newcommand{\katznote}[1]{}
-\newcommand{\rananote}[1]{} \newcommand{\note}[1]{} \newcommand{\jhanote}[1]{}
-\newcommand{\jonnote}[1]{} \newcommand{\alnote}[1]{} \newcommand{\nchnote}[1]{}
-\newcommand{\ysnote}[1]{} \fi
-
-\begin{document} \runningheads{S.~Jha et al.}{Understanding Applications and
-Infrastructure for D3 Science}
-
-%\title{3DPAS: Distributed Dynamic Data-intensive Programming % Abstractions and
-Systems for Large-Scale Science and Engineering}
+\include{header}
+
+\begin{document}
+\runningheads{S.~Jha et al.}{Understanding Applications and Infrastructure for D3 Science}
+
+%\title{3DPAS: Distributed Dynamic Data-intensive Programming
+%  Abstractions and Systems for Large-Scale Science and Engineering}
 
@@
 
-%\alnote{relate infrastructures back to applications: in particular, how do
-specific purpose built infrastructures as the LHCG or ESGF fit here?}
-
-We define {\em infrastructure} as the hardware and software that is provided to
-support an application, rather than being explicitly created or hosted by the
-user themselves. In many cases the precise boundary between infrastructure and
-application is blurred, e.g., a catalog is a component with functionality that
-may be similar to that of similar components in other applications, but is often
-implemented specifically by an application rather than being provided as part of
+%\alnote{relate infrastructures back to applications: in particular, how do specific purpose built infrastructures as the LHCG or ESGF fit here?}
+
+We define {\em infrastructure} as the hardware and software that is
+provided to support an application, rather than being explicitly
+created or hosted by the user themselves. In many cases the precise boundary between
+infrastructure and application is blurred, e.g., a catalog is
+a component with functionality that may be similar to that of similar
+components in other applications, but is often implemented
+specifically by an application rather than being provided as part of
 the infrastructure. This may also be the case when reconfiguring an
 infrastructure service for an application, when it is more onerous than
-reimplementing the service specifically for that application. We also introduce
-the concept of an ``infrastructure'' application as a collection of
-infrastructure components that are packaged so that they can be used by many
-applications in defined ways. The difference between these and ``traditional''
-applications is often delineated by who owns and administers the application,
-with traditional applications ``owned'' entirely by the user, and infrastructure
-applications owned by someone other than the user but provided for use by
-others. Infrastructure can be specialist as opposed to generic, for instance the
-purpose-built infrastructure in ESGF (\S\ref{climateDan}) that is provided to
-support the climate community.
-
-In this section, we focus on the {\em software} infrastructure that is provided
-to support D3 applications, specifically components that deal with distributed
-or dynamic data. The components that make up the infrastructure can be
-categorized based on the function provided, namely (i) infrastructure that
-supports {\it data management}, (ii) infrastructure that supports {\it data
-  analysis}, and (iii) infrastructure that supports the \emph{coordination
-  within applications}. Each of these components may be present at three
-different levels of the infrastructure: programming frameworks, services and
-tools (e.g., MapReduce, workflow engines); middleware and platform services
-(e.g., databases, message queues); and system-level software capabilities (e.g.,
-notifications, file system consistency).
+reimplementing the service specifically for that
+application. We also introduce the concept of an ``infrastructure'' application as a collection
+of infrastructure components that are packaged so that they can be used by many
+applications in defined ways. The difference between these and ``traditional'' applications
+is often delineated by who owns and administers the application, with traditional applications
+``owned'' entirely by the user, and infrastructure applications owned by someone other than the
+user but provided for use by others. Infrastructure can be specialist as opposed to generic, for instance
+the purpose-built infrastructure in ESGF (\S\ref{climateDan}) that is provided to support the climate community.
+
+In this section, we focus on the {\em software} infrastructure that is
+provided to support D3 applications, specifically components that deal
+with distributed or dynamic data. The components that make up the infrastructure can be categorized
+based on the function provided, namely (i) infrastructure that supports {\it data management},
+(ii) infrastructure that supports {\it data analysis}, and (iii) infrastructure that supports the structure of the applications.
+Each of these components may be present at three different levels of the infrastructure: programming frameworks, services and tools (e.g., MapReduce, workflow engines); middleware and platform services (e.g., databases, message queues); and system-level
+software capabilities (e.g., notifications, file system consistency).
 
@@
 
-This section brings together some key observations from the applications and
-infrastructure analysis carried out in this paper.  %as take away messages.
-Section~\ref{sec:observations} summarize the dynamic and distributed
-characteristics of data associated with the applications we've studied in this
-work, attempting to infer more general characteristics that %can be applied may
-apply to other similar applications.  In \S\ref{sec:discussion}, we suggest two
-metrics that may be associated with such applications, primarily as a basis to
-compare how compute and data processing (in terms of their number and
-distribution) may be used to find common features between them. Such feature
-analysis is useful, we believe, for two main reasons: (i) to identify
-similarities between the applications covered in \S\ref{sec:scenarios} with
-others that are in the design phase or under development, so that common
-requirements can be identified; and (ii) to identify gaps in current software
-tools and programming systems that such applications use, thereby leading to the
-development of new tools and systems. As our application and infrastructure
-coverage is, by necessity, limited, identifying common features also enables the
-lessons learned to be generalized and applied more widely. We subsequently
-discuss a set of architectures for D3 science, and pose some questions we
-believe are pertinent as the community begins considering the next-generation
-architectures for D3 applications. In \S\ref{sec:conclusion} we conclude by
-providing a number of observations to inform applications and
+This section brings together some key observations
+from the applications and infrastructure
+analysis carried out in this paper.
+%as take away messages. 
+Section~\ref{sec:discussion}
+provides a discussion of dynamic and distributed characteristics of
+data associated with the applications we analyze in this work,
+attempting to infer more general characteristics that %can be applied
+may apply
+to other similar applications.  We also suggest two metrics that may
+suggest two metrics that may
+be associated with such applications, primarily as a basis to compare
+how compute and data processing (in terms of their number and
+distribution) may be used to find common features between them. Such
+feature analysis is useful, we believe, for two main reasons: (i) to
+identify similarities between the applications covered in
+\S\ref{sec:scenarios} with others that are in the design phase or
+under development, so that common requirements can be identified; and
+(ii) to identify gaps in current software tools and programming
+systems that such applications use, thereby leading to the
+development of new tools and systems. As our application and
+infrastructure coverage is, by necessity, limited, identifying common
+features also enables the lessons learned to be generalized and
+applied more widely. We subsequently discuss a set of architectures
+for D3 science, and pose some questions we believe are pertinent as
+the community begins considering the next-generation
+architectures for D3 applications. In
+\S\ref{sec:conclusion} we conclude by providing a number of
+observations to inform applications and
 cyberinfrastructure developers, based on common trends found in our study.
 
-An important motivation for this work is to capture the state of the art in the
-design and development of distributed dynamic applications and the associated
-infrastructure currently used for their execution. We have aimed to try to
-discern common approaches and challenges, and to identify any obvious gaps (or
-opportunities). Using a common vocabulary and terminology to describe otherwise
-distinct and unrelated applications allows potential application
-users/developers to benefit from the insight that a common analytical framework
-provides. For example, seeing how applications with similar patterns (e.g.,
-pipelines) have addressed issues of dynamism and distribution enables common
-solutions and facilitates the emergence of best practices even in the absence of
-formal methods.  Similarly, our analysis sends a useful message to domain and
-application scientists that the challenges and barriers they face are not unique
-even if current solutions require a level of customization.
+An important motivation for this work is to capture the state of the
+art in the design and development of distributed dynamic applications
+and the associated infrastructure currently used for their
+execution. We have aimed to try to discern common
+approaches and challenges, and to identify any obvious gaps (or
+opportunities). Using a common vocabulary and terminology
+to describe otherwise distinct and unrelated applications allows
+potential application
+users/developers to benefit from the insight that a common
+analytical framework provides. For example, seeing how
+applications with similar patterns (e.g., pipelines) have addressed
+issues of dynamism and distribution enables common solutions and
+facilitates the emergence of best practices even in the absence of formal
+methods.  Similarly, our analysis sends a useful message to domain and
+application scientists that the challenges and barriers they face are not
+unique even if current solutions require a level of customization.
 

### 2015-05-29 old-sections/3.2oldnotes.tex
adding other, mostly old material

\subsection{Old Notes from 3.2}

\note{Old classification:
\begin{itemize}
	\item distributed by design (prev. naturally distributed): the data sources
	themselves are naturally distributed, e.g. scientific instruments or sensors
	collecting observational data.
    \item replicated: the data is replicated and therefore distributed to
	multiple locations
	\item partitioned: the data is partitioned and distributed across
	multiple locations
    \note{possible replacements for the three above - problem: all applications have them!} \note{What is a data source? How many data sources are in Atlas?}
	\item coupled \alnote{alternatively aggregation?}: multiple data sources (M)
	are combined at one or more locations (N), M$>$N
   \jhanote{do we propose natural, replicated and or partitioned as the 3
   types? for coupled can be used as a qualifier for all three of the
   above.}
   \item replicated: multiple data sources (M) are expanded to
          one or more locations (N): M$ <$ N \jhanote{can this be
          merged with above replicated?}
\end{itemize}
}

\note{View point: We should probably aim to classify applications with respect to the data processing piece? How do we describe application with multiple
distributed execution units and the dataflow in between?}

\note{Is the production of the data inside or outside of the application? When
is data production infrastructure and when is it application? DK: Floating
sensors are an instrument that happened to physically distributed. Are sensor as a whole designed to function at all}
\note{What kind of control do you have over the data source?}

\note{federated data? is partition a form of federated? federation consists of partitioning, replication and uniformity. Uniformity not covered federation. properties: autonomy, heterogeneity}

\note{replication could have a value: how many copies? average number of sites
that store a data element; distributedness: average distance to data; does it
make sense to merge partitioning and replicated?}

\note{new properties: coupling and integration (federation is a potential way to do integration).}

\note{federation: loosely coupled, partitioning == loosely coupled? Who is
carrying out the integration of sources? user, application, infrastructure}

\note{data storage vs. data production: Is natural the point where the data
storage is?}

\note{notion of geographic distribution}

\note{This needs to be incorporated in prior subsections: Types of Distributed Data Infrastructure}

\note{volume of data is almost always increasing. Is this not dynamic? We need
to refine the timescale property}

\note{distributed properties: subset (partial replication?) vs. partitioning}

\note{4 could be an entry in every item (except FUSION - since it is always
1 machine). Applies not to special high-end supercomputer. For multi-purpose,
distributed setup infrastructure is an element: data-world analog for a static
execution model. We need to talk about data's use of infrastructure... Would the
data be able to exploit infrastructural changes?
DK: costs for data storage to high to actually do that
}

\katznote{Fusion is not one machine.  4 doesn't apply to a number of applications - it only applies to the applications which have been designed to do this}

\note{view point: application consumer of data? We need to clarify the
viewpoint for each application.}

\note{need to define relationship between production and consumption of data}

\note{not all apps are input-proc-post-proc; there are also feedback systems...}

\note{Use of ``natural'' needs to be defined more carefully. Table needs to be re-visited. }

\note{How to handle feedback to sensors?}

### 2015-05-29 old-sections/3dpas_intro_public_do-not-edit.tex
adding other, mostly old material

%\documentclass[acmjacm,acmnow]{acmtrans2m}
%\documentclass[conference,final]{IEEEtran}
%\documentclass[10pt,letterpaper]{article}
\documentclass[preprint,12pt]{article}

\usepackage{graphicx}
\usepackage{url}
\usepackage{color}
\usepackage{pdfsync}

% \acmVolume{}
% \acmNumber{}
% \acmYear{}
% \acmMonth{}

%\usepackage{sectsty}
%\usepackage{setspace}

% \usepackage{ifpdf}
% \usepackage{wrapfig}
% %\usepackage[cm]{fullpage}
% \usepackage{textcomp}
% \usepackage{srcltx}
% \usepackage{fancyhdr}
% \usepackage{setspace}
\usepackage{lscape}
% \usepackage{longtable}
% \usepackage{paralist}

%\usepackage{hyperref}
%\usepackage{pdfsync}

% \usepackage{ams}

% \ifpdf    %   \usepackage[pdftex,
%               colorlinks=true,
%               linkcolor=red,
%               citecolor=red,
%               filecolor=red,
%               ]{hyperref}
% \else
%   \usepackage[hypertex]{hyperref}
% \fi

% Space saving
%\usepackage[small,compact]{titlesec}
%\usepackage[small,it]{caption}

% \renewcommand\floatpagefraction{.9}
% \renewcommand\topfraction{.9}
% \renewcommand\bottomfraction{.9}
%\renewcommand\textfrare{.1}

\setcounter{totalnumber}{50}
\setcounter{topnumber}{50}
\setcounter{bottomnumber}{50}

%\textwidth = 6in
%\textheight = 8.5 in
%\oddsidemargin = 0.0 in
%\evensidemargin = 0.0 in

% \topmargin = 0.0 in
% \headheight = 0.0 in
% \headsep = 0.0 in
%\parskip = 0.0in
%\parindent = 0.5cm
\parskip = 0.04in
% \parindent = 0.0cm
% \textfloatsep = 0.1in

% \usepackage[draft,pdftex,colorlinks=true, linkcolor=blue,citecolor=blue,
%        urlcolor=blue]{hyperref}

% \ifpdf
% \pdfinfo{ /Author (Shantenu Jha et al.)  /Title (Programming Abstractions for Large-Scale Distributed Applications) }
% \fi

\newcommand{\projectnamefull}{\textit{Programming Abstractions for
    Large-Scale Distributed Applications} }

\newcommand{\upup}{\vspace*{-0.5em}}
\newcommand{\up}{\vspace*{-0.25em}}
\newcommand{\I}[1]{\textit{#1}}
\newcommand{\B}[1]{\textbf{#1}}
\newcommand{\T}[1]{\texttt{#1}}
\newcommand{\BI}[1]{\B{\I{#1}}}

% \ifpdf
%   \DeclareGraphicsExtensions{.pdf, .png, .jpg}
% \else
%   \DeclareGraphicsExtensions{.ps, .eps}
% \fi

\long\def\comment#1{{\bf \textcolor{magenta}{\bf #1}}}
\long\def\ccomment#1{{\bf \textcolor{blue}{\bf #1}}}
\newcommand{\C}{\comment}
\newcommand{\CC}{\ccomment}

\newcommand{\yes}{$\bullet$}

\long\def\finalcomment#1{{\bf \textcolor{red}{\bf #1}}}
\newcommand{\CF}{\finalcomment}

\newif\ifdraft
\drafttrue
\ifdraft
 \newcommand{\katznote}[1]{ {\textcolor{magenta}    { ***Dan:      #1 }}}
 \newcommand{\rananote}[1]{ {\textcolor{blue}    { ***Omer:     #1 }}}
 \newcommand{\note}[1]{ {\textcolor{red}    { #1 }}}
\newcommand{\manishnote}[1]{{\textcolor{green}   { ***Manish:   #1 }}}
\newcommand{\jhanote}[1]{  {\textcolor{blue}     { ***Shantenu: #1 }}}
 \newcommand{\jonnote}[1]{  {\textcolor{red}      { ***Jon: #1 }}}
\else
 \newcommand{\katznote}[1]{}
 \newcommand{\rananote}[1]{}
 \newcommand{\note}[1]{}
 \newcommand{\manishnote}[1]{}
 \newcommand{\jhanote}[1]{}
 \newcommand{\jonnote}[1]{}
\fi

\title{3DPAS: Distributed Dynamic Data-intensive Programming
  Abstractions and Systems}

%\author{J.D. Blower$^{10}$, Neil Chue-Hong$^{}$, Simon Dobson$^{}$,
%  Shantenu Jha$^{1,2,3}$, Daniel S. Katz$^{4,1,5}$, Omer Rana$^{8}$}

\author{ Shantenu Jha, J.D. Blower, Neil Chue-Hong, \\[-0.1em]
Simon Dobson, Daniel S. Katz, Omer Rana}

%Manish Parashar$^{6,7}$,
%   Jon Weissman$^{9}$,

%   \small $^1$Center for Computation \& Technology,
%   Louisiana State University\\[-0.1em]
%   \small $^2$Department of Computer Science,
%   Louisiana State University\\[-0.1em]
%   \small $^3$e-Science Institute,   University of Edinburgh\\[-0.1em]
%   \small $^4$Computation Institute,   University of Chicago\\[-0.1em]
%  \small $^5$Department of Electrical and Computer Engineering,
%   Louisiana State University\\[-0.1em]
%   \small $^6$NSF Center for Autonomic Computing, Rutgers University\\[-0.1em]
%   \small $^7$Department of  Electrical and Computer Engineering, Rutgers University\\[-0.1em]
%   \small $^8$Department of Computer Science,
%   Cardiff University\\[-0.1em]
%   \small $^9$Department of Computer Science,
%   University of Minnesota\\[-0.3em]
%   \small $^10$Reading e-Science Centre, University of Reading, UK
%}
\begin{document}
\maketitle

\begin{abstract}
  Many problems at the forefront of science, engineering, medicine,
  and the social sciences, are increasingly complex and
  interdisciplinary due to the plethora of data sources and
  computational methods available today.  A common feature across many
  of these problem domains is the amount and diversity of data and
  computation that must be integrated to yield insights. Such data is
  increasingly large-scale and distributed;
  the data-sets, their location and scheduling decision
  are time-dependent. % Applications need access
  % to these disparate data-sets and to the methods that operate on
  % them.  Data-intensive applications need first-class support for
  % distributed heterogeneous platform execution.
  The 3DPAS Research theme (funded by the eSI, Edinburgh and US NSF),
  aims to investigate applications that have these characteristics, in
  other words it operates at the triple point of {\it dynamic} and
  {\it distributed} and {\it data-intensive}, (3D) attributes.  This
  evolving document represents work in progress of the 3DPAS Research
  Theme.
\end{abstract}

%\input{notes}

\section{Introduction: Context, Scope and Outline}

\jhanote{Overall introduction, leading into some specific examples.}

\section{Understanding Distributed Dynamic Data}

Why bother about distributed, dynamic and data-intensive applications?
\jhanote{(Shantenu, Omer, Neil, Simon, and Dan)}

\subsection{Importance of Distributed}

\subsection{Importance of Dynamic and Examples}

There is a general appreciation that the challenges of data-intensive
applications span beyond large-volumes and management/curation.  One
specific attribute that we believe is pervasive but not necessarily
obvious is “dynamic data”. The belief is based upon the fact that data
is often distributed, and as a consequence many attributes of
distributed systems permeate to data. Additionally the requirement of
many applications impose a requirement that data is dynamic.

\subsection{Types of Dynamic Data}
Some example of such dynamism:

\begin{itemize}
\item Change of data operated upon.
\item Variability of data rate -- arrival, consumption or ability to generate (due to dependencies)
\item Data placement and scheduling issues are time dependent
\item The nature of coordination between the different execution units
changes
\end{itemize}

\jhanote{Possibly merge?}

\begin{enumerate}

\item Streaming data -- data comes and goes
  \url{http://traffic.berkeley.edu/} and
  \url{http://lagrange.ce.berkeley.edu/fsn/}
\item Data set that is being operating upon is changing
\item Data has to be transformed
\item Placement decisions time-dependent
\end{enumerate}

\jhanote{We should highlight which of the above dynamic case are to be
  found in applications}

\katznote{Question: is there anything unique about scientific work on
  streaming data?  Or is all work on streaming data basically them
  same, whether it's in science or finance, for example?}

\jhanote{Good Q: I think other than scale -- of data volume, and
  degree of variability they are not too dissimilar}

\subsection{Characterising Dynamic Applications}

Amdahl’s number has been useful in characterising “static”
data-intensive applications and systems. We believe there is scope to
generalize Amdahl’s number to capture attributes of dynamic data
applications.

\subsubsection{Dynamic Data in a Web Environment}

The NERC\footnote{UK Natural Environment Research Council} Virtual
Observatory\footnote{http://www.nerc.ac.uk/research/programmes/virtualobservatory/}
is a new initiative to create a pilot cloud-based environment in which
different practitioners will discover, visualize and process data
relating to the UK's system of rivers and soils.  It will provide
access to many different kinds of distributed hydrological data,
including sensor data and numerical models, many of which produce
rapidly-changing, dynamic information.  There are many scientific
challenges to be faced, notably including the coupling of the models
and observations; this includes the possibility that sensors can be
dynamically tasked in response to model forecasts, or that models may
adapt dynamically to new information from sensors.  Principles of the
Web and ``linked data'' may help to solve some of the challenges in
informatics, but many technical challenges will remain regarding the
handling of dynamic data, including keeping careful records of its
provenance in order to record the steps taken in making a decision.

\emph{Just some placeholder notes from Jon.  Could belong elsewhere.}

Web Services are commonly used to distribute data through HTTP.  A key
characteristic of dynamic data, of course, is that it changes with
time.  It is of great importance for the user to be confident that she
is accessing the correct version of the data; this will often be the
most up-to-date data, or may be a specific historical version.  In a
Web environment, servers and proxy servers (which are sometimes beyond
the control of the data provider or the data user) may retain caches
of data in a strategy to reduce server load and network traffic.  For
this reason, the HTTP protocol provides mechanisms for defining expiry
times for data resources, specifying when a resource ought to be
cleared from the cache.

In the environmental and geospatial communities the Open Geospatial
Consortium specifications are being widely adopted alongside existing
protocols such as OPeNDAP.  Unfortunately these protocols, although
built atop HTTP, do not make it easy for this versioning to be
implemented correctly.  The OPeNDAP protocol does not have the concept
of a dataset version, meaning that clients have no reliable or
efficient means to detect that a dataset has changed.  The OGC
protocols in general [TODO: look into this], \emph{do} provide
versioning at the level of the service endpoint, but provide no
information on when a resource should be considered expired.  Clients
are therefore forced to poll the server to check for updates.

Therefore we see that in a ``pull'' environment such as the Web, it is
necessary for servers to advertise not only the version of a resource,
but also the expiry time in order to handle dynamic data correctly.
The frequency of updates is also useful information for service
consumers.  Expiry times can be specified using the mechanisms of HTTP
(although this capability is often not used in practice), but resource
versioning and update frequency must be handled at a higher level.

[All this must be fairly well-known and almost certainly covered in
the literature, with the possible exception of the material on OPeNDAP
and OGC.  Any references we could cite?]

\subsection{Other points to cover}

\begin{itemize}
\item Lay out several scenarios first, and then formulate/define Dynamic Data
\item Dynamic: transformed by the workflow scenario (state is captured by immutable objects)
\item How does this differ from Scientific Databases
\item How does it relate to Scientific Data Management and Scientific data flow
\item Carve out the specifics of this theme -- and how it relates to such existing approaches (a diagram would be nice)
\item Underlying data model (operations) are highly domain specific -- make clear what can be domain specific and domain independent
\item Relationship to ``legacy''
\item Dynamic data can arise at different levels (application vs. middleware (infrastructure)). Identify which is the primary consideration here, and what is different.
\item Reference to EU 2030 Objectives (part of the Project Europe 2030
  Report) -- underpins ``Data Intensive Research'' from an EU
  perspective (strategy) + relate this to NSF ``CIF21'' vision
\end{itemize}

How does Dynamic Data relate to Distributed \& ``Big'' Data? Relate to
DPA and Malcolm et al.'s theme (Data Lifecycle, Complexity, etc)

\section{Application Scenarios} {\bf Dan, Shantenu (co-leads), All}

Application Scenario Characteristics for Dynamic Data (Various)

Identify: (1) what is the case scenario within the application; (2) what data is dynamic; (3)  how the data is being used in the context of each of these scenarios.

Currently trying to do this by asking the following questions:
\begin{enumerate}
\item What is the purpose of the application?
\item How is the application used to do this?
\item What infrastructure is used? (including compute, data, network, instruments, etc.)
\item What dynamic data is used in the application?
\begin{enumerate}
\item What are the types of data,
\item What is the size of the data set(s)?
\end{enumerate}
\item How does the application get the data?
\item What are the time (or quality) constraints on the application?
\end{enumerate}

\katznote{new suggested questions from Don Middleton: How much diverse data integration is involved?  how diverse is the data?}

\subsection{Silvia's Biosciences app: NGS, medical imaging (Silvia)\label{bioSilvia}}

\begin{itemize}
\item Infrastructure monitoring for the execution of workflows to
  manage failure
\item the application is sequence alignment (split data (this is where there is dynamic decision making) to execute the alignment in parallel)
\item DTI Imaging: two cases are interesting: DTI atlas (split, run, merge, split, merge - reduction from 10000 tasks to 1 output); PCA for classification of patients/control
\end{itemize}

\subsubsection{Silvia's contribution}

NGS sequence alignment on the AMC e-bioscience infrastructure

1. What is the purpose of the application?

The application performs analysis of DNA sequencing data for various goals in life science research, e.g., Mutation screening, Virus discovery, Genome-wide Associations (GWA), Linkage analysis, Comparison of bacterial genomes, microRNA expression and Exome sequencing. The data analysis is composed of various methods that are implemented by a bioinformatician, whereas the data is typically owned by the life scientist.

2. How is the application used to do this?

Note: Here we focus on data analysis from the perspective of the bioinformatician, excluding steps related to data acquisition (transformation of raw images into sequences of amino acids) and posterior biostatistics and epidemiological analysis.

The data analysis is implemented as a pipeline of generic and customized components, mostly third-party public software tools. The pipeline may differ for each specific study, depending on the data acquisition and the research question. Typically it includes data preparation (file conversion, reformatting and filtering); alignment (comparison) with some reference database as the human genome; post-processing (output conversion, reformatting, statistical analysis) and visualization. Sequence alignment is the most computing-intensive step, which became prohibitive for sequential execution in next generation sequence experiments. All data is stored in files that are often completely read in memory for fast processing. Compressed formats are usually adopted. Although the files are large, the data can be easily decomposed because the processing is done on each sequence (`string') at a time.

3. What infrastructure is used? (including compute, data, network, instruments, etc.)

\begin{figure*}[htbp]
  \begin{center}
    \includegraphics[width=\textwidth]{figures/silvia.pdf}
    \caption{Figure from Silvia}
    \label{Fig:silvia}
  \end{center}
\end{figure*}

see Figure~\ref{Fig:silvia}.

Data:

Data acquisition: Roche, Solid and Illumina next generation sequencers

Local storage: dedicated server at the AMC network (ftp)

Grid storage: Storage Elements of the Dutch e-Science Grid (www.biggrid.nl)

Data transport:

From sequencer to server: AMC network (XXX) or off-line media

Between data server and grid storage: public network (XXX);  lightpath (near future)

Between grid storage and worker nodes: public (XXX) or site (XXX) network

Computing:

Local Computing: clusters and servers at the AMC and workstations (laptops, desktops) anywhere
Grid Computing: Dutch Grid (www.biggrid.nl), part of EGI

Originally all steps were performed in some local server or on the user's workstation. The visualization is still performed in the user's (windows) workstation, but all the other steps are currently performed on the Dutch grid infrastructure. The pipeline is described as a grid workflow using the GWENDIA language (\url{http://dx.doi.org/10.1145/1645164.1645171}) and enacted on the Dutch Grid using the MOTEUR2 workflow engine (\url{http://modalis.polytech.unice.fr/softwares/moteur/start}).

PS: although only the alignment step is computation-intensive, it is more practical to perform all the others on the grid to minimize data transport to/from the user's workstation, which is normally on low speed networks. That is, the computation is statically and manually taken to the data in this case. This can lead to sub-optimal resource usage.

4. What dynamic data is used in the application?

Note: data is not really `dynamic' in this case -- the sequences are acquired once and analyzed many times. However the execution of this application on a distributed infrastracuture has dynamic aspects when optimized resource usage is considered:

\begin{itemize}
\item Split/merge the data according to available resources
\item Compute resource selection based on data location, processing profile and/or network capacity
\item Resource selection based on security constraints (access control, trust)
\item Collection and processing of monitoring information to steer grid enactment.
\end{itemize}

(a) What are the types of data,

DNA sequences (strings) stored in open formats (FASTA, SFF, SAM, BAM)
Bundles of results from alignment tools (custom format, binary?)

(b) What is the size of the data set(s)?

one example for BWA alignment tool (short sequences) follows:

input data: Sequencing Data in the csFasta format, normally 25-35~GB;
Quality files in the .qual format, normally 50-80~GB;
Reference DB in the Fasta BS format: 3.2~GB (human genome) 140~MB (one chromosome)

Intermediate files:
Reference BWA index, normally 4.5GB (human genome) 240~MB (one chromosome);
Sequencing Data in the FastQ format (fastq.gz), normally 20-30~GB

Output:
Results in .sai format (direct output of BWA), normally 2-3~GB;
Results in .sam format, normally 55-75~GB;
Results in .bam format, normally 20-30~GB

5. How does the application get the data?

From a local file. The legacy code is wrapped to stage the inputs and outputs in/out the worker node.

All data is stored in files that are often completely read in memory for fast processing. Compressed formats are usually adopted. Although the files are large, the data can be easily decomposed because the processing is done on each sequence (`string') at a time

6. What are the time (or quality) constraints on the application?

In research there are no strict constraints on the time to complete the analysis.
In patient care, time could determine life, death or money, but we have no concrete case yet.

Quality in this case refers to the accuracy of the results obtained. This is essential for research and patient care.

\subsection{WLCG (focus on ATLAS) (Steve Fisher) \label{WLCGSteve}}
\begin{itemize}
\item Most of the dynamism is in infrastructure, not in the application (e.g. use of PilotJob)
\item Application dynamism: location of file (e.g. if file turns out to be not local, look up in registry-- but not do this again)
\item No specific QoS issue concerned with data transfer times
\end{itemize}

\subsubsection{New answers from Steve}

3DPAS - Particle Physics - the example of Atlas physics channel selection.

   1.     What is the purpose of the application?

A physics group is interested in a specific physics channel (underlying physical process) as recorded by the Atlas experiment at the LHC (Large Hadron Collider) at CERN in Geneva. The set of available data grows steadily over time as more data are collected and reconstructed.

   2.    How is the application used to do this?

 Both physics groups and individuals run jobs as they see fit, either reading the reconstructed data or reading intermediate datasets created by themselves or their colleagues. If, in the future, resources should prove inadequate then adjustments to working practices to coordinate processing may be made to increase physics output.

   3.      What infrastructure is used? (including compute, data, network, instruments, etc.)

The WLCG (Worldwide LHC (Large Hadron Collider) Computing Grid) is used with a mixture of gLite software, experiment specific software and physics group specific software. WLCG has 250,000 cores distributed over 140 sites and with 100PB of disk. CERN is connected to the eleven `Tier 1' national sites by a dedicated optical network. This may be extended in the future.

   4.      What dynamic data is used in the application?

The data being processed can be considered as dynamic as it grows steadily as new raw data are collected and the basic reconstruction program is run to produce information about specific particles involved in a collision. This reconstruction is rerun periodically (two or three times per year is expected) as calibrations of the detector are improved. Reconstruction algorithms are also improved as are the physics groups' selection codes. One output of the reconstruction is `tag' data - summary information written to a database to facilitate event selection. This is not yet used very much. Calibration data are held on an RDBMS with Frontier/Squid front ends.

         a.             What are the types of data,

          The bulk of the data is individual event data at different stages of refinement. This is held in a in-house representation of C++ serialized objects. Though this format is not standard it is used by all LHC experiments.

         b.            What is the size of the data set(s)?

          Roughly 20~TB of data are collected every day. Multiple generations of processed data are kept. Many data files are replicated on multiple sites; this is done both by copying it to where it is likely to be useful and partly dynamically as the need arises.

   5.      How does the application get the data?

A copy of the file is made to the SE (Storage Element) at the site where the job will run. From there it is either copied to a local disk or read from the SE.

   6.      What are the time (or quality) constraints on the application?

If some data are missed during processing the only effect is a reduction of statistical precision in the final analysis. Accurate book keeping is important to know which data have been processed. If the selection code has been improved then naturally the physics groups would like to see the results as soon as possible.

%\subsubsection{Old answers from Steve}
%
%   1.       What is the purpose of the application?
%
%A physics group is interested in a specific physics channel (underlying physical process) as recorded by the Atlas experiment at the LHC (Large Hadron Collider) at CERN in Geneva. The set of available data grows steadily over time as more data are collected and reconstructed. It would not be efficient to allow individual physics groups to process this large volume of data in an uncoordinated manner as it requires reading all available data.
%
%   2.       How is the application used to do this?
%
%Physics groups register their analysis code to be run on the next scheduled analysis ``train''. The train leaves every two weeks and reads through all the data, runs the physics selection code of each group and writes out the selected events and output from the computation to produce relatively small data sets for each physics group. The groups then use these selections to perform more ad-hoc analysis.
%
%   3.      What infrastructure is used? (including compute, data, network, instruments, etc.)
%
%The WLCG (Worldwide LHC (Large Hadron Collider) Computing Grid) is used with a mixture of gLite software, experiment specific software and physics group specific software. WLCG has 250,000 cores distributed over 140 sites and with 100~PB of disk. There is no wide area dedicated network infrastructure.
%
%   4.      What dynamic data is used in the application?
%
%The data being processed can be considered as dynamic as it grows steadily as new raw data are collected and the basic reconstruction program is run to produce information about specific particles involved in a collision. This reconstruction is rerun periodically (two or three times per year is expected) as calibrations of the detector are improved. Reconstruction algorithms are also improved as are the physics groups' selection code. Calibration data are held on an RDBMS with Frontier/Squid front ends.
%
%         a           What are the types of data,
%
%          The bulk of the data is individual event data at different stages of refinement. The is held in a non-standard representation of C++ serialized objects.
%
%         b.             What is the size of the data set(s)?
%
%          Currently the reconstructed data (excluding copies) is about m TB and is growing at about  n TB per month. Many data files are replicated on multiple sites. This is done both by copying it to where it is likely to be useful and partly dynamically as the need arises.
%
%   5.       How does the application get the data?
%
%This is mostly by copying the file to where the job will run - though some remote file access is also used.
%
%   6.  What are the time (or quality) constraints on the application?
%
%If some data are missed during processing the only effect is a reduction of statistical precision in the final analysis. If the selection code has been improved then naturally the physics groups would like to see the results as soon as possible.
%
%
\subsection{Astrophysics -- Virtual Observatory (Bob) \label{astroBob}}
\begin{itemize}
\item Analysis of data to generate event streams $\rightarrow$ lead to the
  positioning of telescopes
\end{itemize}

\subsubsection{Answers from Bob}

I'll assume that the ``application'' in question is the sum of the two
activities, despite the fact that they are different in many respects,
and I'll try and distinguish between them where appropriate.

Let me know if you want clarification or further information - in some cases I wasn't quite sure what you were after.

1. What is the purpose of the application?

An increasingly important feature of astronomical research is the existence of
systematic sky surveys covering the whole sky (or large fractions thereof) in
many different wavebands. The principal outcome from these surveys are
astronomical source catalogues, which contain values for attributes
characterizing all the celestial sources detected in the sky survey dataset.
These are now typically implemented in relational databases, with SQL interfaces
implemented through web pages, and the principal goal of the Virtual Observatory
(VO) is to standardize access to these, and other astronomical data resources,
so that astronomers can readily perform multi-wavelength analyses, combining all
the extant data on particular sources.

      2. How is the application used to do this?

The VO includes a registry that contains metadata describing all the data
resources published using VO data access standards, enabling the discovery of
datasets relevant to a particular analysis. Web service implementations of data
access protocols enable users to access data in these repositories
programmatically, and distributed scratch space is provided for the storage of
intermediate result sets, so the user does not need to route large data flows
through his/her own workstation. In the future, more and more data analysis
software will be made available through web services compliant with VO
standards, enabling more of the data integration and analysis process to be
combined in workflow.

      3. What infrastructure is used? (including compute, data, network,
      instruments, etc.)

The data resources published to the VO exist in a number of data centers
distributed internationally. Major datasets---e.g. those from large sky
surveys---are typically implemented in relational databases running on high-spec servers
or clusters thereof. The compute requirements of the VO are currently modest,
but are likely to increase over time. Astronomers typically use normal academic
networks, but more specialized networking is required in some cases: e.g., sky
survey data is moved from Cambridge to Edinburgh using the UKLight project's
dedicated optical network, and radio interferometers require high-bandwidth
fibre connections from each antenna to the correlator.

      4. What dynamic data is used in the application?
(a) What are the types of data, (b) What is the size of the data set(s)?

The VO currently includes little dynamic data. The one exception is VOEvent
messages, which are notifications of transient events which can be used to
trigger follow-up observations on other telescopes. These are small XML messages
containing basic information about the position of a transient source and a
small amount of other metadata which the operators of a given telescope can use
to define triggers for follow-up observations of interest to them or their
community.

 5. How does the application get the data?

Assuming this refers solely to the dynamic data, then a number of telescopes
undertaking real-time detection of transient events emit VOEvent messages
through feeds to which other parties can subscribe.

6. What are the time (or quality) constraints on the application?

The time constraints differ for different types of transient source, but in some
cases notification must be made within a few minutes if it is to be useful; the
overheads associated with setting up a follow-up observation mean that much
shorter time scales are practically irrelevant, although they are still
scientifically interesting. The quality constraints centre on providing a
telescope with good enough information about a potential transient source that
it can make a reliable decision about whether to interrupt its existing
observing program to follow-up the event. In most cases, the candidate
transient will be some sort of noise event, so fairly sophisticated filtering is
required to bring the false positive rate down to something acceptable, given
the cost of observing time on cutting edge telescopes.

\subsection{Sensor Network App (Simon) \label{sensorSimon}}

\begin{itemize}
\item Environmental (Marine) -- sensing within a hostile environment,
  data quality + availability of communication infrastructure (Simon)
\item Patient monitoring -- data quality, type of data to transmit
  (Omer)
\end{itemize}

\subsubsection{Answers from Simon}

1. What is the purpose of the application?

Marine sensing covers a range of different applications, including
environmental monitoring, remote exploration, marine life surveys and
habitat assessment. In each case, the main challenges are collecting
data reliably in what is an extremely challenging technical
environment---far more so than in air---and (in the case of biological
surveillance), without interfering with the natural behaviors the
animals exhibit.

2. How is the application used to do this?

At present most applications are quite small-scale. A canonical example
is the monitoring of seal and other sea mammal populations by the
Scottish Oceans Institute (SOI), which works by tagging animals with sensor
packages that can analyze dive behavior, speed, movement and the like,
and report back to base when the animal comes within range of a cellular
network.

3. What infrastructure is used? (including compute, data, network, instruments, etc.)

The data sets collected are not enormous by scientific-data standards,
and are analyzed offline using traditional statistical techniques.
Recent work by SOI has included integration with Google Earth to allow
animal tracks to be visualized on a large scale.

Data are collected remotely, then brought to a central site for analysis.

4. What dynamic data is used in the application?

In the sea mammal case the data isn't all that dynamic, being mainly a
time series collected with only basic adaptation to the resolution etc.
Larger-scale applications in environmental sensing in particular would
make use of more structured and extensive adaptation (e.g.,
changing the sampling frequency and other management characteristics in
response to the data being observed).

(a) What are the types of data, (b) What is the size of the data set(s)?

For sea mammals: positions, motion vectors. For environmental missions:
concentrations and gradients. I'd have to enquire about dataset sizes,
but not excessive.

5. How does the application get the data?

For the sea mammals the animal crawls up a beach, the sensor packages
sees the cellular network and transmits. For environmental missions
we're still wondering about that ourselves :-)

6. What are the time (or quality) constraints on the application?

Not sure I see what you mean here. Generally it's just a matter of
capturing at whatever resolution device the sensor pack can support.

I could probably get more complete -- perhaps *way* too complete :-) --
answers from Mike Fedak in Scottish Oceans if that'd help.

%\subsection{Bioinformatics Applications (Tom Slezak) - do not publish \label{bioTom}}
%
%1. What is the science/engineering problem? What is the purpose of the application?
%
%Comparing metagenomic (DNA sequence) data sets against each other. This could be to distinguish human disease biomarkers among different individuals, determine different soil (or water or air or ...) microbial composition from different samples, etc.
%
%2. How will the application solve the problem?
%
%In general this is a large (for biology -- don't think in terms of exascale!) data problem with large (all-vs-all comparison) computing. There are strategies to make this more scalable that will reduce the computing volume.
%
%(a) Which of the following describes the application
%
%         i) large (exascale) data and large (exascale) computing (e.g., data mining, analytics, bioinformatics)
%
%         ii) large (exascale) data and modest computing (e.g., search)
%
%         iii) modest data and large (exascale) computing (e.g., simulation/modeling)
%
%(b) Can you quantify the amount of computing vs the amount of data?
%
%Potentially peta-bytes or exa-bytes of data. Not sure how to quantify the computing. (Huge memories will be potentially more valuable than huge numbers of cores -- although this is heresy to the physics crowd.)
%
%3. What infrastructure will be used? (including compute, data, network, instruments, etc.)
%
%(a) What will be distributed?  data, compute, both?
%
%Individual data sets will be terabyte-scale. There could be many thousands of such sets to compare. Having them all local may not be feasible (the transfer time from where they are generated by others may be prohibitive. Computing might then have to be distributed. Biology is wrestling with this now.
%
%4. What data be used in the application?
%
%(a) What are the types of data
%
%Genomic sequence, billions of short (35-100?) bases of DNA; unassembled
%
%(b) What is the size of the data set(s)?
%
%Terabyte scale
%
%(c) Will any data be dynamic?
%
%Not for this problem
%
%5. How will the application get the data?
%
%That is a major part of the problem! Data will (in general) be generated from other places around the world. Getting to the data will be an issue.
%
%6. Are there time (or quality) constraints on the application?  (If so, what are they?)
%
%You'd like answers in under a day, generally.
%
%7. Please feel free to also talk about the current state of the application, if it exists today, and any specific gaps that you know will need to be overcome.
%
%Biology is wrestling now with  about 3 orders of magnitude increase in DNA sequence data generation capability in the past 2 years. All algorithms, architectures, storage, and networks have broken or will break soon. Some biology problems (not necessarily the one I describe here) are not good fits for core-rich and memory-poor modern architectures. For some problems (comparing k-mers across metagenomic data sets, for instance) you need terabyte-scale memory, not thousands of cores.
%

\subsection{Climate (Dan) \label{climateDan}}

\subsubsection{CDAT} (from Dan)

The Climate Data Analysis Tools (CDAT)~\cite{CDAT} are a scriptable
(Python) set of packages and tools that are currently heavily used for
analysis of climate data.  CDAT has been developed by LLNL's Program
for Climate Model Diagnosis and Intercomparison (PCMDI), and is often
used with data from the Earth System Grid (ESG)~\cite{ESG}.  The most
common current usage of these tools with the data is that a user
downloads a set of data from ESG to a local resource, then uses the
CDAT to analyze the data.  However, the PCMDI group is now planning to
support a new usage mode, where data transfer can be scripted.
Initial development of this capability is planned for February 2011,
using the Globus online data transfer service~\cite{globusOnline} as a
scriptable component.

\subsubsection{ESGF} (from Don and Dan)

Aimed at supporting international CMIP/IPCC intercomparison activity.
Three stages:

{\bf Data Acquisition (Modeling)}

\begin{itemize}
\item Climate centers run prescribed set of common experiments; produce 2-10 PB of data
\begin{itemize}
\item data can also come from sensors - here, the data would still be post-processed before it would be published.
\end{itemize}
\item Centers can publish their own output data or send it to another center to publish data generated over \~{}2 year period (then post-processed and published over a few more months)
\item Properties: data-intensive, distributed centers, lots of computing
\end{itemize}

{\bf Data storage (Infrastructure/system)}

\begin{itemize}
\item ESGF develops and deploys a federated network of gateways and associated data nodes
\item As models run at each center, output data is post-processed into packages with common formats and prescribed metadata conventions
\item Most centers will deploy data node software stack, and using this, they manage the data from their experiments
  \begin{itemize}
  \item the data node software stack scans the data, makes sure it has the right metadata fields, does QA/QC, build a set of catalogs of the prepared data
  \begin{itemize}
                \item minimally, catalog provides HTTP links to data elements
                \item also can provide GridFTP endpoints
                \item also can provide product services that abstract the dataset in other ways - get whole file, subset, browse, etc.
  \end{itemize}
  \end{itemize}
    \item When the center is happy with the data/catalog in the data node, they publish it to a host/affiliated gateway
\begin{itemize}
         \item  this submits the catalog to the gateway
          \item  them the gateway shares this catalog with other gateways so that all gateways have a consistent view of all the published data
\end{itemize}
    \item  Core archive (1-2 PB, most popular data)
\begin{itemize}
          \item  will be replicated at several sites
         \item  replication activity manually requested/initiated by a gateway owner
          \item  replication: copies catalog to replica site, do data transfer to new data node, then publish  catalog, push catalog to (this) gateway associated with new data node
\end{itemize}
    \item  Currently building notification service
\begin{itemize}
          \item  track which users have downloaded data - if a problem is found in a data set or if new data is added to a data set, notify users who have downloaded/replicated it
\end{itemize}
    \item  Properties: data-intensive, distributed gateways and data nodes, data appears in the system over time (dynamic)
\end{itemize}

{\bf Data processing (Application)}

\begin{itemize}
    \item  Users (\~{}20000 as of Dec. 2010) can:
\begin{itemize}
         \item  browse/search a catalog at any gateway and locate data, which might be hosted by a data node affiliated with another gateway. (can output a wget script that can later fetch the data)
         \item  authenticate, gain access to a group
        \item  download data via http or GridFTP or access product services (uses data retrieval syntax (DRS) - can be scripted)
  \end{itemize}
\end{itemize}

\begin{itemize}
    \item  abstractions that can be used by a programmer: data is stored in NetCDF format, uses CF metadata conventions, can be accessed using DRS
    \item  Some users will analyze data from a single model
    \item  Many applications are multi-model analyses - many users want to look at the same parts of the output of some/all of the models, to understand if and how the models differ
    \item  Some centers will gather some/all of the core archive (plus more, perhaps) on local systems for local users to perform ``power analysis''
\end{itemize}

\subsection{Astro (LSST) (Adam Barker) \label{astroAdam}}

\begin{enumerate}
\item What is the science/engineering problem? What is the purpose of the application?

Find and study variable objects and moving objects.

\item How will the application solve the problem?

Processing of data (data analysis and detection of new sources) from each image should be done while the next image is being taken by the telescope.  This means that if anything unusual is detected, normal observation can be interrupted. Other observing resources can also be notified instantly, so they can observe the same event. As data are collected, they will be added to all the data previously detected from the same location of sky to create a very deep master image. LSST will also build up a database of all known moving objects.

Every time a new image of the sky is obtained, the master image will be subtracted from it. The result is a subtracted image that contains the difference between the current sky image and its average state.

This subtracted image is then processed by a cluster of computers with the following steps:
\begin{enumerate}

\item Query object catalog (knows orbits of all known objects) to find objects which are expected to appear in subtracted image, given the area of sky, time of day.

\item Cross-match expected objects with sources in subtracted image. The result is the unmatched-source catalog, which contains sources that canÕt be matched with a previously known object

\item Use unmatched-source catalogue to compute pairs of detections separated by short time intervals, called tracklets. A tracklet is an observable, short section of orbit. In order to create pairs of detections, all objects in the unmatched-source catalog are queried against the orbit catalog, in an attempt to obtain data about these objects from earlier observations. If earlier observations and current observations can be linked to form tracklets, they are stored in the tracklet catalog.

\item Use tracklet catalog to attempt to link these tracklets over a larger time window. This is achieved by again querying the orbit catalog, this time looking for observations of the same objects even further back in time. If matches can be found and the tracklets can be extrapolated out into longer sections of orbit, then they are added as new orbits to the orbit catalog.

\item Update orbit catalog with redetections of known objects. Each rediscovery provides information that improves the known orbits.

\item Attempt to classify all entries in the orbit catalog (i.e., all the objects which now have known orbits). If any Near Earth Objects are detected to be passing close to the earth, alerts are generated to astronomers so that follow up observations can be scheduled.

\item Compare with other observatories: If an unknown object is found that cannot be classified locally
\begin{enumerate}
\item issue call for participation to set of possible observatories
\item each observatory accepts or rejects the call
\item local observatory chooses from accepting observatories one to obtain data from
\item that observatory takes data and sends it to local observatory
\item local observatory decides what to do - classify the object, contact a human, call for more observations (automatically to other observatories, that decide whether or not to accept, and when)
\end{enumerate}

\end{enumerate}

\begin{enumerate}
\item Which of the following describes the application

large (exascale) data and large (exascale) computing (e.g., data mining, analytics, bioinformatics)

\item Can you quantify the amount of computing vs the amount of data?

LSST will generate 36~GB of data every 30 seconds, and over a 10-hour winter night, will collect up to 30~TB.

\end{enumerate}

\item What infrastructure will be used? (including compute, data, network, instruments, etc.)

A linked network of observatories and data and computing resources.

(a) What will be distributed?  data, compute, both?

both

\item What data be used in the application?

Fits images, and large databases for analysis of the discoveries.

 (a) What are the types of data

 (b) What is the size of the data set(s)?

 (c) Will any data be dynamic?

\item How will the application get the data?

Some will be analyzed at the mountain top, discoveries will be sent to other centers.

\item Are there time (or quality) constraints on the application?  (If so, what are they?)

Time component is critical. As real-time as possible.

\item Please feel free to also talk about the current state of the application, if it exists today, and any specific gaps that you know will need to be overcome.

\end{enumerate}

also, the Palomar Transient Factory is the best example to date:

See:

\url{http://www.astro.caltech.edu/ptf/}

\url{http://xxx.lanl.gov/abs/0906.5350}

\url{http://xxx.lanl.gov/abs/0906.5355}

\url{http://arxiv.org/abs/1011.2199}

\subsection{Interactive exploration of environmental data (Jon) \label{envJon}}
Synopsis of chapter we're writing for Malcolm's book.  Looking at needs of interactive applications as opposed to batch ones.

\subsubsection{quick answers from Jon}

1.     What is the purpose of the application?

We are actually concerned with a class of applications (highly interactive, graphical ones) rather than a specific one.  But generally, the applications involve the intercomparison of different environmental datasets for purposes such as model validation, ground trothing of observations and data assimilation.

2.     How is the application used to do this?

The applications often take the form of graphical tools that share a lot in common with Geographic Information Systems (GIS). However these applications are much better suited to scientific data, which are large and multidimensional.  The general ÒvisionÓ is of a map-based interface, onto which different datasets can be overlain.  A processing step may be initiated by rubber-banding an area of the map and selecting from a list of algorithms that process data within the selected area, perhaps calculating statistics of the data.

3.     What infrastructure is used? (including compute, data, network, in- struments, etc.)

Usually the applications are driven by data served through web services of various kinds.  These web services may be ÒfedÓ by data coming from instruments or from computer simulations.  A key challenge is how to integrate compute services with these data services in an efficient, easy-to-use and transparent way, allowing fast data processing.  The processing algorithms are generally quite simple and so it is frustrating to succumb to the high overhead of some scheduling systems.  The infrastructure probably has more in common with the Web (or perhaps cloud) than the Grid.

     4. What dynamic data is used in the application?
(a) What are the types of data, (b) What is the size of the data set(s)?

Dynamic data are only sometimes used, but may arise from Òreal-timeÓ feeds from instruments or from looking at ÒliveÓ results from a model.  Data from instruments are ÒsmallÓ 1MB-100MB but model output may be very big (GB-TB).

4.     How does the application get the data?

Generally through web services.  These services are designed to allow server-side subsetting of large datasets, attempting to minimize the amount of unwanted data travelling across networks.

5.     What are the time (or quality) constraints on the application?

We try to aim for near-real time interaction with the data, i.e the user should not be waiting more than a few seconds for some kind of response to a request.  We have largely solved this problem for simple visualization, but for linking data access and processing we are not there yet.

\subsection{Power Grids (Shantenu) \label{powerShantenu}}

Consider the energy informatics domain and smart power grids in
particular. For data continuously arriving from 2 million smart meters
in large city, households will soon need to be continuously analyzed
in order to detect impending peak power usage in the smart power grid
and notify the utility to respond by either spinning up additional
power sources or by triggering load curtailment operations to reduce
the demand. This closed loop cyber-physical application, modeled as a
workflow, needs to combine streaming data arriving from sensors with
historic data available in file archives along with structured
collections of weather forecast data that helps the large scale
computational model make an energy use prediction in near real time.

A workflow framework that supports this data model diversity,
including streaming data, structured collections and files, and the
ability to execute reliably and scalably on resources is required.

In general, smart grids - respond to varying data loads, spinning up
additional power sources or triggering load curtailment operations.
E.g., of a closed-loop cyber-physical system modeled as a workflow
needs to combine streaming data arriving from sensors with historic
data available in archives, along with structured collections of
weather forecast data. Need a workflow model that supports such
requirements. Taken from ref (Towards Reliable Performant Workflows
for Streaming Applications on Cloud Platforms).

{\it Requirements: } (i) Applications will have to span desktops,
workstations and clouds, (ii) streaming data models and stream
programming abstractions that are more than TCP sockets, (iii)
reliability of VMs hosting the workflow and avoid costly duplicate
movement of the same logical stream (see paper in WORKS by Zinn et
al).

Scientific workflows are commonplace in eScience applications. Yet,
the lack of integrated support for data models, including streaming
data, structured collections and files, is limiting the ability of
workflows to support emerging applications in energy informatics that
are stream oriented. This is compounded by the absence of Cloud data
services that support reliable and performant streams.  Need a
scientific workflow framework that supports streams as first class
data, and is optimized for performant and reliable execution across
desktop and Cloud platforms.

% \subsection{Astro (CMB) (Julian Borrill) - do not publish \label{astroJulian}}

% 1. What is the science/engineering problem? What is the purpose of the application?

% Cosmic Microwave Background (CMB) data simulation and analysis. The CMB is an image of the Universe as it was only 400,000 years after the Big Bang. Tiny fluctuations in the CMB temperature and polarization encode the fundamental parameters of cosmology and, using the Big Bang as the ultimate particle accelerator, ultra-high energy physics. Extracting this information from the data gathered by current and anticipated CMB observations is an extremely computationally intensive endeavour for which we have developed our massively parallel tools.

% 2. How will the application solve the problem?

% The central computing challenge for any CMB dataset is the simulation and analysis of O($10^4$) synthetic observations, used to correct for biases and quantify uncertainties in the analysis of the real data. We use preconditioned conjugate gradient techniques to solve for the maximum likelihood sky map given the input data (obtained by scanning the sky with hundreds to thousands of detectors for weeks to years) and its piecewise stationary noise statistics. To avoid the IO bottleneck inherent in the traditional simulate/write/read/map paradigm, all simulations are performed on the fly only when requested by the map-making code.

% (a) Which of the following describes the application

%      i) large (exascale) data and large (exascale) computing (e.g., data mining, analytics, bioinformatics)

%      ii) large (exascale) data and modest computing (e.g., search)

%      iii) modest data and large (exascale) computing (e.g., simulation/modeling)

% (iii) O(1 - 10) TB input data and a similar volume output.

% (b) Can you quantify the amount of computing vs the amount of data?

% The computing scales with the number of observation samples, while the communication and IO scale with the number of pixels in the map. For the current Planck satellite experiment these are O($10^12$) and O($10^8$) respectively; for the proposed CMBpol satellite they will be O($10^15$) and O($10^8$) - that is, we expect a 1000x increase in calculation cost over the next 15 years (including intermediate suborbital pathfinder experiments) but fixed communication/IO costs.

% 3. What infrastructure will be used? (including compute, data, network, instruments, etc.)

% We are targeting the largest supercomputers available - Hopper, Tianhe, Blue Waters, etc. It's all about the cycles, although as the concurrency we work at increases previously solved IO and communication scaling bottlenecks typically re-emerge.

% (a) What will be distributed?  data, compute, both?

% Both.

% 4. What data will be used in the application?

% Real and simulated CMB data.

% (a) What are the types of data

% (b) What is the size of the data set(s)?

% (c) Will any data be dynamic?

% See above.

% 5. How will the application get the data?

% Real data are read from disk; simulated data are generated on the fly.

% 6. Are there time (or quality) constraints on the application?  (If so, what are they?)

% Typically we want to be able to simulate and map O($10^2$) realizations of an experiment's data in O(1) wallclock hour to provide 10\% errors during the early analysis stages, and O($10^4$) realizations in O(100) wallclock hours for the final definitive analysis.

% 7. Please feel free to also talk about the current state of the application, if it exists today, and any specific gaps that you know will need to be overcome.

% Moving to the current and future generations of heterogeneous many-core systems will require a significant re-working of our analysis codes, both to deal with communication over hundreds of thousands of cores and to take advantage of accelerator opportunities such as GPUs.

% \subsection{Fusion (Scott Klasky) -- do not publish \label{fusionScott}}

% 1. What is the science/engineering problem? What is the purpose of the
% application?

% With ITER scheduled to operate before 2020, plasma fusion has become
% one of the highest priority research items for DOE. Scalability of
% first-principles fusion codes is an important part of fusion research
% needed to obtain predictive capability of plasma performance with
% first-principles physics fidelity. We want to build a realistic
% full-distribution kinetic code, in realistic ITER geometry, combining
% the strengths of each of the existing gyrokinetic codes (GEM, GTS,
% XGC, and GTC), and utilize the most up-to-date computer science,
% hardware design and applied mathematics technologies developed. The
% application that we will develop will be capable of 1) discovering
% Macroscopic Magneto Hydro Dynamics activity in burning plasmas, 2)
% Understand Energetic particle effects in fusion reactors, 3)
% Understand the heating in a fusion reactor, and 4) Understand the
% interaction of the plasma at the material boundary.

% 2. How will the application solve the problem?

% (a) Which of the following describes the application

%         i) large (exascale) data and large (exascale) computing (e.g., data mining, analytics, bioinformatics)

%         ii) large (exascale) data and modest computing (e.g., search)

%         iii) modest data and large (exascale) computing (e.g., simulation/modeling)

% The applications we currently run (XGC1, XGCP, GTC, GTS) run on all of the leadership class facilities, and require very large resources for the modest problems which we are currently solving. In the next generation (exascale) we will not only need all of the computing power that will be available at the time to understand ITER, we will also need to understand all of the data that will be generated during the simulation. Our techniques have been to examine and understand the data in situ, to reduce the total amount of data produced in the simulation. Today we have produced O(100 TB/week) for a simulation using our ADIOS middleware incorporated in our simulations. Using our staging techniques, we believe that we will only produce O(10 PB) on an exascale machine. (I don't know if this is considered modest dat, or extreme size data)?

% (b) Can you quantify the amount of computing vs the amount of data?

% 1 PB of data -- 10 PB of data, for an exascale size computation run in O(100 hours).
% Today we achieve about 6\% of peak with excellent scalability to 220K
% cores on the Cray XT systems. It is
% difficult to predict how we will run on
% accelerators, and next generation machines.

% 3. What infrastructure will be used? (including compute, data,
% network, instruments, etc.)

% For Validation, we will need to compare our data from the computer to
% ITER (or current fusion reactors). Typically the amount of data used
% in comparison is tiny by comparison (people only look at MB's form
% experiments).

% (a) What will be distributed?  data, compute, both?

% We always run on all available resources, so if we have computer time
% across the country, then our data will be distributed, and so will our
% computation. ITER will have data that will be distributed, but it
% still unknown.

% 4. What data be used in the application?

% I don't understand this question? We will input data (from perhaps a restart), and this data is typically in the ADIOS-BP file format for large data. Smaller data sets are typically stored in HDF5, and the smallest are stored in MDS+.

% (a) What are the types of data

% (b) What is the size of the data set(s)?

% Depends on what we want to understand --- For GTS/GTC we have typically
% looked at the particle data to compute fluxes, etc. and this data can
% be large, but due to the Monte Carlo error, and ways to generated 5D
% distribution functions from the particles, we reduce the particles to
% a manageable amount (300 GB per time step x 1 K time steps). Our
% current I/O is capable of reading this in $<$ 100 s per time step, which
% allows us to analyze the data in about 1-2 days.

% (c) Will any data be dynamic?

% Don't understand the question. Features change and are dynamic. The physics will get more complex as we continue to evolve our codes, and will be more dynamic in nature.

% 5. How will the application get the data?

% Reading the data from disk --- (Perhaps I don't understand the question).  It will also couple to other codes, and the coupling will be in-memory data exchange.

% 6. Are there time (or quality) constraints on the application?  (If so, what are they?)

% We always need to run as fast as possible since we have limited computing time (our team does have about 100M hours FY 11, but this still puts severe restrictions on what we run, so the faster we can run, the more science we can get done.

% 7. Please feel free to also talk about the current state of the application, if it exists today, and any specific gaps that you know will need to be overcome.

% Our biggest gaps will be in accelerating the code on next generation exascale platforms, and in processing as much data as possible in situ.

\subsection{SOA Astronomy applications from Adam\label{astroSOAAdam}}

1. What is the purpose of the application?

Calculates the photometric redshift for a given area of sky [RA, DEC] using two tools: HyperZ and ANNz, both accessible via WS interfaces.

2. How is the application used to do this?

The application is a service-oriented workflow. Services are orchestrated through a pipeline in order to compute the redshift of a given area of sky.

3. What infrastructure is used? (including compute, data, network, instruments, etc.)

The Wide Field Survey Archive (WFS):  image catalogue
Spectroscopic database: used to provide spectroscopic data for an area of sky

SExtractor tool: image processing, extracts all objects of interest (stars, galaxies etc.)
Cross matching tool: compiles one table of all objects of interest over the 5 wavebands

Hyperz: photometric redshift estimation algorithm 1
ANNz: photometric redshift estimation algorithm 2

mySpace: AstroGrid storage service

\katznote{checking more on this point}

4. What dynamic data is used in the application?

The WFS catalogue will be updated regularly, therefore a query at time x will not provide the same data as a query at time y.

(a) What are the types of data, (b) What is the size of the data set(s)?

Data sets (images) will be in the GB range, I can attempt to clarify these details.

5. How does the application get the data?

WS interfaces.

6. What are the time (or quality) constraints on the application?

Enough data needs to be available for a given area of sky from both the WFS archive and the spectroscopic archive.

\subsection{TNO + Thales (Netherlands)  -- Omer (Kees)}

will try to fill in...

\subsection{(Dan's) summaries of contributions}

\subsubsection{Apps that get data from files}

Silvia's Biosciences app (\S\ref{bioSilvia}): data is processed through a customized pipeline of analysis tools; processing of each data element is independent of other elements, so processing is decomposed/parallelized over the data elements.  All data is stored in files.

Steve's WLCG app (\S\ref{WLCGSteve}): there is a hierarchy of systems; data is centrally stored, and locally cached (and copied to where it likely will be used), perhaps at various levels of the hierarchy; processing is done by apps that are independent of each other; processing of one data file is independent of processing of another file, but groups of processing results are collected to obtain statistical outputs about the data.

Bob's VO (astrophysics) app (\S\ref{astroBob}): data is stored in registries with metadata; data access is through web services; there is a workflow concept of access and multiple processing steps; distributed scratch space can be used for storage of intermediate results; there is also a pub/sub mechanism for dynamic events, which are small in terms of the data required to identify an event and in terms of their frequency

Simon's Sensor Network app (\S\ref{sensorSimon}): data collected from sensors that periodically transmit; data brought to a central site; stored data analyzed using statistical techniques, visualized with tools like Google Earth.

Bioinformatics (from Tom Slezak) (source not in this doc): want to compare O(1000s) of terabye-scale data sets (metagenomic) with each other.  Datasets are distributed around the world.  Solution to the problem is not clear.

Climate (from Dan/Don) (\S\ref{climateDan}): Data (output of climate simulations) is distributed in multiple stores.  Current data analysis and visualization tools require the user to bring the data they want to analyze to a local system, then run the tools locally.  New tools will include the capability to add data transfer.  This will allow the set of tools to iterate - transfer data, do analysis, etc.

Astro (from Adam) (\S\ref{astroAdam}): Data taken by telescope - quick analysis done at telescope site for interesting (urgent) events (which may involve comparing new data with previous data).
Can get more data from other observatories if needed, request other observatories to take more data, or call a human. Data then transferred to archive site, which may be the observatory. At archive site, data analyzed , reduced, classified (some of this may be farmed out to grid resources).  Detailed analysis of new data vs. archived data done.  Reanalysis of all data done periodically.  Data stored in files and databases.

\subsubsection{Apps that get data from streams and files}

Jon's Environmental Data exploration app (\S\ref{envJon}): drive apps from graphical tools, inputs are geographic areas and algorithms to run on the data in those areas; data can come through web services from sensors or simulations (possibly running); provide responses in near-real-time.

Power Grids (from Shantenu) (\S\ref{powerShantenu}): a closed-loop cyber-physical system modeled as a workflow,
needs to combine streaming data arriving from sensors with historic
data available in archives, along with structured collections of
weather forecast data. Needs a workflow model that supports such
requirements.

Astro (from Julian) (\S\ref{astroJulian}): Application builds a sky map.  Reads data from detectors, and gets data from simulations performed on-the-fly.

\subsubsection{Apps that get data from streams}

Fusion (from Scott Klasky) (source not in this doc): multiple physics simulation codes run concurrently on a distributed set of parallel computers.  Data from some codes are streamed to other codes to link them into a single simulation.  Data is transformed in flight into needed inputs.  Data is also examined in situ to understand the overall state of the simulation.  Some data is stored for later data mining and visualization.

SOA Astro (from Adam) (\S\ref{astroSOAAdam}): services are orchestrated through a pipeline, including a data retrieval service.  Data moved through the pipeline, intermediate and final products can be stored in AstroGrid storage service.

\subsection{Additional sources}

\katznote{discuss this at April workshop}

\begin{itemize}
\item XLDB workshop series (characterizing aspects of dynamic data)
\url{http://www-conf.slac.stanford.edu/xldb10/Program.asp}
\item \url{http://www.scidb.org/}
\item Digging into Data (US \& UK - but somewhat global too) -
   \url{http://www.diggingintodata.org/}
\item Ocean Observatories Initiative (OOI, US-funded) -- Dan to talk to them
\item Roger Barga (Microsoft) -- Shantenu
\item John Polak and Haibo Chen (Intelligent Transport) -- Omer
\item Bartosz ``Bartek'' Dobrzelecki (EPCC) -- Health Informatics and Digital Humanities -- Adam Carter
\item Animal tracking (Bird tracking)  Willem Bouten -- Silvia
\item LOFAR (Low Frequency Radio Array) -- called ``ASTROWISE''  -- also
  supports provenance  tracking -- Bart Scheer
\item Astro Wise (Astronomical Wide-field Imaging System for Europe)
  http://www.astro-wise.org/ \jhanote{Who was the contact?}
\end{itemize}

\section{Vectors} {\bf Omer (lead), All)}

Vectors (Axes) for consideration for applications  (Shantenu, Jon, Omer, Neil, Simon, and Dan)
\begin{itemize}
\item  Used to differentiate between different kinds of dynamic data
\item Ratio of compute:data volumes (use this as a scale to place these applications)
\item Timeline to indicate how applications fit in on a time axis in terms of the dynamicity of the generated data
\item Data set size (unit of granularity) to support harvesting of data -- files, portions of files, data elements, grouping of these, etc
\item Data availability -- data not ready when you start the analysis (incomplete processing) -- either: (i) it could be, but is not; (ii) could never be.
\end{itemize}

Infrastructure capabilities and requirements  (Shantenu, Jon, Omer, Neil, Simon, and Dan)
\begin{itemize}
\item What should be provided by infrastructure (middleware, RDBMS, Cloud, Grid etc) vs. application
\item What are application characteristics: QoS-requirement vs. fault tolerance (automated) -- relate this back to the applications identified above
\item In-stream processing vs. step-wise processing of data
\end{itemize}

% Diagram suggestion (perhaps indicate what is currently available and what is required (desired): -- (Shantenu, Omer, Neil, Simon, and Dan)

\begin{table}
\begin{footnotesize}
\begin{center}
% \begin{tabular}{|p{0.25*\textwidth}|p{0.25*\textwidth}|p{0.25*\textwidth}|p{0.25*\textwidth}|}
\begin{tabular}{|l|l|l|l|}
  \hline Application  & Data Input & Processing and Analysis & Post-Processing  \\
  \hline
  \hline Medical imaging & - &  - &  - \\
  \hline WLCG/ATLAS & - &  - &  - \\
  \hline Virtual observatory & - &  - &  - \\
  \hline Environmental (Marine) & - &  - &  - \\
  \hline Bioinformatics (metagenomics) & - &  - &  - \\
  \hline Climate (CDAT and ESGF) & - &  - &  - \\
  \hline Astrophysics (LSST) & - &  - &  - \\
  \hline Astro (Cosmic Microwave Background) & - &  - &  - \\
  \hline Power Grids & - &  - &  - \\
  \hline Fusion & - &  - &  - \\
  \hline
\end{tabular}
\caption{Application Vectors\label{table:vectors}}
%  \jhanote{Omer: Can you relate the last column ``support sequential
%    and concurrent'' back to the text or explain why this field is
%    being explored please?}}
\end{center}
\end{footnotesize}
\end{table}

\section{Cyberinfrastructure} {\bf Shantenu, Neil (co-lead), All}

\subsection{CI to support Dynamic Data}

\subsection{CI to support Distributed Data}

\jhanote{Mining@HOME by Talia would be a good candidate to
  discuss/analyze}

``This paper presents the design and the implementation of an
architecture for the analysis of data streams in distributed
environments. In particular, data stream analysis has been carried out
for the computation of items and itemsets that exceed a frequency
threshold. The mining approach is hybrid, that is, frequent items are
calculated with a single pass, using a sketch algorithm, while
frequent itemsets are calculated offline by a further analysis. The
architecture combines parallel and distributed processing to keep the
pace with the rate of distributed data streams. In order to keep
computation close to data, miners are distributed among the domains
where data streams are generated. The paper also reports the
experiment results obtained with a prototype of the architecture,
tested on a Grid composed of two domains handling two different data
streams. Results confirm the scalability of the approach and show that
it is possible to keep the pace of data production and reduce the
communication overhead.''

``... In some cases streams of data must be analyzed in real time to
provide information about trends, outlier values or regularities that
must be signaled as soon as possible.  The need for online computation
is a notable challenge with respect to classical data mining
algorithms [1], [2].  Important application fields for stream mining
are as diverse as financial applications, network monitoring, security
problems, telecommunication networks, Web applications, sensor
networks, analysis of atmospheric data, etc.  A further difficulty
originates when streams are distributed, and mining models must be
derived not only for the data of a single stream, but for the
integration of multiple and heterogeneous data streams [3]. This
scenario can occur in all the application domains mentioned
before. For example, in a Content Distribution Network, user requests
delivered to a Web system can be forwarded to any of several servers
located in different and possibly distant places, in order to serve
requests more efficiently and balance the load.''

``... the architecture combines the parallel and distributed
paradigms, the first to keep the pace with the rate of a
single data stream, by using multiple miners (processors
or cores), the second to cope with the distributed
nature of data streams. Miners are distributed among
the domains where data streams are generated, in order
to keep computation close to data.''

\subsection{FlumeJava} (from Dan)

``FlumeJava is a distributed, reliable, and available service for
efficiently collecting, aggregating, and moving large amounts of log
data. It has a simple and flexible architecture based on streaming
data flows. It is robust and fault tolerant with tunable reliability
mechanisms and many failover and recovery mechanisms. The system is
centrally managed and allows for intelligent dynamic management. It
uses a simple extensible data model that allows for online analytic
application.''

\url{https://github.com/cloudera/flume}

\url{http://portal.acm.org/citation.cfm?id=1806638}

\section{Programming Systems} {\bf Simon, All}

\subsection{Architecture}

Top-level views of a system is typically as a processing pipeline from
collection, through pre-processing, to storage and then
post-processing and querying. (Collection includes both simulation and
sensing.) Different applications place different emphasis on the
different stages: some collect and store all the data they can, with
only minor pre-processing to remove noise; others must perform
enormous pre-processing to reduce the data volumes to manageable
dimensions. Another point in this spectrum is a classifier system that
uses a data stream to recognise higher-level events which it then
propagates, so it's the events that are interesting, not the data.

Question: where does each application lie in this space?

Question: is it the case that each application can be placed at an
appropriate point \textit{a priori}, or do characteristics change?

Question: is there a problem with multiple applications running on the
same fabric and interfering? -- for example a classifier system
running alongside a collection system and generating storms of events?

A lot of scientific processes are predicated on data being collected
and then analysed off-line, which emphasises both query performance
but also the collection and use of meta-data about the data set (for
example the precision of sensing and the provenance effects of any
pre-processing).

Question: is there a need to retain meta-data across the lifetime of a
system and across the complete architecture?

\subsection{Paradigms}

Often limited to pipeline \textit{versus} some stylised large-scale
process such as map-reduce, within the large architectural
pipeline. Often find scientific workflow systems like Taverna
describing pipelines in terms of components, possibly with specialised
high-performance constructions for some intensive
components. Generally find this in the ``traditional'' space of codes
running on workstations and servers; sensor networks themselves are
immature in this respect and typically focus on robust data
transmission without much in the way of in-network processing
(although this is changing, slowly).

Stream processing is a variant on pipelines where the stream is
processed locally and in real time, typically with only fairly small
operations allowed at each stage (for example calculating time
derivatives of a time series).

Question: what are the issues that face developers in using standard
scientific workflow approaches?

Question: would there be any need to push lots of processing to the
front end, near the sensing? And if so, what would this do to the
later stages?

Question: are there any significant application codes that use
anything other than pipelines, streams and map-reduce?

\section{Analysis and Road Ahead}

\section*{Acknowledgements}
This paper is an outcome of the UK e-Science Institute sponsored
Research Theme on Dynamic Distributed Data-Intensive Programming
Abstractions and Systems

\bibliographystyle{unsrt}
\bibliography{3dpas}

\end{document}

### 2015-05-29 old-sections/3dpas_programming.tex
adding other, mostly old material

\section{Programming Systems and Abstractions}
\label{sec:prog_sys}

% \note{The issue of programming systems for scientific applications is always
% vexing. On the one hand, scientific codes place requirements on
% applications that are often substantially different from commercial
% applications in terms of their data volumes and flow rates -- although
% this is becoming less true as industry discovers the value of ``big
% data''; on the other hand, scientific codes must increasingly
% interoperate with other services on the web, and so must appeal to
% standard protocols, data formats and libraries. The latter can be at
% odds with addressing the former, since it complicates the use of
% radically different programing models and tools.  However, the
% convergence of scientific and commercial interest in large-scale
% dynamic data should lead to improvements in the approaches applied in
% both domains.}

%\subsection{Architecture}

% Top-level views of a system is typically as a processing pipeline from
% collection, through pre-processing, to storage and then
% post-processing and querying. (Collection includes both simulation and
% sensing.) Different applications place different emphasis on the
% different stages: some collect and store all the data they can, with
% only minor pre-processing to remove noise; others must perform
% enormous pre-processing to reduce the data volumes to manageable
% dimensions. Another point in this spectrum is a classifier system that
% uses a data stream to recognise higher-level events which it then
% propagates, so it's the events that are interesting, not the data.

% \note{Scientific codes have long lifespans, as tools are refined and
% extended along with improving theoretical models and experimental
% techniques. Many experiments are inherently long-lived, for example in
% fundamental physics or space science, with the associated analysis
% software often being refined over the length of the mission.}

The design of scientific software evolved from an early emphasis on fully-functioning 
  monolithic applications to a library-based
  approach with a core of reusable code by one or more,
  perhaps changing, user environments. More recently there has been a
  further move to re-use through the use of workflow tools such as
  Taverna~\cite{Taverna}, Kepler~\cite{Kepler}, and
  Trident~\cite{Trident}. A data analysis process is
  expressed as an explicit dataflow or control flow composition of components each serving a
  dedicated function. This has simplified the development and the maintenance of complex
  codes, both by allowing common component re-use and by highlighting
  the structure of the processing pipeline.
% Unsurprisingly, more
  % recent domains such as genomics have adopted such technologies from
  % the start; older domains still have substantial legacy codebases
  % with less transparent structuring} \katznote{previous sentence is
  % far too absolute to be correct.}

% The view of scientific applications as pipelines hides significant
% differences between (and within) domains, since the different stages
% of the pipeline receive different emphasis in different
% applications. Some applications must collect and store all the
% incoming data, with only minor pre-processing to clean the data ahead
% of the main processing. For others, a major constraint is to reduce
% the incoming data stream as quickly as possible, perhaps discarding a
% large fraction of the data before any processing occurs. Taking some
% of the examples from \S\ref{sec:scenarios}, medical imaging and
% astrophysics often fall into the first category, whilst particle
% physics (especially in the case of the Large Hadron Collider) often
% falls dramatically into the second -- to the extent that principled
% data reduction forms one of the major scientific challenges for the
% project.

% \note{One way to integrate legacy code -- or those codes too
%   tightly-coupled for a workflow structure -- is to build pipelines
%   from ``macro'' components. In this view the pipeline connects groups
%   of components that are not themselves pipelines. Conversely one
%   might have multiple pipelines running concurrently over a partially
%   shared set of components, driven by data dependencies and exchanging
%   data in more complex ways than are seen in simple pipelines.}

% There are also significant variations in where ``the pipeline'' begins
% and ends. Many systems are driven by data sets collected and shared
% widely, against which several different applications may be
% applied. Climate science is a good example of this approach. In this
% case the data is ``pre-collected'' in the sense of existing outside
% most of the applications, although its collection may have been a
% complex task. The raw data required for gene sequencing
% (\S\ref{bioSilvia}), for example, while being collected using a
% dedicated and very sophisticated computerized processing system, would
% generally be regarded as ``given'' and not part of the processing
% pipeline, since the pipeline would not affect the collection
% task. Another way to look at this is that data is collected
% \emph{independently} of its further processing.

% Alternatively the application may be driven directly
% from a live data stream, with processing happening on the
% fly. The front-end data collection sub-system is becoming increasingly
% flexible and capable, with the introduction of sensor networks to
% replace more passive approaches, and it may often make sense to
% include the sensor network into the application workflow pipeline. All
% these choices have implications for design and provisioning to which
% we return below.

% % Question: where does each application lie in this space?

% % Question: is it the case that each application can be placed at an
% % appropriate point \textit{a priori}, or do characteristics change?

% The choice of which sort of pipeline a particular application uses may
% well be fixed, or will change only slowly as the experimental
% environment evolves. In particular we see some systems which (for
% example) add simulation alongside experimental data collection to act
% as a verification and validation step.

% \note{Viewed from the perspective of the 3DPAS vectors
% (\S\ref{sec:vectors}), programming systems can be seen to exert
% comprehensive influence. In terms of \textit{communication and data},
% the key question is the complexity and extent of the system being
% discussed: does it extend to the collection of data, or to its
% presentation and interaction, or does it simply focus on the
% ``traditional'' processing steps? Does it provide a simple, coherent
% structure (like a pipeline), or does it require more complex,
% \textit{ad hoc} communications? Similarly, the \textit{coordination}
% vector is influenced both by how a computation adapts to on-going
% changes and to where control of this adaptation resides. The
% \textit{execution environment} vector is more centrally the topic of
% this section, and we explore these issues in more depth below,
% \S\ref{sec:issues} after examining some alternatives to pipelined
% architectures.}

%\subsection{Beyond Simple Pipelines}

% Question: what are the issues that face developers in using standard
% scientific workflow approaches?

The workflow approach is simple and attractive to compose sequential data processing pipelines and
offers the flexibility of modeling more complex constructs. However, more specialized programming
abstractions are becoming popular to address large scale and dynamic data processing.
% abstract-away many of the issues that are of immense relevance to
% efficient scientific computing. We will first 
We consider two alternative paradigms of bulk and stream data processing, before considering the
issues of adaptation and provisioning of these models.
%  and otherwise refining scientific
% workflows. Note that in many cases these alternative paradigms
% actually integrate very well into an overall pipeline, architecture,
% providing components that can fit into a general pipelined framework.

\subsection{MapReduce and Bulk Synchronous Parallel (BSP) Models}

% Often limited to pipeline \textit{versus} some stylised large-scale
% process such as map-reduce, within the large architectural
% pipeline. Often find scientific workflow systems like Taverna
% describing pipelines in terms of components, possibly with specialised
% high-performance constructions for some intensive
% components. Generally find this in the ``traditional'' space of codes
% running on workstations and servers; sensor networks themselves are
% immature in this respect and typically focus on robust data
% transmission without much in the way of in-network processing
% (although this is changing, slowly).

% Many criticisms of traditional programming languages begin from the
% challenges posed by the so-called \emph{von Neumann bottleneck}
% introduced by processing data in small chunks, for example within
% iterative loops of tail-recursive functions. In either case, the
% bottleneck comes from imposing linearity into the way the data is
% processed. Many high-performance approaches, perhaps beginning with
% various vectorizing Fortran compilers, have sought to sidestep the
% bottleneck by providing \emph{bulk} operators that express
% computations over collections of data, that can be more effectively
% parallelized.

The MapReduce paradigm~\cite{MapReduce} is the most prominent and recent example of
using \emph{bulk} operators that express computations over collections of data, that can be more
effectively parallelized.  MapReduce forces programmers to model computation in two
stages: a \emph{map} phase in which a transformation is applied to
every element of a collection independently of every other; and a
\emph{reduce} phase in which the transformed results are aggregated
through a binary function that is associative and commutative. These
properties serve to guarantee that the MapReduce computation can be
efficiently parallelized, by performing the map phase concurrently and
then reducing without the reduction order affecting the final
result. The result is to provide a large amount of ``excess''
parallelism that can utilize as many cores or processors as are
available.

% MapReduce arose in part from earlier work on ``skeletal parallelism''
% that generalize the idea of bulk operators to encompass any
% structure that can perform efficient parallel computation. The
% skeleton is essentially a higher-order function that captures the
% structure of the computation, which is then populated with
% application-specific functions to provide the actual calculation. The
% skeleton ensures that the overall structured computation can be
% computed efficiently, potentially making use of significant code
% generation and optimization. Flask~\cite{FlaskHaskell} is a modern example of this
% approach applied to sensor networks.

Whilst often not suitable as a complete system architecture,
MapReduce can be used in a workflow environment to provide scalability to
processor-intensive components, or for data-intensive applications
where it can address the relatively low scalability of I/O
capacity. It does not necessarily work effectively in a streaming
mode, and cannot be relied upon to adapt its structure according to
different data volumes: the structure is generally parameterized by
the processing resources available regardless of the data, rather than
making requests for appropriate resourcing based on the data
presented~\cite{DataIntensiveMapReduce}. Variants of the basic MapReduce model have been proposed,
such as iterative~\cite{twister} and hierarchical~\cite{hmr} MapReduce, to address some of these limitations. Framework implementations
such as Apache Hadoop~\cite{hadoop}, Microsoft Daytona~\cite{daytona}, and Twister~\cite{twister} are available, among others.

Bulk Synchronous Parallel (BSP) processing~\cite{valiant1990bsp} was developed in the 1990 as a bridging model for
parallel computing. This programming abstraction is garnering attention as a way to reduce the I/O complexity of
MapReduce and offer an easier message passing model than MPI. It is being used for data intensive
applications such as matrix and graph processing, with framework implementations by Apache Hama~\cite{hama},
Google Pregel~\cite{pregel} and others~\cite{goldenorb}. The key approach is to separate processing into supersteps where independent
processes operate concurrently on messages received from the previous superstep. This processing is
interleaved with message passing to processes in the next superstep. When a process completes its
execution of a superstep, it reaches a barrier that is synchronized with all other processes in the
superstep. Once synchronized, the next superstep starts.

\subsection{Stream and Event Processing}

% Stream processing is a variant on pipelines where the stream is
% processed locally and in real time, typically with only fairly small
% operations allowed at each stage (for example calculating time
% derivatives of a time series).

Stream processing is an approach modeled on signal processing, in
which a stream of data is transformed on-the-fly to produce one or
more further streams of values. As such it is a programming abstraction suitable for
continuous data feeds, or when results need to be pushed to users of interactive applications. Since
the data is not assumed to be available 
\textit{en masse}, it can be used in situations where storing the
entire data set is undesirable, unnecessary or impossible. Equally,
the approach is simple to understand and mirrors the composition of workflows. However, since
they operate continuously on a seemingly infinite stream of data, notions of data freshness,
processing latencies, resilience/checkpointing, and incremental completion become more
interesting. These introduce new data and composition abstractions
(e.g., Landmarks~\cite{chandrasekaran:telegraphcq2003}, Dynamic Stream
Rates~\cite{Simmhan:sciencecloud:2011}) and trade-offs during execution.

A simple example of stream processing occurs at the front end of a
pipeline, in data cleansing. A processor might perform simple sanity
checks such as outlier removal based on expert knowledge or on
properties calculated from the stream itself -- for example to remove
any data that lies more than two standard deviations from the running
mean. The analogy to signal processing is quite clear: the stream is
processed as a sequence of samples, used to extract stream-level
values such as moving average (mean), time derivatives and the
like. These values may be used in further processing, or may be output
in their own right alongside the raw data and used to feed other
processing steps.

A higher level of abstraction in stream processing is \emph{complex event processing}, where a
stream is processed to extract ``events'' that 
are notionally occurring at a higher semantic level than the raw
data. An example might be the detection of lights being left on in a room (ambient light sensor
stream) when no one is occupying it (infra-red sensor stream). The significant point is that the raw data
stream pre-processed, classified and packaged as a stream of events with timestamps, identifiers and properties on which
higher-order operations like standing-queries and pattern-matching can be performed. Events can even
be associated with semantic annotations~\cite{Teymourian:EDBT:2010, Zhou:debs:2011}.

There are many examples of stream
~\cite{chandrasekaran:telegraphcq2003, abadi:borealis2005, liew:streamgraph2010} and event
~\cite{ESPER, siddhi, MicrosoftInsight} processing frameworks, and recent works have combined
streaming with more traditional file-based workflows~\cite{ZinnCluster,Herath:CCGrid:2010}.

% Question: are there any significant application codes that use
% anything other than pipelines, streams and map-reduce?

% \note{\subsubsection{Specialized Algorithms} Inevitably many
%   scientific codes do not fit into either of these broad schemata and
%   so must be coded in a more \textit{ad hoc} manner. Pattern-matching
%   for gene sequencing (\S\ref{bioSilvia}) and particle physics
%   (\S\ref{WLCGSteve}) are good examples, where domain-specific
%   knowledge is used to optimize traditional pattern-matching.}

% \note{Programming systems such as MPI~\cite{MPI} provide a basis for distributed
% applications, although it is vital not to under-estimate the
% difficulties involved. While such systems allow developers to generate
% very complex patters of distribution and concurrency, they do so at
% the cost of a massive increase in coding complexity -- often beyond
% the programmers' capacities to manage efficiently. Significant
% performance figures are often elusive, and may be fragile in the sense
% that small changes in approach and algorithm destroy the
% gains. Systems that \emph{can} be phrased according to a structured
% algorithm or processing schema with known performance benefits
% therefore \emph{should} be.}

% \subsection{Issues}
% \label{sec:issues}

% \note{So far little we have said has addressed the essential aspects
%   of dynamic data, in particular the unpredictability and scaling
%   features that seem inherent in dealing with such data sets and
%   sources: the \textit{application adaptation} vector. However, it is
%   clear that dynamic data raises some important questions about the
%   management of scientific codes, many of which cross between the
%   various paradigms chosen for expressing the problems.}

% % Question: is there a problem with multiple applications running on the
% % same fabric and interfering? -- for example a classifier system
% % running alongside a collection system and generating storms of events?

% \note{We may divide the issues raised into two broad categories. The first
% pair of \textit{intrinsic} issues capture concerns such as
% provisioning and adaptation that relate to codes themselves in terms
% of their development, composition and management. The second set of
% \textit{extrinsic} issues are more concerned with software engineering
% factors such as the extent of the pipeline and the representation of a
% dataset's history and provenance.}

\subsection{Provisioning}

Chief among the management questions for dynamic data processing frameworks is the
issue of \emph{provisioning}: allocating sufficient and appropriate
resources to a component to allow it to perform most efficiently. This
must be set against the need to manage the interactions of components,
leading to a difficult resource allocation and scheduling problem.

Provisioning can have two different goals: one of optimally using
resources (maximizing resource usage), while the other to ensure
efficient application execution. While the former goal optimizes
provisioning from an infrastructure provider's perspective, the latter
goal is more driven by the application's quality of service (QoS) requirement -- something that
is necessary to be met for the results of the application to be
useful. The latter can be expressed as in specific terms (e.g. makespan, freshness, throughput), or may
simply be a best effort. Expressing requirements becomes difficult in
itself with dynamic applications that may produce continuous,
intermediary results.

Even if one can appropriately calculate the resource needs of a
component for a given data set -- which is in itself non-trivial --
dynamic data adds the need to deal with an unknown and possibly
changing optimal allocation. Changes to data rates, sizes, number of streams and processing
complexities can impact resource needs. Further, the programming
model itself can be inflexible when it comes to allowing dynamic
provisioning. Algorithms may not allow more nodes to be added midway
through a computation, and some data processing systems require all tasks to
be mapped to resources at the time the execution launches. All these
restrictions impede flexible adaptation.

% Commercial systems have addressed provisioning in part by using
% \emph{virtualization}, splitting a computation between a variable
% number of identical virtual machines that can be spun-up and -down on
% top of a smaller number of more capable servers. Scientific codes
% highlight many of the broader problems and benefits of
% virtualization. It can be hard to determine \textit{a priori} the
% computational needs of a component, which makes it hard to
% provision. Co-locating an unexpectedly large component with others may
% impact on the overall performance of the system. However, virtualizing
% the components as independent virtual machines potentially allows
% components to be frozen and re-located, allowing dynamic
% re-provisioning and optimization. This of course leads to further
% problems in the analysis and planning of such operations, but at least
% provides a technical basis for managing the performance of services
% across their execution lifetimes.

While most attention is given to
scaling-out computations (to handle more data), scaling down to
prevent waste of resources is equally problematic. This is particularly so as 
\emph{cloud computing} gains prominence as a cyber-infrastructure. Cloud platforms take
resource virtualization to another level, 
presenting computing as an elastic service than can be ramped up and down.
However, current cloud computing platforms are
highly targeted at scenarios of commercial interest such a
large volumes of independent transactions. While these may match some
scientific problems (such as some issues described in
 \S\ref{bioSilvia}  and \S\ref{WLCGSteve}), it is by no means suitable for
others. The multi-tenancy of clouds whereby the same resource is shared across customers also
introduces dynamism in the performance offered by the
infrastructure~\cite{iosup:ccgrid:2011}. 
Programming models such as MapReduce and the ease of getting access to resources have increased the
use of clouds for eScience, and cloud vendors such as Amazon have also responded to scientific
computing needs with HPC and GPGPU infrastructure. \ysnote{should we move the preceding line to ``D3 Infrastructure''?}
Significantly more research is needed to determine 
the gaps in existing cloud computing platforms, be it on the cloud provider's end or in developing
novel scientific programming models  to meet the needs of the broader scientific domains.

% \note{\subsubsection{Co-existing Services}}

% \note{Provisioning also extends to the co-existence of multiple services and
% components on the same substrate, where their provisioning needs may
% conflict or interact in peculiar ways. Indeed, much of the interest
% that computer scientists have in virtualization comes from these
% interactions. The variability of provisioning makes these challenges
% even more severe for scientific codes.}

% \note{At another level, service co-existence is strongly linked with
%   service-oriented architectures. Scientific codes that provide common
%   and widely-utilized services, for example in astronomy
%   (\S\ref{astroSOAAdam}), demonstrate the potential that these
%   architectures have. Even in systems for which there are fewer
%   outside users, SOA is a useful structuring mechanism within
%   pipelines and can be deployed generically by workflow engines. The
%   advantages in terms of the availability of re-usable components --
%   whether or not such re-use is foreseen -- are potentially highly
%   significant in fields (such as climate modeling, \S\ref{modisKeith})
%   where variations on analytic themes are commonly employed. SOA also
%   raises the issue that the program may be using existing shared or
%   external services that cannot be relocated because it is not in the
%   workflow author's control. It may or may not be possible to locate
%   the user-controlled service to co-locate with the external shared
%   service.}

\subsection{Adaptive Programming Systems}

% Question: would there be any need to push lots of processing to the
% front end, near the sensing? And if so, what would this do to the
% later stages?

As noted earlier, it is often not clear where a particular scientific
pipeline begins and ends. In particular we are seeing increasing use
of sensor networks and other intelligent front-end data collection
systems, as well as the integration of simulation and stream
processing. These components often have the feature of being
\emph{adaptive}, in that they can vary their exact behavior according
either to external control stimuli or from observations made of the
data they are themselves observing.

Adaptation may also be through users' interactively changing the
direction of the pipeline, either by changing parameters or by
adding or removing tasks in the pipeline. This allows decision logic
to be initiated by the pipeline itself, or using an external
control signal that may come from users or an instrument. This form of
``computational steering'' is especially useful in systems being used
to explore a data set.

What impact does this have on other stages? The advantage of including
such systems is that the pipeline then provides an end-to-end
description of the entire scientific process. This is valuable both
for documentation purposes, but also in pursuing issues of component
and data re-use.

% A lot of scientific processes are predicated on data being collected
% and then analysed off-line, which emphasises both query performance
% but also the collection and use of meta-data about the data set (for
% example the precision of sensing and the provenance effects of any
% pre-processing).

% \note{Adaptivity brings its own challenges, many of which are often captured
% under the notion of \emph{autonomic systems} which sense their own
% operation, decide on appropriate management actions to maintain their
% quality-of-service parameters, and then act to affect their own
% operation~\cite{KephartAutonomics,AC-Survey}. This is substantially
% different from the user-driven adaptations discussed above, focusing
% on managing a process rather than changing it on the fly. A
% substantial amount of research is already on-going in computer
% science, often centered on data centre management and provisioning (see
% above); an emerging strand of work is looking at autonomic techniques
% applied to scientific systems. The importance of these approaches is
% that they allow improvements in the service offered to workflows by
% adaptive components. However, this is accompanied by the danger that
% adaptation may not be scientifically benign, and may affect the
% results of the workflow itself. This leads on to issues of data
% provenance below.}

% \note{Many current techniques in autonomic systems approach the
%   problem very much from a systems perspective rather than from a
%   scientific or workflow perspective. This matters, as it makes it
%   harder for users to affect, control or understand the impact that
%   autonomic adaptation may have on their overall analysis. (An example
%   of this is the adaptive processing of a sensor data stream that --
%   if not recorded -- may affect the validity of later statistical
%   analyses. These are to some extent issues of provenance and are
%   discussed more below.)}

% \note{Including sensing into the pipeline (\S\ref{sensorSimon})
% allows analysis to inform collection. A good
% example is where a feature-recognition analysis detects a feature of
% interest, which can then be examined in more detail. Not all
% application areas are amenable to this approach, for example where
% data is collected \textit{en masse} and at a consistent granularity
% and regularity, perhaps independent of \emph{any} particular
% processing function. For many systems, however, the ability to
% feed-back analysis from later stages is of great
% significance. Understanding the stability and dynamics of such
% feedback loops remains a challenge in autonomic systems, especially
% when the feedback deviates from the simple closed-form specifications
% amenable to control theory: more dynamic approaches from game theory,
% physics and econometrics show promise.}

% \note{At the other end of the pipeline, interactive exploration and
%   computational steering (\S\ref{envJon}) involve the scientist and
%   data analyst in providing guidance as to the way in which
%   computation should proceed, affecting either current or future
%   runs. The challenge here is that the ``user'' at this stage is often
%   not the ``programmer'', and so may only be able to provide guidance
%   at a scientific level, rather than at a algorithmic or architectural
%   level. This having been said, such guidance is potentially extremely
%   valuable and allows insights directly from the level of the
%   ``mission'' of the scientific system. This suggests that data
%   visualization and interaction should be considered core to the
%   processing pipeline, and that attention be paid to how scientific
%   and data insights can be fed-back as autonomic control
%   parameters. This is not an area that has received any significant
%   attention from the autonomic systems community, and so offers a
%   major research avenue.}

% %\subsubsection{Collecting Provenance}

% % Question: is there a need to retain meta-data across the lifetime of a
% % system and across the complete architecture?

% \note{Finally we must consider the fact that data processing in a scientific
% setting is itself a scientific process, which introduces properties
% into the data that must be accounted for in a predictable and
% well-founded manner. This implies that the process is well-defined,
% describable and analyzable to explore its effects on the results
% obtained.}

% \note{A simple example of the effects of different choices comes from data
% cleansing. Deciding which elements are ``noise'' in a data set (or
% data stream) may exert a profound effect on the results of further
% processing. Discarding data that is \textit{not} noise will distort
% the results; discarding too many data points may damage the
% statistical properties of the data; while discarding too few may be
% distorting in other ways. The point is not that the data is
% transformed in ways that change its properties: rather, the issue is
% the \textit{hiding} of these transformations in ways that cannot be
% accounted for in future processing stages or interpretation.}

% \note{There is no way that a programming system or model can automate
% decisions of this kind, which remain essentially scientific. What a
% system \textit{can} do, however, is to ensure that the decisions made
% are recorded and propagated with the data and its results. This is the
% issue of the \emph{provenance} of the data and the results: recording
% the transformations that have been made to a data set or data stream
% and associating this \emph{meta-data} alongside any derived data. We
% can view provenance meta-data as a generalized kind of error bar on
% computations -- or, equally, as the information needed to produce
% standard error bars in a well-founded and traceable
% manner.}

% \note{There has been several years' work in developing a community standard,
% the Open Provenance Model~\cite{OPM}, which is leading to more formal
% standards such as the efforts of the W3C Provenance Working Group to
% provide frameworks for the capture and representation of
% provenance. Scientific workflow systems like Taverna are increasingly
% able to handle provenance attachments automatically and maintain them
% between processing stages~\cite{Taverna}, and there are dozens of stand-alone
% provenance systems as well. One may regard this as another example of
% moving analysis into the computer systems and away from the scientists
% to ensure faithful recording.}

% \subsection{Coverage}
% \label{sec:coverage}

% \begin{table}
%   \begin{scriptsize}
%     \begin{center}
%       \caption{Programming systems used by the applications\label{table:progSystems}}
%       \begin{tabular}{|p{6.3cm}|p{7.2cm}|}
%         \hline Application  & Current approaches \\
%         \hline
%         \hline NGS Analytics & pipelines that ship with
%         sequencers~\cite{illumina}; workflows~\cite{Taverna} \\
%         \hline ATLAS & custom algorithms; job submission tied to WLCG \\
%         \hline LSST & custom layer; scripted workflows; queries \\
%         \hline SOA Astronomy & web services \\
%         \hline Cosmic Microwave Background & \textit{ad hoc} \\
%         \hline Sensor Network Application & focused on collection \\
%         \hline Climate & \textit{ad hoc} \\
%         \hline Interactive Exploration of Environmental Data &
%         visualization tools \\
%         \hline Power Grids & stream processing; data mining \\
%         \hline Fusion (ITER) & scripted pipeline \\
%         \hline Industrial Incident Notification and Response &
%         agent-based  \\
%         \hline MODIS Data Processing & scripted pipeline \\
% %        \hline Floating Sensors & \textit{ad hoc} \\
%         \hline Distributed Network Intrusion Detection & monolithic
%         packages \\
%         \hline
%       \end{tabular}
%     \end{center}
%   \end{scriptsize}
% \end{table}

% We conclude this section by revisiting the application scenarios of
% \S\ref{sec:scenarios} in terms of their programming system and using
% this to pose two questions:

% \begin{enumerate}
% \item What programming systems and techniques are \emph{currently}
%   used in implementing the different systems?, and
% \item What elements identified in this section \emph{could} usefully
%   be deployed or further explored in each system?
% \end{enumerate}

% Table~\ref{table:progSystems} summarizes the programming systems
% currently dominant in each application case study. It is worth noting
% that -- in common with most long-lived systems -- many of the cases
% have accreted substantial code bases over time and so show little in
% the way of systematic architecture. (One reason for the dominance of
% pipelines may be that such architectures are quite suitable for adding
% extra processing steps to existing computations.)

% From this we can make several observations and recommendations. The
% use of scripted or \textit{ad hoc} pipelines can in most cases be
% easily replaced with a structured workflow engine. Doing so brings
% immediate benefits in terms of reproducibility, but also in terms of
% standardization of components and interfaces. Standardization is not
% simply a software issue: it has scientific significance in the sense
% that data produced using well-documented, well-used and well-debugged
% suites of components is often more trusted -- and trustworthy -- that
% data produced using locally-produced components that, while they may
% be more innovative and targeted, also run the risk of undetected
% assumptions and errors.

% A further implication of standardization is that component re-use is
% greatly improved. This is perhaps best illustrated by observing the
% need for collecting data provenance over processing runs. The
% recording of processing of processing steps and the like is completely
% generic and can be carried out by the workflow engine uniformly across
% all workflows. The use of standard component-based software techniques
% such as web services, coupled with web technologies for metadata
% representation, can further assist in re-use, re-purposing and
% integration.

% \note{One place in which \textit{ad hoc} approaches may improve over
% standardized engines is in adaptation: the \textit{coordination}
% vector. A targeted approach can make use of application-specific
% analysis and control, whereas a generic approach needs to be able to
% describe the adaptation strategy for \textit{any} application. This is
% a challenge at the current state of the art in autonomic systems, but
% again, if addressed, would make it easier to apply advanced solutions
% across domains and reduce the barriers to entry in creating adaptive
% codes.}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "3dpas_intro"
%%% End:

### 2015-05-29 old-sections/3dpas_public_paper.tex
adding other, mostly old material

--- 2015-05-29 old-sections/3dpas_intro_public_do-not-edit.tex
@@
 
-\author{ Shantenu Jha, J.D. Blower, Neil Chue-Hong, \\[-0.1em]
-Simon Dobson, Daniel S. Katz, Omer Rana}
+\footnotesize{\author{Editors: J.D. Blower, Neil Chue-Hong, Simon Dobson, \\[-0.1em]
+  Shantenu Jha, Daniel S. Katz, Omer Rana}}
 
@@
 \begin{abstract}
-  Many problems at the forefront of science, engineering, medicine,
+\footnotesize{Many problems at the forefront of science, engineering, medicine,
   and the social sciences, are increasingly complex and
@@
   evolving document represents work in progress of the 3DPAS Research
-  Theme.
+  Theme.}
 \end{abstract}
@@
 
+\begin{figure*}[htbp]
+  \begin{center}
+    \includegraphics[scale=0.33]{figures/3dpas-wordle.pdf}
+    \caption{Wordle of 3DPAS document}
+    \label{Fig:wordle}
+  \end{center}
+\end{figure*}
+
 \section{Introduction: Context, Scope and Outline}
-
-\jhanote{Overall introduction, leading into some specific examples.}
-
-\section{Understanding Distributed Dynamic Data}
-
-Why bother about distributed, dynamic and data-intensive applications?
-\jhanote{(Shantenu, Omer, Neil, Simon, and Dan)}
-
-\subsection{Importance of Distributed}
-
-\subsection{Importance of Dynamic and Examples}
-
-There is a general appreciation that the challenges of data-intensive
-applications span beyond large-volumes and management/curation.  One
-specific attribute that we believe is pervasive but not necessarily
-obvious is “dynamic data”. The belief is based upon the fact that data
-is often distributed, and as a consequence many attributes of
-distributed systems permeate to data. Additionally the requirement of
-many applications impose a requirement that data is dynamic.
-
-\subsection{Types of Dynamic Data}
-Some example of such dynamism:
-
-\begin{itemize}
-\item Change of data operated upon.
-\item Variability of data rate -- arrival, consumption or ability to generate (due to dependencies)
-\item Data placement and scheduling issues are time dependent
-\item The nature of coordination between the different execution units
-changes
-\end{itemize}
-
-\jhanote{Possibly merge?}
-
-\begin{enumerate}
-
-\item Streaming data -- data comes and goes
-  \url{http://traffic.berkeley.edu/} and
-  \url{http://lagrange.ce.berkeley.edu/fsn/}
-\item Data set that is being operating upon is changing
-\item Data has to be transformed
-\item Placement decisions time-dependent
-\end{enumerate}
-
-\jhanote{We should highlight which of the above dynamic case are to be
-  found in applications}
-
-\katznote{Question: is there anything unique about scientific work on
-  streaming data?  Or is all work on streaming data basically them
-  same, whether it's in science or finance, for example?}
-
-\jhanote{Good Q: I think other than scale -- of data volume, and
-  degree of variability they are not too dissimilar}
-
-\subsection{Characterising Dynamic Applications}
-
-Amdahl’s number has been useful in characterising “static”
-data-intensive applications and systems. We believe there is scope to
-generalize Amdahl’s number to capture attributes of dynamic data
-applications.
-
-\subsubsection{Dynamic Data in a Web Environment}
 
@@
 
+\section{Understanding Distributed Dynamic Data}
+
+Why bother about distributed, dynamic and data-intensive applications?
+
+\subsection{Importance of Distributed}
+
+\subsection{Importance of Dynamic and Examples}
+
+\subsubsection{Types of Dynamic Data}
+
+\begin{enumerate}
+\item Streaming data -- data comes and goes
+  \url{http://traffic.berkeley.edu/} and \url{http://lagrange.ce.berkeley.edu/fsn/}
+\item Data set that is being operating upon is changing
+\item Data has to be transformed
+\item Placement decisions time-dependent
+\end{enumerate}
+
+\jhanote{We should highlight which of the above dynamic case are to be
+  found in applications}
+
+\katznote{Question: is there anything unique about scientific work on streaming data?  Or is all work on streaming data basically them same, whether it's in science or finance, for example?}
+
+\subsubsection{Dynamic Data in a Web Environment}\label{sec:dynamicdataweb}
 \emph{Just some placeholder notes from Jon.  Could belong elsewhere.}
 
-Web Services are commonly used to distribute data through HTTP.  A key
-characteristic of dynamic data, of course, is that it changes with
-time.  It is of great importance for the user to be confident that she
-is accessing the correct version of the data; this will often be the
-most up-to-date data, or may be a specific historical version.  In a
-Web environment, servers and proxy servers (which are sometimes beyond
-the control of the data provider or the data user) may retain caches
-of data in a strategy to reduce server load and network traffic.  For
-this reason, the HTTP protocol provides mechanisms for defining expiry
-times for data resources, specifying when a resource ought to be
-cleared from the cache.
-
-In the environmental and geospatial communities the Open Geospatial
-Consortium specifications are being widely adopted alongside existing
-protocols such as OPeNDAP.  Unfortunately these protocols, although
-built atop HTTP, do not make it easy for this versioning to be
-implemented correctly.  The OPeNDAP protocol does not have the concept
-of a dataset version, meaning that clients have no reliable or
-efficient means to detect that a dataset has changed.  The OGC
-protocols in general [TODO: look into this], \emph{do} provide
-versioning at the level of the service endpoint, but provide no
-information on when a resource should be considered expired.  Clients
-are therefore forced to poll the server to check for updates.
-
-Therefore we see that in a ``pull'' environment such as the Web, it is
-necessary for servers to advertise not only the version of a resource,
-but also the expiry time in order to handle dynamic data correctly.
-The frequency of updates is also useful information for service
-consumers.  Expiry times can be specified using the mechanisms of HTTP
-(although this capability is often not used in practice), but resource
-versioning and update frequency must be handled at a higher level.
-
-[All this must be fairly well-known and almost certainly covered in
-the literature, with the possible exception of the material on OPeNDAP
-and OGC.  Any references we could cite?]
+Web Services are commonly used to distribute data through HTTP.
+A key characteristic of dynamic data, of course, is that it changes with time.
+It is of great importance for the user to be confident that she is accessing the correct version of the data; this will often be the most up-to-date data, or may be a specific historical version.
+In a Web environment, servers and proxy servers (which are sometimes beyond the control of the data provider or the data user) may retain caches of data in a strategy to reduce server load and network traffic.
+For this reason, the HTTP protocol provides mechanisms for defining expiry times for data resources, specifying when a resource ought to be cleared from the cache.
+
+In the environmental and geospatial communities the Open Geospatial Consortium specifications are being widely adopted alongside existing protocols such as OPeNDAP.
+Unfortunately these protocols, although built atop HTTP, do not make it easy for this versioning to be implemented correctly.
+The OPeNDAP protocol does not have the concept of a dataset version, meaning that clients have no reliable or efficient means to detect that a dataset has changed.
+The OGC protocols in general [TODO: look into this], \emph{do} provide versioning at the level of the service endpoint, but provide no information on when a resource should be considered expired.
+Clients are therefore forced to poll the server to check for updates.
+
+Therefore we see that in a ``pull'' environment such as the Web, it is necessary for servers to advertise not only the version of a resource, but also the expiry time in order to handle dynamic data correctly.
+The frequency of updates is also useful information for service consumers.
+Expiry times can be specified using the mechanisms of HTTP (although this capability is often not used in practice), but resource versioning and update frequency must be handled at a higher level.
+
+[All this must be fairly well-known and almost certainly covered in the literature, with the possible exception of the material on OPeNDAP and OGC.  Any references we could cite?]
 
@@
 
-\katznote{new suggested questions from Don Middleton: How much diverse data integration is involved?  how diverse is the data?}
-
 \subsection{Silvia's Biosciences app: NGS, medical imaging (Silvia)\label{bioSilvia}}
@@
 
-The application performs analysis of DNA sequencing data for various goals in life science research, e.g., Mutation screening, Virus discovery, Genome-wide Associations (GWA), Linkage analysis, Comparison of bacterial genomes, microRNA expression and Exome sequencing.
-The data analysis is composed of various methods that are implemented by a bioinformatician, whereas the data is typically owned by the life scientist.
+The application performs analysis of DNA sequencing data for various
+goals in life science research, e.g., Mutation screening, Virus
+discovery, Genome-wide Associations (GWA), Linkage analysis,
+Comparison of bacterial genomes, microRNA expression and Exome
+sequencing. The data analysis is composed of various methods that are
+implemented by a bioinformatician, whereas the data is typically owned
+by the life scientist.
 
@@
 
-Note: Here we focus on data analysis from the perspective of the bioinformatician, excluding steps related to data acquisition (transformation of raw images into sequences of amino acids) and posterior biostatistics and epidemiological analysis.
-
-The data analysis is implemented as a pipeline of generic and customized components, mostly third-party public software tools.
-The pipeline may differ for each specific study, depending on the data acquisition and the research question.
-Typically it includes data preparation (file conversion, reformatting and filtering); alignment (comparison) with some reference database as the human genome; post-processing (output conversion, reformatting, statistical analysis) and visualization.
-Sequence alignment is the most computing-intensive step, which became prohibitive for sequential execution in next generation sequence experiments.
-All data is stored in files that are often completely read in memory for fast processing.
-Compressed formats are usually adopted.
-Although the files are large, the data can be easily decomposed because the processing is done on each sequence (`string') at a time.
-
-3. What infrastructure is used? (including compute, data, network, instruments, etc.)
+Note: Here we focus on data analysis from the perspective of the
+bioinformatician, excluding steps related to data acquisition
+(transformation of raw images into sequences of amino acids) and
+posterior biostatistics and epidemiological analysis.
+
+The data analysis is implemented as a pipeline of generic and
+customized components, mostly third-party public software tools. The
+pipeline may differ for each specific study, depending on the data
+acquisition and the research question. Typically it includes data
+preparation (file conversion, reformatting and filtering); alignment
+(comparison) with some reference database as the human genome;
+post-processing (output conversion, reformatting, statistical
+analysis) and visualization. Sequence alignment is the most
+computing-intensive step, which became prohibitive for sequential
+execution in next generation sequence experiments. All data is stored
+in files that are often completely read in memory for fast
+processing. Compressed formats are usually adopted. Although the files
+are large, the data can be easily decomposed because the processing is
+done on each sequence (`string') at a time.
+
+3. What infrastructure is used? (including compute, data, network,
+instruments, etc.)
 
@@
 
-Grid storage: Storage Elements of the Dutch e-Science Grid (www.biggrid.nl)
+Grid storage: Storage Elements of the Dutch e-Science Grid
+(www.biggrid.nl)
 
@@
 
-Between data server and grid storage: public network (XXX);  lightpath (near future)
-
-Between grid storage and worker nodes: public (XXX) or site (XXX) network
+Between data server and grid storage: public network (XXX); lightpath
+(near future)
+
+Between grid storage and worker nodes: public (XXX) or site (XXX)
+network
 
@@
 
-Local Computing: clusters and servers at the AMC and workstations (laptops, desktops) anywhere
-Grid Computing: Dutch Grid (www.biggrid.nl), part of EGI
-
-Originally all steps were performed in some local server or on the user's workstation.
-The visualization is still performed in the user's (windows) workstation, but all the other steps are currently performed on the Dutch grid infrastructure.
-The pipeline is described as a grid workflow using the GWENDIA language (\url{http://dx.doi.org/10.1145/1645164.1645171}) and enacted on the Dutch Grid using the MOTEUR2 workflow engine (\url{http://modalis.polytech.unice.fr/softwares/moteur/start}).
-
-PS: although only the alignment step is computation-intensive, it is more practical to perform all the others on the grid to minimize data transport to/from the user's workstation, which is normally on low speed networks.
-That is, the computation is statically and manually taken to the data in this case.
-This can lead to sub-optimal resource usage.
+Local Computing: clusters and servers at the AMC and workstations
+(laptops, desktops) anywhere Grid Computing: Dutch Grid
+(www.biggrid.nl), part of EGI
+
+Originally all steps were performed in some local server or on the
+user's workstation. The visualization is still performed in the user's
+(windows) workstation, but all the other steps are currently performed
+on the Dutch grid infrastructure. The pipeline is described as a grid
+workflow using the GWENDIA language
+(\url{http://dx.doi.org/10.1145/1645164.1645171}) and enacted on the
+Dutch Grid using the MOTEUR2 workflow engine
+(\url{http://modalis.polytech.unice.fr/softwares/moteur/start}).
+
+PS: although only the alignment step is computation-intensive, it is
+more practical to perform all the others on the grid to minimize data
+transport to/from the user's workstation, which is normally on low
+speed networks. That is, the computation is statically and manually
+taken to the data in this case. This can lead to sub-optimal resource
+usage.
 
@@
 
-Note: data is not really `dynamic' in this case -- the sequences are acquired once and analyzed many times.
-However the execution of this application on a distributed infrastracuture has dynamic aspects when optimized resource usage is considered:
+Note: data is not really `dynamic' in this case -- the sequences are
+acquired once and analyzed many times. However the execution of this
+application on a distributed infrastracuture has dynamic aspects when
+optimized resource usage is considered:
 
@@
 
-All data is stored in files that are often completely read in memory for fast processing.
-Compressed formats are usually adopted.
-Although the files are large, the data can be easily decomposed because the processing is done on each sequence (`string') at a time
+All data is stored in files that are often completely read in memory
+for fast processing. Compressed formats are usually adopted. Although
+the files are large, the data can be easily decomposed because the
+processing is done on each sequence (`string') at a time
 
@@
 
-In research there are no strict constraints on the time to complete the analysis.
-In patient care, time could determine life, death or money, but we have no concrete case yet.
-
-Quality in this case refers to the accuracy of the results obtained. This is essential for research and patient care.
+In research there are no strict constraints on the time to complete
+the analysis.  In patient care, time could determine life, death or
+money, but we have no concrete case yet.
+
+Quality in this case refers to the accuracy of the results
+obtained. This is essential for research and patient care.
 
@@
 
-\subsubsection{New answers from Steve}
-
-3DPAS - Particle Physics - the example of Atlas physics channel selection.
+\subsubsection{Answers from Steve}
+
+3DPAS - Particle Physics - the example of Atlas physics channel
+selection.
 
@@
 
-A physics group is interested in a specific physics channel (underlying physical process) as recorded by the Atlas experiment at the LHC (Large Hadron Collider) at CERN in Geneva.
-The set of available data grows steadily over time as more data are collected and reconstructed.
+A physics group is interested in a specific physics channel
+(underlying physical process) as recorded by the Atlas experiment at
+the LHC (Large Hadron Collider) at CERN in Geneva. The set of
+available data grows steadily over time as more data are collected and
+reconstructed.
 
@@
 
- Both physics groups and individuals run jobs as they see fit, either reading the reconstructed data or reading intermediate datasets created by themselves or their colleagues.
-If, in the future, resources should prove inadequate then adjustments to working practices to coordinate processing may be made to increase physics output.
-
-   3.      What infrastructure is used? (including compute, data, network, instruments, etc.)
-
-The WLCG (Worldwide LHC (Large Hadron Collider) Computing Grid) is used with a mixture of gLite software, experiment specific software and physics group specific software.
-WLCG has 250,000 cores distributed over 140 sites and with 100PB of disk.
-CERN is connected to the eleven `Tier 1' national sites by a dedicated optical network.
-This may be extended in the future.
-
-   4.      What dynamic data is used in the application?
-
-The data being processed can be considered as dynamic as it grows steadily as new raw data are collected and the basic reconstruction program is run to produce information about specific particles involved in a collision.
-This reconstruction is rerun periodically (two or three times per year is expected) as calibrations of the detector are improved.
-Reconstruction algorithms are also improved as are the physics groups' selection codes.
-One output of the reconstruction is `tag' data - summary information written to a database to facilitate event selection.
-This is not yet used very much.
-Calibration data are held on an RDBMS with Frontier/Squid front ends.
-
-         a.             What are the types of data,
-
-          The bulk of the data is individual event data at different stages of refinement.
-This is held in a in-house representation of C++ serialized objects.
-Though this format is not standard it is used by all LHC experiments.
-
-         b.            What is the size of the data set(s)?
-
-          Roughly 20~TB of data are collected every day.
-Multiple generations of processed data are kept.
-Many data files are replicated on multiple sites; this is done both by copying it to where it is likely to be useful and partly dynamically as the need arises.
-
-   5.      How does the application get the data?
-
-A copy of the file is made to the SE (Storage Element) at the site where the job will run. From there it is either copied to a local disk or read from the SE.
-
-   6.      What are the time (or quality) constraints on the application?
-
-If some data are missed during processing the only effect is a reduction of statistical precision in the final analysis.
-Accurate book keeping is important to know which data have been processed.
-If the selection code has been improved then naturally the physics groups would like to see the results as soon as possible.
-
-%\subsubsection{Old answers from Steve}
-%
-%   1.       What is the purpose of the application?
-%
-%A physics group is interested in a specific physics channel (underlying physical process) as recorded by the Atlas experiment at the LHC (Large Hadron Collider) at CERN in Geneva.
-The set of available data grows steadily over time as more data are collected and reconstructed.
-It would not be efficient to allow individual physics groups to process this large volume of data in an uncoordinated manner as it requires reading all available data.
-%
-%   2.       How is the application used to do this?
-%
-%Physics groups register their analysis code to be run on the next scheduled analysis ``train''.
-The train leaves every two weeks and reads through all the data, runs the physics selection code of each group and writes out the selected events and output from the computation to produce relatively small data sets for each physics group.
-The groups then use these selections to perform more ad-hoc analysis.
-%
-%   3.      What infrastructure is used? (including compute, data, network, instruments, etc.)
-%
-%The WLCG (Worldwide LHC (Large Hadron Collider) Computing Grid) is used with a mixture of gLite software, experiment specific software and physics group specific software.
-WLCG has 250,000 cores distributed over 140 sites and with 100~PB of disk.
-There is no wide area dedicated network infrastructure.
-%
-%   4.      What dynamic data is used in the application?
-%
-%The data being processed can be considered as dynamic as it grows steadily as new raw data are collected and the basic reconstruction program is run to produce information about specific particles involved in a collision.
-This reconstruction is rerun periodically (two or three times per year is expected) as calibrations of the detector are improved.
-Reconstruction algorithms are also improved as are the physics groups' selection code.
-Calibration data are held on an RDBMS with Frontier/Squid front ends.
-%
-%         a           What are the types of data,
-%
-%          The bulk of the data is individual event data at different stages of refinement. The is held in a non-standard representation of C++ serialized objects.
-%
-%         b.             What is the size of the data set(s)?
-%
-%          Currently the reconstructed data (excluding copies) is about m TB and is growing at about  n TB per month.
-Many data files are replicated on multiple sites.
-This is done both by copying it to where it is likely to be useful and partly dynamically as the need arises.
-%
-%   5.       How does the application get the data?
-%
-%This is mostly by copying the file to where the job will run - though some remote file access is also used.
-%
-%   6.  What are the time (or quality) constraints on the application?
-%
-%If some data are missed during processing the only effect is a reduction of statistical precision in the final analysis.
-If the selection code has been improved then naturally the physics groups would like to see the results as soon as possible.
-%
-%
-\subsection{Astrophysics -- Virtual Observatory (Bob) \label{astroBob}}
-\begin{itemize}
-\item Analysis of data to generate event streams $\rightarrow$ lead to the
-  positioning of telescopes
-\end{itemize}
-
-\subsubsection{Answers from Bob}
-
-I'll assume that the ``application'' in question is the sum of the two
-activities, despite the fact that they are different in many respects,
-and I'll try and distinguish between them where appropriate.
-
-Let me know if you want clarification or further information - in some cases I wasn't quite sure what you were after.
-
-1. What is the purpose of the application?
-
-An increasingly important feature of astronomical research is the existence of
-systematic sky surveys covering the whole sky (or large fractions thereof) in
-many different wavebands. The principal outcome from these surveys are
-astronomical source catalogues, which contain values for attributes
-characterizing all the celestial sources detected in the sky survey dataset.
-These are now typically implemented in relational databases, with SQL interfaces
-implemented through web pages, and the principal goal of the Virtual Observatory
-(VO) is to standardize access to these, and other astronomical data resources,
-so that astronomers can readily perform multi-wavelength analyses, combining all
-the extant data on particular sources.
-
-      2. How is the application used to do this?
-
-The VO includes a registry that contains metadata describing all the data
-resources published using VO data access standards, enabling the discovery of
-datasets relevant to a particular analysis. Web service implementations of data
-access protocols enable users to access data in these repositories
-programmatically, and distributed scratch space is provided for the storage of
-intermediate result sets, so the user does not need to route large data flows
-through his/her own workstation. In the future, more and more data analysis
-software will be made available through web services compliant with VO
-standards, enabling more of the data integration and analysis process to be
-combined in workflow.
+ Both physics groups and individuals run jobs as they see fit, either
+reading the reconstructed data or reading intermediate datasets
+created by themselves or their colleagues. If, in the future,
+resources should prove inadequate then adjustments to working
+practices to coordinate processing may be made to increase physics
+output.
 
@@
 
-The data resources published to the VO exist in a number of data centers
-distributed internationally. Major datasets---e.g. those from large sky
-surveys---are typically implemented in relational databases running on high-spec servers
-or clusters thereof. The compute requirements of the VO are currently modest,
-but are likely to increase over time. Astronomers typically use normal academic
-networks, but more specialized networking is required in some cases: e.g., sky
-survey data is moved from Cambridge to Edinburgh using the UKLight project's
-dedicated optical network, and radio interferometers require high-bandwidth
-fibre connections from each antenna to the correlator.
+The WLCG (Worldwide LHC (Large Hadron Collider) Computing Grid) is
+used with a mixture of gLite software, experiment specific software
+and physics group specific software. WLCG has 250,000 cores
+distributed over 140 sites and with 100PB of disk. CERN is connected
+to the eleven `Tier 1' national sites by a dedicated optical
+network. This may be extended in the future.
 
    4.  What dynamic data is used in the application?
-(a) What are the types of data, (b) What is the size of the data set(s)?
-
-The VO currently includes little dynamic data. The one exception is VOEvent
-messages, which are notifications of transient events which can be used to
-trigger follow-up observations on other telescopes. These are small XML messages
-containing basic information about the position of a transient source and a
-small amount of other metadata which the operators of a given telescope can use
-to define triggers for follow-up observations of interest to them or their
+
+The data being processed can be considered as dynamic as it grows
+steadily as new raw data are collected and the basic reconstruction
+program is run to produce information about specific particles
+involved in a collision. This reconstruction is rerun periodically
+(two or three times per year is expected) as calibrations of the
+detector are improved. Reconstruction algorithms are also improved as
+are the physics groups' selection codes. One output of the
+reconstruction is `tag' data - summary information written to a
+database to facilitate event selection. This is not yet used very
+much. Calibration data are held on an RDBMS with Frontier/Squid front
+ends.
+
+         a.  What are the types of data,
+
+          The bulk of the data is individual event data at different
+stages of refinement. This is held in a in-house representation of C++
+serialized objects. Though this format is not standard it is used by
+all LHC experiments.
+
+         b.  What is the size of the data set(s)?
+
+          Roughly 20~TB of data are collected every day. Multiple
+generations of processed data are kept. Many data files are replicated
+on multiple sites; this is done both by copying it to where it is
+likely to be useful and partly dynamically as the need arises.
+
+   5.  How does the application get the data?
+
+A copy of the file is made to the SE (Storage Element) at the site
+where the job will run. From there it is either copied to a local disk
+or read from the SE.
+
+   6.  What are the time (or quality) constraints on the application?
+
+If some data are missed during processing the only effect is a
+reduction of statistical precision in the final analysis. Accurate
+book keeping is important to know which data have been processed. If
+the selection code has been improved then naturally the physics groups
+would like to see the results as soon as possible.
+
+\subsection{Astrophysics -- Virtual Observatory
+(Bob) \label{astroBob}}
+\begin{itemize}
+\item Analysis of data to generate event streams $\rightarrow$ lead to
+the positioning of telescopes
+\end{itemize}
+
+\subsubsection{Answers from Bob}
+
+I assume that the ``application'' in question is the sum of the two
+activities, despite the fact that they are different in many respects,
+and I try and distinguish between them where appropriate.
+
+1. What is the purpose of the application?
+
+An increasingly important feature of astronomical research is the
+existence of systematic sky surveys covering the whole sky (or large
+fractions thereof) in many different wavebands. The principal outcome
+from these surveys are astronomical source catalogues, which contain
+values for attributes characterizing all the celestial sources
+detected in the sky survey dataset.  These are now typically
+implemented in relational databases, with SQL interfaces implemented
+through web pages, and the principal goal of the Virtual Observatory
+(VO) is to standardize access to these, and other astronomical data
+resources, so that astronomers can readily perform multi-wavelength
+analyses, combining all the extant data on particular sources.
+
+      2. How is the application used to do this?
+
+The VO includes a registry that contains metadata describing all the
+data resources published using VO data access standards, enabling the
+discovery of datasets relevant to a particular analysis. Web service
+implementations of data access protocols enable users to access data
+in these repositories programmatically, and distributed scratch space
+is provided for the storage of intermediate result sets, so the user
+does not need to route large data flows through his/her own
+workstation. In the future, more and more data analysis software will
+be made available through web services compliant with VO standards,
+enabling more of the data integration and analysis process to be
+combined in workflow.
+
+      3. What infrastructure is used? (including compute, data,
+network, instruments, etc.)
+
+The data resources published to the VO exist in a number of data
+centers distributed internationally. Major datasets---e.g. those from
+large sky surveys---are typically implemented in relational databases
+running on high-spec servers or clusters thereof. The compute
+requirements of the VO are currently modest, but are likely to
+increase over time. Astronomers typically use normal academic
+networks, but more specialized networking is required in some cases:
+e.g., sky survey data is moved from Cambridge to Edinburgh using the
+UKLight project's dedicated optical network, and radio interferometers
+require high-bandwidth fibre connections from each antenna to the
+correlator.
+
+      4. What dynamic data is used in the application?  (a) What are
+the types of data, (b) What is the size of the data set(s)?
+
+The VO currently includes little dynamic data. The one exception is
+VOEvent messages, which are notifications of transient events which
+can be used to trigger follow-up observations on other
+telescopes. These are small XML messages containing basic information
+about the position of a transient source and a small amount of other
+metadata which the operators of a given telescope can use to define
+triggers for follow-up observations of interest to them or their
 community.
@@
 
-Assuming this refers solely to the dynamic data, then a number of telescopes
-undertaking real-time detection of transient events emit VOEvent messages
-through feeds to which other parties can subscribe.
+Assuming this refers solely to the dynamic data, then a number of
+telescopes undertaking real-time detection of transient events emit
+VOEvent messages through feeds to which other parties can subscribe.
 
@@
 
-The time constraints differ for different types of transient source, but in some
-cases notification must be made within a few minutes if it is to be useful; the
-overheads associated with setting up a follow-up observation mean that much
-shorter time scales are practically irrelevant, although they are still
-scientifically interesting. The quality constraints centre on providing a
-telescope with good enough information about a potential transient source that
-it can make a reliable decision about whether to interrupt its existing
-observing program to follow-up the event. In most cases, the candidate
-transient will be some sort of noise event, so fairly sophisticated filtering is
-required to bring the false positive rate down to something acceptable, given
-the cost of observing time on cutting edge telescopes.
+The time constraints differ for different types of transient source,
+but in some cases notification must be made within a few minutes if it
+is to be useful; the overheads associated with setting up a follow-up
+observation mean that much shorter time scales are practically
+irrelevant, although they are still scientifically interesting. The
+quality constraints centre on providing a telescope with good enough
+information about a potential transient source that it can make a
+reliable decision about whether to interrupt its existing observing
+program to follow-up the event. In most cases, the candidate transient
+will be some sort of noise event, so fairly sophisticated filtering is
+required to bring the false positive rate down to something
+acceptable, given the cost of observing time on cutting edge
+telescopes.
 
@@
 
-At present most applications are quite small-scale. A canonical example
-is the monitoring of seal and other sea mammal populations by the
-Scottish Oceans Institute (SOI), which works by tagging animals with sensor
-packages that can analyze dive behavior, speed, movement and the like,
-and report back to base when the animal comes within range of a cellular
-network.
-
-3. What infrastructure is used? (including compute, data, network, instruments, etc.)
+At present most applications are quite small-scale. A canonical
+example is the monitoring of seal and other sea mammal populations by
+the Scottish Oceans Institute (SOI), which works by tagging animals
+with sensor packages that can analyze dive behavior, speed, movement
+and the like, and report back to base when the animal comes within
+range of a cellular network.
+
+3. What infrastructure is used? (including compute, data, network,
+instruments, etc.)
 
@@
 
-Data are collected remotely, then brought to a central site for analysis.
+Data are collected remotely, then brought to a central site for
+analysis.
 
@@
 In the sea mammal case the data isn't all that dynamic, being mainly a
-time series collected with only basic adaptation to the resolution etc.
-Larger-scale applications in environmental sensing in particular would
-make use of more structured and extensive adaptation (e.g.,
-changing the sampling frequency and other management characteristics in
-response to the data being observed).
-
-(a) What are the types of data, (b) What is the size of the data set(s)?
-
-For sea mammals: positions, motion vectors. For environmental missions:
-concentrations and gradients. I'd have to enquire about dataset sizes,
-but not excessive.
+time series collected with only basic adaptation to the resolution
+etc.  Larger-scale applications in environmental sensing in particular
+would make use of more structured and extensive adaptation (e.g.,
+changing the sampling frequency and other management characteristics
+in response to the data being observed).
+
+(a) What are the types of data, (b) What is the size of the data
+set(s)?
+
+For sea mammals: positions, motion vectors. For environmental
+missions: concentrations and gradients. I'd have to enquire about
+dataset sizes, but not excessive.
 
@@
 For the sea mammals the animal crawls up a beach, the sensor packages
-sees the cellular network and transmits. For environmental missions
-we're still wondering about that ourselves :-)
+sees the cellular network and transmits. For environmental missions,
+the answer is unclear.
 
@@
 
-Not sure I see what you mean here. Generally it's just a matter of
+Generally it's just a matter of
 capturing at whatever resolution device the sensor pack can support.
 
-I could probably get more complete -- perhaps *way* too complete :-) --
-answers from Mike Fedak in Scottish Oceans if that'd help.
-
-%\subsection{Bioinformatics Applications (Tom Slezak) - do not publish \label{bioTom}}
-%
-%1. What is the science/engineering problem? What is the purpose of the application?
-%
-%Comparing metagenomic (DNA sequence) data sets against each other.
-This could be to distinguish human disease biomarkers among different individuals, determine different soil (or water or air or ...) microbial composition from different samples, etc.
-%
-%2. How will the application solve the problem?
-%
-%In general this is a large (for biology -- don't think in terms of exascale!) data problem with large (all-vs-all comparison) computing.
-There are strategies to make this more scalable that will reduce the computing volume.
-%
-%(a) Which of the following describes the application
-%
-%         i) large (exascale) data and large (exascale) computing (e.g., data mining, analytics, bioinformatics)
-%
-%         ii) large (exascale) data and modest computing (e.g., search)
-%
-%         iii) modest data and large (exascale) computing (e.g., simulation/modeling)
-%
-%(b) Can you quantify the amount of computing vs the amount of data?
-%
-%Potentially peta-bytes or exa-bytes of data.
-Not sure how to quantify the computing.
-(Huge memories will be potentially more valuable than huge numbers of cores -- although this is heresy to the physics crowd.)
-%
-%3. What infrastructure will be used? (including compute, data, network, instruments, etc.)
-%
-%(a) What will be distributed?  data, compute, both?
-%
-%Individual data sets will be terabyte-scale.
-There could be many thousands of such sets to compare.
-Having them all local may not be feasible (the transfer time from where they are generated by others may be prohibitive.
-Computing might then have to be distributed.
-Biology is wrestling with this now.
-%
-%4. What data be used in the application?
-%
-%(a) What are the types of data
-%
-%Genomic sequence, billions of short (35-100?) bases of DNA; unassembled
-%
-%(b) What is the size of the data set(s)?
-%
-%Terabyte scale
-%
-%(c) Will any data be dynamic?
-%
-%Not for this problem
-%
-%5. How will the application get the data?
-%
-%That is a major part of the problem! Data will (in general) be generated from other places around the world. Getting to the data will be an issue.
-%
-%6. Are there time (or quality) constraints on the application?  (If so, what are they?)
-%
-%You'd like answers in under a day, generally.
-%
-%7. Please feel free to also talk about the current state of the application, if it exists today, and any specific gaps that you know will need to be overcome.
-%
-%Biology is wrestling now with  about 3 orders of magnitude increase in DNA sequence data generation capability in the past 2 years.
-All algorithms, architectures, storage, and networks have broken or will break soon.
-Some biology problems (not necessarily the one I describe here) are not good fits for core-rich and memory-poor modern architectures.
-For some problems (comparing k-mers across metagenomic data sets, for instance) you need terabyte-scale memory, not thousands of cores.
-%
-
 \subsection{Climate (Dan) \label{climateDan}}
 
-\subsubsection{CDAT} (from Dan)
-
-The Climate Data Analysis Tools (CDAT)~\cite{CDAT} are a scriptable
-(Python) set of packages and tools that are currently heavily used for
-analysis of climate data.  CDAT has been developed by LLNL's Program
-for Climate Model Diagnosis and Intercomparison (PCMDI), and is often
-used with data from the Earth System Grid (ESG)~\cite{ESG}.  The most
-common current usage of these tools with the data is that a user
-downloads a set of data from ESG to a local resource, then uses the
-CDAT to analyze the data.  However, the PCMDI group is now planning to
-support a new usage mode, where data transfer can be scripted.
-Initial development of this capability is planned for February 2011,
-using the Globus online data transfer service~\cite{globusOnline} as a
-scriptable component.
-
-\subsubsection{ESGF} (from Don and Dan)
-
-Aimed at supporting international CMIP/IPCC intercomparison activity.
-Three stages:
-
-{\bf Data Acquisition (Modeling)}
-
-\begin{itemize}
-\item Climate centers run prescribed set of common experiments; produce 2-10 PB of data
-\begin{itemize}
-\item data can also come from sensors - here, the data would still be post-processed before it would be published.
-\end{itemize}
-\item Centers can publish their own output data or send it to another center to publish data generated over \~{}2 year period (then post-processed and published over a few more months)
-\item Properties: data-intensive, distributed centers, lots of computing
-\end{itemize}
-
-{\bf Data storage (Infrastructure/system)}
-
-\begin{itemize}
-\item ESGF develops and deploys a federated network of gateways and associated data nodes
-\item As models run at each center, output data is post-processed into packages with common formats and prescribed metadata conventions
-\item Most centers will deploy data node software stack, and using this, they manage the data from their experiments
-  \begin{itemize}
-  \item the data node software stack scans the data, makes sure it has the right metadata fields, does QA/QC, build a set of catalogs of the prepared data
-  \begin{itemize}
-                \item minimally, catalog provides HTTP links to data elements
-                \item also can provide GridFTP endpoints
-                \item also can provide product services that abstract the dataset in other ways - get whole file, subset, browse, etc.
-  \end{itemize}
-  \end{itemize}
-    \item When the center is happy with the data/catalog in the data node, they publish it to a host/affiliated gateway
-\begin{itemize}
-         \item  this submits the catalog to the gateway
-          \item  them the gateway shares this catalog with other gateways so that all gateways have a consistent view of all the published data
-\end{itemize}
-    \item  Core archive (1-2 PB, most popular data)
-\begin{itemize}
-          \item  will be replicated at several sites
-         \item  replication activity manually requested/initiated by a gateway owner
-          \item  replication: copies catalog to replica site, do data transfer to new data node, then publish  catalog, push catalog to (this) gateway associated with new data node
-\end{itemize}
-    \item  Currently building notification service
-\begin{itemize}
-          \item  track which users have downloaded data - if a problem is found in a data set or if new data is added to a data set, notify users who have downloaded/replicated it
-\end{itemize}
-    \item  Properties: data-intensive, distributed gateways and data nodes, data appears in the system over time (dynamic)
-\end{itemize}
-
-{\bf Data processing (Application)}
-
-\begin{itemize}
-    \item  Users (\~{}20000 as of Dec. 2010) can:
-\begin{itemize}
-         \item  browse/search a catalog at any gateway and locate data, which might be hosted by a data node affiliated with another gateway. (can output a wget script that can later fetch the data)
-         \item  authenticate, gain access to a group
-        \item  download data via http or GridFTP or access product services (uses data retrieval syntax (DRS) - can be scripted)
-  \end{itemize}
-\end{itemize}
-
-\begin{itemize}
-    \item  abstractions that can be used by a programmer: data is stored in NetCDF format, uses CF metadata conventions, can be accessed using DRS
-    \item  Some users will analyze data from a single model
-    \item  Many applications are multi-model analyses - many users want to look at the same parts of the output of some/all of the models, to understand if and how the models differ
-    \item  Some centers will gather some/all of the core archive (plus more, perhaps) on local systems for local users to perform ``power analysis''
-\end{itemize}
-
-\subsection{Astro (LSST) (Adam Barker) \label{astroAdam}}
-
-\begin{enumerate}
-\item What is the science/engineering problem? What is the purpose of the application?
-
-Find and study variable objects and moving objects.
-
-\item How will the application solve the problem?
-
-Processing of data (data analysis and detection of new sources) from each image should be done while the next image is being taken by the telescope.
-This means that if anything unusual is detected, normal observation can be interrupted.
-Other observing resources can also be notified instantly, so they can observe the same event.
-As data are collected, they will be added to all the data previously detected from the same location of sky to create a very deep master image.
-LSST will also build up a database of all known moving objects.
-
-Every time a new image of the sky is obtained, the master image will be subtracted from it.
-The result is a subtracted image that contains the difference between the current sky image and its average state.
-
-This subtracted image is then processed by a cluster of computers with the following steps:
-\begin{enumerate}
-
-\item Query object catalog (knows orbits of all known objects) to find objects which are expected to appear in subtracted image, given the area of sky, time of day.
-
-\item Cross-match expected objects with sources in subtracted image. The result is the unmatched-source catalog, which contains sources that canÕt be matched with a previously known object
-
-\item Use unmatched-source catalogue to compute pairs of detections separated by short time intervals, called tracklets.
-A tracklet is an observable, short section of orbit.
-In order to create pairs of detections, all objects in the unmatched-source catalog are queried against the orbit catalog, in an attempt to obtain data about these objects from earlier observations.
-If earlier observations and current observations can be linked to form tracklets, they are stored in the tracklet catalog.
-
-\item Use tracklet catalog to attempt to link these tracklets over a larger time window.
-This is achieved by again querying the orbit catalog, this time looking for observations of the same objects even further back in time.
-If matches can be found and the tracklets can be extrapolated out into longer sections of orbit, then they are added as new orbits to the orbit catalog.
-
-\item Update orbit catalog with redetections of known objects. Each rediscovery provides information that improves the known orbits.
-
-\item Attempt to classify all entries in the orbit catalog (i.e., all the objects which now have known orbits).
-If any Near Earth Objects are detected to be passing close to the earth, alerts are generated to astronomers so that follow up observations can be scheduled.
-
-\item Compare with other observatories: If an unknown object is found that cannot be classified locally
-\begin{enumerate}
-\item issue call for participation to set of possible observatories
-\item each observatory accepts or rejects the call
-\item local observatory chooses from accepting observatories one to obtain data from
-\item that observatory takes data and sends it to local observatory
-\item local observatory decides what to do - classify the object, contact a human, call for more observations (automatically to other observatories, that decide whether or not to accept, and when)
-\end{enumerate}
-
-\end{enumerate}
-
-\begin{enumerate}
-\item Which of the following describes the application
-
-large (exascale) data and large (exascale) computing (e.g., data mining, analytics, bioinformatics)
-
-\item Can you quantify the amount of computing vs the amount of data?
-
-LSST will generate 36~GB of data every 30 seconds, and over a 10-hour winter night, will collect up to 30~TB.
-
-\end{enumerate}
-
-\item What infrastructure will be used? (including compute, data, network, instruments, etc.)
-
-A linked network of observatories and data and computing resources.
-
-(a) What will be distributed?  data, compute, both?
-
-both
-
-\item What data be used in the application?
-
-Fits images, and large databases for analysis of the discoveries.
-
- (a) What are the types of data
-
- (b) What is the size of the data set(s)?
-
- (c) Will any data be dynamic?
-
-\item How will the application get the data?
-
-Some will be analyzed at the mountain top, discoveries will be sent to other centers.
-
-\item Are there time (or quality) constraints on the application?  (If so, what are they?)
-
-Time component is critical. As real-time as possible.
-
-\item Please feel free to also talk about the current state of the application, if it exists today, and any specific gaps that you know will need to be overcome.
-
-\end{enumerate}
-
-also, the Palomar Transient Factory is the best example to date:
-
-See:
-
-\url{http://www.astro.caltech.edu/ptf/}
-
-\url{http://xxx.lanl.gov/abs/0906.5350}
-
-\url{http://xxx.lanl.gov/abs/0906.5355}
-
-\url{http://arxiv.org/abs/1011.2199}
-
-\subsection{Interactive exploration of environmental data (Jon) \label{envJon}}
-Synopsis of chapter we're writing for Malcolm's book.  Looking at needs of interactive applications as opposed to batch ones.
+{\bf CDAT}: The Climate Data Analysis Tools (CDAT)~\cite{CDAT} are a
+scriptable (Python) set of packages and tools that are currently
+heavily used for analysis of climate data.  CDAT has been developed by
+LLNL's Program for Climate Model Diagnosis and Intercomparison
+(PCMDI), and is often used with data from the Earth System Grid
+(ESG)~\cite{ESG}.  The most common current usage of these tools with
+the data is that a user downloads a set of data from ESG to a local
+resource, then uses the CDAT to analyze the data.  However, the PCMDI
+group is now planning to support a new usage mode, where data transfer
+can be scripted.  Initial development of this capability is planned
+for February 2011, using the Globus online data transfer
+service~\cite{globusOnline} as a scriptable component.
+
+\subsection{Oceanographic/BADC Applications \label{oceanStephen}}
+Climate from BADC + Oceanographic data (Data Nodes + Gateways)
+(Stephen Pascoe)
+\begin{itemize}
+\item Dynamic aspects not clear -- primarily providing a data storage
+(curated) system
+\item Multiple sources of data -- and potentially coupling between
+simulation and data
+\item Issue of placement and scheduling of the data -- through the use
+of federated registries
+\item Identify how this links it to the ``application genie''
+\end{itemize}
+
+\subsection{Interactive exploration of environmental data
+(Jon) \label{envJon}}
+
+(Here's a synopsis of the chapter I'm writing for Malcolm's book on Data-Intensive Research.
+I'm not sure how to merge this with the answers to the pro-forma questions below.  Also,
+many of these sentences are cut-and-pasted from the book chapter - future copyright issue?)
+
+In the environmental sciences the use of visualization techniques is vital for understanding
+the ever-increasing volume and diversity of data that is being produced by Earth observing
+systems and computer simulations.  The primary purpose of visualization is to gain insight
+\cite{card_readings_1999}, but the majority of current tools in wide use are not adequate
+for achieving this.  The time spent to create visualizations is too great, usually due to the
+crippling overhead of low-level technical tasks such as file-format and metadata interpretation.
+Most scientific tools generate static plots, which are shared in documents or emails, but 
+neither the originating scientist nor the recipient can easily customize or adapt the visualization
+to reveal new information.  The context of how the visualization was generated is usually lost
+in the publication/sharing process, making interpretation and reproduction much more difficult.
+
+Visualization of a single dataset is therefore more tedious and less useful than it should be.
+These problems are compounded if the scientist wishes to examine multiple datasets simultaneously.
+This is an extremely common task in environmental science, which forms an important part of model validation,
+``ground truthing'', quality control and data assimilation.
+
+There is much current interest in developing new, interactive modes of environmental data exploration and visualization, which can
+play a key role at all stages of scientific investigation.   We consider that, for the case of interactive
+data exploration, concerns of performance,
+intuitiveness and user engagement tend to outweigh the desire for sophistication in data processing. 
+We therefore argue that a new generation of "4D GIS" tools, based on the principles of simplicity,
+open standards and user-friendliness is required to gain greater insight from the environmental data
+deluge.  These new tools will not replace existing approaches, but rather complement them.
+
+Increasingly, environmental data are distributed in multiple locations, and are frequently updated
+(often several times a day).  Interactive exploration of such large, dynamic, distributed data is therefore
+a strong technical challenge.  Data infrastructures must support the ability to quickly extract
+subsets of large data, and any data processing must happen quickly, with the minimum of overhead.
+Web services are therefore playing an increasingly-important role, and there is much interest in
+cloud technologies to provide the necessary scalability at the back end.  Grid technologies, being essentially batch-mode
+systems, are less suitable for this kind of application as the scheduling overhead tends to dominate 
+the processing time for simple jobs, seriously hampering interactivity.
+
+Performance is therefore a key challenge, exacerbated by the dynamic nature of the data, which makes
+it hard to achieve scalability through caching (see section~\ref{sec:dynamicdataweb}).  [Note: could probably
+merge material from that section here.]  Security is another serious concern: many environmental datasets
+are held under access control.  In a distributed environment, it is common for machines to access data
+on behalf of end users (e.g.\ a processing service might download its input data from a remote store).
+Finding a means for the user to delegate his/her authority to a multi-web-service infrastructure is
+a key current challenge.  The MashMyData project (http://www.mashmydata.org) is investigating two solutions, based
+respectively on Grid proxy certificates and OAuth.
 
@@
 
-We are actually concerned with a class of applications (highly interactive, graphical ones) rather than a specific one.
-But generally, the applications involve the intercomparison of different environmental datasets for purposes such as model validation, ground trothing of observations and data assimilation.
+We are actually concerned with a class of applications (highly
+interactive, graphical ones) rather than a specific one.  But
+generally, the applications involve the intercomparison of different
+environmental datasets for purposes such as model validation, ground
+trothing of observations and data assimilation.
 
@@
 
-The applications often take the form of graphical tools that share a lot in common with Geographic Information Systems (GIS).
-However these applications are much better suited to scientific data, which are large and multidimensional.
-The general ÒvisionÓ is of a map-based interface, onto which different datasets can be overlain.
-A processing step may be initiated by rubber-banding an area of the map and selecting from a list of algorithms that process data within the selected area, perhaps calculating statistics of the data.
-
-3.     What infrastructure is used? (including compute, data, network, in- struments, etc.)
-
-Usually the applications are driven by data served through web services of various kinds.
-These web services may be ÒfedÓ by data coming from instruments or from computer simulations.
-A key challenge is how to integrate compute services with these data services in an efficient, easy-to-use and transparent way, allowing fast data processing.
-The processing algorithms are generally quite simple and so it is frustrating to succumb to the high overhead of some scheduling systems.
-The infrastructure probably has more in common with the Web (or perhaps cloud) than the Grid.
-
-     4. What dynamic data is used in the application?
-(a) What are the types of data, (b) What is the size of the data set(s)?
-
-Dynamic data are only sometimes used, but may arise from Òreal-timeÓ feeds from instruments or from looking at ÒliveÓ results from a model.
-Data from instruments are ÒsmallÓ 1MB-100MB but model output may be very big (GB-TB).
+The applications often take the form of graphical tools that share a
+lot in common with Geographic Information Systems (GIS). However these
+applications are much better suited to scientific data, which are
+large and multidimensional.  The general “vision” is of a map-based
+interface, onto which different datasets can be overlain.  A
+processing step may be initiated by rubber-banding an area of the map
+and selecting from a list of algorithms that process data within the
+selected area, perhaps calculating statistics of the data.
+
+3.  What infrastructure is used? (including compute, data, network,
+in- struments, etc.)
+
+Usually the applications are driven by data served through web
+services of various kinds.  These web services may be “fed” by data
+coming from instruments or from computer simulations.  A key challenge
+is how to integrate compute services with these data services in an
+efficient, easy-to-use and transparent way, allowing fast data
+processing.  The processing algorithms are generally quite simple and
+so it is frustrating to succumb to the high overhead of some
+scheduling systems.  The infrastructure probably has more in common
+with the Web (or perhaps cloud) than the Grid.
+
+     4. What dynamic data is used in the application?  (a) What are
+the types of data, (b) What is the size of the data set(s)?
+
+Dynamic data are only sometimes used, but may arise from “real-time”
+feeds from instruments or from looking at “live” results from a model.
+Data from instruments are “small” 1MB-100MB but model output may be
+very big (GB-TB).
 
@@
 
-Generally through web services.  These services are designed to allow server-side subsetting of large datasets, attempting to minimize the amount of unwanted data travelling across networks.
+Generally through web services.  These services are designed to allow
+server-side subsetting of large datasets, attempting to minimize the
+amount of unwanted data travelling across networks.
 
@@
 
-We try to aim for near-real time interaction with the data, i.e the user should not be waiting more than a few seconds for some kind of response to a request.
-We have largely solved this problem for simple visualization, but for linking data access and processing we are not there yet.
+We try to aim for near-real time interaction with the data, i.e the
+user should not be waiting more than a few seconds for some kind of
+response to a request.  We have largely solved this problem for simple
+visualization, but for linking data access and processing we are not
+there yet.
 
@@
 
-% \subsection{Astro (CMB) (Julian Borrill) - do not publish \label{astroJulian}}
-
-% 1. What is the science/engineering problem? What is the purpose of the application?
-
-% Cosmic Microwave Background (CMB) data simulation and analysis.
-The CMB is an image of the Universe as it was only 400,000 years after the Big Bang.
-Tiny fluctuations in the CMB temperature and polarization encode the fundamental parameters of cosmology and, using the Big Bang as the ultimate particle accelerator, ultra-high energy physics.
-Extracting this information from the data gathered by current and anticipated CMB observations is an extremely computationally intensive endeavour for which we have developed our massively parallel tools.
-
-% 2. How will the application solve the problem?
-
-% The central computing challenge for any CMB dataset is the simulation and analysis of O($10^4$) synthetic observations, used to correct for biases and quantify uncertainties in the analysis of the real data.
-We use preconditioned conjugate gradient techniques to solve for the maximum likelihood sky map given the input data (obtained by scanning the sky with hundreds to thousands of detectors for weeks to years) and its piecewise stationary noise statistics.
-To avoid the IO bottleneck inherent in the traditional simulate/write/read/map paradigm, all simulations are performed on the fly only when requested by the map-making code.
-
-% (a) Which of the following describes the application
-
-%      i) large (exascale) data and large (exascale) computing (e.g., data mining, analytics, bioinformatics)
-
-%      ii) large (exascale) data and modest computing (e.g., search)
-
-%      iii) modest data and large (exascale) computing (e.g., simulation/modeling)
-
-% (iii) O(1 - 10) TB input data and a similar volume output.
-
-% (b) Can you quantify the amount of computing vs the amount of data?
-
-% The computing scales with the number of observation samples, while the communication and IO scale with the number of pixels in the map.
-For the current Planck satellite experiment these are O($10^12$) and O($10^8$) respectively; for the proposed CMBpol satellite they will be O($10^15$) and O($10^8$) - that is, we expect a 1000x increase in calculation cost over the next 15 years (including intermediate suborbital pathfinder experiments) but fixed communication/IO costs.
-
-% 3. What infrastructure will be used? (including compute, data, network, instruments, etc.)
-
-% We are targeting the largest supercomputers available - Hopper, Tianhe, Blue Waters, etc.
-It's all about the cycles, although as the concurrency we work at increases previously solved IO and communication scaling bottlenecks typically re-emerge.
-
-% (a) What will be distributed?  data, compute, both?
-
-% Both.
-
-% 4. What data will be used in the application?
-
-% Real and simulated CMB data.
-
-% (a) What are the types of data
-
-% (b) What is the size of the data set(s)?
-
-% (c) Will any data be dynamic?
-
-% See above.
-
-% 5. How will the application get the data?
-
-% Real data are read from disk; simulated data are generated on the fly.
-
-% 6. Are there time (or quality) constraints on the application?  (If so, what are they?)
-
-% Typically we want to be able to simulate and map O($10^2$) realizations of an experiment's data in O(1) wallclock hour to provide 10\% errors during the early analysis stages, and O($10^4$) realizations in O(100) wallclock hours for the final definitive analysis.
-
-% 7. Please feel free to also talk about the current state of the application, if it exists today, and any specific gaps that you know will need to be overcome.
-
-% Moving to the current and future generations of heterogeneous many-core systems will require a significant re-working of our analysis codes, both to deal with communication over hundreds of thousands of cores and to take advantage of accelerator opportunities such as GPUs.
-
-% \subsection{Fusion (Scott Klasky) -- do not publish \label{fusionScott}}
-
-% 1. What is the science/engineering problem? What is the purpose of the
-% application?
-
-% With ITER scheduled to operate before 2020, plasma fusion has become
-% one of the highest priority research items for DOE. Scalability of
-% first-principles fusion codes is an important part of fusion research
-% needed to obtain predictive capability of plasma performance with
-% first-principles physics fidelity. We want to build a realistic
-% full-distribution kinetic code, in realistic ITER geometry, combining
-% the strengths of each of the existing gyrokinetic codes (GEM, GTS,
-% XGC, and GTC), and utilize the most up-to-date computer science,
-% hardware design and applied mathematics technologies developed. The
-% application that we will develop will be capable of 1) discovering
-% Macroscopic Magneto Hydro Dynamics activity in burning plasmas, 2)
-% Understand Energetic particle effects in fusion reactors, 3)
-% Understand the heating in a fusion reactor, and 4) Understand the
-% interaction of the plasma at the material boundary.
-
-% 2. How will the application solve the problem?
-
-% (a) Which of the following describes the application
-
-%         i) large (exascale) data and large (exascale) computing (e.g., data mining, analytics, bioinformatics)
-
-%         ii) large (exascale) data and modest computing (e.g., search)
-
-%         iii) modest data and large (exascale) computing (e.g., simulation/modeling)
-
-% The applications we currently run (XGC1, XGCP, GTC, GTS) run on all of the leadership class facilities, and require very large resources for the modest problems which we are currently solving.
-In the next generation (exascale) we will not only need all of the computing power that will be available at the time to understand ITER, we will also need to understand all of the data that will be generated during the simulation.
-Our techniques have been to examine and understand the data in situ, to reduce the total amount of data produced in the simulation.
-Today we have produced O(100 TB/week) for a simulation using our ADIOS middleware incorporated in our simulations.
-Using our staging techniques, we believe that we will only produce O(10 PB) on an exascale machine.
-(I don't know if this is considered modest dat, or extreme size data)?
-
-% (b) Can you quantify the amount of computing vs the amount of data?
-
-% 1 PB of data -- 10 PB of data, for an exascale size computation run in O(100 hours).
-% Today we achieve about 6\% of peak with excellent scalability to 220K
-% cores on the Cray XT systems. It is
-% difficult to predict how we will run on
-% accelerators, and next generation machines.
-
-% 3. What infrastructure will be used? (including compute, data,
-% network, instruments, etc.)
-
-% For Validation, we will need to compare our data from the computer to
-% ITER (or current fusion reactors). Typically the amount of data used
-% in comparison is tiny by comparison (people only look at MB's form
-% experiments).
-
-% (a) What will be distributed?  data, compute, both?
-
-% We always run on all available resources, so if we have computer time
-% across the country, then our data will be distributed, and so will our
-% computation. ITER will have data that will be distributed, but it
-% still unknown.
-
-% 4. What data be used in the application?
-
-% I don't understand this question?
-We will input data (from perhaps a restart), and this data is typically in the ADIOS-BP file format for large data.
-Smaller data sets are typically stored in HDF5, and the smallest are stored in MDS+.
-
-% (a) What are the types of data
-
-% (b) What is the size of the data set(s)?
-
-% Depends on what we want to understand --- For GTS/GTC we have typically
-% looked at the particle data to compute fluxes, etc. and this data can
-% be large, but due to the Monte Carlo error, and ways to generated 5D
-% distribution functions from the particles, we reduce the particles to
-% a manageable amount (300 GB per time step x 1 K time steps). Our
-% current I/O is capable of reading this in $<$ 100 s per time step, which
-% allows us to analyze the data in about 1-2 days.
-
-% (c) Will any data be dynamic?
-
-% Don't understand the question. Features change and are dynamic. The physics will get more complex as we continue to evolve our codes, and will be more dynamic in nature.
-
-% 5. How will the application get the data?
-
-% Reading the data from disk --- (Perhaps I don't understand the question).  It will also couple to other codes, and the coupling will be in-memory data exchange.
-
-% 6. Are there time (or quality) constraints on the application?  (If so, what are they?)
-
-% We always need to run as fast as possible since we have limited computing time (our team does have about 100M hours FY 11, but this still puts severe restrictions on what we run, so the faster we can run, the more science we can get done.
-
-% 7. Please feel free to also talk about the current state of the application, if it exists today, and any specific gaps that you know will need to be overcome.
-
-% Our biggest gaps will be in accelerating the code on next generation exascale platforms, and in processing as much data as possible in situ.
-
-\subsection{SOA Astronomy applications from Adam\label{astroSOAAdam}}
-
-1. What is the purpose of the application?
-
-Calculates the photometric redshift for a given area of sky [RA, DEC] using two tools: HyperZ and ANNz, both accessible via WS interfaces.
-
-2. How is the application used to do this?
-
-The application is a service-oriented workflow. Services are orchestrated through a pipeline in order to compute the redshift of a given area of sky.
-
-3. What infrastructure is used? (including compute, data, network, instruments, etc.)
-
-The Wide Field Survey Archive (WFS):  image catalogue
-Spectroscopic database: used to provide spectroscopic data for an area of sky
-
-SExtractor tool: image processing, extracts all objects of interest (stars, galaxies etc.)
-Cross matching tool: compiles one table of all objects of interest over the 5 wavebands
-
-Hyperz: photometric redshift estimation algorithm 1
-ANNz: photometric redshift estimation algorithm 2
-
-mySpace: AstroGrid storage service
-
-\katznote{checking more on this point}
-
-4. What dynamic data is used in the application?
-
-The WFS catalogue will be updated regularly, therefore a query at time x will not provide the same data as a query at time y.
-
-(a) What are the types of data, (b) What is the size of the data set(s)?
-
-Data sets (images) will be in the GB range, I can attempt to clarify these details.
-
-5. How does the application get the data?
-
-WS interfaces.
-
-6. What are the time (or quality) constraints on the application?
-
-Enough data needs to be available for a given area of sky from both the WFS archive and the spectroscopic archive.
-
-\subsection{TNO + Thales (Netherlands)  -- Omer (Kees)}
-
-will try to fill in...
-
 \subsection{(Dan's) summaries of contributions}
@@
 
-Steve's WLCG app (\S\ref{WLCGSteve}): there is a hierarchy of systems; data is centrally stored, and locally cached (and copied to where it likely will be used), perhaps at various levels of the hierarchy; processing is done by apps that are independent of each other; processing of one data file is independent of processing of another file, but groups of processing results are collected to obtain statistical outputs about the data.
-
-Bob's VO (astrophysics) app (\S\ref{astroBob}): data is stored in registries with metadata; data access is through web services; there is a workflow concept of access and multiple processing steps; distributed scratch space can be used for storage of intermediate results; there is also a pub/sub mechanism for dynamic events, which are small in terms of the data required to identify an event and in terms of their frequency
-
-Simon's Sensor Network app (\S\ref{sensorSimon}): data collected from sensors that periodically transmit; data brought to a central site; stored data analyzed using statistical techniques, visualized with tools like Google Earth.
-
-Bioinformatics (from Tom Slezak) (source not in this doc): want to compare O(1000s) of terabye-scale data sets (metagenomic) with each other.
-Datasets are distributed around the world.
-Solution to the problem is not clear.
-
-Climate (from Dan/Don) (\S\ref{climateDan}): Data (output of climate simulations) is distributed in multiple stores.
+Steve's WLCG app (\S\ref{WLCGSteve}): there is a hierarchy of systems;
+data is centrally stored, and locally cached (and copied to where it
+likely will be used), perhaps at various levels of the hierarchy;
+processing is done by apps that are independent of each other;
+processing of one data file is independent of processing of another
+file, but groups of processing results are collected to obtain
+statistical outputs about the data.
+
+Bob's VO (astrophysics) app (\S\ref{astroBob}): data is stored in
+registries with metadata; data access is through web services; there
+is a workflow concept of access and multiple processing steps;
+distributed scratch space can be used for storage of intermediate
+results; there is also a pub/sub mechanism for dynamic events, which
+are small in terms of the data required to identify an event and in
+terms of their frequency
+
+Simon's Sensor Network app (\S\ref{sensorSimon}): data collected from
+sensors that periodically transmit; data brought to a central site;
+stored data analyzed using statistical techniques, visualized with
+tools like Google Earth.
+
+Bioinformatics: want to compare O(1000s) of terabye-scale data sets (metagenomic) with each other.  Datasets are distributed around the world.  Solution to the problem is not clear.
+
+Climate (from Dan) (\S\ref{climateDan}): Data (output of climate simulations) is distributed in multiple stores.
 Current data analysis and visualization tools require the user to bring the data they want to analyze to a local system, then run the tools locally.
@@
 
-Astro (from Adam) (\S\ref{astroAdam}): Data taken by telescope - quick analysis done at telescope site for interesting (urgent) events (which may involve comparing new data with previous data).
-Can get more data from other observatories if needed, request other observatories to take more data, or call a human.
-Data then transferred to archive site, which may be the observatory.
-At archive site, data analyzed , reduced, classified (some of this may be farmed out to grid resources).
+Astro (LSST): Data taken by telescope - quick analysis done at telescope site for interesting (urgent) events (which may involve comparing new data with previous data).
+Data then transferred to archive site.
+At archive site, data analyzed, reduced, classified.
 Detailed analysis of new data vs.
@@
 
-Jon's Environmental Data exploration app (\S\ref{envJon}): drive apps from graphical tools, inputs are geographic areas and algorithms to run on the data in those areas; data can come through web services from sensors or simulations (possibly running); provide responses in near-real-time.
+Stephen Pascoe's Oceanographic app (\S\ref{oceanStephen}):
+\katznote{need summary here - but not sure what to write...}
+
+Jon's Environmental Data exploration app (\S\ref{envJon}): drive apps
+from graphical tools, inputs are geographic areas and algorithms to
+run on the data in those areas; data can come through web services
+from sensors or simulations (possibly running); provide responses in
+near-real-time.
 
@@
 
-Astro (from Julian) (\S\ref{astroJulian}): Application builds a sky map.  Reads data from detectors, and gets data from simulations performed on-the-fly.
+Astro (CMB): Application builds a sky map.  Reads data from detectors, and gets data from simulations performed on-the-fly.
 
@@
 
-Fusion (from Scott Klasky) (source not in this doc): multiple physics simulation codes run concurrently on a distributed set of parallel computers.
+Fusion: multiple physics simulation codes run concurrently on a distributed set of parallel computers.
 Data from some codes are streamed to other codes to link them into a single simulation.
@@
 
-SOA Astro (from Adam) (\S\ref{astroSOAAdam}): services are orchestrated through a pipeline, including a data retrieval service.
-Data moved through the pipeline, intermediate and final products can be stored in AstroGrid storage service.
-
 \subsection{Additional sources}
-
-\katznote{discuss this at April workshop}
-
-\begin{itemize}
-\item XLDB workshop series (characterizing aspects of dynamic data)
+\begin{itemize}
+\item XLDB workshop series (characterising aspects of dynamic data)
 \url{http://www-conf.slac.stanford.edu/xldb10/Program.asp}
@@
 \item Ocean Observatories Initiative (OOI, US-funded) -- Dan to talk to them
+\item Manish Parashar -- Shantenu or Omer
+\item Brian Lawrence -- Jon (Oceanographic Data)
 \item Roger Barga (Microsoft) -- Shantenu
@@
   supports provenance  tracking -- Bart Scheer
+\item TNO + Thales (Netherlands)  -- Omer (Kees)
 \item Astro Wise (Astronomical Wide-field Imaging System for Europe)
@@
 Vectors (Axes) for consideration for applications  (Shantenu, Jon, Omer, Neil, Simon, and Dan)
+
 \begin{itemize}
@@
 
-Infrastructure capabilities and requirements  (Shantenu, Jon, Omer, Neil, Simon, and Dan)
+Diagram suggestion (perhaps indicate what is currently available and what is required (desired): -- (Shantenu, Omer, Neil, Simon, and Dan)
+
+\section{Infrastructure to support 3DPAS Applications} {\bf Shantenu, Neil (co-lead), All}
+
+% Comments on structure
+%Differentiation between application requirements Show instances of
+%applications running on infrastructure, what does it do, what does
+%it provide - for each of the stages: data ingest, analytics,
+%post-processing/managing reuse / derivatives sharing Have a table of
+%applications, identify what the infrastructure provides at each
+%stage based on what people are actually using; then repeat for ideal
+%scenario, what would they like to have? Summary table covering all
+%applications Identify areas of ìmissingî infrastructure where
+%thereís a cluster of ideals not being fulfilled by whatís currently
+%available
+
+Infrastructure capabilities and requirements  (Shantenu, Jon, Omer,
+Neil, Simon, and Dan)
 \begin{itemize}
@@
 
-% Diagram suggestion (perhaps indicate what is currently available and what is required (desired): -- (Shantenu, Omer, Neil, Simon, and Dan)
-
-\begin{table}
-\begin{footnotesize}
-\begin{center}
-% \begin{tabular}{|p{0.25*\textwidth}|p{0.25*\textwidth}|p{0.25*\textwidth}|p{0.25*\textwidth}|}
-\begin{tabular}{|l|l|l|l|}
-  \hline Application  & Data Input & Processing and Analysis & Post-Processing  \\
-  \hline
-  \hline Medical imaging & - &  - &  - \\
-  \hline WLCG/ATLAS & - &  - &  - \\
-  \hline Virtual observatory & - &  - &  - \\
-  \hline Environmental (Marine) & - &  - &  - \\
-  \hline Bioinformatics (metagenomics) & - &  - &  - \\
-  \hline Climate (CDAT and ESGF) & - &  - &  - \\
-  \hline Astrophysics (LSST) & - &  - &  - \\
-  \hline Astro (Cosmic Microwave Background) & - &  - &  - \\
-  \hline Power Grids & - &  - &  - \\
-  \hline Fusion & - &  - &  - \\
-  \hline
-\end{tabular}
-\caption{Application Vectors\label{table:vectors}}
-%  \jhanote{Omer: Can you relate the last column ``support sequential
-%    and concurrent'' back to the text or explain why this field is
-%    being explored please?}}
-\end{center}
-\end{footnotesize}
-\end{table}
-
-\section{Cyberinfrastructure} {\bf Shantenu, Neil (co-lead), All}
-
-\subsection{CI to support Dynamic Data}
-
-\subsection{CI to support Distributed Data}
+We consider an application to consist of the following stages:
+
+\begin{itemize}
+\item Ingest
+\begin{itemize}
+\item Managing Streams -- identify whether streams are converted to
+files or handled directly.
+\item  Handling events: two issues could be considered here: (i)
+what constitutes an event; (ii) how such an event is subsequently
+used to influence data delivery or processing.
+\item Data delivery (distributed aspects) ñ across shared,
+public networks. Issues with the limited control available to
+support data delivery from point of capture to some analysis engine.
+\item  No common administrative domains -- requiring data to be
+channeled through multiple administrative domains with no central
+(single) control.
+\end{itemize}
+
+\item Analytics (filtering, reduction, analysis) + visualisation (Visual Analytics)
+
+\begin{itemize}
+\item Resource selection, replication, placement.
+\item  Autonomic support (failure management, access and latency issues,
+storage and staging, etc).
+\end{itemize}
+
+\item Preservation (Storage) + Curation
+
+\begin{itemize}
+\item Governance issues (quality, meta-data selection, access control etc)
+\end{itemize}
+
+\end{itemize}
+
+\subsection{Silvia's Biosciences app: NGS, medical imaging
+(Silvia)}
+
+\subsection{WLCG (focus on ATLAS) (Steve Fisher)}
+
+\subsection{Astrophysics -- Virtual Observatory (Bob)}
+
+\subsection{Sensor Network App (Simon)}
+
+\subsection{Climate (Dan)}
+
+\subsection{Metagenomics}
+
+\subsection{Oceanographic/BADC Applications}
+
+\subsection{Interactive exploration of environmental data (Jon)}
+
+\subsection{Power Grids (Shantenu)}
+
+\subsection{CMB}
+
+\subsection{Wiki Analytics (Yike)}
+
+\subsection{Common Themes \& Analysis}
+
+\subsubsection{Infrastructure to support Dynamic Data}
+
+\subsubsection{Infrastructure to support Distributed Data}
 
@@
 
-``This paper presents the design and the implementation of an
+`` This paper presents the design and the implementation of an
 architecture for the analysis of data streams in distributed
@@
 
-``... In some cases streams of data must be analyzed in real time to
+``.. In some cases streams of data must be analyzed in real time to
 provide information about trends, outlier values or regularities that
@@
 
-``... the architecture combines the parallel and distributed
+``..the architecture combines the parallel and distributed
 paradigms, the first to keep the pace with the rate of a
@@
 
-\section{Programming Systems} {\bf Simon, All}
-
-\subsection{Architecture}
-
-Top-level views of a system is typically as a processing pipeline from
-collection, through pre-processing, to storage and then
-post-processing and querying. (Collection includes both simulation and
-sensing.) Different applications place different emphasis on the
-different stages: some collect and store all the data they can, with
-only minor pre-processing to remove noise; others must perform
-enormous pre-processing to reduce the data volumes to manageable
-dimensions. Another point in this spectrum is a classifier system that
-uses a data stream to recognise higher-level events which it then
-propagates, so it's the events that are interesting, not the data.
-
-Question: where does each application lie in this space?
-
-Question: is it the case that each application can be placed at an
-appropriate point \textit{a priori}, or do characteristics change?
-
-Question: is there a problem with multiple applications running on the
-same fabric and interfering? -- for example a classifier system
-running alongside a collection system and generating storms of events?
-
-A lot of scientific processes are predicated on data being collected
-and then analysed off-line, which emphasises both query performance
-but also the collection and use of meta-data about the data set (for
-example the precision of sensing and the provenance effects of any
-pre-processing).
-
-Question: is there a need to retain meta-data across the lifetime of a
-system and across the complete architecture?
-
-\subsection{Paradigms}
-
-Often limited to pipeline \textit{versus} some stylised large-scale
-process such as map-reduce, within the large architectural
-pipeline. Often find scientific workflow systems like Taverna
-describing pipelines in terms of components, possibly with specialised
-high-performance constructions for some intensive
-components. Generally find this in the ``traditional'' space of codes
-running on workstations and servers; sensor networks themselves are
-immature in this respect and typically focus on robust data
-transmission without much in the way of in-network processing
-(although this is changing, slowly).
-
-Stream processing is a variant on pipelines where the stream is
-processed locally and in real time, typically with only fairly small
-operations allowed at each stage (for example calculating time
-derivatives of a time series).
-
-Question: what are the issues that face developers in using standard
-scientific workflow approaches?
-
-Question: would there be any need to push lots of processing to the
-front end, near the sensing? And if so, what would this do to the
-later stages?
-
-Question: are there any significant application codes that use
-anything other than pipelines, streams and map-reduce?
+\section{Programming System} {\bf Simon, All}
+
+\subsection{MapReduce Streaming}
+
+A recent work, MapReduce Online\cite{38}, extends the batch oriented
+MapReduce model to include a streaming model that allows pipelining
+between Map and Reduce tasks.  However, MapReduce by itself is not
+expressive enough compared to scientific workflows. In fact, some
+workflow systems have even included an optimized MapReduce pattern as
+a task available for workflow composers\cite{39}.
 
@@
 \section*{Acknowledgements}
-This paper is an outcome of the UK e-Science Institute sponsored
-Research Theme on Dynamic Distributed Data-Intensive Programming
-Abstractions and Systems
+This paper is an outcome of the UK e-Science Institute and NSF
+sponsored Research Theme on Dynamic Distributed Data-Intensive
+Programming Abstractions and Systems
 

### 2015-05-29 old-sections/old-future.tex
adding other, mostly old material

%the aim of the next subsection.

% What is quite noticeable is that the aspects of dynamism and
% distribution change even within a single application, and so can
% invalidate prior assumptions. \katznote{I don't think I understand the
%   point the previous sentence is trying to make.}

% The techniques developed are being applied ever more
% widely, both in terms of the application domains taking advantage of
% copious computing power and in terms of the volumes of data being
% processed.

% so any
% reusable infrastructure will need to support a diversity of approaches
% if it is to gain broad applicability.

% : aside from the widespread (but far from ubiquitous) use of
% workflows,
% \jhanote{Rather than Science and Trends, I propose ``D3 Science
%   Trends'' -- maybe 1 trend along each ``D''? Then we could do some
%   similar prognostication for infrastructure and programming systems,
%   which would reflect the structure of the document?}

% Data science seems to be positioned in the same manner as network
% science:\jhanote{will network science be obvious? network -- as in
%   physical/layer network?} \katznote{I also have no idea what `network
%   science' means here.} a field that draws upon several disciplines and
% impacts upon many more, without actually being part of any single
% recognized speciality. \jhanote{What about the mouthful term: CDS\&E
%   -- computational and data-enabled science and engineering?}

% Scientific codes tend to be long-lived, and it therefore makes sense
% to design them for maintenance, for evolution, and specifically to
% facilitate the introduction of dynamic aspects over their
% lifetime. Software engineers have developed a host of techniques by
% which this can be accomplished, but these techniques need broader
% visibility---or at least broader application---in scientific
% computing.

\subsubsection{D3 Applications: Development and Software:} 
%\katznote{what is the title of this subsection? see latex source for
%  confusion}\jhanote{fixed}

Dynamism and distribution are not unique to data science, and many of
the techniques developed in enterprise settings may usefully be
deployed.  Autonomic computing and communications can enormously
improve the utilization of physical infrastructure, and can contribute
to supporting dynamism. Autonomic principles can be applied at the
front end (in sensor networks) and the back end (in processing and
storage capabilities) of scientific infrastructure. \katznote{do we
  have examples of this being done we can refer to?}\jhanote{as
  mentioend, this is a candidate for en masse removal} While the
techniques developed for the latter are often not a perfect fit for
scientific applications, being transactional rather than
stream-oriented or stateful, there are certainly significant
opportunities both for deployment and for further research.  The
deployment of autonomic control is made easier if the application code
is structured as components, for example within a workflow, where
there are substantial opportunities for management interventions in
moving, replicating and otherwise adjusting the code ``in flight''.
\jhanote{The above paragraph is shaky and a candidate for removal if
  it cannot be linked with specific application examples or other
  sections of the paper}

% \jhanote{Above are fair points. We should expand this sub-section to
%   mirror Section 4 more closely}

% %\subsubsection{D3 Best Practices}
% \jhanote{I propose we move a discussion of autonomic computing out of
%   Infrastructure section and into best practises section, along with
%   S/W engineering etc.}

% \katznote{I wonder if this section is based on the evidence of this
%   paper, or more on what we think should be done.  I'm uncomfortable
%   with it, at least the first 2 paragraphs.  If nothing else, we
%   shouldn't call this Best Practices---rather, it is ``our''
%   opinions---but overall, I don't think it's the right way to go.}

% \jhanote{I echo Dan's comment that it is unclear whether this follows
%   from the paper, or if these are our (manufactured?) opinions?. There
%   is a need for citation to relevant sections/text of the paper.}

The growth in data science also suggests an elongation of the
lifetimes of both the code and data concerned, in the sense that data
is collected and analyzed with regard to a large-scale and often
long-running experimental program that may make its data widely
available. \katznote{Is the previous a general statement?  Or is it
  specific to something?  I don't think I understand what point is
  being made.}\jhanote{once again, this is entire paragraph is a
  candidate for en-masse removal, hence not responding to specific
  comments within the paragraph.} These developments change the game
for software development and maintenance. The use of proper software
engineering practices---architecture, design, reflection,
componentization, documentation, end-to-end engineering---become
vitally important in dealing with more dynamic scientific
domains. While these considerations can be avoided for single
applications, and often are seen as overheads to development and
experimentation, research in software engineering practice leaves no
room to doubt their effectiveness in constructing and (above all)
maintaining long-lived systems. It is important to remember the maxim
that a system that is not being changed is simply not being used, and
its corollary that a code will spent enormously more of its lifetime
being maintained and evolved than it spent being
developed. \katznote{does this previous maxim make sense here?  Even
  it so, do we need to state it as a maxim? Didn't we say this at the
  end of 6.3.1 already, anyhow?}  \jhanote{Is the maxim even correct?}
Maximizing our ability to manage this evolution makes good economic
and scientific sense.  \jhanote{Candidate to be removed entirely}

The components of scientific computations need themselves to be
managed, individually and collectively. Autonomic computing can
provide this, \emph{if} the relevant monitoring and control points are
available. The telecommunications industry has a long history of
constructing ``managed components'' that export independent functional
and management/monitoring interfaces. While this idea has not gained
enormous traction \katznote{does this mean almost no one uses it?}  in
much of the IT industry, it offers a way of integrating better, more
flexible management and adaptation to dynamism for scientific
applications, and frees the individual component developer from
worrying about global adaptation and decision-making issues. Dividing
the landscape in this way, separating scientific from management
functionalities, will help the development of re-usable
platforms. \jhanote{Again previous paragraph has to be linked to
  specific application and gaps.} \jhanote{Dan, more helpful than
  adding new comments, would be if you could help integrate as per
  existing comments. Else its already flagged as a deletion candidate
  and unlikely I will address new/different comments. Thanks!}

% \jhanote{propose we merge this paragraph with above
%   (industrial/enterprise developments) and the paragraph on autonomic
%   computing into a couple of sentences saying best practises abound in
%   specific contexts..}

% Scientific applications are not alone in having extensive data
% requirements: no matter to what extent this may have been true in the
% past, modern e-commerce generates data at similar rates and volumes as
% many scientific domains. This means that the techniques developed in
% industry to address these issues may be applicable to scientific
% codes, albeit with some modifications to suit different computational
% patterns, privacy constraints and other features.  Cloud computing and
% web services are both prime examples, and their use massively
% simplifies integration with other commodity services.

The rapid adoption of cloud computing in many industries perhaps masks
some of its limitations. Current cloud architectures are quite basic,
often either requiring very specific programming techniques or simply
providing low-level virtual machine management. The former is often
targeted at transactional, stateless computations that are a poor fit
to many scientific applications. The latter provide a certain degree
of elasticity but often provide interfaces without guidance as to
\emph{when} to change resources, or provide management consoles that
cannot be automated within an autonomic workflow. Nevertheless we can
confidently expect cloud computing to address these issues in time,
and further engagement of the scientific community with the providers
and researchers involved would help guide its
evolution. \katznote{again, do we have examples that we can point to
  in this section?}  \jhanote{We could have a bit of a discussion on
  how cloud computing (an increasingly important infrastructure issue)
  will influence D3 science?}

One technology that remains under-used \katznote{Again I would say
  that this is more opinion than evidence supported} is the semantic
web and the associated ideas of linked open data. While these might
seem like ideal technologies for scientific systems, they remain
marginalized. In many cases this may be due to unfamiliarity, the need
for quite sophisticated data handling, and the lack of decent
tooling. However scientific data retains many aspects of privacy,
sharing, and confidentiality that must be respected, and that can have
complex geographic, collaborative, organizational, and temporal
aspects. If these issues are handled in an integrated way---for
example by being specified as part of the overall experimental
workflow---then one might reasonably hope that linked open data would
be more widely adopted. \katznote{I again wonder if we are saying what
  we think should happen here}\jhanote{Yes, I agree. Either we relate
  this back to the applications or reduce in scope}

We also need to recognize the limitations on the reuse of scientific
codes. Aside from some common processing and visualization steps, much
code is too specific to be widely reused across domains (or even
within them). It is important not to overly prioritize the reuse of
components into unsuitable domains. \katznote{here I wonder if we
  should ask why there is so little reuse and suggest how there could
  be more.} \jhanote{also what aspects of the above are unique to
  3DPAS is important; candidate to go too}

### 2015-05-29 old-sections/pecs-whitepaper.tex
adding other, mostly old material

\documentclass[12pt]{article}

\usepackage{graphicx} 
\usepackage{amssymb}
\usepackage{epstopdf}
\usepackage{array}
\usepackage{fancyhdr}
\usepackage{url}
\usepackage{hyperref}
\usepackage{pdfsync}
\pagestyle{fancy}
\usepackage{mdwlist}
\fancyhead{}
\renewcommand{\headrulewidth}{0pt}
\usepackage{hyperref}

% \parskip=0.075in
% \topmargin=-1.25in
% \oddsidemargin=1.0in
% \evensidemargin=1.0in
% \textheight=9.0in
% \textwidth=6.5in
\usepackage{fullpage}
%\usepackage[top=tlength, bottom=blength, left=llength, right=rlength]{geometry}
%http://en.wikibooks.org/wiki/LaTeX/Page_Layout
%\usepackage[margin=1in, paperwidth=5.5in, paperheight=8.5in]{geometry}

\usepackage{fancyhdr}
\setlength{\headheight}{16.0pt}
\pagestyle{fancy}

%\pagestyle{fancyplain}

% \fancyhf{} \chead{\fancyplain{}{{\bf {\it Jha: Cyberinfrastructure
%         Challenges for Dynamic Data Scenarios in X-Scale CDES}}}}

\headheight = 0pt
\headsep = 25pt

\fancyhf{} \fancyhead[OC]{\bf {\it \footnotesize{Jha:
      Cyberinfrastructure Challenges for Dynamic-Data in Extreme-Scale
      CDES}}}

% in Extreme-Scale Computational and Data-Enabled Science}}}
%\rhead{\fancyplain{}{\footnotesize{Computational and Data-Enabled Science}}}
%\rfoot{\fancyplain{}{\thepage}}

\begin{document}

\newcommand{\squishlist}{
  \begin{list}{$\bullet$} { \setlength{\itemsep}{0pt}
      \setlength{\parsep}{3pt} \setlength{\topsep}{3pt}
      \setlength{\partopsep}{0pt} \setlength{\leftmargin}{1.5em}
      \setlength{\labelwidth}{1em} \setlength{\labelsep}{0.5em} } }

  \newcommand{\squishend}{
\end{list}  }
\vspace{-1.0in}

\noindent {\bf Shantenu Jha} is currently the Director of Cyberinfrastructure
Development at the CCT at LSU. He will be a tenure-track Assistant
Professor starting in Spring 2011. Jha's interest are at the triple
point of High-Performance \& Distributed Computing, Computation and
Data-enabled Science (CDES), and Cyberinfrastructure Research \& Development.
He is the lead of the SAGA project (\url{http://www.saga.cct.lsu});
SAGA provides a high-level API for distributed functionality and the
building blocks to develop distributed application (primarily
compute-intensive) and distributed programming systems. Jha is also
the PI for the UK e-Science Institute Research Themes on Distributed
Programming Abstractions (DPA) and the current Distributed Dynamic
Data-Intensive Programming Abstractions and Systems (3DPAS).  He is
the PI of an active NSF-OCI award to establish and facilitate US
collaborations with the UK e-Science Institute 3DPAS Research theme on
Distributed Dynamic Data. Jha is also the PI of the ~\$2M NSF-EPSCOR
Cybertools Project (http://cybertools.loni.org/). Jha is a
contributing co-editor of the chapter on ``Cyberinfrastructure
Requirements and Challengens'' of AFOSR/NSF Co-Sponsored Report on
Dynamic Data Driven Applications Systems~\cite{dddas}.

\noindent  {\bf Vision:\footnote{Much of the underlying theoretical context and
    motivating examples, have been developed in collaboration with
    Jose Fortes (Florida), Daniel Katz (UC/ANL), Manish Parashar
    (Rutgers) and Jon Weissman (Minnesota).  Specifically, Jha has
    been examining the influence of cyberinfrastructure on large-scale
    distributed scientific
    applications\cite{dpa,3dpas,dpa-survey-paper} in conjunction with
    Katz, Weissman and Parashar. } }  % Many problems
% at the forefront of science, engineering, medicine, and the social
% sciences, are increasingly complex and interdisciplinary due to the
% plethora of data sources and computational methods available today.
A common feature across many of scientific and engineering problem
domains is the amount and diversity of data and computation that must
be integrated to yield insights. % For example, large volumes of data
% are being increasingly generated by/from distributed sensors,
% scientific instruments, and simulations.
An increasingly important type of pervasive scientific and engineering
applications are data-driven applications. Such applications involve
computational activities triggered as a consequence of independent
data creation. For example, in large-scale applications such as LEAD,
data streams from the sensors drive the execution of the
computations. Often in response to a predicted or phenomenologically
interesting event, the data source and streams themselves need to be
adapted, e.g., the sampling rates or resolutions are
changed. Additional elements of dynamic compute and data adaptation
arise from spatio-temporal variations in data generation (sensors). In
many dynamic applications, processing of data often needs to take
place in-flight (e.g, via streaming) to meet boundary conditions,
and/or data volume reduction may be required. For example, in the
International Thermonuclear Experimental Reactor (ITER) application,
data has to be transformed while it is being streamed. Understanding
how to support dynamic computations is thus a fundamental, often
critically missing, element in pervasive data-intensive computing.

For a large class of dynamic data applications, data must be
integrated with extreme-scale computations and analytics as part of
end-to-end application workflows. Some specific motivations for the
need for such integration of data with scalable computing capabilities
and techniques are: (1) Exascale simulations will lead to exascale
datasets; exabytes of data will require exaflops of computing
capability to analyze and mine; (2) Large data-set sizes and the need
to operate on them in a time-bounded manner (e.g., real-time video
data-streams for image detection); (3) Large number of data-souces
arising from large geographical coverage (e.g., sensors used in
Hurricane Modelling); (4) Adaptive simulation models coupled with
disparate, varying and refined in realtime data models/sources
(e.g. Gulf of Mexico Oil Spill dispersion trajectory prediction).  We
note that the challenges of managing exaflops of computing driven by
large amounts of data are different from those of exaflops of
computing that generate exabytes of data. The ultimate goal is to
address both sets of challenges as a collective whole, but the path to
that goal will require detailed and specific capabilties for each
individual scenario. For the purposes of the PeCS workshop, we will
pursue the vision, capabilities and challenges inherent in the former,
i.e., coupling extreme scales of computing driven by large amounts of
data.

% Petabyte datasets are already common and terascale to exascale
% datasets are projected in many science fields;

\noindent {\it Challenges in achieving the Vision:} The landscape of
developing and deploying dynamic data-driven applications is dominated
by the existence of partial and often non-overlapping approaches (for
example incomplete solutions for local data versus distributed data,
analysis on static data versus dynamic data, management techniques for
small data versus big data). This is true along at least three
dimensions of cyberinfrastructure -- programming sytems (PS),
middleware and the tools \& services. This limits the capabilities
offered to scientists, effective and efficient application development
and thus the impact of extreme-scale CDES.

\noindent {\it How our vision will have impact: } As we move towards
scientific problems of ever-greater complexity and significance, which
require that we address the next-scale in compute and data, it is
imperative to understand and to design integrated cyberinfrastructure
-- programming systems, middleware, and tools and services -- for
application scenarios that have a critical need to couple dynamic and
distributed data into the analytics control-loop. For example, in
climate science, the next IPCC is expected to generate hundreds of
petabytes of data, which in turn is based upon petabytes of input data
arising from sensors.

\noindent Specific issues that our future research will contribute to:

% in genomics,
% the Joint Genome Institute alone will have about 1 PB of data this
% year and this is doubling each year.

% For dynamic data-intensive scientific applications at extreme-scale \&
% performance. In many of these application scenarios, there is a
% critical need to couple dynamic and distributed data into the
% analytics control-loop.

% for dynamic data-intensive scientific applications at
% extreme-scale \& performance. In many of these application scenarios,
% there is a critical need to couple dynamic and distributed data into
% the analytics control-loop.

\begin{itemize*}\vspace{-0.075in}
\item Analyzing the landscape of applications \&
  cyberinfrastructure %and the programming system 
  for distributed and dynamic-data scenarios

\item Development of cyberinfrastructure to support
  dynamic-data as a first-class concern % of distributed applications.

\item Design and implementation of %general-purpose %and pervasive
  cyberinfrastructure that supports dynamic data, but which adheres
  to open-standards,
  supports many applications types across multiple spatio-temporal scales,
  and can be deployed on a range of production
  distributed systems across scales.

%   Conversely,
%   the implementation and 
%   Conversely,
%   the implementation and deployment on a range of production
%   distributed systems across scales.

\item Role of autonomic decision making in dynamic-data \&
  distributed compute-data co-location challenges. Design \&
  development of middleware to
  support these capabilities 
\end{itemize*}

% Our research agenda will address the following objectives of the PeCS
% workshop:

% \begin{itemize*}\vspace{-0.075in}
% \item Identify software challenges, including middleware and operating
%   systems, for pervasive computing and associated applications;
% \item Explore the nexus between scalability and application
%   characteristics and context with the goal of identifying fundamental
%   insights, models and methods;
% \item Explore new interfaces and modes of interactions between people
%   and pervasive computing devices, applications or environments; 
% % \item In conjunction with the 3DPAS Research Themes,
% %   we will contribute to the following objective: ``Identify open
% %   problems
% %   and fundamental challenges that must be addressed to enable
% %   deployment of pervasive computing systems at massive scale''.
% \end{itemize*}

% SAGA is usable with almost every major middleware globally and is
% supported on nearly ever major compute intensive production grid
% infrastructure (such as NSF TeraGrid, European EGI etc).

% High-end data-intensive applications, the programming systems and
% tools, and both the current and to-be-developed middleware that
% support them.

% Our efforts will build upon sustained collaboration in this domain
% over the past few years. This workshop will enable us to use the
% methodology we have developed for large-scale distributed applications
% for dynamic data-intensive applications and middleware, to explore
% opportunities to leverage existing and emerging technologies to
% address extreme-scale data challenges and identify the gaps..

\vspace{-0.25in}
\begin{footnotesize}
\begin{thebibliography}{9}

\bibitem{dpa} DPA: Distributed Programming Abstraction e-Science
  Institute Research Theme. \\
  \url{http://wiki.esi.ac.uk/Distributed_Programming_Abstractions}

\bibitem{3dpas} 3DPAS: Dynamic, Distributed and Data-Intensive Programming
  Abstractions and Systems.  \\ \url{http://wiki.esi.ac.uk/3DPAS}

% \bibitem{3dpas-hpdc} 3DAPAS: Dynamic, Distributed and Data-Intensive
%   Applications, Programming Abstractions and
%   Systems. \url{https://sites.google.com/site/3dapas/}

\bibitem{dddas} Chapter on Cyberinfrastructure for Dynamically
  Data-Driven Application Scenarios, Edited by Gabrielle Allen and
  Shantenu Jha, AFOSR/NSF Co-Sponsored Workshop on Dynamic Data Driven
  Applications
  Systems. \url{http://www.dddas.org/AFOSR-NSFworkshop2010.html}

\bibitem{dpa-survey-paper} Distributed Computing Practice for
  Large-Scale Science \& Engineering Applications, Under review, by
  Jha, Cole, Katz, Rana, Parashar and Weissman. \\
  \url{https://cct.lsu.edu/~sjha/dpa_publications/dpa_surveypaper.pdf}

% \bibitem Data-Intensive Research Workshop Report, edited by Atkinson,
%   Jha et al.  \url{http://dl.dropbox.com/u/3073925/DIRWS.pdf}

\end{thebibliography} 
\end{footnotesize}
\end{document}

Understanding the primary characteristics, usage scenarios \&
requirements of 3D applications, and then developing the capabilities
to support 3D applications within the same fold as compute intensive
applications will be critical.

### 2015-06-03 3dpas_intro.tex
small wording change

@@
 independent programs, written by different authors, that are trying to
-answer the same type of science question and thus, where a single user
+answer the same type of science question, where a single user
 will only run one of them at a time.  %\katznote{Shantenu, please check

### 2015-08-18 3dpas_intro.tex
tweaks.

--- 2015-05-27 3dpas_intro.tex
@@
 
-\author{Shantenu Jha\affil{1,2}\corrauth\, Neil Chue Hong\affil{2},
-  Simon Dobson\affil{3}, Daniel S. Katz\affil{4,5},
-  Andre~Luckow\affil{1}, Omer Rana\affil{6}, Yogesh Simmhan\affil{7}}
+\author{Shantenu Jha\affil{1}\corrauth\, Daniel S. Katz\affil{4,5},
+  Andre~Luckow\affil{1}, Omer Rana\affil{6}, Yogesh Simmhan\affil{7}, Neil Chue
+  Hong\affil{2}, Simon Dobson\affil{3}}
 
@@
 
-The 3DPAS Research Theme (funded by the eSI, Edinburgh and US NSF)
-investigated applications that have these characteristics; in other
-words, it operated at the triple point of  {\it distributed} and {\it dynamic} and
-{\it data-intensive} (D3) attributes.  This document represents the conclusions
-of the 3DPAS Theme.
+% The 3DPAS Research Theme (funded by the eSI, Edinburgh and US NSF)
+% investigated applications that have these characteristics; in other
+% words, it operated at the triple point of  {\it distributed} and {\it dynamic} and
+% {\it data-intensive} (D3) attributes.  This document represents the conclusions
+% of the 3DPAS Theme.
+
 \end{abstract}
@@
 %as take away messages. 
-Section~\ref{sec:discussion}
-provides a discussion of dynamic and distributed characteristics of
-data associated with the applications we analyze in this work,
+Section~\ref{sec:observations}
+summarize the dynamic and distributed characteristics of
+data associated with the applications we've studied in this work,
 attempting to infer more general characteristics that %can be applied
 may apply
-to other similar applications.  We also suggest two metrics that may
+to other similar applications.  In \S\ref{sec:discussion}, we
 suggest two metrics that may

### 2015-09-02 3dpas_intro.tex
checkpoint.

@@
 \begin{abstract}
+
 %   Many problems at the forefront of science, engineering, medicine,
@@
 
-The landscape of scientific and enterprise
-computing is being fundamentally altered by the current surge in data. The
-resulting upheaval, grandiosely named BigData science by the Lilliputians
-in the data science community, represents a new reality where
-the volumes of data produced by scientific applications are increasing
-rapidly, driven by advancing technologies (e.g., increasing compute
+The landscape of scientific and enterprise computing is being fundamentally
+altered by the current surge in data. The resulting upheaval represents a new
+reality where the volumes of data produced by scientific applications are
+increasing rapidly, driven by advancing technologies (e.g., increasing compute
 capacity and higher resolution sensors) and decreasing costs for data
 acquisition and storage~\cite{tony-propaganda}. With the increase in data
-volumes, types, and sources of data, an important concomitant
-development has been an increase in the requirement for and complexity
-in the processing of data. Data are often naturally distributed and have
-to be processed in situ, and occasionally data that are localized
-have to be distributed in order to be processed, either for reasons
-of performance or for collaborative analytics.  Furthermore when
-processing large volumes of data, data source/destination, data
-production and transfer rates, and data processing capacity can
-all vary over time. Thus properties such as resource scheduling, data
-placement, and transfer decisions, that were statically determinable at
-small scales, emerge as dynamic and important considerations at
-large scales of data and distribution.  The BigData science problem
-is therefore only partly about large volumes of data.
-
-Our focus is on a subspace of the BigData problem space, namely that
-which is characterized by the distributed and dynamic aspects of
-data-intensive science. We introduce the concept of Dynamic
-Distributed Data-intensive science and refer to this subspace of
-BigData science as D3 science.  In order to appreciate the D3 subspace
-we will briefly discuss the importance and role of distribution and
-the pervasiveness of dynamical aspects.
+volumes, types, and sources of data, an important concomitant development has
+been an increase in the requirement for and complexity in the processing of
+data. Data are often naturally distributed and have to be processed in situ, and
+occasionally data that are localized have to be distributed in order to be
+processed, either for reasons of performance or for collaborative analytics.
+Furthermore when processing large volumes of data, data source/destination, data
+production and transfer rates, and data processing capacity can all vary over
+time.
+
+The set of challenges is only partly defined by the large volumes of
+data. Properties such as resource scheduling, data placement, and transfer
+decisions, that were statically determinable at small scales, emerge as dynamic
+and important considerations at large scales of data and distribution.
+
+Our focus is on a subspace of the large set of problems associated with
+data-intensive sciences, namely that which is characterized by the distributed
+and dynamic aspects. We introduce the concept of Dynamic Distributed
+Data-intensive science and refer to this subspace of data-intensive science as
+D3 science.  In order to appreciate the D3 subspace we will briefly discuss the
+importance and role of distribution and the pervasiveness of dynamical aspects.
 
@@
 %   about the importance of treating distributed data}
-
-Although the capacity to store data centrally is
-growing faster than Moore's Law, it is not sufficient in all cases to
-be able to ``dump the data'' in one place (or in the cloud.)
-Oftentimes the costs (data transfer, storage, consequences of
-catastrophic failure) prevent the centralization of data, and
-decentralization---storing or place data in multiple locations---become prudent.
-
-Conversely, not all production of data is centralized; in some cases
-there are multiple producers (and consumers) of data, not to mention
-one-of-a-kind data that must be shared across multiple consumers.
-Furthermore in collaborative science, data output from one analytical
-engine are often the input for the next stage of the analytical
-pipeline.  Last but not least, distribution via replication is a
-tested and effective approach to reliability and fault-tolerance; we
-will argue that it is also potentially a great performance enhancer
-and a means to ensure data consistency too.
 
@@
 
-The challenges of distributed data-intensive applications span beyond
-the management and curation of large volumes of data. A critical
-requirement is to make sense of this data, i.e., transform the data
-into information en route to assembling knowledge. Among other things,
-the transformation from data to information requires extensive amounts
-of computational processing of the data; thus the data-intensive
-nature of the BigData science problem is strongly coupled with
-compute-intensive requirements.  As the data are often distributed, a
-consequence is that many attributes of distributed systems permeate to
-data. Thus almost naturally, distributed data have attributes of
-dynamism, both at the data level as well as the infrastructure
-level.
-
-That is not to say all data-intensive applications have elements of
-both data and compute dynamism and are always distributed, but it is a
-reflection that at extreme scales, dynamism and and distribution
-become increasingly important and often correlated issues.  Thus,
-analyzing the applications along an analytical space spanned by the
+The challenges of distributed data-intensive applications span beyond the
+management and curation of large volumes of data. A critical requirement is to
+make sense of this data, i.e., transform the data into information en route to
+assembling knowledge. Among other things, the transformation from data to
+information requires extensive amounts of computational processing of the data;
+thus the data-intensive nature of the data-intensive science problem is strongly
+coupled with compute-intensive requirements.  As the data are often distributed,
+a consequence is that many attributes of distributed systems permeate to
+data.
+
+% Thus almost naturally, distributed data have attributes of dynamism, both
+% at the data level as well as the infrastructure level.
+
+% That is not to say all data-intensive applications have elements of
+% both data and compute dynamism and are always distributed, but it is a
+% reflection that at extreme scales, dynamism and and distribution
+% become increasingly important and often correlated issues.  Thus,
+
+Analyzing the applications along an analytical space spanned by the
 three ``D''s of data-intensive, distributed, and dynamic provides a
@@
 %  this has now moved upstream}
+
+Although the capacity to store data centrally is growing faster than Moore's
+Law, it is not sufficient in all cases to be able to ``dump the data'' in one
+place (or in the cloud.)  Oftentimes the costs (data transfer, storage,
+consequences of catastrophic failure) prevent the centralization of data, and
+decentralization---storing or place data in multiple locations---become prudent.
+
+Conversely, not all production of data is centralized; in some cases there are
+multiple producers (and consumers) of data, not to mention one-of-a-kind data
+that must be shared across multiple consumers.  Furthermore in collaborative
+science, data output from one analytical engine are often the input for the next
+stage of the analytical pipeline.  Last but not least, distribution via
+replication is a tested and effective approach to reliability and
+fault-tolerance; we will argue that it is also potentially a great performance
+enhancer and a means to ensure data consistency too.
 

### 2015-09-03 3dpas_intro.tex
ongoing/pending working in section 1.

@@
 skillfully architected infrastructure and programming systems.
-
-With recent attention on ``Big Data'' problems, several classification
-schemes have been proposed; one that has garnered attention is the
-3Vs~\cite{gartner3v}. These three
-dimensions (Volume, Velocity and Variety) just describe the
-characteristics of the data. %, rather than describing the interplay between Big
-%Data and applications that operate over them. 
-\jhanote{Does someone
-  have the authentic reference for the 4Vs?} \katznote{see http://www.villanovau.com/university-online-programs/what-is-big-data/ but it's not `authoritative'} In contrast,  D3 
-emphasizes the nature of scientific applications as characterized
-by their data attributes. Furthermore the D3 attributes are considered
-as a collective property, as opposed to  different disjoint attributes.
-Thus we believe there is a fundamental distinction between the D3 and
-3Vs classification schemes.
 

### 2015-09-03 3dpas_intro.tex

@@
 \affilnum{6}Cardiff University\break
-\affilnum{7}University of Southern California\break
+\affilnum{7}Indian Institute of Science\break
 }
@@
 into individual databases daily, and merges them with a master distributed-database on a weekly basis on an
-HPC cluster. Data extraction and mining operations by astronomers are composed using a pseudo-SQL
+HPC cluster~\cite{Simmhan:Building:2009}. Data extraction and mining operations by astronomers are composed using a pseudo-SQL
 language with user defined functions and submitted for execution on the master database using a batch system.
@@
 
-The application~\cite{modisAzure} is used to handle the processes of downloading data from different
+The application~\cite{modisAzure, modis:ipdps:2010} is used to handle the processes of downloading data from different
 sites and integrating them. The integration consists of reprojecting the data from one map projection to another,
@@
 {\em Data Sources} are used in the acquisition stage as inputs of data
-into the application. Broadly speaking they include types such as
+into the application. Broadly speaking these sources include types such as
 files, servers, sensors, instruments, simulations and models.
 Additionally, databases are often considered as an intermediate data
-source in types of system. The dynamic characteristics are primarily
+source in some systems. The dynamic characteristics are primarily
 due to availability. These are often accessed in the Data Generation stage of the application
@@
 of data, e.g., a field survey may bring in a lot of data during summer
-but none during other times. These are commonly used in the Data Storage stage of the application (\figurename~\ref{fig:figures_application-stages}),
+but none during other times. Some storage components like files and databases may also serve as data
+sources. These are commonly used in the Data Storage stage of the application (\figurename~\ref{fig:figures_application-stages}),
 but may also be present at the boundaries between the stages for transient staging of data.
@@
 movement of data between parts of the application, or to and from the
-application. % (often called staging) Data movement may be explicit
+application, % (often called staging) Data movement may be explicit
 and visible to the application (e.g., staging), or implicit (e.g.,
@@
 replicated close to the application). Data movement can be considered
-to be real-time (i.e., as it happens, e.g., sensor data), as real-time
+to be real-time (i.e., as it happens, e.g., sensor data), `as real-time'
 (e.g., a storage buffer has been introduced into the transport system,
@@
 characteristic of the data movement infrastructure depending on the
-application e.g. real-time processing of sensor data for disaster
+application, e.g., real-time processing of sensor data for disaster
 analysis. Other dynamic aspects of data movement include data rates,
@@
 applications stages and can be used by any of the
-stages to guide their selection or identification of incoming data sources and outgoing data sinks.
+stages to guide their selection or identification of incoming data sources and outgoing data sinks
+or storage.
 %\ysnote{rate of growth of data may be an issue.
@@
 \figurename~\ref{fig:figures_application-stages}, we identify a number
-of infrastructure services that can be associated with these stages.
+of infrastructure services for data analysis that can be associated with these stages.
 
@@
 toolboxes and visualization environments to become part of the infrastructure, some of which are
-deployed for specific domains but others which are more generic (e.g., Weka, GraphLab). These components are associated
+deployed for specific domains but others which are more generic (e.g., Weka, GraphLab, Apache Mahout). These components are associated
 with the Data Processing stage.
@@
 messaging to allow complex analyses to be performed on distributed
-data sources. They may also be dynamic if they have the capability to
+data sources. These are often composed as workflows or Directed Acyclic Graphs (DAGs) that capture
+the execution dependency between
+components. They may also be dynamic if they have the capability to
 adapt their behavior based on earlier parts of the workflow or can be
 steered based on external notification received during execution. In
-addition to supporting choreography, often making use of one or more
+addition to supporting choreography, while often making use of one or more
 resource manager/scheduler components, an orchestration component may also
@@
 collections of data which are executed in a loosely-coupled, data-parallel manner.
-The {\em MapReduce} paradigm~\cite{MapReduce} and its variants~\cite{twister,hmr} are the most prominent among these. MapReduce forces programmers to model computation in two stages: 
+The {\em MapReduce} paradigm~\cite{MapReduce} and its variants~\cite{twister,hmr} are the most
+prominent among these, with recent ones like Pregel and Apace Giraph~\cite{pregel} supporting specific domains like
+graph analytics. MapReduce forces programmers to model computation in two stages: 
 a \emph{map} phase in which a transformation is applied to every element of a collection independently of every other;  and a \emph{reduce} phase in which the transformed results 
@@
 Resource selection?}
-Infrastructure for Dynamic Properties & Infrastructure for Distributed Properties
+Infrastructure to support Dynamic Properties & Infrastructure to support Distributed Properties
  \\ \hline
@@
 		\hline	
-\multicolumn{2}{|l|}{{\textbf{Next Generation Sequencing (NGS) Analytics}}} \\ \hline	
+\multicolumn{2}{|l|}{{\textbf{Next Generation Sequencing (NGS) Analytics} (\S\ref{bioSilvia})}} \\ \hline	
 %\nchnote{For example, workload decomposition and distribution must be determined dynamically in order to optimally use resources, e.g., compute resource selection can be based on data location, processing profile and/or network capacity.
@@
 		\hline
-\multicolumn{2}{|l|}{{\textbf{ATLAS (an example of use of the WLCG) Reconstruction app}}} \\ \hline
+\multicolumn{2}{|l|}{{\textbf{ATLAS (an example of use of the WLCG) Reconstruction app} (\S\ref{WLCGSteve})}} \\ \hline
 Manage gradual growth of data using SRM \emph{(Data Storage, Access)}
@@
 		\hline
-\multicolumn{2}{|l|}{{\textbf{LSST}}} \\ \hline
+\multicolumn{2}{|l|}{{\textbf{LSST} (\S\ref{astroAdam})}} \\ \hline
 
@@
 		\hline
-\multicolumn{2}{|l|}{{\textbf{SOA Astronomy Applications}}} \\ \hline
+\multicolumn{2}{|l|}{{\textbf{SOA Astronomy Applications} (\S\ref{astroSOAAdam})}} \\ \hline
 
@@
 		\hline
-\multicolumn{2}{|l|}{{\textbf{Cosmic Microwave Background}}} \\ \hline	
-	      Dynamic resource selection
+\multicolumn{2}{|l|}{{\textbf{Cosmic Microwave Background} (\S\ref{astroJulian})}} \\ \hline	
+	      Dynamic resource selection  \emph{(Data Discovery)}
 %dynamic utilization of resources (4)
@@
 		\hline
-\multicolumn{2}{|l|}{{\textbf{Sensor Network Application}}} \\ \hline	
+\multicolumn{2}{|l|}{{\textbf{Sensor Network Application} (\S\ref{sensorSimon})}} \\ \hline	
 	     Notification of sensor data when in range, data rates and volumes \emph{(Data Movement, Notification)} %data rate, volume dynamic (1-2)
@@
 		\hline
-\multicolumn{2}{|l|}{{\textbf{Climate (ESGF)}}} \\ \hline	
+\multicolumn{2}{|l|}{{\textbf{Climate (ESGF)} (\S\ref{climateDan})}} \\ \hline	
 	      Manage data growth % New data generated over time (ESGF-wide - is this in scope of the application?)
@@
 		\hline
-\multicolumn{2}{|l|}{{\textbf{Interactive Exploration of Environmental Data}}} \\ \hline	
+\multicolumn{2}{|l|}{{\textbf{Interactive Exploration of Environmental Data} (\S\ref{envJon})}} \\ \hline	
 On-demand data subsetting, caching and versioning, real-time instrument data \emph{(Data Discovery, Data
@@
 	    %     \hline
-\multicolumn{2}{|l|}{{\textbf{PowerGrids}}} \\ \hline
+\multicolumn{2}{|l|}{{\textbf{PowerGrids} (\S\ref{power})}} \\ \hline
 
@@
 		\hline
-\multicolumn{2}{|l|}{{\textbf{Fusion (ITER)}}} \\ \hline	
+\multicolumn{2}{|l|}{{\textbf{Fusion (ITER)} (\S\ref{fusionScott})}} \\ \hline	
 	     Stream and transform data between simulation sites on-demand \emph{(Data Movement, Enrichment)}
@@
 		\hline
-\multicolumn{2}{|l|}{{\textbf{Industrial Incident Notification and Response}}} \\ \hline	
+\multicolumn{2}{|l|}{{\textbf{Industrial Incident Notification and Response} (\S\ref{Kees})}} \\ \hline	
 	     Acquire different data sources and rates, Processing determined by data
@@
 &
-Distributed data sources, Analysis by agents across institutions \emph{(Data Sources, Data
+Distributed data sources, Analysis by agents across institutions, Access control \emph{(Data Sources, Data
   Access)} % , Analysis
@@
 		\hline
-\multicolumn{2}{|l|}{{\textbf{MODIS Data Processing}}} \\ \hline
+\multicolumn{2}{|l|}{{\textbf{MODIS Data Processing} (\S\ref{modisKeith})}} \\ \hline
 
@@
 
-\multicolumn{2}{|l|}{{\textbf{Distributed Network Intrusion Detection}}} \\ \hline	
+\multicolumn{2}{|l|}{{\textbf{Distributed Network Intrusion Detection} (\S\ref{detectionJon})}} \\ \hline	
 Analyze dynamic network traffic rates \emph{(Data Movement)} % , Analysis
@@
         \hline LSST & custom layer; scripted workflows; queries \\
-        \hline SOA Astronomy & web services \\
+        \hline SOA Astronomy & web services~\cite{barker} \\
         \hline Cosmic Microwave Background & \textit{ad hoc} \\
@@
         \hline Interactive Exploration of Environmental Data &
-        visualization tools \\
-        \hline Power Grids & stream processing; data mining \\
-        \hline Fusion (ITER) & scripted pipeline \\
+        visualization tools~\cite{blower2} \\
+        \hline Power Grids & stream processing~\cite{Simmhan:sciencecloud:2011}; data mining~\cite{Yin:mapreduce:2012} \\
+        \hline Fusion (ITER) & scripted pipeline~\cite{adios} \\
         \hline Industrial Incident Notification and Response &
-        agent-based  \\
-        \hline MODIS Data Processing & scripted pipeline \\
+        agent-based~\cite{emergency4}  \\
+        \hline MODIS Data Processing & scripted pipeline~\cite{modis:ipdps:2010} \\
 %        \hline Floating Sensors & \textit{ad hoc} \\
@@
 the hardware and software needed to run the applications, we
-note that the specific hardware isn't a primary issue.  The
-applications we've discussed all use a mix of the systems that exist
+note that the specific hardware is not a primary issue.  The
+applications we have discussed all use a mix of the systems that exist
 today, which have aspects of web, grid, and cloud, and can be
@@
 are really the software and services on top of the hardware.  In this
-section, we've discussed infrastructure to support data management and
+section, we have discussed infrastructure to support data management and
 data analysis, as well as the tools and systems used to program
@@
 software and services that the applications use, so the software and programming infrastructure
-we've described here is much more broad than what is used by the
-applications we've studied.
+we have described here is much more broad than what is used by the
+individual applications we have studied.
 
@@
 %as take away messages. 
-Section~\ref{sec:observations}
-summarize the dynamic and distributed characteristics of
-data associated with the applications we've studied in this work,
+Section~\ref{sec:discussion}
+provides a discussion of dynamic and distributed characteristics of
+data associated with the applications we analyze in this work,
 attempting to infer more general characteristics that %can be applied
 may apply
-to other similar applications.  In \S\ref{sec:discussion}, we
+to other similar applications.  We also suggest two metrics that may
 suggest two metrics that may
@@
 
-\subsection{Observations}
-\label{sec:observations}
-
-In this subsection we summarize the source and nature of dynamic data that we observe in
-the application surveyed in \S\ref{sec:scenarios}, and the types of
-data distribution that we observe in such applications.
+\subsection{Discussion}
+\label{sec:discussion}
+
+In this section we attempt to summarize four key themes from previous
+sections: (i) the source and nature of dynamic data that we observe in
+the application surveyed in \S\ref{sec:scenarios}; (ii) the types of
+data distribution that we observe in such applications; (iii) the
+general characteristics that can be associated with D3 science
+applications; (iv) the characteristics of a computational
+infrastructure that may be most suited for D3 science
+applications. Whereas (i) and (ii) are primarily observations that we
+can make on existing applications that we have discussed in
+\S\ref{sec:scenarios}, (iii) and (iv) attempt to synthesize general
+trends that can be applied to other applications we have not
+considered. \katznote{should this be split into two subsections then? Perhaps Observations \& Trends?}
+
+%\katznote{Some text is needed here to introduce the discussion subsection.}
+%\rananote{Added some text above -- does this make sense Dan?}
 
@@
 often, sensor data is also archived and aggregated for historical
-analyses, which leads to higher data volumes.
+analyses, which leads to higher data volumes. \katznote{not sure about the `other' part of this...}
 
@@
 % of distribution changes can be insightful.
-
-\subsection{Discussion}
-\label{sec:discussion}
-
-Now that we have discussed dynamic data and distributed data in the
-context of specific applications, we can use this experience to
-understand the characteristics that can be associated with D3 science
-applications, and the characteristics of a computational
-infrastructure that may be most suited for D3 science
-applications. 
-We believe these are general
-trends that can be applied to other applications we have not
-considered.
 

### 2015-09-12 3dpas_intro.tex
checkpoint.

--- 2015-09-03 3dpas_intro.tex
@@
 
-%   Many problems at the forefront of science, engineering, medicine,
-%   and the social sciences are increasingly complex and
-%   interdisciplinary due to the plethora of data sources and
-%   computational methods available today.  A common feature across many
-%   of these problems is the amount and diversity of data and
-%   computation that must be integrated to yield insights. Such data is
-%   increasingly large-scale and distributed: the datasets, their
-%   locations, and associated scheduling decisions are
-%   time-dependent.
+  A common feature across many science and engineering applications is the
+  amount and diversity of data and computation that must be integrated to yield
+  insights. Often data is increasingly large-scale and distributed. Often the
+  datasets, their locations, availability, analysis and associated scheduling
+  decisions are time-dependent. Collectively these give rise to dynamic
+  distributed data intensive applications. This paper attempts to survey several
+  such applications, to provide a common conceptual framework to understand them
+  as well as examine the infrastructure used in support of applications.
+  Whereas the analysis of large volumes of ``static'' data have received
+  significant attention, the ability to manage large-volumes of dynamic and
+  distributed data is more primitive. This work will ameliorate the state.
 

### 2015-09-12 3dpas_intro.tex
refinement.

@@
   decisions are time-dependent. Collectively these give rise to dynamic
-  distributed data intensive applications. This paper attempts to survey several
-  such applications, to provide a common conceptual framework to understand them
-  as well as examine the infrastructure used in support of applications.
-  Whereas the analysis of large volumes of ``static'' data have received
-  significant attention, the ability to manage large-volumes of dynamic and
-  distributed data is more primitive. This work will ameliorate the state.
+  distributed data intensive applications.  The characteristics, requirements,
+  and software systems for the analysis of large volumes of ``static'' data have
+  received significant attention; dynamic, distributed and data intensive (3D)
+  have received relatively less. This work will ameliorate the state by
+  surveying several representative 3D application scenarios, to provide a common
+  conceptual framework to understand them as well as examine the infrastructure
+  used in support of applications.
 

### 2015-09-12 3dpas_intro.tex
Scattered changes.

--- 2015-09-12 3dpas_intro.tex
@@
   decisions are time-dependent. Collectively these give rise to dynamic
-  distributed data intensive applications. This paper attempts to survey several
-  such applications, to provide a common conceptual framework to understand them
-  as well as examine the infrastructure used in support of applications.
-  Whereas the analysis of large volumes of ``static'' data have received
-  significant attention, the ability to manage large-volumes of dynamic and
-  distributed data is more primitive. This work will ameliorate the state.
+  distributed data intensive applications.  The characteristics, requirements,
+  and software systems for the analysis of large volumes of ``static'' data have
+  received significant attention; dynamic, distributed and data intensive have
+  received relatively less. This work will ameliorate the state by surveying
+  several representative application scenarios, to provide a common conceptual
+  framework to understand them as well as examine the infrastructure used in
+  support of applications.
 
@@
 The landscape of scientific and enterprise computing is being fundamentally
-altered by the current surge in data. The resulting upheaval represents a new
-reality where the volumes of data produced by scientific applications are
-increasing rapidly, driven by advancing technologies (e.g., increasing compute
-capacity and higher resolution sensors) and decreasing costs for data
-acquisition and storage~\cite{tony-propaganda}. With the increase in data
-volumes, types, and sources of data, an important concomitant development has
-been an increase in the requirement for and complexity in the processing of
-data. Data are often naturally distributed and have to be processed in situ, and
-occasionally data that are localized have to be distributed in order to be
+altered by the prodigious volumes of data.  Not only are the volumes of data
+produced by scientific applications increasing rapidly, there is also a
+corresponding increase in the amount of data that needs to be processed and
+analyzed.  With the increase in data volumes, types, and sources of data, an
+important concomitant development has been an increase in the requirement for
+and complexity of the infrastructure needed to support these requirements.
+Furthermore, data are often naturally distributed and have to be processed in
+situ; occasionally data that are localized have to be distributed in order to be
 processed, either for reasons of performance or for collaborative analytics.
-Furthermore when processing large volumes of data, data source/destination, data
-production and transfer rates, and data processing capacity can all vary over
-time.
-
-The set of challenges is only partly defined by the large volumes of
-data. Properties such as resource scheduling, data placement, and transfer
-decisions, that were statically determinable at small scales, emerge as dynamic
-and important considerations at large scales of data and distribution.
-
-Our focus is on a subspace of the large set of problems associated with
-data-intensive sciences, namely that which is characterized by the distributed
-and dynamic aspects. We introduce the concept of Dynamic Distributed
+The scale and significance of varying data source/destination and
+transfer/production rates makes temporal variations important, which in turn
+eventually impact the design and development of infrastructure to support them:
+resource scheduling, data placement, and transfer decisions that were statically
+determinable at small scales, emerge as dynamic and important considerations at
+large scales of data and distribution.
+
+% driven by advancing technologies (e.g., increasing compute capacity and higher
+% resolution sensors) and decreasing costs for data acquisition and
+% storage~\cite{tony-propaganda}.
+
+The focus of this paper is on a subspace of the large set of problems associated
+with data-intensive sciences, namely that which is characterized by {\it both}
+distributed and dynamic aspects. We introduce the concept of Dynamic Distributed
 Data-intensive science and refer to this subspace of data-intensive science as
@@
 \section*{Acknowledgements}
-This paper is an outcome of the UK e-Science Institute sponsored
-Research Theme on Dynamic Distributed Data-Intensive Programming
-Abstractions and Systems. This work is also sponsored by NSF
-OCI-1059635.  We thank Jon Weissman and Jon Blower for useful initial
-discussions, as well as 3DPAS workshop attendees. Discussions of the
-selected applications benefited from the generous help of
-Adam Barker,
-Jon Blower,
-Julian Borrill,
-Silvia Delgado Olabarriaga,
-Steve Fisher,
-Keith Jackson,
-Scott Klasky,
-Bob Mann,
-Don Middleton,
-Kees Nieuwenhuis,
-Manish Parashar,
-Stephen Pascoe,
-and
-Jon Weissman,
-though any errors are the responsibility of the authors.
-Some of the work by Katz was
-supported by the National Science Foundation while working at the
-Foundation; any opinion, finding, and conclusions or recommendations
-expressed in this material are those of the author(s) and do not
-necessarily reflect the views of the National Science Foundation.
+This work was sponsored by NSF OCI-1059635 as well as made possible by the UK
+e-Science Institute sponsored Research Theme on Dynamic Distributed
+Data-Intensive Programming Abstractions and Systems. We thank Jon Weissman and
+Jon Blower for useful initial discussions, as well as 3DPAS workshop
+attendees. Discussions of the selected applications benefited from the generous
+help of Adam Barker, Jon Blower, Julian Borrill, Silvia Delgado Olabarriaga,
+Steve Fisher, Keith Jackson, Scott Klasky, Bob Mann, Don Middleton, Kees
+Nieuwenhuis, Manish Parashar, Stephen Pascoe, and Jon Weissman, though any
+errors are the responsibility of the authors.  Some of the work by Katz was
+supported by the National Science Foundation while working at the Foundation;
+any opinion, finding, and conclusions or recommendations expressed in this
+material are those of the author(s) and do not necessarily reflect the views of
+the National Science Foundation.
 

### 2015-09-12 3dpas_intro.tex
conclusion.

@@
 
-Many fields, such as biology, that until recently were not
-characterized by their intensive use of data are rapidly becoming
-data-driven.  The growth in data has profound implications for the
-design and implementation of services and infrastructure.  For
-example, there is a lifecycle to data that has often been ignored
-until recently: the decreasing cost of storage has made it attractive
-to archive \emph{everything}, just in case it becomes useful. This
-approach is becoming untenable in many areas: some datasets are simply
-too big to archive, and so must be summarized and their raw form
-discarded (possibly in real time). Others are easier to regenerate
-than to store.  Still others are too transient to justify their
-storage, and may in aggregate be too big to store even when being
-small individually.
-
-Additionally, it is useful to observe that the trend toward more
-unstructured and more dynamic data sources lead to higher demands on
-data processing infrastructure.~\cite{snia_2012}
-Any data science architecture needs to
-explicitly recognize, capture, and manage the lifecycle requirements
-of different datasets, within the constraints of available time and
-storage. Associated with the growth in data volumes are increasing
-levels of dynamism and distribution; sometimes distribution is
-logical, but sometimes the data is physically distributed for reasons
-of performance or convenience. At large scales, dynamism becomes
-intrinsic to the application and infrastructure; sometimes it occurs
-because of changing experimental or analytic conditions, sometimes due
-to the availability of infrastructure and resources.
-%We close out this paper with a brief
-%discussion of a range of issues associated with the distribution and
-%dynamism of data.
-
-We have identified many different application approaches, programming
-systems, and infrastructure techniques applicable to dynamic data.
-Consistent with experience in other related fields, there seems no
-reason to expect that these approaches and techniques will disappear
-in the near future. While a convergence in approaches and techniques
-might be expected, this has proved to be elusive; there remains a
-diversity in the design, implementation, and execution of scientific
-codes.  It may even be argued that it is premature to have convergence;
-after all, D3 Science is still an emerging domain.
-
-From the previous discussion we can draw out some conclusions about
-the trends within dynamic, distributed, data-intensive science. For
-instance, in \S\ref{sec:scenarios}, we divided applications into
-groups: traditional and infrastructural applications.  Traditional
-applications have been built by one or more authors in order to solve
-a particular problem or to answer a science question.  Infrastructural
-applications have been built by groups or communities to allow the
-solution of a set of problems, or to answer a set of science
-questions.  In both cases, we can observe that the same set of
-applications stages (as shown in
-\figurename~\ref{fig:figures_application-stages}) can be applied, with
-APIs defined as the inputs and outputs of those stages.  These APIs
-are very informal and loose for the traditional applications, but more
-fixed and conservative for the infrastructural applications.
-% \jhanote{is there a basis for these conjectures?} \katznote{just the
-%   evidence of the applications themselves}
-These APIs limit how new application paradigms %\jhanote{new paradigms of what?}
-can be introduced for the infrastructural applications, unless those
-paradigms happen to fit the existing APIs.  Another way of saying this
-is that the abstractions for dynamism and distribution that can be
-used in the infrastructural applications are those abstractions that
-the developers of the original applications were aware of at the time
-the infrastructural applications were defined.  This leads to a
-possible mismatch, where computer scientists have new ideas that could
-be applied to the class of problems that an infrastructural
-application was designed to solve, but that the infrastructural
-application doesn't allow, due to the fixed APIs.
-New tools may be developed to address a
-narrow set of features while ignoring the larger set of problems that
-prevented earlier tools from being used.  These tools then may be used
-be new applications as those applications are developed; however, they
-are unlikely be used by the existing applications they were meant to
-help.
-
-This is because applications
-(particularly infrastructural applications)
-aren't developed in a vacuum, and that the constraints on the
-applications lead to new tools that are or could be developed not
-being able to be applied in practice.  For example, while
-workflow systems are often sufficiently domain-independent that
-there does not seem to be a tie of a particular system to a particular
-infrastructural application, there are a very large number of workflow
-systems that exist, and it is common in the authors' experience to
-hear about new ones being developed from scratch because ``the
-ones that exist don't quite do what is needed for my application or system.''
+The growth in data volumes and its importance is having profound implications on
+the way applications are designed to the way we develop infrastructure and
+provision associated services.  Many fields, such as biology, that until
+recently were not characterized by their intensive use of data are rapidly
+becoming data-driven.
+
+Associated with the growth in data volumes are increasing levels of dynamism and
+distribution; sometimes distribution is logical, but sometimes the data is
+physically distributed for reasons of performance or convenience. At large
+scales, dynamism becomes intrinsic to the application and infrastructure;
+sometimes it occurs because of changing experimental or analytic conditions,
+sometimes due to the availability of infrastructure and resources. In this
+paper, we have identified many different applications characteristics,
+programming systems, and infrastructure techniques that either support dynamic
+data or produce it.
+
+% For example, there is a lifecycle to data that has often been ignored until
+% recently: the decreasing cost of storage has made it attractive to archive
+% \emph{everything}, just in case it becomes useful. This approach is becoming
+% untenable in many areas: some datasets are simply too big to archive, and so
+% must be summarized and their raw form discarded (possibly in real time). Others
+% are easier to regenerate than to store.  Still others are too transient to
+% justify their storage, and may in aggregate be too big to store even when being
+% small individually.  Additionally, it is useful to observe that the trend toward
+% more unstructured and more dynamic data sources lead to higher demands on data
+% processing infrastructure~\cite{snia_2012}. Any data science architecture needs
+% to explicitly recognize, capture, and manage the lifecycle requirements of
+% different datasets, within the constraints of available time and storage.
+
+% Consistent with experience in other related fields, there seems no reason to
+% expect that these approaches and techniques will disappear in the near
+% future. While a convergence in approaches and techniques might be expected, this
+% has proved to be elusive; there remains a diversity in the design,
+% implementation, and execution of scientific codes.  It may even be argued that
+% it is premature to have convergence; after all, D3 Science is still an emerging
+% domain.
+
+% From the previous discussion we can draw out some conclusions about the trends
+% within dynamic, distributed, data-intensive science. For instance,
+
+In \S\ref{sec:scenarios}, we divided applications into groups: traditional and
+infrastructural applications.  Traditional applications have been built by one
+or more authors in order to solve a particular problem or to answer a science
+question.  Infrastructural applications have been built by groups or communities
+to allow the solution of a set of problems, or to answer a set of science
+questions. All applications in general, but infrastructural applications in
+particular cannot be developed in vacuum; they must be executed on shared
+infrastructure, and are thus sensitive to the external tools and services, as
+well as their provisioning as software-systems.
+
+As part of this survey, we observe that there exist patterns in the described
+applications; however, implementation and deployment specific details can often
+mask these patterns.  For example, applications often differ in the way they
+handle distribution and dynamism, such as managing such as data/compute
+localities. This makes it difficult to support patterns of distribution and
+dynamism in a general-purpose fashion, for anything but the simplest patterns.
+
+% In both cases, we can observe that the same set of applications stages (as shown
+% in \figurename~\ref{fig:figures_application-stages}) can be applied, with APIs
+% defined as the inputs and outputs of those stages.  These APIs are very informal
+% and loose for the traditional applications, but more fixed and conservative for
+% the infrastructural applications. These APIs limit how new application
+% paradigms can be introduced for the infrastructural applications, unless those
+% paradigms happen to fit the existing APIs.  Another way of saying this is that
+% the abstractions for dynamism and distribution that can be used in the
+% infrastructural applications are those abstractions that the developers of the
+% original applications were aware of at the time the infrastructural applications
+% were defined.  This leads to a possible mismatch, where computer scientists have
+% new ideas that could be applied to the class of problems that an infrastructural
+% application was designed to solve, but that the infrastructural application
+% doesn't allow, due to the fixed APIs.  New tools may be developed to address a
+% narrow set of features while ignoring the larger set of problems that prevented
+% earlier tools from being used.  These tools then may be used by new applications
+% as those applications are developed; however, they are unlikely be used by the
+% existing applications they were meant to help.
+
+% constraints on the applications lead to new tools that are or could be developed
+% not being able to be applied in practice.
+
+% For example, while workflow systems are often sufficiently domain-independent
+% that there does not seem to be a tie of a particular system to a particular
+% infrastructural application, there are a very large number of workflow systems
+% that exist, and it is common in the authors' experience to hear about new ones
+% being developed from scratch because ``the ones that exist don't quite do what
+% is needed for my application or system.''
 
@@
 
-As part of this survey, we also observe that there exist patterns in
-the described applications; however, implementation and deployment
-specific details can often mask these patterns.  For example,
-applications often differ in the way they handle distribution and
-dynamism, such as managing such as data/compute localities. This makes
-it difficult to support patterns of distribution and dynamism in a
-general-purpose fashion, for anything but the simplest patterns. In
-\S\ref{sec:quantifyingdistapps} we have attempted to initiate a
-discussion in this context however by considering two
-ratios %(identified as parameters $\alpha$ and $\beta$)
-to relate data volume and distribution with compute operations
-performed by an application. We believe determining values for these
-parameters would provide a useful starting point for comparing
-infrastructure requirements of Big Data applications.
+% In \S\ref{sec:quantifyingdistapps} we have attempted to initiate a discussion in
+% this context however by considering two ratios to relate data volume and
+% distribution with compute operations performed by an application. We believe
+% determining values for these parameters would provide a useful starting point
+% for comparing infrastructure requirements of data-intensive applications and
+% systems.
+
 %\jhanote{I transposed the last two paragraphs}
@@
 \section*{Acknowledgements}
-This work was sponsored by NSF OCI-1059635 as well as made possible by the UK
-e-Science Institute sponsored Research Theme on Dynamic Distributed
-Data-Intensive Programming Abstractions and Systems. We thank Jon Weissman and
-Jon Blower for useful initial discussions, as well as 3DPAS workshop
-attendees. Discussions of the selected applications benefited from the generous
-help of Adam Barker, Jon Blower, Julian Borrill, Silvia Delgado Olabarriaga,
-Steve Fisher, Keith Jackson, Scott Klasky, Bob Mann, Don Middleton, Kees
-Nieuwenhuis, Manish Parashar, Stephen Pascoe, and Jon Weissman, though any
-errors are the responsibility of the authors.  Some of the work by Katz was
-supported by the National Science Foundation while working at the Foundation;
-any opinion, finding, and conclusions or recommendations expressed in this
-material are those of the author(s) and do not necessarily reflect the views of
-the National Science Foundation.
+This work was sponsored by NSF OCI-1059635 and made possible by the UK e-Science
+Institute Research Theme on Dynamic Distributed Data-Intensive Programming
+Abstractions and Systems. We thank Jon Weissman and Jon Blower for useful
+initial discussions, as well as 3DPAS workshop attendees. Discussions of the
+selected applications benefited from the generous help of Adam Barker, Jon
+Blower, Julian Borrill, Silvia Delgado Olabarriaga, Steve Fisher, Keith Jackson,
+Scott Klasky, Bob Mann, Don Middleton, Kees Nieuwenhuis, Manish Parashar,
+Stephen Pascoe, and Jon Weissman, though any errors are the responsibility of
+the authors.  Some of the work by Katz was supported by the National Science
+Foundation while working at the Foundation; any opinion, finding, and
+conclusions or recommendations expressed in this material are those of the
+author(s) and do not necessarily reflect the views of the National Science
+Foundation.
 

### 2015-09-12 3dpas_intro.tex
finished refinement of introduction.

@@
 
-The focus of this paper is on a subspace of the large set of problems associated
-with data-intensive sciences, namely that which is characterized by {\it both}
-distributed and dynamic aspects. We introduce the concept of Dynamic Distributed
-Data-intensive science and refer to this subspace of data-intensive science as
-D3 science.  In order to appreciate the D3 subspace we will briefly discuss the
-importance and role of distribution and the pervasiveness of dynamical aspects.
-
 % \begin{itemize}
@@
 
-Whether data are a priori distributed or localized, a critical
-challenge is to find an efficient and effective means to either
-distribute the initially localized data, or manage the distribution
-(or localization) of a priori distributed data.
+% Whether data are a priori distributed or localized, a critical
+% challenge is to find an efficient and effective means to either
+% distribute the initially localized data, or manage the distribution
+% (or localization) of a priori distributed data.
 
@@
 coupled with compute-intensive requirements.  As the data are often distributed,
-a consequence is that many attributes of distributed systems permeate to
-data.
+a consequence is that many attributes of distributed systems permeate to data.
+
+Subjecting a range of otherwise distinct applications to a
+common analytical framework provides advantages to the infrastructure
+scientists and cyberinfrastructure experts, namely suggestions for
+where effort is both needed and likely to yield maximum returns. In
+other words, as D3 applications evolve, our analysis can provide
+guidance on where to focus effort on infrastructure.  Last but not
+least, the advantages of providing a common but extensible vocabulary
+that can be used by the entire community, application end-users,
+developers and infrastructure scientists alike, in an emerging field
+cannot be overstated. One may argue that such vocabulary and
+terminology has been missing from the practice of distributed
+cyberinfrastructure in general, with less than desirable consequences.
 
@@
 
-Analyzing the applications along an analytical space spanned by the
-three ``D''s of data-intensive, distributed, and dynamic provides a
-general and effective framework.  The diverse set of requirements to
-support D3 scenarios points to the need for a careful understanding of
-D3 characteristics as well as sophisticated, extensible, and
-skillfully architected infrastructure and programming systems.
+The focus of this paper is on a subspace of the large set of problems associated
+with data-intensive sciences, namely that which is characterized by {\it both}
+distributed and dynamic aspects. We introduce the concept of Dynamic Distributed
+Data-intensive science and refer to this subspace of data-intensive science as
+D3 science.  Analyzing the applications along an analytical space spanned by the
+three ``D''s of data-intensive, distributed, and dynamic provides a general and
+effective framework.  The diverse set of requirements to support D3 scenarios
+points to the need for a careful understanding of D3 characteristics as well as
+sophisticated, extensible, and skillfully architected infrastructure and
+programming systems.
+
+%  In order to appreciate the D3 subspace we will briefly discuss the
+% importance and role of distribution and the pervasiveness of dynamical aspects.
 
@@
 %  data-intensive and dynamic aspects of our study.''}  
-\jhanote{we do
-  not discuss explicitly the 4th D, i.e. diversity (courtesy AL) it is
-  assumed but not discussed}
+\jhanote{we do not discuss explicitly the 4th D, i.e. diversity (courtesy AL) it
+  is assumed but not discussed}
 
@@
 
-Distribution can occur on many levels.
 %, as shown in  
 %\figurename~\ref{fig:figures_terminology}.
-In this paper, we use {\em locally distributed} to indicate that
-data/compute resides on multiple nodes within one data center.
-{\em Geographically distributed} describes the distribution of data and/or compute
-across multiple data centers. Further, we define the degree as the
-number of data sources and sites that are used for storing and
-processing the data. A distributed application is an application that
-is composed of entities (i.e., either storage/compute resource or a
-data source) from at least two physical locations. In this work, our
-focus is almost exclusively on geographically distributed data and
-applications.
 
@@
 
-\jhanote{refine next two paragraph...} Analogous to the distinction between programming in
-the large versus system level programming, there is a distinction
-between the ``cyberinfrastructure/systems considerations/design at the
-large-scale'' versus ``CI considerations at the small-scale''. This
-paper is primarily about CI/systems considerations at the large-scale.
-Searching for common solutions and patterns at the large-scale is
-likely to be more fruitful than at small-scales where customized tools
-and optimization of low-level systems features is likely to return
-greater yield.
-
-Thus, subjecting a range of otherwise distinct applications to a
-common analytical framework provides advantages to the infrastructure
-scientists and cyberinfrastructure experts, namely suggestions for
-where effort is both needed and likely to yield maximum returns. In
-other words, as D3 applications evolve, our analysis can provide
-guidance on where to focus effort on infrastructure.  Last but not
-least, the advantages of providing a common but extensible vocabulary
-that can be used by the entire community, application end-users,
-developers and infrastructure scientists alike, in an emerging field
-cannot be overstated. One may argue that such vocabulary and
-terminology has been missing from the practice of distributed
-cyberinfrastructure in general, with less than desirable consequences.
-
 %\katznote{some of this is repeating ideas from the para ``An important
@@
 
-Although the capacity to store data centrally is growing faster than Moore's
-Law, it is not sufficient in all cases to be able to ``dump the data'' in one
-place (or in the cloud.)  Oftentimes the costs (data transfer, storage,
-consequences of catastrophic failure) prevent the centralization of data, and
-decentralization---storing or place data in multiple locations---become prudent.
-
-Conversely, not all production of data is centralized; in some cases there are
-multiple producers (and consumers) of data, not to mention one-of-a-kind data
-that must be shared across multiple consumers.  Furthermore in collaborative
-science, data output from one analytical engine are often the input for the next
-stage of the analytical pipeline.  Last but not least, distribution via
-replication is a tested and effective approach to reliability and
-fault-tolerance; we will argue that it is also potentially a great performance
-enhancer and a means to ensure data consistency too.
+\jhanote{refine next four paragraph...}
+
+% Further, we define the degree as the number of data sources and sites that are
+% used for storing and processing the data. A distributed application is an
+% application that is composed of entities (i.e., either storage/compute resource
+% or a data source) from at least two physical locations. In this work, our focus
+% is almost exclusively on geographically distributed data and applications.
+
+% Although the capacity to store data centrally is growing faster than Moore's
+% Law, it is not sufficient in all cases to be able to ``dump the data'' in one
+% place (or in the cloud.)  Oftentimes the costs (data transfer, storage,
+% consequences of catastrophic failure) prevent the centralization of data, and
+% decentralization---storing or place data in multiple locations---become prudent.
+
+% Conversely, not all production of data is centralized; in some cases there are
+% multiple producers (and consumers) of data, not to mention one-of-a-kind data
+% that must be shared across multiple consumers.  Furthermore in collaborative
+% science, data output from one analytical engine are often the input for the next
+% stage of the analytical pipeline.  Last but not least, distribution via
+% replication is a tested and effective approach to reliability and
+% fault-tolerance; we will argue that it is also potentially a great performance
+% enhancer and a means to ensure data consistency too.
 
@@
 
-We would like the reader to keep in mind three primary context and
-scope-setting points:
+We would like the reader to keep in mind some context and scope-setting points:
 
@@
 
-\item This {\it survey} is not meant to be complete, either in scope
-  of domains covered or types of infrastructure and application
-  surveyed.  They by definition represent the interests and experience
-  of the authors.
-
-\item The focus is on applications and infrastructure that arise from
-  science and engineering projects in academia; this is not exclusive
-  but is more a guiding principle.  While we are cognizant of
-  enterprise applications and infrastructure, and recognize the
-  impressive advances made therein, we also acknowledge that the
-  factors such as scale, distribution, open-source, and
-  interoperability influence the solutions employed and infrastructures
-  used; it is important to therefore focus on problems that emerge as
-  in the space defined by these ``academic constraints''.
-
-\item Almost by definition, any attempt to capture the
-  state-of-the-art of such a dynamic and emerging domain is bound to
-  be fraught with limitations and issues of scoping.  A complete and
-  rigorous survey of the landscape of BigData is bound to be obsolete
-  by the time it emerges; so given the rate of change, we asked what
-  could be done. What we present is a partial analysis motivated by a
-  set of existing and active D3 applications, their trials and
-  tribulations, successes and solutions. We believe there is merit in
-  capturing the state-of-play as is, deriving a research agenda
-  and set of specific yet crosscutting issues, and presenting these
-  to the community.
+\item This {\it survey} is not meant to be complete, either in scope of domains
+  covered or types of infrastructure and application surveyed.  They by
+  definition represent the interests and experience of the authors.
+
+\item The focus is on applications and infrastructure that arise from science
+  and engineering projects in academia; this is not exclusive but is more a
+  guiding principle.  While we are cognizant of enterprise applications and
+  infrastructure, and recognize the impressive advances made therein, we also
+  acknowledge that the factors such as scale, distribution, open-source, and
+  interoperability influence the solutions employed and infrastructures used; it
+  is important to therefore focus on problems that emerge as in the space
+  defined by these ``academic constraints''.
+
+\item Almost by definition, any attempt to capture the state-of-the-art of such
+  a dynamic and emerging domain is bound to be fraught with limitations and
+  issues of scoping.  A complete and rigorous survey is bound to be obsolete by
+  the time it emerges; so given the rate of change, we asked what could be
+  done. What we present is a partial analysis motivated by a set of existing and
+  active D3 applications, their trials and tribulations, successes and
+  solutions. We believe there is merit in capturing the state-of-play as is,
+  deriving a research agenda and set of specific yet crosscutting issues, and
+  presenting these to the community.
+
+\item Distribution can occur on many levels.  In this paper, we use {\em locally
+    distributed} to indicate that data/compute resides on multiple nodes within
+  one data center.  {\em Geographically distributed} describes the distribution
+  of data and/or compute across multiple data centers.  Also, our focus is
+  primarily on infrastructure considerations at the large-scale, as searching
+  for common solutions and patterns at the large-scale is likely to be more
+  fruitful than at small-scales where customized tools and optimization of
+  low-level systems features is likely to return greater yield.
+
+  % Analogous to the distinction between programming in the large versus system
+  % level programming, there is a distinction between the ``infrastructure''
+  % considerations at the large-scale versus the small-scale.
+
 \end{itemize}
@@
 %\end{verbatim}
-
-\subsection{Methodology}
-
-This paper originated in work at the UK e-Science Institute (eSI) at
-the University of Edinburgh, in a research theme examining Dynamic
-Distributed Data-intensive Programming Abstractions and Systems,
-called 3DPAS~\cite{3dpas-theme}.  The description of
-this theme's work was:
-
-\begin{quote}
-Many problems at the forefront of science, engineering, medicine, and the social sciences, are increasingly complex and interdisciplinary due to the plethora of data sources and computational methods available today.
-A common feature across many of these problem domains is the amount and diversity of data and computation that must be integrated to yield insights.
-
-For many complex data-intensive applications, moving the data may have restrictions.
-Increasingly important type of data-intensive applications are data-driven applications.
-For example, data is increasingly large-scale, distributed arising from sensors, scientific instruments \& simulations.
-Such data-driven applications will involve computational activities triggered as a consequence of independent data creation; thus it is imperative for an application to be able to respond to unplanned changes in data load or content.
-Understanding how to support dynamic computations is a fundamental, but currently a critical missing element in data-intensive computing.
-
-The 3DPAS theme seeks to understand the landscape of dynamic, distributed, data-intensive computing: the programming models and abstractions, the run-time and middleware services, and the computational infrastructure.
-It will analyze existing tools and services, identify missing pieces and new abstractions, and propose practical solutions and best practices.
-\end{quote}
-
-The theme held three workshops, and in each, application scientists and computing technologists
-were invited to give presentations on their work, and then to be involved in discussions to
-organize and understand the presented materials.  Between the workshops, the theme
-organizers also drafted outlines of a report that was a predecessor of this paper, for
-discussion in the next workshop.  Overall, there were about 50 workshop attendees.
-
-In order to select applications and gather information about
-them, during the various workshop discussions, when an
-application that ``felt'' new was mentioned, we asked the person who
-mentioned it to provide a written set of answers to the following questions about the
-application:
-
-\begin{enumerate}
-\item What is the purpose of the application?
-\item How is the application used to do this?
-\item What infrastructure is used? (including compute, data, network, instruments, etc.)
-\item What dynamic data is used in the application?
-\begin{enumerate}
-\item What are the types of data,
-\item What is the size of the data set(s)?
-\end{enumerate}
-\item How does the application get the data?
-\item What are the time (or quality) constraints on the application?
-\item How much diverse data integration is involved?
-\item How diverse is the data?
-\item Please feel free to also talk about the current state of the
-application, if it exists today, and any specific gaps that you know
-need to be overcome.
-\end{enumerate}
-
-We then used the text of these answers to describe each of the applications (in \S\ref{sec:scenarios}) in terms of:
-
-\begin{itemize}
-\item What does the application do? What problem does it solve?
-\item How does it do it? How does the application run?  What infrastructure does it need/use?
-\item What are the big data aspects of the application?
-\item What are the distributed aspects of the application?
-\item What are the dynamic aspects  of the application?
-\item What else is important about the application?  What does it do well?  Poorly? How was the app written? What tools does it use?
-\end{itemize}
-
-After the workshops, the theme organizers and a set of attendees who
-were interested in participating started discussing how to organize
-and analyze the applications and the technologies, including
-infrastructure, programming systems and abstractions.  These
-participants are the current authors of this paper, and this paper is
-the result of that process.
-
-As part of the theme, two 3DPAS workshops~\cite{3dapas,D3-escience}
-were organized at major conferences: HPDC 2011 and IEEE eScience 2011.
-These workshops had both peer-reviewed and invited papers.
-
-A combination of invitation-only focused discussion meetings and
-broader community engagement events were used to ensure that the
-applications surveyed and ideas captured in this paper are
-representative of current community thinking as well as of
-intellectual relevance.
 
@@
 
+\section*{Methodology}
+
+This paper originated in work at the UK e-Science Institute (eSI) at
+the University of Edinburgh, in a research theme examining Dynamic
+Distributed Data-intensive Programming Abstractions and Systems,
+called 3DPAS~\cite{3dpas-theme}.  The description of
+this theme's work was:
+
+\begin{quote}
+Many problems at the forefront of science, engineering, medicine, and the social sciences, are increasingly complex and interdisciplinary due to the plethora of data sources and computational methods available today.
+A common feature across many of these problem domains is the amount and diversity of data and computation that must be integrated to yield insights.
+
+For many complex data-intensive applications, moving the data may have restrictions.
+Increasingly important type of data-intensive applications are data-driven applications.
+For example, data is increasingly large-scale, distributed arising from sensors, scientific instruments \& simulations.
+Such data-driven applications will involve computational activities triggered as a consequence of independent data creation; thus it is imperative for an application to be able to respond to unplanned changes in data load or content.
+Understanding how to support dynamic computations is a fundamental, but currently a critical missing element in data-intensive computing.
+
+The 3DPAS theme seeks to understand the landscape of dynamic, distributed, data-intensive computing: the programming models and abstractions, the run-time and middleware services, and the computational infrastructure.
+It will analyze existing tools and services, identify missing pieces and new abstractions, and propose practical solutions and best practices.
+\end{quote}
+
+The theme held three workshops, and in each, application scientists and computing technologists
+were invited to give presentations on their work, and then to be involved in discussions to
+organize and understand the presented materials.  Between the workshops, the theme
+organizers also drafted outlines of a report that was a predecessor of this paper, for
+discussion in the next workshop.  Overall, there were about 50 workshop attendees.
+
+In order to select applications and gather information about
+them, during the various workshop discussions, when an
+application that ``felt'' new was mentioned, we asked the person who
+mentioned it to provide a written set of answers to the following questions about the
+application:
+
+\begin{enumerate}
+\item What is the purpose of the application?
+\item How is the application used to do this?
+\item What infrastructure is used? (including compute, data, network, instruments, etc.)
+\item What dynamic data is used in the application?
+\begin{enumerate}
+\item What are the types of data,
+\item What is the size of the data set(s)?
+\end{enumerate}
+\item How does the application get the data?
+\item What are the time (or quality) constraints on the application?
+\item How much diverse data integration is involved?
+\item How diverse is the data?
+\item Please feel free to also talk about the current state of the
+application, if it exists today, and any specific gaps that you know
+need to be overcome.
+\end{enumerate}
+
+We then used the text of these answers to describe each of the applications (in \S\ref{sec:scenarios}) in terms of:
+
+\begin{itemize}
+\item What does the application do? What problem does it solve?
+\item How does it do it? How does the application run?  What infrastructure does it need/use?
+\item What are the big data aspects of the application?
+\item What are the distributed aspects of the application?
+\item What are the dynamic aspects  of the application?
+\item What else is important about the application?  What does it do well?  Poorly? How was the app written? What tools does it use?
+\end{itemize}
+
+After the workshops, the theme organizers and a set of attendees who
+were interested in participating started discussing how to organize
+and analyze the applications and the technologies, including
+infrastructure, programming systems and abstractions.  These
+participants are the current authors of this paper, and this paper is
+the result of that process.
+
+As part of the theme, two 3DPAS workshops~\cite{3dapas,D3-escience}
+were organized at major conferences: HPDC 2011 and IEEE eScience 2011.
+These workshops had both peer-reviewed and invited papers.
+
+A combination of invitation-only focused discussion meetings and
+broader community engagement events were used to ensure that the
+applications surveyed and ideas captured in this paper are
+representative of current community thinking as well as of
+intellectual relevance.
+
 \end{document}

### 2015-09-12 3dpas_intro.tex
changing order of affiliations

@@
 
-\author{Shantenu Jha\affil{1}\corrauth\, Daniel S. Katz\affil{4,5},
-  Andre~Luckow\affil{1}, Omer Rana\affil{6}, Yogesh Simmhan\affil{7}, Neil Chue
-  Hong\affil{2}, Simon Dobson\affil{3}}
+\author{Shantenu Jha\affil{1}\corrauth\, Daniel S. Katz\affil{2},
+  Andre~Luckow\affil{1}, Omer Rana\affil{3}, Yogesh Simmhan\affil{4}, Neil Chue
+  Hong\affil{5}, Simon Dobson\affil{6}}
 
 \address{\affilnum{1}Rutgers University\break
-\affilnum{2}University of Edinburgh\break
-\affilnum{3}University of St.~Andrews\break
-\affilnum{4}University of Chicago\break
-\affilnum{5}Argonne National Laboratory\break
-\affilnum{6}Cardiff University\break
-\affilnum{7}Indian Institute of Science\break
+\affilnum{2}University of Chicago \& Argonne National Laboratory\break
+\affilnum{3}Cardiff University\break
+\affilnum{4}Indian Institute of Science\break
+\affilnum{5}University of Edinburgh\break
+\affilnum{6}University of St.~Andrews\break
 }

### 2015-09-13 3dpas_intro.tex
removing old comments, adding a new one, fixing a bib entry

@@
   several representative application scenarios, to provide a common conceptual
-  framework to understand them as well as examine the infrastructure used in
+  framework to understand them as well as examining the infrastructure used in
   support of applications.
@@
 
-\ifdraft
-\newpage
-\tableofcontents
+%\ifdraft
 %\newpage
-%\listoftables
-\newpage
-\fi
+%\tableofcontents
+%%\newpage
+%%\listoftables
+%\newpage
+%\fi
 
@@
 
-\jhanote{NCH points out that we are inconsistent in the order of the
-  ``D''s. SJ to fix globally. Good catch Neil!}
+%\jhanote{NCH points out that we are inconsistent in the order of the
+%  ``D''s. SJ to fix globally. Good catch Neil!}
 
@@
 %  data-intensive and dynamic aspects of our study.''}  
-\jhanote{we do not discuss explicitly the 4th D, i.e. diversity (courtesy AL) it
-  is assumed but not discussed}
+%\jhanote{we do not discuss explicitly the 4th D, i.e. diversity (courtesy AL) it
+%  is assumed but not discussed}
 
@@
 
-\jhanote{refine next four paragraph...}
+%\jhanote{refine next four paragraph...}
 
@@
   definition represent the interests and experience of the authors.
+  \katznote{need to add a note about the dates in which application data was collected?  I'm not sure the numbers are correct today, for example, from ATLAS}
 
@@
 
-With ITER~\cite{iter} scheduled to operate before 2020, plasma fusion has become
+With ITER~\cite{iter} scheduled to operate perhaps around 2022, plasma fusion has become
 one of the highest priority research items for the US DOE. Scalability of
@@
 The {\em MapReduce} paradigm~\cite{MapReduce} and its variants~\cite{twister,hmr} are the most
-prominent among these, with recent ones like Pregel and Apace Giraph~\cite{pregel} supporting specific domains like
+prominent among these, with those like Pregel and Apace Giraph~\cite{pregel} supporting specific domains like
 graph analytics. MapReduce forces programmers to model computation in two stages: 
@@
 trends that can be applied to other applications we have not
-considered. \katznote{should this be split into two subsections then? Perhaps Observations \& Trends?}
+considered. 
+%\katznote{should this be split into two subsections then? Perhaps Observations \& Trends?}
 
@@
 often, sensor data is also archived and aggregated for historical
-analyses, which leads to higher data volumes. \katznote{not sure about the `other' part of this...}
+analyses, which leads to higher data volumes. %\katznote{not sure about the `other' part of this...}
 
@@
 
-\jhanote{Note to self: need smoothening}
+%\jhanote{Note to self: need smoothening}
 
@@
 paradigms, similar to what is done in the Power grids application
-(\S\ref{power}). \jhanote{improve description....replace hybrid by
-  bottom-up? }
+(\S\ref{power}). %\jhanote{improve description....replace hybrid by bottom-up? }
 
@@
 
-\katznote{previous 2 paragraphs to be re-read.  Old comments about them
-are available as commented out comments.}
+%\katznote{previous 2 paragraphs to be re-read.  Old comments about them
+%are available as commented out comments.}
 %\jhanote{Irrespective, I find this past paragraph
@@
 
-\section*{Methodology}
+\appendix
+\section*{Appendix: Methodology}
 

### 2015-09-13 3dpas_intro.tex
checkpoint

@@
 The landscape of scientific and enterprise computing is being fundamentally
-altered by the prodigious volumes of data.  Not only are the volumes of data
-produced by scientific applications increasing rapidly, there is also a
-corresponding increase in the amount of data that needs to be processed and
-analyzed.  With the increase in data volumes, types, and sources of data, an
+altered by the prodigious volumes of data.  % Not only are the volumes of data
+% produced by scientific applications increasing rapidly, there is also a
+% corresponding increase in the amount of data that needs to be processed and
+% analyzed.
+With the increase in data volumes, types, and sources of data, an
 important concomitant development has been an increase in the requirement for
@@
 thus the data-intensive nature of the data-intensive science problem is strongly
-coupled with compute-intensive requirements.  As the data are often distributed,
-a consequence is that many attributes of distributed systems permeate to data.
+coupled with compute-intensive requirements.
+
+% As the data are often distributed, a consequence is that many attributes of
+% distributed systems permeate to data.
 

### 2015-09-13 3dpas_intro.tex
further fixes in introduction.

@@
 % analyzed.
-With the increase in data volumes, types, and sources of data, an
-important concomitant development has been an increase in the requirement for
-and complexity of the infrastructure needed to support these requirements.
-Furthermore, data are often naturally distributed and have to be processed in
-situ; occasionally data that are localized have to be distributed in order to be
-processed, either for reasons of performance or for collaborative analytics.
-The scale and significance of varying data source/destination and
-transfer/production rates makes temporal variations important, which in turn
-eventually impact the design and development of infrastructure to support them:
-resource scheduling, data placement, and transfer decisions that were statically
-determinable at small scales, emerge as dynamic and important considerations at
-large scales of data and distribution.
+With the increase in data volumes, types, and sources of data, a concomitant
+development has been an increase in the sophistication and complexity of the
+infrastructure needed to support these requirements. For example, data are often
+naturally distributed and have to be processed in situ; occasionally data that
+are localized have to be distributed in order to be processed, either for
+reasons of performance or for collaborative analytics.  The scale and
+significance of varying data source/destination and transfer/production rates
+makes temporal variations important, which in turn eventually impact the design
+and development of infrastructure to support them: resource scheduling, data
+placement, and transfer decisions that were statically determinable at small
+scales, emerge as dynamic and important considerations at large scales of data
+and distribution.
 
@@
 management and curation of large volumes of data. A critical requirement is to
-make sense of this data, i.e., transform the data into information en route to
-assembling knowledge. Among other things, the transformation from data to
-information requires extensive amounts of computational processing of the data;
-thus the data-intensive nature of the data-intensive science problem is strongly
-coupled with compute-intensive requirements.
+transform the data into information en route to assembling knowledge. Among
+other things, the transformation from data to information requires extensive
+amounts of computational processing of the data; thus the data-intensive nature
+of the data-intensive science problem is strongly coupled with compute-intensive
+requirements.
 
@@
 
-Subjecting a range of otherwise distinct applications to a
-common analytical framework provides advantages to the infrastructure
-scientists and cyberinfrastructure experts, namely suggestions for
-where effort is both needed and likely to yield maximum returns. In
-other words, as D3 applications evolve, our analysis can provide
-guidance on where to focus effort on infrastructure.  Last but not
-least, the advantages of providing a common but extensible vocabulary
-that can be used by the entire community, application end-users,
-developers and infrastructure scientists alike, in an emerging field
-cannot be overstated. One may argue that such vocabulary and
+The focus of this paper is on a subspace of the large set of problems associated
+with data-intensive sciences, namely that which is characterized by {\it both}
+distributed and dynamic aspects. We introduce the concept of Dynamic Distributed
+Data-intensive science and refer to this subspace of data-intensive science as
+D3 science.
+
+Analyzing the applications along an analytical space spanned by the three ``D''s
+of data-intensive, distributed, and dynamic provides a general and effective
+framework.  Subjecting a range of otherwise distinct applications to a common
+analytical framework provides advantages to the infrastructure scientists and
+cyberinfrastructure experts, namely suggestions for where effort is both needed
+and likely to yield maximum returns.  The diverse set of requirements to support
+D3 scenarios points to the need for a careful understanding of D3
+characteristics as well as sophisticated, extensible, and skillfully architected
+infrastructure.
+
+As D3 applications evolve, our analysis can provide guidance on where to focus
+effort on infrastructure.  Last but not least, the advantages of providing a
+common but extensible vocabulary that can be used by the entire community,
+application end-users, developers and infrastructure scientists alike, in an
+emerging field cannot be overstated. One may argue that such vocabulary and
 terminology has been missing from the practice of distributed
@@
 % become increasingly important and often correlated issues.  Thus,
-
-The focus of this paper is on a subspace of the large set of problems associated
-with data-intensive sciences, namely that which is characterized by {\it both}
-distributed and dynamic aspects. We introduce the concept of Dynamic Distributed
-Data-intensive science and refer to this subspace of data-intensive science as
-D3 science.  Analyzing the applications along an analytical space spanned by the
-three ``D''s of data-intensive, distributed, and dynamic provides a general and
-effective framework.  The diverse set of requirements to support D3 scenarios
-points to the need for a careful understanding of D3 characteristics as well as
-sophisticated, extensible, and skillfully architected infrastructure and
-programming systems.
 

### 2015-09-13 3dpas_intro.tex
further refinement in the introduction.

@@
   covered or types of infrastructure and application surveyed.  They by
-  definition represent the interests and experience of the authors.
-  \katznote{need to add a note about the dates in which application data was collected?  I'm not sure the numbers are correct today, for example, from ATLAS}
+  definition represent the interests and experience of the authors. The
+  requirements and barriers of applications cited in this paper specific are
+  mostly representative of the time their representatives were surveyed
+  (2012-2013); this is also true of the quantitative analysis of applications.
+
+  \katznote{need to add a note about the dates in which application data was
+    collected?  I'm not sure the numbers are correct today, for example, from
+    ATLAS}\jhanote{Fixed. CF ATLAS: I'll ask Kaushik/ATLAS to check}
 

### 2015-09-13 3dpas_intro.tex
spell check. partial.

@@
 
-The (grand) challenge in bioinformatics is the need to provide support
+The (grand) challenge in bio-informatics is the need to provide support
 for the analysis of genome sequence data that is becoming available
@@
 own sequencing machines need to deal with large dataset sizes that
-need to be resequenced and reduced for further analysis. Other
+need to be re-sequenced and reduced for further analysis. Other
 researchers may analyze large catalogs of sequenced genome databases
@@
 be employed. This reinforces the need for flexible programming systems
-and associated runtime environments that support collective
+and associated run-time environments that support collective
 (hardware-software) optimization.
@@
 Nowadays, NGS sequencers are even being shipped with servers or small
-clusters for performing the alignment and resequencing as a pipeline
+clusters for performing the alignment and re-sequencing as a pipeline
 right after the reads have been generated~\cite{ilm}.
@@
 Galaxy~\cite{galaxy} (though Galaxy does not support distributed systems currently) have been successful in integrating genome
-analysis tools for bioinformatics. Script based pipelines that combine
-executables are also used often. Very few workflow systems have
+analysis tools for bio-informatics. Script based pipelines that combine
+executable are also used often. Very few workflow systems have
 explicit support for distributed data or dynamic data handling.
@@
 If these sources can be tracked over a few images, an orbit can be determined for them and they can be added to the object catalog.
-Also, the orbit catalog can be updated with redetections of known objects; each rediscovery provides information that improves the known orbits.
+Also, the orbit catalog can be updated with re-detections of known objects; each rediscovery provides information that improves the known orbits.
 
@@
 The application~\cite{modisAzure, modis:ipdps:2010} is used to handle the processes of downloading data from different
-sites and integrating them. The integration consists of reprojecting the data from one map projection to another,
-spatial and temporal resampling, and gap filling. The application itself is a loosely-coupled data
+sites and integrating them. The integration consists of re-projecting the data from one map projection to another,
+spatial and temporal re-sampling, and gap filling. The application itself is a loosely-coupled data
 pipeline that coordinates these processes.
@@
 In addition, scientists can submit Matlab scripts for further custom processing on subsets of
-the data that has been integrated. This includes some summarization processing and plotting
+the data that has been integrated. This includes some summarizing processing and plotting
 routines.  The only time constraints are on the debugging cycle for debugging the scientist-supplied
@@
 that do further custom processing on subsets of the data, which likely
-include some summarization processing (building graphs).  The
+include some summarizing processing (building graphs).  The
 application has been run on a cloud and in an HPC environment.
@@
 
-{\em Bulk Data-parallel} frameworks use ``Bulk'' operators to express computations over 
-collections of data which are executed in a loosely-coupled, data-parallel manner.
-The {\em MapReduce} paradigm~\cite{MapReduce} and its variants~\cite{twister,hmr} are the most
-prominent among these, with those like Pregel and Apace Giraph~\cite{pregel} supporting specific domains like
-graph analytics. MapReduce forces programmers to model computation in two stages: 
-a \emph{map} phase in which a transformation is applied to every element of a collection independently of every other;  and a \emph{reduce} phase in which the transformed results 
-are aggregated through a binary function that is associative and commutative. 
-These properties serve to guarantee that the MapReduce computation can be efficiently 
-parallelized, by performing the map phase concurrently and
-then reducing without the reduction order affecting the final result.
+{\em Bulk Data-parallel} frameworks use ``Bulk'' operators to express
+computations over collections of data which are executed in a loosely-coupled,
+data-parallel manner.  The {\em MapReduce} paradigm~\cite{MapReduce} and its
+variants~\cite{twister,hmr} are the most prominent among these, with those like
+Pregel and Apache Giraph~\cite{pregel} supporting specific domains like graph
+analytics. MapReduce forces programmers to model computation in two stages: a
+\emph{map} phase in which a transformation is applied to every element of a
+collection independently of every other; and a \emph{reduce} phase in which the
+transformed results are aggregated through a binary function that is associative
+and commutative.  These properties serve to guarantee that the MapReduce
+computation can be efficiently parallelized, by performing the map phase
+concurrently and then reducing without the reduction order affecting the final
+result.
 
@@
 the largest volumes of data (O(PB)) typically run on specialized data
-cyberinfrastructure (e.g., the Atlas LHC Grid). Applications such as
+cyberinfrastructure (e.g., the ATLAS LHC Grid). Applications such as
 Fusion and NGS that run on smaller data sets mainly utilize HPC

### 2015-09-14 3dpas_intro.tex
RC1

@@
 \author{Shantenu Jha\affil{1}\corrauth\, Daniel S. Katz\affil{2},
-  Andre~Luckow\affil{1}, Omer Rana\affil{3}, Yogesh Simmhan\affil{4}, Neil Chue
-  Hong\affil{5}, Simon Dobson\affil{6}}
+  Andre~Luckow\affil{1}, Omer Rana\affil{3}, Yogesh Simmhan\affil{4}, \\ Neil Chue
+  Hong\affil{5}}
+
+% \author{Shantenu Jha\affil{1}\corrauth\, Daniel S. Katz\affil{2},
+%   Andre~Luckow\affil{1}, Omer Rana\affil{3}, Yogesh Simmhan\affil{4}, Neil Chue
+%   Hong\affil{5}, Simon Dobson\affil{6}}
+
+%\affilnum{6}University of St.~Andrews\break
 
@@
 \affilnum{5}University of Edinburgh\break
-\affilnum{6}University of St.~Andrews\break
 }
@@
 
-  \katznote{need to add a note about the dates in which application data was
-    collected?  I'm not sure the numbers are correct today, for example, from
-    ATLAS}\jhanote{Fixed. CF ATLAS: I'll ask Kaushik/ATLAS to check}
+  % \katznote{need to add a note about the dates in which application data was
+  %   collected?  I'm not sure the numbers are correct today, for example, from
+  %   ATLAS}\jhanote{Fixed. CF ATLAS: I'll ask Kaushik/ATLAS to check}
 
@@
 Institute Research Theme on Dynamic Distributed Data-Intensive Programming
-Abstractions and Systems. We thank Jon Weissman and Jon Blower for useful
-initial discussions, as well as 3DPAS workshop attendees. Discussions of the
-selected applications benefited from the generous help of Adam Barker, Jon
+Abstractions and Systems. We thank Simon Dobson, Jon Weissman and Jon Blower for
+useful initial discussions, as well as 3DPAS workshop attendees. Discussions of
+the selected applications benefited from the generous help of Adam Barker, Jon
 Blower, Julian Borrill, Silvia Delgado Olabarriaga, Steve Fisher, Keith Jackson,

### 2015-09-15 3dpas_intro.tex
minor changes, undoing some spelling changes

@@
 \author{Shantenu Jha\affil{1}\corrauth\, Daniel S. Katz\affil{2},
-  Andre~Luckow\affil{1}, Omer Rana\affil{3}, Yogesh Simmhan\affil{4}, \\ Neil Chue
+  Andre~Luckow\affil{1}, \\ Omer Rana\affil{3}, Yogesh Simmhan\affil{4}, Neil Chue
   Hong\affil{5}}
@@
 framework.  Subjecting a range of otherwise distinct applications to a common
-analytical framework provides advantages to the infrastructure scientists and
-cyberinfrastructure experts, namely suggestions for where effort is both needed
-and likely to yield maximum returns.  The diverse set of requirements to support
+analytical framework suggests to infrastructure scientists and
+cyberinfrastructure experts where effort is needed
+and is likely to yield maximum returns.  The diverse set of requirements to support
 D3 scenarios points to the need for a careful understanding of D3
@@
 application end-users, developers and infrastructure scientists alike, in an
-emerging field cannot be overstated. One may argue that such vocabulary and
+emerging field cannot be overstated. We believe that such vocabulary and
 terminology has been missing from the practice of distributed
-cyberinfrastructure in general, with less than desirable consequences.
+cyberinfrastructure in general, with negative consequences.
 
@@
 \item This {\it survey} is not meant to be complete, either in scope of domains
-  covered or types of infrastructure and application surveyed.  They by
-  definition represent the interests and experience of the authors. The
-  requirements and barriers of applications cited in this paper specific are
-  mostly representative of the time their representatives were surveyed
+  covered or types of infrastructure and application surveyed.  They
+  represent the interests and experience of the authors. The specific
+  requirements and barriers of applications cited in this paper are
+  representative of the time at which the applications were surveyed
   (2012-2013); this is also true of the quantitative analysis of applications.
@@
 
-The (grand) challenge in bio-informatics is the need to provide support
+The (grand) challenge in bioinformatics is the need to provide support
 for the analysis of genome sequence data that is becoming available
@@
 Galaxy~\cite{galaxy} (though Galaxy does not support distributed systems currently) have been successful in integrating genome
-analysis tools for bio-informatics. Script based pipelines that combine
-executable are also used often. Very few workflow systems have
+analysis tools for bioinformatics. Script-based pipelines that combine
+executables are also often used. Very few workflow systems have
 explicit support for distributed data or dynamic data handling.
@@
 In addition, scientists can submit Matlab scripts for further custom processing on subsets of
-the data that has been integrated. This includes some summarizing processing and plotting
+the data that has been integrated. This includes some summarization processing and plotting
 routines.  The only time constraints are on the debugging cycle for debugging the scientist-supplied
@@
 that do further custom processing on subsets of the data, which likely
-include some summarizing processing (building graphs).  The
+include some summarization processing (building graphs).  The
 application has been run on a cloud and in an HPC environment.

### 2016-02-14 3dpas_intro.tex
Abstract refinement.

@@
 \begin{abstract}
-
   A common feature across many science and engineering applications is the
   amount and diversity of data and computation that must be integrated to yield
-  insights. Often data is increasingly large-scale and distributed. Often the
-  datasets, their locations, availability, analysis and associated scheduling
-  decisions are time-dependent. Collectively these give rise to dynamic
-  distributed data intensive applications.  The characteristics, requirements,
-  and software systems for the analysis of large volumes of ``static'' data have
-  received significant attention; dynamic, distributed and data intensive have
-  received relatively less. This work will ameliorate the state by surveying
-  several representative application scenarios, to provide a common conceptual
-  framework to understand them as well as examining the infrastructure used in
-  support of applications.
-
-%   Applications need access
-%   to these disparate data-sets and to the methods that operate on
-%   them.  Data-intensive applications need first-class support for
-%   distributed heterogeneous platform execution.
-
-% The 3DPAS Research Theme (funded by the eSI, Edinburgh and US NSF)
-% investigated applications that have these characteristics; in other
-% words, it operated at the triple point of  {\it distributed} and {\it dynamic} and
-% {\it data-intensive} (D3) attributes.  This document represents the conclusions
-% of the 3DPAS Theme.
+  insights. Often data is increasingly large-scale and distributed; their
+  location, availability and properties are time-dependent. Collectively these
+  give rise to dynamic distributed data intensive applications. ``Static'' data
+  have received significant attention.  The characteristics, requirements, and
+  software systems for the analysis of large volumes of dynamic, distributed and
+  data intensive have however received relatively less. This paper surveys
+  several representative dynamic distributed data-intensive application
+  scenarios, provides a common conceptual framework to understand them as well
+  as examines the infrastructure used in support of applications.
 

### 2016-02-14 3dpas_intro.tex
introduction - reduced and refined.

@@
 
-The landscape of scientific and enterprise computing is being fundamentally
-altered by the prodigious volumes of data.  % Not only are the volumes of data
+% Not only are the volumes of data
 % produced by scientific applications increasing rapidly, there is also a
@@
 % analyzed.
-With the increase in data volumes, types, and sources of data, a concomitant
-development has been an increase in the sophistication and complexity of the
-infrastructure needed to support these requirements. For example, data are often
-naturally distributed and have to be processed in situ; occasionally data that
+
+The landscape of scientific and enterprise computing is being fundamentally
+altered by the prodigious volumes of data.  Occasionally data are often
+naturally distributed and have to be processed near source; sometimes data that
 are localized have to be distributed in order to be processed, either for
-reasons of performance or for collaborative analytics.  The scale and
-significance of varying data source/destination and transfer/production rates
-makes temporal variations important, which in turn eventually impact the design
-and development of infrastructure to support them: resource scheduling, data
-placement, and transfer decisions that were statically determinable at small
-scales, emerge as dynamic and important considerations at large scales of data
-and distribution.
+reasons of performance or for collaborative analytics. Further, the scale of
+varying data source and destination, as well as the transfer of production rates
+makes temporal variations in data properties increasingly important.  With an
+increase in the importance of distributed and temporal properties of data,
+systems and infrastructure aspects such as resource scheduling, data placement,
+and transfer decisions that were statically determinable at small scales, emerge
+as dynamic and important considerations at large scales of data and
+distribution.
 
@@
 
-The challenges of distributed data-intensive applications span beyond the
-management and curation of large volumes of data. A critical requirement is to
-transform the data into information en route to assembling knowledge. Among
-other things, the transformation from data to information requires extensive
-amounts of computational processing of the data; thus the data-intensive nature
-of the data-intensive science problem is strongly coupled with compute-intensive
-requirements.
+% The challenges of distributed data-intensive applications span beyond the
+% management and curation of large volumes of data. A critical requirement is to
+% transform the data into information en route to assembling knowledge. Among
+% other things, the transformation from data to information requires extensive
+% amounts of computational processing of the data; thus the data-intensive nature
+% of the data-intensive science problem is strongly coupled with compute-intensive
+% requirements.
 
@@
 
-Analyzing the applications along an analytical space spanned by the three ``D''s
-of data-intensive, distributed, and dynamic provides a general and effective
-framework.  Subjecting a range of otherwise distinct applications to a common
-analytical framework suggests to infrastructure scientists and
-cyberinfrastructure experts where effort is needed
-and is likely to yield maximum returns.  The diverse set of requirements to support
+The analytical space spanned by the three ``D''s of data-intensive, distributed,
+and dynamic provides a common vocabulary and effective framework to analyze
+applications and infrastructure alike.  Subjecting a range of otherwise distinct
+applications to a common analytical framework provides an opportunity to search
+for similarities across applications and understand core differences.  The
+diverse % set of requirements to support
 D3 scenarios points to the need for a careful understanding of D3
 characteristics as well as sophisticated, extensible, and skillfully architected
-infrastructure.
-
-As D3 applications evolve, our analysis can provide guidance on where to focus
-effort on infrastructure.  Last but not least, the advantages of providing a
-common but extensible vocabulary that can be used by the entire community,
-application end-users, developers and infrastructure scientists alike, in an
-emerging field cannot be overstated. We believe that such vocabulary and
-terminology has been missing from the practice of distributed
-cyberinfrastructure in general, with negative consequences.
+infrastructure.  A common framework thus informs infrastructure scientists and
+cyberinfrastructure experts where effort is likely to yield maximum returns.
+The advantages of providing a common but extensible vocabulary that can be used
+by the entire community, application end-users, developers and infrastructure
+scientists alike, in an emerging field cannot be overstated. We believe that
+such vocabulary and terminology has been missing from the practice of
+distributed cyberinfrastructure in general, with negative consequences.
+
+% As D3 applications
+% evolve, our analysis can provide guidance on where to focus effort on
+% infrastructure.
+
+% Analyzing the applications along an analytical space spanned by the three ``D''s
+% of data-intensive, distributed, and dynamic provides a general and effective
+% framework.  Subjecting a range of otherwise distinct applications to a common
+% analytical framework suggests to infrastructure scientists and
+% cyberinfrastructure experts where effort is needed and is likely to yield
+% maximum returns.  The diverse set of requirements to support D3 scenarios points
+% to the need for a careful understanding of D3 characteristics as well as
+% sophisticated, extensible, and skillfully architected infrastructure.
 

### 2016-02-14 3dpas_intro.tex
Do miracles happen?

@@
 for similarities across applications and understand core differences.  The
-diverse % set of requirements to support
-D3 scenarios points to the need for a careful understanding of D3
+diverse D3 scenarios points to the need for a careful understanding of D3
 characteristics as well as sophisticated, extensible, and skillfully architected

### 2016-02-14 3dpas_intro.tex
ongoing work in the conclusion section.

@@
 
-This section brings together some key observations
-from the applications and infrastructure
-analysis carried out in this paper.
 %as take away messages.
-Section~\ref{sec:discussion}
-provides a discussion of dynamic and distributed characteristics of
-data associated with the applications we analyze in this work,
-attempting to infer more general characteristics that %can be applied
-may apply
-to other similar applications.  We also suggest two metrics that may
-suggest two metrics that may
-be associated with such applications, primarily as a basis to compare
-how compute and data processing (in terms of their number and
-distribution) may be used to find common features between them. Such
-feature analysis is useful, we believe, for two main reasons: (i) to
-identify similarities between the applications covered in
-\S\ref{sec:scenarios} with others that are in the design phase or
-under development, so that common requirements can be identified; and
-(ii) to identify gaps in current software tools and programming
-systems that such applications use, thereby leading to the
-development of new tools and systems. As our application and
-infrastructure coverage is, by necessity, limited, identifying common
-features also enables the lessons learned to be generalized and
-applied more widely. We subsequently discuss a set of architectures
-for D3 science, and pose some questions we believe are pertinent as
-the community begins considering the next-generation
-architectures for D3 applications. In
-\S\ref{sec:conclusion} we conclude by providing a number of
-observations to inform applications and
+
+This section brings together some key observations from the applications and
+infrastructure analysis carried out in this paper.  Section~\ref{sec:discussion}
+provides a discussion of dynamic and distributed characteristics of data
+associated with the applications we analyze in this work, attempting to infer
+more general characteristics that %can be applied
+may apply to other similar applications.  We also suggest two metrics that may
+suggest two metrics that may be associated with such applications, primarily as
+a basis to compare how compute and data processing (in terms of their number and
+distribution) may be used to find common features between them. Such feature
+analysis is useful, we believe, for two main reasons: (i) to identify
+similarities between the applications covered in \S\ref{sec:scenarios} with
+others that are in the design phase or under development, so that common
+requirements can be identified; and (ii) to identify gaps in current software
+tools and programming systems that such applications use, thereby leading to the
+development of new tools and systems. As our application and infrastructure
+coverage is, by necessity, limited, identifying common features also enables the
+lessons learned to be generalized and applied more widely. We subsequently
+discuss a set of architectures for D3 science, and pose some questions we
+believe are pertinent as the community begins considering the next-generation
+architectures for D3 applications. In \S\ref{sec:conclusion} we conclude by
+providing a number of observations to inform applications and
 cyberinfrastructure developers, based on common trends found in our study.
@@
 
-In this section we attempt to summarize four key themes from previous
-sections: (i) the source and nature of dynamic data that we observe in
-the application surveyed in \S\ref{sec:scenarios}; (ii) the types of
-data distribution that we observe in such applications; (iii) the
-general characteristics that can be associated with D3 science
-applications; (iv) the characteristics of a computational
-infrastructure that may be most suited for D3 science
-applications. Whereas (i) and (ii) are primarily observations that we
-can make on existing applications that we have discussed in
-\S\ref{sec:scenarios}, (iii) and (iv) attempt to synthesize general
-trends that can be applied to other applications we have not
+We summarize four key themes from previous sections: (i) the source and nature
+of dynamic data observed in the applications surveyed in \S\ref{sec:scenarios};
+(ii) the types of data distribution observed in such applications; (iii) the
+general characteristics that can be associated with D3 science applications;
+(iv) the characteristics of a computational infrastructure most suited for D3
+science applications.  Whereas (i) and (ii) are observations that can be made on
+existing applications discussed in \S\ref{sec:scenarios}, (iii) and (iv) attempt
+to synthesize general trends that can be applied to applications we have not
 considered.
+
 %\katznote{should this be split into two subsections then? Perhaps Observations \& Trends?}

### 2016-02-14 3dpas_intro.tex
more refinement.

@@
 
-This section brings together some key observations from the applications and
-infrastructure analysis carried out in this paper.  Section~\ref{sec:discussion}
-provides a discussion of dynamic and distributed characteristics of data
-associated with the applications we analyze in this work, attempting to infer
-more general characteristics that %can be applied
-may apply to other similar applications.  We also suggest two metrics that may
-suggest two metrics that may be associated with such applications, primarily as
-a basis to compare how compute and data processing (in terms of their number and
-distribution) may be used to find common features between them. Such feature
+This section brings together some key observations from the analysis of
+applications and infrastructure carried out in this paper.
+Section~\ref{sec:discussion} provides a discussion of dynamic and distributed
+characteristics of data associated with a set of applications.
+
+% we analyze in this work, attempting to infer more general characteristics
+% that %can be applied
+% may apply to other similar applications.
+%suggest two metrics that may
+
+We suggest two metrics that may be associated with such applications, primarily
+as a basis to compare how compute and data processing (in terms of their number
+and distribution) may be used to find common features between them. Such feature
 analysis is useful, we believe, for two main reasons: (i) to identify
@@
 tools and programming systems that such applications use, thereby leading to the
-development of new tools and systems. As our application and infrastructure
-coverage is, by necessity, limited, identifying common features also enables the
-lessons learned to be generalized and applied more widely. We subsequently
-discuss a set of architectures for D3 science, and pose some questions we
-believe are pertinent as the community begins considering the next-generation
-architectures for D3 applications. In \S\ref{sec:conclusion} we conclude by
-providing a number of observations to inform applications and
-cyberinfrastructure developers, based on common trends found in our study.
-
-An important motivation for this work is to capture the state of the
-art in the design and development of distributed dynamic applications
-and the associated infrastructure currently used for their
-execution. We have aimed to try to discern common
-approaches and challenges, and to identify any obvious gaps (or
-opportunities). Using a common vocabulary and terminology
-to describe otherwise distinct and unrelated applications allows
-potential application
-users/developers to benefit from the insight that a common
-analytical framework provides. For example, seeing how
-applications with similar patterns (e.g., pipelines) have addressed
-issues of dynamism and distribution enables common solutions and
-facilitates the emergence of best practices even in the absence of formal
-methods.  Similarly, our analysis sends a useful message to domain and
-application scientists that the challenges and barriers they face are not
-unique even if current solutions require a level of customization.
+development of new tools and systems.
+
+As our application and infrastructure coverage is, by necessity, limited,
+identifying common features also enables the lessons learned to be generalized
+and applied more widely. We subsequently discuss a set of architectures for D3
+science, and pose some questions we believe are pertinent as the community
+begins considering the next-generation architectures for D3 applications. In
+\S\ref{sec:conclusion} we conclude by providing a number of observations to
+inform applications and cyberinfrastructure developers, based on common trends
+found in our study.
 
@@
 sometimes it occurs because of changing experimental or analytic conditions,
-sometimes due to the availability of infrastructure and resources. In this
-paper, we have identified many different applications characteristics,
-programming systems, and infrastructure techniques that either support dynamic
-data or produce it.
+sometimes due to the availability of infrastructure and resources.
+
+An important motivation for this work is to capture the state of the art in the
+design and development of distributed dynamic applications and the associated
+infrastructure currently used for their execution. We have aimed to try to
+discern common approaches and challenges, and to identify any obvious gaps (or
+opportunities). Using a common vocabulary and terminology to describe otherwise
+distinct and unrelated applications allows potential application
+users/developers to benefit from the insight that a common analytical framework
+provides. For example, seeing how applications with similar patterns (e.g.,
+pipelines) have addressed issues of dynamism and distribution enables common
+solutions and facilitates the emergence of best practices even in the absence of
+formal methods.  Similarly, our analysis sends a useful message to domain and
+application scientists that the challenges and barriers they face are not unique
+even if current solutions require a level of customization.  In this paper, we
+have identified many different applications characteristics, programming
+systems, and infrastructure techniques that either support dynamic data or
+produce it.
 

### 2016-02-14 3dpas_intro.tex
refinements up to 5.1.1

@@
 
-The applications in \S\ref{sec:scenarios} show that the dynamic
-characteristics of many applications arise in two ways, due to data
-generation and placement or due to the coordination
-mechanisms of the application.
-The first of these refers to data dynamism and the second to
-application dynamism (outlined in \S\ref{sec:dynTypes}).
+The applications in \S\ref{sec:scenarios} show that the dynamic characteristics
+of many applications arise in two ways: either due to data generation and
+placement, or due to the certain aspects of the application.  The first
+of these refers to data dynamism and the second to application dynamism
+(outlined in \S\ref{sec:dynTypes}).
 
@@
 observed. Often this data is large in size and must be placed close to
-the user undertaking the analysis. In some instances, it may not be
+the analysis. In some instances, it may not be
 known where (and when) an analysis request may be made, and it is
@@
 interactions between execution units and data sources to a fully
-adaptable interaction using multi-agent systems.  As outlined in the
+adaptable interaction.  As outlined in the
 Industrial Incident Notification and Response application in

### 2016-02-14 3dpas_intro.tex
ongoing fixes.

@@
 
-From \S\ref{sec:distTypes} and
-\figurename~\ref{fig:figures_application}, we find that applications
-with distributed data differ based on the amount or volume of data
-distributed, the degree of distribution (scattered over one site or
-many sites), and the dynamic aspects associated with distribution,
-e.g., latency and relative scheduling flexibility (and thus tolerance
-or constraints on data transfer/aggregation).  Applications processing
-the largest volumes of data (O(PB)) typically run on specialized data
-cyberinfrastructure (e.g., the ATLAS LHC Grid). Applications such as
-Fusion and NGS that run on smaller data sets mainly utilize HPC
-environments. On the other end of the spectrum are highly dynamic
-applications that process sensor data and make near-realtime
-decisions based on streaming data, e.g., the power grid and sensor
-network applications. These applications process much less data;
-often, sensor data is also archived and aggregated for historical
-analyses, which leads to higher data volumes. %\katznote{not sure about the `other' part of this...}
+From \S\ref{sec:distTypes} and \figurename~\ref{fig:figures_application}, we
+find that applications with distributed data vary in the amount or volume of
+data distributed, the degree of distribution (scattered over one site or many
+sites), and the dynamic aspects associated with distribution, e.g., latency and
+relative scheduling flexibility (and thus tolerance or constraints on data
+transfer/aggregation).  Applications processing the largest volumes of data
+(O(PB)) typically run on specialized data cyberinfrastructure (e.g., the ATLAS
+LHC Grid). Applications such as Fusion and NGS that run on smaller data sets
+mainly utilize HPC environments. On the other end of the spectrum are highly
+dynamic applications that process sensor data and make near-realtime decisions
+based on streaming data, e.g., the power grid and sensor network applications.
+
+% These applications process much less data;
+% often, sensor data is also archived and aggregated for historical
+% analyses, which leads to higher data volumes. %\katznote{not sure about the `other' part of this...}
 
@@
 %application in which data can be distributed.
-Because application stages often have a distribution that makes sense
-to that stage, and because that distribution may not be appropriate for all other
-phases, applications must either use one particular distribution
-that may not work well in all phases or transform the data on input to
-some stages.  In particular, infrastructural applications
-require a specific distribution for the data storage, and thus
-the stages that read and process the data will need to make this choice.
-
-As seen, how application data is distributed depends upon which stage
-of an application is being considered. In spite of such variation,
-{\it data topology} can be used to identify the flow of data through
-the application, with nodes representing execution units and edges
-representing communication features.
+
+% Because application stages often have a distribution that makes sense to that
+% stage, and because that distribution may not be appropriate for other
+% phases, applications must either use one particular distribution that may not
+% work well in all phases or transform the data on input to some stages.
+
+Applications must either use one particular distribution, or transform data on
+input to some stages. In particular, infrastructural applications require a
+specific distribution for the data storage, and thus the stages that read and
+process the data will need to make this choice.
+
+As seen, how application data is distributed depends upon which stage of an
+application is being considered. In spite of such variation, {\it data topology}
+can be used to identify the flow of data through the application, with nodes
+representing execution units and edges representing communication features.
+Additionally, how the application reacts to and handles dynamism is an important
+factor and may also indicate that some aspects of the data topology are unknown.
+
 % The overall shape of such a data connectivity pattern gives a sense
@@
 %   paper.}
-Additionally, how the application reacts to and handles dynamism is an
-important factor and may also indicate that some aspects of the data
-topology are unknown.
 

### 2016-02-14 3dpas_intro.tex
ongoing.

@@
 
-Applications must either use one particular distribution, or transform data on
-input to some stages. In particular, infrastructural applications require a
-specific distribution for the data storage, and thus the stages that read and
-process the data will need to make this choice.
-
-As seen, how application data is distributed depends upon which stage of an
-application is being considered. In spite of such variation, {\it data topology}
-can be used to identify the flow of data through the application, with nodes
-representing execution units and edges representing communication features.
-Additionally, how the application reacts to and handles dynamism is an important
-factor and may also indicate that some aspects of the data topology are unknown.
-
-% The overall shape of such a data connectivity pattern gives a sense
-% of {\it distributedness}; the thickness of edges gives a sense of
-% how data intensive the application is; and the change in the shape
-% and thickness (and rate) and number of edges gives a sense of the
-% dynamism of the application. \katznote{it seems odd that we describe
-%   a picture here but don't actually have any such picture in the
-%   paper.}
-
-One aspect of managing data topology is ``data planning," which refers
-to how the data is coordinated from ingest through analysis to
-output. In all cases, there is a coordination of the data with code
-used to analyze or transform it. In some cases these form clear,
-integrated stages where data flows from one to another. In others the
-way that data enters and flows through the application is more loosely
-defined and the execution of code is independent. Some examples
-include the ATLAS experiments, which have a single point of primary
-data generation which fans out to multiple analysis sites followed by
-mass exchange of derived data, and sensor grids, which have multiple
-sources of data generation collected to a single point of analysis
+% Applications must either use one particular distribution, or transform data on
+% input to some stages. In particular,
+
+% Infrastructural applications require a specific distribution for the data
+% storage, and thus  stages that read and process the data will need to make
+% this choice.  As seen,
+
+{\it Data topology} can be used to identify the flow of data through the
+application, with nodes representing execution units and edges representing
+communication features.  One aspect of managing data topology is ``data
+planning," which refers to how the data is coordinated from ingest through
+analysis to output. In all cases, there is a coordination of the data with code
+used to analyze or transform it. In some cases these form clear, integrated
+stages where data flows from one to another. In others the way that data enters
+and flows through the application is more loosely defined and the execution of
+code is independent. Some examples include the ATLAS experiments, which have a
+single point of primary data generation which fans out to multiple analysis
+sites followed by mass exchange of derived data, and sensor grids, which have
+multiple sources of data generation collected to a single point of analysis
 followed by constrained point of reuse.
 
-We observe that pipelines provide a useful abstraction to represent
-the application stages identified in
-\figurename~\ref{fig:figures_application-stages}.  Many workflow tools
-also enable simple data capture and processing pipelines to be
-defined, where each pipeline stage may range from a complete
-application to a subroutine in Matlab (for instance).  It is useful to
-note that pipelines may also be dynamically constructed and adapted
-using context and monitoring information, as undertaken in the
-Industrial Incident Notification and Response application studied in
-this paper. In this instance, the actual pipeline stage to be invoked
-is triggered by the availability of particular types of data, a
-feature also relevant in the Power Grids application.  However, it is
-useful to note that different stages of the pipeline receive different
-emphases in different applications. Some applications must collect and
-store all incoming data, with only minor pre-processing to clean the
-data ahead of the main processing. For others, a major constraint is
-to reduce the incoming data stream as quickly as possible, perhaps
-discarding a large fraction of the data before any processing
-occurs. Reuse of pipeline stages across different applications must
-therefore be undertaken with care.
+% We observe that pipelines provide a useful abstraction to represent
+% the application stages identified in
+% \figurename~\ref{fig:figures_application-stages}.  Many workflow tools
+% also enable simple data capture and processing pipelines to be
+% defined, where each pipeline stage may range from a complete
+% application to a subroutine in Matlab (for instance).  It is useful to
+% note that pipelines may also be dynamically constructed and adapted
+% using context and monitoring information, as undertaken in the
+% Industrial Incident Notification and Response application studied in
+% this paper. In this instance, the actual pipeline stage to be invoked
+% is triggered by the availability of particular types of data, a
+% feature also relevant in the Power Grids application.  However, it is
+% useful to note that different stages of the pipeline receive different
+% emphases in different applications.
+
+Some applications must collect and store all incoming data, with only minor
+pre-processing to clean the data ahead of the main processing. For others, a
+major constraint is to reduce the incoming data stream as quickly as possible,
+perhaps discarding a large fraction of the data before any processing
+occurs.
 

### 2016-02-14 3dpas_intro.tex
checkpoint.

@@
 relative scheduling flexibility (and thus tolerance or constraints on data
-transfer/aggregation).  Applications processing the largest volumes of data
-(O(PB)) typically run on specialized data cyberinfrastructure (e.g., the ATLAS
-LHC Grid). Applications such as Fusion and NGS that run on smaller data sets
-mainly utilize HPC environments. On the other end of the spectrum are highly
+transfer/aggregation).  
+% Applications processing the largest volumes of data typically run on
+% specialized data cyberinfrastructure (e.g., the ATLAS LHC Grid). Applications
+% such as Fusion and NGS that run on smaller data sets mainly utilize HPC
+% environments.
+
+Some applications use specialized data cyberinfrastructure, while some run
+utilize existing HPC environments.  On the other end of the spectrum are highly
 dynamic applications that process sensor data and make near-realtime decisions
@@
 major constraint is to reduce the incoming data stream as quickly as possible,
-perhaps discarding a large fraction of the data before any processing
-occurs.
+perhaps discarding a large fraction of the data before any processing occurs.
 
@@
 denominator, and that it does not capture the time scales over
-which data changes. However, we believe it captures most of the
-non-temporal {\it dynamic} properties.
+which data changes.
+
+% However, we believe it captures most of the non-temporal {\it dynamic}
+% properties.
 
@@
 
-A survey of data-intensive distributed applications on production
-cyberinfrastructure leads us to posit that there are three primary
-macroscopic architectures~\cite{pilot_clouds} and approaches for distributed
-data-intensive applications: (i) localize all data into a ``cloud''
-(or a large analytical engine) -- the paradigm adopted by several
-genome projects and the cancer atlas, (ii) decompose and distribute
-data to an appropriate number of computing/analytical engines as
-available -- the paradigm employed by particle physics for the
-discovery of the Higgs Boson (similar to the application in
-\S\ref{WLCGSteve}), and (iii) a hybrid of the above two paradigms,
-wherein data is decomposed and committed to several infrastructure,
-which in turn could be a combination of either of the first two
-paradigms, similar to what is done in the Power grids application
+An investigation of data-intensive distributed applications on production
+cyberinfrastructure, leads us to believe that there are three primary
+macroscopic architectures and approaches for distributed data-intensive
+applications: (i) localize all data into a ``cloud'' (or a large analytical
+engine) -- the paradigm adopted by several genome projects and the cancer atlas,
+(ii) decompose and distribute data to an appropriate number of
+computing/analytical engines as available -- the paradigm employed by particle
+physics for the discovery of the Higgs Boson (similar to the application in
+\S\ref{WLCGSteve}), and (iii) a hybrid of the above two paradigms, wherein data
+is decomposed and committed to several infrastructure, which in turn could be a
+combination of either of the first two paradigms, similar to what is done in the
+Power grids application
 (\S\ref{power}). %\jhanote{improve description....replace hybrid by bottom-up? }
 
-It is obvious that the first architecture can be used for large
-volumes of data, but surely there are limitations to the scalability
-or validity of this model. How does this limitation change,
-qualitatively or quantitatively, if dynamic data and/or applications
-are considered? This begets the question: how fundamental an issue is
-{\it dynamic}?
+It is obvious that the first architecture can be used for large volumes of data,
+but there are self-evident limitations to the scalability or validity of this
+model. How does this limitation change, qualitatively or quantitatively, if
+dynamic data and/or applications are considered? This begets the question: how
+fundamental an issue is {\it dynamic}?
 
@@
 place. What lessons does the HEP experience teach us for the other
-communities, such as the bioinformatics?  Can an infrastructure
+communities, such as bioinformatics?  Can an infrastructure
 conforming to a particular architecture and developed for a specific
@@
 
+\newpage
 \appendix

### 2016-02-14 3dpas_intro.tex
checkpoint.
updated abstract

@@
   amount and diversity of data and computation that must be integrated to yield
-  insights. Often data is increasingly large-scale and distributed; their
-  location, availability and properties are time-dependent. Collectively these
-  give rise to dynamic distributed data intensive applications. ``Static'' data
-  have received significant attention.  The characteristics, requirements, and
-  software systems for the analysis of large volumes of dynamic, distributed and
-  data intensive have however received relatively less. This paper surveys
+  insights. Often data are increasingly large-scale and distributed; and their
+  location, availability and properties are time-dependent. Collectively, these
+  characteristics give rise to dynamic distributed data intensive applications. While ``static'' data
+  application have received significant attention, the characteristics, requirements, and
+  software systems for the analysis of large volumes of dynamic, distributed, and
+  data intensive have received relatively less. This paper surveys
   several representative dynamic distributed data-intensive application
-  scenarios, provides a common conceptual framework to understand them as well
-  as examines the infrastructure used in support of applications.
+  scenarios, provides a common conceptual framework to understand them, and
+  examines the infrastructure used in support of applications.
 

### 2016-02-14 3dpas_intro.tex
light edits in intro and conclusions

@@
 The landscape of scientific and enterprise computing is being fundamentally
-altered by the prodigious volumes of data.  Occasionally data are often
-naturally distributed and have to be processed near source; sometimes data that
+altered by prodigious volumes of data.  Occasionally data are often
+naturally distributed and have to be processed near their source, and sometimes data that
 are localized have to be distributed in order to be processed, either for
 reasons of performance or for collaborative analytics. Further, the scale of
-varying data source and destination, as well as the transfer of production rates
+varying data source and destination, as well as the transfer of production rates \katznote{what are `transfer of production rates'?}
 makes temporal variations in data properties increasingly important.  With an
@@
 systems and infrastructure aspects such as resource scheduling, data placement,
-and transfer decisions that were statically determinable at small scales, emerge
+and transfer decisions that were statically determinable at small scales emerge
 as dynamic and important considerations at large scales of data and
@@
 
-The analytical space spanned by the three ``D''s of data-intensive, distributed,
-and dynamic provides a common vocabulary and effective framework to analyze
+The analytical space spanned by the three ``D''s of dynamic, distributed,
+and data-intensive provides a common vocabulary and effective framework to analyze
 applications and infrastructure alike.  Subjecting a range of otherwise distinct
@@
 for similarities across applications and understand core differences.  The
-diverse D3 scenarios points to the need for a careful understanding of D3
+diverse D3 scenarios point to the need for a careful understanding of D3
 characteristics as well as sophisticated, extensible, and skillfully architected
@@
 The advantages of providing a common but extensible vocabulary that can be used
-by the entire community, application end-users, developers and infrastructure
-scientists alike, in an emerging field cannot be overstated. We believe that
-such vocabulary and terminology has been missing from the practice of
+by the entire community---application end-users, developers and infrastructure
+scientists alike---in an emerging field cannot be overstated. We believe that
+such a vocabulary and terminology has been missing from the practice of
 distributed cyberinfrastructure in general, with negative consequences.
@@
 are motivated by an attempt to understand infrastructure---existing
-capabilities, trends, and limitations---available to the complexity,
+capabilities, trends, and limitations---available to support the complexity,
 challenges and characteristics of dynamic and distributed data at
@@
 
-We would like the reader to keep in mind some context and scope-setting points:
+We would like the reader to keep in mind some points regarding context and scope:
 
@@
 
-\item This {\it survey} is not meant to be complete, either in scope of domains
-  covered or types of infrastructure and application surveyed.  They
+\item This {\it survey} is not meant to be complete, either in the scope of domains
+  covered or the types of infrastructure and application surveyed.  It
   represent the interests and experience of the authors. The specific
@@
 
-\item The focus is on applications and infrastructure that arise from science
-  and engineering projects in academia; this is not exclusive but is more a
+\item The focus is generally on applications and infrastructure that arise from science
+  and engineering projects in academia in general; this is a
   guiding principle.  While we are cognizant of enterprise applications and
@@
   interoperability influence the solutions employed and infrastructures used; it
-  is important to therefore focus on problems that emerge as in the space
+  is important to therefore focus on problems that emerge in the space
   defined by these ``academic constraints''.
@@
   of data and/or compute across multiple data centers.  Also, our focus is
-  primarily on infrastructure considerations at the large-scale, as searching
-  for common solutions and patterns at the large-scale is likely to be more
-  fruitful than at small-scales where customized tools and optimization of
+  primarily on infrastructure considerations at  large scale, as searching
+  for common solutions and patterns at  large scale is likely to be more
+  fruitful than at small scales where customized tools and optimization of
   low-level systems features is likely to return greater yield.
@@
 
-The rest of this paper consists of
+The remainder of this paper consists of
 the application scenarios (\S\ref{sec:scenarios}),
@@
 The growth in data volumes and its importance is having profound implications on
-the way applications are designed to the way we develop infrastructure and
+the way applications are designed and the way we develop infrastructure and
 provision associated services.  Many fields, such as biology, that until
@@
 distinct and unrelated applications allows potential application
-users/developers to benefit from the insight that a common analytical framework
+users\slash developers to benefit from the insights that a common analytical framework
 provides. For example, seeing how applications with similar patterns (e.g.,

### 2016-02-22 3dpas_intro.tex
addressing DSK's comments.

@@
 The landscape of scientific and enterprise computing is being fundamentally
-altered by prodigious volumes of data.  Occasionally data are often
-naturally distributed and have to be processed near their source, and sometimes data that
+altered by prodigious volumes of data.  Occasionally data are often naturally
+distributed and have to be processed near their source, and sometimes data that
 are localized have to be distributed in order to be processed, either for
-reasons of performance or for collaborative analytics. Further, the scale of
-varying data source and destination, as well as the transfer of production rates \katznote{what are `transfer of production rates'?}
-makes temporal variations in data properties increasingly important.  With an
-increase in the importance of distributed and temporal properties of data,
-systems and infrastructure aspects such as resource scheduling, data placement,
-and transfer decisions that were statically determinable at small scales emerge
-as dynamic and important considerations at large scales of data and
-distribution.
+reasons of performance or for collaborative analytics.  Further, the variation
+of data production rates, data source and destination makes temporal variations
+in data properties increasingly important.  With an increase in the importance
+of distributed and temporal properties of data, systems and infrastructure
+aspects such as resource scheduling, data placement, and transfer decisions that
+were statically determinable at small scales emerge as dynamic and important
+considerations at large scales of data and distribution.
 
@@
 
-Nowadays, NGS sequencers are even being shipped with servers or small
-clusters for performing the alignment and re-sequencing as a pipeline
-right after the reads have been generated~\cite{ilm}.
 %\alnote{What kind of codes are used
@@
 %
-Scientific workflow systems like Taverna~\cite{Taverna:2006}, BioKepler~\cite{biokepler:2011}, and
-Galaxy~\cite{galaxy} (though Galaxy does not support distributed systems currently) have been successful in integrating genome
-analysis tools for bioinformatics. Script-based pipelines that combine
-executables are also often used. Very few workflow systems have
-explicit support for distributed data or dynamic data handling.
-Therefore reusable approaches at extending existing tools to support
-distributed and dynamic data are likely to be useful.
+
+NGS sequencers such Illumina and Ion Torrent provide the option of onsite
+servers or small clusters for performing the alignment and re-sequencing as a
+pipeline right after the reads have been generated.  Scientific workflow systems
+like Taverna~\cite{Taverna:2006}, BioKepler~\cite{biokepler:2011}, and
+Galaxy~\cite{galaxy} (though Galaxy does not support distributed systems
+currently) have been successful in integrating genome analysis tools for
+bioinformatics. Script-based pipelines that combine executables are also often
+used. Very few workflow systems have explicit support for distributed data or
+dynamic data handling.  Therefore reusable approaches at extending existing
+tools to support distributed and dynamic data are likely to be useful.
 %\katznote{Shantenu/Yogesh/Andre, please read this paragraph and tell me if changes
@@
 
-The theme held three workshops, and in each, application scientists and computing technologists
-were invited to give presentations on their work, and then to be involved in discussions to
-organize and understand the presented materials.  Between the workshops, the theme
-organizers also drafted outlines of a report that was a predecessor of this paper, for
-discussion in the next workshop.  Overall, there were about 50 workshop attendees.
+The theme held three workshops, and in each, application scientists and
+computing technologists were invited to give presentations on their work, and
+then to be involved in discussions to organize and understand the presented
+materials.  Between the workshops, the theme organizers also drafted outlines of
+a report that was a predecessor of this paper, for discussion in the next
+workshop.  Overall, there were about 50 workshop attendees.
 

### 2016-02-23 3dpas_intro.tex
v minor in abstract.
minor tweak in acknowledgement.

@@
   location, availability and properties are time-dependent. Collectively, these
-  characteristics give rise to dynamic distributed data intensive applications. While ``static'' data
-  application have received significant attention, the characteristics, requirements, and
-  software systems for the analysis of large volumes of dynamic, distributed, and
-  data intensive have received relatively less. This paper surveys
-  several representative dynamic distributed data-intensive application
-  scenarios, provides a common conceptual framework to understand them, and
-  examines the infrastructure used in support of applications.
+  characteristics give rise to dynamic distributed data intensive
+  applications. While ``static'' data application have received significant
+  attention, the characteristics, requirements, and software systems for the
+  analysis of large volumes of dynamic, distributed, and data intensive have
+  received relatively less. This paper surveys several representative dynamic
+  distributed data-intensive application scenarios, provides a common conceptual
+  framework to understand them, and examines the infrastructure used in support
+  of applications.
 
@@
 \section*{Acknowledgements}
-This work was sponsored by NSF OCI-1059635 and made possible by the UK e-Science
-Institute Research Theme on Dynamic Distributed Data-Intensive Programming
-Abstractions and Systems. We thank Simon Dobson, Jon Weissman and Jon Blower for
+The workshop was sponsored by NSF OCI-1059635 and made possible by the UK
+e-Science Institute Research Theme on Dynamic Distributed Data-Intensive
+Programming Abstractions and Systems.  SJ contribution was also funded by NSF
+CAREER ACI-1253644.  We thank Simon Dobson, Jon Weissman and Jon Blower for
 useful initial discussions, as well as 3DPAS workshop attendees. Discussions of

### 2016-07-22 3dpas_intro.tex
adding rebuttal document - probably should be turned in .tex later made changes for reviewer 1, and indicated that we have so in the rebuttal

--- 2016-02-22 3dpas_intro.tex
@@
 input data needed and some output data produced.  Applications in this
-category are: NGS Analytics (\S\ref{bioSilvia}), CMB
+category are: NGS (next generation sequencing) Analytics (\S\ref{bioSilvia}), CMB (cosmic microwave background)
 (\S\ref{astroJulian}), Fusion (\S\ref{fusionScott}), Industrial
-Incident Notification and Response (\S\ref{Kees}), MODIS Data
+Incident Notification and Response (\S\ref{Kees}), MODIS (moderate resolution imaging spectroradiometer) Data
 Processing (\S\ref{modisKeith}), and Distributed Network Intrusion
@@
 be run by different groups that do not frequently interact.  This type
-of application includes: ATLAS/WLCG (\S\ref{WLCGSteve}), LSST
-(\S\ref{astroAdam}), SOA Astronomy (\S\ref{astroSOAAdam}), Sensor
+of application includes: ATLAS/WLCG (a toroidal large hadron collider apparatus/worldwide large hadron collider computing grid, \S\ref{WLCGSteve}), LSST
+(large synoptic survey telescope, \S\ref{astroAdam}), SOA (service-oriented architecture) Astronomy (\S\ref{astroSOAAdam}), Sensor
 Network Application (\S\ref{sensorSimon}), Climate
@@
 application, we believe a term analogous to the use of Amdahl's
-number~\cite{10.1109/ICDE.2000.839382} could be used to characterize `static'
-data-intensive applications and systems:
-
+number~\cite{10.1109/ICDE.2000.839382} such as 
 \begin{itemize}
-\item[] $\alpha$ = Num. of Compute Operations / Amount of Data
+\item[] Num. of Compute Operations / Amount of Data
   (Consumed, Generated or Transformed)
 \end{itemize}
+  could be used to characterize `static'
+data-intensive applications and systems:
+
+%\begin{itemize}
+%\item[] $\alpha$ = Num. of Compute Operations / Amount of Data
+%  (Consumed, Generated or Transformed)
+%\end{itemize}
 
@@
 
-Similarly, another term can be used to quantify the
+Similarly, another term 
+\begin{itemize}
+\item[] Amount of Data Transferred / Degree of
+  Distribution
+\end{itemize}
+could be used to quantify the
 ``scale'' of distribution, namely the degree of distribution as a
@@
 
-\begin{itemize}
-\item[] $\beta$ = Amount of Data Transferred / Degree of
-  Distribution
-% \item[] $\beta_2$ = Number of Compute Operations / Degree of
+%\begin{itemize}
+%\item[] $\beta$ = Amount of Data Transferred / Degree of
 %  Distribution
-\end{itemize}
+%\end{itemize}
 
@@
 ``class of applications'' as opposed to a specific application. We
-believe, nonetheless, that these measures suggest a possible way to
+believe, nonetheless, that these measures (which we refer to as $\alpha$ and $\beta$ respectively)
+suggest a possible way to
 quantify an application's dynamism and distribution and thereby find

### 2016-07-22 rebuttal.tex
adding rebuttal.tex

\documentclass{article}

\begin{document}
\title{Response to ``Introducing Distributed Dynamic Data-intensive (D3) Science: Understanding Applications and Infrastructure''}
\maketitle

We thank the reviewers for their ...

\section{Reviewer 1}

\begin{enumerate}
\item \emph{Comments to the Author This paper has a really excellent collection of examples and this survey is very valuable. There are two problems. The first is easy to fix and it constitutes the "minor revision" requested: expand each acronym at first use. For example, page 3 has NGS, CMB, MODIS, etc. say "Next Generation Sequencing (NGS)", etc.}

\textbf{Response:} We have made these changes.

\item \emph{The other thing is not something I feel you should revise. It is something that disappointed me. In the analysis of the infrastructure there is little specific detail. For example, tables III and IV there are no details about the specific infrastructure used (except for citations to papers). This leaves the impression that the approaches are all ad hoc. Indeed that may be true as the use of common tools in the sciences is very low.}

\textbf{Response:} Unfortunately, we do believe that this is true.

\item \emph{Also, I like figure 2, but the introduction of alpha and beta are of little value as there is no analysis presented here that uses them. This is mathyness. (sounds like math but it isn't. )}

\textbf{Response:} We have removed the formal defintion of alpha and beta, as we agree that we do not use these.  We added their names in the description since we do plan to use them in future work, and want to be able to point back to this paper as the place where they were introduced.
\end{enumerate}

\section{Reviewer: 2}

%Comments to the Author General Comments

\emph{This paper surveys several representative dynamic distributed data-intensive application scenarios, provides a common conceptual framework to understand them, and examines the infrastructure used in support of applications” I felt that the paper was useful and interesting regarding the first goal (describing data-intensive application scenarios), but was much less successful in the 2nd (providing a common conceptual framework to understand them). Terms are not well defined in the early parts before the survey, and so that makes it harder to understand and much less valuable. However, I think there is a relatively straightforward fix to many of these problems…. move the definitions in sections 3 and 4 before the survey. If this could be done, and the specific comments below addressed, then this would make the survey section (2) much more understandable and valuable. For example, the definition of “dynamic” is (I felt) not clear in section 1, and so this weakened the value I received from reading the survey. However, once I’d read section 3, things were much clearer. When I re-read the survey I got much more from it (but few readers will do this).}

\emph{Overall, I think this could be a valuable paper, surveying and explaining the demands of a wide range of applications if these problems could be addressed.}

\begin{enumerate}
\item  \emph{Here are some specific comments on the text that try to explain this view in more detail (as well as 1 or 2 rewording suggestions): Summary: “Often data are increasingly large-scale and distributed; and their location, availability and properties are time-dependent.” Needs editing to make it flow better Section 1. “software systems for the analysis of large volumes of dynamic, distributed, and data intensive have received relatively less so” [data intensive -> frequently accessed data?] “Occasionally data are often naturally distributed and have to be processed near their source, and sometimes data that are localized have to be distributed in order to be processed, either for reasons of performance or for collaborative analytics.” There are 2 dimensions being described here: 1. data can be created in a set of distributed locations or it can be created in 1 place. 2. data can be processed in 1 place, or it can be processed in distributed locations. The 2 dimensions are orthogonal, and all 4 possibilities occur in real applications. However, the above sentence only captures 2 of the 4 quadrants of the space. “In this paper, we use locally distributed to indicate that data/compute resides on multiple nodes within one data center. Geographically distributed describes the distribution of data and/or compute across multiple data centers” Given the above, I think it would be better in this definition to also separate out data creation location from compute location. Intro: Define precisely what is meant by “dynamic” here as it is key to the paper but isn’t well defined in Section 1 (see recommendation above to fix this by moving definitions from Section 3).}

\textbf{Response:}

\item \emph{Section 2. I didn’t think that the terminology and definitions of “traditional” and “infrastructural” applications made the 2 categories clear. Firstly, “traditional” is a poor choice of term as is not intuitively obvious what types of applications fall into this category. The definition didn’t really help me: “meaning a program that is developed by a user or project to try to answer a science question, usually with some input data needed and some output data produced.” Very few programs don’t require input or output data, so that wasn’t helpful to me. Also, “developed by a [..] project” seemed to also cover the community applications in the other category (if the community was focussed around a project). Also, the term “infrastructural” isn’t helpful for defining a type of application, as it suggests to me a generic, underpinning set of software services that can be used to build an application, rather than the application itself. The overloading of the term “application” in this category didn’t help to clarify the situation for me: “This allows different sub-applications (that we again refer to as ‘applications’)….”. Is the difference that applications in the 2nd category utilise services provided by an infrastructure that is shared by other applications (and users), unlike those in the first category? There may also be the distinction in that only 2nd category applications are distributed.. but that wasn’t clear from the text. The survey of applications in Section 2 is interesting, but suffers from using terms that are not well defined (see above for suggested fix). 2.1 “All algorithms, architectures, storage, and networks have broken or will break soon” Explain what is meant by “broken” and justify this claim. Section 3 see earlier comments on definitions. Section 4. “We also introduce the concept of an “infrastructure” application as a collection of infrastructure components that are packaged so that they can be used by many applications in deﬁned ways” See earlier comments. This doesn’t sound to me like an application, but instead a set of infrastructure support services (“components”) packaged from more basic services/components. Again, the use of application with 2 different meanings in this same sentence doesn’t help the read to understand what is meant here. 4.1 – 4.3 I found these definitions very helpful, but would again recommend that they come before the survey of specific applications so that these terms can be used, and understood by the reader, within the survey.}

\textbf{Response:}

\end{enumerate}

%Date Sent: 21-Jul-2016

\end{document}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2016-07-22 rebuttal.tex
more editing in rebuttal of reviewer comments, getting ready for responses

@@
 \begin{document}
-\title{Response to ``Introducing Distributed Dynamic Data-intensive (D3) Science: Understanding Applications and Infrastructure''}
+\title{Response to reviews of ``Introducing Distributed Dynamic Data-intensive (D3) Science: Understanding Applications and Infrastructure''}
 \maketitle
 
-We thank the reviewers for their ...
+We thank the editor and reviewers for their helpful comments, and have addressed their requests and questions here.
 
@@
 \begin{enumerate}
-\item \emph{Comments to the Author This paper has a really excellent collection of examples and this survey is very valuable.
+\item \emph{This paper has a really excellent collection of examples and this survey is very valuable.
 There are two problems.
@@
 
-\emph{This paper surveys several representative dynamic distributed data-intensive application scenarios, provides a common conceptual framework to understand them, and examines the infrastructure used in support of applications” I felt that the paper was useful and interesting regarding the first goal (describing data-intensive application scenarios), but was much less successful in the 2nd (providing a common conceptual framework to understand them).
+\emph{``This paper surveys several representative dynamic distributed data-intensive application scenarios, provides a common conceptual framework to understand them, and examines the infrastructure used in support of applications'' I felt that the paper was useful and interesting regarding the first goal (describing data-intensive application scenarios), but was much less successful in the 2nd (providing a common conceptual framework to understand them).
 Terms are not well defined in the early parts before the survey, and so that makes it harder to understand and much less valuable.
-However, I think there is a relatively straightforward fix to many of these problems….
+However, I think there is a relatively straightforward fix to many of these problems ...
 move the definitions in sections 3 and 4 before the survey.
 If this could be done, and the specific comments below addressed, then this would make the survey section (2) much more understandable and valuable.
-For example, the definition of “dynamic” is (I felt) not clear in section 1, and so this weakened the value I received from reading the survey.
-However, once I’d read section 3, things were much clearer.
+For example, the definition of ``dynamic'' is (I felt) not clear in section 1, and so this weakened the value I received from reading the survey.
+However, once I'd read section 3, things were much clearer.
 When I re-read the survey I got much more from it (but few readers will do this).}
@@
 
+\emph{Here are some specific comments on the text that try to explain this view in more detail (as well as 1 or 2 rewording suggestions):}
+
 \begin{enumerate}
-\item  \emph{Here are some specific comments on the text that try to explain this view in more detail (as well as 1 or 2 rewording suggestions): Summary: “Often data are increasingly large-scale and distributed; and their location, availability and properties are time-dependent.” Needs editing to make it flow better Section 1.
-“software systems for the analysis of large volumes of dynamic, distributed, and data intensive have received relatively less so” [data intensive -> frequently accessed data?] “Occasionally data are often naturally distributed and have to be processed near their source, and sometimes data that are localized have to be distributed in order to be processed, either for reasons of performance or for collaborative analytics.” There are 2 dimensions being described here: 1.
+
+\item \emph{Summary: ``Often data are increasingly large-scale and distributed; and their location, availability and properties are time-dependent.'' Needs editing to make it flow better}
+
+\textbf{Response:}
+
+\item \emph{Section 1.
+``software systems for the analysis of large volumes of dynamic, distributed, and data intensive have received relatively less soâ'' [data intensive -> frequently accessed data?] ``Occasionally data are often naturally distributed and have to be processed near their source, and sometimes data that are localized have to be distributed in order to be processed, either for reasons of performance or for collaborative analytics.'' There are 2 dimensions being described here: 1.
 data can be created in a set of distributed locations or it can be created in 1 place.
@@
 However, the above sentence only captures 2 of the 4 quadrants of the space.
-“In this paper, we use locally distributed to indicate that data/compute resides on multiple nodes within one data center.
-Geographically distributed describes the distribution of data and/or compute across multiple data centers” Given the above, I think it would be better in this definition to also separate out data creation location from compute location.
-Intro: Define precisely what is meant by “dynamic” here as it is key to the paper but isn’t well defined in Section 1 (see recommendation above to fix this by moving definitions from Section 3).}
+â``In this paper, we use locally distributed to indicate that data/compute resides on multiple nodes within one data center.
+Geographically distributed describes the distribution of data and/or compute across multiple data centersâ'' Given the above, I think it would be better in this definition to also separate out data creation location from compute location.
+Intro: Define precisely what is meant by ``dynamic,'' here as it is key to the paper but isn't well defined in Section 1 (see recommendation above to fix this by moving definitions from Section 3).}
 
@@
 \item \emph{Section 2.
-I didn’t think that the terminology and definitions of “traditional” and “infrastructural” applications made the 2 categories clear.
-Firstly, “traditional” is a poor choice of term as is not intuitively obvious what types of applications fall into this category.
-The definition didn’t really help me: “meaning a program that is developed by a user or project to try to answer a science question, usually with some input data needed and some output data produced.” Very few programs don’t require input or output data, so that wasn’t helpful to me.
-Also, “developed by a [..] project” seemed to also cover the community applications in the other category (if the community was focussed around a project).
-Also, the term “infrastructural” isn’t helpful for defining a type of application, as it suggests to me a generic, underpinning set of software services that can be used to build an application, rather than the application itself.
-The overloading of the term “application” in this category didn’t help to clarify the situation for me: “This allows different sub-applications (that we again refer to as ‘applications’)….”.
+I didn't think that the terminology and definitions of ``traditional'' and ``infrastructural'' applications made the 2 categories clear.
+Firstly, ``traditional'' is a poor choice of term as is not intuitively obvious what types of applications fall into this category.
+The definition didn't really help me: ``meaning a program that is developed by a user or project to try to answer a science question, usually with some input data needed and some output data produced.'' Very few programs don't require input or output data, so that wasn't helpful to me.
+Also, ``developed by a [..] project'' seemed to also cover the community applications in the other category (if the community was focussed around a project).
+Also, the term ``infrastructural'' isn't helpful for defining a type of application, as it suggests to me a generic, underpinning set of software services that can be used to build an application, rather than the application itself.
+The overloading of the term ``application'' in this category didn't help to clarify the situation for me: ``This allows different sub-applications (that we again refer to as `applications')...''.
 Is the difference that applications in the 2nd category utilise services provided by an infrastructure that is shared by other applications (and users), unlike those in the first category?
 There may also be the distinction in that only 2nd category applications are distributed..
-but that wasn’t clear from the text.
+but that wasn't clear from the text.
 The survey of applications in Section 2 is interesting, but suffers from using terms that are not well defined (see above for suggested fix).
-2.1 “All algorithms, architectures, storage, and networks have broken or will break soon” Explain what is meant by “broken” and justify this claim.
-Section 3 see earlier comments on definitions.
-Section 4.
-“We also introduce the concept of an “infrastructure” application as a collection of infrastructure components that are packaged so that they can be used by many applications in deﬁned ways” See earlier comments.
-This doesn’t sound to me like an application, but instead a set of infrastructure support services (“components”) packaged from more basic services/components.
-Again, the use of application with 2 different meanings in this same sentence doesn’t help the read to understand what is meant here.
-4.1 – 4.3 I found these definitions very helpful, but would again recommend that they come before the survey of specific applications so that these terms can be used, and understood by the reader, within the survey.}
+2.1 ``All algorithms, architectures, storage, and networks have broken or will break soon'' Explain what is meant by ``broken'' and justify this claim.}
+
+\textbf{Response:}
+
+\item \emph{Section 3 see earlier comments on definitions.}
+
+\textbf{Response:}
+
+\item \emph{Section 4.
+``We also introduce the concept of an `infrastructure' application as a collection of infrastructure components that are packaged so that they can be used by many applications in defined ways'' See earlier comments.
+This doesn't sound to me like an application, but instead a set of infrastructure support services (``components'') packaged from more basic services/components.
+Again, the use of application with 2 different meanings in this same sentence doesn't help the read to understand what is meant here.
+4.1 -- 4.3 I found these definitions very helpful, but would again recommend that they come before the survey of specific applications so that these terms can be used, and understood by the reader, within the survey.}
 

### 2016-07-23 rebuttal.tex
editing sentence in intro discussed by reviewer 2 breaking up reviewer 2 comments into smaller mostly-independent chunks

@@
 
-\textbf{Response:} We have removed the formal defintion of alpha and beta, as we agree that we do not use these.
+\textbf{Response:} We have removed the formal definition of alpha and beta, as we agree that we do not use these.
 We added their names in the description since we do plan to use them in future work, and want to be able to point back to this paper as the place where they were introduced.
@@
 
+\textbf{Response:} We've edited this.
+
+\hspace{-0.7cm}Section 1.
+
+\item \emph{``software systems for the analysis of large volumes of dynamic, distributed, and data intensive have received relatively less soâ'' [data intensive -> frequently accessed data?]}
+
 \textbf{Response:}
 
-\item \emph{Section 1.
-``software systems for the analysis of large volumes of dynamic, distributed, and data intensive have received relatively less soâ'' [data intensive -> frequently accessed data?] ``Occasionally data are often naturally distributed and have to be processed near their source, and sometimes data that are localized have to be distributed in order to be processed, either for reasons of performance or for collaborative analytics.'' There are 2 dimensions being described here: 1.
+\item \emph{``Occasionally data are often naturally distributed and have to be processed near their source, and sometimes data that are localized have to be distributed in order to be processed, either for reasons of performance or for collaborative analytics.'' There are 2 dimensions being described here: 1.
 data can be created in a set of distributed locations or it can be created in 1 place.
@@
 The 2 dimensions are orthogonal, and all 4 possibilities occur in real applications.
-However, the above sentence only captures 2 of the 4 quadrants of the space.
-â``In this paper, we use locally distributed to indicate that data/compute resides on multiple nodes within one data center.
+However, the above sentence only captures 2 of the 4 quadrants of the space.}
+
+\textbf{Response:}
+
+\item \emph{``In this paper, we use locally distributed to indicate that data/compute resides on multiple nodes within one data center.
 Geographically distributed describes the distribution of data and/or compute across multiple data centersâ'' Given the above, I think it would be better in this definition to also separate out data creation location from compute location.
@@
 
-\item \emph{Section 2.
-I didn't think that the terminology and definitions of ``traditional'' and ``infrastructural'' applications made the 2 categories clear.
+\hspace{-0.7cm}Section 2.
+
+\item \emph{I didn't think that the terminology and definitions of ``traditional'' and ``infrastructural'' applications made the 2 categories clear.
 Firstly, ``traditional'' is a poor choice of term as is not intuitively obvious what types of applications fall into this category.
@@
 There may also be the distinction in that only 2nd category applications are distributed..
-but that wasn't clear from the text.
-The survey of applications in Section 2 is interesting, but suffers from using terms that are not well defined (see above for suggested fix).
-2.1 ``All algorithms, architectures, storage, and networks have broken or will break soon'' Explain what is meant by ``broken'' and justify this claim.}
+but that wasn't clear from the text.}
 
@@
 
-\item \emph{Section 3 see earlier comments on definitions.}
+\item \emph{The survey of applications in Section 2 is interesting, but suffers from using terms that are not well defined (see above for suggested fix).}
 
@@
 
-\item \emph{Section 4.
-``We also introduce the concept of an `infrastructure' application as a collection of infrastructure components that are packaged so that they can be used by many applications in defined ways'' See earlier comments.
+\item \emph{2.1 ``All algorithms, architectures, storage, and networks have broken or will break soon'' Explain what is meant by ``broken'' and justify this claim.}
+
+\textbf{Response:}
+
+\hspace{-0.7cm}Section 3.
+
+\item \emph{see earlier comments on definitions.}
+
+\textbf{Response:}
+
+\hspace{-0.7cm}Section 4.
+
+\item \emph{``We also introduce the concept of an `infrastructure' application as a collection of infrastructure components that are packaged so that they can be used by many applications in defined ways'' See earlier comments.
 This doesn't sound to me like an application, but instead a set of infrastructure support services (``components'') packaged from more basic services/components.
-Again, the use of application with 2 different meanings in this same sentence doesn't help the read to understand what is meant here.
-4.1 -- 4.3 I found these definitions very helpful, but would again recommend that they come before the survey of specific applications so that these terms can be used, and understood by the reader, within the survey.}
+Again, the use of application with 2 different meanings in this same sentence doesn't help the read to understand what is meant here.}
+
+\textbf{Response:}
+
+\item \emph{4.1 -- 4.3 I found these definitions very helpful, but would again recommend that they come before the survey of specific applications so that these terms can be used, and understood by the reader, within the survey.}
 

### 2016-07-23 rebuttal.tex
addressing one more reviewer2 comment adding header stuff in rebuttal for our comments marking more reviewer2 items for discussion

@@
 \documentclass{article}
+
+\usepackage{moreverb}
+
+\newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08emT\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
+
+\usepackage{graphicx}
+\usepackage{url} 
+\usepackage{color} 
+\usepackage{enumerate}
+\usepackage{longtable} 
+\usepackage{textcomp} % \usepackage{srcltx} % \usepackage{fancyhdr} %
+\usepackage{setspace} \usepackage{lscape} % \usepackage{longtable} %
+\usepackage{paralist}
+\setcounter{totalnumber}{50} \setcounter{topnumber}{50}
+\setcounter{bottomnumber}{50}
+
+\newcommand{\I}[1]{\textit{#1}} \newcommand{\B}[1]{\textbf{#1}}
+\newcommand{\T}[1]{\texttt{#1}} \newcommand{\BI}[1]{\B{\I{#1}}}
+
+ \newenvironment{shortlist}{ \vspace*{-0.8em}
+  \begin{itemize} \setlength{\itemsep}{-0.3em} }{
+  \end{itemize} \vspace*{-0.6em} }
+
+ \newenvironment{shortenum}{ \vspace*{-0.8em}
+  \begin{enumerate} \setlength{\itemsep}{-0.3em} }{
+  \end{enumerate} \vspace*{-0.6em} }
+
+\definecolor{orange}{rgb}{1.0,0.3,0.0} \definecolor{violet}{rgb}{0.75,0,1}
+\definecolor{darkgreen}{rgb}{0,0.6,0} \definecolor{cyan}{rgb}{0.2,0.7,0.7}
+\definecolor{blueish}{rgb}{0.2,0.2,0.8}
+
+\long\def\comment#1{{\bf \textcolor{magenta}{\bf #1}}} \long\def\ccomment#1{{\bf
+\textcolor{blue}{\bf #1}}} \newcommand{\C}{\comment} \newcommand{\CC}{\ccomment}
+
+\newcommand{\yes}{$\bullet$}
+
+\long\def\finalcomment#1{{\bf \textcolor{red}{\bf #1}}}
+\newcommand{\CF}{\finalcomment}
+
+\newif\ifdraft 
+\drafttrue 
+\ifdraft 
+\newcommand{\katznote}[1]{{\textcolor{magenta} { ***Dan: #1 }}} 
+\newcommand{\rananote}[1]{{\textcolor{cyan} { ***Omer: #1 }}} 
+\newcommand{\omernote}[1]{ {\textcolor{cyan}{ ***Omer: #1 }}} 
+\newcommand{\note}[1]{ {\textcolor{red} { #1 }}}
+\newcommand{\jhanote}[1]{ {\textcolor{blue} { ***Shantenu: #1 }}}
+\newcommand{\jonnote}[1]{ {\textcolor{red} { ***Jon: #1 }}}
+\newcommand{\alnote}[1]{ {\textcolor{darkgreen} { ***Andre: #1 }}}
+\newcommand{\nchnote}[1]{ {\textcolor{orange} { ***Neil: #1 }}}
+\newcommand{\ysnote}[1]{ {\textcolor{violet} { ***Yogesh: #1 }}}
+\newcommand{\sdnote}[1]{ {\textcolor{blueish} { ***Simon: #1 }}} 
+\else 
+\newcommand{\katznote}[1]{}
+\newcommand{\rananote}[1]{} 
+\newcommand{\note}[1]{} 
+\newcommand{\jhanote}[1]{}
+\newcommand{\jonnote}[1]{} 
+\newcommand{\alnote}[1]{} 
+\newcommand{\nchnote}[1]{}
+\newcommand{\ysnote}[1]{} 
+\fi
+
+\usepackage[pdftex,colorlinks=true, linkcolor=blue,citecolor=blue]{hyperref}
 
@@
 
-\textbf{Response:} We've edited this.
+\textbf{Response:} We've edited this to: ``Data sets are growing larger and becoming distributed; and their location, availability and properties are often time-dependent.''
 
@@
 
-\item \emph{``software systems for the analysis of large volumes of dynamic, distributed, and data intensive have received relatively less soâ'' [data intensive -> frequently accessed data?]}
+\item \emph{``software systems for the analysis of large volumes of dynamic, distributed, and data intensive have received relatively less soâ'' [data intensive $\rightarrow$ frequently accessed data?]}
 
-\textbf{Response:}
+\textbf{Response:} We think this is actually referring to the summary, where we had ``software systems for the analysis of large volumes of dynamic, distributed, and data intensive have received relatively less.''  This was missing the word applications in the midst of the sentence, which made it confusing.
+``Less'' referred to less attention, so ``attention has also been added to make this more clear.'' The sentence is now ``software systems for the analysis of large volumes of dynamic, distributed, and data intensive applications have received relatively less attention.''
 
@@
 
-\textbf{Response:}
+\textbf{Response:} \katznote{attention needed here}
 

### 2016-07-23 3dpas_intro.tex
editing sentence in intro discussed by reviewer 2 breaking up reviewer 2 comments into smaller mostly-independent chunks
addressing one more reviewer2 comment adding header stuff in rebuttal for our comments marking more reviewer2 items for discussion

@@
   amount and diversity of data and computation that must be integrated to yield
-  insights. Often data are increasingly large-scale and distributed; and their
-  location, availability and properties are time-dependent. Collectively, these
+  insights. Data sets are growing larger and becoming distributed; and their
+  location, availability and properties are often time-dependent. Collectively, these
   characteristics give rise to dynamic distributed data intensive
@@
   attention, the characteristics, requirements, and software systems for the
-  analysis of large volumes of dynamic, distributed, and data intensive have
-  received relatively less. This paper surveys several representative dynamic
+  analysis of large volumes of dynamic, distributed, and data intensive applications have
+  received relatively less attention. This paper surveys several representative dynamic
   distributed data-intensive application scenarios, provides a common conceptual

### 2016-07-24 rebuttal.txt

We thank the reviewers for their ...

Referee(s)' Comments to Author:

Reviewer: 1

Comments to the Author This paper has a really excellent collection of examples and this survey is very valuable. There are two problems. The first is easy to fix and it constitutes the "minor revision" requested: expand each acronym at first use. For example, page 3 has NGS, CMB, MODIS, etc. say "Next Generation Sequencing (NGS)", etc.

Response: We have made these changes.

The other thing is not something I feel you should revise. It is something that disappointed me. In the analysis of the infrastructure there is little specific detail. For example, tables III and IV there are no details about the specific infrastructure used (except for citations to papers). This leaves the impression that the approaches are all ad hoc. Indeed that may be true as the use of common tools in the sciences is very low.

Response: Unfortunately, we do believe that this is true.

Also, I like figure 2, but the introduction of alpha and beta are of little value as there is no analysis presented here that uses them. This is mathyness. (sounds like math but it isn't. )

We have removed the formal defintion of alpha and beta, as we agree that we do not use these.  We added their names in the description since we do plan to use them in future work, and want to be able to point back to this paper as the place where they were introduced.

Reviewer: 2

Comments to the Author General Comments

“This paper surveys several representative dynamic distributed data-intensive application scenarios, provides a common conceptual framework to understand them, and examines the infrastructure used in support of applications” I felt that the paper was useful and interesting regarding the first goal (describing data-intensive application scenarios), but was much less successful in the 2nd (providing a common conceptual framework to understand them). Terms are not well defined in the early parts before the survey, and so that makes it harder to understand and much less valuable. However, I think there is a relatively straightforward fix to many of these problems…. move the definitions in sections 3 and 4 before the survey. If this could be done, and the specific comments below addressed, then this would make the survey section (2) much more understandable and valuable. For example, the definition of “dynamic” is (I felt) not clear in section 1, and so this weakened the value I received from reading the survey. However, once I’d read section 3, things were much clearer. When I re-read the survey I got much more from it (but few readers will do this).

Overall, I think this could be a valuable paper, surveying and explaining the demands of a wide range of applications if these problems could be addressed.

Here are some specific comments on the text that try to explain this view in more detail (as well as 1 or 2 rewording suggestions): Summary: “Often data are increasingly large-scale and distributed; and their location, availability and properties are time-dependent.” Needs editing to make it flow better Section 1. “software systems for the analysis of large volumes of dynamic, distributed, and data intensive have received relatively less so” [data intensive -> frequently accessed data?] “Occasionally data are often naturally distributed and have to be processed near their source, and sometimes data that are localized have to be distributed in order to be processed, either for reasons of performance or for collaborative analytics.” There are 2 dimensions being described here: 1. data can be created in a set of distributed locations or it can be created in 1 place. 2. data can be processed in 1 place, or it can be processed in distributed locations. The 2 dimensions are orthogonal, and all 4 possibilities occur in real applications. However, the above sentence only captures 2 of the 4 quadrants of the space. “In this paper, we use locally distributed to indicate that data/compute resides on multiple nodes within one data center. Geographically distributed describes the distribution of data and/or compute across multiple data centers” Given the above, I think it would be better in this definition to also separate out data creation location from compute location. Intro: Define precisely what is meant by “dynamic” here as it is key to the paper but isn’t well defined in Section 1 (see recommendation above to fix this by moving definitions from Section 3).

Section 2. I didn’t think that the terminology and definitions of “traditional” and “infrastructural” applications made the 2 categories clear. Firstly, “traditional” is a poor choice of term as is not intuitively obvious what types of applications fall into this category. The definition didn’t really help me: “meaning a program that is developed by a user or project to try to answer a science question, usually with some input data needed and some output data produced.” Very few programs don’t require input or output data, so that wasn’t helpful to me. Also, “developed by a [..] project” seemed to also cover the community applications in the other category (if the community was focussed around a project). Also, the term “infrastructural” isn’t helpful for defining a type of application, as it suggests to me a generic, underpinning set of software services that can be used to build an application, rather than the application itself. The overloading of the term “application” in this category didn’t help to clarify the situation for me: “This allows different sub-applications (that we again refer to as ‘applications’)….”. Is the difference that applications in the 2nd category utilise services provided by an infrastructure that is shared by other applications (and users), unlike those in the first category? There may also be the distinction in that only 2nd category applications are distributed.. but that wasn’t clear from the text. The survey of applications in Section 2 is interesting, but suffers from using terms that are not well defined (see above for suggested fix). 2.1 “All algorithms, architectures, storage, and networks have broken or will break soon” Explain what is meant by “broken” and justify this claim. Section 3 see earlier comments on definitions. Section 4. “We also introduce the concept of an “infrastructure” application as a collection of infrastructure components that are packaged so that they can be used by many applications in deﬁned ways” See earlier comments. This doesn’t sound to me like an application, but instead a set of infrastructure support services (“components”) packaged from more basic services/components. Again, the use of application with 2 different meanings in this same sentence doesn’t help the read to understand what is meant here. 4.1 – 4.3 I found these definitions very helpful, but would again recommend that they come before the survey of specific applications so that these terms can be used, and understood by the reader, within the survey.
Date Sent:
21-Jul-2016

### 2016-07-28 3dpas_intro.tex
minor.   removing reference to the book-the-changed-the-world, as it didn't quite materialize..

@@
 a spatial data set between components that cooperate to perform a
-single logical task~\cite{dpa_surveypaper,thebookthatchangedtheworld}.
+single logical task~\cite{dpa_surveypaper}.
+% thebookthatchangedtheworld
 Most of the applications in \S\ref{sec:scenarios} are loosely coupled,

### 2016-07-28 3dpas_intro.tex
adding a paragraph at the end of section 1 with info for the reader

@@
 conclusions (\S\ref{sec:conclusions}).
+
+\textit{Guidance for the reader}: Readers who are generally knowledge about distributed
+applications will likely feel comfortable reading the paper from start to end.
+Readers with less distributed applications knowledge may prefer to read
+\S\ref{sec:distdyndata} and \S\ref{sec:infrastructure} before \S\ref{sec:scenarios},
+or might want to read \S\ref{sec:scenarios} first, but jump to \S\ref{sec:distdyndata} and \S\ref{sec:infrastructure} when they find terms which which they are unfamiliar.
+Unfortunately, we do not feel that there is a single ordering of the information in this
+paper that is best for all readers.
 

### 2016-07-28 rebuttal.tex
fixing quote marks
adding partial response to point 4, based on new paragraph adding response to later short point to point to this response

@@
 There are two problems.
-The first is easy to fix and it constitutes the "minor revision" requested: expand each acronym at first use.
+The first is easy to fix and it constitutes the ``minor revision'' requested: expand each acronym at first use.
 For example, page 3 has NGS, CMB, MODIS, etc.
@@
 
+\katznote{partial response:} In addition, recognizing that the two reviewers have different styles of reading
+and different areas of expertise (as do we as authors), we have added a paragraph at the end
+of Section 1 to explain how different types of readers might want to approach this paper.  We do
+not believe that there is one single way of organizing it that would satisfy all readers.
+
 \hspace{-0.7cm}Section 2.
@@
 
-\textbf{Response:} \katznote{attention needed here}
+\textbf{Response:} See earlier response about definitions, in point 4.
 

### 2016-08-15 3dpas_intro.tex
adding Omer's changes, which some very small changes as well

@@
 analysis and characterization of the applications and
-conclusions (\S\ref{sec:conclusions}).
+conclusions (\S\ref{sec:conclusions}). Each of these sections is briefly described below:
+
+\begin{itemize}
+
+\item Section~\ref{sec:scenarios} provides a description of thirteen D3 applications based on our survey.
+For each application, both the {\it dynamic} and {\it distributed} aspects are described.
+The methodology used to carry out the survey is described in the Appendix.
+
+\item Based on the application survey, we derive a set of  definitions for {\it Distribution} and {\it Dynamism} in \S\ref{sec:distdyndata}.
+Note that these definitions are based on the application survey.
+A variety of definitions of these concepts already exist in computer science literature.
+Our intention is
+to provide context for the use of these terms with reference to our survey.
+The distributed and dynamic characteristics of each application surveyed in \S\ref{sec:scenarios} is summarized in this section.
+
+\item In \S\ref{sec:infrastructure}, we subsequently describe the software/systems infrastructure used to support the distributed and dynamic characteristics of applications in \S\ref{sec:scenarios}.
+We outline support for data management, analysis, and coordination, providing examples of actual Cyberinfrastructure made use of by our surveyed applications for these purposes.
+
+\item In \S\ref{sec:conclusions}, we bring together key observations based on the analysis of applications, their distributed and dynamic characteristics, and infrastructure use.
+An attempt is made to compare these applications and suggest the need for an ``architecture for D3 science".
+
+\end{itemize}
 

### 2016-08-17 3dpas_intro.tex
minor: removing claim about everything is broken, but this commit message..
Discussion about "hierarchy" and "granularity" of applications.

@@
 
-Starting with the experience of the attendees of the 3DPAS workshops,
-we have assembled a set of D3 ``applications'' that we describe in
-this section.  We realize that we have overloaded the term application
-here.  In this document, some of our applications are really {\em
-  traditional applications}, meaning a program that is developed by a
-user or project to try to answer a science question, usually with some
-input data needed and some output data produced.  Applications in this
-category are: NGS (next generation sequencing) Analytics (\S\ref{bioSilvia}), CMB (cosmic microwave background)
-(\S\ref{astroJulian}), Fusion (\S\ref{fusionScott}), Industrial
-Incident Notification and Response (\S\ref{Kees}), MODIS (moderate resolution imaging spectroradiometer) Data
-Processing (\S\ref{modisKeith}), and Distributed Network Intrusion
-Detection (\S\ref{detectionJon}).
+Starting with the experience of the attendees of the 3DPAS workshops, we have
+assembled a set of D3 ``applications'' that we describe in this section.  
+%We realize that we have overloaded the term ``application'' here.
+
+The term ``application'' is often overloaded in literature and the computational
+science discourse.  Sometimes the term is a reference to a stand-alone,
+independent executable (e.g., a Molecular Dynamics simulation kernel such as
+AMBER or Gromacs). Sometimes an application refers to science problem to which
+the executable is put to use (e.g., compute the free-energy of binding energy of
+a drug candidate), but sometimes it is also used to reference the end-to-end
+workflow (i.e., a multi-stage computational process requiring a sequence of
+possibly distinct simulations and executables).  To add chaos to the confusion,
+applications are sometimes also as a proxy for projects (e.g., the ATLAS
+project), where the project could in of itself possibly be a series of
+independent applications, each with application in turn a reference to a defined
+science objective achieved using a set of executables.
+
+In addition to the aforementioned hierarchy and granularity to which the term
+application has been applied, the class of applications in any level differ
+widely in scale, sophistication and type as well different objective and
+objects.  Thus, it is inevitable that the non-contextual use of the word
+application be a recipe for befuddlement.
+
+In this document, some of our applications are really {\em traditional
+  applications}, meaning a program that is developed by a user or project to try
+to answer a science question, usually with some input data needed and some
+output data produced.  Applications in this category are: NGS (next generation
+sequencing) Analytics (\S\ref{bioSilvia}), CMB (cosmic microwave background)
+(\S\ref{astroJulian}), Fusion (\S\ref{fusionScott}), Industrial Incident
+Notification and Response (\S\ref{Kees}), MODIS (moderate resolution imaging
+spectroradiometer) Data Processing (\S\ref{modisKeith}), and Distributed Network
+Intrusion Detection (\S\ref{detectionJon}).
 
 Another group of applications are {\em infrastructural applications}.
-Applications of this type are collaborative, in that they depend on a
-context that is often agreed upon by a community, most often in terms
-of how data is stored by the community and how it is accessed, as well
-as, more often than not, a weak consensus on the infrastructure to be
+Applications of this type are collaborative, in that they depend on a context
+that is often agreed upon by a community, most often in terms of how data is
+stored by the community and how it is accessed, as well as, more often than not,
+a weak consensus on the infrastructure to be
 used. %\jhanote{Dan: Please confirm you are OK with this modified
@@
 
-All algorithms, architectures, storage, and networks have broken or
-will break soon.  Some NGS analytics are not good fits for core-rich
-and memory-poor modern architectures. Some problems (e.g., comparing
-k-mers across metagenomic data sets) need terabyte-scale memory, not
-thousands of cores.~\cite{ngs-gap}
+% All algorithms, architectures, storage, and networks have broken or
+% will break soon. 
+Some NGS analytics are not good fits for core-rich and memory-poor modern
+architectures. Some problems (e.g., comparing k-mers across metagenomic data
+sets) need terabyte-scale memory, not thousands of cores.~\cite{ngs-gap}
 

### 2016-08-17 3dpas_intro.tex
ongoing refinement.

@@
 widely in scale, sophistication and type as well different objective and
-objects.  Thus, it is inevitable that the non-contextual use of the word
-application be a recipe for befuddlement.
+objects.  Thus, we warn the reader of the inevitable befuddlement that the
+casual and non-contextual use of the word application can cause.
 
@@
 stored by the community and how it is accessed, as well as, more often than not,
-a weak consensus on the infrastructure to be
-used. %\jhanote{Dan: Please confirm you are OK with this modified
-%  defintion of infrastructural application}
-This allows different
-sub-applications (that we again refer to as `applications') to focus
-on different stages, such as generating data, or processing data.
-Answering a science question now may involve a set of applications
-that need to be run in series, perhaps in different stages, that may
-be run by different groups that do not frequently interact.  This type
-of application includes: ATLAS/WLCG (a toroidal large hadron collider apparatus/worldwide large hadron collider computing grid, \S\ref{WLCGSteve}), LSST
-(large synoptic survey telescope, \S\ref{astroAdam}), SOA (service-oriented architecture) Astronomy (\S\ref{astroSOAAdam}), Sensor
-Network Application (\S\ref{sensorSimon}), Climate
-(\S\ref{climateDan}), Interactive Exploration of Environmental Data
+a weak consensus on the infrastructure to be used.  This allows different
+sub-applications (that we again refer to as `applications') to focus on
+different stages, such as generating data, or processing data.  Answering a
+science question now may involve a set of applications that need to be run in
+series, perhaps in different stages, that may be run by different groups that do
+not frequently interact.  This type of application includes: ATLAS/WLCG (a
+toroidal large hadron collider apparatus/worldwide large hadron collider
+computing grid, \S\ref{WLCGSteve}), LSST (large synoptic survey telescope,
+\S\ref{astroAdam}), SOA (service-oriented architecture) Astronomy
+(\S\ref{astroSOAAdam}), Sensor Network Application (\S\ref{sensorSimon}),
+Climate (\S\ref{climateDan}), Interactive Exploration of Environmental Data
 (\S\ref{envJon}), and Power Grids (\S\ref{power}).

### 2016-08-17 3dpas_intro.tex
more work on applications.

@@
 
-In this document, some of our applications are really {\em traditional
-  applications}, meaning a program that is developed by a user or project to try
-to answer a science question, usually with some input data needed and some
-output data produced.  Applications in this category are: NGS (next generation
+% In this document, some of our applications are really {\em traditional
+% applications}, meaning a program that is developed by a user or project to try
+% to answer a science question, usually with some input data needed and some
+% output data produced.
+
+Against this backdrop, we eschew a formal classification of application types,
+however, to facilitate understanding we categorise applications into two
+categories: the first is that of {\em traditional applications}, where
+% traditional is  applications are {\em traditional applications}, in that 
+a program (or set of programs) is developed by a user or project to try to
+answer a science question (or set of related questions). Traditional application
+might A single standalone computer program, a distributed application or a
+complex workflow could fall under the category of a {\em traditional
+  application}.  Applications in this category are: NGS (next generation
 sequencing) Analytics (\S\ref{bioSilvia}), CMB (cosmic microwave background)
@@
 
-In other words, multiple infrastructural applications are used together in particular
-science domains.  Furthermore, infrastructural
-applications involve more than just the end user (scientist), and
-typically require exerting control at layers below, i.e.,
-``programming the system (infrastructure)'', as opposed to a
-traditional applications, where end user effort and control is often
-confined to a well-defined application kernel.
+In other words, multiple infrastructural applications are used together in
+particular science domains.  Furthermore, infrastructural applications involve
+more than just the end user (scientist), and typically require exerting control
+at layers below, i.e., ``programming the system (infrastructure)'', as opposed
+to a traditional applications, where end user effort and control is often
+confined to a well-defined application kernel and their communication and
+coordination.  Also, for some {\em traditional applications} (e.g., NGS) the
+number of users of an instance of the application is typically small, i.e., the
+predominant use is by the long-tail of science. In contrast, some
+infrastructural applications are correlated with ``big science'' projects.
 

### 2016-08-17 rebuttal.tex
describing Omer's changes
Discussion about "hierarchy" and "granularity" of applications.
responses about applications.

@@
 of Section 1 to explain how different types of readers might want to approach this paper.  We do
-not believe that there is one single way of organizing it that would satisfy all readers.
+not believe that there is one single way of organizing it that would satisfy all readers.  We have also
+added a longer description of each section at this point.
 
@@
 
-\textbf{Response:} \katznote{attention needed here}
+  \textbf{Response:} This was a legacy remark which should have been removed when we deleted contextual discussion. We have removed this over-reaching claim. Apologies for the oversight.
 
@@
 
-\textbf{Response:} \katznote{attention needed here}
+  \textbf{Response:} We have: (i) provided background discussion about the overloading of the term application, (ii) expanded and clarified the discussion of traditional and infrastructural applications.
 

### 2016-08-18 rebuttal.tex
rebuttal update.
fixing formatting, fixing a word

--- 2016-07-28 rebuttal.tex
@@
 
-\textbf{Response:} We have removed the formal definition of alpha and beta, as we agree that we do not use these.
+\textbf{Response:} We have removed the formal definition of alpha and beta, as we agree that we do not use these elsewhere in the paper.
 We added their names in the description since we do plan to use them in future work, and want to be able to point back to this paper as the place where they were introduced.
@@
 
-\item \emph{``software systems for the analysis of large volumes of dynamic, distributed, and data intensive have received relatively less soâ'' [data intensive $\rightarrow$ frequently accessed data?]}
+\item \emph{``software systems for the analysis of large volumes of dynamic, distributed, and data intensive have received relatively less so'' [data intensive $\rightarrow$ frequently accessed data?]}
 
@@
 \item \emph{``In this paper, we use locally distributed to indicate that data/compute resides on multiple nodes within one data center.
-Geographically distributed describes the distribution of data and/or compute across multiple data centersâ'' Given the above, I think it would be better in this definition to also separate out data creation location from compute location.
+Geographically distributed describes the distribution of data and/or compute across multiple data centers'' Given the above, I think it would be better in this definition to also separate out data creation location from compute location.
 Intro: Define precisely what is meant by ``dynamic,'' here as it is key to the paper but isn't well defined in Section 1 (see recommendation above to fix this by moving definitions from Section 3).}
@@
 of Section 1 to explain how different types of readers might want to approach this paper.  We do
-not believe that there is one single way of organizing it that would satisfy all readers.
+not believe that there is one single way of organizing it that would satisfy all readers.  We have also
+added a longer description of each section at this same place.
 

### 2016-08-18 rebuttal.tex

@@
 
-\textbf{Response:} \katznote{attention needed here}
+\textbf{Response:} Although we have retained the terms and only marginally modified the definitions, we have refined the discussion of ``infrastructural'' applications.
 
@@
 
-\textbf{Response:} \katznote{attention needed here}
+\textbf{Response:} \katznote{attention needed here}\jhanote{this should also have been addressed by Omer's work. Omer can you confirm please?}
 
@@
 
-\textbf{Response:} \katznote{attention needed here}
+  \textbf{Response:} \katznote{attention needed here}\jhanote{this should also have been addressed by Omer's work. Omer can you confirm please?}
 

### 2016-08-18 3dpas_intro.tex

@@
 \section*{Acknowledgements}
-The workshop was sponsored by NSF OCI-1059635 and made possible by the UK
+The work was sponsored by NSF OCI-1059635 and made possible by the UK
 e-Science Institute Research Theme on Dynamic Distributed Data-Intensive

### 2016-08-18 3dpas_intro.tex
small edits to applications work from Shantenu

@@
 The term ``application'' is often overloaded in literature and the computational
-science discourse.  Sometimes the term is a reference to a stand-alone,
+science discourse.  Sometimes the term is a reference to a standalone,
 independent executable (e.g., a Molecular Dynamics simulation kernel such as
-AMBER or Gromacs). Sometimes an application refers to science problem to which
-the executable is put to use (e.g., compute the free-energy of binding energy of
+AMBER or Gromacs). Sometimes it refers to science problem to which
+the executable is put to use (e.g., computing the free-energy of binding energy of
 a drug candidate), but sometimes it is also used to reference the end-to-end
@@
 applications are sometimes also as a proxy for projects (e.g., the ATLAS
-project), where the project could in of itself possibly be a series of
-independent applications, each with application in turn a reference to a defined
+project), where the project could itself possibly be a series of
+independent applications, with each application in turn a reference to a defined
 science objective achieved using a set of executables.
@@
 Against this backdrop, we eschew a formal classification of application types,
-however, to facilitate understanding we categorise applications into two
+however, to facilitate understanding we categorize applications into two
 categories: the first is that of {\em traditional applications}, where
@@
 a program (or set of programs) is developed by a user or project to try to
-answer a science question (or set of related questions). Traditional application
-might A single standalone computer program, a distributed application or a
+answer a science question (or set of related questions). 
+A single standalone computer program, a distributed application, or a
 complex workflow could fall under the category of a {\em traditional
-  application}.  Applications in this category are: NGS (next generation
+application}.  Applications in this paper that are in this category are: NGS (next generation
 sequencing) Analytics (\S\ref{bioSilvia}), CMB (cosmic microwave background)
@@
 series, perhaps in different stages, that may be run by different groups that do
-not frequently interact.  This type of application includes: ATLAS/WLCG (a
+not frequently interact.  In this paper, examples of this type of application includes: ATLAS/WLCG (a
 toroidal large hadron collider apparatus/worldwide large hadron collider
@@
 more than just the end user (scientist), and typically require exerting control
-at layers below, i.e., ``programming the system (infrastructure)'', as opposed
-to a traditional applications, where end user effort and control is often
+at lower layers, i.e., ``programming the system (infrastructure)'', as opposed
+to a traditional applications where end user effort and control is often
 confined to a well-defined application kernel and their communication and

### 2016-08-18 rebuttal.tex
meaningless space addition to text commit

@@
 
-\textbf{Response:} Although we have retained the terms and only marginally modified the definitions, we have refined the discussion of ``infrastructural'' applications.
+\textbf{Response:} Although we have retained the terms, we have also modified the definitions and specifically, we have refined the discussion of ``infrastructural'' applications.  We believe that
+understanding the types of applications, which may be a multi-stage process (reading the definitions/discussion, then reading about the applications themselves) is important to our discussion and conclusions that occur after the Applications Section.
 
@@
 
-\textbf{Response:} \katznote{attention needed here}\jhanote{this should also have been addressed by Omer's work. Omer can you confirm please?}
+\textbf{Response:} \katznote{attention needed here}\jhanote{this should also have been addressed by Omer's work.
+Omer can you confirm please?} \katznote{no, Omer's work added a better description of the paper at the end of Section 1.}
 
@@
 
-  \textbf{Response:} This was a legacy remark which should have been removed when we deleted contextual discussion. We have removed this over-reaching claim. Apologies for the oversight.
+  \textbf{Response:} This was a legacy remark which should have been removed when we deleted contextual discussion. We have removed this over-reaching claim.
 
@@
 
-  \textbf{Response:} \katznote{attention needed here}\jhanote{this should also have been addressed by Omer's work. Omer can you confirm please?}
+  \textbf{Response:} \katznote{attention needed here}\jhanote{this should also have been addressed by Omer's work.
+Omer can you confirm please?} \katznote{no, Omer's work added a better description of the paper at the end of Section 1.}
 

### 2016-08-18 3dpas_intro.tex

@@
 The landscape of scientific and enterprise computing is being fundamentally
-altered by prodigious volumes of data.  Occasionally data are often naturally
-distributed and have to be processed near their source, and sometimes data that
-are localized have to be distributed in order to be processed, either for
-reasons of performance or for collaborative analytics.  Further, the variation
-of data production rates, data source and destination makes temporal variations
-in data properties increasingly important.  With an increase in the importance
-of distributed and temporal properties of data, systems and infrastructure
-aspects such as resource scheduling, data placement, and transfer decisions that
+altered by prodigious volumes of data.  
+%Occasionally data are often naturally
+%distributed and have to be processed near their source, and sometimes data that
+%are localized have to be distributed in order to be processed, Data can be naturally distributed because they are
+%being created in a set of distributed locations, rather than in one place. 
+Data can also be transported to be processed in multiple, distributed
+locations, rather than being processed near their source, or be brought together for processing at
+a particular location. These two orthogonal dimensions lead to a number of patterns of data
+generation and processing being present in real applications,
+either for reasons of performance or for collaborative analytics.  
+Further, the variation of data production rates, data source and destination makes 
+temporal variations in data properties increasingly important.  
+With an increase in the importance of distributed and temporal properties of 
+data, systems and infrastructure aspects such as resource scheduling, d
+ata placement, and transfer decisions that
 were statically determinable at small scales emerge as dynamic and important

### 2016-08-18 3dpas_intro.tex
minor changes to Neil's addition

@@
 either for reasons of performance or for collaborative analytics.  
-Further, the variation of data production rates, data source and destination makes 
+Further, the variation of data production rates, data source, and destination makes 
 temporal variations in data properties increasingly important.  
 With an increase in the importance of distributed and temporal properties of 
-data, systems and infrastructure aspects such as resource scheduling, d
-ata placement, and transfer decisions that
+data, systems and infrastructure aspects such as resource scheduling, 
+data placement, and transfer decisions that
 were statically determinable at small scales emerge as dynamic and important

### 2016-08-18 rebuttal.tex
removing added space
Address Item 3 of Reviewer 2
minor changes to Neil's addition
changes re today's call

@@
 
-\textbf{Response:} \katznote{attention needed here}
+\textbf{Response:} We've reworked the first paragraph of Section 1 to address this.
 
@@
 
-\textbf{Response:} \katznote{attention needed here}
+\textbf{Response:}
 
-\katznote{partial response:} In addition, recognizing that the two reviewers have different styles of reading
+\katznote{attention needed here re distributed and dynamic part of comment}
+
+\katznote{Neil will add a sentence in the start of Section 2 to say how we are using distributed and how we are using dynamic, in the text that refers to Figure 1.}
+
+In addition, recognizing that the two reviewers have different styles of reading
 and different areas of expertise (as do we as authors), we have added a paragraph at the end
@@
 not believe that there is one single way of organizing it that would satisfy all readers.  We have also
-added a longer description of each section at this same place.
+added a longer description of each section at this same place.  Also, see our response to point 10 below.
 
@@
 
-\textbf{Response:} \katznote{attention needed here}\jhanote{this should also have been addressed by Omer's work.
-Omer can you confirm please?} \katznote{no, Omer's work added a better description of the paper at the end of Section 1.}
+\textbf{Response:} See above response.
 
@@
 
-  \textbf{Response:} \katznote{attention needed here}\jhanote{this should also have been addressed by Omer's work.
-Omer can you confirm please?} \katznote{no, Omer's work added a better description of the paper at the end of Section 1.}
+  \textbf{Response:} 
+The definitions are provided in the manner they are since they are derived from the applications.  We have clarified this in Section 1.2.  \katznote{Omer, add more here?}
 

### 2016-08-18 rebuttal.tex
Add files via upload

@@
   \textbf{Response:}
-The definitions are provided in the manner they are since they are derived from the applications.  We have clarified this in Section 1.2.  \katznote{Omer, add more here?}
+There are a number of possible definitions related to terms used in sections 4.1--4.3 in the computer science literature.
+Our intention in this work was to derive what these mean based on the application survey we carried out.
+The definitions are therefore provided in the manner they are since they are derived from the applications provided in Section 2.
+We have clarified this in Section 1.2 to provide greater guidance to a reader, justifying the order of the sections.
+\katznote{Omer, add more here?}
 

### 2016-08-18 rebuttal.tex
removing comment to Omer in point 10

@@
   \textbf{Response:}
-There are a number of possible definitions related to terms used in sections 4.1--4.3 in the computer science literature.
+There are a number of possible definitions related to terms used in Sections 4.1--4.3 in the computer science literature.
 Our intention in this work was to derive what these mean based on the application survey we carried out.
-The definitions are therefore provided in the manner they are since they are derived from the applications provided in Section 2.
+The definitions are therefore provided in Section 3 they are since they are derived from the applications provided in Section 2.
 We have clarified this in Section 1.2 to provide greater guidance to a reader, justifying the order of the sections.
-\katznote{Omer, add more here?}
 

### 2016-08-18 rebuttal.tex
Added definitions to address Point 4

--- 2016-08-18 rebuttal.tex
@@
 
-\textbf{Response:} We've reworked the first paragraph of Section 1 to address this.
+\textbf{Response:} We've reworded the first paragraph of Section 1 to address this.
 
@@
 
-\textbf{Response:}
+\textbf{Response:} \katznote{attention needed here}
 
-\katznote{attention needed here re distributed and dynamic part of comment}
+\nchnote{Partial response:} We have added a summary description of what we mean by distributed and dynamic at the start of Section 2 to reflect the fact that definition of the terminology arose and was refined by the discussions at the workshops, and the analysis of the application scenarios presented.
 
-\katznote{Neil will add a sentence in the start of Section 2 to say how we are using distributed and how we are using dynamic, in the text that refers to Figure 1.}
-
-In addition, recognizing that the two reviewers have different styles of reading
+\katznote{partial response:} In addition, recognizing that the two reviewers have different styles of reading
 and different areas of expertise (as do we as authors), we have added a paragraph at the end
@@
 not believe that there is one single way of organizing it that would satisfy all readers.  We have also
-added a longer description of each section at this same place.  Also, see our response to point 10 below.
+added a longer description of each section at this same place.
 
@@
 
-\textbf{Response:} See above response.
+\textbf{Response:} \katznote{attention needed here}\jhanote{this should also have been addressed by Omer's work.
+Omer can you confirm please?} \katznote{no, Omer's work added a better description of the paper at the end of Section 1.}
 
@@
 
-  \textbf{Response:} 
-The definitions are provided in the manner they are since they are derived from the applications.  We have clarified this in Section 1.2.  \katznote{Omer, add more here?}
+  \textbf{Response:} \katznote{attention needed here}\jhanote{this should also have been addressed by Omer's work.
+Omer can you confirm please?} \katznote{no, Omer's work added a better description of the paper at the end of Section 1.}
 

### 2016-08-18 3dpas_intro.tex
Added definitions to address Point 4

@@
 led us to examine the use of data in different parts of the
-applications, where we found some common patterns
+applications, 
+and informed the terminology we use in the rest of the paper.
+We have split the presentation of information on each scenario into three aspects: 
+{\em big data aspects}, concerned with the number and volume of data; 
+{\em distributed aspects}, concerned with the way data generation, storage and processing are separated; and 
+{\em dynamic aspects}, concerned with the spatio-temporal variability of the data or application.
+Doing this has enabled us to identify some common patterns
 across applications.  For example, applications that involve

### 2016-08-18 rebuttal.tex

--- 2016-08-18 rebuttal.tex
@@
 
-\katznote{Neil will add a sentence in the start of Section 2 to say how we are using distributed and how we are using dynamic, in the text that refers to Figure 1.}
+\nchnote{Partial response:} We have added a summary description of what we mean by distributed and dynamic at the start of Section 2 to reflect the fact that definition of the terminology arose and was refined by the discussions at the workshops, and the analysis of the application scenarios presented.
 
-In addition, recognizing that the two reviewers have different styles of reading
+\katznote{partial response:} In addition, recognizing that the two reviewers have different styles of reading
 and different areas of expertise (as do we as authors), we have added a paragraph at the end

### 2016-08-20 3dpas_intro.tex
minor tweek to abstract

@@
   characteristics give rise to dynamic distributed data intensive
-  applications. While ``static'' data application have received significant
+  applications. While ``static'' data applications have received significant
   attention, the characteristics, requirements, and software systems for the
-  analysis of large volumes of dynamic, distributed, and data intensive applications have
+  analysis of large volumes of dynamic, distributed data, and data-intensive applications have
   received relatively less attention. This paper surveys several representative dynamic

### 2016-08-20 3dpas_intro.tex
consistent spelling of cyber-infrastructure

@@
 \item In \S\ref{sec:infrastructure}, we subsequently describe the software/systems infrastructure used to support the distributed and dynamic characteristics of applications in \S\ref{sec:scenarios}.
-We outline support for data management, analysis, and coordination, providing examples of actual Cyberinfrastructure made use of by our surveyed applications for these purposes.
+We outline support for data management, analysis, and coordination, providing examples of actual cyberinfrastructure made use of by our surveyed applications for these purposes.
 

### 2016-08-20 3dpas_intro.tex
consistent spelling of loosely coupled

@@
 % thebookthatchangedtheworld
-Most of the applications in \S\ref{sec:scenarios} are loosely coupled,
+Most of the applications in \S\ref{sec:scenarios} are loosely-coupled,
 with the CMB (\S\ref{astroJulian}) and Fusion (\S\ref{fusionScott})
@@
 
-The {\em MapReduce} paradigm~\cite{MapReduce} and its variants~\cite{twister,hmr} are the most prominent among the bulk data-parallel frameworks which use bulk operators to express computations over collections of data in an loosely coupled manner.
+The {\em MapReduce} paradigm~\cite{MapReduce} and its variants~\cite{twister,hmr} are the most prominent among the bulk data-parallel frameworks which use bulk operators to express computations over collections of data in an loosely-coupled manner.
 Applications such as Next Generation Sequencing (NGS) (\S\ref{bioSilvia}) and Power Grids (\S\ref{power}) use MapReduce for

### 2016-08-20 3dpas_intro.tex
locally versus geographically distributed
remove reference to detailed streaming discussion that does not exist

@@
   according to a spatial or temporal attributes. This can be an opaque
-  (e.g., ID) and or a domain dependent (e.g., spatial) attribute.
+  (e.g., ID) and/or a domain dependent (e.g., spatial) attribute.
 \item \textbf{Streaming}: A form of distributed data processing in
@@
   temporally. Also, typically, the stream of data must be processed in
-  near real-time. Streaming thus also has attributes of dynamism, which is
-  discussed in the next subsection.
+  near real-time. 
+  %Streaming thus also has attributes of dynamism, which is
+  %discussed in the next subsection.
 \end{itemize}
@@
 
-\item Details of the storage system are often not known. Often data is
-  stored in a distributed way, e.g., on a distributed file system,
-  which commonly fits into the medium degree of distribution (the same
-  data center).
+\item Details of the storage system are often not known. Often, data is
+  stored in a distributed way, e.g., on a distributed file system in the same
+  data center (locally distributed).
 

### 2016-08-20 3dpas_intro.tex
consistent spelling of dataset

@@
 from multiple data stores distributed around the world, e.g., to
-compare O(1000s) of TB-scale data sets (metagenomics) with each other.
+compare O(1000s) of TB-scale datasets (metagenomics) with each other.
 Dynamic elements include the choice of executables in each instance of
@@
 application components elements of the same task, for example, sharing
-a spatial data set between components that cooperate to perform a
+a spatial dataset between components that cooperate to perform a
 single logical task~\cite{dpa_surveypaper}.
@@
   \emph{Application} dynamism
-  involves changing the processes or components applied to a data set,
+  involves changing the processes or components applied to a dataset,
   and might encapsulate changes in workflow, \textit{ad hoc} queries,
@@
 Environmental Data application (\S\ref{envJon}) manages various
-sources of data, from archived data sets to real-time feeds.
+sources of data, from archived datasets to real-time feeds.
 Depending on the spatiotemporal properties of data, computation can be
@@
 not assumed to be available \textit{en masse}, it can be used in
-situations where storing the entire data set is undesirable,
+situations where storing the entire dataset is undesirable,
 unnecessary, or impossible.
@@
 from users or an instrument. This form of ``computational steering''
-is especially useful in systems being used to explore a data set. In these cases, it is important to
+is especially useful in systems being used to explore a dataset. In these cases, it is important to
 note that the infrastructure is adaptive, rather than the scenario itself.
@@
 and ends. This represents the difficulty in differentiating between ``traditional'' applications
-and ``infrastructure'' applications. Many systems are driven by data sets collected and shared
+and ``infrastructure'' applications. Many systems are driven by datasets collected and shared
 widely, against which several different applications may be applied.
 Here the infrastructure application is the collection of components used to allow the
-data sets to be accessed and analyzed by the different applications.
+datasets to be accessed and analyzed by the different applications.
 Climate science is a good example of this approach. In this
@@
 % specialized data cyberinfrastructure (e.g., the ATLAS LHC Grid). Applications
-% such as Fusion and NGS that run on smaller data sets mainly utilize HPC
+% such as Fusion and NGS that run on smaller datasets mainly utilize HPC
 % environments.
@@
 \item What are the types of data,
-\item What is the size of the data set(s)?
+\item What is the size of the dataset(s)?
 \end{enumerate}

### 2016-08-21 3dpas_intro.tex

@@
 
-\emph{Application Dynamism: } In many scientific applications the
+\emph{Application dynamism: } In many scientific applications the
 workflow structure is well defined, but either the components used
@@
 
-\subsubsection{Understanding Distributed Dynamic Applications: }
+\subsubsection{Understanding Distributed Dynamic Applications }
 \label{sec:quantifyingdistapps}

### 2016-09-05 rebuttal.tex
processing Yogesh's comments

@@
 
-\nchnote{Partial response:} We have added a summary description of what we mean by distributed and dynamic at the start of Section 2 to reflect the fact that definition of the terminology arose and was refined by the discussions at the workshops, and the analysis of the application scenarios presented.
+\nchnote{Partial response:} We have added a summary description of wha t we mean by distributed and dynamic at the start of Section 2 to reflect the fact that definition of the terminology arose and was refined by the discussions at the workshops, and the analysis of the application scenarios presented.
 

### 2016-09-05 3dpas_intro.tex
processing Yogesh's comments

@@
 
-\keywords{dynamic, distributed, data intensive, scientific applications}
+\keywords{dynamic, distributed, data-intensive, scientific applications}
 
@@
   location, availability and properties are often time-dependent. Collectively, these
-  characteristics give rise to dynamic distributed data intensive
+  characteristics give rise to dynamic distributed data-intensive
   applications. While ``static'' data applications have received significant
@@
 %being created in a set of distributed locations, rather than in one place. 
-Data can also be transported to be processed in multiple, distributed
-locations, rather than being processed near their source, or be brought together for processing at
-a particular location. These two orthogonal dimensions lead to a number of patterns of data
-generation and processing being present in real applications,
-either for reasons of performance or for collaborative analytics.  
-Further, the variation of data production rates, data source, and destination makes 
-temporal variations in data properties increasingly important.  
-With an increase in the importance of distributed and temporal properties of 
-data, systems and infrastructure aspects such as resource scheduling, 
-data placement, and transfer decisions that
-were statically determinable at small scales emerge as dynamic and important
-considerations at large scales of data and distribution.
+This massive data generation can be naturally distributed because they are being created in a set of distributed
+locations, rather than in one place. At the same time, data can also be
+transported to be processed in multiple, distributed locations, rather than
+being processed near their source, or be brought together for centrally
+processing at a particular location. These two orthogonal dimensions lead to a
+number of patterns of data generation and processing being present in real
+applications, either for reasons of performance or for collaborative analytics.
+Further, the variation of data production rates, data source, and destination
+makes temporal variations in data properties increasingly important. With an
+increase in the importance of distributed and temporal properties of data,
+systems and infrastructure aspects such as resource scheduling, data placement,
+and transfer decisions that were statically determinable at small scales emerge
+as dynamic and important considerations at large scales of data and
+distribution.
 
@@
 \item Almost by definition, any attempt to capture the state-of-the-art of such
-  a dynamic and emerging domain is bound to be fraught with limitations and
-  issues of scoping.  A complete and rigorous survey is bound to be obsolete by
+  a fast-changing and emerging domain is bound to be fraught with limitations 
+  and issues of scoping.  A complete and rigorous survey is bound to be 
+  obsolete by
   the time it emerges; so given the rate of change, we asked what could be
@@
 \item Distribution can occur on many levels.  In this paper, we use {\em locally
-    distributed} to indicate that data/compute resides on multiple nodes within
-  one data center.  {\em Geographically distributed} describes the distribution
-  of data and/or compute across multiple data centers.  Also, our focus is
+    distributed} to indicate that data generation and storage, or computation 
+	over it, occurs on multiple nodes within one data center.  
+  {\em Geographically distributed} describes the distribution
+  of data generation, storage and/or compute across multiple data centers.  
+  Also, our focus is
   primarily on infrastructure considerations at  large scale, as searching
@@
   low-level systems features is likely to return greater yield.
+
+ \item We use the term dynamism to refer to the spatio-temporal variability of 
+ data (e.\,g.\ data rates and formats), applications (e.\,g., workflows that 
+ depend on data values) and infrastructures (e.\,g.\ availability).
 
@@
 
-Starting with the experience of the attendees of the 3DPAS workshops, we have
+Starting with the experience of the attendees of the Dynamic Distributed Data Programming Abstractions and Systems workshops (3DPAS), we have
 assembled a set of D3 ``applications'' that we describe in this section.  
@@
 % traditional is  applications are {\em traditional applications}, in that 
-a program (or set of programs) is developed by a user or project to try to
+a program (or set of programs) is developed independently by a user or project to try to
 answer a science question (or set of related questions). 
@@
 a weak consensus on the infrastructure to be used.  This allows different
-sub-applications (that we again refer to as `applications') to focus on
+sub-applications (which themselves are full-fledged ``applications'', but with dependencies on the collaborative infrastructure) to focus on
 different stages, such as generating data, or processing data.  Answering a
@@
 application, we believe a term analogous to the use of Amdahl's
-number~\cite{10.1109/ICDE.2000.839382} such as
-\begin{itemize}
-\item[] Num. of Compute Operations / Amount of Data
-  (Consumed, Generated or Transformed)
-\end{itemize}
-  could be used to characterize `static'
-data-intensive applications and systems:
-
+number~\cite{10.1109/ICDE.2000.839382}, such as $\alpha$ defined as Number\ of
+Compute\ Operation / Amount\ Data (Consumed,\ Generated\ or\ Transformed)
+% \begin{itemize}
+% \item[] Num. of Compute Operations / Amount of Data
+%   (Consumed, Generated or Transformed)
+% \end{itemize}
+could be used to characterize `static' data-intensive applications and systems.
 %\begin{itemize}
@@
 %\end{itemize}
-
 Note that this term is scale-invariant, i.e., a value of 1 could
@@
 
-Similarly, another term
-\begin{itemize}
-\item[] Amount of Data Transferred / Degree of
-  Distribution
-\end{itemize}
+Similarly, another term $\beta$ defined as Amount of Data Transferred / Degree of Distribution
+% \begin{itemize}
+% \item[] Amount of Data Transferred / Degree of
+%   Distribution
+% \end{itemize}
 could be used to quantify the
@@
 function of the amount of data transferred, where the degree of
-distribution can be defined as the number of distinct data locations:
+distribution can be defined as the number of distinct data locations.
 
@@
 ``class of applications'' as opposed to a specific application. We
-believe, nonetheless, that these measures (which we refer to as $\alpha$ and $\beta$ respectively)
-suggest a possible way to
-quantify an application's dynamism and distribution and thereby find
-possible common approaches solutions.  We plan to explore this in our
-future work, and also hope others will consider building on these
+believe, nonetheless, that these measures 
+%(which we refer to as $\alpha$ and $\beta$ respectively)
+suggest a possible way to quantify an application's dynamism and distribution 
+and thereby find possible common approaches solutions.  We plan to explore this 
+in our future work, and also hope others will consider building on these
 initial concepts.

### 2016-09-06 3dpas_intro.tex
removing alpha and beta

@@
 application, we believe a term analogous to the use of Amdahl's
-number~\cite{10.1109/ICDE.2000.839382}, such as $\alpha$ defined as Number\ of
-Compute\ Operation / Amount\ Data (Consumed,\ Generated\ or\ Transformed)
+number~\cite{10.1109/ICDE.2000.839382}, such as [Number\ of
+Compute\ Operation / Amount\ Data (Consumed,\ Generated\ or\ Transformed)]
 % \begin{itemize}
@@
 
-Similarly, another term $\beta$ defined as Amount of Data Transferred / Degree of Distribution
+Similarly, another term [Amount of Data Transferred / Degree of Distribution]
 % \begin{itemize}

### 2016-09-06 3dpas_intro.tex
small tweaks

@@
 
-Starting with the experience of the attendees of the Dynamic Distributed Data Programming Abstractions and Systems workshops (3DPAS), we have
+Starting with the experience of the attendees of the Dynamic Distributed Data Programming Abstractions and Systems (3DPAS) workshops, we have
 assembled a set of D3 ``applications'' that we describe in this section.  
@@
 however, to facilitate understanding we categorize applications into two
-categories: the first is that of {\em traditional applications}, where
+categories. The first is that of {\em traditional applications}, where
 % traditional is  applications are {\em traditional applications}, in that 
@@
 complex workflow could fall under the category of a {\em traditional
-application}.  Applications in this paper that are in this category are: NGS (next generation
+application}.  Applications in this paper that are in this category are NGS (next generation
 sequencing) Analytics (\S\ref{bioSilvia}), CMB (cosmic microwave background)
@@
 series, perhaps in different stages, that may be run by different groups that do
-not frequently interact.  In this paper, examples of this type of application includes: ATLAS/WLCG (a
+not frequently interact.  In this paper, examples of this type of application include
+ATLAS/WLCG (a
 toroidal large hadron collider apparatus/worldwide large hadron collider

### 2016-09-06 rebuttal.tex
finalizing a response

--- 2016-08-18 rebuttal.tex
@@
 
-\katznote{attention needed here re distributed and dynamic part of comment}
-
-\nchnote{Partial response:} We have added a summary description of what we mean by distributed and dynamic at the start of Section 2 to reflect the fact that definition of the terminology arose and was refined by the discussions at the workshops, and the analysis of the application scenarios presented.
-
-\katznote{partial response:} In addition, recognizing that the two reviewers have different styles of reading
+We have added a summary description of wha t we mean by distributed and dynamic at the start of Section 2 to reflect the fact that definition of the terminology arose and was refined by the discussions at the workshops, and the analysis of the application scenarios presented.
+In addition, recognizing that the two reviewers have different styles of reading
 and different areas of expertise (as do we as authors), we have added a paragraph at the end
